### 2017-01-18 writing/manuscript_draft.Rmd
added initial version of outline and manuscript

---
title: "Your time series is (probably) too short"
author: "Easton R. White"
date: "January 18, 2017"
output: html_document
---

## Abstract
Longterm census data is necessary to better understand population dynamics, assess threat statuses, and make future predictions for numerous populations. However, there has been very little work done to assess how long a time series needs to be to answer a particular question. Several databases have emerged to compile long term time series. Here we empirically evaluate the minimum time series length required to answer several questions of biological interest. We address this question using time series data on XX populations of XX species including We find that most time series In the future, we advise

## Introduction 
Observational studies, and censusing populations, has become a cornerstone of modern ecology and conservation biology. Longterm data is necessary to understand population dynamics and to assess extinction risk of species. Even many time series may now be considered "longterm", most are still short (i.e. less than X years). I could have statistic on how long is considered "longterm" in papers. Time series are typically short for a variety of reasons. They are often coupled with an experiment, which may only last a couple of years. In addition, short funding cycles make it difficult to examine populations over longer periods of time. There are of course notable exceptions including the Isle Royale study, Iowa corn study, cod, lynx-hair, and many others.

(lack of statistical tools to evaluate power of time series) - can simulate or subsample data. In order to properly estimate mimumn time seris you need to know     IUt will also depend on your question

In determining the number of samples required for a particular experiment X quantities are intricately linked: statistical power, effect size, alpha, sample size,  Formally, statistical power is the one minus the probability of a type II error, or false negative. In other words, it is a measure of how well you could detect a trend given it is statistically significant. Prior to an experiment, one could set appropriate levels of power, alpha, and the effect size to estimate the sample size required for the experiment. This approach, however, is not as simple for a time series as the data points are clearly non-independent. Therefore, two approaches can be used to estimate the number of time points required to detect a trend. The first approach is a simulation approach. The second approach

Little work has actually gone into evaluating how long a time series needs to be to address questions of interest. The minimum length of a time series is similar to the question of the minimal number of samples required for any experiment. However, samples in a time series cannot be considered independent (citation). The mimumum length of a time series will depend on a number of factors. First, the question will obviously drive a lot of the length reuirments. If you are interested in how a population responds to El Nino, this will need a much longer time series compared to if you wanted to know Second, properties of the species of interest, and their population will affect the minimum time series required, including the presence of internally driven cycles, variance in population size, generation time, 
The IUCN provides minumim time series criteria to assess species threat status. They require a 10 year, or 3 genration, whichever is shorter To determine status.

How sensitivity are our findings to changes in the length of our time series?

Recently, several databases (e.g.) of animal population censuses have emerged that facilate the comparison of many populations.

In this paper, we evaluate the minimal time series length required to answer X questions. 1) How sensitivive are IUCN asssessments to time series lengths? 2) What minimum time series length is required to measure the entire time series trend? 3) How long does a time series need to be to measure the variance in population size (rare event questions)? We demonstrate our approach with X case studies and then show results for X species from the Blah Blah Database.

## Methods

### Data source
We use a series of XXXX time series compiled in @Keith2015. The database includes information on X vertebrate species with a focus on mammals, birds, and fish. We filtered out short time series (less than XX) and were left with XX time series with an average length of XX. The database also includes information on generation length, census type,

For a subset of populations in the database, we were able to connect a database on biological characteristic
Analyses
Our objective is to determine the miniuum time series length (T_min) required to answer various questions. As we describe below, the minimum time series length calculation is particular to each question. We assume that the time series in our database are long enough to include all necessary information (e.g. variability) about the population. In other words, they are a representive sample. We then take random contiguous subsamples of each time series and compare each subset to the measurements on the entire time series. For example, for Isla Royale wolves there are XX years of data available. We can then take contiguous subsets of those XX years. This results in XX subsets of length 2 years, XX subsets of length 3 years, and so forth until one subset of length XX years. We then examine how various measurements on the population change as we increase the subset length.

Question 1: What length of time series is needed to detect a significant trend in abundance?
Here we subsample each time series to determine the T_min required to calculate the overall time series trend. We first take all possible contiguous subsamples of the overall time series. For example, a time series of 40 years would have 39 possible contiguous subsamples of length 2, 38 possible contiguous subsamples of length 3, and so forth. For each subsample we run linear regression to determine the slope parameter and if the slope parameter estimated is significant. Then we can determine the total number of contiguous subsamples of a certain length that detect a significant trend given it is present---this is an estimate of statistical power.

Figure 1 - maybe make one figure for questions 1,2, and 3

We define the minimum time required as the point when the curve begins to saturate above some threshold. For instance, in figure 1 if we want power of at least 0.8, then it would take XX years to be sure that power will be above that threshold. We then define T_min as that specific number of year.

Question 2: How sensitive are IUCN assessments to time series trends?
The IUCN sets threat statuses for a variety of species. Importaly, they have a threshold of 10 years or 3 generation. Therefore, we subsampled all possible 10-year, or greater, time series for each population. For example, species X has been sampled for 40 years. This would lead to XX possible 10-year time series. We then determined what percentage of these 10 year, or greater, time series would lead to the wrong conculsions compared to the full time series.

Question 3: How long does a population need to be sampled to capture the variability in population size?

Populations vary from year-to-year for several reasons including

## Results

```{r,echo=F,eval=F,include=F}
# Pull in analysis of average min time needed **theoretical models** (save analysis as Rdata file or something)

```

```{r,echo=F,eval=F,include=F}
# Pull in analysis of average min time needed for **empirical species** (save analysis as Rdata file or something)

```

- Intro paragraph on what we generally found (average T_min and range), many species needed the full time series

Across all the populations we examined, we found an average minimum time series length required ($T_{min}$) of `r 2+3 #calculation here`.

- What does a theoretically model predict (does min time scale with trend, power, variability, autocorrelation, something else) - definitely depends on what you set for alpha and beta
- Case studie(s)
- For all species
    - Average min time needed
    - Correlates well with trend, variability, and autocorrelation
    - Surprisingly does not correlate well with generation time or any other biological variables (need to do proper regression though)

```{r,echo=F,eval=F,include=F}
# Create plots from min time needed analyses
```

## Potential figures
- Examples of how determining minimum time series works (probably in methods)
- What min time should scale with theoretically (trend, power, variability, etc) - another multi panel figure
- Mean min time by class or system
- Separate results for species, body size, location, biology. - multi panel figure

## Discussion
- Findings
- Extensions to other systems. Useful to use similar species
- Other questions to ask
- Comparisons to past work
- Future recommendations to others

### 2017-01-18 writing/outline.Rmd
added initial version of outline and manuscript

---
title: "Your time series is (probably) too short"
author: "Easton R. White"
date: "January 18, 2017"
output: pdf_document
---

```{r,echo=F,include=F,eval=F}
# General manuscript notes here
```

# Outline

## Titles: "Your time series is (probably) too short" "Minimum time series calculations for population inferences"
## Length: about 150 words for the abstract, 2500 words, not including a short abstract, not more than 20 references, and a total of 3 figures, tables, and/or panels

## Outline

## Abstract (150 words)

## Introduction to problem (600 words)

- ~~Have longer data sets now, Short funding cycles~~
- Little work has empirically investigated how long time series need to be
- Power series
- Empirically answer this question
- What determines minimum time series
    - Deterministic skeleton (dynamics), stochasticity, question being asked, biology of organism, management and conservation practices, power and alpha
- Past work that has looked at this question
- Population dynamics database
- What we plan to do

## Methods (750 words)
- How to subsample
- Theoretical results (maybe in the appendix mostly)
- Min time series calculation for each question
- Database I will be using

## Results (500 words)
- Simulated data with cycles and other types of attractors where we know what time series length you need to sample
- 1 good case study example (maybe this is in methods?) 
   - Isle Royale
- Results from X number of species

## Discussion and generality (800 words)

## Conclusions

## Figures/tables

http://www.frontiersinecology.org/fron/instructions-to-authors/research-communications/

### 2017-01-19 scripts/LPI_data_analysis.R
first commit of script files

@@ -1,0 +1,36 @@
+# Created by Easton White
+# Created on 17-Jan-2017
+# Last edited on 18-Jan-2017
+
+# This code reads in data from the () and cleans it up (mostly making it into a long table format)
+
+library("tidyr")
+library("dplyr")
+
+LPI = read.table('data/LPI_LPR2016data_public_EW_edited.csv',sep=',',header=T)
+
+
+LPI_long <- LPI %>%
+  gather(year, pop_value, starts_with('X'))
+
+LPI_long$year=as.numeric(sub('X','',LPI_long$year))
+
+#head(LPI_long[sort(LPI_long$ID),],50)
+
+newdata=arrange(LPI_long,ID,year)
+
+testing=newdata[,c(1,2,3,19,20)]
+testing$pop_value[testing$pop_value=='NULL']=as.numeric(NA)
+#removed=na.omit(testing)
+
+long_term_series_names=names(sort(table(removed$ID),decreasing=T)[sort(table(removed$ID),decreasing=T)>40])
+
+y=subset(removed,removed$ID %in% long_term_series_names)
+
+
+for (species_code in long_term_series_names[1]){
+  pop = subset(removed,removed$ID==species_code)
+  
+  
+}
+

### 2017-01-19 scripts/calculate_fraction_sign.R

@@ -1,0 +1,51 @@
+
+# Created by Easton White
+# Created on 17-Jan-2017
+# Last edited on 18-Jan-2017
+
+# This code calculates the fraction of linear regressions for subsamples of a time series which are of the correct long term sign (pos vs neg)
+
+require(plyr)
+#repeat slope calculations but only look for positive or negative instead of actual value
+
+#pull in neccessary functions
+source('calculate_slope.R')
+source('create_subsamples.R')
+
+calculate_fraction_correct_sign = function(pop){
+  subset_times=create_subsamples(pop)
+  fraction_correct_sign=matrix(0,nrow=(length(pop)-1),1)
+  for (n in 1:(length(pop)-1)){
+    fraction_correct_sign[n,1]=sum(sign(apply(na.omit(subset_times[[n]]),1,calculate_slope))==sign(calculate_slope(pop))) / (length(pop) - n)
+  }
+  return(fraction_correct_sign)
+}  
+
+
+
+min_time_needed = function(pop, threshold){
+  fraction_correct_sign = calculate_fraction_correct_sign(pop)
+  min_value = tail(which(fraction_correct_sign < threshold),1)+1
+  
+  if (length(min_value)==0){
+    return(2)
+  }else if (min_value ==length(pop)){
+    return(NA)
+  }else{
+    return(min_value)
+  }
+}
+
+
+##########
+# run a loop to find the min time series length rrequrired for each population
+mimimum_time_required = matrix(0,nrow=nrow(pop_info),ncol=1)
+###run loop to make plot
+for (j in names(table(long_dat$ID))){
+  
+  pop=subset(long_dat,long_dat$ID==j)$popvalue
+  
+  mimimum_time_required[which(j==pop_info$ID),1] = min_time_needed(pop,0.8)
+  
+  print(paste(j,":",mimimum_time_required[which(j==pop_info$ID),1]))
+}

### 2017-01-19 scripts/calculate_power_metric.R

@@ -1,0 +1,88 @@
+# Created by Easton White
+# Created on 17-Jan-2017
+# Last edited on 18-Jan-2017
+
+
+# This script is designed to calculate the min time required to have 0.X power
+
+# Calculate power for series of subsamples (what fraction of subsamples have trened that is significant - dictated by alpha)
+require(plyr)
+
+#pull in neccessary functions
+source('scripts/calculate_slope.R')
+source('scripts/create_subsamples.R')
+
+calculate_power_for_subsamples = function(pop,alpha){
+  subset_times=create_subsamples(pop)
+  power=matrix(0,nrow=(length(pop)-1),1)
+  for (n in 1:(length(pop)-1)){
+    power[n,1]=sum(apply(na.omit(subset_times[[n]]),1,calculate_p_value)<alpha)/(length(pop) - n)
+  }
+  return(power)
+}  
+
+
+
+# give population time series, alpha, and a threshold for power (usually 0.8), calculate the minimum time needed to guarentee time series reaches required levels of alpha and power
+min_time_needed = function(pop, alpha,threshold){
+    alpha=alpha
+    power = calculate_power_for_subsamples(pop,alpha)
+  min_value = tail(which(abs((power - as.numeric(tail(power,1)))/as.numeric(tail(power,1)))>(1-threshold)),1)+1
+  
+  if (length(min_value)==0){
+    return(2)
+  }else if (min_value ==length(pop)){
+    return(-999)
+  }else{
+    return(min_value)
+  }
+}
+
+
+
+
+
+## OLD CODE BELOW ##
+
+##########
+# run a loop to find the min time series length requrired for each population
+# mimimum_time_required = matrix(0,nrow=nrow(pop_info),ncol=1)
+# ###run loop to make plot
+# for (j in names(table(long_dat$ID))){
+#   
+#   pop=subset(long_dat,long_dat$ID==j)$popvalue[1:40]/max(subset(long_dat,long_dat$ID==j)$popvalue[1:40])
+#   mimimum_time_required[which(j==pop_info$ID),1] = min_time_needed(pop,0.05,0.8)
+# 
+#   print(paste(which(j==pop_info$ID),j,":",mimimum_time_required[which(j==pop_info$ID),1]))
+# }
+
+
+
[…]

### 2017-01-19 scripts/calculate_slope.R

@@ -1,0 +1,14 @@
+# Created by Easton White
+# Created on 17-Jan-2017
+# Last edited on 18-Jan-2017
+
+# Here are two functions. The first extracts the slope coefficient for linear regression while the second calculates the p_value
+
+calculate_slope= function(x){
+  return(as.numeric(coef(lm(c(x)~c(1:length(x))))[2]))
+}
+
+
+calculate_p_value = function(x){
+  return(summary(lm(c(x)~c(1:length(x))))$coefficients[2,4])
+}

### 2017-01-19 scripts/calculate_slope_metric.R

@@ -1,0 +1,53 @@
+
+# Created by Easton White
+# Created on 17-Jan-2017
+# Last edited on 18-Jan-2017
+
+
+# Calculate time required to be within XX% of the overall slope trend
+require(plyr)
+
+#pull in neccessary functions
+source('calculate_slope.R')
+source('create_subsamples.R')
+
+calculate_average_slopes = function(pop){
+  subset_times=create_subsamples(pop)
+  average_slopes=matrix(0,nrow=(length(pop)-1),1)
+  for (n in 1:(length(pop)-1)){
+    average_slopes[n,1]=mean(apply(na.omit(subset_times[[n]]),1,calculate_slope))
+  }
+  return(average_slopes)
+}  
+
+
+
+
+
+min_time_needed = function(pop, threshold){
+  average_slopes = calculate_average_slopes(pop)
+  min_value = tail(which(abs((average_slopes - as.numeric(tail(average_slopes,1)))/as.numeric(tail(average_slopes,1)))>threshold),1)+1
+  
+  if (length(min_value)==0){
+    return(2)
+  }else if (min_value ==length(pop)){
+    return(-999)
+  }else{
+    return(min_value)
+  }
+}
+
+
+
+##########
+# run a loop to find the min time series length rrequrired for each population
+mimimum_time_required = matrix(0,nrow=nrow(pop_info),ncol=1)
+###run loop to make plot
+for (j in names(table(long_dat$ID))){
+  
+  pop=subset(long_dat,long_dat$ID==j)$popvalue[1:40]/max(subset(long_dat,long_dat$ID==j)$popvalue[1:40])
+
+  mimimum_time_required[which(j==pop_info$ID),1] = min_time_needed(pop,0.05)
+
+  print(paste(which(j==pop_info$ID),j,":",mimimum_time_required[which(j==pop_info$ID),1]))
+}

### 2017-01-19 scripts/combine_databases.R

@@ -1,0 +1,51 @@
+# Created by Easton White
+# Created on 17-Jan-2017
+# Last edited on 18-Jan-2017
+
+
+# This is a script to try and combine multiple population series time together to use biological information from Amniote Database in my other analyses
+
+# Population time series file
+load('For_easton.Rdata')
+dat= dat
+
+# File for biological characteristics of amniotes
+amniote = read.table('Amniote_Database_Aug_2015_erw.csv',sep=',',header=T,na.strings = -999)
+amniote$Binomial = paste(amniote$genus,amniote$species,sep=' ')
+
+# Create bio_info table to store informatino where a species is present in both databases
+bio_info = as.data.frame(matrix(NA,nrow=nrow(dat), ncol=ncol(amniote[,7:35])))
+colnames(bio_info)=names(amniote[,7:35])
+
+dat = cbind(dat,bio_info)
+dat$match =0
+
+for (j in 1:nrow(dat)){
+  if (dat$Binomial[j] %in% amniote$Binomial){
+    dat[j,20:48] = amniote[which(amniote$Binomial %in% dat$Binomial[j]),7:35]
+    dat[j,49]=1
+  }
+    if (j %in% seq(1000,80000,by=5000)){print(j)} #just a counter
+}
+
+# remove coluoms with lots of NA (egg_width_mm forward) #sort(colSums(is.na(dat)),decreasing=T)
+many_NAs=names(sort(colSums(is.na(dat)),decreasing=T))[1:23]
+new_dat = dat[,-which(names(dat) %in% many_NAs)]
+new_dat = new_dat[,-which(names(new_dat) %in% c('year_ass'))]
+
+test=na.omit(new_dat)
+length(table(test$ID)) #number of time series
+
+# Which biological variables are most available for long time series?
+long_time_names=names(table(test$ID))[as.numeric(table(test$ID))>25]
+long_dat = subset(test,test$ID %in% long_time_names)
+length(table(long_dat$ID) )
+
+
+pop_info= long_dat[1,]
+for (j in 1:length(table(long_dat$ID))){
+  index=as.numeric(which(long_dat$ID==long_time_names[j])[1])
+  pop_info[j,] = long_dat[index,]
+}
+
+# All this results in 237 Aves time series and 3 mammal time series

### 2017-01-19 scripts/create-pop-info-table.R

@@ -1,0 +1,61 @@
+# Created by Easton White
+# Created on 17-Jan-2017
+# Last edited on 18-Jan-2017
+
+
+#This code loads in data and then creates a table called pop_info which includes population information on each unique time series
+
+# chuck for loading in data, removing NAs, looking at only long time series
+# also calculates a pop_info table which includes basic information on each population that I later analyze
+
+# Pull in data
+load('data/For_easton.Rdata')
+
+dat=dat[,c(1:7,11)] #
+dat=na.omit(dat)
+
+#this subsets data to find long time series (e.g. >40 years)
+dat=subset(dat,dat$class != 'Aves')
+long_time_names=names(table(dat$ID))[as.numeric(table(dat$ID))>30]
+long_dat = subset(dat,dat$ID %in% long_time_names)
+length(table(long_dat$ID) )
+
+
+#this creates pop_info table for each analyzed pop time series
+pop_info= long_dat[1,]
+  for (j in 1:length(table(long_dat$ID))){
+    index=as.numeric(which(long_dat$ID==long_time_names[j])[1])
+    pop_info[j,] = long_dat[index,]
+  }
+
+#add variables to pop_info table
+pop_info$series_length = as.numeric(table(long_dat$ID))
+pop_info$variance=NA
+pop_info$overall_trend=NA
+pop_info$autocorrelation= NA
+pop_info$trend_p_value=NA
+pop_info$min_time_for_power=NA
+  for (j in 1:nrow(pop_info)){
+    pop= subset(long_dat,long_dat$ID==pop_info$ID[j])[1:30,]
+    pop$popvalue=pop$popvalue/max(pop$popvalue)
+    pop_info$variance[j]= var(pop$popvalue)
+    pop_info$overall_trend[j] = calculate_slope(pop$popvalue)
+    pop_info$autocorrelation[j] = acf(pop$popvalue,plot = F)$acf[2]
+    pop_info$trend_p_value[j]=calculate_p_value(pop$popvalue)
+    
+    source('scripts/calculate_power_metric.R')
+    pop_info$min_time_for_power[j] = min_time_needed(pop$popvalue,0.05,0.8)
+    print(paste(j,':',pop_info$min_time_for_power[j],sep=' '))
+  }
+
+#IUCN calculation
+pop_info$IUCN = 3*pop_info$gen_len
+pop_info$IUCN[pop_info$IUCN<10]=10
+
+
+
[…]

### 2017-01-19 scripts/create_subsamples.R

@@ -1,0 +1,14 @@
+# Created by Easton White
+# Created on 17-Jan-2017
+# Last edited on 18-Jan-2017
+
+# this is a function to create a list of all possible sub samples of a time series. For example, a time series of length ten could have 9 possible 2 year continuous subsamples, 8 possible, 3 year continuous subsamples, and so forth until 1 possible 10 year subsample
+
+require(plyr)
+create_subsamples = function(pop){
+  subsamples=list()
+  for (n in 1:(length(pop)-1)){
+    subsamples[[n]] = ldply(1:length(pop), function(x){pop[x:(x+n)]})
+  }
+  return(subsamples)
+}  

### 2017-01-19 scripts/data-interpolation.R

@@ -1,0 +1,37 @@
+
+# Created by Easton White
+# Created on 17-Jan-2017
+# Last edited on 18-Jan-2017
+
+
+# this code imputes data of a time series between the first and last point for which there is data. Currenly the code runs a loop for the LPI data
+
+
+library('imputeTS')
+# Example of package
+#y=na.interpolation(as.numeric(pop$pop_value), option ="spline")
+#plotNA.imputations(as.numeric(pop$pop_value),y)
+#x <- ts(c(NA,NA,NA,2,3,4,5,6,NA,7,8))
+#y=na.interpolation(x)
+
+
+# For LPI data (This interpolates between all point that have values regardless of number of NAs)
+
+for (index in names(table(testing$ID))[2598:length(names(table(testing$ID)))]){
+  aaa=subset(testing,testing$ID==index)
+  
+  if (sum(is.na(aaa$pop_value))!=length(aaa$pop_value)){
+    start=head(which(is.na(aaa$pop_value)==F),1)
+    end=tail(which(is.na(aaa$pop_value)==F),1)
+  
+    pop=as.numeric(aaa$pop_value[start:end])
+    #na.interpolation(pop,option='spline')
+  
+    aaa$pop_value[start:end]=na.interpolation(pop,option='spline')
+  
+  
+  testing[which(testing$ID==index),]=aaa
+}
+
+print(index)
+}

### 2017-01-19 scripts/statistical-analyses.R

@@ -1,0 +1,25 @@
+# Created by Easton White
+# Created on 17-Jan-2017
+# Last edited on 18-Jan-2017
+
+#Statistical analyses of minimum time needed and what it may correlate with
+# Should probably do poisson or negative binomial with random effects (species or family)
+
+
+# Pull in final data set here to be analyzed
+
+# Run simple model
+with(trial,summary(glm(min_time_for_power~gen_len+habitat + exploitation + invasive + mass + trophic,family = 'poisson')))
+
+# pull in packages that I had used previously with Cocos project
+require(glmmADMB)
+require(MuMIn)
+
+CocosData$SiteCode <- as.factor(CocosData$SiteCode)
+CocosData$CurrentCode <- as.factor(CocosData$CurrentCode)
+pop_info$Binomial <- as.factor(pop_info$Binomial)
+
+model <- glmmadmb(formula=min_time_for_power~gen_len+habitat + exploitation + invasive + mass ,data=pop_info,family="nbinom",zeroInflation=FALSE)
+
+
+model_ranking <- dredge(model,rank="AIC")

### 2017-01-19 scripts/theoretical-models/calculate_power.R

@@ -1,0 +1,21 @@
+# Created by Easton White
+# Created on 17-Jan-2017
+# Last edited on 18-Jan-2017
+
+# This function calculates statistical power by simulating a number of populations and determining what fraction of them produce significant slope coefficients from linear regression given the slope coefficient is truely different from zero. It takes population parameters as imputs to simulate the population for a set number of trials
+
+calculate_power = function(r,phi,sigma,trials,max.time){
+  
+  source('scripts/calculate_slope.R') # calls another function which calculates the slope and p_value of a linear regression on a time series
+  # Parameter values
+  trials   = trials
+  max.time = max.time
+  r        = r
+  phi      = phi
+  sigma    = sigma
+  
+  y=replicate(n=trials,simulate_pop(r,phi,sigma,max.time)) # simulate a number of populations with same parameter values
+  
+  power=sum(apply(y,2,calculate_p_value)<0.05)/trials # evaluates slope coefficients compared to 0.05 (this can be modified)
+  return(power)
+}

### 2017-01-19 scripts/theoretical-models/simulate_pop.R

@@ -1,0 +1,14 @@
+
+# Created by Easton White
+# Created on 17-Jan-2017
+# Last edited on 18-Jan-2017
+
+# This function simulates a population over time given set of input parameters and max time to be run. Assumes initial population of 1000
+
+simulate_pop = function(r,phi,sigma,max.time){
+  pop = matrix(1000,nrow=1,ncol=max.time)
+    for (t in 1:(max.time-1)){
+      pop[t + 1] = pop[t] + r + phi*rnorm(n = 1,mean=0,sd = sigma) # this is the simple population model, other models can also be used here
+    }
+  return(c(pop)) #return vector of population sizes
+}

### 2017-01-19 scripts/theoretical-models/theoretical-approach.R

@@ -1,0 +1,22 @@
+
+# Created by Easton White
+# Created on 17-Jan-2017
+# Last edited on 18-Jan-2017
+
+
+# This code calculates the minimum time needed to achieve certain alpha and power for population model with various imput parameter. The end of this script evaluates min time needed for different parameter values (uses lapply to evaluate a number of values)
+
+
+# simple model: pop[t+1] = pop[t] + beta + phi*rnorm(0,sigma)
+
+
+min_time_needed= function(phi){
+  years_to_sample=seq(10,50,by=1)
+  y=lapply(X=years_to_sample,FUN = calculate_power,phi = phi,r=0.5,sigma = 2,trials=1000)
+  min_time_needed = years_to_sample[tail(which(y<0.8),1)] + 1
+  return(min_time_needed)
+}
+
+
+r_values = seq(0.1,0.9,by=0.1)
+zed=lapply(X=r_values,FUN = min_time_needed)

### 2017-04-18 scripts/theoretical-models/calculate_power.R
made changes to theoretical model approach

@@ -16,6 +16,9 @@ calculate_power = function(r,phi,sigma,trials,max.time){
   
   y=replicate(n=trials,simulate_pop(r,phi,sigma,max.time)) # simulate a number of populations with same parameter values
   
-  power=sum(apply(y,2,calculate_p_value)<0.05)/trials # evaluates slope coefficients compared to 0.05 (this can be modified)
+  power=sum((apply(y,2,calculate_p_value)<0.05) & sign(apply(y,2,calculate_slope))==sign(r))/trials # evaluates slope coefficients compared to 0.05 (this can be modified)
+  # Alternative definition of power which does not require linear regression to have same sign of "true" trend
+  # power=sum((apply(y,2,calculate_p_value)<0.05))/trials 
+  
   return(power)
 }

### 2017-04-18 scripts/theoretical-models/simple_plots.R

@@ -1,0 +1,23 @@
+
+# Created by Easton White
+# Created on 3-Feb-2017
+# Last edited on 3-Feb-2017
+
+# Plots for theoretical output
+
+
+#plot for r
+par(oma=c(0,0,0,0),mar=c(5,5,0.5,0.5))
+plot(r_values,zed,ylab=' ',xlab=' ',type='l',las=1,cex.axis=1.2,lwd=2)
+mtext(text = 'rate of increase (r)',side = 1,cex=1.2,line=3)
+mtext(text = 'minimum time required',side = 2,cex=1.2,line=3)
+
+# plot for p
+plot(1-phi_values,phi_zed,ylab=' ',xlab=' ',type='l',las=1,cex.axis=1.2,lwd=2)
+mtext(text = 'autocorrelation (1 - phi)',side = 1,cex=1.2,line=3)
+mtext(text = 'minimum time required',side = 2,cex=1.2,line=3)
+
+# plot for variance
+plot(sigma_values,zed,ylab=' ',xlab=' ',type='l',las=1,cex.axis=1.2,lwd=2)
+mtext(text = 'variability (sigma)',side = 1,cex=1.2,line=3)
+mtext(text = 'minimum time required',side = 2,cex=1.2,line=3)

### 2017-04-18 scripts/theoretical-models/simulate_pop.R

@@ -5,10 +5,26 @@
 
 # This function simulates a population over time given set of input parameters and max time to be run. Assumes initial population of 1000
 
+# For presentation, I could leave out autocorrelation results
+
 simulate_pop = function(r,phi,sigma,max.time){
+  set.seed(12345)
   pop = matrix(1000,nrow=1,ncol=max.time)
     for (t in 1:(max.time-1)){
-      pop[t + 1] = pop[t] + r + phi*rnorm(n = 1,mean=0,sd = sigma) # this is the simple population model, other models can also be used here
-    }
+      #pop[t + 1] = pop[t] + rnorm(n = 1,mean=r*phi,sd = sigma)
+      #pop[t + 1] = pop[t] + r + phi*rnorm(n = 1,mean=0,sd = sigma) # this is the simple population model, other models can also be used here
+      # if (pop[t]>=0){
+         #pop[t + 1] = pop[t]*rnorm(1,phi*r - ((sigma^2)/2) , sigma) 
+      # }else{
+      #   pop[t+1]=0# From Keith et al 2015
+      # }
+      pop[t + 1] = pop[t] + r + rnorm(n = 1,mean=0,sd = sigma)
+    }    
   return(c(pop)) #return vector of population sizes
 }
+
+
+#y=simulate_pop(4,-0.2,.2,100)
+##par(mfrow=c(1,2))
+#plot(y[1:99],y[2:100])
+#plot(y)

### 2017-04-18 scripts/theoretical-models/theoretical-approach.R

@@ -10,13 +10,51 @@
 # simple model: pop[t+1] = pop[t] + beta + phi*rnorm(0,sigma)
 
 
-min_time_needed= function(phi){
-  years_to_sample=seq(10,50,by=1)
-  y=lapply(X=years_to_sample,FUN = calculate_power,phi = phi,r=0.5,sigma = 2,trials=1000)
+min_time_needed= function(sigma){
+  years_to_sample=seq(2,30,by=1)
+  y=lapply(X=years_to_sample,FUN = calculate_power,phi =0.5,r=0.5,sigma = sigma,trials=100)
   min_time_needed = years_to_sample[tail(which(y<0.8),1)] + 1
   return(min_time_needed)
 }
 
 
-r_values = seq(0.1,0.9,by=0.1)
-zed=lapply(X=r_values,FUN = min_time_needed)
+r_values = seq(0.1,1.5,by=0.1)
+phi_values = seq(0.1,0.9,by=0.1)
+sigma_values=seq(1,12,by=0.5)
+zed=lapply(X=sigma_values,FUN = min_time_needed)
+sigma_zed=zed
+
+##########################################################
+# Run models from r values and sigma values to make nice plot
+parameter_combinations = expand.grid(r_values,sigma_values)
+parameter_combinations$min_time_required = NA
+names(parameter_combinations)= c('r','sigma','min_time_required')
+
+for (i in 1:nrow(parameter_combinations)){
+  years_to_sample=seq(2,25,by=1)
+  y=lapply(X=years_to_sample,FUN = calculate_power,phi =0.5,r=parameter_combinations$r[i],sigma = parameter_combinations$sigma[i],trials=150 )
+  if (length(tail(which(y<0.8),1))>0){
+  parameter_combinations$min_time_required[i] = years_to_sample[tail(which(y<0.8),1)] + 1
+  }
+  print(i)
+}
+
+
+require(ggplot2)
+ggplot(parameter_combinations, aes(x = r, y = sigma, z = min_time_required, fill=min_time_required)) +
+  geom_tile() + 
+  scale_fill_distiller(palette="Spectral", na.value="white") +
+  theme_bw()
+#geom_contour(color = "white", alpha = 0.5) + 
+
+
+
+# Simple plot
+par(mfrow=c(1,2),oma=c(3,0.3,1,1),mar=c(2,6,0,0))
[…]

### 2017-04-18 scripts/IUCN_analysis.R
made changes to several scripts

@@ -1,0 +1,48 @@
+
+
+#IUCN analysis
+
+
+#extract intercept
+calculate_intercept = function(pop){
+  return(as.numeric(coef(lm(pop~c(1:length(pop))))[1]))
+}
+
+
+#IUCN_decline = 0.7*pop_info_declining$intercept/pop_info_declining$IUCN
+
+
+for (j in 1:nrow(pop_info)){
+  pop= subset(long_dat,long_dat$ID==pop_info$ID[j])
+  pop$popvalue=pop$popvalue/max(pop$popvalue)
+  pop_info$intercept[j]=calculate_intercept(pop$popvalue)
+  pop_info$IUCN_decline[j]=pop$popvalue[pop_info$IUCN[j]]/pop_info$intercept[j]
+}
+
+
+pop_info_declining = subset(pop_info,pop_info$overall_trend<0)
+sum(pop_info_declining$IUCN_decline<0.7,na.rm=T)
+# a lot of NAs produced
+
+new=subset(pop_info_declining,pop_info_declining$IUCN_decline<0.7)
+sum(new$min_time_for_power>new$IUCN)/nrow(new)
+
+
+###############
+#IUCN analyses
+pop_info=pop_info[pop_info$trend_p_value<0.05 & sign(pop_info$overall_trend)==-1,]
+long_dat=subset(long_dat,long_dat$ID %in% pop_info$ID)
+long_time_names=subset(long_time_names,long_time_names %in% pop_info$ID)
+
+
+pop=subset(long_dat,long_dat$ID=='10027')
+pop=pop$popvalue
+summary(lm(pop~c(1:length(pop))))
+
+percent_change = function(pop){
+  percent_change=(tail(pop,1) - head(pop,1) )/ head(pop,1)
+  return(percent_change)
+}
+
+
+

### 2017-04-18 scripts/LPI_data_analysis.R

@@ -19,11 +19,11 @@ LPI_long$year=as.numeric(sub('X','',LPI_long$year))
 
 newdata=arrange(LPI_long,ID,year)
 
-testing=newdata[,c(1,2,3,19,20)]
-testing$pop_value[testing$pop_value=='NULL']=as.numeric(NA)
+testing=newdata[,c(1,2,3,10,11,19,20)]
+testing[testing=='NULL']=NA
 #removed=na.omit(testing)
 
-long_term_series_names=names(sort(table(removed$ID),decreasing=T)[sort(table(removed$ID),decreasing=T)>40])
+long_term_series_names=names(sort(table(removed$ID),decreasing=T)[sort(table(removed$ID),decreasing=T)>35])
 
 y=subset(removed,removed$ID %in% long_term_series_names)
 
@@ -33,4 +33,15 @@ for (species_code in long_term_series_names[1]){
   
   
 }
+
+
+
+# Plots to make
+load('example_min_time_LPI_data.Rdata')
+par(mfrow=c(1,1))
+plot(jitter(pop_info$Latitude,amount=2),pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),xlim=c(0,80),las=1,ylab='',xlab='',cex.axis=1.2)
+mtext(text = 'minimum time required',side = 2,line = 3,cex = 1.4)
+mtext(text = 'latitude',side = 1,line = 3,cex = 1.4)
+
+plot(jitter(pop_info$Longitude,amount=2),pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5))
 

### 2017-04-18 scripts/calculate_geometric_growth.R

@@ -1,0 +1,47 @@
+
+
+
+# calculate geometric growth rate
+
+calculate_geometric_growth = function(pop){
+  
+  return((prod(pop[2:length(pop)]/pop[1:(length(pop)-1)]))^(1/(length(pop)-1)))
+  
+}
+
+
+#this gets all plot. Each row output is the slope calculated for each sub sample (e.g. row 1 has X number of slopes for all the subsamples of length X)
+calculate_average_geometric_growth = function(pop){
+  subset_times=create_subsamples(pop)
+  average_slopes=matrix(NA,nrow=(length(pop)-1),ncol=(length(pop)-1))
+  for (n in 1:(length(pop)-1)){
+    average_slopes[n,1:(length(pop)-n)]=apply(na.omit(subset_times[[n]]),1,calculate_geometric_growth)
+  }
+  return(average_slopes)
+  
+}  
+
+
+pop=0.2*(1:100) +100 + rnorm(100,0,5)
+# zed=as.data.frame(calculate_average_geometric_growth(pop))
+# 
+# matplot(2:100,zed,ylim=range(zed,na.rm=T),pch=16,col=rgb(0.5,0.5,0.5,0.5))
+# 
+# plot(rep(2,times=99),zed[2,],ylim=range(zed,na.rm=T),xlim=c(1,100),pch=16,col=rgb(0.5,0.5,0.5,0.5))
+# for (j in 2:99){
+#   
+#   plot(rep(j+1,times=99),zed[j,],ylim=range(zed,na.rm=T),xlim=c(1,100),pch=16,col=rgb(0.5,0.5,0.5,0.5))
+# }
+# 
+# 
+# saveGIF({
+#   #ani.options(nmax = 30)
+#   plot(-1,-1,ylim=range(zed,na.rm=T),pch=16,col=rgb(0.5,0.5,0.5,0.5))
+# for (i in 1:99){
+# for (j in 3:99){
+#   plot(j,zed[j,i],ylim=range(zed,na.rm=T),pch=16,col=rgb(0.5,0.5,0.5,0.5))
+# }
+# }
+# 
+# }, interval = 0.01, movie.name = "bm_demo.gif", ani.width = 600, ani.height = 600)
+

### 2017-04-18 scripts/calculate_power_metric.R

@@ -12,14 +12,32 @@ require(plyr)
 source('scripts/calculate_slope.R')
 source('scripts/create_subsamples.R')
 
+
+#This is a two-sided test (should use 1 tailed test)
+# calculate_power_for_subsamples = function(pop,alpha){
+#   subset_times=create_subsamples(pop)
+#   power=matrix(0,nrow=(length(pop)-1),1)
+#   for (n in 1:(length(pop)-1)){
+#     power[n,1]=sum(apply(na.omit(subset_times[[n]]),1,calculate_p_value)<alpha)/(length(pop) - n)
+#   }
+#   return(power)
+# }  
+
+
+######## Implement one-tailed test
 calculate_power_for_subsamples = function(pop,alpha){
   subset_times=create_subsamples(pop)
   power=matrix(0,nrow=(length(pop)-1),1)
   for (n in 1:(length(pop)-1)){
-    power[n,1]=sum(apply(na.omit(subset_times[[n]]),1,calculate_p_value)<alpha)/(length(pop) - n)
+    power[n,1]=sum(apply(na.omit(subset_times[[n]]),1,calculate_p_value)<alpha  & sign(apply(na.omit(subset_times[[n]]),1,calculate_slope))==sign(calculate_slope(pop))  )/(length(pop) - n)
   }
   return(power)
 }  
+
+
+
+#######
+
 
 
 

### 2017-04-18 scripts/calculate_slope_metric.R

@@ -8,8 +8,8 @@
 require(plyr)
 
 #pull in neccessary functions
-source('calculate_slope.R')
-source('create_subsamples.R')
+source('scripts/calculate_slope.R')
+source('scripts/create_subsamples.R')
 
 calculate_average_slopes = function(pop){
   subset_times=create_subsamples(pop)
@@ -37,17 +37,17 @@ min_time_needed = function(pop, threshold){
   }
 }
 
-
-
-##########
-# run a loop to find the min time series length rrequrired for each population
-mimimum_time_required = matrix(0,nrow=nrow(pop_info),ncol=1)
-###run loop to make plot
-for (j in names(table(long_dat$ID))){
-  
-  pop=subset(long_dat,long_dat$ID==j)$popvalue[1:40]/max(subset(long_dat,long_dat$ID==j)$popvalue[1:40])
-
-  mimimum_time_required[which(j==pop_info$ID),1] = min_time_needed(pop,0.05)
-
-  print(paste(which(j==pop_info$ID),j,":",mimimum_time_required[which(j==pop_info$ID),1]))
-}
+# 
+# 
+# ##########
+# # run a loop to find the min time series length rrequrired for each population
+# mimimum_time_required = matrix(0,nrow=nrow(pop_info),ncol=1)
+# ###run loop to make plot
+# for (j in names(table(long_dat$ID))){
+#   
+#   pop=subset(long_dat,long_dat$ID==j)$popvalue[1:40]/max(subset(long_dat,long_dat$ID==j)$popvalue[1:40])
+# 
+#   mimimum_time_required[which(j==pop_info$ID),1] = min_time_needed(pop,0.05)
+# 
+#   print(paste(which(j==pop_info$ID),j,":",mimimum_time_required[which(j==pop_info$ID),1]))
+# }

### 2017-04-18 scripts/combine_databases.R

@@ -7,10 +7,10 @@
 
 # Population time series file
 load('For_easton.Rdata')
-dat= dat
+dat= long_dat
 
 # File for biological characteristics of amniotes
-amniote = read.table('Amniote_Database_Aug_2015_erw.csv',sep=',',header=T,na.strings = -999)
+amniote = read.table('data/Amniote_Database_Aug_2015_erw.csv',sep=',',header=T,na.strings = -999)
 amniote$Binomial = paste(amniote$genus,amniote$species,sep=' ')
 
 # Create bio_info table to store informatino where a species is present in both databases
@@ -22,22 +22,22 @@ dat$match =0
 
 for (j in 1:nrow(dat)){
   if (dat$Binomial[j] %in% amniote$Binomial){
-    dat[j,20:48] = amniote[which(amniote$Binomial %in% dat$Binomial[j]),7:35]
-    dat[j,49]=1
+    dat[j,9:37] = amniote[which(amniote$Binomial %in% dat$Binomial[j]),7:35]
+    dat[j,38]=1
   }
     if (j %in% seq(1000,80000,by=5000)){print(j)} #just a counter
 }
 
 # remove coluoms with lots of NA (egg_width_mm forward) #sort(colSums(is.na(dat)),decreasing=T)
-many_NAs=names(sort(colSums(is.na(dat)),decreasing=T))[1:23]
-new_dat = dat[,-which(names(dat) %in% many_NAs)]
-new_dat = new_dat[,-which(names(new_dat) %in% c('year_ass'))]
+few_NAs=names(sort(colSums(is.na(dat)),decreasing=F))[1:12]
+new_dat = dat[,which(names(dat) %in% few_NAs)]
+#new_dat = new_dat[,-which(names(new_dat) %in% c('year_ass'))]
 
 test=na.omit(new_dat)
 length(table(test$ID)) #number of time series
 
 # Which biological variables are most available for long time series?
-long_time_names=names(table(test$ID))[as.numeric(table(test$ID))>25]
+long_time_names=names(table(test$ID))[as.numeric(table(test$ID))>30]
 long_dat = subset(test,test$ID %in% long_time_names)
 length(table(long_dat$ID) )
 
@@ -49,3 +49,31 @@ for (j in 1:length(table(long_dat$ID))){
 }
 
[…]

### 2017-04-18 scripts/create-pop-info-table.R

@@ -12,11 +12,11 @@
 load('data/For_easton.Rdata')
 
 dat=dat[,c(1:7,11)] #
-dat=na.omit(dat)
+#dat=na.omit(dat) #this doesn't work well
 
 #this subsets data to find long time series (e.g. >40 years)
-dat=subset(dat,dat$class != 'Aves')
-long_time_names=names(table(dat$ID))[as.numeric(table(dat$ID))>30]
+#dat=subset(dat,dat$class != 'Aves')
+long_time_names=names(table(dat$ID))[as.numeric(table(dat$ID))>35]
 long_dat = subset(dat,dat$ID %in% long_time_names)
 length(table(long_dat$ID) )
 
@@ -35,16 +35,19 @@ pop_info$overall_trend=NA
 pop_info$autocorrelation= NA
 pop_info$trend_p_value=NA
 pop_info$min_time_for_power=NA
+pop_info$pop_value=as.numeric(pop$pop_value)
+long_dat$pop_value=long_dat$pop_value
   for (j in 1:nrow(pop_info)){
-    pop= subset(long_dat,long_dat$ID==pop_info$ID[j])[1:30,]
-    pop$popvalue=pop$popvalue/max(pop$popvalue)
-    pop_info$variance[j]= var(pop$popvalue)
-    pop_info$overall_trend[j] = calculate_slope(pop$popvalue)
-    pop_info$autocorrelation[j] = acf(pop$popvalue,plot = F)$acf[2]
-    pop_info$trend_p_value[j]=calculate_p_value(pop$popvalue)
+    pop= subset(long_dat,long_dat$ID==pop_info$ID[j])
+    pop$pop_value=as.numeric(pop$pop_value)
+    pop$pop_value=pop$pop_value/max(pop$pop_value) #should I standaridize this in a better way
+    pop_info$variance[j]= var(pop$pop_value)
+    pop_info$overall_trend[j] = calculate_slope(pop$pop_value)
+    pop_info$autocorrelation[j] = acf(pop$pop_value,plot = FALSE)$acf[2]
+    pop_info$trend_p_value[j]=calculate_p_value(pop$pop_value)
     
     source('scripts/calculate_power_metric.R')
-    pop_info$min_time_for_power[j] = min_time_needed(pop$popvalue,0.05,0.8)
+    pop_info$min_time_for_power[j] = min_time_needed(pop$pop_value,0.05,0.8)
[…]

### 2017-04-18 scripts/gam_example.R

@@ -1,0 +1,10 @@
+
+
+# Example gam
+gam.object <- gam(popvalue ~ s(year,6),data=moose,family=poisson)
+summary(gam.object)
+plot(gam.object,se=TRUE)
+gam.newdata = as.data.frame(c(1950:2020))
+names(gam.newdata)='year'
+predict(gam.object,type="terms",newdata=gam.newdata)
+plot(1950:2020,predict(gam.object,type="terms",newdata=gam.newdata))

### 2017-04-18 scripts/remove_NA_populations.R

@@ -1,0 +1,29 @@
+
+
+
+# Remove records on interior of time series for each individual population
+load('data/For_easton.Rdata')
+
+dat=dat[,c(1:7,11)] #
+
+dat=testing
+ID_codes = names(table(dat$ID))
+good_rows = NULL
+
+#dat=long_dat
+
+for (j in 2650:length(ID_codes)){
+  #pop = subset(dat,dat$ID==ID_codes[j])[1:35,]
+  pop = subset(dat,dat$ID==ID_codes[j])
+  pop=pop[head(which(is.na(as.numeric(pop$pop_value))==F),1):tail(which(is.na(as.numeric(pop$pop_value))==F),1),]
+  if (nrow(pop) == nrow(na.omit(pop)) &nrow(na.omit(pop))>35){
+    good_rows=cbind(good_rows,row.names(pop))
+  }
+  print(j)
+}
+
+dat = dat[row.names(dat) %in% c(good_rows),]
+
+
+c(bad_IDs); length(bad_IDs)
+#Removes over 500 populations!!

### 2017-04-18 scripts/statistical-analyses-biological.R

@@ -1,0 +1,52 @@
+
+
+
+# Statistical analyses
+
+#bring in data frame
+test = test
+pop_info_test=pop_info
+
+
+
+# Check for highly correlated traits
+
+# Simple GLM with poisson error structure
+poisson_model <-glm(min_time_for_power~ gen_len + litter_or_clutch_size_n + litters_or_clutches_per_y + adult_body_mass_g+ maximum_longevity_y + egg_mass_g +incubation_d,family = 'poisson',data=pop_info)
+
+# Plots of residuals
+EP <- resid(poisson_model,type='pearson')
+ED <- resid(poisson_model,type='deviance')
+mu <- predict(poisson_model, type='response')
+E <- pop_info$min_time_for_power - mu
+EP2 <- E/ sqrt(7.630148 *mu)
+op <- par(mfrow=c(2,2),mar=c(4,4,0,0))
+plot(mu,E,main='Response residuals')
+plot(mu, EP, main='Pearson residuas')
+plot(mu, EP2, main='Pearson residuals scaled')
+plot(mu, ED, main='Deviance residuals')
+par(op)
+
+par(mfrow=c(2,2),mar=c(5,5,1,2),oma=c(0,0,0.5,0))
+plot(poisson_model,cex.lab=1.2,las=1,cex.axis=1.2,main=' ')
+
+
+#R^2 calculation (pg. 218 Zuur)
+R2 <- 100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance)
+
+
+# GLM with negative binomial error structure
+# pull in packages that I had used previously with Cocos project
+require(glmmADMB)
+require(MuMIn)
+
+#CocosData$SiteCode <- as.factor(CocosData$SiteCode)
+#CocosData$CurrentCode <- as.factor(CocosData$CurrentCode)
+pop_info$Binomial <- as.factor(pop_info$Binomial)
+
+#model <- glmmadmb(formula=min_time_for_power~gen_len+habitat + exploitation + invasive + mass ,data=pop_info,family="nbinom",zeroInflation=FALSE)
+model <- glmmadmb(formula=min_time_for_power~ gen_len + litter_or_clutch_size_n + litters_or_clutches_per_y + adult_body_mass_g+ maximum_longevity_y + egg_mass_g +incubation_d,data=pop_info,family="poisson",zeroInflation=FALSE)
+
[…]

### 2017-04-18 scripts/statistical-analyses.R

@@ -1,25 +1,61 @@
 # Created by Easton White
 # Created on 17-Jan-2017
-# Last edited on 18-Jan-2017
+# Last edited on 30-Jan-2017
 
 #Statistical analyses of minimum time needed and what it may correlate with
 # Should probably do poisson or negative binomial with random effects (species or family)
 
 
+# Results so far: It appears the Poisson glm does a pretty good job when examining residual plots. Generation length has positive effect size if you take trend, autocorrelation, and variance into account
+# Using formula from pg. 218 of Zuur, I can explain about 68% of the variation in mimimum time required
+
 # Pull in final data set here to be analyzed
+load('pop_info_example.Rdata')
 
-# Run simple model
-with(trial,summary(glm(min_time_for_power~gen_len+habitat + exploitation + invasive + mass + trophic,family = 'poisson')))
+# Need to look for highly correlated predictor variables and only use those not highly correlated
 
+pop_info$abs_overall_trend = abs(pop_info$overall_trend)
+# Simple linear regression
+linear_model <- with(pop_info, summary(lm(min_time_for_power~ abs_overall_trend + autocorrelation + variance + gen_len)))
+
+# Plots of residuals
+
+# Simple GLM with poisson error structure
+poisson_model <-glm(min_time_for_power~ abs_overall_trend + autocorrelation + variance + gen_len,family = 'poisson',data=pop_info)
+
+# Plots of residuals
+EP <- resid(poisson_model,type='pearson')
+ED <- resid(poisson_model,type='deviance')
+mu <- predict(poisson_model, type='response')
+E <- pop_info$min_time_for_power - mu
+EP2 <- E/ sqrt(7.630148 *mu)
+op <- par(mfrow=c(2,2),mar=c(4,4,0,0))
+plot(mu,E,main='Response residuals')
+plot(mu, EP, main='Pearson residuas')
+plot(mu, EP2, main='Pearson residuals scaled')
+plot(mu, ED, main='Deviance residuals')
+par(op)
+
+par(mfrow=c(2,2),mar=c(5,5,1,2),oma=c(0,0,0.5,0))
+plot(poisson_model,cex.lab=1.2,las=1,cex.axis=1.2,main=' ')
+
+
+#R^2 calculation (pg. 218 Zuur)
+R2 <- 100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance)
+
+
[…]

### 2017-04-18 writing/manuscript_draft.Rmd
changes in intro, methods, and references

@@
 date: "January 18, 2017"
-output: html_document
+output:
+  pdf_document: default
+  word_document: default
+bibliography: White_bib.bib
 ---
@@
 
-## Introduction 
+## Introduction (intro long time series, my general question, definitions of power, general power interpretation, two approaches, what we do in this paper)
+
 Observational studies, and censusing populations, has become a cornerstone of modern ecology and conservation biology.
 Longterm data is necessary to understand population dynamics and to assess extinction risk of species.
-Even many time series may now be considered "longterm", most are still short (i.e.
+Even though many time series may now be considered "longterm", most are still short (i.e.
 less than X years).
-I could have statistic on how long is considered "longterm" in papers.
+*I could have statistic on how long is considered "longterm" in papers*.
 Time series are typically short for a variety of reasons.
@@
 
-(lack of statistical tools to evaluate power of time series) - can simulate or subsample data.
-In order to properly estimate mimumn time seris you need to know     IUt will also depend on your question
+Naturally, one might ask, how long of a time series is actually neccessary?
+This question has important scientific and management implications.
+Scientifically, we need to know how long a time series must be in order to rigously test a specific question.
+Too short of a time series may lead to wrong conclusions given large natural year-to-year variability.
+From a management perspective, it is important to understand when a trend in abundance over time is actually meaningful or not.
+For instance the International Union for the Conservation of Nature (IUCN) suggest...
 
-In determining the number of samples required for a particular experiment X quantities are intricately linked: statistical power, effect size, alpha, sample size,  Formally, statistical power is the one minus the probability of a type II error, or false negative.
-In other words, it is a measure of how well you could detect a trend given it is statistically significant.
-Prior to an experiment, one could set appropriate levels of power, alpha, and the effect size to estimate the sample size required for the experiment.
+In determining the number of samples required for a particular experiment X quantities are intricately linked: statistical power, effect size, alpha, sample size.
+The relationship between these quantitaties depends on the specific statistical test used.
+Formally, statistical power is the one minus the probability of a type II error, or false negative.
+In other words, it is a measure of how well you could detect a trend given it is actually significant.
+Prior to an experiment, one could set appropriate levels of power, $\alpha$, and the effect size to estimate the sample size required for the experiment.
 This approach, however, is not as simple for a time series as the data points are clearly non-independent.
-Therefore, two approaches can be used to estimate the number of time points required to detect a trend.
-The first approach is a simulation approach.
-The second approach
 
-Little work has actually gone into evaluating how long a time series needs to be to address questions of interest.
-The minimum length of a time series is similar to the question of the minimal number of samples required for any experiment.
-However, samples in a time series cannot be considered independent (citation).
-The mimumum length of a time series will depend on a number of factors.
-First, the question will obviously drive a lot of the length reuirments.
-If you are interested in how a population responds to El Nino, this will need a much longer time series compared to if you wanted to know Second, properties of the species of interest, and their population will affect the minimum time series required, including the presence of internally driven cycles, variance in population size, generation time,
-The IUCN provides minumim time series criteria to assess species threat status. They require a 10 year, or 3 genration, whichever is shorter To determine status.
+For time series data, two general approaches can be used.
+First, as can be done with experimental data, we can relax the definition of statistical power (cite Bolker).
+Instead of using classic statistical tests (e.g.
+t-test), and corresponding power calculations, simulation approaches can be used.
+Simple population models can be run with parameter values corresponding to a population of interest.
+Out of a large number of simulations, power can be calculated as simply the fraction of simulations that meet some criteria.
+The specific criteria depend on the question at hand.
+Then one can run simulations for time series of different lengths to determine the minimum time series length required to be confident that you will be able to answer your question.
+In addition to using simulated data, empirical time series can also be used.
+We usually do not have multiple replicates of simular populations, but we can subsample an empirical time series.
+Subsamples of different lengths can then be evaluated to see which fraction of subsamples meet some criteria, again a measure of statistical power.
 
-How sensitivity are our findings to changes in the length of our time series?
-
-Recently, several databases (e.g.) of animal population censuses have emerged that facilate the comparison of many populations.
-
-In this paper, we evaluate the minimal time series length required to answer X questions.
+In this paper, we use both simulated data and empirical time series to determine the minimum number of years required.
+To demostrate these techniques, we answer the specific question of how many years are required to obtain a good estimate of the real slope parameter?...
+<!-- In this paper, we evaluate the minimal time series length required to answer X questions.
 1) How sensitivive are IUCN asssessments to time series lengths?
@@
 We demonstrate our approach with X case studies and then show results for X species from the Blah Blah Database.
+-->
 
@@
 ### Data source
-We use a series of XXXX time series compiled in @Keith2015.
+We use a series of XXXX time series compiled in [@Keith2015].
 The database includes information on X vertebrate species with a focus on mammals, birds, and fish.
@@
 
-For a subset of populations in the database, we were able to connect a database on biological characteristic
-Analyses
-Our objective is to determine the miniuum time series length (T_min) required to answer various questions.
+For a subset of populations in the database, we were able to connect a database on biological characteristics...
+
+### Theoretical approach
+
+One approach to determining the mimumum time series needed is through repitive simulations of population models (cite old paper).
+This is the same approach one might use in sample size calculations for any experimental design too complicated for simple power analyses.
+Essentially, we take a stochastic population model and simulate it for a number of years repetively.
+We can then determine the mimimum time series required to achieve designed levels of significance and power.
+This approach requires us to determine values for our model parameters.
+As an example, we can take the following population model for population size $N$ at time $t$:
+$$
+N(t + 1) = N(t) + r(t) \mbox{ with } r(t) \sim N(\phi \mu, \sigma)
+$$
+where $\epsilon$ is a normally-distributed random noise term with mean $\mu$ and standard deviation $\sigma$.
+The rate of growth $r$ is also the trend strength of increase or decrease.
+Lastly, $\phi$ represents an autocorrelation term where $\phi=0$ indicates no autocorrelation, $\phi>0$ is postively autocorrelated.
+This model represents linear growth with autocorrelation and noise.
+However, it is certainly possible to use any other population model.
+
+By simulating this model repetively for different lengths of time, we can determine the mimumum time series required.
+Here statistical power is simply the number of trials that are themselves significant, as the real trend is different from zero.
+Then the mimumum time required is the mimumum time required to obtain statistical power above some pre-determined threshold.
+Historically, 0.8....s
+
+In figure X, a number of simulated time series are shown for a set number of time periods (t=40).
+It is clear that statistical power increases quickly with increases in $t$ (Fig.
+Xb).
+Once power is greater than XX (the dotted line), that is the mimumum time required ($t_{min}$
+
+![optional caption text](example_theoretical_output.pdf)
+
+```{r,echo=F,eval=F,fig.width=.83}
+setwd("~/Desktop/Research/time-series-project")
+source("scripts/theoretical-models/calculate_power.R")
+source("scripts/theoretical-models/simulate_pop.R")
+
+  years_to_sample=seq(2,100,by=1)
+  ptm <- proc.time()
+  y=lapply(X=years_to_sample,FUN = calculate_power,phi = 0.6,r=2,sigma = .2,trials=300)
+  proc.time() - ptm
+
+  simulations = replicate(n=30,simulate_pop(phi = 0.6,r=2,sigma = .2,max.time=40))
+  min_time_needed = years_to_sample[tail(which(y<0.8),1)] + 1
+
+    par(mfrow=c(1,2),oma=c(1,2,0,0.5),mar=c(4,5,0.5,0.5))
+  matplot(simulations,type='l',col=rgb(0.5,0.5,0.5,0.5),lty=1,ylab='',xlab='',las=1,cex.axis=1.2)
+  mtext(text = 'time' ,side = 1,cex = 1.2,line=3)
+  mtext(text = 'population size' ,side = 2,cex = 1.2,line=4)
+
+  plot(years_to_sample,y,type='l',lwd=2,las=1,ylab='',xlab='',cex.axis=1.2)
+  abline(v=min_time_needed,lty=2,lwd=2,col='red')
+  abline(h=0.8,lty=2,lwd=2,col='red')
+  mtext(text = 'time sampled' ,side = 1,cex = 1.2,line=3)
+  mtext(text = 'statistical power' ,side = 2,cex = 1.2,line=3.5)
+
+```
+
+### Empirical approach
+
+Our objective is to determine the miniuum time series length ($T_{min}$) required to answer a specific question.
 As we describe below, the minimum time series length calculation is particular to each question.
@@
 In other words, they are a representive sample.
-We then take random contiguous subsamples of each time series and compare each subset to the measurements on the entire time series.
-For example, for Isla Royale wolves there are XX years of data available.
-We can then take contiguous subsets of those XX years.
-This results in XX subsets of length 2 years, XX subsets of length 3 years, and so forth until one subset of length XX years.
-We then examine how various measurements on the population change as we increase the subset length.
+We first take all possible contiguous subsamples of each time series.
+For example, a time series of 40 years would have 39 possible contiguous subsamples of length 2, 38 possible contiguous subsamples of length 3, and so forth.
+In this paper, we will try to ask a simple, specific question.
+We seek to determine the $T_{min}$ required to be confident in an estimate of the slope coefficient from linear regression.
+We run a linear regression for each subsampled time series.
+Then we determine the fraction of subsamples of a particular length that have slopes from linear regression which are statisically different from zero given the longterm, or ``true", time series also has a significant slope.
+This is a measure of statistical power.
+Lastly, we determine which subsample length is required to guarentee we have above a certain threshold of statistical power, which here we will assume is 0.8 by convention (citation).
 
-Question 1: What length of time series is needed to detect a significant trend in abundance?
-Here we subsample each time series to determine the T_min required to calculate the overall time series trend.
-We first take all possible contiguous subsamples of the overall time series.
-For example, a time series of 40 years would have 39 possible contiguous subsamples of length 2, 38 possible contiguous subsamples of length 3, and so forth.
-For each subsample we run linear regression to determine the slope parameter and if the slope parameter estimated is significant.
-Then we can determine the total number of contiguous subsamples of a certain length that detect a significant trend given it is present---this is an estimate of statistical power.
+We are examined how minimum time series calculations changed depending on the question of interest.
+In the supplementary material, we show how to make similar calculations to determine the minimnm time series required to estimate overall population growth rate, the variability in population size, and future extinction risk.
 
-Figure 1 - maybe make one figure for questions 1,2, and 3
+```{r, echo=F,message=F,warning=F}
+# Example of subsampling empirical data
+setwd("~/Desktop/Research/time-series-project")
+ load("~/Desktop/Research/time-series-project/pop_info_example_goodNAs.Rdata")
+ par(mfrow=c(1,2))
+pop = subset(long_dat,long_dat$ID=='10696') #13192 for Harp seals
+plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='population size',xlab='time (years)',pch=16,ylim=c(0,2000),las=1)
+abline(lm(pop$popvalue~pop$year),lwd=2,col='red')
+# could insert all possible regressions of length n
+source('scripts/calculate_power_metric.R')
+power_example=calculate_power_for_subsamples(pop$popvalue,0.05)
+plot(power_example,cex.axis=1.2,cex.lab=1.2,ylab='average statistical power',xlab='years sampled',las=1,pch=16)
+# could also insert standard error bars
+abline(h=0.8,col='red',lwd=2,lty=2)
+abline(v=22,col='red',lwd=2,lty=2)
 
-We define the minimum time required as the point when the curve begins to saturate above some threshold.
-For instance, in figure 1 if we want power of at least 0.8, then it would take XX years to be sure that power will be above that threshold.
-We then define T_min as that specific number of year.
+```
 
-Question 2: How sensitive are IUCN assessments to time series trends?
+<!--Question 2: How sensitive are IUCN assessments to time series trends?
 The IUCN sets threat statuses for a variety of species.
@@
 
-Populations vary from year-to-year for several reasons including
+Populations vary from year-to-year for several reasons including-->
 
 ## Results
+
+We evaluated the question of what minimum time series is required to answer a question of interest.
+Here we ask, what is the minimum time series length required to determine, via linear regression, the longterm trend in population abundance?
+We investigated this question through two approaches: simulation-based and empirical analyses.
+In both cases, we found the minimum time series length required depended strongly on the trend strength, population variability, and autocorrelation.
+
+### Simulation approach
+
+We constructed a general population model where the trend over time could be a model parameter.
+We chose a simple model, but any other population model could be used.
+THe specific model choice should be tailored to the population of interest.
+We then ran simulations of different lengths to determine the minimum time series length required to achieve a certain level of power (which we set to 0.8).
+We were also able to alter the parameter values (variability and autocorrelation) of the population model to see how this altered $T_{min}$.
+
+In line with past work (citation), we found the $T_{min}$ increases (i.e. more time is required) with decreases in trend strength or autocorrelation and with increases in variability.
 
@@
 
-```{r,echo=F,eval=F,include=F}
+### Empirical test
+
+```{r,echo=F,fig.cap='duh\\label{fig:min_time_dist}'}
 # Pull in analysis of average min time needed for **empirical species** (save analysis as Rdata file or something)
+  load("~/Desktop/Research/time-series-project/pop_info_example_goodNAs.Rdata")
+
+#only look at pops with significant longterm declines or increases
+pop_info=pop_info[pop_info$trend_p_value<0.05,]
+long_dat=subset(long_dat,long_dat$ID %in% pop_info$ID)
+long_time_names=subset(long_time_names,long_time_names %in% pop_info$ID)
+
+plot(density(pop_info$min_time_for_power),las=1,ylab='',xlab='',cex.axis=1.2,main='')
+mtext(text = 'Frequency',2,line=3,cex=1.2)
+mtext(text = 'Minimum time required',1,line=3,cex=1.2)
+```
+
+```{r poisson_model, echo=F}
+#Poisson regression
+pop_info$abs_overall_trend = abs(pop_info$overall_trend)
+poisson_model <-glm(min_time_for_power~ abs_overall_trend + autocorrelation + variance + gen_len,family = 'poisson',data=pop_info)
+par(mfrow=c(2,2),mar=c(5,5,1,2),oma=c(0,0,0.5,0))
+plot(poisson_model,cex.lab=1.2,las=1,cex.axis=1.2,main=' ')
+
+#R^2 calculation (pg. 218 Zuur)
+R2 <- 100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance)
 
@@
 
-- Intro paragraph on what we generally found (average T_min and range), many species needed the full time series
+```{r,echo=F,fig.cap='duh\\label{fig:correlates}'}
 
-Across all the populations we examined, we found an average minimum time series length required ($T_{min}$) of `r 2+3 #calculation here`.
+  par(mfrow=c(3,1),mar=c(4,1,1,1),oma=c(0,4.5,0,0.5))
+  plot(abs(pop_info$overall_trend),pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
+  mtext('| trend strength |',1,line=3,cex=1.2)
+    plot(pop_info$variance,pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
+    mtext('population variance',1,line=3,cex=1.2)
+      plot(pop_info$autocorrelation,pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
+      mtext('autocorrelation',1,line=3,cex=1.2)
 
+       mtext('minimum time required',2,line=2.5,cex=1.2,outer = T)
+
+```
+
+We examined a database of `r nrow(pop_info)` seperate population time series representing `r length(table(pop_info$Binomial))` species.
+This database consists of species with a variety of life history characteristics.
+We limited our analyses to populations with at least 35 years of continuous sampling.
+
+Across all the populations we examined, we found an average minimum time series length required ($T_{min}$) of `r mean(pop_info$min_time_for_power)` ($\sigma$=`r (var(pop_info$min_time_for_power))^0.5`), with a wide distribution (Fig.
+\ref{fig:min_time_dist}).
+As expected based on theoretical results (cite papers), the minimum time series length required is strongly correlated with trend strength, population variance, and autocorrelation (Fig \ref{fig:correlates}).
+Using a generalized linear model, with a Poisson error structure, all three of these explanatory variables were significant and had large effect sizes (cite model here).
+Combined trend strength, population variance, and autocorrelation account for about `r 100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance)` of the variance in minimum time series length required (as proposed by CITE Zuur).
+
+We also examined a subset of populations where more explantatory variables were known (e.g.
+body mass, longevitiy, location).
+None (need to do regression) of these explanatory variables were predictive of the minimum time series length required.
+Some differences if $t_{min}$ did occur between biological classes, but this can be attributed to the amount populations are know to fluctuate.
+
+<!--
 - What does a theoretically model predict (does min time scale with trend, power, variability, autocorrelation, something else) - definitely depends on what you set for alpha and beta
@@
     - Correlates well with trend, variability, and autocorrelation
-    - Surprisingly does not correlate well with generation time or any other biological variables (need to do proper regression though)
+    - Surprisingly does not correlate well with generation time or any other biological variables (need to do proper regression though)-->
 
-```{r,echo=F,eval=F,include=F}
-# Create plots from min time needed analyses
+```{r,echo=F}
+# Biological correlates plots
+par(mfrow=c(2,2),mar=c(4,5,0.5,0.5))
+
+boxplot(pop_info$min_time_for_power~pop_info$class, lwd = 1, ylab = 'minimum time required',col='white',cex.lab=1.4,cex.axis=1.2,las=1)
+stripchart(pop_info$min_time_for_power~pop_info$class, vertical = TRUE,method = "jitter", add = TRUE, pch = 20, col = rgb(0.5,0.5,0.5,0.5))
+
+boxplot(abs(pop_info$overall_trend)~pop_info$class, lwd = 1, ylab = ' ',col='white',cex.lab=1.4,cex.axis=1.2,las=1)
+stripchart(abs(pop_info$overall_trend)~pop_info$class, vertical = TRUE,method = "jitter", add = TRUE, pch = 20, col = rgb(0.5,0.5,0.5,0.5))
+  mtext('| trend strength |',2,line=3.5,cex=1.2)
+
+boxplot(pop_info$variance~pop_info$class, lwd = 1, ylab = 'population variance',col='white',cex.lab=1.4,cex.axis=1.2,las=1)
+stripchart(pop_info$variance~pop_info$class, vertical = TRUE,method = "jitter", add = TRUE, pch = 20, col = rgb(0.5,0.5,0.5,0.5))
+
+boxplot(pop_info$autocorrelation~pop_info$class, lwd = 1, ylab = 'autocorrelation',col='white',cex.lab=1.4,cex.axis=1.2,las=1)
+stripchart(pop_info$autocorrelation~pop_info$class, vertical = TRUE,method = "jitter", add = TRUE, pch = 20, col = rgb(0.5,0.5,0.5,0.5))
 ```
 
+```{r,echo=F}
+par(mfrow=c(1,1))
+  plot(pop_info$gen_len,pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
+  mtext('generation time (years)',1,line=3,cex=1.2)
+  mtext('minimum time required',2,line=2.5,cex=1.2,outer = F)
+```
+
+```{r,echo=F}
+summary(poisson_model)
+```
+
+<!--
 ## Potential figures
@@
 - Separate results for species, body size, location, biology. - multi panel figure
+
+-->
 
@@
 - Future recommendations to others
+
+The IUCN has an 10 year/3 generation rule...
+If we apply this rule to the species in our database we see that approximently `r 100*sum(pop_info$min_time_for_power>pop_info$IUCN)/nrow(pop_info)`% of populations need a longer minimum time series than that required by IUCN.
+
+## References
+
+# Supplementary material
+
+```{r,echo=F}
+# Example of using population growth rate instead of power
+setwd("~/Desktop/Research/time-series-project")
+source("scripts/calculate_geometric_growth.R")
+
+par(mfrow=c(1,2))
+pop = subset(long_dat,long_dat$ID=='10696') #13192 for Harp seals
+plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='',xlab='',pch=16,ylim=c(0,2000),las=1)
+abline(lm(pop$popvalue~pop$year),lwd=2,col='red')
+mtext(text = 'year',1,line=3,cex=1.4)
+mtext(text = 'population size',2,line=3.5,cex=1.4)
+
+aaa=calculate_average_geometric_growth(pop$popvalue)
+plot(rowMeans(aaa,na.rm=T),cex.axis=1.2,pch=16,ylab='',xlab='',las=1)
+mtext(text = 'years sampled',1,line=3,cex=1.4)
+mtext(text = 'population growth rate',2,line=3.7,cex=1.4)
+```

### 2017-04-18 writing/manuscript_draft.Rmd
added abstract from ESA talk

@@
 ## Abstract
-Longterm census data is necessary to better understand population dynamics, assess threat statuses, and make future predictions for numerous populations.
-However, there has been very little work done to assess how long a time series needs to be to answer a particular question.
-Several databases have emerged to compile long term time series.
-Here we empirically evaluate the minimum time series length required to answer several questions of biological interest.
-We address this question using time series data on XX populations of XX species including We find that most time series In the future, we advise
+Long-term census data is necessary to better understand ecological processes and make management decisions.
+In addition, census data is often expensive, requiring a lot of people-hours or equipment.
+However, little work has actually addressed the length of time series required.
+In other words, when is a time series of census data long enough to address a question of interest?
+Here, I explore two approaches to address this question: simulation-based and an empirical approach.
+I specifically determine the minimum time series length (e.g.
+number of years) required to estimate significant increases or decreases in population abundance.
+Importantly, I examine the ability to detect a trend with a set level of statistical power, which is often neglected in ecological time series analyses.
+Using simple simulations, I demonstrate how the minimum time series length required increases with weaker trends in abundance and with higher variability in population size.
+In addition, I examine 868 populations of vertebrate species to determine the minimum time required to detect changes in their abundance.
+I found that 10-15 years of continuous monitoring are required in order to achieve a high level of statistical power.
+Similar to simulation-approaches, the minimum time required for field census data strongly depends on trend strength, population variability, and temporal autocorrelation.
+Perhaps surprisingly, the minimum time required did not correlate well with biological explanatory variables, like body size or generation length.
+These results point to the importance of sampling populations over long periods of time.
+Here, I stress the need for verifying a study has enough statistical power to accurately determine long-term changes in population abundance.
+Most studies, especially those less than 10-15 years, are probably underpowered and potentially misleading.
 

### 2017-04-18 writing/manuscript_draft.Rmd
added to methods section

@@
 We usually do not have multiple replicates of simular populations, but we can subsample an empirical time series.
-Subsamples of different lengths can then be evaluated to see which fraction of subsamples meet some criteria, again a measure of statistical power.
+Subsamples of different lengths can then be evaluated to see which fraction of subsamples meet some criteria, again a measure of statistical power...
+
+Past work...
 
@@
 
+```{r, echo=F}
+setwd("~/Desktop/Research/time-series-project")
+load('cleaned-data/pop_info_example_goodNAs.Rdata')
+```
+
 ### Data source
-We use a series of XXXX time series compiled in [@Keith2015].
-The database includes information on X vertebrate species with a focus on mammals, birds, and fish.
-We filtered out short time series (less than XX) and were left with XX time series with an average length of XX.
-The database also includes information on generation length, census type,
+We use a database of `r nrow(pop_info)` population time series compiled in [@Keith2015].
+The data is orignally from the Global Population Dynamics Database (citation needed).
+The database includes information on `r length(table(pop_info$Binomial))` vertebrate species with a focus on mammals, birds, and fish.
+We filtered out short time series (less than `r pop_info$series_length[1]`) and were left with XX time series with an average length of XX.
+The database also includes information on generation length, census type,.
+For each time series, we also calculated variables of interest like variance in population size, longterm trend in abundance, and temporal autocorrelation.
+We also calcuated the minimum number of years required to sample according to IUCN standards under rule ...
+(citation needed).
 
@@
 
-Our objective is to determine the miniuum time series length ($T_{min}$) required to answer a specific question.
+Our objective is to determine the miniuum time series length ($T_{min}$) required to answer a specific question of interest.
 As we describe below, the minimum time series length calculation is particular to each question.
+Intuitively, we might expect a longer minimum time series length required given more complicated questions.
+
 We assume that the time series in our database are long enough to include all necessary information (e.g.
@@
 We seek to determine the $T_{min}$ required to be confident in an estimate of the slope coefficient from linear regression.
-We run a linear regression for each subsampled time series.
-Then we determine the fraction of subsamples of a particular length that have slopes from linear regression which are statisically different from zero given the longterm, or ``true", time series also has a significant slope.
+Therefore, we run a linear regression for each subsampled time series.
+Then, we determine the fraction of subsamples of a particular length that have slopes from linear regression which are statisically different from zero given the longterm, or ``true", time series also has a significant slope.
 This is a measure of statistical power.
@@
 
-We are examined how minimum time series calculations changed depending on the question of interest.
+We also examine how minimum time series calculations changed depending on the question of interest.
 In the supplementary material, we show how to make similar calculations to determine the minimnm time series required to estimate overall population growth rate, the variability in population size, and future extinction risk.
+
+In the appendix, we explore other potential questions of interest...
 
@@
 setwd("~/Desktop/Research/time-series-project")
- load("~/Desktop/Research/time-series-project/pop_info_example_goodNAs.Rdata")
+ load("cleaned-data/pop_info_example_goodNAs.Rdata")
  par(mfrow=c(1,2))
@@
 We evaluated the question of what minimum time series is required to answer a question of interest.
-Here we ask, what is the minimum time series length required to determine, via linear regression, the longterm trend in population abundance?
+More specifically we ask, what is the minimum time series length required to determine, via linear regression, the longterm trend in population abundance?
+The minimum time series length required is that which had high enough statistical power.
 We investigated this question through two approaches: simulation-based and empirical analyses.
@@
 We chose a simple model, but any other population model could be used.
-THe specific model choice should be tailored to the population of interest.
+The specific model choice should be tailored to the population of interest, based on species ecology.
 We then ran simulations of different lengths to determine the minimum time series length required to achieve a certain level of power (which we set to 0.8).
-We were also able to alter the parameter values (variability and autocorrelation) of the population model to see how this altered $T_{min}$.
+We were also able to alter the parameter values (trend strength and population variability) of the population model to see how this altered $T_{min}$.
 
@@
 # Pull in analysis of average min time needed for **empirical species** (save analysis as Rdata file or something)
-  load("~/Desktop/Research/time-series-project/pop_info_example_goodNAs.Rdata")
+ setwd("~/Desktop/Research/time-series-project")
+ load("cleaned-data/pop_info_example_goodNAs.Rdata")
 
@@
 
-```{r,echo=F}
+```{r,echo=F,eval=F}
 summary(poisson_model)
@@
 ## Discussion
-- Findings
+- Findings and how they relate to past work
 - Extensions to other systems. Useful to use similar species
-- Other questions to ask
-- Comparisons to past work
-- Future recommendations to others
+- Other questions to ask and results in appendix
+- Future recommendations to others and future work
+- conclusion paragraph
 

### 2017-04-20 writing/manuscript_draft.Rmd
added outlines for sections and checked spelling

@@
 author: "Easton R. White"
-date: "January 18, 2017"
+date: "22-Apr-2017"
 output:
@@
 Here, I explore two approaches to address this question: simulation-based and an empirical approach.
-I specifically determine the minimum time series length (e.g.
-number of years) required to estimate significant increases or decreases in population abundance.
+I specifically determine the minimum time series length required to estimate significant increases or decreases in population abundance.
 Importantly, I examine the ability to detect a trend with a set level of statistical power, which is often neglected in ecological time series analyses.
@@
 Here, I stress the need for verifying a study has enough statistical power to accurately determine long-term changes in population abundance.
-Most studies, especially those less than 10-15 years, are probably underpowered and potentially misleading.
+Most studies, especially those less than 10-15 years, are probably under-powered and potentially misleading.
 
-## Introduction (intro long time series, my general question, definitions of power, general power interpretation, two approaches, what we do in this paper)
+## Introduction
+
+- intro long time series
+- my general question
+- primer on power
+- more general power interpretation
+- 2+ approaches
+- what we do in this paper
 
@@
 
-Naturally, one might ask, how long of a time series is actually neccessary?
+Naturally, one might ask, how long of a time series is actually necessary?
 This question has important scientific and management implications.
-Scientifically, we need to know how long a time series must be in order to rigously test a specific question.
+Scientifically, we need to know how long a time series must be in order to rigorously test a specific question.
 Too short of a time series may lead to wrong conclusions given large natural year-to-year variability.
@@
 
-In determining the number of samples required for a particular experiment X quantities are intricately linked: statistical power, effect size, alpha, sample size.
-The relationship between these quantitaties depends on the specific statistical test used.
+In determining the number of samples required for a particular experiment X quantities are intricately linked: statistical power, effect size, alpha, sample size [@Legg2006].
+The relationship between these quantities depends on the specific statistical test used.
 Formally, statistical power is the one minus the probability of a type II error, or false negative.
@@
 In addition to using simulated data, empirical time series can also be used.
-We usually do not have multiple replicates of simular populations, but we can subsample an empirical time series.
+We usually do not have multiple replicates of similar populations, but we can subsample an empirical time series.
 Subsamples of different lengths can then be evaluated to see which fraction of subsamples meet some criteria, again a measure of statistical power...
 
-Past work...
+Past work has investigated these questions with either theoretical approaches not using data or by examining a single population. As an example of the former, [@Rhodes2011] examined..
 
-In this paper, we use both simulated data and empirical time series to determine the minimum number of years required.
-To demostrate these techniques, we answer the specific question of how many years are required to obtain a good estimate of the real slope parameter?...
+In this paper, we use both simulations and a database of empirical time series to determine the minimum number of years required to address various questions.
+To demonstrate these techniques, we address three specific questions.
 <!-- In this paper, we evaluate the minimal time series length required to answer X questions.
@@
 
+- data source
+- empirical approach
+- theoretical approach
+- stuff for the appendix
+
 ```{r, echo=F}
+# Pull in cleaned database
 setwd("~/Desktop/Research/time-series-project")
@@
 We use a database of `r nrow(pop_info)` population time series compiled in [@Keith2015].
-The data is orignally from the Global Population Dynamics Database (citation needed).
+The data is originally from the Global Population Dynamics Database (citation needed).
 The database includes information on `r length(table(pop_info$Binomial))` vertebrate species with a focus on mammals, birds, and fish.
@@
 For each time series, we also calculated variables of interest like variance in population size, longterm trend in abundance, and temporal autocorrelation.
-We also calcuated the minimum number of years required to sample according to IUCN standards under rule ...
+We also calculated the minimum number of years required to sample according to IUCN standards under rule ...
 (citation needed).
@@
 
-One approach to determining the mimumum time series needed is through repitive simulations of population models (cite old paper).
+One approach to determining the minimum time series needed is through repetitive simulations of population models (cite old paper).
 This is the same approach one might use in sample size calculations for any experimental design too complicated for simple power analyses.
-Essentially, we take a stochastic population model and simulate it for a number of years repetively.
-We can then determine the mimimum time series required to achieve designed levels of significance and power.
+Essentially, we take a stochastic population model and simulate it for a number of years repetitively.
+We can then determine the minimum time series required to achieve designed levels of significance and power.
 This approach requires us to determine values for our model parameters.
@@
 The rate of growth $r$ is also the trend strength of increase or decrease.
-Lastly, $\phi$ represents an autocorrelation term where $\phi=0$ indicates no autocorrelation, $\phi>0$ is postively autocorrelated.
+Lastly, $\phi$ represents an autocorrelation term where $\phi=0$ indicates no autocorrelation, $\phi>0$ is positively autocorrelated.
 This model represents linear growth with autocorrelation and noise.
@@
 
-By simulating this model repetively for different lengths of time, we can determine the mimumum time series required.
+By simulating this model repetitively for different lengths of time, we can determine the minimum time series required.
 Here statistical power is simply the number of trials that are themselves significant, as the real trend is different from zero.
-Then the mimumum time required is the mimumum time required to obtain statistical power above some pre-determined threshold.
+Then the minimum time required is the minimum time required to obtain statistical power above some pre-determined threshold.
 Historically, 0.8....s
@@
 Xb).
-Once power is greater than XX (the dotted line), that is the mimumum time required ($t_{min}$
+Once power is greater than XX (the dotted line), that is the minimum time required ($t_{min}$
 
@@
 
-Our objective is to determine the miniuum time series length ($T_{min}$) required to answer a specific question of interest.
+Our objective is to determine the minimum time series length ($T_{min}$) required to answer a specific question of interest.
 As we describe below, the minimum time series length calculation is particular to each question.
@@
 
-We assume that the time series in our database are long enough to include all necessary information (e.g.
+We assume that each time series in our database are long enough to include all necessary information (e.g.
 variability) about the population.
-In other words, they are a representive sample.
+In other words, they are a Representative sample.
 We first take all possible contiguous subsamples of each time series.
@@
 Therefore, we run a linear regression for each subsampled time series.
-Then, we determine the fraction of subsamples of a particular length that have slopes from linear regression which are statisically different from zero given the longterm, or ``true", time series also has a significant slope.
+Then, we determine the fraction of subsamples of a particular length that have slopes from linear regression which are statistically different from zero given the longterm, or ``true", time series also has a significant slope.
 This is a measure of statistical power.
-Lastly, we determine which subsample length is required to guarentee we have above a certain threshold of statistical power, which here we will assume is 0.8 by convention (citation).
+Lastly, we determine which subsample length is required to guarantee we have above a certain threshold of statistical power, which here we will assume is 0.8 by convention (citation).
 
 We also examine how minimum time series calculations changed depending on the question of interest.
-In the supplementary material, we show how to make similar calculations to determine the minimnm time series required to estimate overall population growth rate, the variability in population size, and future extinction risk.
+In the supplementary material, we show how to make similar calculations to determine the minimum time series required to estimate overall population growth rate, the variability in population size, and future extinction risk.
 
@@
 ## Results
+
+- summary of findings
+- theoretical approach
+  - example plot
+- empirical approach
+  - plots: distribution of min time, what min time correlations with
+  - walk through one empirical example (maybe in appendix)
 
@@
 
-We examined a database of `r nrow(pop_info)` seperate population time series representing `r length(table(pop_info$Binomial))` species.
+We examined a database of `r nrow(pop_info)` separate population time series representing `r length(table(pop_info$Binomial))` species.
 This database consists of species with a variety of life history characteristics.
@@
 
-We also examined a subset of populations where more explantatory variables were known (e.g.
-body mass, longevitiy, location).
+We also examined a subset of populations where more explanatory variables were known (e.g.
+body mass, longevity, location).
 None (need to do regression) of these explanatory variables were predictive of the minimum time series length required.
 Some differences if $t_{min}$ did occur between biological classes, but this can be attributed to the amount populations are know to fluctuate.
-
-<!--
-- What does a theoretically model predict (does min time scale with trend, power, variability, autocorrelation, something else) - definitely depends on what you set for alpha and beta
-- Case studie(s)
-- For all species
-    - Average min time needed
-    - Correlates well with trend, variability, and autocorrelation
-    - Surprisingly does not correlate well with generation time or any other biological variables (need to do proper regression though)-->
 
@@
 ```{r,echo=F,eval=F}
-summary(poisson_model)
+#summary(poisson_model)
 ```
@@
 - Future recommendations to others and future work
+- implications
 - conclusion paragraph
@@
 The IUCN has an 10 year/3 generation rule...
-If we apply this rule to the species in our database we see that approximently `r 100*sum(pop_info$min_time_for_power>pop_info$IUCN)/nrow(pop_info)`% of populations need a longer minimum time series than that required by IUCN.
+If we apply this rule to the species in our database we see that approximately `r 100*sum(pop_info$min_time_for_power>pop_info$IUCN)/nrow(pop_info)`% of populations need a longer minimum time series than that required by IUCN.
+
+### Time series length recommendations from other papers
+
+- greater than 5 years [@Morrison2008]
+- more than 10-15 years [@Rueda-Cediel2015]
+- general advice (variability, correlations) [@Rhodes2011]
+- more than 10 years [@Gerber1999]
+- specific recommendations depend on amount of power and strength of trend (Bart 2004, Hatch 2003) [@Seavy2007]
 
 ## References
-
-# Supplementary material
-
-```{r,echo=F}
-# Example of using population growth rate instead of power
-setwd("~/Desktop/Research/time-series-project")
-source("scripts/calculate_geometric_growth.R")
-
-par(mfrow=c(1,2))
-pop = subset(long_dat,long_dat$ID=='10696') #13192 for Harp seals
-plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='',xlab='',pch=16,ylim=c(0,2000),las=1)
-abline(lm(pop$popvalue~pop$year),lwd=2,col='red')
-mtext(text = 'year',1,line=3,cex=1.4)
-mtext(text = 'population size',2,line=3.5,cex=1.4)
-
-aaa=calculate_average_geometric_growth(pop$popvalue)
-plot(rowMeans(aaa,na.rm=T),cex.axis=1.2,pch=16,ylab='',xlab='',las=1)
-mtext(text = 'years sampled',1,line=3,cex=1.4)
-mtext(text = 'population growth rate',2,line=3.7,cex=1.4)
-```

### 2017-04-21 writing/supp_mat_draft.Rmd
first commit of supp mat draft

---
title: 'Supp mat: time series paper'
author: "Easton R. White"
date: "February 1, 2017"
output: html_document
---

This is where I will build a supplementary material file

## Example of subsamling and power calculations

Here, we illustrate in detail how we performed the subsampling and power calculations for a specific population. As an example, we examine a 35-year time series of (). Simple linear regression indicates a significant decline for this population with a blank and blank. We assume that this significant decline over 35 years is in fact the ``true trend". In statistical jargon, the 35-year trend is an effect that is actually present. We can then use this as a benchmark to see if subsamples of the time series show a similar result.

We first extract all contiguous subsamples of the time series. This leads to 34 two-year subsamples, 33 two-year subsamples, and so forth until a single 35-year subsample. We can call each set of subsamples, of the same length, a set. For each subsample, we conduct linear regression and extract model coefficients and p-values. Then, the fraction of subsamples within a set that show significant trends (signfiiant slope coefficient) is the statistical power. It is important to note that we only consider subsamples to be significant if they are significant in the same direction as the complete 35-year time series.

We can then plot statistical power as a function of time series length. As expected, we can see that power increases with the more years that are sampled.

Then, we determine an appropriate level of statistical power that we find acceptable. Traditionally, this has been at 0.8, however, this is purely historical. Statisical power of 0.8 implies...

With statistical power of 0.8, we then determine the minimum time series length ($t_min$) required to achieve that level of statistical power. Here, $t_min$ is the first point as which all points to the right are above 0.8.

```{r, echo=F}
# show an example
```

## Results from other questions of interest

# Supplementary material

```{r,echo=F}
# Example of using population growth rate instead of power
setwd("~/Desktop/Research/time-series-project")
source("scripts/calculate_geometric_growth.R")

par(mfrow=c(1,2))
pop = subset(long_dat,long_dat$ID=='10696') #13192 for Harp seals
plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='',xlab='',pch=16,ylim=c(0,2000),las=1)
abline(lm(pop$popvalue~pop$year),lwd=2,col='red')
mtext(text = 'year',1,line=3,cex=1.4)
mtext(text = 'population size',2,line=3.5,cex=1.4)

aaa=calculate_average_geometric_growth(pop$popvalue)
plot(rowMeans(aaa,na.rm=T),cex.axis=1.2,pch=16,ylab='',xlab='',las=1)
mtext(text = 'years sampled',1,line=3,cex=1.4)
mtext(text = 'population growth rate',2,line=3.7,cex=1.4)
```

### 2017-05-11 writing/manuscript_draft.Rmd
made changes to formatting and introduction

@@
 ---
-title: "Your time series is (probably) too short"
-author: "Easton R. White"
-date: "22-Apr-2017"
 output:
-  pdf_document: default
-  word_document: default
+  pdf_document:
+    fig_caption: yes
+    keep_tex: yes
+    number_sections: true
+fontsize: 11pt
+geometry: margin=0.75in
 bibliography: White_bib.bib
+header-includes: 
+    \usepackage{float}
+    \usepackage{setspace}             
+    \usepackage{lineno}
+    \usepackage[round]{natbib}
+    \bibpunct[; ]{(}{)}{,}{a}{}{;}
+    \linenumbers
 ---
+
+\begin{spacing}{1.5} %use 2
+	\begin{flushleft}
+Running head: 
+\vspace{3 mm}
+
+Title: Your time series is (probably) too short
+\vspace{7 mm}
+
+Author(s): \textsc{Easton R. White$^{1,2}$} 
+\vspace{3 mm}
+
+Addresses: \\ \emph{$^1$ Center for Population Biology, University of California-Davis, Davis, California 95616 USA}
+\vspace{3 mm}
+
+$^2$Corresponding author: eawhite@ucdavis.edu
+\vspace{3 mm}
+
+Number of words: X,XXX 
+\vspace{3 mm}
+
+To be submitted to: \emph{XXXXX} (Article)
+\vspace{3 mm}
+
+Keywords:
+
+\vspace{3 mm}
+
+\end{flushleft}
+\end{spacing}
+
+\linenumbers
+
+\clearpage
 
@@
 In other words, when is a time series of census data long enough to address a question of interest?
-Here, I explore two approaches to address this question: simulation-based and an empirical approach.
+Here, I explore two approaches to address this question: one simulation-based and an empirical approach.
 I specifically determine the minimum time series length required to estimate significant increases or decreases in population abundance.
@@
 In addition, I examine 868 populations of vertebrate species to determine the minimum time required to detect changes in their abundance.
-I found that 10-15 years of continuous monitoring are required in order to achieve a high level of statistical power.
+Here I show 10-15 years of continuous monitoring are required in order to achieve a high level of statistical power.
 Similar to simulation-approaches, the minimum time required for field census data strongly depends on trend strength, population variability, and temporal autocorrelation.
@@
 
-Observational studies, and censusing populations, has become a cornerstone of modern ecology and conservation biology.
+Observational studies, and censusing populations, has become a cornerstone of modern ecology and conservation biology [@Magurran2010].
 Longterm data is necessary to understand population dynamics and to assess extinction risk of species.
@@
 This question has important scientific and management implications.
-Scientifically, we need to know how long a time series must be in order to rigorously test a specific question.
+Scientifically, we need to know how long a time series is required to address a specific question.
 Too short of a time series may lead to wrong conclusions given large natural year-to-year variability.
 From a management perspective, it is important to understand when a trend in abundance over time is actually meaningful or not.
-For instance the International Union for the Conservation of Nature (IUCN) suggest...
+For instance the International Union for the Conservation of Nature (IUCN) suggest...In addition, because sampling is typically expensive, we do not want to over sample [@Gerber1999].
 
-In determining the number of samples required for a particular experiment X quantities are intricately linked: statistical power, effect size, alpha, sample size [@Legg2006].
-The relationship between these quantities depends on the specific statistical test used.
-Formally, statistical power is the one minus the probability of a type II error, or false negative.
+In determining the number of samples required for a particular experiment four quantities are intricately linked: statistical power, effect size, alpha, sample size [@Legg2006].
+The exact relationship between these quantities depends on the specific statistical test.
+Formally, statistical power is one minus the probability of a type II error, or false negative.
 In other words, it is a measure of how well you could detect a trend given it is actually significant.
@@
 Instead of using classic statistical tests (e.g.
-t-test), and corresponding power calculations, simulation approaches can be used.
-Simple population models can be run with parameter values corresponding to a population of interest.
-Out of a large number of simulations, power can be calculated as simply the fraction of simulations that meet some criteria.
+t-test), and corresponding power calculations, simulation approaches can be used [@Gerrodette1987].
+Simple population models can be simulated with parameter values corresponding to a population of interest.
+Out of a large number of simulations, power can be calculated as the fraction of simulations that meet some criteria.
 The specific criteria depend on the question at hand.
@@
 In this paper, we use both simulations and a database of empirical time series to determine the minimum number of years required to address various questions.
-To demonstrate these techniques, we address three specific questions.
+To demonstrate these techniques, we address three specific questions: (1) what is the minimum time series length required to identify significant changes over time using linear regression, (2)
 <!-- In this paper, we evaluate the minimal time series length required to answer X questions.
@@
 
-- data source
-- empirical approach
-- theoretical approach
-- stuff for the appendix
-
-```{r, echo=F}
-# Pull in cleaned database
-setwd("~/Desktop/Research/time-series-project")
-load('cleaned-data/pop_info_example_goodNAs.Rdata')
-```
-
-### Data source
-We use a database of `r nrow(pop_info)` population time series compiled in [@Keith2015].
-The data is originally from the Global Population Dynamics Database (citation needed).
-The database includes information on `r length(table(pop_info$Binomial))` vertebrate species with a focus on mammals, birds, and fish.
-We filtered out short time series (less than `r pop_info$series_length[1]`) and were left with XX time series with an average length of XX.
-The database also includes information on generation length, census type,.
-For each time series, we also calculated variables of interest like variance in population size, longterm trend in abundance, and temporal autocorrelation.
-We also calculated the minimum number of years required to sample according to IUCN standards under rule ...
-(citation needed).
-
-For a subset of populations in the database, we were able to connect a database on biological characteristics...
-
 ### Theoretical approach
 
-One approach to determining the minimum time series needed is through repetitive simulations of population models (cite old paper).
-This is the same approach one might use in sample size calculations for any experimental design too complicated for simple power analyses.
+One approach to determining the minimum time series needed is through repetitive simulations of a population model [@Gerrodette1987].
+This is the same approach one might use in sample size calculations for any experimental design too complicated for simple power analyses [@Johnson2015].
+We only briefly discuss this approach as it has been desribed elsewhere (citations).
 Essentially, we take a stochastic population model and simulate it for a number of years repetitively.
@@
 
+<!--
 ![optional caption text](example_theoretical_output.pdf)
+-->
 
-```{r,echo=F,eval=F,fig.width=.83}
+```{r,echo=F,fig.cap='theoretical approach\\label{fig:theoretical_approach}'}
 setwd("~/Desktop/Research/time-series-project")
@@
   years_to_sample=seq(2,100,by=1)
-  ptm <- proc.time()
-  y=lapply(X=years_to_sample,FUN = calculate_power,phi = 0.6,r=2,sigma = .2,trials=300)
-  proc.time() - ptm
+  power=lapply(X=years_to_sample,FUN = calculate_power,phi = 0.6,r=0.2,sigma = .5,trials=50)#should use 300 trials
 
-  simulations = replicate(n=30,simulate_pop(phi = 0.6,r=2,sigma = .2,max.time=40))
-  min_time_needed = years_to_sample[tail(which(y<0.8),1)] + 1
+  simulations = replicate(n=30,simulate_pop(phi = 0.6,r=0.2,sigma = .5,max.time=40))
+  min_time_needed = years_to_sample[tail(which(power<0.8),1)] + 1
 
+  # Plotting commands
   par(mfrow=c(1,2),oma=c(1,2,0,0.5),mar=c(4,5,0.5,0.5))
@@
 
-  plot(years_to_sample,y,type='l',lwd=2,las=1,ylab='',xlab='',cex.axis=1.2)
+    plot(years_to_sample,power,type='l',lwd=2,las=1,ylab='',xlab='',cex.axis=1.2)
       abline(v=min_time_needed,lty=2,lwd=2,col='red')
@@
 ```
+
+```{r, echo=F}
+# Pull in original database from Keith et al. 2015 and the cleaned version after our series of filtering steps
+ setwd("~/Desktop/Research/time-series-project")
+#Load keith database
+load('data/Keith_et_al_2015_data.Rdata')
+
+#Load our cleaned data
+load('cleaned-data/pop_info_example_goodNAs.Rdata')
+```
+
+### Data source
+We use a database of `r length(table(dat$ID))` population time series compiled in [@Keith2015].
+The data is originally from the Global Population Dynamics Database [@GPDD2010].
+We filtered out short time series (less than `r pop_info$series_length[1]`) and were left with `r nrow(pop_info)` time.
+The database includes information on `r length(table(pop_info$Binomial))` vertebrate species with a focus on mammals, birds, and fish.
+The database also includes information on generation length and census specifications.
+For each time series, we also calculated variables of interest like variance in population size, longterm trend in abundance (slope coefficient from simple linear regression), and temporal autocorrelation.
+Lastly, we also calculated the minimum number of years required to sample according to IUCN standards under rule ...
+(citation needed).
+
+For a subset of populations in the database (X = XX), we were able to connect a database on biological characteristics like body size and generation length. We used
 
@@
 
-In line with past work (citation), we found the $T_{min}$ increases (i.e. more time is required) with decreases in trend strength or autocorrelation and with increases in variability.
-
-```{r,echo=F,eval=F,include=F}
-# Pull in analysis of average min time needed **theoretical models** (save analysis as Rdata file or something)
-
-```
+In line with past work, we found the $T_{min}$ increases (i.e. more time is required) with decreases in trend strength or autocorrelation and with increases in variability (citation).
 
@@
 
-```{r,echo=F,fig.cap='duh\\label{fig:min_time_dist}'}
+```{r,echo=F,fig.cap='Distribution of the minimum time required in order to detect a significant trend (at the 0.05 level) in abundance given power of 0.8.\\label{fig:fig:min_time_dist}'}
 # Pull in analysis of average min time needed for **empirical species** (save analysis as Rdata file or something)
@@
 par(mfrow=c(2,2),mar=c(5,5,1,2),oma=c(0,0,0.5,0))
-plot(poisson_model,cex.lab=1.2,las=1,cex.axis=1.2,main=' ')
+#plot(poisson_model,cex.lab=1.2,las=1,cex.axis=1.2,main=' ')
 
@@
 
+# Plot minimum time required against trend strength, population variance, and autocorrelation  
   par(mfrow=c(3,1),mar=c(4,1,1,1),oma=c(0,4.5,0,0.5))
@@
 - Extensions to other systems. Useful to use similar species
+- Biological variables not good predictors, but we also used all long-lived vertebrate species
 - Other questions to ask and results in appendix

### 2017-05-11 writing/manuscript_draft.tex
made changes to formatting and introduction

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=0.75in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{}
  \pretitle{\vspace{\droptitle}}
  \posttitle{}
  \author{}
  \preauthor{}\postauthor{}
  \date{}
  \predate{}\postdate{}

\usepackage{float} \usepackage{setspace} \usepackage{lineno}
\usepackage[round]{natbib} \bibpunct[; ]{(}{)}{,}{a}{}{;} \linenumbers

\begin{document}

\begin{spacing}{1.5} %use 2
    \begin{flushleft}
Running head: 
\vspace{3 mm}

Title: Your time series is (probably) too short
\vspace{7 mm}

Author(s): \textsc{Easton R. White$^{1,2}$} 
\vspace{3 mm}

Addresses: \\ \emph{$^1$ Center for Population Biology, University of California-Davis, Davis, California 95616 USA}
\vspace{3 mm}

$^2$Corresponding author: eawhite@ucdavis.edu
\vspace{3 mm}

Number of words: X,XXX 
\vspace{3 mm}

To be submitted to: \emph{XXXXX} (Article)
\vspace{3 mm}

Keywords:

\vspace{3 mm}

\end{flushleft}
\end{spacing}

\linenumbers

\clearpage

\subsection{Abstract}\label{abstract}

Long-term census data is necessary to better understand ecological
processes and make management decisions. In addition, census data is
often expensive, requiring a lot of people-hours or equipment. However,
little work has actually addressed the length of time series required.
In other words, when is a time series of census data long enough to
address a question of interest? Here, I explore two approaches to
address this question: one simulation-based and an empirical approach. I
specifically determine the minimum time series length required to
estimate significant increases or decreases in population abundance.
Importantly, I examine the ability to detect a trend with a set level of
statistical power, which is often neglected in ecological time series
analyses. Using simple simulations, I demonstrate how the minimum time
series length required increases with weaker trends in abundance and
with higher variability in population size. In addition, I examine 868
populations of vertebrate species to determine the minimum time required
to detect changes in their abundance. Here I show 10-15 years of
continuous monitoring are required in order to achieve a high level of
statistical power. Similar to simulation-approaches, the minimum time
required for field census data strongly depends on trend strength,
population variability, and temporal autocorrelation. Perhaps
surprisingly, the minimum time required did not correlate well with
biological explanatory variables, like body size or generation length.
These results point to the importance of sampling populations over long
periods of time. Here, I stress the need for verifying a study has
enough statistical power to accurately determine long-term changes in
population abundance. Most studies, especially those less than 10-15
years, are probably under-powered and potentially misleading.

\subsection{Introduction}\label{introduction}

\begin{itemize}
\tightlist
\item
  intro long time series
\item
  my general question
\item
  primer on power
\item
  more general power interpretation
\item
  2+ approaches
\item
  what we do in this paper
\end{itemize}

Observational studies, and censusing populations, has become a
cornerstone of modern ecology and conservation biology (Magurran et al.
2010). Longterm data is necessary to understand population dynamics and
to assess extinction risk of species. Even though many time series may
now be considered ``longterm'', most are still short (i.e.~less than X
years). \emph{I could have statistic on how long is considered
``longterm'' in papers}. Time series are typically short for a variety
of reasons. They are often coupled with an experiment, which may only
last a couple of years. In addition, short funding cycles make it
difficult to examine populations over longer periods of time. There are
of course notable exceptions including the Isle Royale study, Iowa corn
study, cod, lynx-hair, and many others.

Naturally, one might ask, how long of a time series is actually
necessary? This question has important scientific and management
implications. Scientifically, we need to know how long a time series is
required to address a specific question. Too short of a time series may
lead to wrong conclusions given large natural year-to-year variability.
From a management perspective, it is important to understand when a
trend in abundance over time is actually meaningful or not. For instance
the International Union for the Conservation of Nature (IUCN)
suggest\ldots{}In addition, because sampling is typically expensive, we
do not want to over sample (Gerber, DeMaster, and Kareiva 1999).

In determining the number of samples required for a particular
experiment four quantities are intricately linked: statistical power,
effect size, alpha, sample size (Legg and Nagy 2006). The exact
relationship between these quantities depends on the specific
statistical test. Formally, statistical power is one minus the
probability of a type II error, or false negative. In other words, it is
a measure of how well you could detect a trend given it is actually
significant. Prior to an experiment, one could set appropriate levels of
power, \(\alpha\), and the effect size to estimate the sample size
required for the experiment. This approach, however, is not as simple
for a time series as the data points are clearly non-independent.

For time series data, two general approaches can be used. First, as can
be done with experimental data, we can relax the definition of
statistical power (cite Bolker). Instead of using classic statistical
tests (e.g.~t-test), and corresponding power calculations, simulation
approaches can be used (Gerrodette 1987). Simple population models can
be simulated with parameter values corresponding to a population of
interest. Out of a large number of simulations, power can be calculated
as the fraction of simulations that meet some criteria. The specific
criteria depend on the question at hand. Then one can run simulations
for time series of different lengths to determine the minimum time
series length required to be confident that you will be able to answer
your question. In addition to using simulated data, empirical time
series can also be used. We usually do not have multiple replicates of
similar populations, but we can subsample an empirical time series.
Subsamples of different lengths can then be evaluated to see which
fraction of subsamples meet some criteria, again a measure of
statistical power\ldots{}

Past work has investigated these questions with either theoretical
approaches not using data or by examining a single population. As an
example of the former, (Rhodes and Jonzen 2011) examined..

In this paper, we use both simulations and a database of empirical time
series to determine the minimum number of years required to address
various questions. To demonstrate these techniques, we address three
specific questions: (1) what is the minimum time series length required
to identify significant changes over time using linear regression, (2)

\subsection{Methods}\label{methods}

\subsubsection{Theoretical approach}\label{theoretical-approach}

One approach to determining the minimum time series needed is through
repetitive simulations of a population model (Gerrodette 1987). This is
the same approach one might use in sample size calculations for any
experimental design too complicated for simple power analyses (Johnson
et al. 2015). We only briefly discuss this approach as it has been
desribed elsewhere (citations). Essentially, we take a stochastic
population model and simulate it for a number of years repetitively. We
can then determine the minimum time series required to achieve designed
levels of significance and power. This approach requires us to determine
values for our model parameters. As an example, we can take the
following population model for population size \(N\) at time \(t\): \[
N(t + 1) = N(t) + r(t) \mbox{ with } r(t) \sim N(\phi \mu, \sigma)
\] where \(\epsilon\) is a normally-distributed random noise term with
mean \(\mu\) and standard deviation \(\sigma\). The rate of growth \(r\)
is also the trend strength of increase or decrease. Lastly, \(\phi\)
represents an autocorrelation term where \(\phi=0\) indicates no
autocorrelation, \(\phi>0\) is positively autocorrelated. This model
represents linear growth with autocorrelation and noise. However, it is
certainly possible to use any other population model.

By simulating this model repetitively for different lengths of time, we
can determine the minimum time series required. Here statistical power
is simply the number of trials that are themselves significant, as the
real trend is different from zero. Then the minimum time required is the
minimum time required to obtain statistical power above some
pre-determined threshold. Historically, 0.8\ldots{}.s

In figure X, a number of simulated time series are shown for a set
number of time periods (t=40). It is clear that statistical power
increases quickly with increases in \(t\) (Fig. Xb). Once power is
greater than XX (the dotted line), that is the minimum time required
(\(t_{min}\)

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-1-1.pdf}
\caption{theoretical approach\label{fig:theoretical_approach}}
\end{figure}

\subsubsection{Data source}\label{data-source}

We use a database of 2444 population time series compiled in (Keith et
al. 2015). The data is originally from the Global Population Dynamics
Database (NERC Centre for Population Biology 2010). We filtered out
short time series (less than 35) and were left with 1019 time. The
database includes information on 499 vertebrate species with a focus on
mammals, birds, and fish. The database also includes information on
generation length and census specifications. For each time series, we
also calculated variables of interest like variance in population size,
longterm trend in abundance (slope coefficient from simple linear
regression), and temporal autocorrelation. Lastly, we also calculated
the minimum number of years required to sample according to IUCN
standards under rule \ldots{} (citation needed).

For a subset of populations in the database (X = XX), we were able to
connect a database on biological characteristics like body size and
generation length. We used

\subsubsection{Empirical approach}\label{empirical-approach}

Our objective is to determine the minimum time series length
(\(T_{min}\)) required to answer a specific question of interest. As we
describe below, the minimum time series length calculation is particular
to each question. Intuitively, we might expect a longer minimum time
series length required given more complicated questions.

We assume that each time series in our database are long enough to
include all necessary information (e.g.~variability) about the
population. In other words, they are a Representative sample. We first
take all possible contiguous subsamples of each time series. For
example, a time series of 40 years would have 39 possible contiguous
subsamples of length 2, 38 possible contiguous subsamples of length 3,
and so forth. In this paper, we will try to ask a simple, specific
question. We seek to determine the \(T_{min}\) required to be confident
in an estimate of the slope coefficient from linear regression.
Therefore, we run a linear regression for each subsampled time series.
Then, we determine the fraction of subsamples of a particular length
that have slopes from linear regression which are statistically
different from zero given the longterm, or ``true``, time series also
has a significant slope. This is a measure of statistical power. Lastly,
we determine which subsample length is required to guarantee we have
above a certain threshold of statistical power, which here we will
assume is 0.8 by convention (citation).

We also examine how minimum time series calculations changed depending
on the question of interest. In the supplementary material, we show how
to make similar calculations to determine the minimum time series
required to estimate overall population growth rate, the variability in
population size, and future extinction risk.

In the appendix, we explore other potential questions of
interest\ldots{}

\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-3-1.pdf}

\subsection{Results}\label{results}

\begin{itemize}
\tightlist
\item
  summary of findings
\item
  theoretical approach
\item
  example plot
\item
  empirical approach
\item
  plots: distribution of min time, what min time correlations with
\item
  walk through one empirical example (maybe in appendix)
\end{itemize}

We evaluated the question of what minimum time series is required to
answer a question of interest. More specifically we ask, what is the
minimum time series length required to determine, via linear regression,
the longterm trend in population abundance? The minimum time series
length required is that which had high enough statistical power. We
investigated this question through two approaches: simulation-based and
empirical analyses. In both cases, we found the minimum time series
length required depended strongly on the trend strength, population
variability, and autocorrelation.

\subsubsection{Simulation approach}\label{simulation-approach}

We constructed a general population model where the trend over time
could be a model parameter. We chose a simple model, but any other
population model could be used. The specific model choice should be
tailored to the population of interest, based on species ecology. We
then ran simulations of different lengths to determine the minimum time
series length required to achieve a certain level of power (which we set
to 0.8). We were also able to alter the parameter values (trend strength
and population variability) of the population model to see how this
altered \(T_{min}\).

In line with past work, we found the \(T_{min}\) increases (i.e.~more
time is required) with decreases in trend strength or autocorrelation
and with increases in variability (citation).

\subsubsection{Empirical test}\label{empirical-test}

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{Distribution of the minimum time required in order to detect a
significant trend (at the 0.05 level) in abundance given power of
0.8.\label{fig:fig:min_time_dist}}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{duh\label{fig:correlates}}
\end{figure}

We examined a database of 868 separate population time series
representing 475 species. This database consists of species with a
variety of life history characteristics. We limited our analyses to
populations with at least 35 years of continuous sampling.

Across all the populations we examined, we found an average minimum time
series length required (\(T_{min}\)) of 16.6463134
(\(\sigma\)=8.6664727), with a wide distribution (Fig.
\ref{fig:min_time_dist}). As expected based on theoretical results (cite
papers), the minimum time series length required is strongly correlated
with trend strength, population variance, and autocorrelation (Fig
\ref{fig:correlates}). Using a generalized linear model, with a Poisson
error structure, all three of these explanatory variables were
significant and had large effect sizes (cite model here). Combined trend
strength, population variance, and autocorrelation account for about
70.7793122 of the variance in minimum time series length required (as
proposed by CITE Zuur).

We also examined a subset of populations where more explanatory
variables were known (e.g.~body mass, longevity, location). None (need
to do regression) of these explanatory variables were predictive of the
minimum time series length required. Some differences if \(t_{min}\) did
occur between biological classes, but this can be attributed to the
amount populations are know to fluctuate.

\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-6-1.pdf}

\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-7-1.pdf}

\subsection{Discussion}\label{discussion}

\begin{itemize}
\tightlist
\item
  Findings and how they relate to past work
\item
  Extensions to other systems. Useful to use similar species
\item
  Biological variables not good predictors, but we also used all
  long-lived vertebrate species
\item
  Other questions to ask and results in appendix
\item
  Future recommendations to others and future work
\item
  implications
\item
  conclusion paragraph
\end{itemize}

The IUCN has an 10 year/3 generation rule\ldots{} If we apply this rule
to the species in our database we see that approximately 50.5760369\% of
populations need a longer minimum time series than that required by
IUCN.

\subsubsection{Time series length recommendations from other
papers}\label{time-series-length-recommendations-from-other-papers}

\begin{itemize}
\tightlist
\item
  greater than 5 years (Morrison and Hik 2008)
\item
  more than 10-15 years (Rueda-Cediel et al. 2015)
\item
  general advice (variability, correlations) (Rhodes and Jonzen 2011)
\item
  more than 10 years (Gerber, DeMaster, and Kareiva 1999)
\item
  specific recommendations depend on amount of power and strength of
  trend (Bart 2004, Hatch 2003) (Seavy and Reynolds 2007)
\end{itemize}

\subsection*{References}\label{references}
\addcontentsline{toc}{subsection}{References}

\hypertarget{refs}{}
\hypertarget{ref-Gerber1999}{}
Gerber, L R, D P DeMaster, and P M Kareiva. 1999. ``Gray whales and the
value of monitering data in implementing the U.S. endangered species
act.'' \emph{Conservation Biology} 13 (5): 1215--9.

\hypertarget{ref-Gerrodette1987}{}
Gerrodette, Tim. 1987. ``A power analysis for detecting trends.''
doi:\href{https://doi.org/10.2307/1939220}{10.2307/1939220}.

\hypertarget{ref-Johnson2015}{}
Johnson, Paul C D, Sarah J E Barry, Heather M. Ferguson, and Pie Müller.
2015. ``Power analysis for generalized linear mixed models in ecology
and evolution.'' \emph{Methods in Ecology and Evolution} 6 (2): 133--42.
doi:\href{https://doi.org/10.1111/2041-210X.12306}{10.1111/2041-210X.12306}.

\hypertarget{ref-Keith2015}{}
Keith, David, H. Resit Akçakaya, Stuart H.M. Butchart, Ben Collen,
Nicholas K. Dulvy, Elizabeth E. Holmes, Jeffrey A. Hutchings, et al.
2015. ``Temporal correlations in population trends: Conservation
implications from time-series analysis of diverse animal taxa.''
\emph{Biological Conservation} 192. Elsevier B.V.: 247--57.
doi:\href{https://doi.org/10.1016/j.biocon.2015.09.021}{10.1016/j.biocon.2015.09.021}.

\hypertarget{ref-Legg2006}{}
Legg, Colin J, and Laszlo Nagy. 2006. ``Why most conservation monitoring
is , but need not be , a waste of time.'' \emph{Journal of Environmental
Management} 78: 194--99.
doi:\href{https://doi.org/10.1016/j.jenvman.2005.04.016}{10.1016/j.jenvman.2005.04.016}.

\hypertarget{ref-Magurran2010}{}
Magurran, Anne E, Stephen R Baillie, Stephen T Buckland, Jan Mcp Dick,
David A Elston, E Marian Scott, Rognvald I Smith, Paul J Somerfield, and
Allan D Watt. 2010. ``Long-term datasets in biodiversity research and
monitoring : assessing change in ecological communities through time.''
\emph{Trends in Ecology and Evolution} 25: 574--82.
doi:\href{https://doi.org/10.1016/j.tree.2010.06.016}{10.1016/j.tree.2010.06.016}.

\hypertarget{ref-Morrison2008}{}
Morrison, Sharn, and David S. Hik. 2008. ``When? Where? and for how
long? Census design considerations for an Alpine Lagomorpg, the Collared
pika.'' In \emph{Lagomorph Biology}, 103--13. Springer Berlin
Heidelberg.
doi:\href{https://doi.org/10.1007/978-3-540-72446-9}{10.1007/978-3-540-72446-9}.

\hypertarget{ref-GPDD2010}{}
NERC Centre for Population Biology, Imperial College. 2010. ``The Global
Population Dynamics Database Version 2.''

\hypertarget{ref-Rhodes2011}{}
Rhodes, Jonathan R., and Niclas Jonzen. 2011. ``Monitoring temporal
trends in spatially structured populations: how should sampling effort
be allocated between space and time?'' \emph{Ecography} 34 (6): 1040--8.
doi:\href{https://doi.org/10.1111/j.1600-0587.2011.06370.x}{10.1111/j.1600-0587.2011.06370.x}.

\hypertarget{ref-Rueda-Cediel2015}{}
Rueda-Cediel, Pamela, Kurt E Anderson, Tracey J Regan, Janet Franklin,
and M Regan. 2015. ``Combined Influences of Model Choice , Data Quality
, and Data Quantity When Estimating Population Trends.'' \emph{PLoSONE}
10 (7): e0132255.
doi:\href{https://doi.org/10.1371/journal.pone.0132255}{10.1371/journal.pone.0132255}.

\hypertarget{ref-Seavy2007}{}
Seavy, Nathaniel E., and Michelle H. Reynolds. 2007. ``Is statistical
power to detect trends a good assessment of population monitoring?''
\emph{Biological Conservation} 140 (1-2): 187--91.
doi:\href{https://doi.org/10.1016/j.biocon.2007.08.007}{10.1016/j.biocon.2007.08.007}.

\end{document}

### 2017-05-11 writing/supp_mat_draft.Rmd
made changes to formatting and introduction

@@
 ---
-title: 'Supp mat: time series paper'
-author: "Easton R. White"
-date: "February 1, 2017"
-output: html_document
+output:
+  pdf_document:
+    fig_caption: yes
+    keep_tex: yes
+    number_sections: true
+    citation_package: natbib
+fontsize: 11pt
+geometry: margin=0.75in
+bibliography: White_bib.bib
+header-includes: 
+    \usepackage{float}
+    \renewcommand{\thepage}{S\arabic{page}} 
+    \renewcommand{\thesection}{S\arabic{section}}  
+    \renewcommand{\thetable}{S\arabic{table}}  
+    \renewcommand{\thefigure}{S\arabic{figure}}
 ---
 
-This is where I will build a supplementary material file
+\vspace{2cm}
+
+\begin{center}
+ \textbf{Supplementary Material: Article title here}
+
+Authors: Easton R. White$^{1*}$
+\vspace{3 mm}
+
+Address: \\ \emph{$^1$ Center for Population Biology, University of California-Davis, Davis, California 95616 USA}
+
+*Corresponding author: eawhite@ucdavis.edu
+
+Dec XX, 2016
+ \end{center}
+
+\vspace{2cm}
+
+This supplementary material includes:
+
+1. Mathematical description of our model
+2. Inverse pattern-orientated estimation procedure
+3. Effects of spatial structure and patch heterogeneity
+4. Mainland-Island effect
+5. Invasion from southern area of study site
+6. Parameter sensitivity
+7. Effect of unsampled patches
+8. Long term dynamics
+9. Individual patch dynamics over time
+
+Code for all the figures and tables can be found on [Github](https://github.com/erwhite1).
+
+\vspace{2cm}
 
@@
 
-# Supplementary material
+# Supplementary material \ref{fig:growth_rate}
 
-```{r,echo=F}
+```{r,echo=F,fig.cap='duh\\label{fig:growth_rate}'}
 # Example of using population growth rate instead of power
-setwd("~/Desktop/Research/time-series-project")
-source("scripts/calculate_geometric_growth.R")
+#setwd("~/Desktop/Research/time-series-project")
+#source("scripts/calculate_geometric_growth.R")
 
 par(mfrow=c(1,2))
-pop = subset(long_dat,long_dat$ID=='10696') #13192 for Harp seals
-plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='',xlab='',pch=16,ylim=c(0,2000),las=1)
-abline(lm(pop$popvalue~pop$year),lwd=2,col='red')
-mtext(text = 'year',1,line=3,cex=1.4)
-mtext(text = 'population size',2,line=3.5,cex=1.4)
+#pop = subset(long_dat,long_dat$ID=='10696') #13192 for Harp seals
+#plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='',xlab='',pch=16,ylim=c(0,2000),las=1)
+#abline(lm(pop$popvalue~pop$year),lwd=2,col='red')
+#mtext(text = 'year',1,line=3,cex=1.4)
+#mtext(text = 'population size',2,line=3.5,cex=1.4)
+plot(1,1)
 
-aaa=calculate_average_geometric_growth(pop$popvalue)
-plot(rowMeans(aaa,na.rm=T),cex.axis=1.2,pch=16,ylab='',xlab='',las=1)
-mtext(text = 'years sampled',1,line=3,cex=1.4)
-mtext(text = 'population growth rate',2,line=3.7,cex=1.4)
+#aaa=calculate_average_geometric_growth(pop$popvalue)
+#plot(rowMeans(aaa,na.rm=T),cex.axis=1.2,pch=16,ylab='',xlab='',las=1)
+#mtext(text = 'years sampled',1,line=3,cex=1.4)
+#mtext(text = 'population growth rate',2,line=3.7,cex=1.4)
 ```

### 2017-05-11 writing/supp_mat_draft.tex
made changes to formatting and introduction

--- 2017-05-11 writing/manuscript_draft.tex
@@
 \urlstyle{same}  % don't use monospace font for urls
+\usepackage{natbib}
+\bibliographystyle{plainnat}
 \usepackage{graphicx,grffile}
@@
 
-\usepackage{float} \usepackage{setspace} \usepackage{lineno}
-\usepackage[round]{natbib} \bibpunct[; ]{(}{)}{,}{a}{}{;} \linenumbers
+\usepackage{float} \renewcommand{\thepage}{S\arabic{page}}
+\renewcommand{\thesection}{S\arabic{section}}
+\renewcommand{\thetable}{S\arabic{table}}
+\renewcommand{\thefigure}{S\arabic{figure}}
 
@@
 
-\begin{spacing}{1.5} %use 2
-    \begin{flushleft}
-Running head: 
+\vspace{2cm}
+
+\begin{center}
+ \textbf{Supplementary Material: Article title here}
+
+Authors: Easton R. White$^{1*}$ , John D. Nagy$^{2,3}$, and Andrew T. Smith$^4$ 
 \vspace{3 mm}
 
-Title: Your time series is (probably) too short
-\vspace{7 mm}
+Address: \\ \emph{$^1$ Center for Population Biology, University of California-Davis, Davis, California 95616 USA; \\ $^2$ Department of Life Science, Scottsdale Community College, Scottsdale, Arizona 85253 USA; \\ $^3$ School of Mathematical and Statistical Sciences, Arizona State University, Tempe, Arizona 85287 USA; \\ $^4$ School of Life Sciences, Arizona State University, Tempe, Arizona 85287 USA}
 
-Author(s): \textsc{Easton R. White$^{1,2}$} 
-\vspace{3 mm}
+*Corresponding author: eawhite@ucdavis.edu
 
-Addresses: \\ \emph{$^1$ Center for Population Biology, University of California-Davis, Davis, California 95616 USA}
-\vspace{3 mm}
+Dec XX, 2016
+ \end{center}
 
-$^2$Corresponding author: eawhite@ucdavis.edu
-\vspace{3 mm}
+\vspace{2cm}
 
-Number of words: X,XXX 
-\vspace{3 mm}
+This supplementary material includes:
 
-To be submitted to: \emph{XXXXX} (Article)
-\vspace{3 mm}
-
-Keywords:
-
-\vspace{3 mm}
-
-\end{flushleft}
-\end{spacing}
-
-\linenumbers
-
-\clearpage
-
-\subsection{Abstract}\label{abstract}
-
-Long-term census data is necessary to better understand ecological
-processes and make management decisions. In addition, census data is
-often expensive, requiring a lot of people-hours or equipment. However,
-little work has actually addressed the length of time series required.
-In other words, when is a time series of census data long enough to
-address a question of interest? Here, I explore two approaches to
-address this question: one simulation-based and an empirical approach. I
-specifically determine the minimum time series length required to
-estimate significant increases or decreases in population abundance.
-Importantly, I examine the ability to detect a trend with a set level of
-statistical power, which is often neglected in ecological time series
-analyses. Using simple simulations, I demonstrate how the minimum time
-series length required increases with weaker trends in abundance and
-with higher variability in population size. In addition, I examine 868
-populations of vertebrate species to determine the minimum time required
-to detect changes in their abundance. Here I show 10-15 years of
-continuous monitoring are required in order to achieve a high level of
-statistical power. Similar to simulation-approaches, the minimum time
-required for field census data strongly depends on trend strength,
-population variability, and temporal autocorrelation. Perhaps
-surprisingly, the minimum time required did not correlate well with
-biological explanatory variables, like body size or generation length.
-These results point to the importance of sampling populations over long
-periods of time. Here, I stress the need for verifying a study has
-enough statistical power to accurately determine long-term changes in
-population abundance. Most studies, especially those less than 10-15
-years, are probably under-powered and potentially misleading.
-
-\subsection{Introduction}\label{introduction}
-
-\begin{itemize}
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
 \tightlist
 \item
-  intro long time series
+  Mathematical description of our model
 \item
-  my general question
+  Inverse pattern-orientated estimation procedure
 \item
-  primer on power
+  Effects of spatial structure and patch heterogeneity
 \item
-  more general power interpretation
+  Mainland-Island effect
 \item
-  2+ approaches
+  Invasion from southern area of study site
 \item
-  what we do in this paper
-\end{itemize}
+  Parameter sensitivity
+\item
+  Effect of unsampled patches
+\item
+  Long term dynamics
+\item
+  Individual patch dynamics over time
+\end{enumerate}
 
-Observational studies, and censusing populations, has become a
-cornerstone of modern ecology and conservation biology (Magurran et al.
-2010). Longterm data is necessary to understand population dynamics and
-to assess extinction risk of species. Even though many time series may
-now be considered ``longterm'', most are still short (i.e.~less than X
-years). \emph{I could have statistic on how long is considered
-``longterm'' in papers}. Time series are typically short for a variety
-of reasons. They are often coupled with an experiment, which may only
-last a couple of years. In addition, short funding cycles make it
-difficult to examine populations over longer periods of time. There are
-of course notable exceptions including the Isle Royale study, Iowa corn
-study, cod, lynx-hair, and many others.
+Code for all the figures and tables can be found on
+\href{https://github.com/erwhite1}{Github}.
 
-Naturally, one might ask, how long of a time series is actually
-necessary? This question has important scientific and management
-implications. Scientifically, we need to know how long a time series is
-required to address a specific question. Too short of a time series may
-lead to wrong conclusions given large natural year-to-year variability.
-From a management perspective, it is important to understand when a
-trend in abundance over time is actually meaningful or not. For instance
-the International Union for the Conservation of Nature (IUCN)
-suggest\ldots{}In addition, because sampling is typically expensive, we
-do not want to over sample (Gerber, DeMaster, and Kareiva 1999).
+\vspace{2cm}
 
-In determining the number of samples required for a particular
-experiment four quantities are intricately linked: statistical power,
-effect size, alpha, sample size (Legg and Nagy 2006). The exact
-relationship between these quantities depends on the specific
-statistical test. Formally, statistical power is one minus the
-probability of a type II error, or false negative. In other words, it is
-a measure of how well you could detect a trend given it is actually
-significant. Prior to an experiment, one could set appropriate levels of
-power, \(\alpha\), and the effect size to estimate the sample size
-required for the experiment. This approach, however, is not as simple
-for a time series as the data points are clearly non-independent.
+\subsection{Example of subsamling and power
+calculations}\label{example-of-subsamling-and-power-calculations}
 
-For time series data, two general approaches can be used. First, as can
-be done with experimental data, we can relax the definition of
-statistical power (cite Bolker). Instead of using classic statistical
-tests (e.g.~t-test), and corresponding power calculations, simulation
-approaches can be used (Gerrodette 1987). Simple population models can
-be simulated with parameter values corresponding to a population of
-interest. Out of a large number of simulations, power can be calculated
-as the fraction of simulations that meet some criteria. The specific
-criteria depend on the question at hand. Then one can run simulations
-for time series of different lengths to determine the minimum time
-series length required to be confident that you will be able to answer
-your question. In addition to using simulated data, empirical time
-series can also be used. We usually do not have multiple replicates of
-similar populations, but we can subsample an empirical time series.
-Subsamples of different lengths can then be evaluated to see which
-fraction of subsamples meet some criteria, again a measure of
-statistical power\ldots{}
+Here, we illustrate in detail how we performed the subsampling and power
+calculations for a specific population. As an example, we examine a
+35-year time series of (). Simple linear regression indicates a
+significant decline for this population with a blank and blank. We
+assume that this significant decline over 35 years is in fact the ``true
+trend``. In statistical jargon, the 35-year trend is an effect that is
+actually present. We can then use this as a benchmark to see if
+subsamples of the time series show a similar result.
 
-Past work has investigated these questions with either theoretical
-approaches not using data or by examining a single population. As an
-example of the former, (Rhodes and Jonzen 2011) examined..
+We first extract all contiguous subsamples of the time series. This
+leads to 34 two-year subsamples, 33 two-year subsamples, and so forth
+until a single 35-year subsample. We can call each set of subsamples, of
+the same length, a set. For each subsample, we conduct linear regression
+and extract model coefficients and p-values. Then, the fraction of
+subsamples within a set that show significant trends (signfiiant slope
+coefficient) is the statistical power. It is important to note that we
+only consider subsamples to be significant if they are significant in
+the same direction as the complete 35-year time series.
 
-In this paper, we use both simulations and a database of empirical time
-series to determine the minimum number of years required to address
-various questions. To demonstrate these techniques, we address three
-specific questions: (1) what is the minimum time series length required
-to identify significant changes over time using linear regression, (2)
+We can then plot statistical power as a function of time series length.
+As expected, we can see that power increases with the more years that
+are sampled.
 
-\subsection{Methods}\label{methods}
+Then, we determine an appropriate level of statistical power that we
+find acceptable. Traditionally, this has been at 0.8, however, this is
+purely historical. Statisical power of 0.8 implies\ldots{}
 
-\subsubsection{Theoretical approach}\label{theoretical-approach}
+With statistical power of 0.8, we then determine the minimum time series
+length (\(t_min\)) required to achieve that level of statistical power.
+Here, \(t_min\) is the first point as which all points to the right are
+above 0.8.
 
-One approach to determining the minimum time series needed is through
-repetitive simulations of a population model (Gerrodette 1987). This is
-the same approach one might use in sample size calculations for any
-experimental design too complicated for simple power analyses (Johnson
-et al. 2015). We only briefly discuss this approach as it has been
-desribed elsewhere (citations). Essentially, we take a stochastic
-population model and simulate it for a number of years repetitively. We
-can then determine the minimum time series required to achieve designed
-levels of significance and power. This approach requires us to determine
-values for our model parameters. As an example, we can take the
-following population model for population size \(N\) at time \(t\): \[
-N(t + 1) = N(t) + r(t) \mbox{ with } r(t) \sim N(\phi \mu, \sigma)
-\] where \(\epsilon\) is a normally-distributed random noise term with
-mean \(\mu\) and standard deviation \(\sigma\). The rate of growth \(r\)
-is also the trend strength of increase or decrease. Lastly, \(\phi\)
-represents an autocorrelation term where \(\phi=0\) indicates no
-autocorrelation, \(\phi>0\) is positively autocorrelated. This model
-represents linear growth with autocorrelation and noise. However, it is
-certainly possible to use any other population model.
+\subsection{Results from other questions of
+interest}\label{results-from-other-questions-of-interest}
 
-By simulating this model repetitively for different lengths of time, we
-can determine the minimum time series required. Here statistical power
-is simply the number of trials that are themselves significant, as the
-real trend is different from zero. Then the minimum time required is the
-minimum time required to obtain statistical power above some
-pre-determined threshold. Historically, 0.8\ldots{}.s
-
-In figure X, a number of simulated time series are shown for a set
-number of time periods (t=40). It is clear that statistical power
-increases quickly with increases in \(t\) (Fig. Xb). Once power is
-greater than XX (the dotted line), that is the minimum time required
-(\(t_{min}\)
+\section{\texorpdfstring{Supplementary material
+\ref{fig:growth_rate}}{Supplementary material }}\label{supplementary-material}
 
@@
 \centering
-\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-1-1.pdf}
-\caption{theoretical approach\label{fig:theoretical_approach}}
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-2-1.pdf}
+\caption{duh\label{fig:growth_rate}}
 \end{figure}
 
-\subsubsection{Data source}\label{data-source}
-
-We use a database of 2444 population time series compiled in (Keith et
-al. 2015). The data is originally from the Global Population Dynamics
-Database (NERC Centre for Population Biology 2010). We filtered out
-short time series (less than 35) and were left with 1019 time. The
-database includes information on 499 vertebrate species with a focus on
-mammals, birds, and fish. The database also includes information on
-generation length and census specifications. For each time series, we
-also calculated variables of interest like variance in population size,
-longterm trend in abundance (slope coefficient from simple linear
-regression), and temporal autocorrelation. Lastly, we also calculated
-the minimum number of years required to sample according to IUCN
-standards under rule \ldots{} (citation needed).
-
-For a subset of populations in the database (X = XX), we were able to
-connect a database on biological characteristics like body size and
-generation length. We used
-
-\subsubsection{Empirical approach}\label{empirical-approach}
-
-Our objective is to determine the minimum time series length
-(\(T_{min}\)) required to answer a specific question of interest. As we
-describe below, the minimum time series length calculation is particular
-to each question. Intuitively, we might expect a longer minimum time
-series length required given more complicated questions.
-
-We assume that each time series in our database are long enough to
-include all necessary information (e.g.~variability) about the
-population. In other words, they are a Representative sample. We first
-take all possible contiguous subsamples of each time series. For
-example, a time series of 40 years would have 39 possible contiguous
-subsamples of length 2, 38 possible contiguous subsamples of length 3,
-and so forth. In this paper, we will try to ask a simple, specific
-question. We seek to determine the \(T_{min}\) required to be confident
-in an estimate of the slope coefficient from linear regression.
-Therefore, we run a linear regression for each subsampled time series.
-Then, we determine the fraction of subsamples of a particular length
-that have slopes from linear regression which are statistically
-different from zero given the longterm, or ``true``, time series also
-has a significant slope. This is a measure of statistical power. Lastly,
-we determine which subsample length is required to guarantee we have
-above a certain threshold of statistical power, which here we will
-assume is 0.8 by convention (citation).
-
-We also examine how minimum time series calculations changed depending
-on the question of interest. In the supplementary material, we show how
-to make similar calculations to determine the minimum time series
-required to estimate overall population growth rate, the variability in
-population size, and future extinction risk.
-
-In the appendix, we explore other potential questions of
-interest\ldots{}
-
-\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-3-1.pdf}
-
-\subsection{Results}\label{results}
-
-\begin{itemize}
-\tightlist
-\item
-  summary of findings
-\item
-  theoretical approach
-\item
-  example plot
-\item
-  empirical approach
-\item
-  plots: distribution of min time, what min time correlations with
-\item
-  walk through one empirical example (maybe in appendix)
-\end{itemize}
-
-We evaluated the question of what minimum time series is required to
-answer a question of interest. More specifically we ask, what is the
-minimum time series length required to determine, via linear regression,
-the longterm trend in population abundance? The minimum time series
-length required is that which had high enough statistical power. We
-investigated this question through two approaches: simulation-based and
-empirical analyses. In both cases, we found the minimum time series
-length required depended strongly on the trend strength, population
-variability, and autocorrelation.
-
-\subsubsection{Simulation approach}\label{simulation-approach}
-
-We constructed a general population model where the trend over time
-could be a model parameter. We chose a simple model, but any other
-population model could be used. The specific model choice should be
-tailored to the population of interest, based on species ecology. We
-then ran simulations of different lengths to determine the minimum time
-series length required to achieve a certain level of power (which we set
-to 0.8). We were also able to alter the parameter values (trend strength
-and population variability) of the population model to see how this
-altered \(T_{min}\).
-
-In line with past work, we found the \(T_{min}\) increases (i.e.~more
-time is required) with decreases in trend strength or autocorrelation
-and with increases in variability (citation).
-
-\subsubsection{Empirical test}\label{empirical-test}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-4-1.pdf}
-\caption{Distribution of the minimum time required in order to detect a
-significant trend (at the 0.05 level) in abundance given power of
-0.8.\label{fig:fig:min_time_dist}}
-\end{figure}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
-\caption{duh\label{fig:correlates}}
-\end{figure}
-
-We examined a database of 868 separate population time series
-representing 475 species. This database consists of species with a
-variety of life history characteristics. We limited our analyses to
-populations with at least 35 years of continuous sampling.
-
-Across all the populations we examined, we found an average minimum time
-series length required (\(T_{min}\)) of 16.6463134
-(\(\sigma\)=8.6664727), with a wide distribution (Fig.
-\ref{fig:min_time_dist}). As expected based on theoretical results (cite
-papers), the minimum time series length required is strongly correlated
-with trend strength, population variance, and autocorrelation (Fig
-\ref{fig:correlates}). Using a generalized linear model, with a Poisson
-error structure, all three of these explanatory variables were
-significant and had large effect sizes (cite model here). Combined trend
-strength, population variance, and autocorrelation account for about
-70.7793122 of the variance in minimum time series length required (as
-proposed by CITE Zuur).
-
-We also examined a subset of populations where more explanatory
-variables were known (e.g.~body mass, longevity, location). None (need
-to do regression) of these explanatory variables were predictive of the
-minimum time series length required. Some differences if \(t_{min}\) did
-occur between biological classes, but this can be attributed to the
-amount populations are know to fluctuate.
-
-\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-6-1.pdf}
-
-\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-7-1.pdf}
-
-\subsection{Discussion}\label{discussion}
-
-\begin{itemize}
-\tightlist
-\item
-  Findings and how they relate to past work
-\item
-  Extensions to other systems. Useful to use similar species
-\item
-  Biological variables not good predictors, but we also used all
-  long-lived vertebrate species
-\item
-  Other questions to ask and results in appendix
-\item
-  Future recommendations to others and future work
-\item
-  implications
-\item
-  conclusion paragraph
-\end{itemize}
-
-The IUCN has an 10 year/3 generation rule\ldots{} If we apply this rule
-to the species in our database we see that approximately 50.5760369\% of
-populations need a longer minimum time series than that required by
-IUCN.
-
-\subsubsection{Time series length recommendations from other
-papers}\label{time-series-length-recommendations-from-other-papers}
-
-\begin{itemize}
-\tightlist
-\item
-  greater than 5 years (Morrison and Hik 2008)
-\item
-  more than 10-15 years (Rueda-Cediel et al. 2015)
-\item
-  general advice (variability, correlations) (Rhodes and Jonzen 2011)
-\item
-  more than 10 years (Gerber, DeMaster, and Kareiva 1999)
-\item
-  specific recommendations depend on amount of power and strength of
-  trend (Bart 2004, Hatch 2003) (Seavy and Reynolds 2007)
-\end{itemize}
-
-\subsection*{References}\label{references}
-\addcontentsline{toc}{subsection}{References}
-
-\hypertarget{refs}{}
-\hypertarget{ref-Gerber1999}{}
-Gerber, L R, D P DeMaster, and P M Kareiva. 1999. ``Gray whales and the
-value of monitering data in implementing the U.S. endangered species
-act.'' \emph{Conservation Biology} 13 (5): 1215--9.
-
-\hypertarget{ref-Gerrodette1987}{}
-Gerrodette, Tim. 1987. ``A power analysis for detecting trends.''
-doi:\href{https://doi.org/10.2307/1939220}{10.2307/1939220}.
-
-\hypertarget{ref-Johnson2015}{}
-Johnson, Paul C D, Sarah J E Barry, Heather M. Ferguson, and Pie Müller.
-2015. ``Power analysis for generalized linear mixed models in ecology
-and evolution.'' \emph{Methods in Ecology and Evolution} 6 (2): 133--42.
-doi:\href{https://doi.org/10.1111/2041-210X.12306}{10.1111/2041-210X.12306}.
-
-\hypertarget{ref-Keith2015}{}
-Keith, David, H. Resit Akçakaya, Stuart H.M. Butchart, Ben Collen,
-Nicholas K. Dulvy, Elizabeth E. Holmes, Jeffrey A. Hutchings, et al.
-2015. ``Temporal correlations in population trends: Conservation
-implications from time-series analysis of diverse animal taxa.''
-\emph{Biological Conservation} 192. Elsevier B.V.: 247--57.
-doi:\href{https://doi.org/10.1016/j.biocon.2015.09.021}{10.1016/j.biocon.2015.09.021}.
-
-\hypertarget{ref-Legg2006}{}
-Legg, Colin J, and Laszlo Nagy. 2006. ``Why most conservation monitoring
-is , but need not be , a waste of time.'' \emph{Journal of Environmental
-Management} 78: 194--99.
-doi:\href{https://doi.org/10.1016/j.jenvman.2005.04.016}{10.1016/j.jenvman.2005.04.016}.
-
-\hypertarget{ref-Magurran2010}{}
-Magurran, Anne E, Stephen R Baillie, Stephen T Buckland, Jan Mcp Dick,
-David A Elston, E Marian Scott, Rognvald I Smith, Paul J Somerfield, and
-Allan D Watt. 2010. ``Long-term datasets in biodiversity research and
-monitoring : assessing change in ecological communities through time.''
-\emph{Trends in Ecology and Evolution} 25: 574--82.
-doi:\href{https://doi.org/10.1016/j.tree.2010.06.016}{10.1016/j.tree.2010.06.016}.
-
-\hypertarget{ref-Morrison2008}{}
-Morrison, Sharn, and David S. Hik. 2008. ``When? Where? and for how
-long? Census design considerations for an Alpine Lagomorpg, the Collared
-pika.'' In \emph{Lagomorph Biology}, 103--13. Springer Berlin
-Heidelberg.
-doi:\href{https://doi.org/10.1007/978-3-540-72446-9}{10.1007/978-3-540-72446-9}.
-
-\hypertarget{ref-GPDD2010}{}
-NERC Centre for Population Biology, Imperial College. 2010. ``The Global
-Population Dynamics Database Version 2.''
-
-\hypertarget{ref-Rhodes2011}{}
-Rhodes, Jonathan R., and Niclas Jonzen. 2011. ``Monitoring temporal
-trends in spatially structured populations: how should sampling effort
-be allocated between space and time?'' \emph{Ecography} 34 (6): 1040--8.
-doi:\href{https://doi.org/10.1111/j.1600-0587.2011.06370.x}{10.1111/j.1600-0587.2011.06370.x}.
-
-\hypertarget{ref-Rueda-Cediel2015}{}
-Rueda-Cediel, Pamela, Kurt E Anderson, Tracey J Regan, Janet Franklin,
-and M Regan. 2015. ``Combined Influences of Model Choice , Data Quality
-, and Data Quantity When Estimating Population Trends.'' \emph{PLoSONE}
-10 (7): e0132255.
-doi:\href{https://doi.org/10.1371/journal.pone.0132255}{10.1371/journal.pone.0132255}.
-
-\hypertarget{ref-Seavy2007}{}
-Seavy, Nathaniel E., and Michelle H. Reynolds. 2007. ``Is statistical
-power to detect trends a good assessment of population monitoring?''
-\emph{Biological Conservation} 140 (1-2): 187--91.
-doi:\href{https://doi.org/10.1016/j.biocon.2007.08.007}{10.1016/j.biocon.2007.08.007}.
+\bibliography{White_bib.bib}
 

### 2017-05-11 scripts/theoretical-models/simulate_pop.R
modified simulate_pop function to not run off a single random seed

@@ -8,7 +8,7 @@
 # For presentation, I could leave out autocorrelation results
 
 simulate_pop = function(r,phi,sigma,max.time){
-  set.seed(12345)
+  #set.seed(12345)
   pop = matrix(1000,nrow=1,ncol=max.time)
     for (t in 1:(max.time-1)){
       #pop[t + 1] = pop[t] + rnorm(n = 1,mean=r*phi,sd = sigma)

### 2017-05-11 writing/supp_mat_draft.Rmd
updated supp mat with outline and r code

--- 2017-04-21 writing/supp_mat_draft.Rmd
@@
 ---
-title: 'Supp mat: time series paper'
-author: "Easton R. White"
-date: "February 1, 2017"
-output: html_document
+output:
+  pdf_document:
+    fig_caption: yes
+    keep_tex: yes
+    number_sections: true
+    citation_package: natbib
+fontsize: 12pt
+geometry: margin=0.75in
+bibliography: White_bib.bib
+header-includes: 
+    \usepackage{float}
+    \renewcommand{\thepage}{S\arabic{page}} 
+    \renewcommand{\thesection}{S\arabic{section}}  
+    \renewcommand{\thetable}{S\arabic{table}}  
+    \renewcommand{\thefigure}{S\arabic{figure}}
 ---
 
-This is where I will build a supplementary material file
+\vspace{2cm}
 
-## Example of subsamling and power calculations
+\begin{center}
+ \textbf{Supplementary Material: Article title here}
+
+Authors: Easton R. White$^{1*}$
+\vspace{3 mm}
+
+Address: \\ \emph{$^1$ Center for Population Biology, University of California-Davis, Davis, California 95616 USA}
+
+*Corresponding author: eawhite@ucdavis.edu
+
+Dec XX, 2016
+ \end{center}
+
+\vspace{2cm}
+
+This supplementary material includes:
+
+1. Mathematical description of our model
+2. Inverse pattern-orientated estimation procedure
+
+Code for all the figures and tables can be found on [Github](https://github.com/erwhite1).
+
+\vspace{2cm}
+
+## Detailed example of subsampling and power calculations
 
@@
 ```{r, echo=F}
-# show an example
+# show an example of subsampling and power calculation routines
 ```
@@
 
-# Supplementary material
+### Growth rate calculations
 
-```{r,echo=F}
-# Example of using population growth rate instead of power
-setwd("~/Desktop/Research/time-series-project")
+```{r, echo=F, fig.cap='duh\\label{fig:growth_rate}'}
+#Example of using population growth rate instead of power
+setwd("/Users/eastonwhite/Desktop/Research/time-series-project")
 source("scripts/calculate_geometric_growth.R")
+load('cleaned-data/pop_info_example_goodNAs.Rdata')
 
@@
 
+  source('scripts/create_subsamples.R')
   aaa=calculate_average_geometric_growth(pop$popvalue)
@@
 ```
+
+### Using Generalized additive model to identify significant trends
+
+TEXT HERE ON IDEA
+
+```{r,echo=F,fig.cap='results here from GAM.\\label{fig:gam_example}'}
+# show gam example side-by-side plot
+```
+
+```{r,echo=F,fig.cap='results here from GAM.\\label{fig:gam_result}'}
+# show gam results for minimium time required
+```
+
+### References

### 2017-05-11 writing/supp_mat_draft.tex
updated supp mat with outline and r code

@@
-\documentclass[11pt,]{article}
+\documentclass[12pt,]{article}
 \usepackage{lmodern}
@@
 
-Authors: Easton R. White$^{1*}$ , John D. Nagy$^{2,3}$, and Andrew T. Smith$^4$ 
+Authors: Easton R. White$^{1*}$
 \vspace{3 mm}
 
-Address: \\ \emph{$^1$ Center for Population Biology, University of California-Davis, Davis, California 95616 USA; \\ $^2$ Department of Life Science, Scottsdale Community College, Scottsdale, Arizona 85253 USA; \\ $^3$ School of Mathematical and Statistical Sciences, Arizona State University, Tempe, Arizona 85287 USA; \\ $^4$ School of Life Sciences, Arizona State University, Tempe, Arizona 85287 USA}
+Address: \\ \emph{$^1$ Center for Population Biology, University of California-Davis, Davis, California 95616 USA}
 
@@
   Inverse pattern-orientated estimation procedure
-\item
-  Effects of spatial structure and patch heterogeneity
-\item
-  Mainland-Island effect
-\item
-  Invasion from southern area of study site
-\item
-  Parameter sensitivity
-\item
-  Effect of unsampled patches
-\item
-  Long term dynamics
-\item
-  Individual patch dynamics over time
 \end{enumerate}
@@
 
-\subsection{Example of subsamling and power
-calculations}\label{example-of-subsamling-and-power-calculations}
+\subsection{Detailed example of subsampling and power
+calculations}\label{detailed-example-of-subsampling-and-power-calculations}
 
@@
 
-\section{\texorpdfstring{Supplementary material
-\ref{fig:growth_rate}}{Supplementary material }}\label{supplementary-material}
+\subsubsection{Growth rate calculations}\label{growth-rate-calculations}
+
+\begin{verbatim}
+## Loading required package: plyr
+\end{verbatim}
 
@@
 
+\subsubsection{Using Generalized additive model to identify significant
+trends}\label{using-generalized-additive-model-to-identify-significant-trends}
+
+TEXT HERE ON IDEA
+
+\subsubsection{References}\label{references}
+
 \bibliography{White_bib.bib}

### 2017-05-12 writing/manuscript_draft.Rmd
first draft of the discussion section

@@
 ## Discussion
+
+<!-- 
 - Findings and how they relate to past work
@@
 If we apply this rule to the species in our database we see that approximately `r 100*sum(pop_info$min_time_for_power>pop_info$IUCN)/nrow(pop_info)`% of populations need a longer minimum time series than that required by IUCN.
+-->
+
+We show two different approaches to estimate the minimum time series length required required to address a particular question of interest.
+The simulation-based approach has been suggested by others, especially in situations where analysis is more complicated than a simple power analysis [@Gerrodette1987].
+We focus on a empirical approach to estimate the minimum time series length requried.
+
+The first question we asked was, what is $t_{min}$ for determining whether a population was increasing or decreasing over time according to linear regression.
+We found that for `r nrow(pop_info)` populations, at least 10-15 years of continuous monitoring were neccessary.
+In line with theoretical predictions (citations), we also found $t_{min}$ was strongly correlated with the trend strength, variability in population size, and temporal autocorrelation.
+
+There have been previous attempts to address how long a time series should be to assess trends in abundance. For example,...
+
+### Limitations
+
+Our work has obvious limitations in determining the minimum time series length required.
+First, $t_{min}$ depends strongly on the question being asked and the availability of long-term census data.
+For our empirical approach, the subsampling of the full time series allows for estimates of power, but the individual subsamples are clearly not independent of one another.
+(sentence about alpha and power limititations here) The perfect scenario would be to have a population model built and parameterized for each population of interest.
+Then, model simulations could be used to estimate the minimum time series reuqired to address a specific question of interest.
+Clearly, this is not always practical, especially if running analyses for a wide array of species as we do here.
+In addition, our statistical models suggest that $t_{min}$ does not correlate with any biological variables of interest, at least for the question of linear regression.
+Therefore, it is not possible to use these results to predict $t_{min}$ for another population, even if the population is of a species with a similar life-history to one in our database.
+
+### Conclusions
+
+- what we found, implications, future work
+
+We use a database of `r nrow(pop_info)` populations to determine the minimum time series length required.
+We show that to assess longterm trends in abundance, 10-15 years of continuos monitoring are required.
+In line with theoretical predictions, we also show that $t_{min}$ is strongly correlated with the overall trend in abundance, variability in abundance, and the temporal autocorrelation.
+Our work implies that for many populations, census data less than 10-15 years are probably not reliable.
+This stresses the importance of longterm monitoring programs.
+(where do IUCN recommendations come from?) From both a scientific and management perspective estimates of $t_{min}$ are important.
+If a time series is too short, we are likely to make incorrect conclusions about longterm trends.
+In addition, a time series that is too long may have been a poor use of scare funds.
+Future work can examine other species, with a wider range of life history characteristics, or address other questions besides longterm trends in abundance.
 

### 2017-05-13 writing/manuscript_draft.Rmd
added to discussion section

@@
 - Other questions to ask and results in appendix
+
 - Future recommendations to others and future work
@@
 
+We also found that $t_{min}$ did not correlate with any biological variables of interest (cite figure or supp mat).
+We initially hypothesized that species with longer lifespans or generation times may require more years to sample.
+(Our result could have been the effect of two reasons) First, the database we used may not include a diverse enough set of species with different life history traits.
+Second, the question we poised, whether a population is increasing or decreases, is only worried about population surveys.
+Therefore, biological variables may be more important if the question itself was different...
+
+### Other questions addressed
+
+In the appendix we asked two additional questions to make estimates of $t_{min}$. First, we asked how many years of sampling are neccessary to estimate a geometric growth rate?
+
+### Past work on this issue
 There have been previous attempts to address how long a time series should be to assess trends in abundance. For example,...
+
+An important related question, is the optimal allocation of sampling effort in space versus time (citations).
+In a theoretical investigation of this question, @Rhodes2011 found that the optimal allocation of sampling depended strongly on temporal and spatial autocorrelation.
+If spatial population dynamics were highly correlated, then it was better to sample more temporally, and vice versa.
+Our work supports this idea as populations with strong temporal autocorrelation needed less years of sampling (cite figure here).
+@Morrison2008 also studied the optimal allocation of sampling effort in space versus time, but used emprical data from a longterm census of the collared pika (\emph{Ochotona collaris}) found in the Yukon.
+They estimated longterm growth rates among three subpopulations over a 10-year period.
+They found that censuses less than 5 years may be misleading and that extrapolating from one population to another, even when nearby, may be difficult.
+Therefore, they argued more sampling should be conducted spatially than temporally [@Morrison2008].
 

### 2017-05-13 writing/manuscript_draft.tex
added to discussion section

@@
 
+We show two different approaches to estimate the minimum time series
+length required required to address a particular question of interest.
+The simulation-based approach has been suggested by others, especially
+in situations where analysis is more complicated than a simple power
+analysis (Gerrodette 1987). We focus on a empirical approach to estimate
+the minimum time series length requried.
+
+The first question we asked was, what is \(t_{min}\) for determining
+whether a population was increasing or decreasing over time according to
+linear regression. We found that for 868 populations, at least 10-15
+years of continuous monitoring were neccessary. In line with theoretical
+predictions (citations), we also found \(t_{min}\) was strongly
+correlated with the trend strength, variability in population size, and
+temporal autocorrelation.
+
+We also found that \(t_{min}\) did not correlate with any biological
+variables of interest (cite figure or supp mat). We initially
+hypothesized that species with longer lifespans or generation times may
+require more years to sample. (Our result could have been the effect of
+two reasons) First, the database we used may not include a diverse
+enough set of species with different life history traits. Second, the
+question we poised, whether a population is increasing or decreases, is
+only worried about population surveys. Therefore, biological variables
+may be more important if the question itself was different\ldots{}
+
+\subsubsection{Other questions
+addressed}\label{other-questions-addressed}
+
+In the appendix we asked two additional questions to make estimates of
+\(t_{min}\). First, we asked how many years of sampling are neccessary
+to estimate a geometric growth rate?
+
+\subsubsection{Past work on this issue}\label{past-work-on-this-issue}
+
+There have been previous attempts to address how long a time series
+should be to assess trends in abundance. For example,\ldots{}
+
+An important related question, is the optimal allocation of sampling
+effort in space versus time (citations). In a theoretical investigation
+of this question, Rhodes and Jonzen (2011) found that the optimal
+allocation of sampling depended strongly on temporal and spatial
+autocorrelation. If spatial population dynamics were highly correlated,
+then it was better to sample more temporally, and vice versa. Our work
+supports this idea as populations with strong temporal autocorrelation
+needed less years of sampling (cite figure here). Morrison and Hik
+(2008) also studied the optimal allocation of sampling effort in space
+versus time, but used emprical data from a longterm census of the
+collared pika (\emph{Ochotona collaris}) found in the Yukon. They
+estimated longterm growth rates among three subpopulations over a
+10-year period. They found that censuses less than 5 years may be
+misleading and that extrapolating from one population to another, even
+when nearby, may be difficult. Therefore, they argued more sampling
+should be conducted spatially than temporally (Morrison and Hik 2008).
+
+\subsubsection{Limitations}\label{limitations}
+
+Our work has obvious limitations in determining the minimum time series
+length required. First, \(t_{min}\) depends strongly on the question
+being asked and the availability of long-term census data. For our
+empirical approach, the subsampling of the full time series allows for
+estimates of power, but the individual subsamples are clearly not
+independent of one another. (sentence about alpha and power
+limititations here) The perfect scenario would be to have a population
+model built and parameterized for each population of interest. Then,
+model simulations could be used to estimate the minimum time series
+reuqired to address a specific question of interest. Clearly, this is
+not always practical, especially if running analyses for a wide array of
+species as we do here. In addition, our statistical models suggest that
+\(t_{min}\) does not correlate with any biological variables of
+interest, at least for the question of linear regression. Therefore, it
+is not possible to use these results to predict \(t_{min}\) for another
+population, even if the population is of a species with a similar
+life-history to one in our database.
+
+\subsubsection{Conclusions}\label{conclusions}
+
 \begin{itemize}
@@
 \item
-  Findings and how they relate to past work
-\item
-  Extensions to other systems. Useful to use similar species
-\item
-  Biological variables not good predictors, but we also used all
-  long-lived vertebrate species
-\item
-  Other questions to ask and results in appendix
-\item
-  Future recommendations to others and future work
-\item
-  implications
-\item
-  conclusion paragraph
+  what we found, implications, future work
 \end{itemize}
 
-The IUCN has an 10 year/3 generation rule\ldots{} If we apply this rule
-to the species in our database we see that approximately 50.5760369\% of
-populations need a longer minimum time series than that required by
-IUCN.
+We use a database of 868 populations to determine the minimum time
+series length required. We show that to assess longterm trends in
+abundance, 10-15 years of continuos monitoring are required. In line
+with theoretical predictions, we also show that \(t_{min}\) is strongly
+correlated with the overall trend in abundance, variability in
+abundance, and the temporal autocorrelation. Our work implies that for
+many populations, census data less than 10-15 years are probably not
+reliable. This stresses the importance of longterm monitoring programs.
+(where do IUCN recommendations come from?) From both a scientific and
+management perspective estimates of \(t_{min}\) are important. If a time
+series is too short, we are likely to make incorrect conclusions about
+longterm trends. In addition, a time series that is too long may have
+been a poor use of scare funds. Future work can examine other species,
+with a wider range of life history characteristics, or address other
+questions besides longterm trends in abundance.
 

### 2017-05-15 writing/manuscript_draft.Rmd
added related work in discussion

@@
 
+<!--
 - intro long time series
@@
 - what we do in this paper
+-->
 
-Observational studies, and censusing populations, has become a cornerstone of modern ecology and conservation biology [@Magurran2010].
+Observational studies and population censuses have become a cornerstone of modern ecology and conservation biology [@Magurran2010].
 Longterm data is necessary to understand population dynamics and to assess extinction risk of species.
@@
 Past work has investigated these questions with either theoretical approaches not using data or by examining a single population. As an example of the former, [@Rhodes2011] examined..
+- need to expand this paragraph in line with what I wrote in the discussion...
 
@@
 The simulation-based approach has been suggested by others, especially in situations where analysis is more complicated than a simple power analysis [@Gerrodette1987].
-We focus on a empirical approach to estimate the minimum time series length requried.
+Here, we focus on a empirical approach to estimate the minimum time series length requried to address a given question.
 
-The first question we asked was, what is $t_{min}$ for determining whether a population was increasing or decreasing over time according to linear regression.
-We found that for `r nrow(pop_info)` populations, at least 10-15 years of continuous monitoring were neccessary.
-In line with theoretical predictions (citations), we also found $t_{min}$ was strongly correlated with the trend strength, variability in population size, and temporal autocorrelation.
+First we asked, what is $t_{min}$ for determining whether a population was increasing or decreasing over time according to linear regression?
+We found that for `r nrow(pop_info)` populations, at least 10-15 years of continuous monitoring were neccessary (Fig.
+\ref{fig:fig:min_time_dist}).
+In line with theoretical predictions (citations), we also found $t_{min}$ was strongly correlated with the trend strength, variability in population size, and temporal autocorrelation (cite figure).
 
@@
 There have been previous attempts to address how long a time series should be to assess trends in abundance. For example,...
+
+@Rueda-Cediel2015 found that 10-15 years was adequate to characterize population trends in a short lived animal. They used a matrix model of a particular species...
+
+@Gerber1999 investigated the minimum time series required to estimate population growth of an endagered, but recovering, species.
+They used a longterm census to retroactively determine the minimum time series required to assess threat status.
+They found that only XX years were needed, XX years before the delisting decision was actually made.
+This highlights the importance of estimating the minimunm time series required as an earlier decision would have saved time and money [@Gerber1999].
+
+There has also been interest in determining the minimum time series length in the bird community (citations).
+(Bart 2004, Hatch 2003) [@Seavy2007] Hatch 2003 used seabird monitoring data to estimate the minimum time needed to achieve appropriate levels of power.
+He found that the minimum time required ranged between 11 and 69 years depending on species, trend strength, and study design (cite Hatch).
+Similarly, Bart et al 2004 recommended a standard of 80% power to detect a 50% decline over 20 years in landbird populations.
+They found that only about 42% of populations in the Breeding Bird Survey met their criteria.
+@Seavy2007 asked whether statistical power was even a useful framework for assessing longterm population trends.
+They used 24 years of census data on Red-tailed Tropicbirds (\emph{Phaethon rubricauda}) in Hawaii and showed that to detect a 50% decline over 10 years almost always resulting in high statistical power (above 0.8).
+Therefore, they cautionsioned against only using power analyses to design monitoring schemes and instead argued for precision based metrics ..
 
@@
 - specific recommendations depend on amount of power and strength of trend (Bart 2004, Hatch 2003) [@Seavy2007]
+- paper on syncrony from James
+
+## Supporting Information
+
+- list each bit of supporting information. Github reference
+
+## Acknowledgements
+
+ERW was partially supported by a National Science Foundation Graduate Fellowship.
+We would like to thank members of the Ecological Theory group at the University of California, Davis for their insight.
+We would also like to thank X anonymous reviewers for their helpful comments.
 

### 2017-05-15 writing/manuscript_draft.tex
added related work in discussion

@@
 
-Observational studies, and censusing populations, has become a
-cornerstone of modern ecology and conservation biology (Magurran et al.
-2010). Longterm data is necessary to understand population dynamics and
-to assess extinction risk of species. Even though many time series may
-now be considered ``longterm'', most are still short (i.e.~less than X
+Observational studies and population censuses have become a cornerstone
+of modern ecology and conservation biology (Magurran et al. 2010).
+Longterm data is necessary to understand population dynamics and to
+assess extinction risk of species. Even though many time series may now
+be considered ``longterm'', most are still short (i.e.~less than X
 years). \emph{I could have statistic on how long is considered
@@
 approaches not using data or by examining a single population. As an
-example of the former, (Rhodes and Jonzen 2011) examined..
+example of the former, (Rhodes and Jonzen 2011) examined.. - need to
+expand this paragraph in line with what I wrote in the
+discussion\ldots{}
 
@@
 in situations where analysis is more complicated than a simple power
-analysis (Gerrodette 1987). We focus on a empirical approach to estimate
-the minimum time series length requried.
+analysis (Gerrodette 1987). Here, we focus on a empirical approach to
+estimate the minimum time series length requried to address a given
+question.
 
-The first question we asked was, what is \(t_{min}\) for determining
-whether a population was increasing or decreasing over time according to
-linear regression. We found that for 868 populations, at least 10-15
-years of continuous monitoring were neccessary. In line with theoretical
-predictions (citations), we also found \(t_{min}\) was strongly
-correlated with the trend strength, variability in population size, and
-temporal autocorrelation.
+First we asked, what is \(t_{min}\) for determining whether a population
+was increasing or decreasing over time according to linear regression?
+We found that for 868 populations, at least 10-15 years of continuous
+monitoring were neccessary (Fig. \ref{fig:fig:min_time_dist}). In line
+with theoretical predictions (citations), we also found \(t_{min}\) was
+strongly correlated with the trend strength, variability in population
+size, and temporal autocorrelation (cite figure).
 
@@
 should be to assess trends in abundance. For example,\ldots{}
+
+Rueda-Cediel et al. (2015) found that 10-15 years was adequate to
+characterize population trends in a short lived animal. They used a
+matrix model of a particular species\ldots{}
+
+Gerber, DeMaster, and Kareiva (1999) investigated the minimum time
+series required to estimate population growth of an endagered, but
+recovering, species. They used a longterm census to retroactively
+determine the minimum time series required to assess threat status. They
+found that only XX years were needed, XX years before the delisting
+decision was actually made. This highlights the importance of estimating
+the minimunm time series required as an earlier decision would have
+saved time and money (Gerber, DeMaster, and Kareiva 1999).
+
+There has also been interest in determining the minimum time series
+length in the bird community (citations). (Bart 2004, Hatch 2003) (Seavy
+and Reynolds 2007) Hatch 2003 used seabird monitoring data to estimate
+the minimum time needed to achieve appropriate levels of power. He found
+that the minimum time required ranged between 11 and 69 years depending
+on species, trend strength, and study design (cite Hatch). Similarly,
+Bart et al 2004 recommended a standard of 80\% power to detect a 50\%
+decline over 20 years in landbird populations. They found that only
+about 42\% of populations in the Breeding Bird Survey met their
+criteria. Seavy and Reynolds (2007) asked whether statistical power was
+even a useful framework for assessing longterm population trends. They
+used 24 years of census data on Red-tailed Tropicbirds
+(\emph{Phaethon rubricauda}) in Hawaii and showed that to detect a 50\%
+decline over 10 years almost always resulting in high statistical power
+(above 0.8). Therefore, they cautionsioned against only using power
+analyses to design monitoring schemes and instead argued for precision
+based metrics ..
 
@@
   trend (Bart 2004, Hatch 2003) (Seavy and Reynolds 2007)
+\item
+  paper on syncrony from James
 \end{itemize}
+
+\subsection{Supporting Information}\label{supporting-information}
+
+\begin{itemize}
+\tightlist
+\item
+  list each bit of supporting information. Github reference
+\end{itemize}
+
+\subsection{Acknowledgements}\label{acknowledgements}
+
+ERW was partially supported by a National Science Foundation Graduate
+Fellowship. We would like to thank members of the Ecological Theory
+group at the University of California, Davis for their insight. We would
+also like to thank X anonymous reviewers for their helpful comments.
 

### 2017-06-03 writing/manuscript_draft.Rmd
added to methods and intro of paper

@@
     number_sections: true
-fontsize: 11pt
-geometry: margin=0.75in
+fontsize: 12pt
+geometry: margin=1in
 bibliography: White_bib.bib
@@
 Here, I stress the need for verifying a study has enough statistical power to accurately determine long-term changes in population abundance.
-Most studies, especially those less than 10-15 years, are probably under-powered and potentially misleading.
+Most studies, especially those less than 10-15 years, are probably underpowered and potentially misleading.
 
-## Introduction
+## Introduction - need to clean up and reduce text
 
@@
 In addition, short funding cycles make it difficult to examine populations over longer periods of time.
-There are of course notable exceptions including the Isle Royale study, Iowa corn study, cod, lynx-hair, and many others.
+However, there are of course notable exceptions including the Isle Royale study, Iowa corn study, cod, lynx-hair, and many others.
 
@@
 Scientifically, we need to know how long a time series is required to address a specific question.
-Too short of a time series may lead to wrong conclusions given large natural year-to-year variability.
+Too short of a time series may lead to wrong conclusions given large natural year-to-year variability (citations).
 From a management perspective, it is important to understand when a trend in abundance over time is actually meaningful or not.
-For instance the International Union for the Conservation of Nature (IUCN) suggest...In addition, because sampling is typically expensive, we do not want to over sample [@Gerber1999].
+For example, the IUCN Red List Categories and Criteria suggest under Criterian A2, a species qualifies as vulnerable if it has experienced a 30% decline over 10 years, or 3 generations [@IUCN2012].
+In addition, because sampling is typically expensive, we do not want to sample for longer than is neccessary [@Gerber1999].
 
@@
 The exact relationship between these quantities depends on the specific statistical test.
-Formally, statistical power is one minus the probability of a type II error, or false negative.
+Formally, statistical power ($1-\beta$) is one minus the probability of a type II error ($\beta$), or false negative.
 In other words, it is a measure of how well you could detect a trend given it is actually significant.
-Prior to an experiment, one could set appropriate levels of power, $\alpha$, and the effect size to estimate the sample size required for the experiment.
-This approach, however, is not as simple for a time series as the data points are clearly non-independent.
+Prior to an experiment, one could set appropriate levels of power, significance threshold ($\alpha$), and the effect size to estimate the sample size required for the experiment.
+This approach, however, is not straight-forward for a time series, as data points are clearly non-independent.
 
@@
 Instead of using classic statistical tests (e.g.
-t-test), and corresponding power calculations, simulation approaches can be used [@Gerrodette1987].
+t-test), and corresponding formulas for determinig sample size, simulation approaches can be used [@Gerrodette1987].
 Simple population models can be simulated with parameter values corresponding to a population of interest.
@@
 The specific criteria depend on the question at hand.
-Then one can run simulations for time series of different lengths to determine the minimum time series length required to be confident that you will be able to answer your question.
+Then time series simulations of different lengths are run to determine the minimum time series length required to be confident that you will be able to answer your question (cite Bolker).
 In addition to using simulated data, empirical time series can also be used.
-We usually do not have multiple replicates of similar populations, but we can subsample an empirical time series.
-Subsamples of different lengths can then be evaluated to see which fraction of subsamples meet some criteria, again a measure of statistical power...
+We usually do not have multiple replicates of similar populations, but we can subsample an empirical time series (cite).
+Subsamples of different lengths can then be evaluated to see which fraction of subsamples meet some criteria, again a measure of statistical power.
+Similar to the simulation approach, we can use these measures of power to determine the minimum time series required for a particular question of interest.
 
-Past work has investigated these questions with either theoretical approaches not using data or by examining a single population. As an example of the former, [@Rhodes2011] examined..
-- need to expand this paragraph in line with what I wrote in the discussion...
+Past work has investigated questions related to the minimum time series required to estimate trends in population size over time.
+However, most work has either focused on theoretical approaches or by examined a single population.
+As an example of the former, @Rhodes2011 examined the optimal allocation of effort between spatial and temporal replicates.
+Using simple populations models, they found that the allocation of effort depends on environmental variation, spatial and temporal autocorrelation, and observer error.
+@Rueda-Cediel2015 also used a modeling approach, but parameterized it for a threatened snail, \emph{Tasmaphena lamproides}.
+They found that for this short-lived organims, that 15 years was adequate to assess longterm trends in abundance.
+However, these studies, and other past work (citation), have focused either on theoretical aspects of monitoring design or focused on single species.
 
@@
 
-One approach to determining the minimum time series needed is through repetitive simulations of a population model [@Gerrodette1987].
-This is the same approach one might use in sample size calculations for any experimental design too complicated for simple power analyses [@Johnson2015].
+One approach to determining the minimum time series length needed is through repetitive simulations of a population model [@Gerrodette1987].
+This is the same approach one might use in sample size calculations for any experimental design too complicated for simple power analyses [@Johnson2015] (and cite Bolker).
 We only briefly discuss this approach as it has been desribed elsewhere (citations).
 Essentially, we take a stochastic population model and simulate it for a number of years repetitively.
-We can then determine the minimum time series required to achieve designed levels of significance and power.
-This approach requires us to determine values for our model parameters.
+We can then determine the minimum time series required to achieve designed levels of significance and power for a particular question.
+This approach requires us to determine values for our model parameters (e.g.
+growth rate).
 As an example, we can take the following population model for population size $N$ at time $t$:
@@
 The rate of growth $r$ is also the trend strength of increase or decrease.
-Lastly, $\phi$ represents an autocorrelation term where $\phi=0$ indicates no autocorrelation, $\phi>0$ is positively autocorrelated.
+<!--Lastly, $\phi$ represents an autocorrelation term where $\phi=0$ indicates no autocorrelation, $\phi>0$ is positively autocorrelated.
 This model represents linear growth with autocorrelation and noise.
-However, it is certainly possible to use any other population model.
+However, it is certainly possible to use any other population model.-->
 
 By simulating this model repetitively for different lengths of time, we can determine the minimum time series required.
-Here statistical power is simply the number of trials that are themselves significant, as the real trend is different from zero.
-Then the minimum time required is the minimum time required to obtain statistical power above some pre-determined threshold.
-Historically, 0.8....s
+Here statistical power is simply the fraction of simulations with significant trends in abundance from simple linear regression.
+Then, the minimum time required is the minimum time required to obtain statistical power above some pre-determined threshold (historically at 0.8).
 
-In figure X, a number of simulated time series are shown for a set number of time periods (t=40).
+In figure \ref{fig:theoretical_approach}a, a number of simulated time series are shown for a set number of time periods (t=40).
 It is clear that statistical power increases quickly with increases in $t$ (Fig.
-Xb).
+\ref{fig:theoretical_approach}b).
 Once power is greater than XX (the dotted line), that is the minimum time required ($t_{min}$
 
-<!--
-![optional caption text](example_theoretical_output.pdf)
--->
-
-```{r,echo=F,fig.cap='theoretical approach\\label{fig:theoretical_approach}'}
+```{r,echo=F,fig.cap='(a) example of simulated time series for 40 time periods and (b) statistical power versus the simulated time series length.
+The horizontal, dashed line is our desired statistical power of 0.8.
+The vertical, dashed line is the minimum time required to achieve our desired statistical power.\\label{fig:theoretical_approach}'}
 setwd("~/Desktop/Research/time-series-project")
@@
   setwd("~/Desktop/Research/time-series-project")
-#Load keith database
+
+#Load keith et al. database (from GPDD database)
   load('data/Keith_et_al_2015_data.Rdata')
 
-#Load our cleaned data
+#Load our cleaned data (see other code for cleaning steps)
   load('cleaned-data/pop_info_example_goodNAs.Rdata')
@@
 For each time series, we also calculated variables of interest like variance in population size, longterm trend in abundance (slope coefficient from simple linear regression), and temporal autocorrelation.
-Lastly, we also calculated the minimum number of years required to sample according to IUCN standards under rule ...
-(citation needed).
+Lastly, we also calculated the minimum number of years required (10 years or 3 generations) under IUCN Red List Criterian A2 to classify a species as vulnerable [@IUCN2012].
 
@@
 As we describe below, the minimum time series length calculation is particular to each question.
-Intuitively, we might expect a longer minimum time series length required given more complicated questions.
+Intuitively, we might expect a longer minimum time series length ($T_{min}$) required given more complicated questions (citations).
 
-We assume that each time series in our database are long enough to include all necessary information (e.g.
+We assume that each time series in our database is long enough to include all necessary information (e.g.
 variability) about the population.
-In other words, they are a Representative sample.
+In other words, they are a representative sample.
 We first take all possible contiguous subsamples of each time series.
-For example, a time series of 40 years would have 39 possible contiguous subsamples of length 2, 38 possible contiguous subsamples of length 3, and so forth.
+For example, a time series of 40 years would have 39 possible contiguous subsamples of length 2, 38 possible contiguous subsamples of length 3, and continueing until 1 possible contiguous subsample of length 40.
 In this paper, we will try to ask a simple, specific question.
@@
 Therefore, we run a linear regression for each subsampled time series.
-Then, we determine the fraction of subsamples of a particular length that have slopes from linear regression which are statistically different from zero given the longterm, or ``true", time series also has a significant slope.
-This is a measure of statistical power.
-Lastly, we determine which subsample length is required to guarantee we have above a certain threshold of statistical power, which here we will assume is 0.8 by convention (citation).
+Then, we determine the fraction of subsamples of a particular length that have estimated slope coefficients which are statistically different from zero given the longterm, or ``true", time series also has a significant slope.
+This is a measure of statistical power (often denoted as $1- \beta$, where $\beta$ is the probability of a type II error).
+Lastly, we determine which subsample length is required to guarantee we have above a certain threshold of statistical power, which we will assume is 0.8 by convention (citation).
+A power of 0.8 would imply that given a trend exists, there is a 0.8 probability of detecting the trend.
+Complementary, there would be a 0.2 probability of a false negative.
 
@@
 In the supplementary material, we show how to make similar calculations to determine the minimum time series required to estimate overall population growth rate, the variability in population size, and future extinction risk.
-
-In the appendix, we explore other potential questions of interest...
+Again, we took subsamples of time series data to determine $T_{min}$ for each of these questions.
 
@@
 ## Results
-
+<!--
 - summary of findings
@@
   - walk through one empirical example (maybe in appendix)
+-->
 
@@
 We chose a simple model, but any other population model could be used.
-The specific model choice should be tailored to the population of interest, based on species ecology.
+The specific model choice should be tailored to the population of interest, based on its ecology.
 We then ran simulations of different lengths to determine the minimum time series length required to achieve a certain level of power (which we set to 0.8).
@@
 
-```{r,echo=F,fig.cap='Distribution of the minimum time required in order to detect a significant trend (at the 0.05 level) in abundance given power of 0.8.\\label{fig:fig:min_time_dist}'}
+```{r,echo=F,fig.cap='Distribution of the minimum time required in order to detect a significant trend (at the 0.05 level) in abundance given power of 0.8.\\label{fig:min_time_dist}'}
 # Pull in analysis of average min time needed for **empirical species** (save analysis as Rdata file or something)
@@
 
-```{r poisson_model, echo=F}
+```{r poisson_model, echo=F,include=F}
 #Poisson regression
@@
 
-```{r,echo=F,fig.cap='duh\\label{fig:correlates}'}
+```{r,echo=F,fig.cap='Minimum time required to estimate change in abundance correlated with (a) trend strength (absolute value of slope coefficient estimated from linear regression), (b) population variance (interannual variability in population size), and (c) temporal autocorrelation.\\label{fig:correlates}'}
 
@@
   mtext('| trend strength |',1,line=3,cex=1.2)
+  mtext('(a)',3,line=-1.5,at=0.033,cex=1.2,font = 2)
     plot(pop_info$variance,pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
-    mtext('population variance',1,line=3,cex=1.2)
+    mtext('population variance',1,line=3,cex=1.3)
+    mtext('(b)',3,line=-1.5,at=0.128,cex=1.2,font = 2)
       plot(pop_info$autocorrelation,pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
       mtext('autocorrelation',1,line=3,cex=1.2)
-
+       mtext('(c)',3,line=-1.5,at=0.98,cex=1.2,font = 2)
        mtext('minimum time required',2,line=2.5,cex=1.2,outer = T)
@@
 
-Across all the populations we examined, we found an average minimum time series length required ($T_{min}$) of `r mean(pop_info$min_time_for_power)` ($\sigma$=`r (var(pop_info$min_time_for_power))^0.5`), with a wide distribution (Fig.
+Across all the populations we examined, we found an average minimum time series length required ($T_{min}$) of `r round(mean(pop_info$min_time_for_power),1)` ($\sigma$=`r round((var(pop_info$min_time_for_power))^0.5,1)`), with a wide distribution (Fig.
 \ref{fig:min_time_dist}).
-As expected based on theoretical results (cite papers), the minimum time series length required is strongly correlated with trend strength, population variance, and autocorrelation (Fig \ref{fig:correlates}).
-Using a generalized linear model, with a Poisson error structure, all three of these explanatory variables were significant and had large effect sizes (cite model here).
-Combined trend strength, population variance, and autocorrelation account for about `r 100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance)` of the variance in minimum time series length required (as proposed by CITE Zuur).
+In addition, as expected based on theoretical results (cite papers), the minimum time series length required is strongly correlated with trend strength, population variance, and autocorrelation (Fig \ref{fig:correlates}).
+Using a generalized linear model, with a Poisson error structure, all three of these explanatory variables were significant and had large effect sizes (cite model from supp mat here).
+Combined trend strength, population variance, and autocorrelation account for about `r round(100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance),1)` percent of the variance in minimum time series length required (as proposed by CITE Zuur).
 
@@
 body mass, longevity, location).
-None (need to do regression) of these explanatory variables were predictive of the minimum time series length required.
-Some differences if $t_{min}$ did occur between biological classes, but this can be attributed to the amount populations are know to fluctuate.
+None (need to do regression) of these explanatory variables were predictive of the minimum time series length required (supp mat).
+Some differences if $t_{min}$ did occur between biological class (supp mat), but this can be attributed to the amount populations are known to fluctuate (i.e.
+the variability in population size).
 

### 2017-06-03 writing/manuscript_draft.tex
added to methods and intro of paper

@@
 }{}
-\usepackage[margin=0.75in]{geometry}
+\usepackage[margin=1in]{geometry}
 \usepackage{hyperref}
@@
 population abundance. Most studies, especially those less than 10-15
-years, are probably under-powered and potentially misleading.
+years, are probably underpowered and potentially misleading.
 
-\subsection{Introduction}\label{introduction}
-
-\begin{itemize}
-\tightlist
-\item
-  intro long time series
-\item
-  my general question
-\item
-  primer on power
-\item
-  more general power interpretation
-\item
-  2+ approaches
-\item
-  what we do in this paper
-\end{itemize}
+\subsection{Introduction - need to clean up and reduce
+text}\label{introduction---need-to-clean-up-and-reduce-text}
 
@@
 last a couple of years. In addition, short funding cycles make it
-difficult to examine populations over longer periods of time. There are
-of course notable exceptions including the Isle Royale study, Iowa corn
-study, cod, lynx-hair, and many others.
+difficult to examine populations over longer periods of time. However,
+there are of course notable exceptions including the Isle Royale study,
+Iowa corn study, cod, lynx-hair, and many others.
 
@@
 required to address a specific question. Too short of a time series may
-lead to wrong conclusions given large natural year-to-year variability.
-From a management perspective, it is important to understand when a
-trend in abundance over time is actually meaningful or not. For instance
-the International Union for the Conservation of Nature (IUCN)
-suggest\ldots{}In addition, because sampling is typically expensive, we
-do not want to over sample (Gerber, DeMaster, and Kareiva 1999).
+lead to wrong conclusions given large natural year-to-year variability
+(citations). From a management perspective, it is important to
+understand when a trend in abundance over time is actually meaningful or
+not. For example, the IUCN Red List Categories and Criteria suggest
+under Criterian A2, a species qualifies as vulnerable if it has
+experienced a 30\% decline over 10 years, or 3 generations (IUCN 2012).
+In addition, because sampling is typically expensive, we do not want to
+sample for longer than is neccessary (Gerber, DeMaster, and Kareiva
+1999).
 
@@
 relationship between these quantities depends on the specific
-statistical test. Formally, statistical power is one minus the
-probability of a type II error, or false negative. In other words, it is
-a measure of how well you could detect a trend given it is actually
-significant. Prior to an experiment, one could set appropriate levels of
-power, \(\alpha\), and the effect size to estimate the sample size
-required for the experiment. This approach, however, is not as simple
-for a time series as the data points are clearly non-independent.
+statistical test. Formally, statistical power (\(1-\beta\)) is one minus
+the probability of a type II error (\(\beta\)), or false negative. In
+other words, it is a measure of how well you could detect a trend given
+it is actually significant. Prior to an experiment, one could set
+appropriate levels of power, significance threshold (\(\alpha\)), and
+the effect size to estimate the sample size required for the experiment.
+This approach, however, is not straight-forward for a time series, as
+data points are clearly non-independent.
 
@@
 statistical power (cite Bolker). Instead of using classic statistical
-tests (e.g.~t-test), and corresponding power calculations, simulation
-approaches can be used (Gerrodette 1987). Simple population models can
-be simulated with parameter values corresponding to a population of
-interest. Out of a large number of simulations, power can be calculated
-as the fraction of simulations that meet some criteria. The specific
-criteria depend on the question at hand. Then one can run simulations
-for time series of different lengths to determine the minimum time
-series length required to be confident that you will be able to answer
-your question. In addition to using simulated data, empirical time
-series can also be used. We usually do not have multiple replicates of
-similar populations, but we can subsample an empirical time series.
-Subsamples of different lengths can then be evaluated to see which
-fraction of subsamples meet some criteria, again a measure of
-statistical power\ldots{}
+tests (e.g.~t-test), and corresponding formulas for determinig sample
+size, simulation approaches can be used (Gerrodette 1987). Simple
+population models can be simulated with parameter values corresponding
+to a population of interest. Out of a large number of simulations, power
+can be calculated as the fraction of simulations that meet some
+criteria. The specific criteria depend on the question at hand. Then
+time series simulations of different lengths are run to determine the
+minimum time series length required to be confident that you will be
+able to answer your question (cite Bolker). In addition to using
+simulated data, empirical time series can also be used. We usually do
+not have multiple replicates of similar populations, but we can
+subsample an empirical time series (cite). Subsamples of different
+lengths can then be evaluated to see which fraction of subsamples meet
+some criteria, again a measure of statistical power. Similar to the
+simulation approach, we can use these measures of power to determine the
+minimum time series required for a particular question of interest.
 
-Past work has investigated these questions with either theoretical
-approaches not using data or by examining a single population. As an
-example of the former, (Rhodes and Jonzen 2011) examined.. - need to
-expand this paragraph in line with what I wrote in the
-discussion\ldots{}
+Past work has investigated questions related to the minimum time series
+required to estimate trends in population size over time. However, most
+work has either focused on theoretical approaches or by examined a
+single population. As an example of the former, Rhodes and Jonzen (2011)
+examined the optimal allocation of effort between spatial and temporal
+replicates. Using simple populations models, they found that the
+allocation of effort depends on environmental variation, spatial and
+temporal autocorrelation, and observer error. Rueda-Cediel et al. (2015)
+also used a modeling approach, but parameterized it for a threatened
+snail, \emph{Tasmaphena lamproides}. They found that for this
+short-lived organims, that 15 years was adequate to assess longterm
+trends in abundance. However, these studies, and other past work
+(citation), have focused either on theoretical aspects of monitoring
+design or focused on single species.
 
@@
 
-One approach to determining the minimum time series needed is through
-repetitive simulations of a population model (Gerrodette 1987). This is
-the same approach one might use in sample size calculations for any
-experimental design too complicated for simple power analyses (Johnson
-et al. 2015). We only briefly discuss this approach as it has been
-desribed elsewhere (citations). Essentially, we take a stochastic
-population model and simulate it for a number of years repetitively. We
-can then determine the minimum time series required to achieve designed
-levels of significance and power. This approach requires us to determine
-values for our model parameters. As an example, we can take the
-following population model for population size \(N\) at time \(t\): \[
+One approach to determining the minimum time series length needed is
+through repetitive simulations of a population model (Gerrodette 1987).
+This is the same approach one might use in sample size calculations for
+any experimental design too complicated for simple power analyses
+(Johnson et al. 2015) (and cite Bolker). We only briefly discuss this
+approach as it has been desribed elsewhere (citations). Essentially, we
+take a stochastic population model and simulate it for a number of years
+repetitively. We can then determine the minimum time series required to
+achieve designed levels of significance and power for a particular
+question. This approach requires us to determine values for our model
+parameters (e.g.~growth rate). As an example, we can take the following
+population model for population size \(N\) at time \(t\): \[
 N(t + 1) = N(t) + r(t) \mbox{ with } r(t) \sim N(\phi \mu, \sigma)
@@
 mean \(\mu\) and standard deviation \(\sigma\). The rate of growth \(r\)
-is also the trend strength of increase or decrease. Lastly, \(\phi\)
-represents an autocorrelation term where \(\phi=0\) indicates no
-autocorrelation, \(\phi>0\) is positively autocorrelated. This model
-represents linear growth with autocorrelation and noise. However, it is
-certainly possible to use any other population model.
+is also the trend strength of increase or decrease.
 
@@
 can determine the minimum time series required. Here statistical power
-is simply the number of trials that are themselves significant, as the
-real trend is different from zero. Then the minimum time required is the
-minimum time required to obtain statistical power above some
-pre-determined threshold. Historically, 0.8\ldots{}.s
+is simply the fraction of simulations with significant trends in
+abundance from simple linear regression. Then, the minimum time required
+is the minimum time required to obtain statistical power above some
+pre-determined threshold (historically at 0.8).
 
-In figure X, a number of simulated time series are shown for a set
-number of time periods (t=40). It is clear that statistical power
-increases quickly with increases in \(t\) (Fig. Xb). Once power is
-greater than XX (the dotted line), that is the minimum time required
-(\(t_{min}\)
+In figure \ref{fig:theoretical_approach}a, a number of simulated time
+series are shown for a set number of time periods (t=40). It is clear
+that statistical power increases quickly with increases in \(t\) (Fig.
+\ref{fig:theoretical_approach}b). Once power is greater than XX (the
+dotted line), that is the minimum time required (\(t_{min}\)
 
@@
 \includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-1-1.pdf}
-\caption{theoretical approach\label{fig:theoretical_approach}}
+\caption{(a) example of simulated time series for 40 time periods and
+(b) statistical power versus the simulated time series length. The
+horizontal, dashed line is our desired statistical power of 0.8. The
+vertical, dashed line is the minimum time required to achieve our
+desired statistical power.\label{fig:theoretical_approach}}
 \end{figure}
@@
 regression), and temporal autocorrelation. Lastly, we also calculated
-the minimum number of years required to sample according to IUCN
-standards under rule \ldots{} (citation needed).
+the minimum number of years required (10 years or 3 generations) under
+IUCN Red List Criterian A2 to classify a species as vulnerable (IUCN
+2012).
 
@@
 to each question. Intuitively, we might expect a longer minimum time
-series length required given more complicated questions.
+series length (\(T_{min}\)) required given more complicated questions
+(citations).
 
-We assume that each time series in our database are long enough to
+We assume that each time series in our database is long enough to
 include all necessary information (e.g.~variability) about the
-population. In other words, they are a Representative sample. We first
+population. In other words, they are a representative sample. We first
 take all possible contiguous subsamples of each time series. For
@@
 subsamples of length 2, 38 possible contiguous subsamples of length 3,
-and so forth. In this paper, we will try to ask a simple, specific
-question. We seek to determine the \(T_{min}\) required to be confident
-in an estimate of the slope coefficient from linear regression.
-Therefore, we run a linear regression for each subsampled time series.
-Then, we determine the fraction of subsamples of a particular length
-that have slopes from linear regression which are statistically
-different from zero given the longterm, or ``true``, time series also
-has a significant slope. This is a measure of statistical power. Lastly,
-we determine which subsample length is required to guarantee we have
-above a certain threshold of statistical power, which here we will
-assume is 0.8 by convention (citation).
+and continueing until 1 possible contiguous subsample of length 40. In
+this paper, we will try to ask a simple, specific question. We seek to
+determine the \(T_{min}\) required to be confident in an estimate of the
+slope coefficient from linear regression. Therefore, we run a linear
+regression for each subsampled time series. Then, we determine the
+fraction of subsamples of a particular length that have estimated slope
+coefficients which are statistically different from zero given the
+longterm, or ``true``, time series also has a significant slope. This is
+a measure of statistical power (often denoted as \(1- \beta\), where
+\(\beta\) is the probability of a type II error). Lastly, we determine
+which subsample length is required to guarantee we have above a certain
+threshold of statistical power, which we will assume is 0.8 by
+convention (citation). A power of 0.8 would imply that given a trend
+exists, there is a 0.8 probability of detecting the trend.
+Complementary, there would be a 0.2 probability of a false negative.
 
@@
 required to estimate overall population growth rate, the variability in
-population size, and future extinction risk.
-
-In the appendix, we explore other potential questions of
-interest\ldots{}
+population size, and future extinction risk. Again, we took subsamples
+of time series data to determine \(T_{min}\) for each of these
+questions.
 
@@
 \subsection{Results}\label{results}
-
-\begin{itemize}
-\tightlist
-\item
-  summary of findings
-\item
-  theoretical approach
-\item
-  example plot
-\item
-  empirical approach
-\item
-  plots: distribution of min time, what min time correlations with
-\item
-  walk through one empirical example (maybe in appendix)
-\end{itemize}
 
@@
 population model could be used. The specific model choice should be
-tailored to the population of interest, based on species ecology. We
-then ran simulations of different lengths to determine the minimum time
+tailored to the population of interest, based on its ecology. We then
+ran simulations of different lengths to determine the minimum time
 series length required to achieve a certain level of power (which we set
@@
 significant trend (at the 0.05 level) in abundance given power of
-0.8.\label{fig:fig:min_time_dist}}
+0.8.\label{fig:min_time_dist}}
 \end{figure}
@@
 \includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
-\caption{duh\label{fig:correlates}}
+\caption{Minimum time required to estimate change in abundance
+correlated with (a) trend strength (absolute value of slope coefficient
+estimated from linear regression), (b) population variance (interannual
+variability in population size), and (c) temporal
+autocorrelation.\label{fig:correlates}}
 \end{figure}
@@
 Across all the populations we examined, we found an average minimum time
-series length required (\(T_{min}\)) of 16.6463134
-(\(\sigma\)=8.6664727), with a wide distribution (Fig.
-\ref{fig:min_time_dist}). As expected based on theoretical results (cite
-papers), the minimum time series length required is strongly correlated
-with trend strength, population variance, and autocorrelation (Fig
-\ref{fig:correlates}). Using a generalized linear model, with a Poisson
-error structure, all three of these explanatory variables were
-significant and had large effect sizes (cite model here). Combined trend
-strength, population variance, and autocorrelation account for about
-70.7793122 of the variance in minimum time series length required (as
-proposed by CITE Zuur).
+series length required (\(T_{min}\)) of 16.6 (\(\sigma\)=8.7), with a
+wide distribution (Fig. \ref{fig:min_time_dist}). In addition, as
+expected based on theoretical results (cite papers), the minimum time
+series length required is strongly correlated with trend strength,
+population variance, and autocorrelation (Fig \ref{fig:correlates}).
+Using a generalized linear model, with a Poisson error structure, all
+three of these explanatory variables were significant and had large
+effect sizes (cite model from supp mat here). Combined trend strength,
+population variance, and autocorrelation account for about 70.8 percent
+of the variance in minimum time series length required (as proposed by
+CITE Zuur).
 
@@
 to do regression) of these explanatory variables were predictive of the
-minimum time series length required. Some differences if \(t_{min}\) did
-occur between biological classes, but this can be attributed to the
-amount populations are know to fluctuate.
+minimum time series length required (supp mat). Some differences if
+\(t_{min}\) did occur between biological class (supp mat), but this can
+be attributed to the amount populations are known to fluctuate (i.e.~the
+variability in population size).
 
@@
 
+\hypertarget{ref-IUCN2012}{}
+IUCN. 2012. ``IUCN Red List Categories and Criteria: Version 3.1.''
+doi:\href{https://doi.org/10.9782-8317-0633-5}{10.9782-8317-0633-5}.
+
 \hypertarget{ref-Johnson2015}{}

### 2017-06-05 writing/manuscript_draft.Rmd
changes to figures and supp mat

@@
 #Load our cleaned data (see other code for cleaning steps)
-  load('cleaned-data/pop_info_example_goodNAs.Rdata')
+  load('cleaned-data/pop_info_example_goodNAs.Rdata') #Do i need to filter out non-significant populations as well (where no trend was present)
 ```
@@
 The data is originally from the Global Population Dynamics Database [@GPDD2010].
-We filtered out short time series (less than `r pop_info$series_length[1]`) and were left with `r nrow(pop_info)` time.
+We filtered out short time series (less than `r pop_info$series_length[1]`), and those with missing data, and were left with `r nrow(pop_info)` time series.
 The database includes information on `r length(table(pop_info$Binomial))` vertebrate species with a focus on mammals, birds, and fish.
@@
 We evaluated the question of what minimum time series is required to answer a question of interest.
-More specifically we ask, what is the minimum time series length required to determine, via linear regression, the longterm trend in population abundance?
+More specifically we asked, what is the minimum time series length required to determine, via linear regression, the longterm trend in population abundance?
 The minimum time series length required is that which had high enough statistical power.
@@
 
+### Questions in the appendix
+
+Describe some of the results that we include in the appendix
+
 ## Discussion
@@
 The simulation-based approach has been suggested by others, especially in situations where analysis is more complicated than a simple power analysis [@Gerrodette1987].
+These simulations support past work...
+
 Here, we focus on a empirical approach to estimate the minimum time series length requried to address a given question.
@@
 
-### Past work on this issue
+### Past work on this issue - can I integrate this better into the rest of the section. How much should go in introduction?
 There have been previous attempts to address how long a time series should be to assess trends in abundance. For example,...
@@
 We use a database of `r nrow(pop_info)` populations to determine the minimum time series length required.
-We show that to assess longterm trends in abundance, 10-15 years of continuos monitoring are required.
-In line with theoretical predictions, we also show that $t_{min}$ is strongly correlated with the overall trend in abundance, variability in abundance, and the temporal autocorrelation.
+We show that to assess longterm trends in abundance, 10-15 years of continuos monitoring are required (Fig.
+\ref{fig:min_time_dist}.
+In line with theoretical predictions, we also show that $t_{min}$ is strongly correlated with the overall trend in abundance, variability in abundance, and the temporal autocorrelation (Fig.
+\ref{fig:correlates}).
 Our work implies that for many populations, census data less than 10-15 years are probably not reliable.
@@
 If a time series is too short, we are likely to make incorrect conclusions about longterm trends.
-In addition, a time series that is too long may have been a poor use of scare funds.
+In addition, a time series that is too long may be a poor use of already limited funds [@Gerber1999].
 Future work can examine other species, with a wider range of life history characteristics, or address other questions besides longterm trends in abundance.
 
+<!--
 ### Time series length recommendations from other papers
@@
 - paper on syncrony from James
+-->
 

### 2017-06-05 writing/manuscript_draft.tex
changes to figures and supp mat

@@
 Database (NERC Centre for Population Biology 2010). We filtered out
-short time series (less than 35) and were left with 1019 time. The
-database includes information on 499 vertebrate species with a focus on
-mammals, birds, and fish. The database also includes information on
-generation length and census specifications. For each time series, we
-also calculated variables of interest like variance in population size,
-longterm trend in abundance (slope coefficient from simple linear
-regression), and temporal autocorrelation. Lastly, we also calculated
-the minimum number of years required (10 years or 3 generations) under
-IUCN Red List Criterian A2 to classify a species as vulnerable (IUCN
-2012).
+short time series (less than 35), and those with missing data, and were
+left with 1019 time series. The database includes information on 499
+vertebrate species with a focus on mammals, birds, and fish. The
+database also includes information on generation length and census
+specifications. For each time series, we also calculated variables of
+interest like variance in population size, longterm trend in abundance
+(slope coefficient from simple linear regression), and temporal
+autocorrelation. Lastly, we also calculated the minimum number of years
+required (10 years or 3 generations) under IUCN Red List Criterian A2 to
+classify a species as vulnerable (IUCN 2012).
 
@@
 We evaluated the question of what minimum time series is required to
-answer a question of interest. More specifically we ask, what is the
+answer a question of interest. More specifically we asked, what is the
 minimum time series length required to determine, via linear regression,
@@
 
+\subsubsection{Questions in the
+appendix}\label{questions-in-the-appendix}
+
+Describe some of the results that we include in the appendix
+
 \subsection{Discussion}\label{discussion}
@@
 in situations where analysis is more complicated than a simple power
-analysis (Gerrodette 1987). Here, we focus on a empirical approach to
-estimate the minimum time series length requried to address a given
-question.
+analysis (Gerrodette 1987). These simulations support past work\ldots{}
+
+Here, we focus on a empirical approach to estimate the minimum time
+series length requried to address a given question.
 
@@
 
-\subsubsection{Past work on this issue}\label{past-work-on-this-issue}
+\subsubsection{Past work on this issue - can I integrate this better
+into the rest of the section. How much should go in
+introduction?}\label{past-work-on-this-issue---can-i-integrate-this-better-into-the-rest-of-the-section.-how-much-should-go-in-introduction}
 
@@
 series length required. We show that to assess longterm trends in
-abundance, 10-15 years of continuos monitoring are required. In line
-with theoretical predictions, we also show that \(t_{min}\) is strongly
-correlated with the overall trend in abundance, variability in
-abundance, and the temporal autocorrelation. Our work implies that for
-many populations, census data less than 10-15 years are probably not
-reliable. This stresses the importance of longterm monitoring programs.
-(where do IUCN recommendations come from?) From both a scientific and
-management perspective estimates of \(t_{min}\) are important. If a time
-series is too short, we are likely to make incorrect conclusions about
-longterm trends. In addition, a time series that is too long may have
-been a poor use of scare funds. Future work can examine other species,
-with a wider range of life history characteristics, or address other
-questions besides longterm trends in abundance.
-
-\subsubsection{Time series length recommendations from other
-papers}\label{time-series-length-recommendations-from-other-papers}
-
-\begin{itemize}
-\tightlist
-\item
-  greater than 5 years (Morrison and Hik 2008)
-\item
-  more than 10-15 years (Rueda-Cediel et al. 2015)
-\item
-  general advice (variability, correlations) (Rhodes and Jonzen 2011)
-\item
-  more than 10 years (Gerber, DeMaster, and Kareiva 1999)
-\item
-  specific recommendations depend on amount of power and strength of
-  trend (Bart 2004, Hatch 2003) (Seavy and Reynolds 2007)
-\item
-  paper on syncrony from James
-\end{itemize}
+abundance, 10-15 years of continuos monitoring are required (Fig.
+\ref{fig:min_time_dist}. In line with theoretical predictions, we also
+show that \(t_{min}\) is strongly correlated with the overall trend in
+abundance, variability in abundance, and the temporal autocorrelation
+(Fig. \ref{fig:correlates}). Our work implies that for many populations,
+census data less than 10-15 years are probably not reliable. This
+stresses the importance of longterm monitoring programs. (where do IUCN
+recommendations come from?) From both a scientific and management
+perspective estimates of \(t_{min}\) are important. If a time series is
+too short, we are likely to make incorrect conclusions about longterm
+trends. In addition, a time series that is too long may be a poor use of
+already limited funds (Gerber, DeMaster, and Kareiva 1999). Future work
+can examine other species, with a wider range of life history
+characteristics, or address other questions besides longterm trends in
+abundance.
 

### 2017-06-05 writing/supp_mat_draft.Rmd
changes to figures and supp mat

@@
 fontsize: 12pt
-geometry: margin=0.75in
+geometry: margin=1in
 bibliography: White_bib.bib
@@
 
-Dec XX, 2016
+June 22, 2017
  \end{center}
@@
 
-1. Mathematical description of our model
-2. Inverse pattern-orientated estimation procedure
+1. Detailed example of subsampling and power calculations
+2. Figures and results from other questions of interest
+3. Additional figures?
+4. Additional references
 
-Code for all the figures and tables can be found on [Github](https://github.com/erwhite1).
+Code for all the figures and tables can be found at [](https://github.com/erwhite1).
 
@@
 
-```{r, echo=F, fig.cap='duh\\label{fig:growth_rate}'}
+Instead of detecting a trend over time with linear regression, we could also calculate the geometric growth rate of the population.
+In figure \ref{fig:growth_rate}, we show how to calculate growth rates for subsamples of a time series.
+First, we created subsamples of each possible length from the full 35 year time series, as we did in the main next.
+Next, we calculated the mean and standard deviation of growth rates for each possible time series length (Fig.
+\ref{fig:growth_rate}b).
+Lastly, we calculated the percent error $\mbox{percent error} = 100 \times \left| \frac{\mbox{observed} - \mbox{theoretical}}{\mbox{theoretical}} \right|$ between the mean of each time series length (observed) and the overall population growth rate (theoretical).
+In Fig.
+\ref{fig:growth_rate}c), we show the percent error as a function of time series length.
+Here we define the minimum time required as the minimum number of years to achieve less than 20% error.
+
+```{r, echo=F, fig.cap='Example of calculating minimum time required for growth rate estimation.
+(a) abundance over time, (b) mean and standard deviation of growth rate for subsamples of entire time series, and (c) the percent error between mean estimated growth rate and the true longterm growth rate.
+The vertical bar denotes the minimum time required to estimate growth rate within 10% error.\\label{fig:growth_rate}'}
 #Example of using population growth rate instead of power
@@
 
-par(mfrow=c(1,2))
-  pop = subset(long_dat,long_dat$ID=='10696') #13192 for Harp seals
-  plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='',xlab='',pch=16,ylim=c(0,2000),las=1)
-    abline(lm(pop$popvalue~pop$year),lwd=2,col='red')
+par(mfrow=c(1,3),oma=c(0,1,0,0))
+
+  pop = subset(long_dat,long_dat$ID=='10007') #13192 for Harp seals
+  source('scripts/create_subsamples.R')
+  source('scripts/percent_error.R')
+  aaa=calculate_average_geometric_growth(pop$popvalue)
+  plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='',xlab='',pch=16,las=1)
+    #abline(lm(pop$popvalue~pop$year),lwd=2,col='red')
     mtext(text = 'year',1,line=3,cex=1.4)
@@
 
-  source('scripts/create_subsamples.R')
-  aaa=calculate_average_geometric_growth(pop$popvalue)
   plot(rowMeans(aaa,na.rm=T),cex.axis=1.2,pch=16,ylab='',xlab='',las=1)
@@
     mtext(text = 'population growth rate',2,line=3.7,cex=1.4)
+
+    abline(h= -0.01*rowMeans(aaa,na.rm = T)[34] +rowMeans(aaa,na.rm = T)[34] )
+    abline(h= 0.01*rowMeans(aaa,na.rm = T)[34] +rowMeans(aaa,na.rm = T)[34] )
+
+    arrows(1:34,rowMeans(aaa,na.rm=T),1:34,rowMeans(aaa,na.rm=T)+sd(apply(aaa,MARGIN = 1,FUN=var,na.rm=T),na.rm=T),angle = 90,length = 0.1)
+    arrows(1:34,rowMeans(aaa,na.rm=T),1:34,rowMeans(aaa,na.rm=T)-sd(apply(aaa,MARGIN = 1,FUN=var,na.rm=T),na.rm=T),angle = 90,length = 0.1)
+
+    plot(percent_error(rowMeans(aaa,na.rm=T),rowMeans(aaa,na.rm=T)[34]),cex.axis=1.2,pch=16,ylab='',xlab='',las=1)
+    mtext(text = 'years sampled',1,line=3,cex=1.4)
+    mtext(text = 'percent error in growth rate',2,line=3.7,cex=1.4)
+    abline(h=0.1,lty=2)
+    abline(v=tail(which(percent_error(rowMeans(aaa,na.rm=T),rowMeans(aaa,na.rm=T)[34])>0.1),1)+1,lty=2)
 ```
+
+We applied the same calculations to `r nrow(pop_info)` populations, as we did in the main text.
+We then obtain a distribution of the minimum time required to measure the ``true" longterm growth rate (Fig.
+\reg{fig:min_time_growth_dist}).
+We see a bimodal distribution with many populations required 30+ years to estimate the longterm growth rate.
+The large number of short year required is due to cases where the entire time series is consistently increasing or decreasing at the same rate.
+
+```{r, echo=F,fig.cap='Histogram of the minimum time required in order to estimate the longterm growth rate within 20% error.\\label{fig:min_time_growth_dist}'}
+# Growth rates calculations for each species
+setwd("/Users/eastonwhite/Desktop/Research/time-series-project")
+load('cleaned-data/pop_info_example_goodNAs.Rdata')
+
+ for (j in 1:nrow(pop_info)){
+    pop= subset(long_dat,long_dat$ID==pop_info$ID[j])[1:35,]
+    pop$popvalue=as.numeric(pop$popvalue)
+    pop$popvalue=pop$popvalue/max(pop$popvalue) #should I standaridize this in a better way
+
+   # source('scripts/calculate_power_metric.R')
+    aaa=calculate_average_geometric_growth(pop$popvalue)
+    if (length(which(percent_error(rowMeans(aaa,na.rm=T),rowMeans(aaa,na.rm=T)[34])>0.2))==0){
+      pop_info$min_time_growth_rate[j] = 2
+    }else{
+      pop_info$min_time_growth_rate[j] = tail(which(percent_error(rowMeans(aaa,na.rm=T),rowMeans(aaa,na.rm=T)[34])>0.2),1)+1}
+    #print(paste(j,':',pop_info$min_time_for_power[j],sep=' '))
+ }
+
+# Create plot
+par(mfrow=c(1,1))
+hist(pop_info$min_time_growth_rate,las=1,ylab='',xlab='',cex.axis=1.2,main='',breaks=20,xlim=c(0,40),ylim=c(0,500))
+mtext(text = 'Frequency',2,line=3,cex=1.2)
+mtext(text = 'Minimum time required',1,line=3,cex=1.2)
+
+```
+
+\clearpage
 
@@
 
-TEXT HERE ON IDEA
+In the main text, we examined the minimum time required to identify a trend in abundance via linear regression.
+This approach allowed us to identify increases or decreases, but a linear model may not always be a good fit.
+Generalized additive models (GAMs) are more general than general linear models and allow more flexible.
+GAMs are models where a response variable depends on unknown smooth functions of explanatory variables.
+GAMs, therefore, can identify relationships between response variables and explantory variables that are non-linear and perhaps more complicated.
+The downsite of GAMs...
+
+Here, we conduct the same analyses in the main text, but instead calculate the minimum time series required to detect trends over time according to a GAM model.
+We hypothesize that GAMs should require less time to detect a trend as they are more flexible than linear regression.
+We provide an example that shows statistical power increases with more time sampled (Fig.
+\ref{fig:gam_example}).
 
@@
 # show gam example side-by-side plot
+plot(1,1)
 ```
+
+We then fit GAM models for XX populations. We found... (Fig. \ref{fig:gam_result}). This is shorter/longer than the main text question.
 
@@
 # show gam results for minimium time required
+setwd("/Users/eastonwhite/Desktop/Research/time-series-project")
+load('cleaned-data/pop_info_example_goodNAs.Rdata')
+
+ for (j in 1:nrow(pop_info)){
+    pop= subset(long_dat,long_dat$ID==pop_info$ID[j])[1:35,]
+    pop$popvalue=as.numeric(pop$popvalue)
+    pop$popvalue=pop$popvalue/max(pop$popvalue) #should I standaridize this in a better way
+
+    source('scripts/calculate_power_metric_gam.R')
+    pop_info$min_time_for_power_gam[j] = min_time_needed(pop$popvalue,0.05,0.8)
+    print(paste(j,':',pop_info$min_time_for_power[j],sep=' '))
+ }
+
+# Create plot
+par(mfrow=c(1,1))
+hist(pop_info$min_time_growth_rate,las=1,ylab='',xlab='',cex.axis=1.2,main='',breaks=20,xlim=c(0,40),ylim=c(0,500))
+mtext(text = 'Frequency',2,line=3,cex=1.2)
+mtext(text = 'Minimum time required',1,line=3,cex=1.2)
+
 ```
 
+### Minimum time to detect density-dependence in time series
+
+For decades, the importance of density-dependence in natural populations has been an important question in ecology (citations).
+To address this question, @Knape2012 used data from the Global Population Dynamics Database [@GPDD2010].
+They fitted 627 time series with a Gompertz population model and found evidence for density-dependence in 45% of the time series.
+Each time series in their database contained at least 15 datapoints, and ranged from 15 to 157.
+A follow up question to their work would be, what time series length is required to confidently estimate density-dependence.
+
+\clearpage
 ### References

### 2017-06-05 writing/supp_mat_draft.tex
changes to figures and supp mat

@@
 \item
-  Mathematical description of our model
+  Detailed example of subsampling and power calculations
 \item
-  Inverse pattern-orientated estimation procedure
+  Figures and results from other questions of interest
+\item
+  Additional figures?
+\item
+  Additional references
 \end{enumerate}
 
-Code for all the figures and tables can be found on
-\href{https://github.com/erwhite1}{Github}.
+Code for all the figures and tables can be found at
+\href{https://github.com/erwhite1}{}.
 
@@
 
+Instead of detecting a trend over time with linear regression, we could
+also calculate the geometric growth rate of the population. In figure
+\ref{fig:growth_rate}, we show how to calculate growth rates for
+subsamples of a time series. First, we created subsamples of each
+possible length from the full 35 year time series, as we did in the main
+next. Next, we calculated the mean and standard deviation of growth
+rates for each possible time series length (Fig.
+\ref{fig:growth_rate}b). Lastly, we calculated the percent error
+\(\mbox{percent error} = 100 \times \left| \frac{\mbox{observed} - \mbox{theoretical}}{\mbox{theoretical}} \right|\)
+between the mean of each time series length (observed) and the overall
+population growth rate (theoretical). In Fig. \ref{fig:growth_rate}c),
+we show the percent error as a function of time series length. Here we
+define the minimum time required as the minimum number of years to
+achieve less than 20\% error.
+
 \begin{verbatim}
@@
 \includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-2-1.pdf}
-\caption{duh\label{fig:growth_rate}}
+\caption{Example of calculating minimum time required for growth rate
+estimation. (a) abundance over time, (b) mean and standard deviation of
+growth rate for subsamples of entire time series, and (c) the percent
+error between mean estimated growth rate and the true longterm growth
+rate. The vertical bar denotes the minimum time required to estimate
+growth rate within 10\% error.\label{fig:growth_rate}}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-3-1.pdf}
+\caption{Histogram of the minimum time required in order to estimate the
+longterm growth rate within 20\% error.\label{fig:min_time_dist}}
 \end{figure}
@@
 
-TEXT HERE ON IDEA
+In the main text, we examined the minimum time required to identify a
+trend in abundance via linear regression. This approach allowed us to
+identify increases or decreases, but a linear model may not always be a
+good fit. Generalized additive models (GAMs) are more general than
+general linear models and allow more flexible. GAMs are models where a
+response variable depends on unknown smooth functions of explanatory
+variables. GAMs, therefore, can identify relationships between response
+variables and explantory variables that are non-linear and perhaps more
+complicated. The downsite of GAMs\ldots{}
+
+Here, we conduct the same analyses in the main text, but instead
+calculate the minimum time series required to detect trends over time
+according to a GAM model. We hypothesize that GAMs should require less
+time to detect a trend as they are more flexible than linear regression.
+We provide an example that shows statistical power increases with more
+time sampled (Fig. \ref{fig:gam_example}).
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-4-1.pdf}
+\caption{results here from GAM.\label{fig:gam_example}}
+\end{figure}
+
+We then fit GAM models for XX populations. We found\ldots{} (Fig.
+\ref{fig:gam_result}). This is shorter/longer than the main text
+question.
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
+\caption{results here from GAM.\label{fig:gam_result}}
+\end{figure}
+
+\subsubsection{Minimum time to detect density-dependence in time
+series}\label{minimum-time-to-detect-density-dependence-in-time-series}
+
+For decades, the importance of density-dependence in natural populations
+has been an important question in ecology (citations). To address this
+question, \citet{Knape2012} used data from the Global Population
+Dynamics Database \citep{GPDD2010}. They fitted 627 time series with a
+Gompertz population model and found evidence for density-dependence in
+45\% of the time series. Each time series in their database contained at
+least 15 datapoints, and ranged from 15 to 157. A follow up question to
+their work would be, what time series length is required to confidently
+estimate density-dependence.
 

### 2017-06-05 scripts/calculate_p_value_gam.R
modified code to make GAM calculations

@@ -1,0 +1,16 @@
+
+
+# Created by Easton White
+# Created on 17-Jan-2017
+# Last edited on 11-May-2017
+
+# Calculate p value for GAM model
+
+#SHould I use gaussian or poisson family? Poisson does not work when I have non-count data
+require(mgcv)
+calculate_p_value_gam = function(pop){
+  smoothing_parameter=3
+  year=1:length(pop)
+  return(summary(gam(pop ~ s(year,k=smoothing_parameter),family=gaussian))$s.table[4])
+}
+

### 2017-06-05 scripts/calculate_power_metric_gam.R

@@ -1,0 +1,53 @@
+
+
+
+# Created by Easton White
+# Created on 17-Jan-2017
+# Last edited on 2-June-2017
+
+
+# Functions for calculating power metric and minimum time series required for GAM example
+
+# Calculate power for series of subsamples (what fraction of subsamples have trened that is significant - dictated by alpha)
+require(plyr)
+
+#pull in neccessary functions
+source('scripts/calculate_p_value_gam.R')
+source('scripts/create_subsamples.R')
+
+
+######## Implement two-tailed test
+calculate_power_for_subsamples = function(pop,alpha){
+  subset_times=create_subsamples(pop)
+  power=matrix(0,nrow=(length(pop)-2),1)
+  for (n in 3:(length(pop)-2)){
+    power[n,1]=sum(apply(na.omit(subset_times[[n]]),1,calculate_p_value_gam)<alpha)/(length(pop) - n)
+  }
+  return(power)
+}  
+
+
+
+#######
+
+
+
+
+# give population time series, alpha, and a threshold for power (usually 0.8), calculate the minimum time needed to guarentee time series reaches required levels of alpha and power
+min_time_needed = function(pop, alpha,threshold){
+  alpha=alpha
+  power = calculate_power_for_subsamples(pop,alpha)
+  min_value = tail(which(abs((power - as.numeric(tail(power,1)))/as.numeric(tail(power,1)))>(1-threshold)),1)+1
+  
+  if (length(min_value)==0){
+    return(2)
+  }else if (min_value ==length(pop)-2){
+    return(-999)
+  }else{
+    return(min_value)
+  }
+}
+
+
+
+

### 2017-06-05 scripts/percent_error.R
wrote function to calculate percent error

@@ -1,0 +1,10 @@
+
+
+
+## Percent error function
+
+
+percent_error = function(observation, theoretical){
+  
+  return(100*abs((observation - theoretical) / theoretical))
+}

### 2017-06-05 scripts/scripts_to_clean_data/LPI_data_analysis.R
created folder for code to clean data

@@ -1,0 +1,47 @@
+# Created by Easton White
+# Created on 17-Jan-2017
+# Last edited on 18-Jan-2017
+
+# This code reads in data from the () and cleans it up (mostly making it into a long table format)
+
+library("tidyr")
+library("dplyr")
+
+LPI = read.table('data/LPI_LPR2016data_public_EW_edited.csv',sep=',',header=T)
+
+
+LPI_long <- LPI %>%
+  gather(year, pop_value, starts_with('X'))
+
+LPI_long$year=as.numeric(sub('X','',LPI_long$year))
+
+#head(LPI_long[sort(LPI_long$ID),],50)
+
+newdata=arrange(LPI_long,ID,year)
+
+testing=newdata[,c(1,2,3,10,11,19,20)]
+testing[testing=='NULL']=NA
+#removed=na.omit(testing)
+
+long_term_series_names=names(sort(table(removed$ID),decreasing=T)[sort(table(removed$ID),decreasing=T)>35])
+
+y=subset(removed,removed$ID %in% long_term_series_names)
+
+
+for (species_code in long_term_series_names[1]){
+  pop = subset(removed,removed$ID==species_code)
+  
+  
+}
+
+
+
+# Plots to make
+load('example_min_time_LPI_data.Rdata')
+par(mfrow=c(1,1))
+plot(jitter(pop_info$Latitude,amount=2),pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),xlim=c(0,80),las=1,ylab='',xlab='',cex.axis=1.2)
+mtext(text = 'minimum time required',side = 2,line = 3,cex = 1.4)
+mtext(text = 'latitude',side = 1,line = 3,cex = 1.4)
+
+plot(jitter(pop_info$Longitude,amount=2),pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5))
+

### 2017-06-05 scripts/scripts_to_clean_data/combine_databases.R

@@ -1,0 +1,79 @@
+# Created by Easton White
+# Created on 17-Jan-2017
+# Last edited on 18-Jan-2017
+
+
+# This is a script to try and combine multiple population series time together to use biological information from Amniote Database in my other analyses
+
+# Population time series file
+load('For_easton.Rdata')
+dat= long_dat
+
+# File for biological characteristics of amniotes
+amniote = read.table('data/Amniote_Database_Aug_2015_erw.csv',sep=',',header=T,na.strings = -999)
+amniote$Binomial = paste(amniote$genus,amniote$species,sep=' ')
+
+# Create bio_info table to store informatino where a species is present in both databases
+bio_info = as.data.frame(matrix(NA,nrow=nrow(dat), ncol=ncol(amniote[,7:35])))
+colnames(bio_info)=names(amniote[,7:35])
+
+dat = cbind(dat,bio_info)
+dat$match =0
+
+for (j in 1:nrow(dat)){
+  if (dat$Binomial[j] %in% amniote$Binomial){
+    dat[j,9:37] = amniote[which(amniote$Binomial %in% dat$Binomial[j]),7:35]
+    dat[j,38]=1
+  }
+    if (j %in% seq(1000,80000,by=5000)){print(j)} #just a counter
+}
+
+# remove coluoms with lots of NA (egg_width_mm forward) #sort(colSums(is.na(dat)),decreasing=T)
+few_NAs=names(sort(colSums(is.na(dat)),decreasing=F))[1:12]
+new_dat = dat[,which(names(dat) %in% few_NAs)]
+#new_dat = new_dat[,-which(names(new_dat) %in% c('year_ass'))]
+
+test=na.omit(new_dat)
+length(table(test$ID)) #number of time series
+
+# Which biological variables are most available for long time series?
+long_time_names=names(table(test$ID))[as.numeric(table(test$ID))>30]
+long_dat = subset(test,test$ID %in% long_time_names)
+length(table(long_dat$ID) )
+
+
+pop_info= long_dat[1,]
+for (j in 1:length(table(long_dat$ID))){
+  index=as.numeric(which(long_dat$ID==long_time_names[j])[1])
+  pop_info[j,] = long_dat[index,]
+}
+
[…]

### 2017-06-05 scripts/scripts_to_clean_data/create-pop-info-table.R

@@ -1,0 +1,73 @@
+# Created by Easton White
+# Created on 17-Jan-2017
+# Last edited on 18-Jan-2017
+
+
+#This code loads in data and then creates a table called pop_info which includes population information on each unique time series
+
+# chuck for loading in data, removing NAs, looking at only long time series
+# also calculates a pop_info table which includes basic information on each population that I later analyze
+
+# Pull in original data from Keith et al 2015.
+load("data/Keith_et_al_2015_data.Rdata")
+
+dat=dat[,c(1:7,11)] #
+
+# This script removes time series with NAs present. This may take a few moments
+source('scripts/scripts_to_clean_data/remove_NA_populations.R')
+
+
+#this subsets data to find long time series (e.g. >=35 years)
+#dat=subset(dat,dat$class != 'Aves')
+long_time_names=names(table(dat$ID))[as.numeric(table(dat$ID))>34]
+long_dat = subset(dat,dat$ID %in% long_time_names)
+length(table(long_dat$ID) )
+
+
+#this creates pop_info table for each analyzed pop time series
+pop_info= long_dat[1,]
+  for (j in 1:length(table(long_dat$ID))){
+    index=as.numeric(which(long_dat$ID==long_time_names[j])[1])
+    pop_info[j,] = long_dat[index,]
+  }
+
+#add variables to pop_info table
+pop_info$series_length = as.numeric(table(long_dat$ID))
+pop_info$variance=NA
+pop_info$overall_trend=NA
+pop_info$autocorrelation= NA
+pop_info$trend_p_value=NA
+pop_info$min_time_for_power=NA
+pop_info$popvalue=NA
+long_dat$pop_value=long_dat$pop_value
+source('scripts/calculate_slope.R')
+  for (j in 1:nrow(pop_info)){
+    pop= subset(long_dat,long_dat$ID==pop_info$ID[j])
+    pop = pop[(nrow(pop)-34):nrow(pop),]
+    pop$popvalue=as.numeric(pop$popvalue)
+    pop$popvalue=pop$popvalue/max(pop$popvalue) #should I standaridize this in a better way
+    pop_info$variance[j]= var(pop$popvalue)
+    pop_info$overall_trend[j] = calculate_slope(pop$popvalue)
+    pop_info$autocorrelation[j] = acf(pop$popvalue,plot = FALSE)$acf[2]
[…]

### 2017-06-05 scripts/scripts_to_clean_data/remove_NA_populations.R

@@ -1,0 +1,28 @@
+# Created by Easton White
+# Created on 17-Jan-2017
+# Last edited on 18-Jan-2017
+
+
+# Remove records on interior of time series for each individual population
+#load('data/Keith_et_al_2015_data.Rdata')
+#dat=dat[,c(1:7,11)] #
+
+ID_codes = names(table(dat$ID))
+good_rows = NULL
+
+for (j in 1:length(ID_codes)){
+  #pop = subset(dat,dat$ID==ID_codes[j])[1:35,]
+  pop = subset(dat,dat$ID==ID_codes[j])
+  if (nrow(na.omit(pop))>34){
+    first_none_NA = head(which(is.na(as.numeric(pop$popvalue))==F),1)
+    last_none_NA = tail(which(is.na(as.numeric(pop$popvalue))==F),1)
+    pop=pop[first_none_NA:last_none_NA,]
+      if (nrow(pop) == nrow(na.omit(pop))){
+        good_rows=cbind(good_rows,row.names(pop))
+      }
+  }
+ # print(j)
+}
+
+dat = dat[row.names(dat) %in% c(good_rows),]
+

### 2017-06-05 writing/supp_mat_draft.Rmd
edited supp mat to include cleaned data

@@
 source("scripts/calculate_geometric_growth.R")
-load('cleaned-data/pop_info_example_goodNAs.Rdata')
+load('cleaned-data/cleaned_timeseries_database.Rdata')
 
@@
 setwd("/Users/eastonwhite/Desktop/Research/time-series-project")
-load('cleaned-data/pop_info_example_goodNAs.Rdata')
+load('cleaned-data/cleaned_timeseries_database.Rdata')
 

### 2017-06-09 writing/manuscript_draft.Rmd
shortened abstract and set double spacing

@@
     \usepackage{float}
-    \usepackage{setspace}             
+    \usepackage{setspace}\doublespacing           
     \usepackage{lineno}
@@
 
-\begin{spacing}{1.5} %use 2
-	\begin{flushleft}
 Running head: 
@@
 
-\end{flushleft}
-\end{spacing}
-
 \linenumbers
-
-\clearpage
 
@@
 Long-term census data is necessary to better understand ecological processes and make management decisions.
-In addition, census data is often expensive, requiring a lot of people-hours or equipment.
-However, little work has actually addressed the length of time series required.
-In other words, when is a time series of census data long enough to address a question of interest?
+However, census data is often expensive, requiring a lot of people-hours and equipment.
+Little work has addressed the question, when is a time series of census data long enough to address a question of interest?
 Here, I explore two approaches to address this question: one simulation-based and an empirical approach.
 I specifically determine the minimum time series length required to estimate significant increases or decreases in population abundance.
-Importantly, I examine the ability to detect a trend with a set level of statistical power, which is often neglected in ecological time series analyses.
-Using simple simulations, I demonstrate how the minimum time series length required increases with weaker trends in abundance and with higher variability in population size.
+<!--Importantly, I examine the ability to detect a trend with a set level of statistical power, which is often neglected in ecological time series analyses.--> Using a simulation approach, I demonstrate how the minimum time series length required increases with weaker trends in abundance and with higher variability in population size.
 In addition, I examine 868 populations of vertebrate species to determine the minimum time required to detect changes in their abundance.
-Here I show 10-15 years of continuous monitoring are required in order to achieve a high level of statistical power.
+I show that 10-15 years of continuous monitoring are required in order to achieve a high level of statistical power.
 Similar to simulation-approaches, the minimum time required for field census data strongly depends on trend strength, population variability, and temporal autocorrelation.
-Perhaps surprisingly, the minimum time required did not correlate well with biological explanatory variables, like body size or generation length.
-These results point to the importance of sampling populations over long periods of time.
-Here, I stress the need for verifying a study has enough statistical power to accurately determine long-term changes in population abundance.
-Most studies, especially those less than 10-15 years, are probably underpowered and potentially misleading.
+<!--Perhaps surprisingly, the minimum time required did not correlate well with biological explanatory variables, like body size or generation length.--> These results point to the importance of sampling populations over long periods of time.
+<!--Here, I stress the need for verifying a study has enough statistical power to accurately determine long-term changes in population abundance.--> Most studies, especially those less than 10-15 years, are probably underpowered and potentially misleading.
 

### 2017-06-09 writing/manuscript_draft.tex
shortened abstract and set double spacing

@@
 
-\usepackage{float} \usepackage{setspace} \usepackage{lineno}
+\usepackage{float}
+\usepackage{setspace}\doublespacing \usepackage{lineno}
 \usepackage[round]{natbib} \bibpunct[; ]{(}{)}{,}{a}{}{;} \linenumbers
@@
 
-\begin{spacing}{1.5} %use 2
-    \begin{flushleft}
-Running head: 
+Running head: \vspace{3 mm}
+
+Title: Your time series is (probably) too short \vspace{7 mm}
+
+Author(s): \textsc{Easton R. White$^{1,2}$} \vspace{3 mm}
+
+Addresses: \textbackslash{}
+\emph{$^1$ Center for Population Biology, University of California-Davis, Davis, California 95616 USA}
 \vspace{3 mm}
 
-Title: Your time series is (probably) too short
-\vspace{7 mm}
-
-Author(s): \textsc{Easton R. White$^{1,2}$} 
+\(^2\)Corresponding author:
+\href{mailto:eawhite@ucdavis.edu}{\nolinkurl{eawhite@ucdavis.edu}}
 \vspace{3 mm}
 
-Addresses: \\ \emph{$^1$ Center for Population Biology, University of California-Davis, Davis, California 95616 USA}
-\vspace{3 mm}
+Number of words: X,XXX \vspace{3 mm}
 
-$^2$Corresponding author: eawhite@ucdavis.edu
-\vspace{3 mm}
-
-Number of words: X,XXX 
-\vspace{3 mm}
-
-To be submitted to: \emph{XXXXX} (Article)
-\vspace{3 mm}
+To be submitted to: \emph{XXXXX} (Article) \vspace{3 mm}
 
@@
 Long-term census data is necessary to better understand ecological
-processes and make management decisions. In addition, census data is
-often expensive, requiring a lot of people-hours or equipment. However,
-little work has actually addressed the length of time series required.
-In other words, when is a time series of census data long enough to
-address a question of interest? Here, I explore two approaches to
-address this question: one simulation-based and an empirical approach. I
-specifically determine the minimum time series length required to
-estimate significant increases or decreases in population abundance.
-Importantly, I examine the ability to detect a trend with a set level of
-statistical power, which is often neglected in ecological time series
-analyses. Using simple simulations, I demonstrate how the minimum time
-series length required increases with weaker trends in abundance and
-with higher variability in population size. In addition, I examine 868
-populations of vertebrate species to determine the minimum time required
-to detect changes in their abundance. Here I show 10-15 years of
-continuous monitoring are required in order to achieve a high level of
-statistical power. Similar to simulation-approaches, the minimum time
+processes and make management decisions. However, census data is often
+expensive, requiring a lot of people-hours and equipment. Little work
+has addressed the question, when is a time series of census data long
+enough to address a question of interest? Here, I explore two approaches
+to address this question: one simulation-based and an empirical
+approach. I specifically determine the minimum time series length
+required to estimate significant increases or decreases in population
+abundance. Using a simulation approach, I demonstrate how the minimum
+time series length required increases with weaker trends in abundance
+and with higher variability in population size. In addition, I examine
+868 populations of vertebrate species to determine the minimum time
+required to detect changes in their abundance. I show that 10-15 years
+of continuous monitoring are required in order to achieve a high level
+of statistical power. Similar to simulation-approaches, the minimum time
 required for field census data strongly depends on trend strength,
-population variability, and temporal autocorrelation. Perhaps
-surprisingly, the minimum time required did not correlate well with
-biological explanatory variables, like body size or generation length.
-These results point to the importance of sampling populations over long
-periods of time. Here, I stress the need for verifying a study has
-enough statistical power to accurately determine long-term changes in
-population abundance. Most studies, especially those less than 10-15
-years, are probably underpowered and potentially misleading.
+population variability, and temporal autocorrelation. These results
+point to the importance of sampling populations over long periods of
+time. Most studies, especially those less than 10-15 years, are probably
+underpowered and potentially misleading.
 

### 2017-06-13 writing/manuscript_draft.Rmd
moved results from ms to supp mat and expanded supp mat text

@@
 ---
+bibliography: White_bib.bib
+fontsize: 12pt
+geometry: margin=1in
+header-includes: \usepackage{float} \usepackage{setspace}\doublespacing \usepackage{lineno}
+  \usepackage[round]{natbib} \bibpunct[; ]{(}{)}{,}{a}{}{;} \linenumbers
 output:
@@
     keep_tex: yes
-    number_sections: true
-fontsize: 12pt
-geometry: margin=1in
-bibliography: White_bib.bib
-header-includes: 
-    \usepackage{float}
-    \usepackage{setspace}\doublespacing           
-    \usepackage{lineno}
-    \usepackage[round]{natbib}
-    \bibpunct[; ]{(}{)}{,}{a}{}{;}
-    \linenumbers
+    number_sections: yes
+  word_document: default
 ---
@@
 
-Addresses: \\ \emph{$^1$ Center for Population Biology, University of California-Davis, Davis, California 95616 USA}
+Addresses:\emph{$^1$ Center for Population Biology, University of California-Davis, Davis, California 95616 USA}
 \vspace{3 mm}
@@
 
-Observational studies and population censuses have become a cornerstone of modern ecology and conservation biology [@Magurran2010].
-Longterm data is necessary to understand population dynamics and to assess extinction risk of species.
+Observational studies and population censuses have become a cornerstone of modern ecological research and conservation biology [@Magurran2010].
+Longterm data is necessary to both understand population dynamics and to assess extinction risk of species.
 Even though many time series may now be considered "longterm", most are still short (i.e.
@@
 They are often coupled with an experiment, which may only last a couple of years.
-In addition, short funding cycles make it difficult to examine populations over longer periods of time.
+In addition, short funding cycles make it difficult to examine populations over longer periods of time (citations).
 However, there are of course notable exceptions including the Isle Royale study, Iowa corn study, cod, lynx-hair, and many others.
@@
 For example, the IUCN Red List Categories and Criteria suggest under Criterian A2, a species qualifies as vulnerable if it has experienced a 30% decline over 10 years, or 3 generations [@IUCN2012].
-In addition, because sampling is typically expensive, we do not want to sample for longer than is neccessary [@Gerber1999].
+For both scientific and management questions, because sampling is typically expensive, we do not want to sample for longer than is neccessary [@Gerber1999].
 
@@
 
-For time series data, two general approaches can be used.
+<!--For time series data, two general approaches can be used.
 First, as can be done with experimental data, we can relax the definition of statistical power (cite Bolker).
@@
 Subsamples of different lengths can then be evaluated to see which fraction of subsamples meet some criteria, again a measure of statistical power.
-Similar to the simulation approach, we can use these measures of power to determine the minimum time series required for a particular question of interest.
+Similar to the simulation approach, we can use these measures of power to determine the minimum time series required for a particular question of interest.-->
 
-Past work has investigated questions related to the minimum time series required to estimate trends in population size over time.
-However, most work has either focused on theoretical approaches or by examined a single population.
-As an example of the former, @Rhodes2011 examined the optimal allocation of effort between spatial and temporal replicates.
+Past work has investigated questions related to the minimum time series required to estimate trends in population size over time [@Giron-Nava2017].
+@Rhodes2011 examined the optimal allocation of effort between spatial and temporal replicates.
 Using simple populations models, they found that the allocation of effort depends on environmental variation, spatial and temporal autocorrelation, and observer error.
@@
 In this paper, we use both simulations and a database of empirical time series to determine the minimum number of years required to address various questions.
-To demonstrate these techniques, we address three specific questions: (1) what is the minimum time series length required to identify significant changes over time using linear regression, (2)
+In the main manuscript, we determine the minimum time series length to determine if a population is incresing or decreasing according to a significant slope coefficient in simple linear regression.
 <!-- In this paper, we evaluate the minimal time series length required to answer X questions.
@@
 $$
-N(t + 1) = N(t) + r(t) \mbox{ with } r(t) \sim N(\phi \mu, \sigma)
+N(t + 1) = N(t) + r(t) \mbox{ with } r(t) \sim N(\mu, \sigma)
 $$
-where $\epsilon$ is a normally-distributed random noise term with mean $\mu$ and standard deviation $\sigma$.
-The rate of growth $r$ is also the trend strength of increase or decrease.
-<!--Lastly, $\phi$ represents an autocorrelation term where $\phi=0$ indicates no autocorrelation, $\phi>0$ is positively autocorrelated.
-This model represents linear growth with autocorrelation and noise.
-However, it is certainly possible to use any other population model.-->
+where $\epsilon$ is a normally-distributed random noise term with mean $\mu$ and standard deviation $\sigma$. The rate of growth $r$ is also the trend strength of increase or decrease.
 
@@
 
-In figure \ref{fig:theoretical_approach}a, a number of simulated time series are shown for a set number of time periods (t=40).
-It is clear that statistical power increases quickly with increases in $t$ (Fig.
+In Figure \ref{fig:theoretical_approach}a, a number of simulated time series are shown for a set number of time periods ($t=40$).
+It is clear that statistical power increases quickly with increases in length of time sampled (Fig.
 \ref{fig:theoretical_approach}b).
-Once power is greater than XX (the dotted line), that is the minimum time required ($t_{min}$
+Once power is greater than 0.8 (the dotted line), that is the minimum time required ($T_{min}$).
 
@@
   # Plotting commands
-  par(mfrow=c(1,2),oma=c(1,2,0,0.5),mar=c(4,5,0.5,0.5))
+  par(mfrow=c(2,2),oma=c(1,2,0,0.5),mar=c(4,5,0.5,0.5))
     matplot(simulations,type='l',col=rgb(0.5,0.5,0.5,0.5),lty=1,ylab='',xlab='',las=1,cex.axis=1.2)
@@
 
+    # Plot versus trend strength
+      r_values = seq(1,3,by=0.2)
+      sigma_values=seq(1,7,by=0.2)
+
+      parameter_combinations = expand.grid(r_values,5)
+      parameter_combinations$min_time_required = NA
+      names(parameter_combinations)= c('r','sigma','min_time_required')
+
+      for (i in 1:nrow(parameter_combinations)){
+        years_to_sample=seq(2,50,by=1)
+        y=lapply(X=years_to_sample,FUN = calculate_power,phi =0.5,r=parameter_combinations$r[i],sigma = parameter_combinations$sigma[i],trials=100)
+        if (length(tail(which(y<0.8),1))>0){
+        parameter_combinations$min_time_required[i] = years_to_sample[tail(which(y<0.8),1)] + 1
+        }
+        #print(i)
+      }
+
+      plot(parameter_combinations$r,parameter_combinations$min_time_required,type='l',lwd=2,las=1,ylab='',xlab='',cex.axis=1.2)
+      mtext(text = 'trend strength (slope coefficient)' ,side = 1,cex = 1.2,line=3)
+      mtext(text = 'minimum time required' ,side = 2,cex = 1.2,line=3.5)
+
+      # Plot versus sigma values
+      parameter_combinations = expand.grid(1.5,sigma_values)
+      parameter_combinations$min_time_required = NA
+      names(parameter_combinations)= c('r','sigma','min_time_required')
+
+      for (i in 1:nrow(parameter_combinations)){
+        years_to_sample=seq(2,50,by=1)
+        y=lapply(X=years_to_sample,FUN = calculate_power,phi =0.5,r=parameter_combinations$r[i],sigma = parameter_combinations$sigma[i],trials=100)
+        if (length(tail(which(y<0.8),1))>0){
+        parameter_combinations$min_time_required[i] = years_to_sample[tail(which(y<0.8),1)] + 1
+        }
+        #print(i)
+      }
+
+      plot(parameter_combinations$sigma,parameter_combinations$min_time_required,type='l',lwd=2,las=1,ylab='',xlab='',cex.axis=1.2)
+      mtext(text = 'population variability' ,side = 1,cex = 1.2,line=3)
+      mtext(text = 'minimum time required' ,side = 2,cex = 1.2,line=3.5)
 ```
@@
   load('cleaned-data/pop_info_example_goodNAs.Rdata') #Do i need to filter out non-significant populations as well (where no trend was present)
+  pop_info=pop_info[pop_info$trend_p_value<0.05,]
+  long_dat=subset(long_dat,long_dat$ID %in% pop_info$ID)
+  long_time_names=subset(long_time_names,long_time_names %in% pop_info$ID)
 ```
@@
 
+```{r,echo=F}
+# Pull in results of biological characteristics comparison to min time series length required
+```
 For a subset of populations in the database (X = XX), we were able to connect a database on biological characteristics like body size and generation length. We used
@@
 
-Our objective is to determine the minimum time series length ($T_{min}$) required to answer a specific question of interest.
+<!--Our objective is to determine the minimum time series length ($T_{min}$) required to answer a specific question of interest.
 As we describe below, the minimum time series length calculation is particular to each question.
-Intuitively, we might expect a longer minimum time series length ($T_{min}$) required given more complicated questions (citations).
+Intuitively, we might expect a longer minimum time series length ($T_{min}$) required given more complicated questions (citations).-->
 
@@
 We first take all possible contiguous subsamples of each time series.
-For example, a time series of 40 years would have 39 possible contiguous subsamples of length 2, 38 possible contiguous subsamples of length 3, and continueing until 1 possible contiguous subsample of length 40.
-In this paper, we will try to ask a simple, specific question.
-We seek to determine the $T_{min}$ required to be confident in an estimate of the slope coefficient from linear regression.
+For example, a time series of 40 years would have 39 possible contiguous subsamples of length 2, 38 possible contiguous subsamples of length 3, and continuing until 1 possible contiguous subsample of length 40 [@Gerber1999, @Giron-Nava2017].
+Then we determined the $T_{min}$ required to be confident (i.e.
+have high enough statistical power) in an estimate of the slope coefficient from linear regression.
 Therefore, we run a linear regression for each subsampled time series.
@@
 
-We also examine how minimum time series calculations changed depending on the question of interest.
-In the supplementary material, we show how to make similar calculations to determine the minimum time series required to estimate overall population growth rate, the variability in population size, and future extinction risk.
-Again, we took subsamples of time series data to determine $T_{min}$ for each of these questions.
-
-```{r, echo=F,message=F,warning=F}
+```{r, echo=F,message=F,warning=F,fig.cap='(a) Population size of ANIMAL NAME over time.
+The red line is the best fit line from linear regresssion.
+(b) statistical power for different subsets of the time series in panel a.\\label{fig:empirical_approach_example}'}
 # Example of subsampling empirical data
@@
   par(mfrow=c(1,2))
-pop = subset(long_dat,long_dat$ID=='10696') #13192 for Harp seals
-plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='population size',xlab='time (years)',pch=16,ylim=c(0,2000),las=1)
+    pop = subset(long_dat,long_dat$ID=='13192') #13192 for Harp seals
+    plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='population size',xlab='time (years)',pch=16,las=1)
     abline(lm(pop$popvalue~pop$year),lwd=2,col='red')
+    mtext(text = '(a)',side = 3,line = -1.2,at = 1965,font = 2,cex=1.2)
 # could insert all possible regressions of length n
@@
     power_example=calculate_power_for_subsamples(pop$popvalue,0.05)
-plot(power_example,cex.axis=1.2,cex.lab=1.2,ylab='average statistical power',xlab='years sampled',las=1,pch=16)
+    plot(power_example,cex.axis=1.2,cex.lab=1.2,ylab='statistical power',xlab='years sampled',las=1,pch=16)
+    mtext(text = '(b)',side = 3,line = -1.2,at = 7,font = 2,cex=1.2)
 # could also insert standard error bars
     abline(h=0.8,col='red',lwd=2,lty=2)
-abline(v=22,col='red',lwd=2,lty=2)
+    abline(v=head(which(power_example>0.8),1),col='red',lwd=2,lty=2)
 
 ```
+
+We also examine how minimum time series calculations changed depending on the question of interest.
+In the supplementary material, we show how to make similar calculations to determine the minimum time series required to estimate overall population growth rate, the variability in population size, and future extinction risk.
+Again, we took subsamples of time series data to determine $T_{min}$ for each of these questions.
 
@@
 
-We evaluated the question of what minimum time series is required to answer a question of interest.
+We determined the minimum time series length required to address a particular question of interest.
 More specifically we asked, what is the minimum time series length required to determine, via linear regression, the longterm trend in population abundance?
@@
 We investigated this question through two approaches: simulation-based and empirical analyses.
-In both cases, we found the minimum time series length required depended strongly on the trend strength, population variability, and autocorrelation.
+We briefly desribe results from the simulation approach and then discuss our emprirical approach.
 
@@
 We chose a simple model, but any other population model could be used.
-The specific model choice should be tailored to the population of interest, based on its ecology.
-We then ran simulations of different lengths to determine the minimum time series length required to achieve a certain level of power (which we set to 0.8).
+The specific model choice should be tailored to the population of interest.
+We then ran simulated time series of different lengths.
+From these simulations we determined the minimum time series length required to achieve a certain level of power (which we set to 0.8).
 We were also able to alter the parameter values (trend strength and population variability) of the population model to see how this altered $T_{min}$.
 
-In line with past work, we found the $T_{min}$ increases (i.e. more time is required) with decreases in trend strength or autocorrelation and with increases in variability (citation).
+In line with past work, we found the $T_{min}$ increases (i.e. more time is required) with decreases in trend strength and with increases in variability (citation).
 
@@
 ```{r poisson_model, echo=F,include=F}
+# I should put this in the appendix
 #Poisson regression
@@
   mtext('| trend strength |',1,line=3,cex=1.2)
-  mtext('(a)',3,line=-1.5,at=0.033,cex=1.2,font = 2)
+  mtext('(a)',3,line=-1.5,at=0.0318,cex=1.2,font = 2)
     plot(pop_info$variance,pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
@@
       mtext('autocorrelation',1,line=3,cex=1.2)
-       mtext('(c)',3,line=-1.5,at=0.98,cex=1.2,font = 2)
+       mtext('(c)',3,line=-1.5,at=0.97,cex=1.2,font = 2)
        mtext('minimum time required',2,line=2.5,cex=1.2,outer = T)
@@
 In addition, as expected based on theoretical results (cite papers), the minimum time series length required is strongly correlated with trend strength, population variance, and autocorrelation (Fig \ref{fig:correlates}).
-Using a generalized linear model, with a Poisson error structure, all three of these explanatory variables were significant and had large effect sizes (cite model from supp mat here).
-Combined trend strength, population variance, and autocorrelation account for about `r round(100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance),1)` percent of the variance in minimum time series length required (as proposed by CITE Zuur).
+Using a generalized linear model, with a Poisson error structure, all three of these explanatory variables were significant and had large effect sizes (see Fig.
+SX).
+Combined trend strength, population variance, and autocorrelation account for about `r round(100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance),1)` of the explained deviance, which is similar to $R^2$ [@Zuur2009], in minimum time series length required.
 
@@
 None (need to do regression) of these explanatory variables were predictive of the minimum time series length required (supp mat).
-Some differences if $t_{min}$ did occur between biological class (supp mat), but this can be attributed to the amount populations are known to fluctuate (i.e.
+Some differences if $T_{min}$ did occur between biological class (supp mat), but this can be attributed to the amount populations are known to fluctuate (i.e.
 the variability in population size).
 
-```{r,echo=F}
+```{r,echo=F,include=FALSE, eval=FALSE,fig.cap='(a) Minimum time required to estimate change in abundance for species class, (b) Longterm trend (estimated slope coefficient) by species class, (c) interannual variability in population size by species class, and (d) temporal autocorrelation by species class.\\label{fig:biological_correlates}'}
 # Biological correlates plots
@@
 
-```{r,echo=F}
+```{r,echo=F,include=FALSE,eval=FALSE,fig.cap='Minimum time series required versus the generation length of each species\\label{fig:generation_length}'}
 par(mfrow=c(1,1))
@@
 The simulation-based approach has been suggested by others, especially in situations where analysis is more complicated than a simple power analysis [@Gerrodette1987].
-These simulations support past work...
+Our simulations support past work...
 
@@
 
-First we asked, what is $t_{min}$ for determining whether a population was increasing or decreasing over time according to linear regression?
+First we asked, what is $T_{min}$ for determining whether a population was increasing or decreasing over time according to linear regression?
 We found that for `r nrow(pop_info)` populations, at least 10-15 years of continuous monitoring were neccessary (Fig.
 \ref{fig:fig:min_time_dist}).
-In line with theoretical predictions (citations), we also found $t_{min}$ was strongly correlated with the trend strength, variability in population size, and temporal autocorrelation (cite figure).
+In line with theoretical predictions (citations), we also found $T_{min}$ was strongly correlated with the trend strength, variability in population size, and temporal autocorrelation (cite figure).
 
-We also found that $t_{min}$ did not correlate with any biological variables of interest (cite figure or supp mat).
+We also found that $T_{min}$ did not correlate with any biological variables of interest (cite figure or supp mat).
 We initially hypothesized that species with longer lifespans or generation times may require more years to sample.
@@
 
-In the appendix we asked two additional questions to make estimates of $t_{min}$. First, we asked how many years of sampling are neccessary to estimate a geometric growth rate?
+In the appendix we asked two additional questions to make estimates of $T_{min}$. First, we asked how many years of sampling are neccessary to estimate a geometric growth rate?
 
@@
 Our work has obvious limitations in determining the minimum time series length required.
-First, $t_{min}$ depends strongly on the question being asked and the availability of long-term census data.
+First, $T_{min}$ depends strongly on the question being asked and the availability of long-term census data.
 For our empirical approach, the subsampling of the full time series allows for estimates of power, but the individual subsamples are clearly not independent of one another.
@@
 Clearly, this is not always practical, especially if running analyses for a wide array of species as we do here.
-In addition, our statistical models suggest that $t_{min}$ does not correlate with any biological variables of interest, at least for the question of linear regression.
-Therefore, it is not possible to use these results to predict $t_{min}$ for another population, even if the population is of a species with a similar life-history to one in our database.
+In addition, our statistical models suggest that $T_{min}$ does not correlate with any biological variables of interest, at least for the question of linear regression.
+Therefore, it is not possible to use these results to predict $T_{min}$ for another population, even if the population is of a species with a similar life-history to one in our database.
 
@@
 \ref{fig:min_time_dist}.
-In line with theoretical predictions, we also show that $t_{min}$ is strongly correlated with the overall trend in abundance, variability in abundance, and the temporal autocorrelation (Fig.
+In line with theoretical predictions, we also show that $T_{min}$ is strongly correlated with the overall trend in abundance, variability in abundance, and the temporal autocorrelation (Fig.
 \ref{fig:correlates}).
@@
 This stresses the importance of longterm monitoring programs.
-(where do IUCN recommendations come from?) From both a scientific and management perspective estimates of $t_{min}$ are important.
+(where do IUCN recommendations come from?) From both a scientific and management perspective estimates of $T_{min}$ are important.
 If a time series is too short, we are likely to make incorrect conclusions about longterm trends.
@@
 
-- list each bit of supporting information. Github reference
+In the supporting material, we provide an expanded methods sections, additional figures, and additional analyses of other time series questions.
+All code and data can be found at www.github.com/erwhite1
 

### 2017-06-13 writing/manuscript_draft.tex
moved results from ms to supp mat and expanded supp mat text

@@
 
-Addresses: \textbackslash{}
-\emph{$^1$ Center for Population Biology, University of California-Davis, Davis, California 95616 USA}
+Addresses:\emph{$^1$ Center for Population Biology, University of California-Davis, Davis, California 95616 USA}
 \vspace{3 mm}
@@
 Observational studies and population censuses have become a cornerstone
-of modern ecology and conservation biology (Magurran et al. 2010).
-Longterm data is necessary to understand population dynamics and to
-assess extinction risk of species. Even though many time series may now
-be considered ``longterm'', most are still short (i.e.~less than X
-years). \emph{I could have statistic on how long is considered
+of modern ecological research and conservation biology (Magurran et al.
+2010). Longterm data is necessary to both understand population dynamics
+and to assess extinction risk of species. Even though many time series
+may now be considered ``longterm'', most are still short (i.e.~less than
+X years). \emph{I could have statistic on how long is considered
 ``longterm'' in papers}. Time series are typically short for a variety
@@
 last a couple of years. In addition, short funding cycles make it
-difficult to examine populations over longer periods of time. However,
-there are of course notable exceptions including the Isle Royale study,
-Iowa corn study, cod, lynx-hair, and many others.
+difficult to examine populations over longer periods of time
+(citations). However, there are of course notable exceptions including
+the Isle Royale study, Iowa corn study, cod, lynx-hair, and many others.
 
@@
 experienced a 30\% decline over 10 years, or 3 generations (IUCN 2012).
-In addition, because sampling is typically expensive, we do not want to
-sample for longer than is neccessary (Gerber, DeMaster, and Kareiva
-1999).
+For both scientific and management questions, because sampling is
+typically expensive, we do not want to sample for longer than is
+neccessary (Gerber, DeMaster, and Kareiva 1999).
 
@@
 
-For time series data, two general approaches can be used. First, as can
-be done with experimental data, we can relax the definition of
-statistical power (cite Bolker). Instead of using classic statistical
-tests (e.g.~t-test), and corresponding formulas for determinig sample
-size, simulation approaches can be used (Gerrodette 1987). Simple
-population models can be simulated with parameter values corresponding
-to a population of interest. Out of a large number of simulations, power
-can be calculated as the fraction of simulations that meet some
-criteria. The specific criteria depend on the question at hand. Then
-time series simulations of different lengths are run to determine the
-minimum time series length required to be confident that you will be
-able to answer your question (cite Bolker). In addition to using
-simulated data, empirical time series can also be used. We usually do
-not have multiple replicates of similar populations, but we can
-subsample an empirical time series (cite). Subsamples of different
-lengths can then be evaluated to see which fraction of subsamples meet
-some criteria, again a measure of statistical power. Similar to the
-simulation approach, we can use these measures of power to determine the
-minimum time series required for a particular question of interest.
-
 Past work has investigated questions related to the minimum time series
-required to estimate trends in population size over time. However, most
-work has either focused on theoretical approaches or by examined a
-single population. As an example of the former, Rhodes and Jonzen (2011)
-examined the optimal allocation of effort between spatial and temporal
-replicates. Using simple populations models, they found that the
-allocation of effort depends on environmental variation, spatial and
-temporal autocorrelation, and observer error. Rueda-Cediel et al. (2015)
-also used a modeling approach, but parameterized it for a threatened
-snail, \emph{Tasmaphena lamproides}. They found that for this
+required to estimate trends in population size over time. Rhodes and
+Jonzen (2011) examined the optimal allocation of effort between spatial
+and temporal replicates. Using simple populations models, they found
+that the allocation of effort depends on environmental variation,
+spatial and temporal autocorrelation, and observer error. Rueda-Cediel
+et al. (2015) also used a modeling approach, but parameterized it for a
+threatened snail, \emph{Tasmaphena lamproides}. They found that for this
 short-lived organims, that 15 years was adequate to assess longterm
@@
 series to determine the minimum number of years required to address
-various questions. To demonstrate these techniques, we address three
-specific questions: (1) what is the minimum time series length required
-to identify significant changes over time using linear regression, (2)
+various questions. In the main manuscript, we determine the minimum time
+series length to determine if a population is incresing or decreasing
+according to a significant slope coefficient in simple linear
+regression.
 
@@
 In figure \ref{fig:theoretical_approach}a, a number of simulated time
-series are shown for a set number of time periods (t=40). It is clear
-that statistical power increases quickly with increases in \(t\) (Fig.
-\ref{fig:theoretical_approach}b). Once power is greater than XX (the
-dotted line), that is the minimum time required (\(t_{min}\)
+series are shown for a set number of time periods (\(t=40\)). It is
+clear that statistical power increases quickly with increases in length
+of time sampled (Fig. \ref{fig:theoretical_approach}b). Once power is
+greater than XX (the dotted line), that is the minimum time required
+(\(t_{min}\)). We can also examine how the minimum time required changes
+with effect size and variability (\textbf{make graph of this here})
 
@@
 
-Our objective is to determine the minimum time series length
-(\(T_{min}\)) required to answer a specific question of interest. As we
-describe below, the minimum time series length calculation is particular
-to each question. Intuitively, we might expect a longer minimum time
-series length (\(T_{min}\)) required given more complicated questions
-(citations).
-
 We assume that each time series in our database is long enough to
@@
 subsamples of length 2, 38 possible contiguous subsamples of length 3,
-and continueing until 1 possible contiguous subsample of length 40. In
-this paper, we will try to ask a simple, specific question. We seek to
-determine the \(T_{min}\) required to be confident in an estimate of the
-slope coefficient from linear regression. Therefore, we run a linear
-regression for each subsampled time series. Then, we determine the
-fraction of subsamples of a particular length that have estimated slope
-coefficients which are statistically different from zero given the
-longterm, or ``true``, time series also has a significant slope. This is
-a measure of statistical power (often denoted as \(1- \beta\), where
-\(\beta\) is the probability of a type II error). Lastly, we determine
-which subsample length is required to guarantee we have above a certain
-threshold of statistical power, which we will assume is 0.8 by
-convention (citation). A power of 0.8 would imply that given a trend
-exists, there is a 0.8 probability of detecting the trend.
-Complementary, there would be a 0.2 probability of a false negative.
+and continueing until 1 possible contiguous subsample of length 40. Then
+we determined the \(T_{min}\) required to be confident (i.e.~have high
+enough statistical power) in an estimate of the slope coefficient from
+linear regression. Therefore, we run a linear regression for each
+subsampled time series. Then, we determine the fraction of subsamples of
+a particular length that have estimated slope coefficients which are
+statistically different from zero given the longterm, or ``true``, time
+series also has a significant slope. This is a measure of statistical
+power (often denoted as \(1- \beta\), where \(\beta\) is the probability
+of a type II error). Lastly, we determine which subsample length is
+required to guarantee we have above a certain threshold of statistical
+power, which we will assume is 0.8 by convention (citation). A power of
+0.8 would imply that given a trend exists, there is a 0.8 probability of
+detecting the trend. Complementary, there would be a 0.2 probability of
+a false negative.
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-4-1.pdf}
+\caption{(a) Population size of ANIMAL NAME over time. The red line is
+the best fit line from linear regresssion. (b) statistical power for
+different subsets of the time series in panel
+a.\label{fig:empirical_approach_example}}
+\end{figure}
 
@@
 
-\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-3-1.pdf}
-
 \subsection{Results}\label{results}
 
-We evaluated the question of what minimum time series is required to
-answer a question of interest. More specifically we asked, what is the
+We determined the minimum time series length required to address a
+particular question of interest. More specifically we asked, what is the
 minimum time series length required to determine, via linear regression,
@@
 investigated this question through two approaches: simulation-based and
-empirical analyses. In both cases, we found the minimum time series
-length required depended strongly on the trend strength, population
-variability, and autocorrelation.
+empirical analyses. We briefly desribe results from the simulation
+approach and then discuss our emprirical approach.
 
@@
 population model could be used. The specific model choice should be
-tailored to the population of interest, based on its ecology. We then
-ran simulations of different lengths to determine the minimum time
-series length required to achieve a certain level of power (which we set
-to 0.8). We were also able to alter the parameter values (trend strength
-and population variability) of the population model to see how this
-altered \(T_{min}\).
+tailored to the population of interest. We then ran simulated time
+series of different lengths. From these simulations we determined the
+minimum time series length required to achieve a certain level of power
+(which we set to 0.8). We were also able to alter the parameter values
+(trend strength and population variability) of the population model to
+see how this altered \(T_{min}\).
 
 In line with past work, we found the \(T_{min}\) increases (i.e.~more
-time is required) with decreases in trend strength or autocorrelation
-and with increases in variability (citation).
+time is required) with decreases in trend strength and with increases in
+variability (citation).
 
@@
 \centering
-\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-4-1.pdf}
+\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
 \caption{Distribution of the minimum time required in order to detect a
@@
 \centering
-\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
+\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-6-1.pdf}
 \caption{Minimum time required to estimate change in abundance
@@
 three of these explanatory variables were significant and had large
-effect sizes (cite model from supp mat here). Combined trend strength,
-population variance, and autocorrelation account for about 70.8 percent
-of the variance in minimum time series length required (as proposed by
-CITE Zuur).
+effect sizes (see Fig. SX). Combined trend strength, population
+variance, and autocorrelation account for about 70.8 of the explained
+deviance, which is similar to \(R^2\) (Zuur et al. 2009), in minimum
+time series length required.
 
@@
 
-\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-6-1.pdf}
+\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-7-1.pdf}
 
-\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-7-1.pdf}
+\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-8-1.pdf}
 
@@
 in situations where analysis is more complicated than a simple power
-analysis (Gerrodette 1987). These simulations support past work\ldots{}
+analysis (Gerrodette 1987). Our simulations support past work\ldots{}
 
@@
 
+\hypertarget{ref-Zuur2009}{}
+Zuur, Alain F., Elena N. Ieno, Neil J. Walker, Anatoly A. Saveliev, and
+Graham M. Smith. 2009. \emph{Mixed Effects Models and Extensions in
+Ecology with R}. New York: Springer.
+
 \end{document}

### 2017-06-13 writing/supp_mat_draft.Rmd
moved results from ms to supp mat and expanded supp mat text

@@
 1. Detailed example of subsampling and power calculations
-2. Figures and results from other questions of interest
-3. Additional figures?
+2. Additional figures
+3. Figures and results from other questions of interest
 4. Additional references
@@
 ```
+
+## Additional results from the main manuscript
+
+```{r, echo=F}
+setwd("~/Desktop/Research/time-series-project")
+load('cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
+```
+
+In the main manuscript, we examined the minimum time required to detect a significant trend in abundance over time using linear regression.
+As detailed in the main manuscript the minimum time required was around 15, but there was a wide distribution.
+Therefore, we were interested in potential explanatory variables of the minimum time required.
+In the main manuscript, we examined characteristics of the time series itself, like variability, autocorrelation, and the trend in abundance over time.
+Here, we combined our time series database with a database on life history characteristics of amniotes from (cite other database here).
+There was life history information available for `r nrow(LPI_pop_info)` populations representing `r length(table(LPI_pop_info$Binomial))` different species, all of which were in the Aves class.
+
+We then correlated minimum time required for each population with its corresponding life history characteristics.
+In figure \ref{fig:biological_correlates} we examined miminmum time required versus generation length (years), litter size (n), adult body mass (grams), maximum longevity (years), egg mass (grams), and incubation (days).
+None of these variables had much explanatory power in accounting for the variance in the minimum time required.
+see table \ref{table:model_output}}
+
+```{r, echo=F, fig.cap='Mimimum time required versus (a) generation length (years), (b) litter size (n), (c) adult body mass (grams), (d) maximum longevity (years), (e) egg mass (grams), and (f) incubation (days).
+The lines in each plot represent the best fit line from linear regression.\\label{fig:biological_correlates}',fig.height=6}
+
+par(mfrow=c(3,2),oma=c(5,4,1,1),mar=c(5,3,0,0))
+
+n=1
+x_axis_name_list = c('generation length (years)','litter size (n)','adult body mass (grams)','maximum longevity (years)','egg mass (grams)','incubation (days)')
+for (x_axis_name in names(LPI_pop_info)[20:25]){
+  plot(LPI_pop_info[[x_axis_name]],LPI_pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
+  abline(lm(LPI_pop_info$min_time_for_power~LPI_pop_info[[x_axis_name]]),lwd=3)
+
+  mtext(text = x_axis_name_list[n],side = 1,line = 3,cex = 1.2)
+  mtext(text = paste('(',letters[n],')',sep=''),side = 3,line=-1.5,adj = 0.98,cex=1.2)
+  n=n+1
+}
+
+mtext(text = 'minimum time required',side = 2,line = 2,cex = 1.4,outer = TRUE)
+
+```
+
+\pagebreak
+
+In the main text, we explained how the minimum time required strongly correlated with the trend strength, temporal autocorrelation, and variance in population size.
+Here we use a generalized linear model framework with a Poisson error structure to determine drivers of the minimum time required.
+In figure, \ref{fig:poisson_model} we show a set of residual plots for the regression.
+We then show the coefficient estimates and levels of significance in table \ref{table:model_output}.
+
+```{r poisson_model, echo=F,include=T,fig.cap='Output of poisson regression...\\label{fig:poisson_model}'}
+setwd("~/Desktop/Research/time-series-project")
+load("cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
+
+# GLM with Poisson error structure (include all terms - it is best not to include abs_overall_trend with variance as they are strongly corrrelated to one another)
+pop_info$abs_overall_trend = abs(pop_info$overall_trend)
+poisson_model <-glm(min_time_for_power~ abs_overall_trend + autocorrelation + variance + gen_len,family = 'poisson',data=pop_info)
+
+  par(mfrow=c(2,2),mar=c(5,5,1,2),oma=c(0,0,0.5,0))
+    plot(poisson_model,cex.lab=1.2,las=1,cex.axis=1.2,main=' ',pch=16, col=rgb(0.5,0.5,0.5,0.4))
+#R^2 calculation (pg. 218 Zuur et al. 2009 mixed effects book)
+R2_fullmodel <- 100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance)
+
+poisson_model_no_variance <-glm(min_time_for_power~ abs_overall_trend + autocorrelation + gen_len,family = 'poisson',data=pop_info)
+R2_no_variance <- 100*(poisson_model_no_variance$null.deviance - poisson_model_no_variance$deviance)/(poisson_model_no_variance$null.deviance)
+
+poisson_model_no_trend <-glm(min_time_for_power~ autocorrelation + variance + gen_len,family = 'poisson',data=pop_info)
+R2_no_trend <- 100*(poisson_model_no_trend$null.deviance - poisson_model_no_trend$deviance)/(poisson_model_no_trend$null.deviance)
+
+```
+
+```{r,echo=F,message=F,warning=F}
+
+require(knitr)
+knitr::kable(summary(poisson_model)$coef,caption = 'Stuff here\\label{table:model_output}')
+
+```
+
+Our model with trend strength, autocorrelation, variance, and generation length accounted for `r round(R2_fullmodel,2)`% of the variation in the minimum time required.
+However, we also found trend strength and variance to be strongly correlated with one another.
+Therefore, we ran two additional models with either trend strength or variance, but not both together.
+This resulted in lower explained deviance (analogue to $R^2$) of `r round(R2_no_variance,2)`% and `r round(R2_no_trend,2)`%, respectively.
+
+\pagebreak
 
@@
 We then obtain a distribution of the minimum time required to measure the ``true" longterm growth rate (Fig.
-\reg{fig:min_time_growth_dist}).
+\ref{fig:min_time_growth_dist}).
 We see a bimodal distribution with many populations required 30+ years to estimate the longterm growth rate.
@@
 
-```{r, echo=F,fig.cap='Histogram of the minimum time required in order to estimate the longterm growth rate within 20% error.\\label{fig:min_time_growth_dist}'}
+```{r, echo=F,eval=F,fig.cap='Histogram of the minimum time required in order to estimate the longterm growth rate within 20% error.\\label{fig:min_time_growth_dist}'}
 # Growth rates calculations for each species
@@
 
-```{r,echo=F,fig.cap='results here from GAM.\\label{fig:gam_result}'}
+```{r,echo=F,eval=F,fig.cap='results here from GAM.\\label{fig:gam_result}'}
 # show gam results for minimium time required

### 2017-06-13 writing/supp_mat_draft.tex
moved results from ms to supp mat and expanded supp mat text

@@
 \bibliographystyle{plainnat}
+\usepackage{longtable,booktabs}
 \usepackage{graphicx,grffile}
@@
 \item
+  Additional figures
+\item
   Figures and results from other questions of interest
-\item
-  Additional figures?
 \item
@@
 
+\subsection{Additional results from the main
+manuscript}\label{additional-results-from-the-main-manuscript}
+
+In the main manuscript, we examined the minimum time required to detect
+a significant trend in abundance over time using linear regression. As
+detailed in the main manuscript the minimum time required was around 15,
+but there was a wide distribution. Therefore, we were interested in
+potential explanatory variables of the minimum time required. In the
+main manuscript, we examined characteristics of the time series itself,
+like variability, autocorrelation, and the trend in abundance over time.
+Here, we combined our time series database with a database on life
+history characteristics of amniotes from (cite other database here).
+There was life history information available for 547 populations
+representing 315 different species, all of which were in the Aves class.
+
+We then correlated minimum time required for each population with its
+corresponding life history characteristics. In figure
+\ref{fig:biological_correlates} we examined miminmum time required
+versus generation length (years), litter size (n), adult body mass
+(grams), maximum longevity (years), egg mass (grams), and incubation
+(days). None of these variables had much explanatory power in accounting
+for the variance in the minimum time required. see table
+\ref{table:model_output}\}
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-3-1.pdf}
+\caption{Mimimum time required versus (a) generation length (years), (b)
+litter size (n), (c) adult body mass (grams), (d) maximum longevity
+(years), (e) egg mass (grams), and (f) incubation (days). The lines in
+each plot represent the best fit line from linear
+regression.\label{fig:biological_correlates}}
+\end{figure}
+
+\pagebreak
+
+In the main text, we explained how the minimum time required strongly
+correlated with the trend strength, temporal autocorrelation, and
+variance in population size. Here we use a generalized linear model
+framework with a Poisson error structure to determine drivers of the
+minimum time required. In figure, \ref{fig:poisson_model} we show a set
+of residual plots for the regression. We then show the coefficient
+estimates and levels of significance in table \ref{table:model_output}.
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{supp_mat_draft_files/figure-latex/poisson_model-1.pdf}
+\caption{Output of poisson regression\ldots{}\label{fig:poisson_model}}
+\end{figure}
+
+\begin{longtable}[]{@{}lrrrr@{}}
+\caption{Stuff here\label{table:model_output}}\tabularnewline
+\toprule
+& Estimate & Std. Error & z value &
+Pr(\textgreater{}\textbar{}z\textbar{})\tabularnewline
+\midrule
+\endfirsthead
+\toprule
+& Estimate & Std. Error & z value &
+Pr(\textgreater{}\textbar{}z\textbar{})\tabularnewline
+\midrule
+\endhead
+(Intercept) & 3.7141297 & 0.0274350 & 135.379203 &
+0.0000000\tabularnewline
+abs\_overall\_trend & -104.4974288 & 3.0997732 & -33.711314 &
+0.0000000\tabularnewline
+autocorrelation & -0.4905888 & 0.0431062 & -11.380936 &
+0.0000000\tabularnewline
+variance & 18.7756472 & 0.6409982 & 29.291262 & 0.0000000\tabularnewline
+gen\_len & 0.0053584 & 0.0025995 & 2.061352 & 0.0392695\tabularnewline
+\bottomrule
+\end{longtable}
+
+Our model with trend strength, autocorrelation, variance, and generation
+length accounted for 72.55\% of the variation in the minimum time
+required. However, we also found trend strength and variance to be
+strongly correlated with one another. Therefore, we ran two additional
+models with either trend strength or variance, but not both together.
+This resulted in lower explained deviance (analogue to \(R^2\)) of
+53.08\% and 45.04\%, respectively.
+
+\pagebreak
+
 \subsection{Results from other questions of
@@
 \centering
-\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-2-1.pdf}
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
 \caption{Example of calculating minimum time required for growth rate
@@
 
-\begin{figure}[htbp]
-\centering
-\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-3-1.pdf}
-\caption{Histogram of the minimum time required in order to estimate the
-longterm growth rate within 20\% error.\label{fig:min_time_dist}}
-\end{figure}
+We applied the same calculations to 1032 populations, as we did in the
+main text. We then obtain a distribution of the minimum time required to
+measure the ``true" longterm growth rate (Fig.
+\ref{fig:min_time_growth_dist}). We see a bimodal distribution with many
+populations required 30+ years to estimate the longterm growth rate. The
+large number of short year required is due to cases where the entire
+time series is consistently increasing or decreasing at the same rate.
+
+\clearpage
 
@@
 \centering
-\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-4-1.pdf}
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-7-1.pdf}
 \caption{results here from GAM.\label{fig:gam_example}}
@@
 question.
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
-\caption{results here from GAM.\label{fig:gam_result}}
-\end{figure}
 
@@
 
-\subsubsection{References}\label{references}
+\clearpage
+\#\#\# References
 

### 2017-06-13 scripts/combine_databases.R
modified code to combine cleaned time series database and amniote life history database

@@ -1,79 +1,0 @@
-# Created by Easton White
-# Created on 17-Jan-2017
-# Last edited on 18-Jan-2017
-
-
-# This is a script to try and combine multiple population series time together to use biological information from Amniote Database in my other analyses
-
-# Population time series file
-load('For_easton.Rdata')
-dat= long_dat
-
-# File for biological characteristics of amniotes
-amniote = read.table('data/Amniote_Database_Aug_2015_erw.csv',sep=',',header=T,na.strings = -999)
-amniote$Binomial = paste(amniote$genus,amniote$species,sep=' ')
-
-# Create bio_info table to store informatino where a species is present in both databases
-bio_info = as.data.frame(matrix(NA,nrow=nrow(dat), ncol=ncol(amniote[,7:35])))
-colnames(bio_info)=names(amniote[,7:35])
-
-dat = cbind(dat,bio_info)
-dat$match =0
-
-for (j in 1:nrow(dat)){
-  if (dat$Binomial[j] %in% amniote$Binomial){
-    dat[j,9:37] = amniote[which(amniote$Binomial %in% dat$Binomial[j]),7:35]
-    dat[j,38]=1
-  }
-    if (j %in% seq(1000,80000,by=5000)){print(j)} #just a counter
-}
-
-# remove coluoms with lots of NA (egg_width_mm forward) #sort(colSums(is.na(dat)),decreasing=T)
-few_NAs=names(sort(colSums(is.na(dat)),decreasing=F))[1:12]
-new_dat = dat[,which(names(dat) %in% few_NAs)]
-#new_dat = new_dat[,-which(names(new_dat) %in% c('year_ass'))]
-
-test=na.omit(new_dat)
-length(table(test$ID)) #number of time series
-
-# Which biological variables are most available for long time series?
-long_time_names=names(table(test$ID))[as.numeric(table(test$ID))>30]
-long_dat = subset(test,test$ID %in% long_time_names)
-length(table(long_dat$ID) )
-
-
-pop_info= long_dat[1,]
-for (j in 1:length(table(long_dat$ID))){
-  index=as.numeric(which(long_dat$ID==long_time_names[j])[1])
-  pop_info[j,] = long_dat[index,]
-}
-
[…]

### 2017-06-13 scripts/scripts_to_clean_data/combine_databases.R
modified scripts to clean table and to combine databases

@@ -1,15 +1,16 @@
 # Created by Easton White
 # Created on 17-Jan-2017
-# Last edited on 18-Jan-2017
+# Last edited on 10-Jun-2017
 
 
 # This is a script to try and combine multiple population series time together to use biological information from Amniote Database in my other analyses
 
-# Population time series file
-load('For_easton.Rdata')
+makePlot=FALSE
+# Load cleaned population time series database (n=878). This includes a column for min time required to estimate significant slope (alpha<0.05) coefficient for linear regression with power>0.8
+load('cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata')
 dat= long_dat
 
-# File for biological characteristics of amniotes
+# File for biological characteristics of amniotes (where is this from again?)
 amniote = read.table('data/Amniote_Database_Aug_2015_erw.csv',sep=',',header=T,na.strings = -999)
 amniote$Binomial = paste(amniote$genus,amniote$species,sep=' ')
 
@@ -18,8 +19,9 @@ bio_info = as.data.frame(matrix(NA,nrow=nrow(dat), ncol=ncol(amniote[,7:35])))
 colnames(bio_info)=names(amniote[,7:35])
 
 dat = cbind(dat,bio_info)
-dat$match =0
+dat$match =0 # Create column to see if data matches
 
+# inefficient way to create matches between the two databases
 for (j in 1:nrow(dat)){
   if (dat$Binomial[j] %in% amniote$Binomial){
     dat[j,9:37] = amniote[which(amniote$Binomial %in% dat$Binomial[j]),7:35]
@@ -28,41 +30,35 @@ for (j in 1:nrow(dat)){
     if (j %in% seq(1000,80000,by=5000)){print(j)} #just a counter
 }
 
-# remove coluoms with lots of NA (egg_width_mm forward) #sort(colSums(is.na(dat)),decreasing=T)
-few_NAs=names(sort(colSums(is.na(dat)),decreasing=F))[1:12]
+# remove columns with lots of NA (e.g. )
+few_NAs=names(sort(colSums(is.na(dat)),decreasing=F))[1:15]
 new_dat = dat[,which(names(dat) %in% few_NAs)]
[…]

### 2017-06-13 scripts/scripts_to_clean_data/create-pop-info-table.R

@@ -46,16 +46,16 @@ source('scripts/calculate_slope.R')
     pop = pop[(nrow(pop)-34):nrow(pop),]
     pop$popvalue=as.numeric(pop$popvalue)
     pop$popvalue=pop$popvalue/max(pop$popvalue) #should I standaridize this in a better way
-    pop_info$variance[j]= var(pop$popvalue)
-    pop_info$overall_trend[j] = calculate_slope(pop$popvalue)
-    pop_info$autocorrelation[j] = acf(pop$popvalue,plot = FALSE)$acf[2]
-    pop_info$trend_p_value[j]=calculate_p_value(pop$popvalue)
-    pop_info$popvalue = pop$popvalue[1] #initial abundance
-    pop_info$series_length = nrow(pop)
+    # pop_info$variance[j]= var(pop$popvalue)
+    # pop_info$overall_trend[j] = calculate_slope(pop$popvalue)
+    # pop_info$autocorrelation[j] = acf(pop$popvalue,plot = FALSE)$acf[2]
+    # pop_info$trend_p_value[j]=calculate_p_value(pop$popvalue)
+    # pop_info$popvalue = pop$popvalue[1] #initial abundance
+    # pop_info$series_length = nrow(pop)
     
-   # source('scripts/calculate_power_metric.R')
-   #  pop_info$min_time_for_power[j] = min_time_needed(pop$popvalue,0.05,0.8)
-   # print(paste(j,':',pop_info$min_time_for_power[j],sep=' '))
+   source('scripts/calculate_power_metric.R')
+     pop_info$min_time_for_power[j] = min_time_needed(pop$popvalue,0.05,0.8)
+    print(paste(j,':',pop_info$min_time_for_power[j],sep=' '))
   }
 
 #IUCN calculation

### 2017-06-13 scripts/gam_example.R
modified script to run GAM code

@@ -1,9 +1,15 @@
 
 
 # Example gam
-gam.object <- gam(popvalue ~ s(year,6),data=moose,family=poisson)
+require(gam)
+gam.object <- gam(popvalue ~ s(year,3),data=short_moose,family=poisson)
 summary(gam.object)
 plot(gam.object,se=TRUE)
+points(short_moose$year,(short_moose$popvalue-mean(short_moose$popvalue))/mean(short_moose$popvalue),col='red',pch=16)
+
+
+short_short_moose=short_moose[1:10]
+
 gam.newdata = as.data.frame(c(1950:2020))
 names(gam.newdata)='year'
 predict(gam.object,type="terms",newdata=gam.newdata)

### 2017-06-13 scripts/theoretical-models/theoretical-approach.R

@@ -11,14 +11,14 @@
 
 
 min_time_needed= function(sigma){
-  years_to_sample=seq(2,30,by=1)
+  years_to_sample=seq(2,100,by=1)
   y=lapply(X=years_to_sample,FUN = calculate_power,phi =0.5,r=0.5,sigma = sigma,trials=100)
   min_time_needed = years_to_sample[tail(which(y<0.8),1)] + 1
   return(min_time_needed)
 }
 
 
-r_values = seq(0.1,1.5,by=0.1)
+r_values = seq(1,3,by=0.5)
 phi_values = seq(0.1,0.9,by=0.1)
 sigma_values=seq(1,12,by=0.5)
 zed=lapply(X=sigma_values,FUN = min_time_needed)
@@ -26,13 +26,13 @@ sigma_zed=zed
 
 ##########################################################
 # Run models from r values and sigma values to make nice plot
-parameter_combinations = expand.grid(r_values,sigma_values)
+parameter_combinations = expand.grid(r_values,5)
 parameter_combinations$min_time_required = NA
 names(parameter_combinations)= c('r','sigma','min_time_required')
 
 for (i in 1:nrow(parameter_combinations)){
-  years_to_sample=seq(2,25,by=1)
-  y=lapply(X=years_to_sample,FUN = calculate_power,phi =0.5,r=parameter_combinations$r[i],sigma = parameter_combinations$sigma[i],trials=150 )
+  years_to_sample=seq(2,50,by=1)
+  y=lapply(X=years_to_sample,FUN = calculate_power,phi =0.5,r=parameter_combinations$r[i],sigma = parameter_combinations$sigma[i],trials=50)
   if (length(tail(which(y<0.8),1))>0){
   parameter_combinations$min_time_required[i] = years_to_sample[tail(which(y<0.8),1)] + 1
   }

### 2017-06-14 writing/supp_mat_draft.Rmd
cleaned up code for supp mat

@@
 
-Our model with trend strength, autocorrelation, variance, and generation length accounted for `r round(R2_fullmodel,2)`% of the variation in the minimum time required.
+Our model with trend strength, autocorrelation, variance, and generation length accounted for `r round(R2_fullmodel,2)`% of the variation in the minimum time required (Table \ref{table:model_output}).
 However, we also found trend strength and variance to be strongly correlated with one another.
@@
 ## Results from other questions of interest
+
+### Simulations with more complicated population model
+
+In the main text, we showed how a simple population model could be simulated repeatidly to estimate the power obtained with time series of increasing length.
+The model in the main text simulated linear population growth with only a slope coefficient, y-intercept, and noise parameter required.
+This model is purely phenemilogical and does not include any species life history.
+Here, we use the same routine as the main text, but simulate from a more biologically-realistic population model, the (MODEL NAME HERE):
+
+  \begin{equation}
+    y = x
+  \end{equation}
+
+We simulate this model repetively for different lengths of time series.
+Statistical power is the fraction of time series of a set length that showed signicant trends in abundance, given a true trend in abundance.
+*How do I determine significance here?
+Through linear regression again, or do I look at non-zero estimation of the parameter values?*
+
+### Empirical approach with more complicated population model (use model similar to de Valpine work?)
 
@@
 GAMs, therefore, can identify relationships between response variables and explantory variables that are non-linear and perhaps more complicated.
-The downsite of GAMs...
+The downside of GAMs is they typically require more data and are also prone to overfitting (citations).
 

### 2017-06-14 writing/supp_mat_draft.tex
cleaned up code for supp mat

@@
 length accounted for 72.55\% of the variation in the minimum time
-required. However, we also found trend strength and variance to be
-strongly correlated with one another. Therefore, we ran two additional
-models with either trend strength or variance, but not both together.
-This resulted in lower explained deviance (analogue to \(R^2\)) of
-53.08\% and 45.04\%, respectively.
+required (Table \ref{table:model_output}). However, we also found trend
+strength and variance to be strongly correlated with one another.
+Therefore, we ran two additional models with either trend strength or
+variance, but not both together. This resulted in lower explained
+deviance (analogue to \(R^2\)) of 53.08\% and 45.04\%, respectively.
 
@@
 interest}\label{results-from-other-questions-of-interest}
+
+\subsubsection{Simulations with more complicated population
+model}\label{simulations-with-more-complicated-population-model}
+
+In the main text, we showed how a simple population model could be
+simulated repeatidly to estimate the power obtained with time series of
+increasing length. The model in the main text simulated linear
+population growth with only a slope coefficient, y-intercept, and noise
+parameter required. This model is purely phenemilogical and does not
+include any species life history. Here, we use the same routine as the
+main text, but simulate from a more biologically-realistic population
+model, the (MODEL NAME HERE):
+
+\begin{equation}
+    y = x
+  \end{equation}
+
+We simulate this model repetively for different lengths of time series.
+Statistical power is the fraction of time series of a set length that
+showed signicant trends in abundance, given a true trend in abundance.
+\emph{How do I determine significance here? Through linear regression
+again, or do I look at non-zero estimation of the parameter values?}
+
+\subsubsection{Empirical approach with more complicated population model
+(use model similar to de Valpine
+work?)}\label{empirical-approach-with-more-complicated-population-model-use-model-similar-to-de-valpine-work}
 
@@
 variables and explantory variables that are non-linear and perhaps more
-complicated. The downsite of GAMs\ldots{}
+complicated. The downside of GAMs is they typically require more data
+and are also prone to overfitting (citations).
 

### 2017-06-14 scripts/LPI_data_analysis.R

@@ -1,47 +1,0 @@
-# Created by Easton White
-# Created on 17-Jan-2017
-# Last edited on 18-Jan-2017
-
-# This code reads in data from the () and cleans it up (mostly making it into a long table format)
-
-library("tidyr")
-library("dplyr")
-
-LPI = read.table('data/LPI_LPR2016data_public_EW_edited.csv',sep=',',header=T)
-
-
-LPI_long <- LPI %>%
-  gather(year, pop_value, starts_with('X'))
-
-LPI_long$year=as.numeric(sub('X','',LPI_long$year))
-
-#head(LPI_long[sort(LPI_long$ID),],50)
-
-newdata=arrange(LPI_long,ID,year)
-
-testing=newdata[,c(1,2,3,10,11,19,20)]
-testing[testing=='NULL']=NA
-#removed=na.omit(testing)
-
-long_term_series_names=names(sort(table(removed$ID),decreasing=T)[sort(table(removed$ID),decreasing=T)>35])
-
-y=subset(removed,removed$ID %in% long_term_series_names)
-
-
-for (species_code in long_term_series_names[1]){
-  pop = subset(removed,removed$ID==species_code)
-  
-  
-}
-
-
-
-# Plots to make
-load('example_min_time_LPI_data.Rdata')
-par(mfrow=c(1,1))
-plot(jitter(pop_info$Latitude,amount=2),pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),xlim=c(0,80),las=1,ylab='',xlab='',cex.axis=1.2)
-mtext(text = 'minimum time required',side = 2,line = 3,cex = 1.4)
-mtext(text = 'latitude',side = 1,line = 3,cex = 1.4)
-
-plot(jitter(pop_info$Longitude,amount=2),pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5))
-

### 2017-06-14 scripts/create-pop-info-table.R

@@ -1,66 +1,0 @@
-# Created by Easton White
-# Created on 17-Jan-2017
-# Last edited on 18-Jan-2017
-
-
-#This code loads in data and then creates a table called pop_info which includes population information on each unique time series
-
-# chuck for loading in data, removing NAs, looking at only long time series
-# also calculates a pop_info table which includes basic information on each population that I later analyze
-
-# Pull in data
-load('data/For_easton.Rdata')
-
-dat=dat[,c(1:7,11)] #
-#dat=na.omit(dat) #this doesn't work well
-
-#this subsets data to find long time series (e.g. >40 years)
-#dat=subset(dat,dat$class != 'Aves')
-long_time_names=names(table(dat$ID))[as.numeric(table(dat$ID))>35]
-long_dat = subset(dat,dat$ID %in% long_time_names)
-length(table(long_dat$ID) )
-
-
-#this creates pop_info table for each analyzed pop time series
-pop_info= long_dat[1,]
-  for (j in 1:length(table(long_dat$ID))){
-    index=as.numeric(which(long_dat$ID==long_time_names[j])[1])
-    pop_info[j,] = long_dat[index,]
-  }
-
-#add variables to pop_info table
-pop_info$series_length = as.numeric(table(long_dat$ID))
-pop_info$variance=NA
-pop_info$overall_trend=NA
-pop_info$autocorrelation= NA
-pop_info$trend_p_value=NA
-pop_info$min_time_for_power=NA
-pop_info$pop_value=as.numeric(pop$pop_value)
-long_dat$pop_value=long_dat$pop_value
-  for (j in 1:nrow(pop_info)){
-    pop= subset(long_dat,long_dat$ID==pop_info$ID[j])
-    pop$pop_value=as.numeric(pop$pop_value)
-    pop$pop_value=pop$pop_value/max(pop$pop_value) #should I standaridize this in a better way
-    pop_info$variance[j]= var(pop$pop_value)
-    pop_info$overall_trend[j] = calculate_slope(pop$pop_value)
-    pop_info$autocorrelation[j] = acf(pop$pop_value,plot = FALSE)$acf[2]
-    pop_info$trend_p_value[j]=calculate_p_value(pop$pop_value)
-    
-    source('scripts/calculate_power_metric.R')
-    pop_info$min_time_for_power[j] = min_time_needed(pop$pop_value,0.05,0.8)
[…]

### 2017-06-14 scripts/remove_NA_populations.R

@@ -1,29 +1,0 @@
-
-
-
-# Remove records on interior of time series for each individual population
-load('data/For_easton.Rdata')
-
-dat=dat[,c(1:7,11)] #
-
-dat=testing
-ID_codes = names(table(dat$ID))
-good_rows = NULL
-
-#dat=long_dat
-
-for (j in 2650:length(ID_codes)){
-  #pop = subset(dat,dat$ID==ID_codes[j])[1:35,]
-  pop = subset(dat,dat$ID==ID_codes[j])
-  pop=pop[head(which(is.na(as.numeric(pop$pop_value))==F),1):tail(which(is.na(as.numeric(pop$pop_value))==F),1),]
-  if (nrow(pop) == nrow(na.omit(pop)) &nrow(na.omit(pop))>35){
-    good_rows=cbind(good_rows,row.names(pop))
-  }
-  print(j)
-}
-
-dat = dat[row.names(dat) %in% c(good_rows),]
-
-
-c(bad_IDs); length(bad_IDs)
-#Removes over 500 populations!!

### 2017-06-28 writing/ER_White_time_series_power_cover_letter.tex
added to discussion and created cover letter

\documentclass[]{article}

\begin{document}
\noindent 28-June-2017

\medskip
\medskip
\medskip
\noindent Dear editorial board of \emph{Biological Conservation}:

\medskip
\medskip

%"The cover letter should explain how the manuscript fits the scope of the journal, and more specifically how it advances the field, while having broad appeal. If the manuscript relates to any previous submission to an ESA journal, that must be explained as well.  Longer submissions those between 30 and 50 manuscript pages) should be accompanied by a detailed justification for the length. There is a required text box for the cover letter.  Uploading a cover letter as an attachment is optional"

\indent \indent Please find enclosed the manuscript: ``Your time series is (probably) too short", by Easton R. White, to be submitted as a \emph{Short Communication} article to \emph{Biological Conservation} for consideration of publication. There is no financial interest to report. I certify that the submission is my own original work and is not under review at any other publication. A version of the submission is available as a preprint on Peerj at

Longterm census monitoring has become a cornerstone of modern ecological and conservation science. However, little work has examined the question, what is the minimum time series length required to adequately address a particular question of interest? Unfortunately, classic power analyses fail to address needs of questions related to time series analyses. I examine this question building on a few papers, in large part due to work in \emph{Biological Conservation}. In this manuscript, I use both simulation and empirical approaches to determine the minimum time required to address questions related to longterm population increases or decreases. I use an database of 878 population time series to determine that at least 10-15 years of continuous modeling are required to determine changes in abundance over time. Both simulation and empirical approaches noted the many considerations for designing a monitoring program.

I believe that my findings are of general interest to the readers of \emph{Biological Conservation} for its applications to both basic ecological research and management. This manuscript provides detailed examples of how to perform power analyses for time series using both simulation and empirical approaches. All of the R code and data used in this manuscript are freely available on Github. My work highlights the importance of longterm monitoring programs for a variety of species. The 878 populations in the manuscript represent a wide-array of life history traits and locations. This manuscript provides practical advice for designing longterm monitoring programs. 
\medskip

\noindent For potential reviewers, I would recommend:

\noindent 	- Michelle H. Reynolds, US Geological Survey, mreynolds@usgs.gov, 808-985-6416 \\
	- Jonathan R. Rhodes, University of Queensland, j.rhodes@uq.edu.au, 61 7 336 56838 \\
	- Paul C.D. Johnson, University of Glasgow, Paul.Johnson@glasgow.ac.uk, 01413306638 \\
	- Kurt E. Anderson, University of California, Riverside, kurt.anderson@ucr.edu \\

\medskip
\noindent I look forward to hearing from you at your earliest convenience.
\medskip
\medskip

Sincerely yours,

\medskip
\medskip
\medskip
\medskip

Easton White \\ 
Center for Population Biology, University of California, Davis, CA, USA, eawhite@ucdavis.edu, +1 480-203-7931

\end{document}

### 2017-06-28 writing/manuscript_draft.Rmd
added to discussion and created cover letter

@@
 
-Number of words: X,XXX 
+Number of words: X,XXX (less than 4,000 words or 20 pages)
 \vspace{3 mm}
 
-To be submitted to: \emph{XXXXX} (Article)
+To be submitted to: \emph{Biological Conservation} (Short Communication)
 \vspace{3 mm}
 
-Keywords:
+Available as a preprint at: www.peerj.com
+\vspace{3 mm}
+
+Keywords: ecological time series, experimental design, monitoring, power analysis, statistical power, sampling design
 
@@
 
+\pagebreak
+
 \linenumbers
 
-## Abstract
+# Abstract
 Long-term census data is necessary to better understand ecological processes and make management decisions.
@@
 Little work has addressed the question, when is a time series of census data long enough to address a question of interest?
-Here, I explore two approaches to address this question: one simulation-based and an empirical approach.
+Here, we explore two approaches to address this question: through simulations and through an empirical approach.
 I specifically determine the minimum time series length required to estimate significant increases or decreases in population abundance.
-<!--Importantly, I examine the ability to detect a trend with a set level of statistical power, which is often neglected in ecological time series analyses.--> Using a simulation approach, I demonstrate how the minimum time series length required increases with weaker trends in abundance and with higher variability in population size.
-In addition, I examine 868 populations of vertebrate species to determine the minimum time required to detect changes in their abundance.
-I show that 10-15 years of continuous monitoring are required in order to achieve a high level of statistical power.
+Importantly, I examine the ability to detect a trend with a set level of statistical power, which is often neglected in ecological time series analyses.
+Using a simulation approach, we demonstrate how the minimum time series length required increases with weaker trends in abundance and with higher population variability.
+In addition, we examine 868 populations of vertebrate species to determine the minimum time required to detect changes in their abundance.
+We show that 10-15 years of continuous monitoring are required in order to achieve a high level of statistical power.
 Similar to simulation-approaches, the minimum time required for field census data strongly depends on trend strength, population variability, and temporal autocorrelation.
 <!--Perhaps surprisingly, the minimum time required did not correlate well with biological explanatory variables, like body size or generation length.--> These results point to the importance of sampling populations over long periods of time.
-<!--Here, I stress the need for verifying a study has enough statistical power to accurately determine long-term changes in population abundance.--> Most studies, especially those less than 10-15 years, are probably underpowered and potentially misleading.
+<!--Here, We stress the need for verifying a study has enough statistical power to accurately determine long-term changes in population abundance.--> Most studies, especially those less than 10-15 years, are probably underpowered and potentially misleading.
 
-## Introduction - need to clean up and reduce text
+# Introduction
 
@@
 In addition, short funding cycles make it difficult to examine populations over longer periods of time (citations).
-However, there are of course notable exceptions including the Isle Royale study, Iowa corn study, cod, lynx-hair, and many others.
+There are, of course, notable exceptions including the Isle Royale study, Iowa corn study, cod, lynx-hair, and many others (citations).
 
 Naturally, one might ask, how long of a time series is actually necessary?
-This question has important scientific and management implications.
-Scientifically, we need to know how long a time series is required to address a specific question.
+This question has important implications for both research and management.
+Scientifically, we need to know the time series length required to address a specific question.
 Too short of a time series may lead to wrong conclusions given large natural year-to-year variability (citations).
 From a management perspective, it is important to understand when a trend in abundance over time is actually meaningful or not.
-For example, the IUCN Red List Categories and Criteria suggest under Criterian A2, a species qualifies as vulnerable if it has experienced a 30% decline over 10 years, or 3 generations [@IUCN2012].
+The IUCN Red List Categories and Criteria suggest under Criterian A2, a species qualifies as vulnerable if it has experienced a 30% decline over 10 years, or 3 generations [@IUCN2012].
 For both scientific and management questions, because sampling is typically expensive, we do not want to sample for longer than is neccessary [@Gerber1999].
 
-In determining the number of samples required for a particular experiment four quantities are intricately linked: statistical power, effect size, alpha, sample size [@Legg2006].
+In determining the number of samples required for a particular experiment four quantities are intricately linked: statistical power (1-$\beta$), effect size, $\alpha$, and sample size [@Legg2006].
 The exact relationship between these quantities depends on the specific statistical test.
@@
 Prior to an experiment, one could set appropriate levels of power, significance threshold ($\alpha$), and the effect size to estimate the sample size required for the experiment.
-This approach, however, is not straight-forward for a time series, as data points are clearly non-independent.
+This approach, however, is not straight-forward for a time series, or more complicated scenarios [@Johnson2015], as data points are clearly non-independent.
 
-<!--For time series data, two general approaches can be used.
+For time series data, two general approaches can be used.
 First, as can be done with experimental data, we can relax the definition of statistical power (cite Bolker).
@@
 Subsamples of different lengths can then be evaluated to see which fraction of subsamples meet some criteria, again a measure of statistical power.
-Similar to the simulation approach, we can use these measures of power to determine the minimum time series required for a particular question of interest.-->
+Similar to the simulation approach, we can use these measures of power to determine the minimum time series required for a particular question of interest.
 
 Past work has investigated questions related to the minimum time series required to estimate trends in population size over time [@Giron-Nava2017].
-@Rhodes2011 examined the optimal allocation of effort between spatial and temporal replicates.
+For example, @Rhodes2011 examined the optimal allocation of effort between spatial and temporal replicates.
 Using simple populations models, they found that the allocation of effort depends on environmental variation, spatial and temporal autocorrelation, and observer error.
@@
 They found that for this short-lived organims, that 15 years was adequate to assess longterm trends in abundance.
-However, these studies, and other past work (citation), have focused either on theoretical aspects of monitoring design or focused on single species.
+However, these studies, and other past work, have focused either on theoretical aspects of monitoring design or focused on single species.
 
 In this paper, we use both simulations and a database of empirical time series to determine the minimum number of years required to address various questions.
-In the main manuscript, we determine the minimum time series length to determine if a population is incresing or decreasing according to a significant slope coefficient in simple linear regression.
+More specifically, we estimate the minimum time series length required to asssess longterm changes in abundance via simple linear regression.
 <!-- In this paper, we evaluate the minimal time series length required to answer X questions.
@@
 
-## Methods
+# Methods
 
-### Theoretical approach
+## Simulation approach
 
@@
 
-In Figure \ref{fig:theoretical_approach}a, a number of simulated time series are shown for a set number of time periods ($t=40$).
+In Fig.
+\ref{fig:theoretical_approach}a, a number of simulated time series are shown for a set number of time periods ($t=40$).
 It is clear that statistical power increases quickly with increases in length of time sampled (Fig.
@@
 The horizontal, dashed line is our desired statistical power of 0.8.
-The vertical, dashed line is the minimum time required to achieve our desired statistical power.\\label{fig:theoretical_approach}'}
+The vertical, dashed line is the minimum time required to achieve our desired statistical power.
+(c) Minimum time required for different values of the trend strength ($r$) and (d) minimum time required for different levels of population variability ($\\sigma$).\\label{fig:theoretical_approach}'}
 setwd("~/Desktop/Research/time-series-project")
@@
   years_to_sample=seq(2,100,by=1)
-  power=lapply(X=years_to_sample,FUN = calculate_power,phi = 0.6,r=0.2,sigma = .5,trials=50)#should use 300 trials
+  power=lapply(X=years_to_sample,FUN = calculate_power,phi = 0.6,r=0.2,sigma = .5,trials=50)#should use 300+ trials
 
@@
       mtext(text = 'population size' ,side = 2,cex = 1.2,line=4)
+      mtext(text = '(a)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
 
@@
       mtext(text = 'statistical power' ,side = 2,cex = 1.2,line=3.5)
+      mtext(text = '(b)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
 
     # Plot versus trend strength
-      r_values = seq(1,3,by=0.2)
-      sigma_values=seq(1,7,by=0.2)
+      r_values = seq(1,3,by=1) #by=0.2
+      sigma_values=seq(1,7,by=1) #by=0.2
 
@@
 
-      plot(parameter_combinations$r,parameter_combinations$min_time_required,type='l',lwd=2,las=1,ylab='',xlab='',cex.axis=1.2)
+      plot(parameter_combinations$r,parameter_combinations$min_time_required,type='l',lwd=2,las=1,ylab='',xlab='',cex.axis=1.2,ylim=c(0,50))
       mtext(text = 'trend strength (slope coefficient)' ,side = 1,cex = 1.2,line=3)
       mtext(text = 'minimum time required' ,side = 2,cex = 1.2,line=3.5)
+      mtext(text = '(c)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
 
@@
 
-      plot(parameter_combinations$sigma,parameter_combinations$min_time_required,type='l',lwd=2,las=1,ylab='',xlab='',cex.axis=1.2)
+      plot(parameter_combinations$sigma,parameter_combinations$min_time_required,type='l',lwd=2,las=1,ylab='',xlab='',cex.axis=1.2,ylim=c(0,50))
       mtext(text = 'population variability' ,side = 1,cex = 1.2,line=3)
       mtext(text = 'minimum time required' ,side = 2,cex = 1.2,line=3.5)
+      mtext(text = '(d)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
 ```
 
+<!-- code to pull in databases-->
 ```{r, echo=F}
@@
 #Load our cleaned data (see other code for cleaning steps)
-  load('cleaned-data/pop_info_example_goodNAs.Rdata') #Do i need to filter out non-significant populations as well (where no trend was present)
-  pop_info=pop_info[pop_info$trend_p_value<0.05,]
-  long_dat=subset(long_dat,long_dat$ID %in% pop_info$ID)
-  long_time_names=subset(long_time_names,long_time_names %in% pop_info$ID)
+  load('cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata') #Do i need to filter out non-significant populations as well (where no trend was present)
+ # pop_info=pop_info[pop_info$trend_p_value<0.05,]
+ # long_dat=subset(long_dat,long_dat$ID %in% pop_info$ID)
+ # long_time_names=subset(long_time_names,long_time_names %in% pop_info$ID)
 ```
 
-### Data source
+## Data source
 We use a database of `r length(table(dat$ID))` population time series compiled in [@Keith2015].
 The data is originally from the Global Population Dynamics Database [@GPDD2010].
-We filtered out short time series (less than `r pop_info$series_length[1]`), and those with missing data, and were left with `r nrow(pop_info)` time series.
+We filtered out short time series (less than `r pop_info$series_length[1]` years), and those with missing data, and were left with `r nrow(pop_info)` time series.
 The database includes information on `r length(table(pop_info$Binomial))` vertebrate species with a focus on mammals, birds, and fish.
@@
 # Pull in results of biological characteristics comparison to min time series length required
+setwd("~/Desktop/Research/time-series-project")
+load('cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
 ```
-For a subset of populations in the database (X = XX), we were able to connect a database on biological characteristics like body size and generation length. We used
+For a subset of populations in the database ($n =$ `r nrow(LPI_pop_info)`), we were able to connect a database on biological characteristics (**cite database here**) like body size and generation length.
+All `r nrow(LPI_pop_info)` populations are within the Aves class.
+In the supplementary material, we explore how the minimum time required does not strongly correlate with any biological characteristics.
 
-### Empirical approach
+## Empirical approach
 
@@
 variability) about the population.
-In other words, they are a representative sample.
+In other words, each time series is a representative sample.
 We first take all possible contiguous subsamples of each time series.
@@
 
-```{r, echo=F,message=F,warning=F,fig.cap='(a) Population size of ANIMAL NAME over time.
-The red line is the best fit line from linear regresssion.
-(b) statistical power for different subsets of the time series in panel a.\\label{fig:empirical_approach_example}'}
+<!-- example of how subsampling and empirical approach works -->
+```{r, echo=F,message=F,warning=F,fig.cap='(a) Population size of Harp Seals over time.
+The line is the best fit line from linear regresssion.
+(b) Statistical power for different subsets of the time series in panel a.\\label{fig:empirical_approach_example}'}
 # Example of subsampling empirical data
  setwd("~/Desktop/Research/time-series-project")
- load("cleaned-data/pop_info_example_goodNAs.Rdata")
-  par(mfrow=c(1,2))
+ load("cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
+  par(mfrow=c(1,2),oma=c(4,4,0.5,0.5))
     pop = subset(long_dat,long_dat$ID=='13192') #13192 for Harp seals
-    plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='population size',xlab='time (years)',pch=16,las=1)
+    plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='',xlab='time (years)',pch=16,las=1)
     abline(lm(pop$popvalue~pop$year),lwd=2,col='red')
-    mtext(text = '(a)',side = 3,line = -1.2,at = 1965,font = 2,cex=1.2)
+    mtext(text = '(a)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
+    mtext(text='population size', side=2,line=4,cex=1.2)
 # could insert all possible regressions of length n
@@
     plot(power_example,cex.axis=1.2,cex.lab=1.2,ylab='statistical power',xlab='years sampled',las=1,pch=16)
-    mtext(text = '(b)',side = 3,line = -1.2,at = 7,font = 2,cex=1.2)
+    mtext(text = '(b)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
 # could also insert standard error bars
@@
 
-## Results
+# Results
 <!--
@@
 
-### Simulation approach
+## Simulation approach
 
@@
 
-### Empirical test
+## Empirical test
 
@@
  setwd("~/Desktop/Research/time-series-project")
- load("cleaned-data/pop_info_example_goodNAs.Rdata")
+ load("cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
 
@@
 
-```{r,echo=F,fig.cap='Minimum time required to estimate change in abundance correlated with (a) trend strength (absolute value of slope coefficient estimated from linear regression), (b) population variance (interannual variability in population size), and (c) temporal autocorrelation.\\label{fig:correlates}'}
+```{r,echo=F, fig.cap='Minimum time required to estimate change in abundance correlated with (a) trend strength (absolute value of slope coefficient estimated from linear regression), (b) population variance (interannual variability in population size), and (c) temporal autocorrelation.\\label{fig:correlates}'}
 
@@
   mtext('| trend strength |',1,line=3,cex=1.2)
-  mtext('(a)',3,line=-1.5,at=0.0318,cex=1.2,font = 2)
+  mtext('(a)',3,line=-1.5,adj = 0.985,cex=1.2,font = 2)
     plot(pop_info$variance,pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
     mtext('population variance',1,line=3,cex=1.3)
-    mtext('(b)',3,line=-1.5,at=0.128,cex=1.2,font = 2)
+    mtext('(b)',3,line=-1.5,adj = 0.985,cex=1.2,font = 2)
       plot(pop_info$autocorrelation,pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
       mtext('autocorrelation',1,line=3,cex=1.2)
-       mtext('(c)',3,line=-1.5,at=0.97,cex=1.2,font = 2)
+       mtext('(c)',3,line=-1.5,adj = 0.985,cex=1.2,font = 2)
        mtext('minimum time required',2,line=2.5,cex=1.2,outer = T)
@@
 SX).
-Combined trend strength, population variance, and autocorrelation account for about `r round(100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance),1)` of the explained deviance, which is similar to $R^2$ [@Zuur2009], in minimum time series length required.
+Combined, trend strength, population variance, and autocorrelation account for about `r round(100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance),1)` of the explained deviance, which is similar to $R^2$ [@Zuur2009], in minimum time series length required.
 
@@
 body mass, longevity, location).
-None (need to do regression) of these explanatory variables were predictive of the minimum time series length required (supp mat).
-Some differences if $T_{min}$ did occur between biological class (supp mat), but this can be attributed to the amount populations are known to fluctuate (i.e.
+None of these explanatory variables were predictive of the minimum time series length required (**supp mat**).
+Some differences if $T_{min}$ did occur between biological class (**supp mat**), but this can be attributed to the amount populations are known to fluctuate (i.e.
 the variability in population size).
-
-```{r,echo=F,include=FALSE, eval=FALSE,fig.cap='(a) Minimum time required to estimate change in abundance for species class, (b) Longterm trend (estimated slope coefficient) by species class, (c) interannual variability in population size by species class, and (d) temporal autocorrelation by species class.\\label{fig:biological_correlates}'}
-# Biological correlates plots
-par(mfrow=c(2,2),mar=c(4,5,0.5,0.5))
-
-boxplot(pop_info$min_time_for_power~pop_info$class, lwd = 1, ylab = 'minimum time required',col='white',cex.lab=1.4,cex.axis=1.2,las=1)
-stripchart(pop_info$min_time_for_power~pop_info$class, vertical = TRUE,method = "jitter", add = TRUE, pch = 20, col = rgb(0.5,0.5,0.5,0.5))
-
-boxplot(abs(pop_info$overall_trend)~pop_info$class, lwd = 1, ylab = ' ',col='white',cex.lab=1.4,cex.axis=1.2,las=1)
-stripchart(abs(pop_info$overall_trend)~pop_info$class, vertical = TRUE,method = "jitter", add = TRUE, pch = 20, col = rgb(0.5,0.5,0.5,0.5))
-  mtext('| trend strength |',2,line=3.5,cex=1.2)
-
-boxplot(pop_info$variance~pop_info$class, lwd = 1, ylab = 'population variance',col='white',cex.lab=1.4,cex.axis=1.2,las=1)
-stripchart(pop_info$variance~pop_info$class, vertical = TRUE,method = "jitter", add = TRUE, pch = 20, col = rgb(0.5,0.5,0.5,0.5))
-
-boxplot(pop_info$autocorrelation~pop_info$class, lwd = 1, ylab = 'autocorrelation',col='white',cex.lab=1.4,cex.axis=1.2,las=1)
-stripchart(pop_info$autocorrelation~pop_info$class, vertical = TRUE,method = "jitter", add = TRUE, pch = 20, col = rgb(0.5,0.5,0.5,0.5))
-```
 
@@
 <!--
-## Potential figures
+# Potential figures
 - Examples of how determining minimum time series works (probably in methods)
@@
 
-### Questions in the appendix
+## Questions in the appendix
 
@@
 
-## Discussion
+# Discussion
 
@@
 
-We show two different approaches to estimate the minimum time series length required required to address a particular question of interest.
-The simulation-based approach has been suggested by others, especially in situations where analysis is more complicated than a simple power analysis [@Gerrodette1987].
-Our simulations support past work...
+We explored two different approaches to estimate the minimum time series length required to address a particular question of interest.
+We specifically asked, what is the minimum time series length required to determine longterm changes in abundance using linear regression?
+The simulation-based approach has been suggested by others, especially in situations more complicated than that suited for classic power analysis [@Gerrodette1987, Johnson2015].
+Our simulations support past work that longer time series are needed when the trend strength (i.e.
+rate of increase or decrease) is weak or when population variability is high [@Gerrodette1987].
 
-Here, we focus on a empirical approach to estimate the minimum time series length requried to address a given question.
+Here, we focus on a empirical approach to estimate the minimum time series length requried to assess changes in abundance over time.
+We examined long time series and then subsampled each to determine the minimum time required to achieve a desired level of significance and power for linear regression.
+Statistical power is important as it provides on information as to the neccessary samples required to dertermine a significant trend [@Legg2006].
+We found that for `r nrow(pop_info)` populations, at least 10-15 years of continuous monitoring were neccessary (Fig.
+\ref{fig:min_time_dist}).
+This timeframe is in line with past work on a short-lived snail species [@Rueda-Cediel2015] and a long-lived whale species [@Gerber1999].
+In line with theoretical predictions (citations), we also found $T_{min}$ was strongly correlated with the trend strength, variability in population size, and temporal autocorrelation (Fig.
+\ref{fig:correlates}).
+Against our prior expectations, we also found that $T_{min}$ did not correlate with any biological variables of interest (cite figure or supp mat).
+We initially hypothesized that species with longer lifespans or generation times may require a longer sampling period.
+Our result could have result of at least two effects.
+First, the database we used may not include a diverse enough set of species with different life history traits.
+Second, the question we poised, whether a population is increasing or decreases, is only worried about population surveys.
+Therefore, biological variables may be more important if the question itself was different.
+ANOTHER SENTENCE HERE
 
-First we asked, what is $T_{min}$ for determining whether a population was increasing or decreasing over time according to linear regression?
-We found that for `r nrow(pop_info)` populations, at least 10-15 years of continuous monitoring were neccessary (Fig.
-\ref{fig:fig:min_time_dist}).
-In line with theoretical predictions (citations), we also found $T_{min}$ was strongly correlated with the trend strength, variability in population size, and temporal autocorrelation (cite figure).
-
-We also found that $T_{min}$ did not correlate with any biological variables of interest (cite figure or supp mat).
-We initially hypothesized that species with longer lifespans or generation times may require more years to sample.
-(Our result could have been the effect of two reasons) First, the database we used may not include a diverse enough set of species with different life history traits.
-Second, the question we poised, whether a population is increasing or decreases, is only worried about population surveys.
-Therefore, biological variables may be more important if the question itself was different...
-
-### Other questions addressed
-
-In the appendix we asked two additional questions to make estimates of $T_{min}$. First, we asked how many years of sampling are neccessary to estimate a geometric growth rate?
-
-### Past work on this issue - can I integrate this better into the rest of the section. How much should go in introduction?
-There have been previous attempts to address how long a time series should be to assess trends in abundance. For example,...
-
-@Rueda-Cediel2015 found that 10-15 years was adequate to characterize population trends in a short lived animal. They used a matrix model of a particular species...
-
-@Gerber1999 investigated the minimum time series required to estimate population growth of an endagered, but recovering, species.
-They used a longterm census to retroactively determine the minimum time series required to assess threat status.
-They found that only XX years were needed, XX years before the delisting decision was actually made.
-This highlights the importance of estimating the minimunm time series required as an earlier decision would have saved time and money [@Gerber1999].
-
-There has also been interest in determining the minimum time series length in the bird community (citations).
-(Bart 2004, Hatch 2003) [@Seavy2007] Hatch 2003 used seabird monitoring data to estimate the minimum time needed to achieve appropriate levels of power.
-He found that the minimum time required ranged between 11 and 69 years depending on species, trend strength, and study design (cite Hatch).
-Similarly, Bart et al 2004 recommended a standard of 80% power to detect a 50% decline over 20 years in landbird populations.
+There has also been interest in determining the minimum time series length in the bird community [@Hatch2003, @Bart2004, @Seavy2007].
+@Hatch2003 used seabird monitoring data to estimate the minimum time needed to achieve appropriate levels of power.
+He found that the minimum time required ranged between 11 and 69 years depending on species, trend strength, and study design.
+Similarly, @Bart2004 recommended a standard of 80% power to detect a 50% decline over 20 years in landbird populations.
 They found that only about 42% of populations in the Breeding Bird Survey met their criteria.
@@
 They used 24 years of census data on Red-tailed Tropicbirds (\emph{Phaethon rubricauda}) in Hawaii and showed that to detect a 50% decline over 10 years almost always resulting in high statistical power (above 0.8).
-Therefore, they cautionsioned against only using power analyses to design monitoring schemes and instead argued for precision based metrics ..
+Therefore, they cautioned against only using power analyses to design monitoring schemes and instead argued for precision based metrics ..
 
@@
 
-### Limitations
+## Other questions addressed
 
-Our work has obvious limitations in determining the minimum time series length required.
-First, $T_{min}$ depends strongly on the question being asked and the availability of long-term census data.
+In the appendix we asked two additional questions to make estimates of $T_{min}$. First, we asked how many years of sampling are neccessary to estimate a geometric growth rate?
+
+<!--
+@Rueda-Cediel2015 found that 10-15 years was adequate to characterize population trends in a short lived animal. They used a matrix model of a particular species...
+
+@Gerber1999 investigated the minimum time series required to estimate population growth of an endagered, but recovering, species.
+They used a longterm census to retroactively determine the minimum time series required to assess threat status.
+They found that only XX years were needed, XX years before the delisting decision was actually made.
+This highlights the importance of estimating the minimunm time series required as an earlier decision would have saved time and money [@Gerber1999].
+-->
+
+## Limitations
+
+Our work has some limitations in determining the minimum time series length required.
+First, $T_{min}$ depends strongly on the question of interest.
 For our empirical approach, the subsampling of the full time series allows for estimates of power, but the individual subsamples are clearly not independent of one another.
-(sentence about alpha and power limititations here) The perfect scenario would be to have a population model built and parameterized for each population of interest.
+In addition, estimates of $T_{min}$ depend on chosen values of $\alpha$ and $\beta$ (see appendix).
+<!-- include sentence on how T_min depends on these parameters --> In an ideal world, we would build a specific population model parameterized for each population of interest.
 Then, model simulations could be used to estimate the minimum time series reuqired to address a specific question of interest.
@@
 
-### Conclusions
-
-- what we found, implications, future work
+## Conclusions
 
 We use a database of `r nrow(pop_info)` populations to determine the minimum time series length required.
-We show that to assess longterm trends in abundance, 10-15 years of continuos monitoring are required (Fig.
-\ref{fig:min_time_dist}.
+We show that to identify longterm changes in abundance, 10-15 years of continuos monitoring are required (Fig.
+\ref{fig:min_time_dist}).
 In line with theoretical predictions, we also show that $T_{min}$ is strongly correlated with the overall trend in abundance, variability in abundance, and the temporal autocorrelation (Fig.
@@
 This stresses the importance of longterm monitoring programs.
-(where do IUCN recommendations come from?) From both a scientific and management perspective estimates of $T_{min}$ are important.
+<!--(where do IUCN recommendations come from?)--> From both a scientific and management perspective estimates of $T_{min}$ are important.
 If a time series is too short, we are likely to make incorrect conclusions about longterm trends.
 In addition, a time series that is too long may be a poor use of already limited funds [@Gerber1999].
-Future work can examine other species, with a wider range of life history characteristics, or address other questions besides longterm trends in abundance.
+Future work can examine other species, with a wider range of life history characteristics.
+In addition, similar approaches can be used to determine the minimum time series length required to address other questions.
 
 <!--
-### Time series length recommendations from other papers
+## Time series length recommendations from other papers
 
@@
 
-## Supporting Information
+# Supporting Information
 
@@
 
-## Acknowledgements
+# Acknowledgements
 
@@
 
-## References
+# References
+
+<div id="refs"></div>
+
+# Figures
+
+stuff here
+
+# Supplementary material

### 2017-06-28 writing/manuscript_draft.tex
added to discussion and created cover letter

@@
 
-Number of words: X,XXX \vspace{3 mm}
+Number of words: X,XXX (less than 4,000 words or 20 pages) \vspace{3 mm}
 
-To be submitted to: \emph{XXXXX} (Article) \vspace{3 mm}
+To be submitted to: \emph{Biological Conservation} (Short Communication)
+\vspace{3 mm}
 
-Keywords:
+Available as a preprint at: www.peerj.com \vspace{3 mm}
+
+Keywords: ecological time series, experimental design, monitoring, power
+analysis, statistical power, sampling design
 
@@
 
+\pagebreak
+
 \linenumbers
 
-\subsection{Abstract}\label{abstract}
+\section{Abstract}\label{abstract}
 
@@
 has addressed the question, when is a time series of census data long
-enough to address a question of interest? Here, I explore two approaches
-to address this question: one simulation-based and an empirical
-approach. I specifically determine the minimum time series length
-required to estimate significant increases or decreases in population
-abundance. Using a simulation approach, I demonstrate how the minimum
-time series length required increases with weaker trends in abundance
-and with higher variability in population size. In addition, I examine
-868 populations of vertebrate species to determine the minimum time
-required to detect changes in their abundance. I show that 10-15 years
-of continuous monitoring are required in order to achieve a high level
-of statistical power. Similar to simulation-approaches, the minimum time
-required for field census data strongly depends on trend strength,
-population variability, and temporal autocorrelation. These results
-point to the importance of sampling populations over long periods of
-time. Most studies, especially those less than 10-15 years, are probably
-underpowered and potentially misleading.
+enough to address a question of interest? Here, we explore two
+approaches to address this question: through simulations and through an
+empirical approach. I specifically determine the minimum time series
+length required to estimate significant increases or decreases in
+population abundance. Importantly, I examine the ability to detect a
+trend with a set level of statistical power, which is often neglected in
+ecological time series analyses. Using a simulation approach, we
+demonstrate how the minimum time series length required increases with
+weaker trends in abundance and with higher population variability. In
+addition, we examine 868 populations of vertebrate species to determine
+the minimum time required to detect changes in their abundance. We show
+that 10-15 years of continuous monitoring are required in order to
+achieve a high level of statistical power. Similar to
+simulation-approaches, the minimum time required for field census data
+strongly depends on trend strength, population variability, and temporal
+autocorrelation. These results point to the importance of sampling
+populations over long periods of time. Most studies, especially those
+less than 10-15 years, are probably underpowered and potentially
+misleading.
 
-\subsection{Introduction - need to clean up and reduce
-text}\label{introduction---need-to-clean-up-and-reduce-text}
+\section{Introduction}\label{introduction}
 
@@
 difficult to examine populations over longer periods of time
-(citations). However, there are of course notable exceptions including
-the Isle Royale study, Iowa corn study, cod, lynx-hair, and many others.
+(citations). There are, of course, notable exceptions including the Isle
+Royale study, Iowa corn study, cod, lynx-hair, and many others
+(citations).
 
 Naturally, one might ask, how long of a time series is actually
-necessary? This question has important scientific and management
-implications. Scientifically, we need to know how long a time series is
+necessary? This question has important implications for both research
+and management. Scientifically, we need to know the time series length
 required to address a specific question. Too short of a time series may
@@
 understand when a trend in abundance over time is actually meaningful or
-not. For example, the IUCN Red List Categories and Criteria suggest
-under Criterian A2, a species qualifies as vulnerable if it has
-experienced a 30\% decline over 10 years, or 3 generations (IUCN 2012).
-For both scientific and management questions, because sampling is
-typically expensive, we do not want to sample for longer than is
-neccessary (Gerber, DeMaster, and Kareiva 1999).
+not. The IUCN Red List Categories and Criteria suggest under Criterian
+A2, a species qualifies as vulnerable if it has experienced a 30\%
+decline over 10 years, or 3 generations (IUCN 2012). For both scientific
+and management questions, because sampling is typically expensive, we do
+not want to sample for longer than is neccessary (Gerber, DeMaster, and
+Kareiva 1999).
 
 In determining the number of samples required for a particular
-experiment four quantities are intricately linked: statistical power,
-effect size, alpha, sample size (Legg and Nagy 2006). The exact
-relationship between these quantities depends on the specific
-statistical test. Formally, statistical power (\(1-\beta\)) is one minus
-the probability of a type II error (\(\beta\)), or false negative. In
-other words, it is a measure of how well you could detect a trend given
-it is actually significant. Prior to an experiment, one could set
-appropriate levels of power, significance threshold (\(\alpha\)), and
-the effect size to estimate the sample size required for the experiment.
-This approach, however, is not straight-forward for a time series, as
-data points are clearly non-independent.
+experiment four quantities are intricately linked: statistical power
+(1-\(\beta\)), effect size, \(\alpha\), and sample size (Legg and Nagy
+2006). The exact relationship between these quantities depends on the
+specific statistical test. Formally, statistical power (\(1-\beta\)) is
+one minus the probability of a type II error (\(\beta\)), or false
+negative. In other words, it is a measure of how well you could detect a
+trend given it is actually significant. Prior to an experiment, one
+could set appropriate levels of power, significance threshold
+(\(\alpha\)), and the effect size to estimate the sample size required
+for the experiment. This approach, however, is not straight-forward for
+a time series, or more complicated scenarios (P. C. D. Johnson et al.
+2015), as data points are clearly non-independent.
+
+For time series data, two general approaches can be used. First, as can
+be done with experimental data, we can relax the definition of
+statistical power (cite Bolker). Instead of using classic statistical
+tests (e.g.~t-test), and corresponding formulas for determinig sample
+size, simulation approaches can be used (Gerrodette 1987). Simple
+population models can be simulated with parameter values corresponding
+to a population of interest. Out of a large number of simulations, power
+can be calculated as the fraction of simulations that meet some
+criteria. The specific criteria depend on the question at hand. Then
+time series simulations of different lengths are run to determine the
+minimum time series length required to be confident that you will be
+able to answer your question (cite Bolker). In addition to using
+simulated data, empirical time series can also be used. We usually do
+not have multiple replicates of similar populations, but we can
+subsample an empirical time series (cite). Subsamples of different
+lengths can then be evaluated to see which fraction of subsamples meet
+some criteria, again a measure of statistical power. Similar to the
+simulation approach, we can use these measures of power to determine the
+minimum time series required for a particular question of interest.
 
 Past work has investigated questions related to the minimum time series
-required to estimate trends in population size over time. Rhodes and
-Jonzen (2011) examined the optimal allocation of effort between spatial
-and temporal replicates. Using simple populations models, they found
-that the allocation of effort depends on environmental variation,
-spatial and temporal autocorrelation, and observer error. Rueda-Cediel
-et al. (2015) also used a modeling approach, but parameterized it for a
-threatened snail, \emph{Tasmaphena lamproides}. They found that for this
-short-lived organims, that 15 years was adequate to assess longterm
-trends in abundance. However, these studies, and other past work
-(citation), have focused either on theoretical aspects of monitoring
-design or focused on single species.
+required to estimate trends in population size over time (Giron-Nava et
+al. 2017). Rhodes and Jonzen (2011) examined the optimal allocation of
+effort between spatial and temporal replicates. Using simple populations
+models, they found that the allocation of effort depends on
+environmental variation, spatial and temporal autocorrelation, and
+observer error. Rueda-Cediel et al. (2015) also used a modeling
+approach, but parameterized it for a threatened snail,
+\emph{Tasmaphena lamproides}. They found that for this short-lived
+organims, that 15 years was adequate to assess longterm trends in
+abundance. However, these studies, and other past work (citation), have
+focused either on theoretical aspects of monitoring design or focused on
+single species.
 
@@
 series to determine the minimum number of years required to address
-various questions. In the main manuscript, we determine the minimum time
-series length to determine if a population is incresing or decreasing
-according to a significant slope coefficient in simple linear
-regression.
+various questions. Here, we determine the minimum time series length to
+determine if a population is incresing or decreasing according to a
+significant slope coefficient in simple linear regression.
 
-\subsection{Methods}\label{methods}
+\section{Methods}\label{methods}
 
-\subsubsection{Theoretical approach}\label{theoretical-approach}
+\subsection{Simulation approach}\label{simulation-approach}
 
@@
 This is the same approach one might use in sample size calculations for
-any experimental design too complicated for simple power analyses
-(Johnson et al. 2015) (and cite Bolker). We only briefly discuss this
+any experimental design too complicated for simple power analyses (P. C.
+D. Johnson et al. 2015) (and cite Bolker). We only briefly discuss this
 approach as it has been desribed elsewhere (citations). Essentially, we
@@
 population model for population size \(N\) at time \(t\): \[
-N(t + 1) = N(t) + r(t) \mbox{ with } r(t) \sim N(\phi \mu, \sigma)
+N(t + 1) = N(t) + r(t) \mbox{ with } r(t) \sim N(\mu, \sigma)
 \] where \(\epsilon\) is a normally-distributed random noise term with
@@
 
-In figure \ref{fig:theoretical_approach}a, a number of simulated time
+In Fig. \ref{fig:theoretical_approach}a, a number of simulated time
 series are shown for a set number of time periods (\(t=40\)). It is
@@
 of time sampled (Fig. \ref{fig:theoretical_approach}b). Once power is
-greater than XX (the dotted line), that is the minimum time required
-(\(t_{min}\)). We can also examine how the minimum time required changes
-with effect size and variability (\textbf{make graph of this here})
+greater than 0.8 (the dotted line), that is the minimum time required
+(\(T_{min}\)).
 
@@
 vertical, dashed line is the minimum time required to achieve our
-desired statistical power.\label{fig:theoretical_approach}}
+desired statistical power. (c) Minimum time required for different
+values of the trend strength (\(r\)) and (d) minimum time required for
+different levels of population variability
+(\(\sigma\)).\label{fig:theoretical_approach}}
 \end{figure}
 
-\subsubsection{Data source}\label{data-source}
+\subsection{Data source}\label{data-source}
 
@@
 Database (NERC Centre for Population Biology 2010). We filtered out
-short time series (less than 35), and those with missing data, and were
-left with 1019 time series. The database includes information on 499
+short time series (less than 35 years), and those with missing data, and
+were left with 878 time series. The database includes information on 478
 vertebrate species with a focus on mammals, birds, and fish. The
@@
 
-For a subset of populations in the database (X = XX), we were able to
-connect a database on biological characteristics like body size and
-generation length. We used
+For a subset of populations in the database (\(n =\) 547), we were able
+to connect a database on biological characteristics (\textbf{cite
+database here}) like body size and generation length. All 547
+populations are within the Aves class. In the supplementary material, we
+explore how the minimum time required does not strongly correlate with
+any biological characteristics.
 
-\subsubsection{Empirical approach}\label{empirical-approach}
+\subsection{Empirical approach}\label{empirical-approach}
 
@@
 include all necessary information (e.g.~variability) about the
-population. In other words, they are a representative sample. We first
-take all possible contiguous subsamples of each time series. For
-example, a time series of 40 years would have 39 possible contiguous
+population. In other words, each time series is a representative sample.
+We first take all possible contiguous subsamples of each time series.
+For example, a time series of 40 years would have 39 possible contiguous
 subsamples of length 2, 38 possible contiguous subsamples of length 3,
-and continueing until 1 possible contiguous subsample of length 40. Then
-we determined the \(T_{min}\) required to be confident (i.e.~have high
+and continuing until 1 possible contiguous subsample of length 40
+(Gerber, DeMaster, and Kareiva 1999, Giron-Nava et al. (2017)). Then we
+determined the \(T_{min}\) required to be confident (i.e.~have high
 enough statistical power) in an estimate of the slope coefficient from
@@
 \includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-4-1.pdf}
-\caption{(a) Population size of ANIMAL NAME over time. The red line is
-the best fit line from linear regresssion. (b) statistical power for
+\caption{(a) Population size of Harp Seals over time. The line is the
+best fit line from linear regresssion. (b) Statistical power for
 different subsets of the time series in panel
@@
 
-\subsection{Results}\label{results}
+\section{Results}\label{results}
 
@@
 
-\subsubsection{Simulation approach}\label{simulation-approach}
+\subsection{Simulation approach}\label{simulation-approach-1}
 
@@
 
-\subsubsection{Empirical test}\label{empirical-test}
+\subsection{Empirical test}\label{empirical-test}
 
@@
 
-We examined a database of 868 separate population time series
-representing 475 species. This database consists of species with a
+We examined a database of 878 separate population time series
+representing 478 species. This database consists of species with a
 variety of life history characteristics. We limited our analyses to
@@
 Across all the populations we examined, we found an average minimum time
-series length required (\(T_{min}\)) of 16.6 (\(\sigma\)=8.7), with a
+series length required (\(T_{min}\)) of 16.5 (\(\sigma\)=8.4), with a
 wide distribution (Fig. \ref{fig:min_time_dist}). In addition, as
@@
 three of these explanatory variables were significant and had large
-effect sizes (see Fig. SX). Combined trend strength, population
-variance, and autocorrelation account for about 70.8 of the explained
+effect sizes (see Fig. SX). Combined, trend strength, population
+variance, and autocorrelation account for about 72.6 of the explained
 deviance, which is similar to \(R^2\) (Zuur et al. 2009), in minimum
@@
 We also examined a subset of populations where more explanatory
-variables were known (e.g.~body mass, longevity, location). None (need
-to do regression) of these explanatory variables were predictive of the
-minimum time series length required (supp mat). Some differences if
-\(t_{min}\) did occur between biological class (supp mat), but this can
-be attributed to the amount populations are known to fluctuate (i.e.~the
+variables were known (e.g.~body mass, longevity, location). None of
+these explanatory variables were predictive of the minimum time series
+length required (\textbf{supp mat}). Some differences if \(T_{min}\) did
+occur between biological class (\textbf{supp mat}), but this can be
+attributed to the amount populations are known to fluctuate (i.e.~the
 variability in population size).
 
-\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-7-1.pdf}
-
-\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-8-1.pdf}
-
-\subsubsection{Questions in the
-appendix}\label{questions-in-the-appendix}
+\subsection{Questions in the appendix}\label{questions-in-the-appendix}
 
@@
 
-\subsection{Discussion}\label{discussion}
+\section{Discussion}\label{discussion}
 
-We show two different approaches to estimate the minimum time series
-length required required to address a particular question of interest.
-The simulation-based approach has been suggested by others, especially
-in situations where analysis is more complicated than a simple power
-analysis (Gerrodette 1987). Our simulations support past work\ldots{}
+We explored two different approaches to estimate the minimum time series
+length required to address a particular question of interest. We
+specifically asked, what is the minimum time series length required to
+determine longterm changes in abundance using linear regression? The
+simulation-based approach has been suggested by others, especially in
+situations more complicated than that suited for classic power analysis
+(Gerrodette 1987, Johnson2015). Our simulations support past work that
+longer time series are needed when the trend strength (i.e.~rate of
+increase or decrease) is weak or when population variability is high
+(Gerrodette 1987).
 
 Here, we focus on a empirical approach to estimate the minimum time
-series length requried to address a given question.
-
-First we asked, what is \(t_{min}\) for determining whether a population
-was increasing or decreasing over time according to linear regression?
-We found that for 868 populations, at least 10-15 years of continuous
-monitoring were neccessary (Fig. \ref{fig:fig:min_time_dist}). In line
-with theoretical predictions (citations), we also found \(t_{min}\) was
-strongly correlated with the trend strength, variability in population
-size, and temporal autocorrelation (cite figure).
-
-We also found that \(t_{min}\) did not correlate with any biological
-variables of interest (cite figure or supp mat). We initially
-hypothesized that species with longer lifespans or generation times may
-require more years to sample. (Our result could have been the effect of
-two reasons) First, the database we used may not include a diverse
-enough set of species with different life history traits. Second, the
-question we poised, whether a population is increasing or decreases, is
-only worried about population surveys. Therefore, biological variables
-may be more important if the question itself was different\ldots{}
-
-\subsubsection{Other questions
-addressed}\label{other-questions-addressed}
-
-In the appendix we asked two additional questions to make estimates of
-\(t_{min}\). First, we asked how many years of sampling are neccessary
-to estimate a geometric growth rate?
-
-\subsubsection{Past work on this issue - can I integrate this better
-into the rest of the section. How much should go in
-introduction?}\label{past-work-on-this-issue---can-i-integrate-this-better-into-the-rest-of-the-section.-how-much-should-go-in-introduction}
-
-There have been previous attempts to address how long a time series
-should be to assess trends in abundance. For example,\ldots{}
-
-Rueda-Cediel et al. (2015) found that 10-15 years was adequate to
-characterize population trends in a short lived animal. They used a
-matrix model of a particular species\ldots{}
-
-Gerber, DeMaster, and Kareiva (1999) investigated the minimum time
-series required to estimate population growth of an endagered, but
-recovering, species. They used a longterm census to retroactively
-determine the minimum time series required to assess threat status. They
-found that only XX years were needed, XX years before the delisting
-decision was actually made. This highlights the importance of estimating
-the minimunm time series required as an earlier decision would have
-saved time and money (Gerber, DeMaster, and Kareiva 1999).
+series length requried to assess changes in abundance over time. We
+examined long time series and then subsampled each to determine the
+minimum time required to achieve a desired level of significance and
+power for linear regression. Statistical power is important as it
+provides on information as to the neccessary samples required to
+dertermine a significant trend (Legg and Nagy 2006). We found that for
+878 populations, at least 10-15 years of continuous monitoring were
+neccessary (Fig. \ref{fig:min_time_dist}). This timeframe is in line
+with past work on a short-lived snail species (Rueda-Cediel et al. 2015)
+and a long-lived whale species (Gerber, DeMaster, and Kareiva 1999). In
+line with theoretical predictions (citations), we also found \(T_{min}\)
+was strongly correlated with the trend strength, variability in
+population size, and temporal autocorrelation (Fig.
+\ref{fig:correlates}). Against our prior expectations, we also found
+that \(T_{min}\) did not correlate with any biological variables of
+interest (cite figure or supp mat). We initially hypothesized that
+species with longer lifespans or generation times may require a longer
+sampling period. Our result could have result of at least two effects.
+First, the database we used may not include a diverse enough set of
+species with different life history traits. Second, the question we
+poised, whether a population is increasing or decreases, is only worried
+about population surveys. Therefore, biological variables may be more
+important if the question itself was different. ANOTHER SENTENCE HERE
 
 There has also been interest in determining the minimum time series
-length in the bird community (citations). (Bart 2004, Hatch 2003) (Seavy
-and Reynolds 2007) Hatch 2003 used seabird monitoring data to estimate
+length in the bird community (Hatch 2003, ({\textbf{???}}), Seavy and
+Reynolds (2007)). Hatch (2003) used seabird monitoring data to estimate
 the minimum time needed to achieve appropriate levels of power. He found
 that the minimum time required ranged between 11 and 69 years depending
-on species, trend strength, and study design (cite Hatch). Similarly,
-Bart et al 2004 recommended a standard of 80\% power to detect a 50\%
+on species, trend strength, and study design. Similarly,
+({\textbf{???}}) recommended a standard of 80\% power to detect a 50\%
 decline over 20 years in landbird populations. They found that only
@@
 decline over 10 years almost always resulting in high statistical power
-(above 0.8). Therefore, they cautionsioned against only using power
-analyses to design monitoring schemes and instead argued for precision
-based metrics ..
+(above 0.8). Therefore, they cautioned against only using power analyses
+to design monitoring schemes and instead argued for precision based
+metrics ..
 
@@
 
-\subsubsection{Limitations}\label{limitations}
+\subsection{Other questions addressed}\label{other-questions-addressed}
 
-Our work has obvious limitations in determining the minimum time series
-length required. First, \(t_{min}\) depends strongly on the question
-being asked and the availability of long-term census data. For our
-empirical approach, the subsampling of the full time series allows for
-estimates of power, but the individual subsamples are clearly not
-independent of one another. (sentence about alpha and power
-limititations here) The perfect scenario would be to have a population
-model built and parameterized for each population of interest. Then,
-model simulations could be used to estimate the minimum time series
-reuqired to address a specific question of interest. Clearly, this is
-not always practical, especially if running analyses for a wide array of
-species as we do here. In addition, our statistical models suggest that
-\(t_{min}\) does not correlate with any biological variables of
-interest, at least for the question of linear regression. Therefore, it
-is not possible to use these results to predict \(t_{min}\) for another
-population, even if the population is of a species with a similar
-life-history to one in our database.
+In the appendix we asked two additional questions to make estimates of
+\(T_{min}\). First, we asked how many years of sampling are neccessary
+to estimate a geometric growth rate?
 
-\subsubsection{Conclusions}\label{conclusions}
+\subsection{Limitations}\label{limitations}
 
-\begin{itemize}
-\tightlist
-\item
-  what we found, implications, future work
-\end{itemize}
+Our work has some limitations in determining the minimum time series
+length required. First, \(T_{min}\) depends strongly on the question of
+interest. For our empirical approach, the subsampling of the full time
+series allows for estimates of power, but the individual subsamples are
+clearly not independent of one another. In addition, estimates of
+\(T_{min}\) depend on chosen values of \(\alpha\) and \(\beta\) (see
+appendix). In an ideal world, we would build a specific population model
+parameterized for each population of interest. Then, model simulations
+could be used to estimate the minimum time series reuqired to address a
+specific question of interest. Clearly, this is not always practical,
+especially if running analyses for a wide array of species as we do
+here. In addition, our statistical models suggest that \(T_{min}\) does
+not correlate with any biological variables of interest, at least for
+the question of linear regression. Therefore, it is not possible to use
+these results to predict \(T_{min}\) for another population, even if the
+population is of a species with a similar life-history to one in our
+database.
 
-We use a database of 868 populations to determine the minimum time
-series length required. We show that to assess longterm trends in
+\subsection{Conclusions}\label{conclusions}
+
+We use a database of 878 populations to determine the minimum time
+series length required. We show that to identify longterm changes in
 abundance, 10-15 years of continuos monitoring are required (Fig.
-\ref{fig:min_time_dist}. In line with theoretical predictions, we also
-show that \(t_{min}\) is strongly correlated with the overall trend in
+\ref{fig:min_time_dist}). In line with theoretical predictions, we also
+show that \(T_{min}\) is strongly correlated with the overall trend in
 abundance, variability in abundance, and the temporal autocorrelation
@@
 census data less than 10-15 years are probably not reliable. This
-stresses the importance of longterm monitoring programs. (where do IUCN
-recommendations come from?) From both a scientific and management
-perspective estimates of \(t_{min}\) are important. If a time series is
-too short, we are likely to make incorrect conclusions about longterm
-trends. In addition, a time series that is too long may be a poor use of
-already limited funds (Gerber, DeMaster, and Kareiva 1999). Future work
-can examine other species, with a wider range of life history
-characteristics, or address other questions besides longterm trends in
-abundance.
+stresses the importance of longterm monitoring programs. From both a
+scientific and management perspective estimates of \(T_{min}\) are
+important. If a time series is too short, we are likely to make
+incorrect conclusions about longterm trends. In addition, a time series
+that is too long may be a poor use of already limited funds (Gerber,
+DeMaster, and Kareiva 1999). Future work can examine other species, with
+a wider range of life history characteristics. In addition, similar
+approaches can be used to determine the minimum time series length
+required to address other questions.
 
-\subsection{Supporting Information}\label{supporting-information}
+\section{Supporting Information}\label{supporting-information}
 
-\begin{itemize}
-\tightlist
-\item
-  list each bit of supporting information. Github reference
-\end{itemize}
+In the supporting material, we provide an expanded methods sections,
+additional figures, and additional analyses of other time series
+questions. All code and data can be found at www.github.com/erwhite1
 
-\subsection{Acknowledgements}\label{acknowledgements}
+\section{Acknowledgements}\label{acknowledgements}
 
@@
 
-\subsection*{References}\label{references}
-\addcontentsline{toc}{subsection}{References}
+\section{References}\label{references}
 
@@
 doi:\href{https://doi.org/10.2307/1939220}{10.2307/1939220}.
+
+\hypertarget{ref-Giron-Nava2017}{}
+Giron-Nava, Alfredo, Chase C James, Andrew F Johnson, David Dannecker,
+Bethany Kolody, Adrienne Lee, Maitreyi Nagarkar, et al. 2017.
+``Quantitative argument for long-term ecological monitoring.''
+\emph{Marine Ecology Progress Series} 572: 269--74.
+
+\hypertarget{ref-Hatch2003}{}
+Hatch, S A. 2003. ``Statistical power for detecting trends with
+applications to seabirds monitoring.'' \emph{Biological Conservation}
+111: 317--29.
 
@@
 
+\section{Figures}\label{figures}
+
+stuff here
+
+\section{Supplementary material}\label{supplementary-material}
+
 \end{document}

### 2017-06-28 writing/supp_mat_draft.Rmd
added to discussion and created cover letter

@@
 1. Detailed example of subsampling and power calculations
-2. Additional figures
+2. Additional results
+  - sensitivity of alpha and beta
+  - other
 3. Figures and results from other questions of interest
@@
 This resulted in lower explained deviance (analogue to $R^2$) of `r round(R2_no_variance,2)`% and `r round(R2_no_trend,2)`%, respectively.
+
+\pagebreak
+
+### Minimum time series and biological class
+
+```{r,echo=F,include=FALSE, eval=FALSE,fig.cap='(a) Minimum time required to estimate change in abundance for species class, (b) Longterm trend (estimated slope coefficient) by species class, (c) interannual variability in population size by species class, and (d) temporal autocorrelation by species class.\\label{fig:biological_correlates}'}
+# Biological correlates plots
+par(mfrow=c(2,2),mar=c(4,5,0.5,0.5))
+
+boxplot(pop_info$min_time_for_power~pop_info$class, lwd = 1, ylab = 'minimum time required',col='white',cex.lab=1.4,cex.axis=1.2,las=1)
+stripchart(pop_info$min_time_for_power~pop_info$class, vertical = TRUE,method = "jitter", add = TRUE, pch = 20, col = rgb(0.5,0.5,0.5,0.5))
+
+boxplot(abs(pop_info$overall_trend)~pop_info$class, lwd = 1, ylab = ' ',col='white',cex.lab=1.4,cex.axis=1.2,las=1)
+stripchart(abs(pop_info$overall_trend)~pop_info$class, vertical = TRUE,method = "jitter", add = TRUE, pch = 20, col = rgb(0.5,0.5,0.5,0.5))
+  mtext('| trend strength |',2,line=3.5,cex=1.2)
+
+boxplot(pop_info$variance~pop_info$class, lwd = 1, ylab = 'population variance',col='white',cex.lab=1.4,cex.axis=1.2,las=1)
+stripchart(pop_info$variance~pop_info$class, vertical = TRUE,method = "jitter", add = TRUE, pch = 20, col = rgb(0.5,0.5,0.5,0.5))
+
+boxplot(pop_info$autocorrelation~pop_info$class, lwd = 1, ylab = 'autocorrelation',col='white',cex.lab=1.4,cex.axis=1.2,las=1)
+stripchart(pop_info$autocorrelation~pop_info$class, vertical = TRUE,method = "jitter", add = TRUE, pch = 20, col = rgb(0.5,0.5,0.5,0.5))
+```
+
+\pagebreak
+
+### Sensitivity analysis of $\alpha$ and $\beta$
+
+Estimates of $T_{min}$ depend strongly on the values used for the significance level ($\alpha$) and the probability of type II error ($\beta$), both of which are set by the practioner.
+Here we explore how estimates of $T_{min}$ are affected by changes in each of these parameters.
+
+```{r,echo=F}
+# Bring in theoretical code
+# Build plot of $T_min$ vs alpha and another versus beta
+# Explain results after building figures
+
+```
 

### 2017-06-29 writing/manuscript_draft.Rmd
added sensitivity analysis to alpha and beta

@@
 t-test), and corresponding formulas for determinig sample size, simulation approaches can be used [@Gerrodette1987].
-Simple population models can be simulated with parameter values corresponding to a population of interest.
+Simple models can be simulated with parameter values corresponding to a population of interest.
 Out of a large number of simulations, power can be calculated as the fraction of simulations that meet some criteria.
 The specific criteria depend on the question at hand.
-Then time series simulations of different lengths are run to determine the minimum time series length required to be confident that you will be able to answer your question (cite Bolker).
+Then, time series simulations of different lengths are run to determine the minimum time series length required to be confident that you will be able to answer your question (cite Bolker).
 In addition to using simulated data, empirical time series can also be used.
-We usually do not have multiple replicates of similar populations, but we can subsample an empirical time series (cite).
+We usually do not have multiple replicates of similar populations, but we can subsample an empirical time series [@Gerber1999].
 Subsamples of different lengths can then be evaluated to see which fraction of subsamples meet some criteria, again a measure of statistical power.
@@
 Using simple populations models, they found that the allocation of effort depends on environmental variation, spatial and temporal autocorrelation, and observer error.
-@Rueda-Cediel2015 also used a modeling approach, but parameterized it for a threatened snail, \emph{Tasmaphena lamproides}.
-They found that for this short-lived organims, that 15 years was adequate to assess longterm trends in abundance.
-However, these studies, and other past work, have focused either on theoretical aspects of monitoring design or focused on single species.
+@Rueda-Cediel2015 also used a modeling approach, but parameterized amodel specifically for a threatened snail, \emph{Tasmaphena lamproides}.
+They found that for this short-lived organim, 15 years was adequate to assess longterm trends in abundance.
+However, these studies, and other past work, have typically focused either on theoretical aspects of monitoring design or focused on single species.
 
@@
 More specifically, we estimate the minimum time series length required to asssess longterm changes in abundance via simple linear regression.
+We examine 878 population time series to estimate this minimum time series length required.
 <!-- In this paper, we evaluate the minimal time series length required to answer X questions.
@@
 By simulating this model repetitively for different lengths of time, we can determine the minimum time series required.
-Here statistical power is simply the fraction of simulations with significant trends in abundance from simple linear regression.
-Then, the minimum time required is the minimum time required to obtain statistical power above some pre-determined threshold (historically at 0.8).
+Here statistical power is simply the fraction of simulations with significant trends in abundance.
+A significant trend in abundance is determined by estimating the significance of the slope parameter from simple linear regression.
+Then, the minimum time required ($T_{min}$) is the number of years required to obtain statistical power above some pre-determined threshold (historically at 0.8).
+Statistical power of 0.8 would indicate that, if there was a true trend in abundance, we would have a 0.8 probability of detecting the trend.
 
@@
 Lastly, we determine which subsample length is required to guarantee we have above a certain threshold of statistical power, which we will assume is 0.8 by convention (citation).
-A power of 0.8 would imply that given a trend exists, there is a 0.8 probability of detecting the trend.
-Complementary, there would be a 0.2 probability of a false negative.
-
-<!-- example of how subsampling and empirical approach works -->
-```{r, echo=F,message=F,warning=F,fig.cap='(a) Population size of Harp Seals over time.
-The line is the best fit line from linear regresssion.
-(b) Statistical power for different subsets of the time series in panel a.\\label{fig:empirical_approach_example}'}
-# Example of subsampling empirical data
- setwd("~/Desktop/Research/time-series-project")
- load("cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
-  par(mfrow=c(1,2),oma=c(4,4,0.5,0.5))
-    pop = subset(long_dat,long_dat$ID=='13192') #13192 for Harp seals
-    plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='',xlab='time (years)',pch=16,las=1)
-    abline(lm(pop$popvalue~pop$year),lwd=2,col='red')
-    mtext(text = '(a)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
-    mtext(text='population size', side=2,line=4,cex=1.2)
-# could insert all possible regressions of length n
-source('scripts/calculate_power_metric.R')
-    power_example=calculate_power_for_subsamples(pop$popvalue,0.05)
-    plot(power_example,cex.axis=1.2,cex.lab=1.2,ylab='statistical power',xlab='years sampled',las=1,pch=16)
-    mtext(text = '(b)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
-# could also insert standard error bars
-    abline(h=0.8,col='red',lwd=2,lty=2)
-    abline(v=head(which(power_example>0.8),1),col='red',lwd=2,lty=2)
-
-```
+<!--In Figure \ref{fig:empirical_approach_example}b, statistical power increases with more time sampled until our desired power of 0.8 at 16 years.
+Therefore, 16 years would be our estimate for $T_{min}$.-->
 

### 2017-06-29 writing/manuscript_draft.tex
added sensitivity analysis to alpha and beta

@@
 required to estimate trends in population size over time (Giron-Nava et
-al. 2017). Rhodes and Jonzen (2011) examined the optimal allocation of
-effort between spatial and temporal replicates. Using simple populations
-models, they found that the allocation of effort depends on
-environmental variation, spatial and temporal autocorrelation, and
-observer error. Rueda-Cediel et al. (2015) also used a modeling
-approach, but parameterized it for a threatened snail,
+al. 2017). For example, Rhodes and Jonzen (2011) examined the optimal
+allocation of effort between spatial and temporal replicates. Using
+simple populations models, they found that the allocation of effort
+depends on environmental variation, spatial and temporal
+autocorrelation, and observer error. Rueda-Cediel et al. (2015) also
+used a modeling approach, but parameterized it for a threatened snail,
 \emph{Tasmaphena lamproides}. They found that for this short-lived
 organims, that 15 years was adequate to assess longterm trends in
-abundance. However, these studies, and other past work (citation), have
-focused either on theoretical aspects of monitoring design or focused on
-single species.
+abundance. However, these studies, and other past work, have focused
+either on theoretical aspects of monitoring design or focused on single
+species.
 
@@
 series to determine the minimum number of years required to address
-various questions. Here, we determine the minimum time series length to
-determine if a population is incresing or decreasing according to a
-significant slope coefficient in simple linear regression.
+various questions. More specifically, we estimate the minimum time
+series length required to asssess longterm changes in abundance via
+simple linear regression.
 

### 2017-06-29 writing/supp_mat_draft.Rmd
added sensitivity analysis to alpha and beta

--- 2017-06-14 writing/supp_mat_draft.Rmd
@@
 
+<!--
 This supplementary material includes:
@@
 1. Detailed example of subsampling and power calculations
-2. Additional figures
+  - figure of subsampling routine
+2. Additional results
+  - statistical model results of main text
+  - biological correlates plot
+  - minimum time required and biological class
+  - sensitivity of alpha and power
 3. Figures and results from other questions of interest
+  - simulations with more realistic population model
+  - empirical example with more realistic population model
+  - minimum time required to estimate longterm geometric growth rate
+  - empirical approach with longterm geometric growth rate
+  - estimating minimum time required for generalized additive models
+
 4. Additional references
+-->
+
+\tableofcontents
 
@@
 
-## Detailed example of subsampling and power calculations
+# Detailed example of subsampling and power calculations
 
@@
 For each subsample, we conduct linear regression and extract model coefficients and p-values.
-Then, the fraction of subsamples within a set that show significant trends (signfiiant slope coefficient) is the statistical power.
+Then, the fraction of subsamples within a set that show significant trends (significant slope coefficient) is the statistical power.
 It is important to note that we only consider subsamples to be significant if they are significant in the same direction as the complete 35-year time series.
@@
 
-Then, we determine an appropriate level of statistical power that we find acceptable. Traditionally, this has been at 0.8, however, this is purely historical. Statisical power of 0.8 implies...
+Then, we determine an appropriate level of statistical power that we find acceptable. Traditionally, this has been at 0.8, however, this is purely historical. Statistical power of 0.8 implies...
 
@@
 
-```{r, echo=F}
-# show an example of subsampling and power calculation routines
-```
-
-## Additional results from the main manuscript
-
-```{r, echo=F}
+<!-- example of how subsampling and empirical approach works -->
+```{r, echo=F,message=F,warning=F,fig.cap='(a) Population size of Harp Seals over time.
+The line is the best fit line from linear regresssion.
+(b) Statistical power for different subsets of the time series in panel a.\\label{fig:empirical_approach_example}'}
+# Example of subsampling empirical data
  setwd("~/Desktop/Research/time-series-project")
-load('cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
-```
-
-In the main manuscript, we examined the minimum time required to detect a significant trend in abundance over time using linear regression.
-As detailed in the main manuscript the minimum time required was around 15, but there was a wide distribution.
-Therefore, we were interested in potential explanatory variables of the minimum time required.
-In the main manuscript, we examined characteristics of the time series itself, like variability, autocorrelation, and the trend in abundance over time.
-Here, we combined our time series database with a database on life history characteristics of amniotes from (cite other database here).
-There was life history information available for `r nrow(LPI_pop_info)` populations representing `r length(table(LPI_pop_info$Binomial))` different species, all of which were in the Aves class.
-
-We then correlated minimum time required for each population with its corresponding life history characteristics.
-In figure \ref{fig:biological_correlates} we examined miminmum time required versus generation length (years), litter size (n), adult body mass (grams), maximum longevity (years), egg mass (grams), and incubation (days).
-None of these variables had much explanatory power in accounting for the variance in the minimum time required.
-see table \ref{table:model_output}}
-
-```{r, echo=F, fig.cap='Mimimum time required versus (a) generation length (years), (b) litter size (n), (c) adult body mass (grams), (d) maximum longevity (years), (e) egg mass (grams), and (f) incubation (days).
-The lines in each plot represent the best fit line from linear regression.\\label{fig:biological_correlates}',fig.height=6}
-
-par(mfrow=c(3,2),oma=c(5,4,1,1),mar=c(5,3,0,0))
-
-n=1
-x_axis_name_list = c('generation length (years)','litter size (n)','adult body mass (grams)','maximum longevity (years)','egg mass (grams)','incubation (days)')
-for (x_axis_name in names(LPI_pop_info)[20:25]){
-  plot(LPI_pop_info[[x_axis_name]],LPI_pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
-  abline(lm(LPI_pop_info$min_time_for_power~LPI_pop_info[[x_axis_name]]),lwd=3)
-
-  mtext(text = x_axis_name_list[n],side = 1,line = 3,cex = 1.2)
-  mtext(text = paste('(',letters[n],')',sep=''),side = 3,line=-1.5,adj = 0.98,cex=1.2)
-  n=n+1
-}
-
-mtext(text = 'minimum time required',side = 2,line = 2,cex = 1.4,outer = TRUE)
+ load("cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
+  par(mfrow=c(1,2),oma=c(4,4,0.5,0.5))
+    pop = subset(long_dat,long_dat$ID=='13192') #13192 for Harp seals
+    plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='',xlab='time (years)',pch=16,las=1)
+    abline(lm(pop$popvalue~pop$year),lwd=2,col='red')
+    mtext(text = '(a)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
+    mtext(text='population size', side=2,line=4,cex=1.2)
+# could insert all possible regressions of length n
+source('scripts/calculate_power_metric.R')
+    power_example=calculate_power_for_subsamples(pop$popvalue,0.05)
+    plot(power_example,cex.axis=1.2,cex.lab=1.2,ylab='statistical power',xlab='years sampled',las=1,pch=16)
+    mtext(text = '(b)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
+# could also insert standard error bars
+    abline(h=0.8,col='red',lwd=2,lty=2)
+    abline(v=head(which(power_example>0.8),1),col='red',lwd=2,lty=2)
 
@@
 
-\pagebreak
+In Figure \ref{fig:empirical_approach_example}b, statistical power increases with more time sampled until our desired power of 0.8 at 16 years. Therefore, 16 years would be our estimate for $T_{min}$.
+
+# Additional results from the main manuscript
+
+## Statistical model of minimum time required and time series characteristics
 
@@
 
-## Results from other questions of interest
+## Minimum time required and biological correlates
+```{r, echo=F}
+setwd("~/Desktop/Research/time-series-project")
+load('cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
+```
 
-### Simulations with more complicated population model
+In the main manuscript, we examined the minimum time required to detect a significant trend in abundance over time using linear regression.
+As detailed in the main manuscript the minimum time required was around 15, but there was a wide distribution.
+Therefore, we were interested in potential explanatory variables of the minimum time required.
+In the main manuscript, we examined characteristics of the time series itself, like variability, autocorrelation, and the trend in abundance over time.
+Here, we combined our time series database with a database on life history characteristics of amniotes from (cite other database here).
+There was life history information available for `r nrow(LPI_pop_info)` populations representing `r length(table(LPI_pop_info$Binomial))` different species, all of which were in the Aves class.
 
-In the main text, we showed how a simple population model could be simulated repeatidly to estimate the power obtained with time series of increasing length.
+We then correlated minimum time required for each population with its corresponding life history characteristics.
+In figure \ref{fig:biological_correlates} we examined minimum time required versus generation length (years), litter size (n), adult body mass (grams), maximum longevity (years), egg mass (grams), and incubation (days).
+None of these variables had much explanatory power in accounting for the variance in the minimum time required.
+see table \ref{table:model_output}}
+
+```{r, echo=F, fig.cap='Mimimum time required versus (a) generation length (years), (b) litter size (n), (c) adult body mass (grams), (d) maximum longevity (years), (e) egg mass (grams), and (f) incubation (days).
+The lines in each plot represent the best fit line from linear regression.\\label{fig:biological_correlates}',fig.height=6}
+
+par(mfrow=c(3,2),oma=c(5,4,1,1),mar=c(5,3,0,0))
+
+n=1
+x_axis_name_list = c('generation length (years)','litter size (n)','adult body mass (grams)','maximum longevity (years)','egg mass (grams)','incubation (days)')
+for (x_axis_name in names(LPI_pop_info)[20:25]){
+  plot(LPI_pop_info[[x_axis_name]],LPI_pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
+  abline(lm(LPI_pop_info$min_time_for_power~LPI_pop_info[[x_axis_name]]),lwd=3)
+
+  mtext(text = x_axis_name_list[n],side = 1,line = 3,cex = 1.2)
+  mtext(text = paste('(',letters[n],')',sep=''),side = 3,line=-1.5,adj = 0.98,cex=1.2)
+  n=n+1
+}
+
+mtext(text = 'minimum time required',side = 2,line = 2,cex = 1.4,outer = TRUE)
+
+```
+
+\pagebreak
+
+## Minimum time series and biological class
+
+When we examined differences in the minimum time required there were similar patterns with biological classes.
+For instance, more time is required for species within the Actinopterygii class compared to other species (Fig.
+\ref{fig:class}a).
+These differences between biological classes can be explained by differences in population variability, with species in Actinopterygii have larger variability in population size from year-to-year.
+
+```{r,echo=F,fig.cap='(a) Minimum time required to estimate change in abundance for species class, (b) longterm trend (estimated slope coefficient) by species class, (c) interannual variability in population size by species class, and (d) temporal autocorrelation by species class.\\label{fig:class}',fig.height=6}
+
+# Pull in data
+setwd("~/Desktop/Research/time-series-project")
+load('cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
+# Biological correlates plots
+par(mfrow=c(2,2),oma=c(8,1,1.5,0.5),mar=c(0,4,0,1))
+
+boxplot(pop_info$min_time_for_power~pop_info$class, lwd = 1, ylab = 'minimum time required',col='white',cex.lab=1.4,cex.axis=1.2,las=2,xaxt="n")
+stripchart(pop_info$min_time_for_power~pop_info$class, vertical = TRUE,method = "jitter", add = TRUE, pch = 20, col = rgb(0.5,0.5,0.5,0.5))
+
+boxplot(abs(pop_info$overall_trend)~pop_info$class, lwd = 1, ylab = ' ',col='white',cex.lab=1.4,cex.axis=1.2,las=2,xaxt="n")
+stripchart(abs(pop_info$overall_trend)~pop_info$class, vertical = TRUE,method = "jitter", add = TRUE, pch = 20, col = rgb(0.5,0.5,0.5,0.5))
+  mtext('| trend strength |',2,line=3.5,cex=1.2)
+
+boxplot(pop_info$variance~pop_info$class, lwd = 1, ylab = 'population variance',col='white',cex.lab=1.4,cex.axis=1.2,las=2)
+stripchart(pop_info$variance~pop_info$class, vertical = TRUE,method = "jitter", add = TRUE, pch = 20, col = rgb(0.5,0.5,0.5,0.5))
+
+boxplot(pop_info$autocorrelation~pop_info$class, lwd = 1, ylab = 'autocorrelation',col='white',cex.lab=1.4,cex.axis=1.2,las=2)
+stripchart(pop_info$autocorrelation~pop_info$class, vertical = TRUE,method = "jitter", add = TRUE, pch = 20, col = rgb(0.5,0.5,0.5,0.5))
+```
+
+\pagebreak
+
+## Sensitivity analysis of $\alpha$ and power
+
+Estimates of $T_{min}$ depend strongly on the values used for the significance level ($\alpha$) and the probability of type II error ($\beta$), both of which are set by the practitioner.
+Here we explore how estimates of $T_{min}$ are affected by changes in each of these parameters.
+We see that the minimum time required increases with increases in statistical power or decreases with increases in the threshold for statistical significance (Fig \ref{fig:min_time_vs_alpha_beta}).
+
+```{r,echo=F,message=F,error=F,fig.cap='Minimum time required to assess longterm trends in abundance for values of statistical significance ($\\alpha$) and power ($1-\\beta$).\\label{fig:min_time_vs_alpha_beta}'}
+# Bring in theoretical code
+# Build plot of $T_min$ vs alpha and another versus beta
+# Explain results after building figures
+setwd("~/Desktop/Research/time-series-project")
+source('scripts/calculate_slope.R') 
+source('scripts/theoretical-models/simulate_pop.R')
+# calculate_power = function(r,phi,sigma,trials,max.time,alpha){
+#   trials   = trials
+#   max.time = max.time
+#   r        = r
+#   phi      = phi
+#   sigma    = sigma
+#   
+#   y=replicate(n=trials,simulate_pop(r,phi,sigma,max.time)) 
+#   
+#   power=sum((apply(y,2,calculate_p_value)<alpha) & sign(apply(y,2,calculate_slope))==sign(r))/trials 
+#   
+#   return(power)
+# }
+# 
+# 
+# min_time_needed= function(alpha,power){
+#   years_to_sample=seq(2,30,by=1)
+#   y=lapply(X=years_to_sample,FUN = calculate_power,phi =0.5,r=0.5,sigma = 1,trials=100,alpha=alpha)
+#   min_time_needed = years_to_sample[tail(which(y<power),1)] + 1
+#   return(min_time_needed)
+# }
+# 
+# alpha_values=seq(0.001,0.2,by=0.01)
+# power_values=seq(0.65,0.95,by=0.01)
+# parameter_combinations = expand.grid(alpha_values,power_values)
+# names(parameter_combinations) = c('alpha','power')
+# parameter_combinations$min_time_required = NA
+# 
+# for (i in 1:nrow(parameter_combinations)){
+# 
+#   parameter_combinations$min_time_required[i] =  min_time_needed(parameter_combinations[i,1],parameter_combinations[i,2])
+#  
+#   #print(i)
+# }
+
+#write.csv(parameter_combinations,file = 'analysis-outputs/min_time_vs_alpha_and_power.csv',quote = F,row.names = F)
+
+# Pull in data to prevent R from running for a long time to create data from scratch
+parameter_combinations = read.csv('analysis-outputs/min_time_vs_alpha_and_power.csv',header=T)
+
+require(ggplot2)
+ggplot(parameter_combinations, aes(x = alpha, y = power, z = min_time_required, fill=min_time_required)) +
+  geom_tile() + 
+  labs(title = " ", x = expression(alpha), y = expression(paste('power ',(1-beta))), fill = expression(T[min])) +
+  scale_fill_distiller(palette="Spectral", na.value="white") +
+  theme_bw()
+```
+
+\pagebreak
+
+# Results from other questions of interest
+
+## Simulations with more complicated population model
+
+In the main text, we showed how a simple population model could be simulated repeatedly to estimate the power obtained with time series of increasing length.
 The model in the main text simulated linear population growth with only a slope coefficient, y-intercept, and noise parameter required.
-This model is purely phenemilogical and does not include any species life history.
+This model is purely phenomenological and does not include any species life history.
 Here, we use the same routine as the main text, but simulate from a more biologically-realistic population model, the (MODEL NAME HERE):
@@
 
-We simulate this model repetively for different lengths of time series.
-Statistical power is the fraction of time series of a set length that showed signicant trends in abundance, given a true trend in abundance.
+```{r,echo=F}
+#Include simulations of more interesting model here
+```
+
+We simulate this model repetitively for different lengths of time series.
+Statistical power is the fraction of time series of a set length that showed significant trends in abundance, given a true trend in abundance.
 *How do I determine significance here?
@@
 
-### Empirical approach with more complicated population model (use model similar to de Valpine work?)
+## Empirical approach with more complicated population model (use model similar to de Valpine work?)
 
-### Growth rate calculations
+## Growth rate calculations
 
@@
 \ref{fig:growth_rate}c), we show the percent error as a function of time series length.
-Here we define the minimum time required as the minimum number of years to achieve less than 20% error.
+Here we define the minimum time required as the minimum number of years to achieve less than 10% error.
 
@@
 
-### Using Generalized additive model to identify significant trends
+## Using Generalized additive model to identify significant trends
 
@@
 GAMs are models where a response variable depends on unknown smooth functions of explanatory variables.
-GAMs, therefore, can identify relationships between response variables and explantory variables that are non-linear and perhaps more complicated.
+GAMs, therefore, can identify relationships between response variables and explanatory variables that are non-linear and perhaps more complicated.
 The downside of GAMs is they typically require more data and are also prone to overfitting (citations).
@@
 
-```{r,echo=F,fig.cap='results here from GAM.\\label{fig:gam_example}'}
+```{r,echo=F,message=F,warning=F,fig.cap='results here from GAM.\\label{fig:gam_example}'}
 # show gam example side-by-side plot
-plot(1,1)
+ setwd("~/Desktop/Research/time-series-project")
+ load("cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
+ source('scripts/calculate_power_metric_gam.R')
+  par(mfrow=c(1,2),oma=c(4,4,0.5,0.5))
+    pop = subset(long_dat,long_dat$ID=='7558') #13192 for Harp seals, 7558 for Thunnus obesus
+    plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='',xlab='time (years)',pch=16,las=1)
+    abline(lm(pop$popvalue~pop$year),lwd=2,col='red')
+    mtext(text = '(a)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
+    mtext(text='population size', side=2,line=4,cex=1.2)
+# could insert all possible regressions of length n
+
+    power_example=calculate_power_for_subsamples(pop$popvalue,0.05)
+    plot(power_example,cex.axis=1.2,cex.lab=1.2,ylab='statistical power',xlab='years sampled',las=1,pch=16)
+    mtext(text = '(b)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
+# could also insert standard error bars
+    abline(h=0.8,col='red',lwd=2,lty=2)
+    abline(v=head(which(power_example>0.8),1),col='red',lwd=2,lty=2)
 ```
@@
 
-### Minimum time to detect density-dependence in time series
+## Minimum time to detect density-dependence in time series
 
@@
 They fitted 627 time series with a Gompertz population model and found evidence for density-dependence in 45% of the time series.
-Each time series in their database contained at least 15 datapoints, and ranged from 15 to 157.
+Each time series in their database contained at least 15 data points, and ranged from 15 to 157.
 A follow up question to their work would be, what time series length is required to confidently estimate density-dependence.
@@
 \clearpage
-### References
+
+# References

### 2017-06-29 writing/supp_mat_draft.tex
added sensitivity analysis to alpha and beta

@@
 
-This supplementary material includes:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Detailed example of subsampling and power calculations
-\item
-  Additional figures
-\item
-  Figures and results from other questions of interest
-\item
-  Additional references
-\end{enumerate}
+\tableofcontents
 
@@
 
-\subsection{Detailed example of subsampling and power
+\section{Detailed example of subsampling and power
 calculations}\label{detailed-example-of-subsampling-and-power-calculations}
@@
 and extract model coefficients and p-values. Then, the fraction of
-subsamples within a set that show significant trends (signfiiant slope
+subsamples within a set that show significant trends (significant slope
 coefficient) is the statistical power. It is important to note that we
@@
 find acceptable. Traditionally, this has been at 0.8, however, this is
-purely historical. Statisical power of 0.8 implies\ldots{}
+purely historical. Statistical power of 0.8 implies\ldots{}
 
@@
 
-\subsection{Additional results from the main
+\begin{figure}[htbp]
+\centering
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-1-1.pdf}
+\caption{(a) Population size of Harp Seals over time. The line is the
+best fit line from linear regresssion. (b) Statistical power for
+different subsets of the time series in panel
+a.\label{fig:empirical_approach_example}}
+\end{figure}
+
+In Figure \ref{fig:empirical_approach_example}b, statistical power
+increases with more time sampled until our desired power of 0.8 at 16
+years. Therefore, 16 years would be our estimate for \(T_{min}\).
+
+\section{Additional results from the main
 manuscript}\label{additional-results-from-the-main-manuscript}
 
-In the main manuscript, we examined the minimum time required to detect
-a significant trend in abundance over time using linear regression. As
-detailed in the main manuscript the minimum time required was around 15,
-but there was a wide distribution. Therefore, we were interested in
-potential explanatory variables of the minimum time required. In the
-main manuscript, we examined characteristics of the time series itself,
-like variability, autocorrelation, and the trend in abundance over time.
-Here, we combined our time series database with a database on life
-history characteristics of amniotes from (cite other database here).
-There was life history information available for 547 populations
-representing 315 different species, all of which were in the Aves class.
-
-We then correlated minimum time required for each population with its
-corresponding life history characteristics. In figure
-\ref{fig:biological_correlates} we examined miminmum time required
-versus generation length (years), litter size (n), adult body mass
-(grams), maximum longevity (years), egg mass (grams), and incubation
-(days). None of these variables had much explanatory power in accounting
-for the variance in the minimum time required. see table
-\ref{table:model_output}\}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-3-1.pdf}
-\caption{Mimimum time required versus (a) generation length (years), (b)
-litter size (n), (c) adult body mass (grams), (d) maximum longevity
-(years), (e) egg mass (grams), and (f) incubation (days). The lines in
-each plot represent the best fit line from linear
-regression.\label{fig:biological_correlates}}
-\end{figure}
-
-\pagebreak
+\subsection{Statistical model of minimum time required and time series
+characteristics}\label{statistical-model-of-minimum-time-required-and-time-series-characteristics}
 
@@
 
-\subsection{Results from other questions of
+\subsection{Minimum time required and biological
+correlates}\label{minimum-time-required-and-biological-correlates}
+
+In the main manuscript, we examined the minimum time required to detect
+a significant trend in abundance over time using linear regression. As
+detailed in the main manuscript the minimum time required was around 15,
+but there was a wide distribution. Therefore, we were interested in
+potential explanatory variables of the minimum time required. In the
+main manuscript, we examined characteristics of the time series itself,
+like variability, autocorrelation, and the trend in abundance over time.
+Here, we combined our time series database with a database on life
+history characteristics of amniotes from (cite other database here).
+There was life history information available for 547 populations
+representing 315 different species, all of which were in the Aves class.
+
+We then correlated minimum time required for each population with its
+corresponding life history characteristics. In figure
+\ref{fig:biological_correlates} we examined minimum time required versus
+generation length (years), litter size (n), adult body mass (grams),
+maximum longevity (years), egg mass (grams), and incubation (days). None
+of these variables had much explanatory power in accounting for the
+variance in the minimum time required. see table
+\ref{table:model_output}\}
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-4-1.pdf}
+\caption{Mimimum time required versus (a) generation length (years), (b)
+litter size (n), (c) adult body mass (grams), (d) maximum longevity
+(years), (e) egg mass (grams), and (f) incubation (days). The lines in
+each plot represent the best fit line from linear
+regression.\label{fig:biological_correlates}}
+\end{figure}
+
+\pagebreak
+
+\subsection{Minimum time series and biological
+class}\label{minimum-time-series-and-biological-class}
+
+When we examined differences in the minimum time required there were
+similar patterns with biological classes. For instance, more time is
+required for species within the Actinopterygii class compared to other
+species (Fig. \ref{fig:class}a). These differences between biological
+classes can be explained by differences in population variability, with
+species in Actinopterygii have larger variability in population size
+from year-to-year.
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
+\caption{(a) Minimum time required to estimate change in abundance for
+species class, (b) longterm trend (estimated slope coefficient) by
+species class, (c) interannual variability in population size by species
+class, and (d) temporal autocorrelation by species
+class.\label{fig:class}}
+\end{figure}
+
+\pagebreak
+
+\subsection{\texorpdfstring{Sensitivity analysis of \(\alpha\) and
+power}{Sensitivity analysis of \textbackslash{}alpha and power}}\label{sensitivity-analysis-of-alpha-and-power}
+
+Estimates of \(T_{min}\) depend strongly on the values used for the
+significance level (\(\alpha\)) and the probability of type II error
+(\(\beta\)), both of which are set by the practitioner. Here we explore
+how estimates of \(T_{min}\) are affected by changes in each of these
+parameters. We see that the minimum time required increases with
+increases in statistical power or decreases with increases in the
+threshold for statistical significance (Fig
+\ref{fig:min_time_vs_alpha_beta}).
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-6-1.pdf}
+\caption{Minimum time required to assess longterm trends in abundance
+for values of statistical significance (\(\alpha\)) and power
+(\(1-\beta\)).\label{fig:min_time_vs_alpha_beta}}
+\end{figure}
+
+\pagebreak
+
+\section{Results from other questions of
 interest}\label{results-from-other-questions-of-interest}
 
-\subsubsection{Simulations with more complicated population
+\subsection{Simulations with more complicated population
 model}\label{simulations-with-more-complicated-population-model}
@@
 In the main text, we showed how a simple population model could be
-simulated repeatidly to estimate the power obtained with time series of
+simulated repeatedly to estimate the power obtained with time series of
 increasing length. The model in the main text simulated linear
 population growth with only a slope coefficient, y-intercept, and noise
-parameter required. This model is purely phenemilogical and does not
+parameter required. This model is purely phenomenological and does not
 include any species life history. Here, we use the same routine as the
@@
 
-We simulate this model repetively for different lengths of time series.
-Statistical power is the fraction of time series of a set length that
-showed signicant trends in abundance, given a true trend in abundance.
-\emph{How do I determine significance here? Through linear regression
-again, or do I look at non-zero estimation of the parameter values?}
+We simulate this model repetitively for different lengths of time
+series. Statistical power is the fraction of time series of a set length
+that showed significant trends in abundance, given a true trend in
+abundance. \emph{How do I determine significance here? Through linear
+regression again, or do I look at non-zero estimation of the parameter
+values?}
 
-\subsubsection{Empirical approach with more complicated population model
+\subsection{Empirical approach with more complicated population model
 (use model similar to de Valpine
@@
 
-\subsubsection{Growth rate calculations}\label{growth-rate-calculations}
+\subsection{Growth rate calculations}\label{growth-rate-calculations}
 
@@
 define the minimum time required as the minimum number of years to
-achieve less than 20\% error.
-
-\begin{verbatim}
-## Loading required package: plyr
-\end{verbatim}
+achieve less than 10\% error.
 
@@
 \centering
-\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-8-1.pdf}
 \caption{Example of calculating minimum time required for growth rate
@@
 
-\subsubsection{Using Generalized additive model to identify significant
+\subsection{Using Generalized additive model to identify significant
 trends}\label{using-generalized-additive-model-to-identify-significant-trends}
@@
 variables. GAMs, therefore, can identify relationships between response
-variables and explantory variables that are non-linear and perhaps more
+variables and explanatory variables that are non-linear and perhaps more
 complicated. The downside of GAMs is they typically require more data
@@
 \centering
-\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-7-1.pdf}
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-10-1.pdf}
 \caption{results here from GAM.\label{fig:gam_example}}
@@
 
-\subsubsection{Minimum time to detect density-dependence in time
+\subsection{Minimum time to detect density-dependence in time
 series}\label{minimum-time-to-detect-density-dependence-in-time-series}
@@
 45\% of the time series. Each time series in their database contained at
-least 15 datapoints, and ranged from 15 to 157. A follow up question to
+least 15 data points, and ranged from 15 to 157. A follow up question to
 their work would be, what time series length is required to confidently
@@
 \clearpage
-\#\#\# References
 
+\renewcommand\refname{References}
 \bibliography{White_bib.bib}

### 2017-07-05 scripts/scripts_to_clean_data/create-pop-info-table.R
edited file to only include most 35 recent years of each time series

@@ -39,24 +39,30 @@ pop_info$autocorrelation= NA
 pop_info$trend_p_value=NA
 pop_info$min_time_for_power=NA
 pop_info$popvalue=NA
-long_dat$pop_value=long_dat$pop_value
+time_series_data=NULL
 source('scripts/calculate_slope.R')
   for (j in 1:nrow(pop_info)){
     pop= subset(long_dat,long_dat$ID==pop_info$ID[j])
     pop = pop[(nrow(pop)-34):nrow(pop),]
     pop$popvalue=as.numeric(pop$popvalue)
     pop$popvalue=pop$popvalue/max(pop$popvalue) #should I standaridize this in a better way
-    # pop_info$variance[j]= var(pop$popvalue)
-    # pop_info$overall_trend[j] = calculate_slope(pop$popvalue)
-    # pop_info$autocorrelation[j] = acf(pop$popvalue,plot = FALSE)$acf[2]
-    # pop_info$trend_p_value[j]=calculate_p_value(pop$popvalue)
-    # pop_info$popvalue = pop$popvalue[1] #initial abundance
-    # pop_info$series_length = nrow(pop)
+    pop_info$year[j] = pop$year[1]
+    pop_info$variance[j]= var(pop$popvalue)
+    pop_info$overall_trend[j] = calculate_slope(pop$popvalue)
+    pop_info$autocorrelation[j] = acf(pop$popvalue,plot = FALSE)$acf[2]
+    pop_info$trend_p_value[j]=calculate_p_value(pop$popvalue)
+    pop_info$popvalue = pop$popvalue[1] #initial abundance
+    pop_info$series_length = nrow(pop)
     
-   source('scripts/calculate_power_metric.R')
+   #source('scripts/calculate_power_metric.R')
      pop_info$min_time_for_power[j] = min_time_needed(pop$popvalue,0.05,0.8)
     print(paste(j,':',pop_info$min_time_for_power[j],sep=' '))
+    
+    
+    time_series_data = rbind(time_series_data,pop)
   }
+
+long_dat = time_series_data
 
 #IUCN calculation
 pop_info$IUCN = 3*pop_info$gen_len

### 2017-07-07 writing/manuscript_draft.Rmd
added section to reference figure in supp mat

@@
   \usepackage[round]{natbib} \bibpunct[; ]{(}{)}{,}{a}{}{;} \linenumbers
+
 output:
@@
 Observational studies and population censuses have become a cornerstone of modern ecological research and conservation biology [@Magurran2010].
-Longterm data is necessary to both understand population dynamics and to assess extinction risk of species.
-Even though many time series may now be considered "longterm", most are still short (i.e.
+long-term data is necessary to both understand population dynamics and to assess extinction risk of species.
+Even though many time series may now be considered "long-term", most are still short (i.e.
 less than X years).
-*I could have statistic on how long is considered "longterm" in papers*.
+*I could have statistic on how long is considered "long-term" in papers*.
 Time series are typically short for a variety of reasons.
@@
 @Rueda-Cediel2015 also used a modeling approach, but parameterized amodel specifically for a threatened snail, \emph{Tasmaphena lamproides}.
-They found that for this short-lived organim, 15 years was adequate to assess longterm trends in abundance.
+They found that for this short-lived organim, 15 years was adequate to assess long-term trends in abundance.
 However, these studies, and other past work, have typically focused either on theoretical aspects of monitoring design or focused on single species.
@@
 In this paper, we use both simulations and a database of empirical time series to determine the minimum number of years required to address various questions.
-More specifically, we estimate the minimum time series length required to asssess longterm changes in abundance via simple linear regression.
+More specifically, we estimate the minimum time series length required to asssess long-term changes in abundance via simple linear regression.
 We examine 878 population time series to estimate this minimum time series length required.
@@
 
-```{r,echo=F,fig.cap='(a) example of simulated time series for 40 time periods and (b) statistical power versus the simulated time series length.
+```{r,echo=F,eval=F,fig.cap='(a) example of simulated time series for 40 time periods and (b) statistical power versus the simulated time series length.
 The horizontal, dashed line is our desired statistical power of 0.8.
@@
 The database also includes information on generation length and census specifications.
-For each time series, we also calculated variables of interest like variance in population size, longterm trend in abundance (slope coefficient from simple linear regression), and temporal autocorrelation.
+For each time series, we also calculated variables of interest like variance in population size, long-term trend in abundance (slope coefficient from simple linear regression), and temporal autocorrelation.
 Lastly, we also calculated the minimum number of years required (10 years or 3 generations) under IUCN Red List Criterian A2 to classify a species as vulnerable [@IUCN2012].
@@
 ```
-For a subset of populations in the database ($n =$ `r nrow(LPI_pop_info)`), we were able to connect a database on biological characteristics (**cite database here**) like body size and generation length.
+For a subset of populations in the database ($n =$ `r nrow(LPI_pop_info)`), we were able to connect a database on biological characteristics [@Myhrvold2015] like body size and generation length.
 All `r nrow(LPI_pop_info)` populations are within the Aves class.
@@
 Therefore, we run a linear regression for each subsampled time series.
-Then, we determine the fraction of subsamples of a particular length that have estimated slope coefficients which are statistically different from zero given the longterm, or ``true", time series also has a significant slope.
+Then, we determine the fraction of subsamples of a particular length that have estimated slope coefficients which are statistically different from zero given the long-term, or ``true", time series also has a significant slope.
 This is a measure of statistical power (often denoted as $1- \beta$, where $\beta$ is the probability of a type II error).
@@
 We also examine how minimum time series calculations changed depending on the question of interest.
-In the supplementary material, we show how to make similar calculations to determine the minimum time series required to estimate overall population growth rate, the variability in population size, and future extinction risk.
+In the supplementary material, we show how to make similar calculations to determine the minimum time series required to estimate overall population growth rate (Fig.
+\ref{fig:growth_rate}), the variability in population size, and future extinction risk.
 Again, we took subsamples of time series data to determine $T_{min}$ for each of these questions.
@@
 We determined the minimum time series length required to address a particular question of interest.
-More specifically we asked, what is the minimum time series length required to determine, via linear regression, the longterm trend in population abundance?
+More specifically we asked, what is the minimum time series length required to determine, via linear regression, the long-term trend in population abundance?
 The minimum time series length required is that which had high enough statistical power.
@@
 
-#only look at pops with significant longterm declines or increases
-pop_info=pop_info[pop_info$trend_p_value<0.05,]
-long_dat=subset(long_dat,long_dat$ID %in% pop_info$ID)
-long_time_names=subset(long_time_names,long_time_names %in% pop_info$ID)
+#only look at pops with significant long-term declines or increases
+#pop_info=pop_info[pop_info$trend_p_value<0.05,]
+#long_dat=subset(long_dat,long_dat$ID %in% pop_info$ID)
+#long_time_names=subset(long_time_names,long_time_names %in% pop_info$ID)
 
@@
 We explored two different approaches to estimate the minimum time series length required to address a particular question of interest.
-We specifically asked, what is the minimum time series length required to determine longterm changes in abundance using linear regression?
+We specifically asked, what is the minimum time series length required to determine long-term changes in abundance using linear regression?
 The simulation-based approach has been suggested by others, especially in situations more complicated than that suited for classic power analysis [@Gerrodette1987, Johnson2015].
@@
 They found that only about 42% of populations in the Breeding Bird Survey met their criteria.
-@Seavy2007 asked whether statistical power was even a useful framework for assessing longterm population trends.
+@Seavy2007 asked whether statistical power was even a useful framework for assessing long-term population trends.
 They used 24 years of census data on Red-tailed Tropicbirds (\emph{Phaethon rubricauda}) in Hawaii and showed that to detect a 50% decline over 10 years almost always resulting in high statistical power (above 0.8).
@@
 Our work supports this idea as populations with strong temporal autocorrelation needed less years of sampling (cite figure here).
-@Morrison2008 also studied the optimal allocation of sampling effort in space versus time, but used emprical data from a longterm census of the collared pika (\emph{Ochotona collaris}) found in the Yukon.
-They estimated longterm growth rates among three subpopulations over a 10-year period.
+@Morrison2008 also studied the optimal allocation of sampling effort in space versus time, but used emprical data from a long-term census of the collared pika (\emph{Ochotona collaris}) found in the Yukon.
+They estimated long-term growth rates among three subpopulations over a 10-year period.
 They found that censuses less than 5 years may be misleading and that extrapolating from one population to another, even when nearby, may be difficult.
@@
 @Gerber1999 investigated the minimum time series required to estimate population growth of an endagered, but recovering, species.
-They used a longterm census to retroactively determine the minimum time series required to assess threat status.
+They used a long-term census to retroactively determine the minimum time series required to assess threat status.
 They found that only XX years were needed, XX years before the delisting decision was actually made.
@@
 We use a database of `r nrow(pop_info)` populations to determine the minimum time series length required.
-We show that to identify longterm changes in abundance, 10-15 years of continuos monitoring are required (Fig.
+We show that to identify long-term changes in abundance, 10-15 years of continuos monitoring are required (Fig.
 \ref{fig:min_time_dist}).
@@
 Our work implies that for many populations, census data less than 10-15 years are probably not reliable.
-This stresses the importance of longterm monitoring programs.
+This stresses the importance of long-term monitoring programs.
 <!--(where do IUCN recommendations come from?)--> From both a scientific and management perspective estimates of $T_{min}$ are important.
-If a time series is too short, we are likely to make incorrect conclusions about longterm trends.
+If a time series is too short, we are likely to make incorrect conclusions about long-term trends.
 In addition, a time series that is too long may be a poor use of already limited funds [@Gerber1999].
@@
 
-stuff here
+stuff here \ref{fig:empirical_approach_example}
+
+\clearpage
 
 # Supplementary material
+
+ \setcounter{figure}{0}
+ \renewcommand{\thefigure}{S\arabic{figure}}
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-2-1.pdf}
+\caption{(a) Population size of Bigeye tuna (\emph{Thunnus obesus}) over
+time. The line is the best fit line from linear regresssion. (b)
+Statistical power for different subsets of the time series in panel
+a.\label{fig:empirical_approach_example}}
+\end{figure}
+
+\begin{longtable}[]{@{}lrrrr@{}}
+\caption{Output of generalized linear model to examine predictors of the
+minimum time required for determing long-term population
+trends.\label{table:model_output}}\tabularnewline
+\toprule
+& Estimate & Std. Error & z value &
+Pr(\textgreater{}\textbar{}z\textbar{})\tabularnewline
+\midrule
+\endfirsthead
+\toprule
+& Estimate & Std. Error & z value &
+Pr(\textgreater{}\textbar{}z\textbar{})\tabularnewline
+\midrule
+\endhead
+(Intercept) & 3.7405513 & 0.0242742 & 154.09595 & 0\tabularnewline
+abs\_overall\_trend & -104.0533454 & 3.0941097 & -33.62949 &
+0\tabularnewline
+autocorrelation & -0.4899533 & 0.0431769 & -11.34757 & 0\tabularnewline
+variance & 18.7366110 & 0.6404458 & 29.25557 & 0\tabularnewline
+\bottomrule
+\end{longtable}
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
+\caption{Mimimum time required versus (a) generation length (years), (b)
+litter size (n), (c) adult body mass (grams), (d) maximum longevity
+(years), (e) egg mass (grams), and (f) incubation (days). The lines in
+each plot represent the best fit line from linear
+regression.\label{fig:biological_correlates}}
+\end{figure}

### 2017-07-07 writing/manuscript_draft.tex
added section to reference figure in supp mat

@@
 of modern ecological research and conservation biology (Magurran et al.
-2010). Longterm data is necessary to both understand population dynamics
-and to assess extinction risk of species. Even though many time series
-may now be considered ``longterm'', most are still short (i.e.~less than
-X years). \emph{I could have statistic on how long is considered
-``longterm'' in papers}. Time series are typically short for a variety
-of reasons. They are often coupled with an experiment, which may only
-last a couple of years. In addition, short funding cycles make it
-difficult to examine populations over longer periods of time
+2010). long-term data is necessary to both understand population
+dynamics and to assess extinction risk of species. Even though many time
+series may now be considered ``long-term'', most are still short
+(i.e.~less than X years). \emph{I could have statistic on how long is
+considered ``long-term'' in papers}. Time series are typically short for
+a variety of reasons. They are often coupled with an experiment, which
+may only last a couple of years. In addition, short funding cycles make
+it difficult to examine populations over longer periods of time
 (citations). There are, of course, notable exceptions including the Isle
@@
 for the experiment. This approach, however, is not straight-forward for
-a time series, or more complicated scenarios (P. C. D. Johnson et al.
+a time series, or more complicated scenarios (P. C. Johnson et al.
 2015), as data points are clearly non-independent.
@@
 tests (e.g.~t-test), and corresponding formulas for determinig sample
-size, simulation approaches can be used (Gerrodette 1987). Simple
-population models can be simulated with parameter values corresponding
-to a population of interest. Out of a large number of simulations, power
-can be calculated as the fraction of simulations that meet some
-criteria. The specific criteria depend on the question at hand. Then
-time series simulations of different lengths are run to determine the
-minimum time series length required to be confident that you will be
-able to answer your question (cite Bolker). In addition to using
-simulated data, empirical time series can also be used. We usually do
-not have multiple replicates of similar populations, but we can
-subsample an empirical time series (cite). Subsamples of different
-lengths can then be evaluated to see which fraction of subsamples meet
-some criteria, again a measure of statistical power. Similar to the
+size, simulation approaches can be used (Gerrodette 1987). Simple models
+can be simulated with parameter values corresponding to a population of
+interest. Out of a large number of simulations, power can be calculated
+as the fraction of simulations that meet some criteria. The specific
+criteria depend on the question at hand. Then, time series simulations
+of different lengths are run to determine the minimum time series length
+required to be confident that you will be able to answer your question
+(cite Bolker). In addition to using simulated data, empirical time
+series can also be used. We usually do not have multiple replicates of
+similar populations, but we can subsample an empirical time series
+(Gerber, DeMaster, and Kareiva 1999). Subsamples of different lengths
+can then be evaluated to see which fraction of subsamples meet some
+criteria, again a measure of statistical power. Similar to the
 simulation approach, we can use these measures of power to determine the
@@
 autocorrelation, and observer error. Rueda-Cediel et al. (2015) also
-used a modeling approach, but parameterized it for a threatened snail,
-\emph{Tasmaphena lamproides}. They found that for this short-lived
-organims, that 15 years was adequate to assess longterm trends in
-abundance. However, these studies, and other past work, have focused
-either on theoretical aspects of monitoring design or focused on single
-species.
+used a modeling approach, but parameterized amodel specifically for a
+threatened snail, \emph{Tasmaphena lamproides}. They found that for this
+short-lived organim, 15 years was adequate to assess long-term trends in
+abundance. However, these studies, and other past work, have typically
+focused either on theoretical aspects of monitoring design or focused on
+single species.
 
@@
 various questions. More specifically, we estimate the minimum time
-series length required to asssess longterm changes in abundance via
-simple linear regression.
+series length required to asssess long-term changes in abundance via
+simple linear regression. We examine 878 population time series to
+estimate this minimum time series length required.
 
@@
 any experimental design too complicated for simple power analyses (P. C.
-D. Johnson et al. 2015) (and cite Bolker). We only briefly discuss this
+Johnson et al. 2015) (and cite Bolker). We only briefly discuss this
 approach as it has been desribed elsewhere (citations). Essentially, we
@@
 is simply the fraction of simulations with significant trends in
-abundance from simple linear regression. Then, the minimum time required
-is the minimum time required to obtain statistical power above some
-pre-determined threshold (historically at 0.8).
+abundance. A significant trend in abundance is determined by estimating
+the significance of the slope parameter from simple linear regression.
+Then, the minimum time required (\(T_{min}\)) is the number of years
+required to obtain statistical power above some pre-determined threshold
+(historically at 0.8). Statistical power of 0.8 would indicate that, if
+there was a true trend in abundance, we would have a 0.8 probability of
+detecting the trend.
 
@@
 (\(T_{min}\)).
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-1-1.pdf}
-\caption{(a) example of simulated time series for 40 time periods and
-(b) statistical power versus the simulated time series length. The
-horizontal, dashed line is our desired statistical power of 0.8. The
-vertical, dashed line is the minimum time required to achieve our
-desired statistical power. (c) Minimum time required for different
-values of the trend strength (\(r\)) and (d) minimum time required for
-different levels of population variability
-(\(\sigma\)).\label{fig:theoretical_approach}}
-\end{figure}
 
@@
 specifications. For each time series, we also calculated variables of
-interest like variance in population size, longterm trend in abundance
+interest like variance in population size, long-term trend in abundance
 (slope coefficient from simple linear regression), and temporal
@@
 For a subset of populations in the database (\(n =\) 547), we were able
-to connect a database on biological characteristics (\textbf{cite
-database here}) like body size and generation length. All 547
-populations are within the Aves class. In the supplementary material, we
-explore how the minimum time required does not strongly correlate with
-any biological characteristics.
+to connect a database on biological characteristics (Myhrvold et al.
+2015) like body size and generation length. All 547 populations are
+within the Aves class. In the supplementary material, we explore how the
+minimum time required does not strongly correlate with any biological
+characteristics.
 
@@
 a particular length that have estimated slope coefficients which are
-statistically different from zero given the longterm, or ``true``, time
+statistically different from zero given the long-term, or ``true``, time
 series also has a significant slope. This is a measure of statistical
@@
 required to guarantee we have above a certain threshold of statistical
-power, which we will assume is 0.8 by convention (citation). A power of
-0.8 would imply that given a trend exists, there is a 0.8 probability of
-detecting the trend. Complementary, there would be a 0.2 probability of
-a false negative.
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-4-1.pdf}
-\caption{(a) Population size of Harp Seals over time. The line is the
-best fit line from linear regresssion. (b) Statistical power for
-different subsets of the time series in panel
-a.\label{fig:empirical_approach_example}}
-\end{figure}
+power, which we will assume is 0.8 by convention (citation).
 
@@
 to make similar calculations to determine the minimum time series
-required to estimate overall population growth rate, the variability in
-population size, and future extinction risk. Again, we took subsamples
-of time series data to determine \(T_{min}\) for each of these
-questions.
+required to estimate overall population growth rate (Fig.
+\ref{fig:growth_rate}), the variability in population size, and future
+extinction risk. Again, we took subsamples of time series data to
+determine \(T_{min}\) for each of these questions.
 
@@
 minimum time series length required to determine, via linear regression,
-the longterm trend in population abundance? The minimum time series
+the long-term trend in population abundance? The minimum time series
 length required is that which had high enough statistical power. We
@@
 \centering
-\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
+\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-4-1.pdf}
 \caption{Distribution of the minimum time required in order to detect a
@@
 \centering
-\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-6-1.pdf}
+\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
 \caption{Minimum time required to estimate change in abundance
@@
 specifically asked, what is the minimum time series length required to
-determine longterm changes in abundance using linear regression? The
+determine long-term changes in abundance using linear regression? The
 simulation-based approach has been suggested by others, especially in
@@
 criteria. Seavy and Reynolds (2007) asked whether statistical power was
-even a useful framework for assessing longterm population trends. They
+even a useful framework for assessing long-term population trends. They
 used 24 years of census data on Red-tailed Tropicbirds
@@
 (2008) also studied the optimal allocation of sampling effort in space
-versus time, but used emprical data from a longterm census of the
+versus time, but used emprical data from a long-term census of the
 collared pika (\emph{Ochotona collaris}) found in the Yukon. They
-estimated longterm growth rates among three subpopulations over a
+estimated long-term growth rates among three subpopulations over a
 10-year period. They found that censuses less than 5 years may be
@@
 We use a database of 878 populations to determine the minimum time
-series length required. We show that to identify longterm changes in
+series length required. We show that to identify long-term changes in
 abundance, 10-15 years of continuos monitoring are required (Fig.
@@
 census data less than 10-15 years are probably not reliable. This
-stresses the importance of longterm monitoring programs. From both a
+stresses the importance of long-term monitoring programs. From both a
 scientific and management perspective estimates of \(T_{min}\) are
 important. If a time series is too short, we are likely to make
-incorrect conclusions about longterm trends. In addition, a time series
+incorrect conclusions about long-term trends. In addition, a time series
 that is too long may be a poor use of already limited funds (Gerber,
@@
 \hypertarget{ref-Johnson2015}{}
-Johnson, Paul C D, Sarah J E Barry, Heather M. Ferguson, and Pie Müller.
-2015. ``Power analysis for generalized linear mixed models in ecology
-and evolution.'' \emph{Methods in Ecology and Evolution} 6 (2): 133--42.
+Johnson, Paul C.D., Sarah J.E. Barry, Heather M. Ferguson, and Pie
+Müller. 2015. ``Power analysis for generalized linear mixed models in
+ecology and evolution.'' \emph{Methods in Ecology and Evolution} 6 (2):
+133--42.
 doi:\href{https://doi.org/10.1111/2041-210X.12306}{10.1111/2041-210X.12306}.
@@
 
+\hypertarget{ref-Myhrvold2015}{}
+Myhrvold, Nathan P., Elita Baldridge, Benjamin Chan, Dhileep Sivam,
+Daniel L. Freeman, and S.K. Morgan Ernest. 2015. ``An amniote
+life-history database to perform comparative analyses with birds,
+mammals, and reptiles.'' \emph{Ecology} 96 (11): 3109.
+
 \hypertarget{ref-GPDD2010}{}
@@
 
-stuff here
+stuff here \ref{fig:empirical_approach_example}
+
+\clearpage
 
@@
 
+\setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-2-1.pdf}
+\caption{(a) Population size of Bigeye tuna (\emph{Thunnus obesus}) over
+time. The line is the best fit line from linear regresssion. (b)
+Statistical power for different subsets of the time series in panel
+a.\label{fig:empirical_approach_example}}
+\end{figure}
+
+\begin{longtable}[]{@{}lrrrr@{}}
+\caption{Output of generalized linear model to examine predictors of the
+minimum time required for determing long-term population
+trends.\label{table:model_output}}\tabularnewline
+\toprule
+& Estimate & Std. Error & z value &
+Pr(\textgreater{}\textbar{}z\textbar{})\tabularnewline
+\midrule
+\endfirsthead
+\toprule
+& Estimate & Std. Error & z value &
+Pr(\textgreater{}\textbar{}z\textbar{})\tabularnewline
+\midrule
+\endhead
+(Intercept) & 3.7405513 & 0.0242742 & 154.09595 & 0\tabularnewline
+abs\_overall\_trend & -104.0533454 & 3.0941097 & -33.62949 &
+0\tabularnewline
+autocorrelation & -0.4899533 & 0.0431769 & -11.34757 & 0\tabularnewline
+variance & 18.7366110 & 0.6404458 & 29.25557 & 0\tabularnewline
+\bottomrule
+\end{longtable}
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
+\caption{Mimimum time required versus (a) generation length (years), (b)
+litter size (n), (c) adult body mass (grams), (d) maximum longevity
+(years), (e) egg mass (grams), and (f) incubation (days). The lines in
+each plot represent the best fit line from linear
+regression.\label{fig:biological_correlates}}
+\end{figure}
+
 \end{document}

### 2017-07-07 scripts/LemonShark_PopulationModel.R
added Lemon shark pop model script from White et al 2014

@@ -1,0 +1,100 @@
+
+#This script runs a population model for lemon sharks in a nursery lagoon in Bimini, Bahamas. 
+#This model is desribed in White et al 2014 "Modeling lemon shark population dynamics" https://peerj.com/preprints/364/
+#Currently the script using a Poisson process to calculate litter sizes but this can be modified to use a litter size distribution from a genetic study (see below and paper)
+
+#this code has not been optimized for speed
+
+#Created by Easton R. White 
+#Created: 17-Apr-2012
+#Last edited: 5-Jul-2017
+
+power = function(max.time){
+
+  max.time = max.time
+#Intial condition for Lemon shark population size
+IC= c(30,30,10,15,15,5,5,3,3,3,3,3,2,2,2,2,2,2,1,1,1,1,2,2,2,2)
+
+#set number of runs for model
+no.runs= 1000
+slope_p_value = vector("numeric",length=no.runs)
+estimated_slope = vector("numeric",length=no.runs)
+
+#keep track of variance and mean of Juvenile population size over a series of runs with the model
+#J_var= matrix(0,nrow=1,ncol=no.runs)
+#J_mean= matrix(0,nrow=1,ncol=no.runs)
+
+
+#Probablities for different litter sizes (see paper for details) based on genetic data
+#birthprob=c(0.02272727,0.14772727,0.113636,0.143939,0.0757575,0.1022727,0.06060606,0.0492424,0.09848484,0.0643939,0.04166666,0.034090909,0.02272727,0.00378787,0.0151515,0,0,0.0037878)
+
+
+for (g in 1:(no.runs)){
+
+
+	#set up matrix (Lemons) to track number of individuals in each age class over time and matrix (Stable_age) to track the age distribution through time
+	Lemons = matrix(IC,nrow= 26,ncol=max.time)
+	Stable_age =matrix(0,nrow=26,ncol=max.time)
+	
+	#initialize matrices
+	Lemons[1:26,1] = IC
+#	Stable_age[1:26,1]=IC/sum(IC)
+	
+	#keep track of Juveniles, Subadults, and Adults
+	J =  sum(Lemons[1:3,1])    #Juveniles and newborns in lagoon
[…]

### 2017-07-07 writing/supp_mat_draft.Rmd
cleaned up supp mat

--- 2017-06-14 writing/supp_mat_draft.Rmd
@@
     number_sections: true
-    citation_package: natbib
 fontsize: 12pt
@@
     \renewcommand{\thefigure}{S\arabic{figure}}
+    \usepackage[round]{natbib} \bibpunct[; ]{(}{)}{,}{a}{}{;}
 ---
@@
 
-Authors: Easton R. White$^{1*}$
+Easton R. White$^{*}$
 \vspace{3 mm}
 
-Address: \\ \emph{$^1$ Center for Population Biology, University of California-Davis, Davis, California 95616 USA}
+Address: \\ \emph{Center for Population Biology, University of California-Davis, Davis, California 95616 USA}
 
@@
 
-This supplementary material includes:
+\tableofcontents
 
-1. Detailed example of subsampling and power calculations
-2. Additional figures
-3. Figures and results from other questions of interest
-4. Additional references
+\vspace{1cm}
 
-Code for all the figures and tables can be found at [](https://github.com/erwhite1).
+Code for all the figures and tables can be found at (https://github.com/erwhite1).
 
@@
 
-## Detailed example of subsampling and power calculations
+\clearpage
 
-Here, we illustrate in detail how we performed the subsampling and power calculations for a specific population.
-As an example, we examine a 35-year time series of ().
-Simple linear regression indicates a significant decline for this population with a blank and blank.
-We assume that this significant decline over 35 years is in fact the ``true trend".
-In statistical jargon, the 35-year trend is an effect that is actually present.
-We can then use this as a benchmark to see if subsamples of the time series show a similar result.
+# Detailed example of subsampling and power calculations
+
+```{r,echo=F}
+ setwd("~/Desktop/Research/time-series-project")
+ load("cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
+    pop = subset(long_dat,long_dat$ID=='7558')
+    model_example = lm(pop$popvalue~c(1:nrow(pop)))
+```
+
+Here, we illustrate how we performed the subsampling and power calculations for a specific population.
+We provide more detail here than in the main manuscript.
+As an example, we examine a 35-year time series of Bigeye tuna (\emph{Thunnus obesus}), one species in the Global Population Dynamics Database [@GPDD2010].
+<!--Simple linear regression indicates a significant increase for this population with an estimated intercept of `r as.character(round(as.numeric(coef(model_example)[1])))` and a slope coefficient of `r as.numeric(coef(model_example)[2])`.--> We assume that this significant increase over 35 years is in fact the ``true trend".
+In statistical jargon, the 35-year trend is an effect that is actually present, we can reject the null hypothesis of no trend.
+We can then use this as a benchmark to see if subsamples of the time series also indicate a significant increase.
 
@@
 For each subsample, we conduct linear regression and extract model coefficients and p-values.
-Then, the fraction of subsamples within a set that show significant trends (signfiiant slope coefficient) is the statistical power.
+Then, the fraction of subsamples within a set that show significant trends (significant slope coefficient) is the statistical power.
 It is important to note that we only consider subsamples to be significant if they are significant in the same direction as the complete 35-year time series.
+In other words, we are conducting a one-tailed test.
 
-We can then plot statistical power as a function of time series length. As expected, we can see that power increases with the more years that are sampled.
+We can then plot statistical power as a function of time series length (Fig.
+\ref{fig:empirical_approach_example}b).
+As expected, we can see that power increases with the more years that are sampled.
+Then, we determine an appropriate level of statistical power that we find acceptable.
+Traditionally, this has been at 0.8, however, this is purely historical.
+Statistical power of 0.8 implies if a true trend is present, or a real change in abundance, then we will detect this trend 0.8 fraction of the time.
 
-Then, we determine an appropriate level of statistical power that we find acceptable. Traditionally, this has been at 0.8, however, this is purely historical. Statisical power of 0.8 implies...
+<!-- example of how subsampling and empirical approach works -->
+```{r, echo=F,message=F,warning=F,fig.cap='(a) Population size of Bigeye tuna (\\emph{Thunnus obesus}) over time.
+The line is the best fit line from linear regresssion.
+(b) Statistical power for different subsets of the time series in panel a.\\label{fig:empirical_approach_example}'}
+# Example of subsampling empirical data
+ setwd("~/Desktop/Research/time-series-project")
+ load("cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
+  par(mfrow=c(1,2),oma=c(4,4,0.5,0.5))
+    pop = subset(long_dat,long_dat$ID=='7558') #13192 for Harp seals, 7558 for Thunnus obesus
+    plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='',xlab='time (years)',pch=16,las=1)
+    abline(lm(pop$popvalue~pop$year),lwd=2,col='red')
+    mtext(text = '(a)',side = 3,line = -1.2,adj=0.95,font = 2,cex=1.2)
+    mtext(text='population size', side=2,line=4,cex=1.2)
+# could insert all possible regressions of length n
+source('scripts/calculate_power_metric.R')
+    power_example=calculate_power_for_subsamples(pop$popvalue,0.05)
+    plot(power_example,cex.axis=1.2,cex.lab=1.2,ylab='statistical power',xlab='years sampled',las=1,pch=16)
+    mtext(text = '(b)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
+# could also insert standard error bars
+    abline(h=0.8,col='red',lwd=2,lty=2)
+    abline(v=head(which(power_example>0.8),1),col='red',lwd=2,lty=2)
 
-With statistical power of 0.8, we then determine the minimum time series length ($t_min$) required to achieve that level of statistical power.
-Here, $t_min$ is the first point as which all points to the right are above 0.8.
-
-```{r, echo=F}
-# show an example of subsampling and power calculation routines
 ```
 
-## Additional results from the main manuscript
+With statistical power of 0.8, we then determine the minimum time series length ($T_{min}$) required to achieve that level of statistical power.
+Here, $T_{min}$ is the first point in Fig.
+\ref{fig:empirical_approach_example}b where following points are also above 0.8.
+In this example, $T_{min}$ is `r head(which(power_example>0.8),1)`.
+Therefore, a minimum of 22 years of continuous monitoring are required to have 0.8 statistical power in determining long-term changes in abundance.
 
+# Additional results from the main manuscript
+
+## Statistical model of minimum time required and time series characteristics
+
+In the main text, we explained how the minimum time required strongly correlated with the trend strength, temporal autocorrelation, and variance in population size.
+Here, we use a generalized linear model framework with a Poisson error structure to determine drivers of the minimum time required.
+In figure, \ref{fig:poisson_model} we show a set of residual plots for the regression.
+We then show the coefficient estimates and levels of significance in table \ref{table:model_output}.
+
+```{r poisson_model, echo=F,include=T,fig.cap='Output of generalized linear model with a Poisson error structure for predicting the minimum time required with explanatory variables of the absolute value of the slope coefficient (or trend strength), temporal autocorrelation, and variability in population size.\\label{fig:poisson_model}'}
+setwd("~/Desktop/Research/time-series-project")
+load("cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
+
+# GLM with Poisson error structure (include all terms - it is best not to include abs_overall_trend with variance as they are strongly corrrelated to one another)
+pop_info$abs_overall_trend = abs(pop_info$overall_trend)
+poisson_model <-glm(min_time_for_power~ abs_overall_trend + autocorrelation + variance,family = 'poisson',data=pop_info)
+
+  par(mfrow=c(2,2),mar=c(5,5,1,2),oma=c(0,0,0.5,0))
+    plot(poisson_model,cex.lab=1.2,las=1,cex.axis=1.2,main=' ',pch=16, col=rgb(0.5,0.5,0.5,0.4))
+#R^2 calculation (pg. 218 Zuur et al. 2009 mixed effects book)
+R2_fullmodel <- 100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance)
+
+poisson_model_no_variance <-glm(min_time_for_power~ abs_overall_trend + autocorrelation,family = 'poisson',data=pop_info)
+R2_no_variance <- 100*(poisson_model_no_variance$null.deviance - poisson_model_no_variance$deviance)/(poisson_model_no_variance$null.deviance)
+
+poisson_model_no_trend <-glm(min_time_for_power~ autocorrelation + variance,family = 'poisson',data=pop_info)
+R2_no_trend <- 100*(poisson_model_no_trend$null.deviance - poisson_model_no_trend$deviance)/(poisson_model_no_trend$null.deviance)
+
+```
+
+```{r,echo=F,message=F,warning=F}
+
+require(knitr)
+knitr::kable(summary(poisson_model)$coef,caption = 'Output of generalized linear model to examine predictors of the minimum time required for determing long-term population trends.\\label{table:model_output}')
+
+```
+
+Our model with trend strength, autocorrelation, variance, and generation length accounted for `r round(R2_fullmodel,2)`% of the variation in the minimum time required (Table \ref{table:model_output}).
+However, we also found trend strength and variance to be strongly correlated with one another.
+Therefore, we ran two additional models with either trend strength or variance, but not both together.
+This resulted in lower explained deviance (analogue to $R^2$) of `r round(R2_no_variance,2)`% and `r round(R2_no_trend,2)`%, respectively.
+
+\pagebreak
+
+## Minimum time required and biological correlates
 ```{r, echo=F}
@@
 In the main manuscript, we examined characteristics of the time series itself, like variability, autocorrelation, and the trend in abundance over time.
-Here, we combined our time series database with a database on life history characteristics of amniotes from (cite other database here).
+Here, we combined our time series database with a database on life history characteristics of amniotes from @Myhrvold2015.
 There was life history information available for `r nrow(LPI_pop_info)` populations representing `r length(table(LPI_pop_info$Binomial))` different species, all of which were in the Aves class.
@@
 We then correlated minimum time required for each population with its corresponding life history characteristics.
-In figure \ref{fig:biological_correlates} we examined miminmum time required versus generation length (years), litter size (n), adult body mass (grams), maximum longevity (years), egg mass (grams), and incubation (days).
+In figure \ref{fig:biological_correlates} we examined minimum time required versus generation length (years), litter size (n), adult body mass (grams), maximum longevity (years), egg mass (grams), and incubation (days).
 None of these variables had much explanatory power in accounting for the variance in the minimum time required.
@@
 The lines in each plot represent the best fit line from linear regression.\\label{fig:biological_correlates}',fig.height=6}
-
+setwd("~/Desktop/Research/time-series-project")
+load('cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
 par(mfrow=c(3,2),oma=c(5,4,1,1),mar=c(5,3,0,0))
@@
 x_axis_name_list = c('generation length (years)','litter size (n)','adult body mass (grams)','maximum longevity (years)','egg mass (grams)','incubation (days)')
-for (x_axis_name in names(LPI_pop_info)[20:25]){
+for (x_axis_name in names(LPI_pop_info)[21:26]){
   plot(LPI_pop_info[[x_axis_name]],LPI_pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
@@
 
-In the main text, we explained how the minimum time required strongly correlated with the trend strength, temporal autocorrelation, and variance in population size.
-Here we use a generalized linear model framework with a Poisson error structure to determine drivers of the minimum time required.
-In figure, \ref{fig:poisson_model} we show a set of residual plots for the regression.
-We then show the coefficient estimates and levels of significance in table \ref{table:model_output}.
+## Minimum time series and biological class
 
-```{r poisson_model, echo=F,include=T,fig.cap='Output of poisson regression...\\label{fig:poisson_model}'}
+When we examined differences in the minimum time required there were similar patterns with biological classes.
+For instance, more time is required for species within the Actinopterygii class compared to other species (Fig.
+\ref{fig:class}a).
+These differences between biological classes can be explained by differences in population variability, with species in Actinopterygii have larger variability in population size from year-to-year.
+
+```{r,echo=F,fig.cap='(a) Minimum time required to estimate change in abundance for species class, (b) long-term trend (estimated slope coefficient) by species class, (c) interannual variability in population size by species class, and (d) temporal autocorrelation by species class.\\label{fig:class}',fig.height=6}
+
+# Pull in data
 setwd("~/Desktop/Research/time-series-project")
-load("cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
+load("~/Desktop/Research/time-series-project/cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
+# Biological correlates plots
+par(mfrow=c(2,2),oma=c(8,1,1.5,0.5),mar=c(0,4,0,1))
 
-# GLM with Poisson error structure (include all terms - it is best not to include abs_overall_trend with variance as they are strongly corrrelated to one another)
-pop_info$abs_overall_trend = abs(pop_info$overall_trend)
-poisson_model <-glm(min_time_for_power~ abs_overall_trend + autocorrelation + variance + gen_len,family = 'poisson',data=pop_info)
+boxplot(pop_info$min_time_for_power~pop_info$class, lwd = 1, ylab = 'minimum time required',col='white',cex.lab=1.4,cex.axis=1.2,las=2,xaxt="n")
+mtext(text = '(a)',side = 3,line = -1.2,adj=0.95,font = 2,cex=1.2)
+stripchart(pop_info$min_time_for_power~pop_info$class, vertical = TRUE,method = "jitter", add = TRUE, pch = 20, col = rgb(0.5,0.5,0.5,0.5))
 
-  par(mfrow=c(2,2),mar=c(5,5,1,2),oma=c(0,0,0.5,0))
-    plot(poisson_model,cex.lab=1.2,las=1,cex.axis=1.2,main=' ',pch=16, col=rgb(0.5,0.5,0.5,0.4))
-#R^2 calculation (pg. 218 Zuur et al. 2009 mixed effects book)
-R2_fullmodel <- 100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance)
+boxplot(abs(pop_info$overall_trend)~pop_info$class, lwd = 1, ylab = ' ',col='white',cex.lab=1.4,cex.axis=1.2,las=2,xaxt="n")
+stripchart(abs(pop_info$overall_trend)~pop_info$class, vertical = TRUE,method = "jitter", add = TRUE, pch = 20, col = rgb(0.5,0.5,0.5,0.5))
+  mtext('| slope coefficient |',2,line=3.5,cex=1.2)
+  mtext(text = '(b)',side = 3,line = -1.2,adj=0.95,font = 2,cex=1.2)
 
-poisson_model_no_variance <-glm(min_time_for_power~ abs_overall_trend + autocorrelation + gen_len,family = 'poisson',data=pop_info)
-R2_no_variance <- 100*(poisson_model_no_variance$null.deviance - poisson_model_no_variance$deviance)/(poisson_model_no_variance$null.deviance)
+boxplot(pop_info$variance~pop_info$class, lwd = 1, ylab = 'population variance',col='white',cex.lab=1.4,cex.axis=1.2,las=2)
+stripchart(pop_info$variance~pop_info$class, vertical = TRUE,method = "jitter", add = TRUE, pch = 20, col = rgb(0.5,0.5,0.5,0.5))
+mtext(text = '(c)',side = 3,line = -1.2,adj=0.95,font = 2,cex=1.2)
 
-poisson_model_no_trend <-glm(min_time_for_power~ autocorrelation + variance + gen_len,family = 'poisson',data=pop_info)
-R2_no_trend <- 100*(poisson_model_no_trend$null.deviance - poisson_model_no_trend$deviance)/(poisson_model_no_trend$null.deviance)
+boxplot(pop_info$autocorrelation~pop_info$class, lwd = 1, ylab = 'autocorrelation',col='white',cex.lab=1.4,cex.axis=1.2,las=2,ylim=c(-0.2,1.2))
+stripchart(pop_info$autocorrelation~pop_info$class, vertical = TRUE,method = "jitter", add = TRUE, pch = 20, col = rgb(0.5,0.5,0.5,0.5))
+mtext(text = '(d)',side = 3,line = -1.2,adj=0.95,font = 2,cex=1.2)
+```
+
+\pagebreak
+
+## Sensitivity analysis of significance level and power
+
+Estimates of $T_{min}$ depend strongly on the values used for the significance level ($\alpha$) and the probability of type II error ($\beta$), both of which are set by the practitioner.
+Here we explore how estimates of $T_{min}$ are affected by changes in each of these parameters.
+We see that the minimum time required increases with increases in statistical power or decreases with increases in the threshold for statistical significance (Fig \ref{fig:min_time_vs_alpha_beta}).
+
+```{r,echo=F,message=F,error=F,fig.cap='Minimum time required to assess long-term trends in abundance for values of statistical significance ($\\alpha$) and power ($1-\\beta$).\\label{fig:min_time_vs_alpha_beta}'}
+# Bring in theoretical code
+# Build plot of $T_min$ vs alpha and another versus beta
+# Explain results after building figures
+setwd("~/Desktop/Research/time-series-project")
+source('scripts/calculate_slope.R') 
+source('scripts/theoretical-models/simulate_pop.R')
+# calculate_power = function(r,phi,sigma,trials,max.time,alpha){
+#   trials   = trials
+#   max.time = max.time
+#   r        = r
+#   phi      = phi
+#   sigma    = sigma
+#   
+#   y=replicate(n=trials,simulate_pop(r,phi,sigma,max.time)) 
+#   
+#   power=sum((apply(y,2,calculate_p_value)<alpha) & sign(apply(y,2,calculate_slope))==sign(r))/trials 
+#   
+#   return(power)
+# }
+# 
+# 
+# min_time_needed= function(alpha,power){
+#   years_to_sample=seq(2,30,by=1)
+#   y=lapply(X=years_to_sample,FUN = calculate_power,phi =0.5,r=0.5,sigma = 1,trials=100,alpha=alpha)
+#   min_time_needed = years_to_sample[tail(which(y<power),1)] + 1
+#   return(min_time_needed)
+# }
+# 
+# alpha_values=seq(0.001,0.2,by=0.01)
+# power_values=seq(0.65,0.95,by=0.01)
+# parameter_combinations = expand.grid(alpha_values,power_values)
+# names(parameter_combinations) = c('alpha','power')
+# parameter_combinations$min_time_required = NA
+# 
+# for (i in 1:nrow(parameter_combinations)){
+# 
+#   parameter_combinations$min_time_required[i] =  min_time_needed(parameter_combinations[i,1],parameter_combinations[i,2])
+#  
+#   #print(i)
+# }
+
+#write.csv(parameter_combinations,file = 'analysis-outputs/min_time_vs_alpha_and_power.csv',quote = F,row.names = F)
+
+# Pull in data to prevent R from running for a long time to create data from scratch
+parameter_combinations = read.csv('analysis-outputs/min_time_vs_alpha_and_power.csv',header=T)
+
+require(ggplot2)
+ggplot(parameter_combinations, aes(x = alpha, y = power, z = min_time_required, fill=min_time_required)) +
+  geom_tile() + 
+  labs(title = " ", x = expression(alpha), y = expression(paste('power ',(1-beta))), fill = expression(T[min])) +
+  scale_fill_distiller(palette="Spectral", na.value="white") +
+  theme_bw()
+```
+
+# Minimum time calculations testing exponential growth
+
+In the main text we evaluated the minimum time required to determine long-term trends in abundance via linear regression.
+This process examined linear trends in abundance over time.
+Here we examine the minimum time required to estimate long-term trends that are either exponential growth or decay.
+We use the same methods as described in the previous sections, but we take the $log$ of population density, or abundance.
+
+```{r,echo=F,message=F,error=F,fig.cap='Distribution of the minimum time required in order to detect a significant trend (at the 0.05 level) in log(abundance) given power of 0.8.\\label{fig:min_time_dist_log_pop}'}
+setwd("~/Desktop/Research/time-series-project")
+# load('cleaned-data/cleaned_timeseries_database.Rdata')
+# 
+# source('scripts/calculate_slope.R')
+# source('scripts/calculate_power_metric.R')
+# 
+# for (j in 1:nrow(pop_info)){
+#     pop= subset(long_dat,long_dat$ID==pop_info$ID[j])
+#     pop = pop[(nrow(pop)-34):nrow(pop),]
+#     pop$popvalue=as.numeric(pop$popvalue)
+#     pop$popvalue=pop$popvalue/max(pop$popvalue) #should I standaridize this in a better way
+#     ## Take the log of population size and add a small number
+#     pop$popvalue = log(pop$popvalue + 0.001)
+#     
+#     pop_info$year[j] = pop$year[1]
+#     pop_info$variance[j]= var(pop$popvalue)
+#     pop_info$overall_trend[j] = calculate_slope(pop$popvalue)
+#     pop_info$autocorrelation[j] = acf(pop$popvalue,plot = FALSE)$acf[2]
+#     pop_info$trend_p_value[j]=calculate_p_value(pop$popvalue)
+#     pop_info$popvalue = pop$popvalue[1] #initial abundance
+#     pop_info$series_length = nrow(pop)
+#     
+# 
+#      pop_info$min_time_for_power[j] = min_time_needed(pop$popvalue,0.05,0.8)
+#     print(paste(j,':',pop_info$min_time_for_power[j],sep=' '))
+#     
+# }
+# 
+# pop_info=pop_info[pop_info$trend_p_value<0.05,]
+# long_dat=subset(long_dat,long_dat$ID %in% pop_info$ID)
+# long_time_names=subset(long_time_names,long_time_names %in% pop_info$ID)
+# 
+# save(pop_info,long_dat,long_time_names,file = 'cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression_for_log_pop.Rdata')
+
+# LOAD in previously run script from this code chuck and make histogram plot
+load('cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression_for_log_pop.Rdata')
+
+plot(density(pop_info$min_time_for_power),las=1,ylab='',xlab='',cex.axis=1.2,main='')
+mtext(text = 'Frequency',2,line=3,cex=1.2)
+mtext(text = 'Minimum time required',1,line=3,cex=1.2)
 
@@
 
-```{r,echo=F,message=F,warning=F}
+We see that the distribution of $T_{min}$ is almost identical to that in the main manuscript.
+This is perhaps not surprising as most time series that would significantly increase or decrease linearly would probably also significantly increase or decrease at an exponential rate.
+Further, the calculations here and in the main manuscript both use linear regression, and both estimate the same number of parameters.
 
-require(knitr)
-knitr::kable(summary(poisson_model)$coef,caption = 'Stuff here\\label{table:model_output}')
+\pagebreak
+
+# Simulations with more complicated population model
+
+In the main text, we showed how a simple population model could be simulated repeatedly to estimate the power obtained with time series of increasing length.
+The model in the main text simulated linear population growth with only a slope coefficient, y-intercept, and noise parameter required.
+This model is purely phenomenological and does not include any species life history.
+Here, we use the same routine as the main text, but simulate from a more biologically-realistic population model.
+We simulate the model described in @White2014.
+The model is a stochastic, age-structured population model that includes density-dependence for lemon sharks in Bimini, Bahamas.
+
+```{r, echo=F,fig.cap='Statistical power for different length of time series simulations for a lemon shark population in Bimini, Bahamas.\\label{fig:shark_example}'}
+
+setwd("~/Desktop/Research/time-series-project")
+#source('scripts/LemonShark_PopulationModel.R')
+years_to_sample=seq(5,50,by=1)
+#shark_power=sapply(years_to_sample,power)
+
+load('analysis-outputs/example_power_for_shark_model.Rdata')
+par(mfrow=c(1,1),oma=c(3,3,0.5,0.5),mar=c(2,2,0,0))
+   plot(years_to_sample,shark_power,type='l',lwd=2,las=1,ylab='',xlab='',cex.axis=1.2)
+      abline(v=years_to_sample[tail(which(shark_power<0.8),1)+1],lty=2,lwd=2,col='red')
+      abline(h=0.8,lty=2,lwd=2,col='red')
+      mtext(text = 'time sampled' ,side = 1,cex = 1.2,line=3)
+      mtext(text = 'statistical power' ,side = 2,cex = 1.2,line=3.5)
+      #mtext(text = '(b)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
+
+#save(shark_power,file='analysis-outputs/example_power_for_shark_model.Rdata')
 
@@
 
-Our model with trend strength, autocorrelation, variance, and generation length accounted for `r round(R2_fullmodel,2)`% of the variation in the minimum time required (Table \ref{table:model_output}).
-However, we also found trend strength and variance to be strongly correlated with one another.
-Therefore, we ran two additional models with either trend strength or variance, but not both together.
-This resulted in lower explained deviance (analogue to $R^2$) of `r round(R2_no_variance,2)`% and `r round(R2_no_trend,2)`%, respectively.
+Here, we parameterize the model for a situation where adult mortality rate is high enough to cause a population decline.
+In the same way we simulated models in the main text, we simulate this more biologically-realistic model for different lengths of time.
+For each length of time, we calculate the statistical power.
+Similar to results with the simpler model, statistical power generally increases with longer sampling time (Fig.
+\ref{fig:shark_example}).
+In this example, the minimum time required ($T_{min}$) to obtain at least 0.8 statistical power, given a significance level of 0.05, is `r years_to_sample[tail(which(shark_power<0.8),1)+1]` years.
 
-\pagebreak
-
-## Results from other questions of interest
-
-### Simulations with more complicated population model
-
-In the main text, we showed how a simple population model could be simulated repeatidly to estimate the power obtained with time series of increasing length.
+<!--
+## Optional section looking at Gompertz model
+In the main text, we showed how a simple population model could be simulated repeatedly to estimate the power obtained with time series of increasing length.
 The model in the main text simulated linear population growth with only a slope coefficient, y-intercept, and noise parameter required.
-This model is purely phenemilogical and does not include any species life history.
-Here, we use the same routine as the main text, but simulate from a more biologically-realistic population model, the (MODEL NAME HERE):
+This model is purely phenomenological and does not include any species life history.
+Here, we use the same routine as the main text, but simulate from a more biologically-realistic population model, the stochastic Gompertz population model:
 
   \begin{equation}
-    y = x
+    N_{t+1} = N_t e^{a - b \mbox{ log} N_t + \epsilon_t}
   \end{equation}
 
-We simulate this model repetively for different lengths of time series.
-Statistical power is the fraction of time series of a set length that showed signicant trends in abundance, given a true trend in abundance.
+Here, $N_t$ is the population density in year $t$, $a$ is an intercept term, and $b$ is a parameter measuring the strength of density-dependence.
+Lastly, $\epsilon_t$ is process error that is assumed to be normally distributed.
+@Knape2012 used this model to investigate the frequency of density-dependence in time series of populations.
+
+We simulate this model repetitively for different lengths of time series.
+Statistical power is the fraction of time series of a set length that showed significant trends in abundance, given a true trend in abundance.
 *How do I determine significance here?
-Through linear regression again, or do I look at non-zero estimation of the parameter values?*
+Through linear regression again, or do I look at non-zero estimation of the parameter values?*-->
 
-### Empirical approach with more complicated population model (use model similar to de Valpine work?)
-
-### Growth rate calculations
+# Minimum time required to estimate geometric growth rate
 
@@
 ```{r, echo=F, fig.cap='Example of calculating minimum time required for growth rate estimation.
-(a) abundance over time, (b) mean and standard deviation of growth rate for subsamples of entire time series, and (c) the percent error between mean estimated growth rate and the true longterm growth rate.
-The vertical bar denotes the minimum time required to estimate growth rate within 10% error.\\label{fig:growth_rate}'}
+(a) European herring gull (\\emph{Larus argentatus}) scaled density over time, (b) mean and standard deviation of growth rate for subsamples of entire time series, and (c) the percent error between mean estimated growth rate and the true long-term growth rate.
+The vertical bar denotes the minimum time required to estimate growth rate within 20% error.\\label{fig:growth_rate}'}
 #Example of using population growth rate instead of power
@@
 
-  pop = subset(long_dat,long_dat$ID=='10007') #13192 for Harp seals
+  pop = subset(long_dat,long_dat$ID=='10008') #13192 for Harp seals, 10007 is California Gull, 10008 for European herring gull (Larus argentatus)
   source('scripts/create_subsamples.R')
@@
     mtext(text = 'population size',2,line=3.5,cex=1.4)
+    mtext(text = '(a)',side = 3,line = -1.2,adj=0.95,font = 2,cex=1.2)
 
@@
     arrows(1:34,rowMeans(aaa,na.rm=T),1:34,rowMeans(aaa,na.rm=T)-sd(apply(aaa,MARGIN = 1,FUN=var,na.rm=T),na.rm=T),angle = 90,length = 0.1)
+    mtext(text = '(b)',side = 3,line = -1.2,adj=0.95,font = 2,cex=1.2)
 
@@
     mtext(text = 'percent error in growth rate',2,line=3.7,cex=1.4)
-    abline(h=0.1,lty=2)
-    abline(v=tail(which(percent_error(rowMeans(aaa,na.rm=T),rowMeans(aaa,na.rm=T)[34])>0.1),1)+1,lty=2)
+    abline(h=0.2,lty=2)
+    abline(v=tail(which(percent_error(rowMeans(aaa,na.rm=T),rowMeans(aaa,na.rm=T)[34])>0.1),1),lty=2)
+    mtext(text = '(c)',side = 3,line = -1.2,adj=0.95,font = 2,cex=1.2)
 ```
@@
 We applied the same calculations to `r nrow(pop_info)` populations, as we did in the main text.
-We then obtain a distribution of the minimum time required to measure the ``true" longterm growth rate (Fig.
+We then obtain a distribution of the minimum time required to measure the ``true" long-term growth rate (Fig.
 \ref{fig:min_time_growth_dist}).
-We see a bimodal distribution with many populations required 30+ years to estimate the longterm growth rate.
-The large number of short year required is due to cases where the entire time series is consistently increasing or decreasing at the same rate.
+We see a bimodal distribution with many populations required 30+ years to estimate the long-term growth rate.
+The large number of short years required is due to cases where the entire time series is consistently increasing or decreasing at the same rate.
 
-```{r, echo=F,eval=F,fig.cap='Histogram of the minimum time required in order to estimate the longterm growth rate within 20% error.\\label{fig:min_time_growth_dist}'}
+```{r, echo=F,eval=T,fig.cap='Histogram of the minimum time required in order to estimate the long-term growth rate within 20% error.\\label{fig:min_time_growth_dist}'}
 # Growth rates calculations for each species
 setwd("/Users/eastonwhite/Desktop/Research/time-series-project")
-load('cleaned-data/cleaned_timeseries_database.Rdata')
+# load('cleaned-data/cleaned_timeseries_database.Rdata')
+# 
+# source('scripts/create_subsamples.R')
+# source('scripts/calculate_geometric_growth.R')
+# source('scripts/percent_error.R')
+#  for (j in 1:nrow(pop_info)){
+#     pop= subset(long_dat,long_dat$ID==pop_info$ID[j])
+#     pop$popvalue=as.numeric(pop$popvalue)
+#     pop$popvalue=pop$popvalue/max(pop$popvalue) #should I standaridize this in a better way
+#     
+#    # source('scripts/calculate_power_metric.R')
+#     aaa=calculate_average_geometric_growth(pop$popvalue)
+#     if (length(which(percent_error(rowMeans(aaa,na.rm=T),rowMeans(aaa,na.rm=T)[34])>0.2))==0){
+#       pop_info$min_time_growth_rate[j] = 2
+#     }else{
+#       pop_info$min_time_growth_rate[j] = tail(which(percent_error(rowMeans(aaa,na.rm=T),rowMeans(aaa,na.rm=T)[34])>0.2),1)+1}
+#     #print(paste(j,':',pop_info$min_time_for_power[j],sep=' '))
+#  }
+# save(pop_info,file = 'analysis-outputs/min_time_for_growth_rate0_2.Rdata')
 
- for (j in 1:nrow(pop_info)){
-    pop= subset(long_dat,long_dat$ID==pop_info$ID[j])[1:35,]
-    pop$popvalue=as.numeric(pop$popvalue)
-    pop$popvalue=pop$popvalue/max(pop$popvalue) #should I standaridize this in a better way
-
-   # source('scripts/calculate_power_metric.R')
-    aaa=calculate_average_geometric_growth(pop$popvalue)
-    if (length(which(percent_error(rowMeans(aaa,na.rm=T),rowMeans(aaa,na.rm=T)[34])>0.2))==0){
-      pop_info$min_time_growth_rate[j] = 2
-    }else{
-      pop_info$min_time_growth_rate[j] = tail(which(percent_error(rowMeans(aaa,na.rm=T),rowMeans(aaa,na.rm=T)[34])>0.2),1)+1}
-    #print(paste(j,':',pop_info$min_time_for_power[j],sep=' '))
- }
+load('analysis-outputs/min_time_for_growth_rate0_2.Rdata')
 
@@
 hist(pop_info$min_time_growth_rate,las=1,ylab='',xlab='',cex.axis=1.2,main='',breaks=20,xlim=c(0,40),ylim=c(0,500))
-mtext(text = 'Frequency',2,line=3,cex=1.2)
-mtext(text = 'Minimum time required',1,line=3,cex=1.2)
+mtext(text = 'frequency',2,line=3,cex=1.2)
+mtext(text = 'minimum time required',1,line=3,cex=1.2)
 
@@
 
-### Using Generalized additive model to identify significant trends
+# Using Generalized additive model to identify significant trends
 
@@
 
-```{r,echo=F,fig.cap='results here from GAM.\\label{fig:gam_example}'}
+```{r,echo=F,message=F,warning=F,fig.cap='(a) Time series for Bigeye tuna (\\emph{Thunnus obesus}) with corresponding fitted GAM model in red and (b) statistical power as a function of the number of years sampled.
+The horizontal ine at 0.8 indicates the minimum threshold for statistical power and the vertical line denotes the minimum time required to achieve 0.8 statistical power.\\label{fig:gam_example}'}
 # show gam example side-by-side plot
-plot(1,1)
+ setwd("~/Desktop/Research/time-series-project")
+ load("cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
+ source('scripts/calculate_power_metric_gam.R')
+  par(mfrow=c(1,2),oma=c(4,4,0.5,0.5))
+    pop = subset(long_dat,long_dat$ID=='7558') #13192 for Harp seals, 7558 for Thunnus obesus
+    plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='',xlab='time (years)',pch=16,las=1)
+    example_gam_model=gam(pop$popvalue ~ s(pop$year,k=3),family=gaussian)
+    points(pop$year,example_gam_model$fitted.values,col='red',type='l',lwd=2)
+    mtext(text = '(a)',side = 3,line = -1.2,adj=0.95,font = 2,cex=1.2)
+    mtext(text='population size', side=2,line=4,cex=1.2)
+# could insert all possible regressions of length n
+
+    power_example=calculate_power_for_subsamples(pop$popvalue,0.05)
+    plot(power_example,cex.axis=1.2,cex.lab=1.2,ylab='statistical power',xlab='years sampled',las=1,pch=16)
+    mtext(text = '(b)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
+# could also insert standard error bars
+    abline(h=0.8,col='red',lwd=2,lty=2)
+    abline(v=head(which(power_example>0.8),1),col='red',lwd=2,lty=2)
 ```
 
-We then fit GAM models for XX populations. We found... (Fig. \ref{fig:gam_result}). This is shorter/longer than the main text question.
-
-```{r,echo=F,eval=F,fig.cap='results here from GAM.\\label{fig:gam_result}'}
+```{r,echo=F,fig.cap='Distribution of the minimum time required in order to detect a significant trend (at the 0.05 level) in abundance according to a GAM model given statistical power of 0.8.
+The smoothing parameter was set to 3 for each population.\\label{fig:min_time_dist_gam}'}
 # show gam results for minimium time required
 setwd("/Users/eastonwhite/Desktop/Research/time-series-project")
-load('cleaned-data/cleaned_timeseries_database.Rdata')
+load("cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
+main_text = pop_info
+# load('cleaned-data/cleaned_timeseries_database.Rdata')
+#  source('scripts/calculate_power_metric_gam.R')
+#  for (j in 1:nrow(pop_info)){
+#     pop= subset(long_dat,long_dat$ID==pop_info$ID[j])
+#     pop$popvalue=as.numeric(pop$popvalue)
+#     pop$popvalue=pop$popvalue/max(pop$popvalue) #should I standaridize this in a better way
+#   
+#     pop$trend_p_value = calculate_p_value_gam(pop$popvalue)
+#  
+#     pop_info$min_time_for_power_gam[j] = min_time_needed(pop$popvalue,0.05,0.8)
+#     print(paste(j,':',pop_info$min_time_for_power[j],sep=' '))
+#  }
+# 
+# #pop_info$min_time_for_power_gam[pop_info$min_time_for_power_gam==-999]=33
+# 
+# pop_info=pop_info[pop_info$trend_p_value<0.05,]
+# long_dat=subset(long_dat,long_dat$ID %in% pop_info$ID)
+# long_time_names=subset(long_time_names,long_time_names %in% pop_info$ID)
+# 
+# save(pop_info,long_dat,long_time_names,file='analysis-outputs/min_time_required_GAM.Rdata')
 
- for (j in 1:nrow(pop_info)){
-    pop= subset(long_dat,long_dat$ID==pop_info$ID[j])[1:35,]
-    pop$popvalue=as.numeric(pop$popvalue)
-    pop$popvalue=pop$popvalue/max(pop$popvalue) #should I standaridize this in a better way
-
-    source('scripts/calculate_power_metric_gam.R')
-    pop_info$min_time_for_power_gam[j] = min_time_needed(pop$popvalue,0.05,0.8)
-    print(paste(j,':',pop_info$min_time_for_power[j],sep=' '))
- }
-
+load('analysis-outputs/min_time_required_GAM.Rdata')
 # Create plot
 par(mfrow=c(1,1))
-hist(pop_info$min_time_growth_rate,las=1,ylab='',xlab='',cex.axis=1.2,main='',breaks=20,xlim=c(0,40),ylim=c(0,500))
+min_time_gam = pop_info$min_time_for_power_gam[pop_info$min_time_for_power_gam!=-999]
+plot(density(min_time_gam),las=1,ylab='',xlab='',cex.axis=1.2,main='',xlim=c(0,40))
 mtext(text = 'Frequency',2,line=3,cex=1.2)
 mtext(text = 'Minimum time required',1,line=3,cex=1.2)
-
 ```
 
-### Minimum time to detect density-dependence in time series
+We then fit GAM models for `r length(pop_info$min_time_for_power_gam[pop_info$min_time_for_power_gam!=-999])` populations.
+We found a similar distribution of minimum time required as in the main text for linear regression (Fig.
+\ref{fig:min_time_dist_gam}).
+However, in line with our hypothesis, the GAM models did result in a lower mean minimum time required of `r round(mean(pop_info$min_time_for_power_gam[pop_info$min_time_for_power_gam!=-999]),2)` years compared to the results from the main text of `r round(mean(main_text$min_time_for_power),2)` years.
+
+<!--
+## Minimum time to detect density-dependence in time series
 
@@
 They fitted 627 time series with a Gompertz population model and found evidence for density-dependence in 45% of the time series.
-Each time series in their database contained at least 15 datapoints, and ranged from 15 to 157.
+Each time series in their database contained at least 15 data points, and ranged from 15 to 157.
 A follow up question to their work would be, what time series length is required to confidently estimate density-dependence.
+-->
 
 \clearpage
-### References
+
+# References

### 2017-07-07 writing/supp_mat_draft.tex
cleaned up supp mat

@@
 \urlstyle{same}  % don't use monospace font for urls
-\usepackage{natbib}
-\bibliographystyle{plainnat}
 \usepackage{longtable,booktabs}
@@
 \renewcommand{\thetable}{S\arabic{table}}
-\renewcommand{\thefigure}{S\arabic{figure}}
+\renewcommand{\thefigure}{S\arabic{figure}} \usepackage[round]{natbib}
+\bibpunct[; ]{(}{)}{,}{a}{}{;}
 
@@
 
+\vspace{1cm}
+
 Code for all the figures and tables can be found at
-\href{https://github.com/erwhite1}{}.
+(\url{https://github.com/erwhite1}).
 
 \vspace{2cm}
+
+\clearpage
 
@@
 
-Here, we illustrate in detail how we performed the subsampling and power
-calculations for a specific population. As an example, we examine a
-35-year time series of (). Simple linear regression indicates a
-significant decline for this population with a blank and blank. We
-assume that this significant decline over 35 years is in fact the ``true
-trend``. In statistical jargon, the 35-year trend is an effect that is
-actually present. We can then use this as a benchmark to see if
-subsamples of the time series show a similar result.
+Here, we illustrate how we performed the subsampling and power
+calculations for a specific population. We provide more detail here than
+in the main manuscript. As an example, we examine a 35-year time series
+of Bigeye tuna (\emph{Thunnus obesus}), one species in the Global
+Population Dynamics Database (NERC Centre for Population Biology 2010).
+We assume that this significant increase over 35 years is in fact the
+``true trend``. In statistical jargon, the 35-year trend is an effect
+that is actually present, we can reject the null hypothesis of no trend.
+We can then use this as a benchmark to see if subsamples of the time
+series also indicate a significant increase.
 
@@
 only consider subsamples to be significant if they are significant in
-the same direction as the complete 35-year time series.
+the same direction as the complete 35-year time series. In other words,
+we are conducting a one-tailed test.
 
-We can then plot statistical power as a function of time series length.
-As expected, we can see that power increases with the more years that
-are sampled.
-
-Then, we determine an appropriate level of statistical power that we
-find acceptable. Traditionally, this has been at 0.8, however, this is
-purely historical. Statistical power of 0.8 implies\ldots{}
-
-With statistical power of 0.8, we then determine the minimum time series
-length (\(t_min\)) required to achieve that level of statistical power.
-Here, \(t_min\) is the first point as which all points to the right are
-above 0.8.
+We can then plot statistical power as a function of time series length
+(Fig. \ref{fig:empirical_approach_example}b). As expected, we can see
+that power increases with the more years that are sampled. Then, we
+determine an appropriate level of statistical power that we find
+acceptable. Traditionally, this has been at 0.8, however, this is purely
+historical. Statistical power of 0.8 implies if a true trend is present,
+or a real change in abundance, then we will detect this trend 0.8
+fraction of the time.
 
@@
 \centering
-\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-1-1.pdf}
-\caption{(a) Population size of Harp Seals over time. The line is the
-best fit line from linear regresssion. (b) Statistical power for
-different subsets of the time series in panel
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-2-1.pdf}
+\caption{(a) Population size of Bigeye tuna (\emph{Thunnus obesus}) over
+time. The line is the best fit line from linear regresssion. (b)
+Statistical power for different subsets of the time series in panel
 a.\label{fig:empirical_approach_example}}
@@
 
-In Figure \ref{fig:empirical_approach_example}b, statistical power
-increases with more time sampled until our desired power of 0.8 at 16
-years. Therefore, 16 years would be our estimate for \(T_{min}\).
+With statistical power of 0.8, we then determine the minimum time series
+length (\(T_{min}\)) required to achieve that level of statistical
+power. Here, \(T_{min}\) is the first point in Fig.
+\ref{fig:empirical_approach_example}b where following points are also
+above 0.8. In this example, \(T_{min}\) is 22. Therefore, a minimum of
+22 years of continuous monitoring are required to have 0.8 statistical
+power in determining long-term changes in abundance.
 
@@
 correlated with the trend strength, temporal autocorrelation, and
-variance in population size. Here we use a generalized linear model
+variance in population size. Here, we use a generalized linear model
 framework with a Poisson error structure to determine drivers of the
@@
 \includegraphics{supp_mat_draft_files/figure-latex/poisson_model-1.pdf}
-\caption{Output of poisson regression\ldots{}\label{fig:poisson_model}}
+\caption{Output of generalized linear model with a Poisson error
+structure for predicting the minimum time required with explanatory
+variables of the absolute value of the slope coefficient (or trend
+strength), temporal autocorrelation, and variability in population
+size.\label{fig:poisson_model}}
 \end{figure}
@@
 \begin{longtable}[]{@{}lrrrr@{}}
-\caption{Stuff here\label{table:model_output}}\tabularnewline
+\caption{Output of generalized linear model to examine predictors of the
+minimum time required for determing long-term population
+trends.\label{table:model_output}}\tabularnewline
 \toprule
@@
 \endhead
-(Intercept) & 3.7141297 & 0.0274350 & 135.379203 &
-0.0000000\tabularnewline
-abs\_overall\_trend & -104.4974288 & 3.0997732 & -33.711314 &
-0.0000000\tabularnewline
-autocorrelation & -0.4905888 & 0.0431062 & -11.380936 &
-0.0000000\tabularnewline
-variance & 18.7756472 & 0.6409982 & 29.291262 & 0.0000000\tabularnewline
-gen\_len & 0.0053584 & 0.0025995 & 2.061352 & 0.0392695\tabularnewline
+(Intercept) & 3.7405513 & 0.0242742 & 154.09595 & 0\tabularnewline
+abs\_overall\_trend & -104.0533454 & 3.0941097 & -33.62949 &
+0\tabularnewline
+autocorrelation & -0.4899533 & 0.0431769 & -11.34757 & 0\tabularnewline
+variance & 18.7366110 & 0.6404458 & 29.25557 & 0\tabularnewline
 \bottomrule
@@
 Our model with trend strength, autocorrelation, variance, and generation
-length accounted for 72.55\% of the variation in the minimum time
+length accounted for 72.45\% of the variation in the minimum time
 required (Table \ref{table:model_output}). However, we also found trend
@@
 variance, but not both together. This resulted in lower explained
-deviance (analogue to \(R^2\)) of 53.08\% and 45.04\%, respectively.
+deviance (analogue to \(R^2\)) of 53.03\% and 45.03\%, respectively.
 
@@
 Here, we combined our time series database with a database on life
-history characteristics of amniotes from (cite other database here).
-There was life history information available for 547 populations
-representing 315 different species, all of which were in the Aves class.
+history characteristics of amniotes from Myhrvold et al. (2015). There
+was life history information available for 547 populations representing
+315 different species, all of which were in the Aves class.
 
@@
 \centering
-\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-4-1.pdf}
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
 \caption{Mimimum time required versus (a) generation length (years), (b)
@@
 \centering
-\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-6-1.pdf}
 \caption{(a) Minimum time required to estimate change in abundance for
-species class, (b) longterm trend (estimated slope coefficient) by
+species class, (b) long-term trend (estimated slope coefficient) by
 species class, (c) interannual variability in population size by species
@@
 
-\subsection{\texorpdfstring{Sensitivity analysis of \(\alpha\) and
-power}{Sensitivity analysis of \textbackslash{}alpha and power}}\label{sensitivity-analysis-of-alpha-and-power}
+\subsection{Sensitivity analysis of significance level and
+power}\label{sensitivity-analysis-of-significance-level-and-power}
 
@@
 \centering
-\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-6-1.pdf}
-\caption{Minimum time required to assess longterm trends in abundance
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-7-1.pdf}
+\caption{Minimum time required to assess long-term trends in abundance
 for values of statistical significance (\(\alpha\)) and power
@@
 
+\section{Minimum time calculations testing exponential
+growth}\label{minimum-time-calculations-testing-exponential-growth}
+
+In the main text we evaluated the minimum time required to determine
+long-term trends in abundance via linear regression. This process
+examined linear trends in abundance over time. Here we examine the
+minimum time required to estimate long-term trends that are either
+exponential growth or decay. We use the same methods as described in the
+previous sections, but we take the \(log\) of population density, or
+abundance.
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-8-1.pdf}
+\caption{Distribution of the minimum time required in order to detect a
+significant trend (at the 0.05 level) in log(abundance) given power of
+0.8.\label{fig:min_time_dist_log_pop}}
+\end{figure}
+
+We see that the distribution of \(T_{min}\) is almost identical to that
+in the main manuscript. This is perhaps not surprising as most time
+series that would signficantly increase or decrease linearly would
+probably also significantly increase or decrease at an exponential rate.
+Further, the calculations here and in the main manuscript both use
+linear regression, and both estimate the same number of parameters.
+
 \pagebreak
 
-\section{Results from other questions of
-interest}\label{results-from-other-questions-of-interest}
-
-\subsection{Simulations with more complicated population
+\section{Simulations with more complicated population
 model}\label{simulations-with-more-complicated-population-model}
@@
 main text, but simulate from a more biologically-realistic population
-model, the (MODEL NAME HERE):
+model. We simulate the model described in White, Nagy, and Gruber
+(2014). The model is a stochastic, age-structured population model that
+includes density-dependence for lemon sharks in Bimini, Bahamas.
 
-\begin{equation}
-    y = x
-  \end{equation}
+\begin{figure}[htbp]
+\centering
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-9-1.pdf}
+\caption{Statistical power for different length of time series
+simulations for a lemon shark population in Bimini,
+Bahamas.\label{fig:shark_example}}
+\end{figure}
 
-We simulate this model repetitively for different lengths of time
-series. Statistical power is the fraction of time series of a set length
-that showed significant trends in abundance, given a true trend in
-abundance. \emph{How do I determine significance here? Through linear
-regression again, or do I look at non-zero estimation of the parameter
-values?}
+Here, we parameterize the model for a situation where adult mortality
+rate is high enough to cause a population decline. In the same way we
+simulated models in the main text, we simulate this more
+biologically-realistic model for different lengths of time. For each
+length of time, we calculate the statistical power. Similar to results
+with the simpler model, statistical power generally increases with
+longer sampling time (Fig. \ref{fig:shark_example}). In this example,
+the minimum time required (\(T_{min}\)) to obtain at least 0.8
+statistical power, given a significance level of 0.05, is 27 years.
 
-\subsection{Empirical approach with more complicated population model
-(use model similar to de Valpine
-work?)}\label{empirical-approach-with-more-complicated-population-model-use-model-similar-to-de-valpine-work}
-
-\subsection{Growth rate calculations}\label{growth-rate-calculations}
+\section{Minimum time required to estimate geometric growth
+rate}\label{minimum-time-required-to-estimate-geometric-growth-rate}
 
@@
 define the minimum time required as the minimum number of years to
-achieve less than 10\% error.
+achieve less than 20\% error.
 
@@
 \centering
-\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-8-1.pdf}
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-10-1.pdf}
 \caption{Example of calculating minimum time required for growth rate
-estimation. (a) abundance over time, (b) mean and standard deviation of
-growth rate for subsamples of entire time series, and (c) the percent
-error between mean estimated growth rate and the true longterm growth
-rate. The vertical bar denotes the minimum time required to estimate
-growth rate within 10\% error.\label{fig:growth_rate}}
+estimation. (a) European herring gull (\emph{Larus argentatus}) scaled
+density over time, (b) mean and standard deviation of growth rate for
+subsamples of entire time series, and (c) the percent error between mean
+estimated growth rate and the true long-term growth rate. The vertical
+bar denotes the minimum time required to estimate growth rate within
+20\% error.\label{fig:growth_rate}}
 \end{figure}
@@
 main text. We then obtain a distribution of the minimum time required to
-measure the ``true" longterm growth rate (Fig.
+measure the ``true" long-term growth rate (Fig.
 \ref{fig:min_time_growth_dist}). We see a bimodal distribution with many
-populations required 30+ years to estimate the longterm growth rate. The
-large number of short year required is due to cases where the entire
-time series is consistently increasing or decreasing at the same rate.
+populations required 30+ years to estimate the long-term growth rate.
+The large number of short years required is due to cases where the
+entire time series is consistently increasing or decreasing at the same
+rate.
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-11-1.pdf}
+\caption{Histogram of the minimum time required in order to estimate the
+long-term growth rate within 20\%
+error.\label{fig:min_time_growth_dist}}
+\end{figure}
 
@@
 
-\subsection{Using Generalized additive model to identify significant
+\section{Using Generalized additive model to identify significant
 trends}\label{using-generalized-additive-model-to-identify-significant-trends}
@@
 \centering
-\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-10-1.pdf}
-\caption{results here from GAM.\label{fig:gam_example}}
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-12-1.pdf}
+\caption{(a) Time series for Bigeye tuna (\emph{Thunnus obesus}) with
+corresponding fitted GAM model in red and (b) statistical power as a
+function of the number of years sampled. The horizontal ine at 0.8
+indicates the minimum threshold for statistical power and the vertical
+line denotes the minimum time required to achieve 0.8 statistical
+power.\label{fig:gam_example}}
 \end{figure}
 
-We then fit GAM models for XX populations. We found\ldots{} (Fig.
-\ref{fig:gam_result}). This is shorter/longer than the main text
-question.
+\begin{figure}[htbp]
+\centering
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-13-1.pdf}
+\caption{Distribution of the minimum time required in order to detect a
+significant trend (at the 0.05 level) in abundance according to a GAM
+model given statistical power of 0.8. The smoothing parameter was set to
+3 for each population.\label{fig:min_time_dist_gam}}
+\end{figure}
 
-\subsection{Minimum time to detect density-dependence in time
-series}\label{minimum-time-to-detect-density-dependence-in-time-series}
-
-For decades, the importance of density-dependence in natural populations
-has been an important question in ecology (citations). To address this
-question, \citet{Knape2012} used data from the Global Population
-Dynamics Database \citep{GPDD2010}. They fitted 627 time series with a
-Gompertz population model and found evidence for density-dependence in
-45\% of the time series. Each time series in their database contained at
-least 15 data points, and ranged from 15 to 157. A follow up question to
-their work would be, what time series length is required to confidently
-estimate density-dependence.
+We then fit GAM models for 851 populations. We found a similar
+distribution of minimum time required as in the main text for linear
+regression (Fig. \ref{fig:min_time_dist_gam}). However, in line with our
+hypothesis, the GAM models did result in a lower mean minimum time
+required of 14.65 years compared to the results from the main text of
+16.45 years.
 
@@
 
-\renewcommand\refname{References}
-\bibliography{White_bib.bib}
+\section*{References}\label{references}
+\addcontentsline{toc}{section}{References}
+
+\hypertarget{refs}{}
+\hypertarget{ref-Myhrvold2015}{}
+Myhrvold, Nathan P., Elita Baldridge, Benjamin Chan, Dhileep Sivam,
+Daniel L. Freeman, and S.K. Morgan Ernest. 2015. ``An amniote
+life-history database to perform comparative analyses with birds,
+mammals, and reptiles.'' \emph{Ecology} 96 (11): 3109.
+
+\hypertarget{ref-GPDD2010}{}
+NERC Centre for Population Biology, Imperial College. 2010. ``The Global
+Population Dynamics Database Version 2.''
+
+\hypertarget{ref-White2014}{}
+White, Easton R, John D Nagy, and Samuel H Gruber. 2014. ``Modeling the
+population dynamics of lemon sharks.'' \emph{Biology Direct} 9 (23):
+1--18.
 

### 2017-07-07 scripts/scripts_to_clean_data/combine_databases.R
changed cleaning code to only retain most 35 recent years of each time series

@@ -60,15 +60,15 @@ save(LPI_pop_info,new_dat,file = 'cleaned-data/combined_databases_with_min_time_
 # Simple plots
 if (makePlot=TRUE){
 par(mfrow=c(1,3),oma=c(4,4,1,1),mar=c(1,3,0,0))
-with(LPI_pop_info,plot(gen_len,min_time_required,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2))
+with(LPI_pop_info,plot(gen_len,min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2))
 mtext(text = 'minimum time required',side = 2,line = 3,cex = 1.4)
 mtext(text = 'generation length (years)',side = 1,line = 3,cex = 1.4)
 
-with(LPI_pop_info,plot(litter_or_clutch_size_n,min_time_required,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2))
+with(LPI_pop_info,plot(litter_or_clutch_size_n,min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2))
 #mtext(text = 'minimum time required',side = 2,line = 3,cex = 1.4)
 mtext(text = 'litter size',side = 1,line = 3,cex = 1.4)
 
-with(LPI_pop_info,plot(log(adult_body_mass_g),min_time_required,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2))
+with(LPI_pop_info,plot(log(adult_body_mass_g),min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2))
 #mtext(text = 'minimum time required',side = 2,line = 3,cex = 1.4)
 mtext(text = 'log(adult body mass (grams))',side = 1,line = 3,cex = 1.4)
 }

### 2017-07-07 scripts/scripts_to_clean_data/create-pop-info-table.R

@@ -40,8 +40,11 @@ pop_info$trend_p_value=NA
 pop_info$min_time_for_power=NA
 pop_info$popvalue=NA
 time_series_data=NULL
+
 source('scripts/calculate_slope.R')
-  for (j in 1:nrow(pop_info)){
+source('scripts/calculate_power_metric.R')
+
+for (j in 1:nrow(pop_info)){
     pop= subset(long_dat,long_dat$ID==pop_info$ID[j])
     pop = pop[(nrow(pop)-34):nrow(pop),]
     pop$popvalue=as.numeric(pop$popvalue)
@@ -54,13 +57,13 @@ source('scripts/calculate_slope.R')
     pop_info$popvalue = pop$popvalue[1] #initial abundance
     pop_info$series_length = nrow(pop)
     
-   #source('scripts/calculate_power_metric.R')
+
      pop_info$min_time_for_power[j] = min_time_needed(pop$popvalue,0.05,0.8)
     print(paste(j,':',pop_info$min_time_for_power[j],sep=' '))
     
     
     time_series_data = rbind(time_series_data,pop)
-  }
+}
 
 long_dat = time_series_data
 
@@ -68,12 +71,15 @@ long_dat = time_series_data
 pop_info$IUCN = 3*pop_info$gen_len
 pop_info$IUCN[pop_info$IUCN<10]=10
 
-# Save output here
-save(pop_info,long_dat,long_time_names,file = 'cleaned-data/cleaned_timeseries_database.Rdata')
+
+
+
+##### SAVE OUTPUT HERE ######
+save(pop_info[,c(1:13,15)],long_dat,long_time_names,file = 'cleaned-data/cleaned_timeseries_database.Rdata')
 
 #Only look at populations with certain level of the overall trend
-#pop_info=pop_info[pop_info$trend_p_value<0.05,]
-#long_dat=subset(long_dat,long_dat$ID %in% pop_info$ID)
-#long_time_names=subset(long_time_names,long_time_names %in% pop_info$ID)
+pop_info=pop_info[pop_info$trend_p_value<0.05,]
+long_dat=subset(long_dat,long_dat$ID %in% pop_info$ID)
+long_time_names=subset(long_time_names,long_time_names %in% pop_info$ID)
 
-
+save(pop_info,long_dat,long_time_names,file = 'cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata')

### 2017-07-07 writing/manuscript_draft.Rmd
changed code to allow relative directories

@@
 # Introduction
-
-<!--
-- intro long time series
-- my general question
-- primer on power
-- more general power interpretation
-- 2+ approaches
-- what we do in this paper
--->
 
@@
 (c) Minimum time required for different values of the trend strength ($r$) and (d) minimum time required for different levels of population variability ($\\sigma$).\\label{fig:theoretical_approach}'}
-setwd("~/Desktop/Research/time-series-project")
-source("scripts/theoretical-models/calculate_power.R")
-source("scripts/theoretical-models/simulate_pop.R")
+source("../scripts/theoretical-models/calculate_power.R")
+source("../scripts/theoretical-models/simulate_pop.R")
 
@@
 # Pull in original database from Keith et al. 2015 and the cleaned version after our series of filtering steps
-  setwd("~/Desktop/Research/time-series-project")
 
 #Load keith et al. database (from GPDD database)
-  load('data/Keith_et_al_2015_data.Rdata')
+  load('../data/Keith_et_al_2015_data.Rdata')
 
 #Load our cleaned data (see other code for cleaning steps)
-  load('cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata') #Do i need to filter out non-significant populations as well (where no trend was present)
+  load('../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata') #Do i need to filter out non-significant populations as well (where no trend was present)
  # pop_info=pop_info[pop_info$trend_p_value<0.05,]
@@
 # Pull in results of biological characteristics comparison to min time series length required
-setwd("~/Desktop/Research/time-series-project")
-load('cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
+load('../cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
 ```
@@
 # Pull in analysis of average min time needed for **empirical species** (save analysis as Rdata file or something)
- setwd("~/Desktop/Research/time-series-project")
- load("cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
+ load("../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
 
@@
 ```
-
-```{r,echo=F,eval=F}
-#summary(poisson_model)
-```
-
-<!--
-# Potential figures
-- Examples of how determining minimum time series works (probably in methods)
-- What min time should scale with theoretically (trend, power, variability, etc) - another multi panel figure
-- Mean min time by class or system
-- Separate results for species, body size, location, biology. - multi panel figure
-
--->
 
@@
 \end{figure}
-
-\begin{longtable}[]{@{}lrrrr@{}}
-\caption{Output of generalized linear model to examine predictors of the
-minimum time required for determing long-term population
-trends.\label{table:model_output}}\tabularnewline
-\toprule
-& Estimate & Std. Error & z value &
-Pr(\textgreater{}\textbar{}z\textbar{})\tabularnewline
-\midrule
-\endfirsthead
-\toprule
-& Estimate & Std. Error & z value &
-Pr(\textgreater{}\textbar{}z\textbar{})\tabularnewline
-\midrule
-\endhead
-(Intercept) & 3.7405513 & 0.0242742 & 154.09595 & 0\tabularnewline
-abs\_overall\_trend & -104.0533454 & 3.0941097 & -33.62949 &
-0\tabularnewline
-autocorrelation & -0.4899533 & 0.0431769 & -11.34757 & 0\tabularnewline
-variance & 18.7366110 & 0.6404458 & 29.25557 & 0\tabularnewline
-\bottomrule
-\end{longtable}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
-\caption{Mimimum time required versus (a) generation length (years), (b)
-litter size (n), (c) adult body mass (grams), (d) maximum longevity
-(years), (e) egg mass (grams), and (f) incubation (days). The lines in
-each plot represent the best fit line from linear
-regression.\label{fig:biological_correlates}}
-\end{figure}

### 2017-07-07 writing/manuscript_draft.tex
changed code to allow relative directories

@@
 
-\begin{longtable}[]{@{}lrrrr@{}}
-\caption{Output of generalized linear model to examine predictors of the
-minimum time required for determing long-term population
-trends.\label{table:model_output}}\tabularnewline
-\toprule
-& Estimate & Std. Error & z value &
-Pr(\textgreater{}\textbar{}z\textbar{})\tabularnewline
-\midrule
-\endfirsthead
-\toprule
-& Estimate & Std. Error & z value &
-Pr(\textgreater{}\textbar{}z\textbar{})\tabularnewline
-\midrule
-\endhead
-(Intercept) & 3.7405513 & 0.0242742 & 154.09595 & 0\tabularnewline
-abs\_overall\_trend & -104.0533454 & 3.0941097 & -33.62949 &
-0\tabularnewline
-autocorrelation & -0.4899533 & 0.0431769 & -11.34757 & 0\tabularnewline
-variance & 18.7366110 & 0.6404458 & 29.25557 & 0\tabularnewline
-\bottomrule
-\end{longtable}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
-\caption{Mimimum time required versus (a) generation length (years), (b)
-litter size (n), (c) adult body mass (grams), (d) maximum longevity
-(years), (e) egg mass (grams), and (f) incubation (days). The lines in
-each plot represent the best fit line from linear
-regression.\label{fig:biological_correlates}}
-\end{figure}
-
 \end{document}

### 2017-07-07 writing/supp_mat_draft.Rmd
changed code to allow relative directories

@@
 ```{r,echo=F}
- setwd("~/Desktop/Research/time-series-project")
- load("cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
+ #
+ load("../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
     pop = subset(long_dat,long_dat$ID=='7558')
@@
 # Example of subsampling empirical data
- setwd("~/Desktop/Research/time-series-project")
- load("cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
+ load("../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
   par(mfrow=c(1,2),oma=c(4,4,0.5,0.5))
@@
 # could insert all possible regressions of length n
-source('scripts/calculate_power_metric.R')
+source("../scripts/calculate_power_metric.R",chdir=T)
     power_example=calculate_power_for_subsamples(pop$popvalue,0.05)
@@
 ```{r poisson_model, echo=F,include=T,fig.cap='Output of generalized linear model with a Poisson error structure for predicting the minimum time required with explanatory variables of the absolute value of the slope coefficient (or trend strength), temporal autocorrelation, and variability in population size.\\label{fig:poisson_model}'}
-setwd("~/Desktop/Research/time-series-project")
-load("cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
+
+load("../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
 
@@
 ```{r, echo=F}
-setwd("~/Desktop/Research/time-series-project")
-load('cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
+
+load('../cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
 ```
@@
 The lines in each plot represent the best fit line from linear regression.\\label{fig:biological_correlates}',fig.height=6}
-setwd("~/Desktop/Research/time-series-project")
-load('cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
+
+load('../cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
 par(mfrow=c(3,2),oma=c(5,4,1,1),mar=c(5,3,0,0))
@@
 # Pull in data
-setwd("~/Desktop/Research/time-series-project")
-load("~/Desktop/Research/time-series-project/cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
+
+load("../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
 # Biological correlates plots
@@
 # Explain results after building figures
-setwd("~/Desktop/Research/time-series-project")
-source('scripts/calculate_slope.R') 
-source('scripts/theoretical-models/simulate_pop.R')
+
+source('../scripts/calculate_slope.R') 
+source('../scripts/theoretical-models/simulate_pop.R')
 # calculate_power = function(r,phi,sigma,trials,max.time,alpha){
@@
 # Pull in data to prevent R from running for a long time to create data from scratch
-parameter_combinations = read.csv('analysis-outputs/min_time_vs_alpha_and_power.csv',header=T)
+parameter_combinations = read.csv('../analysis-outputs/min_time_vs_alpha_and_power.csv',header=T)
 
@@
 ```{r,echo=F,message=F,error=F,fig.cap='Distribution of the minimum time required in order to detect a significant trend (at the 0.05 level) in log(abundance) given power of 0.8.\\label{fig:min_time_dist_log_pop}'}
-setwd("~/Desktop/Research/time-series-project")
-# load('cleaned-data/cleaned_timeseries_database.Rdata')
+
+# load('../cleaned-data/cleaned_timeseries_database.Rdata')
 # 
-# source('scripts/calculate_slope.R')
-# source('scripts/calculate_power_metric.R')
+# source('../scripts/calculate_slope.R')
+# source('../scripts/calculate_power_metric.R')
 # 
@@
 # 
-# save(pop_info,long_dat,long_time_names,file = 'cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression_for_log_pop.Rdata')
+# save(pop_info,long_dat,long_time_names,file = '../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression_for_log_pop.Rdata')
 
 # LOAD in previously run script from this code chuck and make histogram plot
-load('cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression_for_log_pop.Rdata')
+load('../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression_for_log_pop.Rdata')
 
@@
 
-setwd("~/Desktop/Research/time-series-project")
-#source('scripts/LemonShark_PopulationModel.R')
+#source('../scripts/LemonShark_PopulationModel.R')
 years_to_sample=seq(5,50,by=1)
@@
 
-load('analysis-outputs/example_power_for_shark_model.Rdata')
+load('../analysis-outputs/example_power_for_shark_model.Rdata')
 par(mfrow=c(1,1),oma=c(3,3,0.5,0.5),mar=c(2,2,0,0))
@@
 
-#save(shark_power,file='analysis-outputs/example_power_for_shark_model.Rdata')
+#save(shark_power,file='../analysis-outputs/example_power_for_shark_model.Rdata')
 
@@
 #Example of using population growth rate instead of power
-setwd("/Users/eastonwhite/Desktop/Research/time-series-project")
-source("scripts/calculate_geometric_growth.R")
-load('cleaned-data/cleaned_timeseries_database.Rdata')
+source("../scripts/calculate_geometric_growth.R")
+load('../cleaned-data/cleaned_timeseries_database.Rdata')
 
@@
   pop = subset(long_dat,long_dat$ID=='10008') #13192 for Harp seals, 10007 is California Gull, 10008 for European herring gull (Larus argentatus)
-  source('scripts/create_subsamples.R')
-  source('scripts/percent_error.R')
+  source('../scripts/create_subsamples.R')
+  source('../scripts/percent_error.R')
   aaa=calculate_average_geometric_growth(pop$popvalue)
@@
 # Growth rates calculations for each species
-setwd("/Users/eastonwhite/Desktop/Research/time-series-project")
-# load('cleaned-data/cleaned_timeseries_database.Rdata')
+# load('../cleaned-data/cleaned_timeseries_database.Rdata')
 # 
-# source('scripts/create_subsamples.R')
-# source('scripts/calculate_geometric_growth.R')
-# source('scripts/percent_error.R')
+# source('../scripts/create_subsamples.R')
+# source('../scripts/calculate_geometric_growth.R')
+# source('../scripts/percent_error.R')
 #  for (j in 1:nrow(pop_info)){
@@
 #     
-#    # source('scripts/calculate_power_metric.R')
+#    # source('../scripts/calculate_power_metric.R')
 #     aaa=calculate_average_geometric_growth(pop$popvalue)
@@
 #  }
-# save(pop_info,file = 'analysis-outputs/min_time_for_growth_rate0_2.Rdata')
+# save(pop_info,file = '../analysis-outputs/min_time_for_growth_rate0_2.Rdata')
 
-load('analysis-outputs/min_time_for_growth_rate0_2.Rdata')
+load('../analysis-outputs/min_time_for_growth_rate0_2.Rdata')
 
@@
 # show gam example side-by-side plot
- setwd("~/Desktop/Research/time-series-project")
- load("cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
- source('scripts/calculate_power_metric_gam.R')
+
+ load("../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
+ source('../scripts/calculate_power_metric_gam.R')
   par(mfrow=c(1,2),oma=c(4,4,0.5,0.5))
@@
 # show gam results for minimium time required
-setwd("/Users/eastonwhite/Desktop/Research/time-series-project")
-load("cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
+load("../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
 main_text = pop_info
-# load('cleaned-data/cleaned_timeseries_database.Rdata')
-#  source('scripts/calculate_power_metric_gam.R')
+# load('../cleaned-data/cleaned_timeseries_database.Rdata')
+#  source('../scripts/calculate_power_metric_gam.R')
 #  for (j in 1:nrow(pop_info)){
@@
 # 
-# save(pop_info,long_dat,long_time_names,file='analysis-outputs/min_time_required_GAM.Rdata')
+# save(pop_info,long_dat,long_time_names,file='../analysis-outputs/min_time_required_GAM.Rdata')
 
-load('analysis-outputs/min_time_required_GAM.Rdata')
+load('../analysis-outputs/min_time_required_GAM.Rdata')
 # Create plot

### 2017-07-07 writing/supp_mat_draft.tex
changed code to allow relative directories

@@
 in the main manuscript. This is perhaps not surprising as most time
-series that would signficantly increase or decrease linearly would
+series that would significantly increase or decrease linearly would
 probably also significantly increase or decrease at an exponential rate.

### 2017-07-07 scripts/calculate_power_metric.R

@@ -9,8 +9,8 @@
 require(plyr)
 
 #pull in neccessary functions
-source('scripts/calculate_slope.R')
-source('scripts/create_subsamples.R')
+source('../scripts/calculate_slope.R')
+source('../scripts/create_subsamples.R')
 
 
 #This is a two-sided test (should use 1 tailed test)

### 2017-07-07 scripts/calculate_power_metric_gam.R

@@ -12,8 +12,8 @@
 require(plyr)
 
 #pull in neccessary functions
-source('scripts/calculate_p_value_gam.R')
-source('scripts/create_subsamples.R')
+source('../scripts/calculate_p_value_gam.R')
+source('../scripts/create_subsamples.R')
 
 
 ######## Implement two-tailed test

### 2017-07-07 scripts/theoretical-models/calculate_power.R

@@ -6,7 +6,7 @@
 
 calculate_power = function(r,phi,sigma,trials,max.time){
   
-  source('scripts/calculate_slope.R') # calls another function which calculates the slope and p_value of a linear regression on a time series
+  source('../scripts/calculate_slope.R') # calls another function which calculates the slope and p_value of a linear regression on a time series
   # Parameter values
   trials   = trials
   max.time = max.time

### 2017-07-07 writing/manuscript_draft.Rmd
added citations and modified code for figure 1

@@
 Here, we explore two approaches to address this question: through simulations and through an empirical approach.
-I specifically determine the minimum time series length required to estimate significant increases or decreases in population abundance.
-Importantly, I examine the ability to detect a trend with a set level of statistical power, which is often neglected in ecological time series analyses.
+We specifically determine the minimum time series length required to estimate significant increases or decreases in population abundance.
+Importantly, We examine the ability to detect a trend with a set level of statistical power, which is often neglected in ecological time series analyses.
 Using a simulation approach, we demonstrate how the minimum time series length required increases with weaker trends in abundance and with higher population variability.
@@
 
-Observational studies and population censuses have become a cornerstone of modern ecological research and conservation biology [@Magurran2010].
-long-term data is necessary to both understand population dynamics and to assess extinction risk of species.
-Even though many time series may now be considered "long-term", most are still short (i.e.
-less than X years).
-*I could have statistic on how long is considered "long-term" in papers*.
+Observational studies and population censuses have become a cornerstone of modern ecological research and conservation biology [@Magurran2010; @Hughes2017].
+Long-term data is necessary to both understand population dynamics and to assess extinction risk of species.
+Even though many time series may now be considered "long-term", most are still short.
 Time series are typically short for a variety of reasons.
 They are often coupled with an experiment, which may only last a couple of years.
-In addition, short funding cycles make it difficult to examine populations over longer periods of time (citations).
-There are, of course, notable exceptions including the Isle Royale study, Iowa corn study, cod, lynx-hair, and many others (citations).
+In addition, short funding cycles make it difficult to examine populations over longer periods of time [@Hughes2017]].
+There are, of course, notable exceptions where long-term monitoring has occured, for example the continuous plankton recorder [@Giron-Nava2017].
 
@@
 For time series data, two general approaches can be used.
-First, as can be done with experimental data, we can relax the definition of statistical power (cite Bolker).
+First, as can be done with experimental data, we can relax the definition of statistical power [@Bolker2008].
 Instead of using classic statistical tests (e.g.
-t-test), and corresponding formulas for determinig sample size, simulation approaches can be used [@Gerrodette1987].
+t-test), and corresponding formulas for determinig sample size, simulation approaches can be used [@Gerrodette1987, Bolker2008].
 Simple models can be simulated with parameter values corresponding to a population of interest.
@@
 The specific criteria depend on the question at hand.
-Then, time series simulations of different lengths are run to determine the minimum time series length required to be confident that you will be able to answer your question (cite Bolker).
+Then, time series simulations of different lengths are run to determine the minimum time series length required to be confident that you will be able to answer your question [@Bolker2008].
 In addition to using simulated data, empirical time series can also be used.
@@
 
-```{r,echo=F,eval=F,fig.cap='(a) example of simulated time series for 40 time periods and (b) statistical power versus the simulated time series length.
+```{r,echo=F,eval=T,fig.cap='(a) example of simulated time series for 40 time periods and (b) statistical power versus the simulated time series length.
 The horizontal, dashed line is our desired statistical power of 0.8.
@@
 
-  years_to_sample=seq(2,100,by=1)
-  power=lapply(X=years_to_sample,FUN = calculate_power,phi = 0.6,r=0.2,sigma = .5,trials=50)#should use 300+ trials
+  # years_to_sample=seq(2,100,by=1)
+  # power=lapply(X=years_to_sample,FUN = calculate_power,phi = 0.6,r=0.2,sigma = .5,trials=100)#should use 300+ trials
+  # 
+  # simulations = replicate(n=30,simulate_pop(phi = 0.6,r=0.2,sigma = .5,max.time=100))
+  # min_time_needed = years_to_sample[tail(which(power<0.8),1)] + 1
+  # 
+  # save(simulations,years_to_sample,power,min_time_needed,file='../analysis-outputs/example_simulations_fig1.Rdata')
+  load('../analysis-outputs/example_simulations_fig1.Rdata')
 
-  simulations = replicate(n=30,simulate_pop(phi = 0.6,r=0.2,sigma = .5,max.time=40))
-  min_time_needed = years_to_sample[tail(which(power<0.8),1)] + 1
-
+  #load command
   # Plotting commands
@@
 
-    # Plot versus trend strength
-      r_values = seq(1,3,by=1) #by=0.2
-      sigma_values=seq(1,7,by=1) #by=0.2
-
-      parameter_combinations = expand.grid(r_values,5)
-      parameter_combinations$min_time_required = NA
-      names(parameter_combinations)= c('r','sigma','min_time_required')
-
-      for (i in 1:nrow(parameter_combinations)){
-        years_to_sample=seq(2,50,by=1)
-        y=lapply(X=years_to_sample,FUN = calculate_power,phi =0.5,r=parameter_combinations$r[i],sigma = parameter_combinations$sigma[i],trials=100)
-        if (length(tail(which(y<0.8),1))>0){
-        parameter_combinations$min_time_required[i] = years_to_sample[tail(which(y<0.8),1)] + 1
-        }
-        #print(i)
-      }
+    #   
+    # # Plot versus trend strength
+    #   r_values = seq(1,3,by=0.1) #by=0.2
+    #   sigma_values=seq(1,7,by=0.3) #by=0.2
+    # 
+    #   parameter_combinations = expand.grid(r_values,5)
+    #   parameter_combinations$min_time_required = NA
+    #   names(parameter_combinations)= c('r','sigma','min_time_required')
+    #   
+    #   for (i in 1:nrow(parameter_combinations)){
+    #     years_to_sample=seq(2,50,by=1)
+    #     y=lapply(X=years_to_sample,FUN = calculate_power,phi =0.5,r=parameter_combinations$r[i],sigma = parameter_combinations$sigma[i],trials=150)
+    #     if (length(tail(which(y<0.8),1))>0){
+    #     parameter_combinations$min_time_required[i] = years_to_sample[tail(which(y<0.8),1)] + 1
+    #     }
+    #     #print(i)
+    #   }
+    #   
+    #   save(parameter_combinations,file='../analysis-outputs/min_time_vs_trend_strength.Rdata')
+      load('../analysis-outputs/min_time_vs_trend_strength.Rdata')
 
@@
 
-      # Plot versus sigma values
-      parameter_combinations = expand.grid(1.5,sigma_values)
-      parameter_combinations$min_time_required = NA
-      names(parameter_combinations)= c('r','sigma','min_time_required')
+      # 
+      # 
+      # 
+      # # Plot versus sigma values
+      # parameter_combinations = expand.grid(1.5,sigma_values)
+      # parameter_combinations$min_time_required = NA
+      # names(parameter_combinations)= c('r','sigma','min_time_required')
+      # 
+      # for (i in 1:nrow(parameter_combinations)){
+      #   years_to_sample=seq(2,50,by=1)
+      #   y=lapply(X=years_to_sample,FUN = calculate_power,phi =0.5,r=parameter_combinations$r[i],sigma = parameter_combinations$sigma[i],trials=150)
+      #   if (length(tail(which(y<0.8),1))>0){
+      #   parameter_combinations$min_time_required[i] = years_to_sample[tail(which(y<0.8),1)] + 1
+      #   }
+      #   #print(i)
+      # }
 
-      for (i in 1:nrow(parameter_combinations)){
-        years_to_sample=seq(2,50,by=1)
-        y=lapply(X=years_to_sample,FUN = calculate_power,phi =0.5,r=parameter_combinations$r[i],sigma = parameter_combinations$sigma[i],trials=100)
-        if (length(tail(which(y<0.8),1))>0){
-        parameter_combinations$min_time_required[i] = years_to_sample[tail(which(y<0.8),1)] + 1
-        }
-        #print(i)
-      }
-
+      #save(parameter_combinations,file='../analysis-outputs/min_time_vs_pop_var.Rdata')
+      load('../analysis-outputs/min_time_vs_pop_var.Rdata')
       plot(parameter_combinations$sigma,parameter_combinations$min_time_required,type='l',lwd=2,las=1,ylab='',xlab='',cex.axis=1.2,ylim=c(0,50))
@@
 We first take all possible contiguous subsamples of each time series.
-For example, a time series of 40 years would have 39 possible contiguous subsamples of length 2, 38 possible contiguous subsamples of length 3, and continuing until 1 possible contiguous subsample of length 40 [@Gerber1999, @Giron-Nava2017].
+For example, a time series of 40 years would have 39 possible contiguous subsamples of length 2, 38 possible contiguous subsamples of length 3, and continuing until 1 possible contiguous subsample of length 40 [@Gerber1999; @Giron-Nava2017].
 Then we determined the $T_{min}$ required to be confident (i.e.
@@
 Then, we determine the fraction of subsamples of a particular length that have estimated slope coefficients which are statistically different from zero given the long-term, or ``true", time series also has a significant slope.
-This is a measure of statistical power (often denoted as $1- \beta$, where $\beta$ is the probability of a type II error).
+This is a measure of statistical power (often denoted as $1 - \beta$, where $\beta$ is the probability of a type II error).
 Lastly, we determine which subsample length is required to guarantee we have above a certain threshold of statistical power, which we will assume is 0.8 by convention (citation).
-<!--In Figure \ref{fig:empirical_approach_example}b, statistical power increases with more time sampled until our desired power of 0.8 at 16 years.
-Therefore, 16 years would be our estimate for $T_{min}$.-->
+
+<!--
+## Questions addressed in the appendix
 
 We also examine how minimum time series calculations changed depending on the question of interest.
-In the supplementary material, we show how to make similar calculations to determine the minimum time series required to estimate overall population growth rate (Fig.
-\ref{fig:growth_rate}), the variability in population size, and future extinction risk.
-Again, we took subsamples of time series data to determine $T_{min}$ for each of these questions.
-
-<!--Question 2: How sensitive are IUCN assessments to time series trends?
-The IUCN sets threat statuses for a variety of species.
-Importaly, they have a threshold of 10 years or 3 generation.
-Therefore, we subsampled all possible 10-year, or greater, time series for each population.
-For example, species X has been sampled for 40 years.
-This would lead to XX possible 10-year time series.
-We then determined what percentage of these 10 year, or greater, time series would lead to the wrong conculsions compared to the full time series.
-
-Question 3: How long does a population need to be sampled to capture the variability in population size?
-
-Populations vary from year-to-year for several reasons including-->
+In the supplementary material, we use similar calculations to determine the minimum time series required to estimate overall population growth rate (Fig.
+\ref{fig:growth_rate}).
+We also explore how the method can be applied to more biologically-realistic population models (Fig.
+\ref{fig:shark_example}.
+More specifically, we determine the minimum time required to estimate long-term population trends using a stochastic, age-structured model of lemon shark population dynamics in the Bahamas [@White2014].
+Lastly, we determine the minimum time required to estimate long-term trends according to generalized additive models, instead of the simple linear models used here.
+-->
 
 # Results
-<!--
-- summary of findings
-- theoretical approach
-  - example plot
-- empirical approach
-  - plots: distribution of min time, what min time correlations with
-  - walk through one empirical example (maybe in appendix)
--->
 
@@
 
-## Questions in the appendix
+## Questions in the supplmentary material
 
-Describe some of the results that we include in the appendix
+In the supplmentary material, we explore several additional questions related to statistical power and estimating $T_{min}$.
+First, we showed that other metrics besides estimating the slope coefficient from linear regression could be used.
+More specifically, we show how the same approach desribed here can be used to determine the minimum time required to estimate a population's geometric growth rate (Figs.
+\ref{fig:growth_rate},\ref{fig:min_time_growth_dist}).
+We also explored how the simulation approach can be applied to more biologically-realistic population models (Fig.
+\ref{fig:shark_example}).
+More specifically, we determine the minimum time required to estimate long-term population trends using a stochastic, age-structured model of lemon shark population dynamics in the Bahamas [@White2014].
+We found that over 27 years of continous monitoring were needed in this particular scenario (Fig.
+\ref{fig:shark_example}.
+Lastly, we determine the minimum time required to estimate long-term trends according to generalized additive models, instead of the simple linear models used here (Fig.
+\ref{fig:gam_example}).
+We found that although we obtain a similar distribution of minimum times required for GAMs, the minimum time required for GAMs is on average shorter than for linear regression (Fig.
+\ref{fig:min_time_dist_gam}).
+We expand on these results in the supplementary material.
 
@@
 
-<!-- 
-- Findings and how they relate to past work
-- Extensions to other systems. Useful to use similar species
-- Biological variables not good predictors, but we also used all long-lived vertebrate species
-- Other questions to ask and results in appendix
-
-- Future recommendations to others and future work
-- implications
-- conclusion paragraph
-
-The IUCN has an 10 year/3 generation rule...
-If we apply this rule to the species in our database we see that approximately `r 100*sum(pop_info$min_time_for_power>pop_info$IUCN)/nrow(pop_info)`% of populations need a longer minimum time series than that required by IUCN.
--->
-
 We explored two different approaches to estimate the minimum time series length required to address a particular question of interest.
 We specifically asked, what is the minimum time series length required to determine long-term changes in abundance using linear regression?
-The simulation-based approach has been suggested by others, especially in situations more complicated than that suited for classic power analysis [@Gerrodette1987, Johnson2015].
+The simulation-based approach has been suggested by others, especially in situations more complicated than that suited for classic power analysis [@Gerrodette1987;@Johnson2015;@Bolker2008].
 Our simulations support past work that longer time series are needed when the trend strength (i.e.
@@
 
-Here, we focus on a empirical approach to estimate the minimum time series length requried to assess changes in abundance over time.
+Here, we focus on an empirical approach to estimate the minimum time series length requried to assess changes in abundance over time.
 We examined long time series and then subsampled each to determine the minimum time required to achieve a desired level of significance and power for linear regression.
-Statistical power is important as it provides on information as to the neccessary samples required to dertermine a significant trend [@Legg2006].
+Statistical power is important as it provides on information as to the neccessary samples required to determine a significant trend [@Legg2006].
 We found that for `r nrow(pop_info)` populations, at least 10-15 years of continuous monitoring were neccessary (Fig.
@@
 This timeframe is in line with past work on a short-lived snail species [@Rueda-Cediel2015] and a long-lived whale species [@Gerber1999].
-In line with theoretical predictions (citations), we also found $T_{min}$ was strongly correlated with the trend strength, variability in population size, and temporal autocorrelation (Fig.
+In line with theoretical predictions [@Rhodes2011], we also found $T_{min}$ was strongly correlated with the trend strength, variability in population size, and temporal autocorrelation (Fig.
 \ref{fig:correlates}).
-Against our prior expectations, we also found that $T_{min}$ did not correlate with any biological variables of interest (cite figure or supp mat).
+Contrary to our prior expectations, we also found that $T_{min}$ did not correlate with any biological variables of interest (Fig.
+\ref{fig:biological_correlates}).
 We initially hypothesized that species with longer lifespans or generation times may require a longer sampling period.
-Our result could have result of at least two effects.
+Our result could have been a result of at least two effects.
 First, the database we used may not include a diverse enough set of species with different life history traits.
-Second, the question we poised, whether a population is increasing or decreases, is only worried about population surveys.
-Therefore, biological variables may be more important if the question itself was different.
-ANOTHER SENTENCE HERE
+Second, the question we poised, whether a population is increasing or decreasing, was specifically concerned with trends in population density over time.
+Therefore, life-history characteristics may be more important for other questions (e.g.
+extinction risk) where estimates minimum time required are needed.
 
-There has also been interest in determining the minimum time series length in the bird community [@Hatch2003, @Bart2004, @Seavy2007].
-@Hatch2003 used seabird monitoring data to estimate the minimum time needed to achieve appropriate levels of power.
+There has been interest in determining the minimum time series length in the bird research community [@Hatch2003; @Bart2004; @Seavy2007].
+For example, @Hatch2003 used seabird monitoring data to estimate the minimum time needed to achieve appropriate levels of power.
 He found that the minimum time required ranged between 11 and 69 years depending on species, trend strength, and study design.
@@
 They used 24 years of census data on Red-tailed Tropicbirds (\emph{Phaethon rubricauda}) in Hawaii and showed that to detect a 50% decline over 10 years almost always resulting in high statistical power (above 0.8).
-Therefore, they cautioned against only using power analyses to design monitoring schemes and instead argued for precision based metrics ..
+Therefore, they cautioned against only using power analyses to design monitoring schemes and instead argued for metrics that would increase precision.
+For example, @Seavy2007 suggest improving randomization, reducing bias, and increases detection probability when designing and evaluating monitoring programs.
 
-An important related question, is the optimal allocation of sampling effort in space versus time (citations).
+An important related question, is the optimal allocation of sampling effort in space versus time.
 In a theoretical investigation of this question, @Rhodes2011 found that the optimal allocation of sampling depended strongly on temporal and spatial autocorrelation.
 If spatial population dynamics were highly correlated, then it was better to sample more temporally, and vice versa.
-Our work supports this idea as populations with strong temporal autocorrelation needed less years of sampling (cite figure here).
+Our work supports this idea as populations with strong temporal autocorrelation needed less years of sampling (Fig.
+\ref{fig:correlates}).
 @Morrison2008 also studied the optimal allocation of sampling effort in space versus time, but used emprical data from a long-term census of the collared pika (\emph{Ochotona collaris}) found in the Yukon.
 They estimated long-term growth rates among three subpopulations over a 10-year period.
-They found that censuses less than 5 years may be misleading and that extrapolating from one population to another, even when nearby, may be difficult.
-Therefore, they argued more sampling should be conducted spatially than temporally [@Morrison2008].
-
-## Other questions addressed
-
-In the appendix we asked two additional questions to make estimates of $T_{min}$. First, we asked how many years of sampling are neccessary to estimate a geometric growth rate?
+They found that censuses less than 5 years may be misleading and that extrapolating from one population to another, even when nearby geographically, may be untenable.
 
@@
 Our work has some limitations in determining the minimum time series length required.
-First, $T_{min}$ depends strongly on the question of interest.
-For our empirical approach, the subsampling of the full time series allows for estimates of power, but the individual subsamples are clearly not independent of one another.
-In addition, estimates of $T_{min}$ depend on chosen values of $\alpha$ and $\beta$ (see appendix).
-<!-- include sentence on how T_min depends on these parameters --> In an ideal world, we would build a specific population model parameterized for each population of interest.
-Then, model simulations could be used to estimate the minimum time series reuqired to address a specific question of interest.
-Clearly, this is not always practical, especially if running analyses for a wide array of species as we do here.
-In addition, our statistical models suggest that $T_{min}$ does not correlate with any biological variables of interest, at least for the question of linear regression.
+First, $T_{min}$ is particular to the specific question of interest.
+For example, in the supplementary material, we used generalized additive models to detect population trends and came up with a similar, but quantivaely different value for $T_{min}$ (Fig.
+\ref{fig:min_time_dist_gam}).
+An additional complication is that for our empirical approach, the subsampling of the full time series allows for estimates of power, but the individual subsamples are clearly not independent of one another.
+Further, estimates of $T_{min}$ depend on chosen values of $\alpha$ and $\beta$ (Fig.
+\ref{fig:min_time_vs_alpha_beta}).
+In an ideal setting, we would build a specific population model parameterized for each population of interest.
+Then, model simulations could be used to estimate the minimum time series required to address each specific question of interest.
+Clearly, this is not always practical, especially if conducting analyses for a wide array of species as we do here.
+In addition, our statistical models suggest that $T_{min}$ does not correlate with any biological variables of interest, at least for the question of linear regression (Fig.
+\ref{fig:biological_correlates}).
 Therefore, it is not possible to use these results to predict $T_{min}$ for another population, even if the population is of a species with a similar life-history to one in our database.
@@
 
-We use a database of `r nrow(pop_info)` populations to determine the minimum time series length required.
+We use a database of `r nrow(pop_info)` populations to determine the minimum time series length required to determine population trends.
+This goes beyond previous work that either focused on theoretical investigations or a limited number of species.
 We show that to identify long-term changes in abundance, 10-15 years of continuos monitoring are required (Fig.
 \ref{fig:min_time_dist}).
-In line with theoretical predictions, we also show that $T_{min}$ is strongly correlated with the overall trend in abundance, variability in abundance, and the temporal autocorrelation (Fig.
+In line with theoretical predictions [@Gerrodette1987], we also show that $T_{min}$ is strongly correlated with the long-term population trend (i.e.
+rate of increase), variability in population size, and the temporal autocorrelation (Fig.
 \ref{fig:correlates}).
@@
 This stresses the importance of long-term monitoring programs.
-<!--(where do IUCN recommendations come from?)--> From both a scientific and management perspective estimates of $T_{min}$ are important.
-If a time series is too short, we are likely to make incorrect conclusions about long-term trends.
+From both a scientific and management perspective estimates of $T_{min}$ are important.
+If a time series is too short we lack statistical power to reliably detect long-term population trends.
 In addition, a time series that is too long may be a poor use of already limited funds [@Gerber1999].
-Future work can examine other species, with a wider range of life history characteristics.
-In addition, similar approaches can be used to determine the minimum time series length required to address other questions.
+Future work should examine other species, with a wider range of life history characteristics.
+In addition, similar approaches can be used to determine the minimum time series length required to address additional questions of interest.
 
@@
 
-In the supporting material, we provide an expanded methods sections, additional figures, and additional analyses of other time series questions.
+In the supporting material, we provide an expanded methods sections, additional figures, minimum time calculations for determining exponetial growth, simulations with a more complicated population model, minimum time calculations for determining long-term growth rates, and the use of generalized additive models to identify population trends.
 All code and data can be found at www.github.com/erwhite1
@@
 We would like to thank members of the Ecological Theory group at the University of California, Davis for their insight.
-We would also like to thank X anonymous reviewers for their helpful comments.
+We would also like to thank T.
+Dallas and E.
+Malcolm-White<!--X anonymous reviewers -->for their helpful comments.
 
@@
 \centering
-\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-2-1.pdf}
 \caption{(a) Population size of Bigeye tuna (\emph{Thunnus obesus}) over
@@
 \end{figure}
+
+\begin{figure}[htbp]
+\centering
+\caption{Output of generalized linear model with a Poisson error
+structure for predicting the minimum time required with explanatory
+variables of the absolute value of the slope coefficient (or trend
+strength), temporal autocorrelation, and variability in population
+size.\label{fig:poisson_model}}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\caption{Mimimum time required versus (a) generation length (years), (b)
+litter size (n), (c) adult body mass (grams), (d) maximum longevity
+(years), (e) egg mass (grams), and (f) incubation (days). The lines in
+each plot represent the best fit line from linear
+regression.\label{fig:biological_correlates}}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\caption{(a) Minimum time required to estimate change in abundance for
+species class, (b) long-term trend (estimated slope coefficient) by
+species class, (c) interannual variability in population size by species
+class, and (d) temporal autocorrelation by species
+class.\label{fig:class}}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\caption{Minimum time required to assess long-term trends in abundance
+for values of statistical significance (\(\alpha\)) and power
+(\(1-\beta\)).\label{fig:min_time_vs_alpha_beta}}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\caption{Distribution of the minimum time required in order to detect a
+significant trend (at the 0.05 level) in log(abundance) given power of
+0.8.\label{fig:min_time_dist_log_pop}}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\caption{Statistical power for different length of time series
+simulations for a lemon shark population in Bimini,
+Bahamas.\label{fig:shark_example}}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\caption{Example of calculating minimum time required for growth rate
+estimation. (a) European herring gull (\emph{Larus argentatus}) scaled
+density over time, (b) mean and standard deviation of growth rate for
+subsamples of entire time series, and (c) the percent error between mean
+estimated growth rate and the true long-term growth rate. The vertical
+bar denotes the minimum time required to estimate growth rate within
+20\% error.\label{fig:growth_rate}}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\caption{Histogram of the minimum time required in order to estimate the
+long-term growth rate within 20\%
+error.\label{fig:min_time_growth_dist}}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\caption{(a) Time series for Bigeye tuna (\emph{Thunnus obesus}) with
+corresponding fitted GAM model in red and (b) statistical power as a
+function of the number of years sampled. The horizontal ine at 0.8
+indicates the minimum threshold for statistical power and the vertical
+line denotes the minimum time required to achieve 0.8 statistical
+power.\label{fig:gam_example}}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\caption{Distribution of the minimum time required in order to detect a
+significant trend (at the 0.05 level) in abundance according to a GAM
+model given statistical power of 0.8. The smoothing parameter was set to
+3 for each population.\label{fig:min_time_dist_gam}}
+\end{figure}

### 2017-07-07 writing/manuscript_draft.tex
added citations and modified code for figure 1

@@
 approaches to address this question: through simulations and through an
-empirical approach. I specifically determine the minimum time series
+empirical approach. We specifically determine the minimum time series
 length required to estimate significant increases or decreases in
-population abundance. Importantly, I examine the ability to detect a
+population abundance. Importantly, We examine the ability to detect a
 trend with a set level of statistical power, which is often neglected in
@@
 of modern ecological research and conservation biology (Magurran et al.
-2010). long-term data is necessary to both understand population
-dynamics and to assess extinction risk of species. Even though many time
-series may now be considered ``long-term'', most are still short
-(i.e.~less than X years). \emph{I could have statistic on how long is
-considered ``long-term'' in papers}. Time series are typically short for
-a variety of reasons. They are often coupled with an experiment, which
-may only last a couple of years. In addition, short funding cycles make
-it difficult to examine populations over longer periods of time
-(citations). There are, of course, notable exceptions including the Isle
-Royale study, Iowa corn study, cod, lynx-hair, and many others
-(citations).
+2010; Hughes et al. 2017). Long-term data is necessary to both
+understand population dynamics and to assess extinction risk of species.
+Even though many time series may now be considered ``long-term'', most
+are still short. Time series are typically short for a variety of
+reasons. They are often coupled with an experiment, which may only last
+a couple of years. In addition, short funding cycles make it difficult
+to examine populations over longer periods of time (Hughes et al.
+2017){]}. There are, of course, notable exceptions where long-term
+monitoring has occured, for example the continuous plankton recorder
+(Giron-Nava et al. 2017).
 
@@
 be done with experimental data, we can relax the definition of
-statistical power (cite Bolker). Instead of using classic statistical
+statistical power (Bolker 2008). Instead of using classic statistical
 tests (e.g.~t-test), and corresponding formulas for determinig sample
-size, simulation approaches can be used (Gerrodette 1987). Simple models
-can be simulated with parameter values corresponding to a population of
-interest. Out of a large number of simulations, power can be calculated
-as the fraction of simulations that meet some criteria. The specific
-criteria depend on the question at hand. Then, time series simulations
-of different lengths are run to determine the minimum time series length
-required to be confident that you will be able to answer your question
-(cite Bolker). In addition to using simulated data, empirical time
-series can also be used. We usually do not have multiple replicates of
-similar populations, but we can subsample an empirical time series
-(Gerber, DeMaster, and Kareiva 1999). Subsamples of different lengths
-can then be evaluated to see which fraction of subsamples meet some
-criteria, again a measure of statistical power. Similar to the
-simulation approach, we can use these measures of power to determine the
-minimum time series required for a particular question of interest.
+size, simulation approaches can be used (Gerrodette 1987, Bolker2008).
+Simple models can be simulated with parameter values corresponding to a
+population of interest. Out of a large number of simulations, power can
+be calculated as the fraction of simulations that meet some criteria.
+The specific criteria depend on the question at hand. Then, time series
+simulations of different lengths are run to determine the minimum time
+series length required to be confident that you will be able to answer
+your question (Bolker 2008). In addition to using simulated data,
+empirical time series can also be used. We usually do not have multiple
+replicates of similar populations, but we can subsample an empirical
+time series (Gerber, DeMaster, and Kareiva 1999). Subsamples of
+different lengths can then be evaluated to see which fraction of
+subsamples meet some criteria, again a measure of statistical power.
+Similar to the simulation approach, we can use these measures of power
+to determine the minimum time series required for a particular question
+of interest.
 
@@
 
+\begin{figure}[htbp]
+\centering
+\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-1-1.pdf}
+\caption{(a) example of simulated time series for 40 time periods and
+(b) statistical power versus the simulated time series length. The
+horizontal, dashed line is our desired statistical power of 0.8. The
+vertical, dashed line is the minimum time required to achieve our
+desired statistical power. (c) Minimum time required for different
+values of the trend strength (\(r\)) and (d) minimum time required for
+different levels of population variability
+(\(\sigma\)).\label{fig:theoretical_approach}}
+\end{figure}
+
 \subsection{Data source}\label{data-source}
@@
 and continuing until 1 possible contiguous subsample of length 40
-(Gerber, DeMaster, and Kareiva 1999, Giron-Nava et al. (2017)). Then we
+(Gerber, DeMaster, and Kareiva 1999; Giron-Nava et al. 2017). Then we
 determined the \(T_{min}\) required to be confident (i.e.~have high
@@
 series also has a significant slope. This is a measure of statistical
-power (often denoted as \(1- \beta\), where \(\beta\) is the probability
-of a type II error). Lastly, we determine which subsample length is
-required to guarantee we have above a certain threshold of statistical
-power, which we will assume is 0.8 by convention (citation).
-
-We also examine how minimum time series calculations changed depending
-on the question of interest. In the supplementary material, we show how
-to make similar calculations to determine the minimum time series
-required to estimate overall population growth rate (Fig.
-\ref{fig:growth_rate}), the variability in population size, and future
-extinction risk. Again, we took subsamples of time series data to
-determine \(T_{min}\) for each of these questions.
+power (often denoted as \(1 - \beta\), where \(\beta\) is the
+probability of a type II error). Lastly, we determine which subsample
+length is required to guarantee we have above a certain threshold of
+statistical power, which we will assume is 0.8 by convention (citation).
 
@@
 
-\subsection{Questions in the appendix}\label{questions-in-the-appendix}
+\subsection{Questions in the supplmentary
+material}\label{questions-in-the-supplmentary-material}
 
-Describe some of the results that we include in the appendix
+In the supplmentary material, we explore several additional questions
+related to statistical power and estimating \(T_{min}\). First, we
+showed that other metrics besides estimating the slope coefficient from
+linear regression could be used. More specifically, we show how the same
+approach desribed here can be used to determine the minimum time
+required to estimate a population's geometric growth rate (Figs.
+\ref{fig:growth_rate},\ref{fig:min_time_growth_dist}). We also explored
+how the simulation approach can be applied to more
+biologically-realistic population models (Fig. \ref{fig:shark_example}).
+More specifically, we determine the minimum time required to estimate
+long-term population trends using a stochastic, age-structured model of
+lemon shark population dynamics in the Bahamas (White, Nagy, and Gruber
+2014). We found that over 27 years of continous monitoring were needed
+in this particular scenario (Fig. \ref{fig:shark_example}. Lastly, we
+determine the minimum time required to estimate long-term trends
+according to generalized additive models, instead of the simple linear
+models used here (Fig. \ref{fig:gam_example}). We found that although we
+obtain a similar distribution of minimum times required for GAMs, the
+minimum time required for GAMs is on average shorter than for linear
+regression (Fig. \ref{fig:min_time_dist_gam}). We expand on these
+results in the supplementary material.
 
@@
 situations more complicated than that suited for classic power analysis
-(Gerrodette 1987, Johnson2015). Our simulations support past work that
-longer time series are needed when the trend strength (i.e.~rate of
-increase or decrease) is weak or when population variability is high
-(Gerrodette 1987).
+(Gerrodette 1987; P. C. Johnson et al. 2015; Bolker 2008). Our
+simulations support past work that longer time series are needed when
+the trend strength (i.e.~rate of increase or decrease) is weak or when
+population variability is high (Gerrodette 1987).
 
-Here, we focus on a empirical approach to estimate the minimum time
+Here, we focus on an empirical approach to estimate the minimum time
 series length requried to assess changes in abundance over time. We
@@
 provides on information as to the neccessary samples required to
-dertermine a significant trend (Legg and Nagy 2006). We found that for
+determine a significant trend (Legg and Nagy 2006). We found that for
 878 populations, at least 10-15 years of continuous monitoring were
@@
 and a long-lived whale species (Gerber, DeMaster, and Kareiva 1999). In
-line with theoretical predictions (citations), we also found \(T_{min}\)
-was strongly correlated with the trend strength, variability in
-population size, and temporal autocorrelation (Fig.
-\ref{fig:correlates}). Against our prior expectations, we also found
+line with theoretical predictions (Rhodes and Jonzen 2011), we also
+found \(T_{min}\) was strongly correlated with the trend strength,
+variability in population size, and temporal autocorrelation (Fig.
+\ref{fig:correlates}). Contrary to our prior expectations, we also found
 that \(T_{min}\) did not correlate with any biological variables of
-interest (cite figure or supp mat). We initially hypothesized that
-species with longer lifespans or generation times may require a longer
-sampling period. Our result could have result of at least two effects.
-First, the database we used may not include a diverse enough set of
-species with different life history traits. Second, the question we
-poised, whether a population is increasing or decreases, is only worried
-about population surveys. Therefore, biological variables may be more
-important if the question itself was different. ANOTHER SENTENCE HERE
+interest (Fig. \ref{fig:biological_correlates}). We initially
+hypothesized that species with longer lifespans or generation times may
+require a longer sampling period. Our result could have been a result of
+at least two effects. First, the database we used may not include a
+diverse enough set of species with different life history traits.
+Second, the question we poised, whether a population is increasing or
+decreasing, was specifically concerned with trends in population density
+over time. Therefore, life-history characteristics may be more important
+for other questions (e.g.~extinction risk) where estimates minimum time
+required are needed.
 
-There has also been interest in determining the minimum time series
-length in the bird community (Hatch 2003, ({\textbf{???}}), Seavy and
-Reynolds (2007)). Hatch (2003) used seabird monitoring data to estimate
-the minimum time needed to achieve appropriate levels of power. He found
-that the minimum time required ranged between 11 and 69 years depending
-on species, trend strength, and study design. Similarly,
-({\textbf{???}}) recommended a standard of 80\% power to detect a 50\%
+There has been interest in determining the minimum time series length in
+the bird research community (Hatch 2003; Bart et al. 2004; Seavy and
+Reynolds 2007). For example, Hatch (2003) used seabird monitoring data
+to estimate the minimum time needed to achieve appropriate levels of
+power. He found that the minimum time required ranged between 11 and 69
+years depending on species, trend strength, and study design. Similarly,
+Bart et al. (2004) recommended a standard of 80\% power to detect a 50\%
 decline over 20 years in landbird populations. They found that only
@@
 (above 0.8). Therefore, they cautioned against only using power analyses
-to design monitoring schemes and instead argued for precision based
-metrics ..
+to design monitoring schemes and instead argued for metrics that would
+increase precision. For example, Seavy and Reynolds (2007) suggest
+improving randomization, reducing bias, and increases detection
+probability when designing and evaluating monitoring programs.
 
 An important related question, is the optimal allocation of sampling
-effort in space versus time (citations). In a theoretical investigation
-of this question, Rhodes and Jonzen (2011) found that the optimal
-allocation of sampling depended strongly on temporal and spatial
-autocorrelation. If spatial population dynamics were highly correlated,
-then it was better to sample more temporally, and vice versa. Our work
-supports this idea as populations with strong temporal autocorrelation
-needed less years of sampling (cite figure here). Morrison and Hik
-(2008) also studied the optimal allocation of sampling effort in space
-versus time, but used emprical data from a long-term census of the
-collared pika (\emph{Ochotona collaris}) found in the Yukon. They
-estimated long-term growth rates among three subpopulations over a
-10-year period. They found that censuses less than 5 years may be
-misleading and that extrapolating from one population to another, even
-when nearby, may be difficult. Therefore, they argued more sampling
-should be conducted spatially than temporally (Morrison and Hik 2008).
-
-\subsection{Other questions addressed}\label{other-questions-addressed}
-
-In the appendix we asked two additional questions to make estimates of
-\(T_{min}\). First, we asked how many years of sampling are neccessary
-to estimate a geometric growth rate?
+effort in space versus time. In a theoretical investigation of this
+question, Rhodes and Jonzen (2011) found that the optimal allocation of
+sampling depended strongly on temporal and spatial autocorrelation. If
+spatial population dynamics were highly correlated, then it was better
+to sample more temporally, and vice versa. Our work supports this idea
+as populations with strong temporal autocorrelation needed less years of
+sampling (Fig. \ref{fig:correlates}). Morrison and Hik (2008) also
+studied the optimal allocation of sampling effort in space versus time,
+but used emprical data from a long-term census of the collared pika
+(\emph{Ochotona collaris}) found in the Yukon. They estimated long-term
+growth rates among three subpopulations over a 10-year period. They
+found that censuses less than 5 years may be misleading and that
+extrapolating from one population to another, even when nearby
+geographically, may be untenable.
 
@@
 Our work has some limitations in determining the minimum time series
-length required. First, \(T_{min}\) depends strongly on the question of
-interest. For our empirical approach, the subsampling of the full time
-series allows for estimates of power, but the individual subsamples are
-clearly not independent of one another. In addition, estimates of
-\(T_{min}\) depend on chosen values of \(\alpha\) and \(\beta\) (see
-appendix). In an ideal world, we would build a specific population model
-parameterized for each population of interest. Then, model simulations
-could be used to estimate the minimum time series reuqired to address a
-specific question of interest. Clearly, this is not always practical,
-especially if running analyses for a wide array of species as we do
-here. In addition, our statistical models suggest that \(T_{min}\) does
-not correlate with any biological variables of interest, at least for
-the question of linear regression. Therefore, it is not possible to use
-these results to predict \(T_{min}\) for another population, even if the
-population is of a species with a similar life-history to one in our
+length required. First, \(T_{min}\) is particular to the specific
+question of interest. For example, in the supplementary material, we
+used generalized additive models to detect population trends and came up
+with a similar, but quantivaely different value for \(T_{min}\) (Fig.
+\ref{fig:min_time_dist_gam}). An additional complication is that for our
+empirical approach, the subsampling of the full time series allows for
+estimates of power, but the individual subsamples are clearly not
+independent of one another. Further, estimates of \(T_{min}\) depend on
+chosen values of \(\alpha\) and \(\beta\) (Fig.
+\ref{fig:min_time_vs_alpha_beta}). In an ideal setting, we would build a
+specific population model parameterized for each population of interest.
+Then, model simulations could be used to estimate the minimum time
+series required to address each specific question of interest. Clearly,
+this is not always practical, especially if conducting analyses for a
+wide array of species as we do here. In addition, our statistical models
+suggest that \(T_{min}\) does not correlate with any biological
+variables of interest, at least for the question of linear regression
+(Fig. \ref{fig:biological_correlates}). Therefore, it is not possible to
+use these results to predict \(T_{min}\) for another population, even if
+the population is of a species with a similar life-history to one in our
 database.
@@
 We use a database of 878 populations to determine the minimum time
-series length required. We show that to identify long-term changes in
+series length required to determine population trends. This goes beyond
+previous work that either focused on theoretical investigations or a
+limited number of species. We show that to identify long-term changes in
 abundance, 10-15 years of continuos monitoring are required (Fig.
-\ref{fig:min_time_dist}). In line with theoretical predictions, we also
-show that \(T_{min}\) is strongly correlated with the overall trend in
-abundance, variability in abundance, and the temporal autocorrelation
-(Fig. \ref{fig:correlates}). Our work implies that for many populations,
+\ref{fig:min_time_dist}). In line with theoretical predictions
+(Gerrodette 1987), we also show that \(T_{min}\) is strongly correlated
+with the long-term population trend (i.e.~rate of increase), variability
+in population size, and the temporal autocorrelation (Fig.
+\ref{fig:correlates}). Our work implies that for many populations,
 census data less than 10-15 years are probably not reliable. This
@@
 scientific and management perspective estimates of \(T_{min}\) are
-important. If a time series is too short, we are likely to make
-incorrect conclusions about long-term trends. In addition, a time series
+important. If a time series is too short we lack statistical power to
+reliably detect long-term population trends. In addition, a time series
 that is too long may be a poor use of already limited funds (Gerber,
-DeMaster, and Kareiva 1999). Future work can examine other species, with
-a wider range of life history characteristics. In addition, similar
+DeMaster, and Kareiva 1999). Future work should examine other species,
+with a wider range of life history characteristics. In addition, similar
 approaches can be used to determine the minimum time series length
-required to address other questions.
+required to address additional questions of interest.
 
@@
 In the supporting material, we provide an expanded methods sections,
-additional figures, and additional analyses of other time series
-questions. All code and data can be found at www.github.com/erwhite1
+additional figures, minimum time calculations for determining exponetial
+growth, simulations with a more complicated population model, minimum
+time calculations for determining long-term growth rates, and the use of
+generalized additive models to identify population trends. All code and
+data can be found at www.github.com/erwhite1
 
@@
 group at the University of California, Davis for their insight. We would
-also like to thank X anonymous reviewers for their helpful comments.
+also like to thank T. Dallas and E. Malcolm-Whitefor their helpful
+comments.
 
@@
 \hypertarget{refs}{}
+\hypertarget{ref-Bart2004}{}
+Bart, Jonathon, Kenneth P. Burnham, Erica H. Dunn, Charles M. Francis,
+and C. John Ralph. 2004. ``Goals and strategies for estimating trends in
+landbird abundance.'' \emph{Journal of Wildlife Management} 68 (3):
+611--26.
+
+\hypertarget{ref-Bolker2008}{}
+Bolker, Benjamin M. 2008. \emph{Ecological Models and Data in R}. 1st
+ed. Princeton, New Jersey: Princeton University Press.
+
 \hypertarget{ref-Gerber1999}{}
@@
 111: 317--29.
+
+\hypertarget{ref-Hughes2017}{}
+Hughes, Brent B, Rodrigo Beas-luna, Allison K Barner, Kimberly Brewitt,
+Daniel R Brumbaugh, Elizabeth B. Cerny-Chipman, Sarah L. Close, et al.
+2017. ``Long-term studies contribute disproportionately to ecology and
+policy.'' \emph{BioScience} 67 (3): 271--81.
+doi:\href{https://doi.org/10.1093/biosci/biw185}{10.1093/biosci/biw185}.
 
@@
 Legg, Colin J, and Laszlo Nagy. 2006. ``Why most conservation monitoring
-is , but need not be , a waste of time.'' \emph{Journal of Environmental
+is, but need not be, a waste of time.'' \emph{Journal of Environmental
 Management} 78: 194--99.
@@
 \hypertarget{ref-Morrison2008}{}
-Morrison, Sharn, and David S. Hik. 2008. ``When? Where? and for how
-long? Census design considerations for an Alpine Lagomorpg, the Collared
+Morrison, Sharn, and David S. Hik. 2008. ``When? Where? And for how
+long? Census design considerations for an Alpine Lagomorph, the Collared
 pika.'' In \emph{Lagomorph Biology}, 103--13. Springer Berlin
@@
 Rueda-Cediel, Pamela, Kurt E Anderson, Tracey J Regan, Janet Franklin,
-and M Regan. 2015. ``Combined Influences of Model Choice , Data Quality
-, and Data Quantity When Estimating Population Trends.'' \emph{PLoSONE}
-10 (7): e0132255.
+and M Regan. 2015. ``Combined influences of model choice, data quality,
+and data quantity when estimating population trends.'' \emph{PLoSONE} 10
+(7): e0132255.
 doi:\href{https://doi.org/10.1371/journal.pone.0132255}{10.1371/journal.pone.0132255}.
@@
 doi:\href{https://doi.org/10.1016/j.biocon.2007.08.007}{10.1016/j.biocon.2007.08.007}.
+
+\hypertarget{ref-White2014}{}
+White, Easton R, John D Nagy, and Samuel H Gruber. 2014. ``Modeling the
+population dynamics of lemon sharks.'' \emph{Biology Direct} 9 (23):
+1--18.
 
@@
 
+\begin{figure}[htbp]
+\centering
+\caption{Output of generalized linear model with a Poisson error
+structure for predicting the minimum time required with explanatory
+variables of the absolute value of the slope coefficient (or trend
+strength), temporal autocorrelation, and variability in population
+size.\label{fig:poisson_model}}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\caption{Mimimum time required versus (a) generation length (years), (b)
+litter size (n), (c) adult body mass (grams), (d) maximum longevity
+(years), (e) egg mass (grams), and (f) incubation (days). The lines in
+each plot represent the best fit line from linear
+regression.\label{fig:biological_correlates}}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\caption{(a) Minimum time required to estimate change in abundance for
+species class, (b) long-term trend (estimated slope coefficient) by
+species class, (c) interannual variability in population size by species
+class, and (d) temporal autocorrelation by species
+class.\label{fig:class}}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\caption{Minimum time required to assess long-term trends in abundance
+for values of statistical significance (\(\alpha\)) and power
+(\(1-\beta\)).\label{fig:min_time_vs_alpha_beta}}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\caption{Distribution of the minimum time required in order to detect a
+significant trend (at the 0.05 level) in log(abundance) given power of
+0.8.\label{fig:min_time_dist_log_pop}}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\caption{Statistical power for different length of time series
+simulations for a lemon shark population in Bimini,
+Bahamas.\label{fig:shark_example}}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\caption{Example of calculating minimum time required for growth rate
+estimation. (a) European herring gull (\emph{Larus argentatus}) scaled
+density over time, (b) mean and standard deviation of growth rate for
+subsamples of entire time series, and (c) the percent error between mean
+estimated growth rate and the true long-term growth rate. The vertical
+bar denotes the minimum time required to estimate growth rate within
+20\% error.\label{fig:growth_rate}}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\caption{Histogram of the minimum time required in order to estimate the
+long-term growth rate within 20\%
+error.\label{fig:min_time_growth_dist}}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\caption{(a) Time series for Bigeye tuna (\emph{Thunnus obesus}) with
+corresponding fitted GAM model in red and (b) statistical power as a
+function of the number of years sampled. The horizontal ine at 0.8
+indicates the minimum threshold for statistical power and the vertical
+line denotes the minimum time required to achieve 0.8 statistical
+power.\label{fig:gam_example}}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\caption{Distribution of the minimum time required in order to detect a
+significant trend (at the 0.05 level) in abundance according to a GAM
+model given statistical power of 0.8. The smoothing parameter was set to
+3 for each population.\label{fig:min_time_dist_gam}}
+\end{figure}
+
 \end{document}

### 2017-07-07 writing/manuscript_draft.Rmd
updated citations and spelling throughout document

@@
 
-Title: Your time series is (probably) too short
+Title: Statistical power and detecting population trends: the need for long-term monitoring programs
 \vspace{7 mm}
@@
 
-To be submitted to: \emph{Biological Conservation} (Short Communication)
+To be submitted to: <!--\emph{Biological Conservation} (Short Communication)-->
 \vspace{3 mm}
@@
 # Abstract
-Long-term census data is necessary to better understand ecological processes and make management decisions.
-However, census data is often expensive, requiring a lot of people-hours and equipment.
+Long-term census data are necessary to better understand ecological processes and make management decisions.
+However, census data are often expensive, requiring a lot of people-hours and equipment.
 Little work has addressed the question, when is a time series of census data long enough to address a question of interest?
-Here, we explore two approaches to address this question: through simulations and through an empirical approach.
+Here, we explore two approaches to address this question: simulations and an empirical approach.
 We specifically determine the minimum time series length required to estimate significant increases or decreases in population abundance.
-Importantly, We examine the ability to detect a trend with a set level of statistical power, which is often neglected in ecological time series analyses.
+Importantly, we examine the ability to detect a trend with a set level of statistical power, which is often neglected in ecological time series analyses.
 Using a simulation approach, we demonstrate how the minimum time series length required increases with weaker trends in abundance and with higher population variability.
@@
 <!--Perhaps surprisingly, the minimum time required did not correlate well with biological explanatory variables, like body size or generation length.--> These results point to the importance of sampling populations over long periods of time.
-<!--Here, We stress the need for verifying a study has enough statistical power to accurately determine long-term changes in population abundance.--> Most studies, especially those less than 10-15 years, are probably underpowered and potentially misleading.
+<!--Here, We stress the need for verifying a study has enough statistical power to accurately determine long-term changes in population abundance.--> Most studies, especially those less than 10-15 years, are probably under-powered and potentially misleading.
 
@@
 Observational studies and population censuses have become a cornerstone of modern ecological research and conservation biology [@Magurran2010; @Hughes2017].
-Long-term data is necessary to both understand population dynamics and to assess extinction risk of species.
+Long-term data are necessary to both understand population dynamics and to assess extinction risk of species.
 Even though many time series may now be considered "long-term", most are still short.
@@
 They are often coupled with an experiment, which may only last a couple of years.
-In addition, short funding cycles make it difficult to examine populations over longer periods of time [@Hughes2017]].
-There are, of course, notable exceptions where long-term monitoring has occured, for example the continuous plankton recorder [@Giron-Nava2017].
+In addition, short funding cycles make it difficult to examine populations over longer periods of time [@Hughes2017].
+There are, of course, notable exceptions where long-term monitoring has occurred, for example the continuous plankton recorder [@Giron-Nava2017].
 
@@
 From a management perspective, it is important to understand when a trend in abundance over time is actually meaningful or not.
-The IUCN Red List Categories and Criteria suggest under Criterian A2, a species qualifies as vulnerable if it has experienced a 30% decline over 10 years, or 3 generations [@IUCN2012].
-For both scientific and management questions, because sampling is typically expensive, we do not want to sample for longer than is neccessary [@Gerber1999].
+The IUCN Red List Categories and Criteria suggest under Criterion A2, a species qualifies as vulnerable if it has experienced a 30% decline over 10 years, or 3 generations [@IUCN2012].
+For both scientific and management questions, because sampling is typically expensive, we do not want to sample for longer than is necessary [@Gerber1999].
 
@@
 Instead of using classic statistical tests (e.g.
-t-test), and corresponding formulas for determinig sample size, simulation approaches can be used [@Gerrodette1987, Bolker2008].
+t-test), and corresponding formulas for determining sample size, simulation approaches can be used [@Gerrodette1987; @Bolker2008; @Johnson2015].
 Simple models can be simulated with parameter values corresponding to a population of interest.
@@
 Using simple populations models, they found that the allocation of effort depends on environmental variation, spatial and temporal autocorrelation, and observer error.
-@Rueda-Cediel2015 also used a modeling approach, but parameterized amodel specifically for a threatened snail, \emph{Tasmaphena lamproides}.
-They found that for this short-lived organim, 15 years was adequate to assess long-term trends in abundance.
+@Rueda-Cediel2015 also used a modeling approach, but parameterized a model specifically for a threatened snail, \emph{Tasmaphena lamproides}.
+They found that for this short-lived organism, 15 years was adequate to assess long-term trends in abundance.
 However, these studies, and other past work, have typically focused either on theoretical aspects of monitoring design or focused on single species.
@@
 In this paper, we use both simulations and a database of empirical time series to determine the minimum number of years required to address various questions.
-More specifically, we estimate the minimum time series length required to asssess long-term changes in abundance via simple linear regression.
-We examine 878 population time series to estimate this minimum time series length required.
-<!-- In this paper, we evaluate the minimal time series length required to answer X questions.
-1) How sensitivive are IUCN asssessments to time series lengths?
-2) What minimum time series length is required to measure the entire time series trend?
-3) How long does a time series need to be to measure the variance in population size (rare event questions)?
-We demonstrate our approach with X case studies and then show results for X species from the Blah Blah Database.
--->
+More specifically, we estimate the minimum time series length required to assess long-term changes in abundance via simple linear regression.
+We examine 878 population time series to estimate this minimum time series length required for detecting population trends.
 
@@
 This is the same approach one might use in sample size calculations for any experimental design too complicated for simple power analyses [@Johnson2015] (and cite Bolker).
-We only briefly discuss this approach as it has been desribed elsewhere (citations).
+We only briefly discuss this approach as it has been described elsewhere (citations).
 Essentially, we take a stochastic population model and simulate it for a number of years repetitively.
@@
 A significant trend in abundance is determined by estimating the significance of the slope parameter from simple linear regression.
-Then, the minimum time required ($T_{min}$) is the number of years required to obtain statistical power above some pre-determined threshold (historically at 0.8).
+Then, the minimum time required ($T_{min}$) is the number of years required to obtain statistical power above some pre-determined threshold (historically at 0.8, @Cohen1992).
 Statistical power of 0.8 would indicate that, if there was a true trend in abundance, we would have a 0.8 probability of detecting the trend.
@@
 The vertical, dashed line is the minimum time required to achieve our desired statistical power.
-(c) Minimum time required for different values of the trend strength ($r$) and (d) minimum time required for different levels of population variability ($\\sigma$).\\label{fig:theoretical_approach}'}
+(c) Minimum time required ($T_{min}$) for simulations with different values of the trend strength ($r$) and (d) different levels of population variability ($\\sigma$).
+In each case, the minimum time required is the minimum number of years to achieve 0.8 statistical power given a significance level of 0.05.\\label{fig:theoretical_approach}'}
 source("../scripts/theoretical-models/calculate_power.R")
@@
   # Plotting commands
-  par(mfrow=c(2,2),oma=c(1,2,0,0.5),mar=c(4,5,0.5,0.5))
+  par(mfrow=c(2,2),oma=c(1,2,0.5,0.5),mar=c(4,5,2.5,0.5))
     matplot(simulations,type='l',col=rgb(0.5,0.5,0.5,0.5),lty=1,ylab='',xlab='',las=1,cex.axis=1.2)
       mtext(text = 'time' ,side = 1,cex = 1.2,line=3)
-      mtext(text = 'population size' ,side = 2,cex = 1.2,line=4)
+      mtext(text = 'population size' ,side = 2,cex = 1.2,line=3.5)
       mtext(text = '(a)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
@@
       mtext(text = 'time sampled' ,side = 1,cex = 1.2,line=3)
-      mtext(text = 'statistical power' ,side = 2,cex = 1.2,line=3.5)
+      mtext(text = 'statistical power' ,side = 2,cex = 1.2,line=3)
       mtext(text = '(b)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
@@
       mtext(text = 'trend strength (slope coefficient)' ,side = 1,cex = 1.2,line=3)
-      mtext(text = 'minimum time required' ,side = 2,cex = 1.2,line=3.5)
+      mtext(text = expression(T[min]) ,side = 2,cex = 1.2,line=3)
       mtext(text = '(c)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
@@
       mtext(text = 'population variability' ,side = 1,cex = 1.2,line=3)
-      mtext(text = 'minimum time required' ,side = 2,cex = 1.2,line=3.5)
+      mtext(text = expression(T[min]) ,side = 2,cex = 1.2,line=3)
       mtext(text = '(d)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
@@
 We use a database of `r length(table(dat$ID))` population time series compiled in [@Keith2015].
-The data is originally from the Global Population Dynamics Database [@GPDD2010].
+The data are originally from the Global Population Dynamics Database [@GPDD2010].
 We filtered out short time series (less than `r pop_info$series_length[1]` years), and those with missing data, and were left with `r nrow(pop_info)` time series.
@@
 For each time series, we also calculated variables of interest like variance in population size, long-term trend in abundance (slope coefficient from simple linear regression), and temporal autocorrelation.
-Lastly, we also calculated the minimum number of years required (10 years or 3 generations) under IUCN Red List Criterian A2 to classify a species as vulnerable [@IUCN2012].
+Lastly, we also calculated the minimum number of years required (10 years or 3 generations) under IUCN Red List Criterion A2 to classify a species as vulnerable [@IUCN2012].
 
@@
 All `r nrow(LPI_pop_info)` populations are within the Aves class.
-In the supplementary material, we explore how the minimum time required does not strongly correlate with any biological characteristics.
+In the supplementary material, we explore how the minimum time required does not strongly correlate with any biological characteristics (Fig.
+\ref{fig:biological_correlates}).
 
@@
 Therefore, we run a linear regression for each subsampled time series.
-Then, we determine the fraction of subsamples of a particular length that have estimated slope coefficients which are statistically different from zero given the long-term, or ``true", time series also has a significant slope.
+Then, we determine the fraction of subsamples of a particular length that have estimated slope coefficients which are statistically different from zero given the long-term, or ``true'', time series also has a significant slope.
 This is a measure of statistical power (often denoted as $1 - \beta$, where $\beta$ is the probability of a type II error).
-Lastly, we determine which subsample length is required to guarantee we have above a certain threshold of statistical power, which we will assume is 0.8 by convention (citation).
+Lastly, we determine which subsample length is required to guarantee we have above a certain threshold of statistical power, which we will assume is 0.8 by convention [@Cohen1992].
 
@@
 We investigated this question through two approaches: simulation-based and empirical analyses.
-We briefly desribe results from the simulation approach and then discuss our emprirical approach.
+We briefly describe results from the simulation approach and then discuss our empirical approach.
 
@@
 
-```{r,echo=F, fig.cap='Minimum time required to estimate change in abundance correlated with (a) trend strength (absolute value of slope coefficient estimated from linear regression), (b) population variance (interannual variability in population size), and (c) temporal autocorrelation.\\label{fig:correlates}'}
+```{r,echo=F, fig.cap='Minimum time required to estimate change in abundance correlated with (a) trend strength (absolute value of slope coefficient estimated from linear regression), (b) population variance (interannual variability in population size), and (c) temporal autocorrelation.\\label{fig:correlates}',fig.height=5}
 
 # Plot minimum time required against trend strength, population variance, and autocorrelation  
-  par(mfrow=c(3,1),mar=c(4,1,1,1),oma=c(0,4.5,0,0.5))
+  par(mfrow=c(3,1),mar=c(4,2,2,1),oma=c(0,4.5,0,0.5))
   plot(abs(pop_info$overall_trend),pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
@@
 body mass, longevity, location).
-None of these explanatory variables were predictive of the minimum time series length required (**supp mat**).
-Some differences if $T_{min}$ did occur between biological class (**supp mat**), but this can be attributed to the amount populations are known to fluctuate (i.e.
+None of these explanatory variables were predictive of the minimum time series length required (Fig.
+\ref{fig:biological_correlates}).
+Some differences if $T_{min}$ did occur between biological class (Fig.
+\ref{fig:class}), but this can be attributed to the amount populations are known to fluctuate (i.e.
 the variability in population size).
@@
 
-In the supplmentary material, we explore several additional questions related to statistical power and estimating $T_{min}$.
+In the supplementary material, we explore several additional questions related to statistical power and estimating $T_{min}$.
 First, we showed that other metrics besides estimating the slope coefficient from linear regression could be used.
-More specifically, we show how the same approach desribed here can be used to determine the minimum time required to estimate a population's geometric growth rate (Figs.
+More specifically, we show how the same approach described here can be used to determine the minimum time required to estimate a population's geometric growth rate (Figs.
 \ref{fig:growth_rate},\ref{fig:min_time_growth_dist}).
@@
 More specifically, we determine the minimum time required to estimate long-term population trends using a stochastic, age-structured model of lemon shark population dynamics in the Bahamas [@White2014].
-We found that over 27 years of continous monitoring were needed in this particular scenario (Fig.
+We found that over 27 years of continuous monitoring were needed in this particular scenario (Fig.
 \ref{fig:shark_example}.
@@
 
-Here, we focus on an empirical approach to estimate the minimum time series length requried to assess changes in abundance over time.
+Here, we focus on an empirical approach to estimate the minimum time series length required to assess changes in abundance over time.
 We examined long time series and then subsampled each to determine the minimum time required to achieve a desired level of significance and power for linear regression.
-Statistical power is important as it provides on information as to the neccessary samples required to determine a significant trend [@Legg2006].
-We found that for `r nrow(pop_info)` populations, at least 10-15 years of continuous monitoring were neccessary (Fig.
+Statistical power is important as it provides on information as to the necessary samples required to determine a significant trend [@Legg2006].
+We found that for `r nrow(pop_info)` populations, at least 10-15 years of continuous monitoring were necessary (Fig.
 \ref{fig:min_time_dist}).
-This timeframe is in line with past work on a short-lived snail species [@Rueda-Cediel2015] and a long-lived whale species [@Gerber1999].
+This time frame is in line with past work on a short-lived snail species [@Rueda-Cediel2015] and a long-lived whale species [@Gerber1999].
 In line with theoretical predictions [@Rhodes2011], we also found $T_{min}$ was strongly correlated with the trend strength, variability in population size, and temporal autocorrelation (Fig.
@@
 First, $T_{min}$ is particular to the specific question of interest.
-For example, in the supplementary material, we used generalized additive models to detect population trends and came up with a similar, but quantivaely different value for $T_{min}$ (Fig.
+For example, in the supplementary material, we used generalized additive models to detect population trends and came up with a similar, but quantitatively different value for $T_{min}$ (Fig.
 \ref{fig:min_time_dist_gam}).
@@
 This goes beyond previous work that either focused on theoretical investigations or a limited number of species.
-We show that to identify long-term changes in abundance, 10-15 years of continuos monitoring are required (Fig.
+We show that to identify long-term changes in abundance, 10-15 years of continuous monitoring are required (Fig.
 \ref{fig:min_time_dist}).
@@
 
-In the supporting material, we provide an expanded methods sections, additional figures, minimum time calculations for determining exponetial growth, simulations with a more complicated population model, minimum time calculations for determining long-term growth rates, and the use of generalized additive models to identify population trends.
+In the supporting material, we provide an expanded methods sections, additional figures, minimum time calculations for determining exponential growth, simulations with a more complicated population model, minimum time calculations for determining long-term growth rates, and the use of generalized additive models to identify population trends.
 All code and data can be found at www.github.com/erwhite1
@@
 \caption{(a) Population size of Bigeye tuna (\emph{Thunnus obesus}) over
-time. The line is the best fit line from linear regresssion. (b)
+time. The line is the best fit line from linear regression. (b)
 Statistical power for different subsets of the time series in panel
@@
 \centering
-\caption{Mimimum time required versus (a) generation length (years), (b)
+\caption{Minimum time required versus (a) generation length (years), (b)
 litter size (n), (c) adult body mass (grams), (d) maximum longevity
@@
 species class, (b) long-term trend (estimated slope coefficient) by
-species class, (c) interannual variability in population size by species
+species class, (c) inter-annual variability in population size by species
 class, and (d) temporal autocorrelation by species
@@
 corresponding fitted GAM model in red and (b) statistical power as a
-function of the number of years sampled. The horizontal ine at 0.8
+function of the number of years sampled. The horizontal line at 0.8
 indicates the minimum threshold for statistical power and the vertical

### 2017-07-07 writing/manuscript_draft.tex
updated citations and spelling throughout document

@@
 
-Title: Your time series is (probably) too short \vspace{7 mm}
+Title: Statistical power and detecting population trends: the need for
+long-term monitoring programs \vspace{7 mm}
 
@@
 
-To be submitted to: \emph{Biological Conservation} (Short Communication)
-\vspace{3 mm}
+To be submitted to: \vspace{3 mm}
 
@@
 
-Long-term census data is necessary to better understand ecological
-processes and make management decisions. However, census data is often
+Long-term census data are necessary to better understand ecological
+processes and make management decisions. However, census data are often
 expensive, requiring a lot of people-hours and equipment. Little work
@@
 enough to address a question of interest? Here, we explore two
-approaches to address this question: through simulations and through an
-empirical approach. We specifically determine the minimum time series
-length required to estimate significant increases or decreases in
-population abundance. Importantly, We examine the ability to detect a
-trend with a set level of statistical power, which is often neglected in
-ecological time series analyses. Using a simulation approach, we
-demonstrate how the minimum time series length required increases with
-weaker trends in abundance and with higher population variability. In
-addition, we examine 868 populations of vertebrate species to determine
-the minimum time required to detect changes in their abundance. We show
-that 10-15 years of continuous monitoring are required in order to
-achieve a high level of statistical power. Similar to
-simulation-approaches, the minimum time required for field census data
-strongly depends on trend strength, population variability, and temporal
-autocorrelation. These results point to the importance of sampling
-populations over long periods of time. Most studies, especially those
-less than 10-15 years, are probably underpowered and potentially
-misleading.
+approaches to address this question: simulations and an empirical
+approach. We specifically determine the minimum time series length
+required to estimate significant increases or decreases in population
+abundance. Importantly, we examine the ability to detect a trend with a
+set level of statistical power, which is often neglected in ecological
+time series analyses. Using a simulation approach, we demonstrate how
+the minimum time series length required increases with weaker trends in
+abundance and with higher population variability. In addition, we
+examine 868 populations of vertebrate species to determine the minimum
+time required to detect changes in their abundance. We show that 10-15
+years of continuous monitoring are required in order to achieve a high
+level of statistical power. Similar to simulation-approaches, the
+minimum time required for field census data strongly depends on trend
+strength, population variability, and temporal autocorrelation. These
+results point to the importance of sampling populations over long
+periods of time. Most studies, especially those less than 10-15 years,
+are probably under-powered and potentially misleading.
 
@@
 of modern ecological research and conservation biology (Magurran et al.
-2010; Hughes et al. 2017). Long-term data is necessary to both
+2010; Hughes et al. 2017). Long-term data are necessary to both
 understand population dynamics and to assess extinction risk of species.
@@
 a couple of years. In addition, short funding cycles make it difficult
-to examine populations over longer periods of time (Hughes et al.
-2017){]}. There are, of course, notable exceptions where long-term
-monitoring has occured, for example the continuous plankton recorder
-(Giron-Nava et al. 2017).
+to examine populations over longer periods of time (Hughes et al. 2017).
+There are, of course, notable exceptions where long-term monitoring has
+occurred, for example the continuous plankton recorder (Giron-Nava et
+al. 2017).
 
@@
 understand when a trend in abundance over time is actually meaningful or
-not. The IUCN Red List Categories and Criteria suggest under Criterian
+not. The IUCN Red List Categories and Criteria suggest under Criterion
 A2, a species qualifies as vulnerable if it has experienced a 30\%
@@
 and management questions, because sampling is typically expensive, we do
-not want to sample for longer than is neccessary (Gerber, DeMaster, and
+not want to sample for longer than is necessary (Gerber, DeMaster, and
 Kareiva 1999).
@@
 statistical power (Bolker 2008). Instead of using classic statistical
-tests (e.g.~t-test), and corresponding formulas for determinig sample
-size, simulation approaches can be used (Gerrodette 1987, Bolker2008).
-Simple models can be simulated with parameter values corresponding to a
-population of interest. Out of a large number of simulations, power can
-be calculated as the fraction of simulations that meet some criteria.
-The specific criteria depend on the question at hand. Then, time series
-simulations of different lengths are run to determine the minimum time
-series length required to be confident that you will be able to answer
-your question (Bolker 2008). In addition to using simulated data,
-empirical time series can also be used. We usually do not have multiple
-replicates of similar populations, but we can subsample an empirical
-time series (Gerber, DeMaster, and Kareiva 1999). Subsamples of
-different lengths can then be evaluated to see which fraction of
-subsamples meet some criteria, again a measure of statistical power.
-Similar to the simulation approach, we can use these measures of power
-to determine the minimum time series required for a particular question
-of interest.
+tests (e.g.~t-test), and corresponding formulas for determining sample
+size, simulation approaches can be used (Gerrodette 1987; Bolker 2008;
+P. C. Johnson et al. 2015). Simple models can be simulated with
+parameter values corresponding to a population of interest. Out of a
+large number of simulations, power can be calculated as the fraction of
+simulations that meet some criteria. The specific criteria depend on the
+question at hand. Then, time series simulations of different lengths are
+run to determine the minimum time series length required to be confident
+that you will be able to answer your question (Bolker 2008). In addition
+to using simulated data, empirical time series can also be used. We
+usually do not have multiple replicates of similar populations, but we
+can subsample an empirical time series (Gerber, DeMaster, and Kareiva
+1999). Subsamples of different lengths can then be evaluated to see
+which fraction of subsamples meet some criteria, again a measure of
+statistical power. Similar to the simulation approach, we can use these
+measures of power to determine the minimum time series required for a
+particular question of interest.
 
@@
 autocorrelation, and observer error. Rueda-Cediel et al. (2015) also
-used a modeling approach, but parameterized amodel specifically for a
+used a modeling approach, but parameterized a model specifically for a
 threatened snail, \emph{Tasmaphena lamproides}. They found that for this
-short-lived organim, 15 years was adequate to assess long-term trends in
-abundance. However, these studies, and other past work, have typically
-focused either on theoretical aspects of monitoring design or focused on
-single species.
+short-lived organism, 15 years was adequate to assess long-term trends
+in abundance. However, these studies, and other past work, have
+typically focused either on theoretical aspects of monitoring design or
+focused on single species.
 
@@
 various questions. More specifically, we estimate the minimum time
-series length required to asssess long-term changes in abundance via
+series length required to assess long-term changes in abundance via
 simple linear regression. We examine 878 population time series to
-estimate this minimum time series length required.
+estimate this minimum time series length required for detecting
+population trends.
 
@@
 Johnson et al. 2015) (and cite Bolker). We only briefly discuss this
-approach as it has been desribed elsewhere (citations). Essentially, we
+approach as it has been described elsewhere (citations). Essentially, we
 take a stochastic population model and simulate it for a number of years
@@
 required to obtain statistical power above some pre-determined threshold
-(historically at 0.8). Statistical power of 0.8 would indicate that, if
-there was a true trend in abundance, we would have a 0.8 probability of
-detecting the trend.
+(historically at 0.8, Cohen (1992)). Statistical power of 0.8 would
+indicate that, if there was a true trend in abundance, we would have a
+0.8 probability of detecting the trend.
 
@@
 vertical, dashed line is the minimum time required to achieve our
-desired statistical power. (c) Minimum time required for different
-values of the trend strength (\(r\)) and (d) minimum time required for
-different levels of population variability
-(\(\sigma\)).\label{fig:theoretical_approach}}
+desired statistical power. (c) Minimum time required (\(T_{min}\)) for
+simulations with different values of the trend strength (\(r\)) and (d)
+different levels of population variability (\(\sigma\)). In each case,
+the minimum time required is the minimum number of years to achieve 0.8
+statistical power given a significance level of
+0.05.\label{fig:theoretical_approach}}
 \end{figure}
@@
 We use a database of 2444 population time series compiled in (Keith et
-al. 2015). The data is originally from the Global Population Dynamics
+al. 2015). The data are originally from the Global Population Dynamics
 Database (NERC Centre for Population Biology 2010). We filtered out
@@
 autocorrelation. Lastly, we also calculated the minimum number of years
-required (10 years or 3 generations) under IUCN Red List Criterian A2 to
+required (10 years or 3 generations) under IUCN Red List Criterion A2 to
 classify a species as vulnerable (IUCN 2012).
@@
 minimum time required does not strongly correlate with any biological
-characteristics.
+characteristics (Fig. \ref{fig:biological_correlates}).
 
@@
 a particular length that have estimated slope coefficients which are
-statistically different from zero given the long-term, or ``true``, time
+statistically different from zero given the long-term, or ``true'', time
 series also has a significant slope. This is a measure of statistical
@@
 length is required to guarantee we have above a certain threshold of
-statistical power, which we will assume is 0.8 by convention (citation).
+statistical power, which we will assume is 0.8 by convention (Cohen
+1992).
 
@@
 investigated this question through two approaches: simulation-based and
-empirical analyses. We briefly desribe results from the simulation
-approach and then discuss our emprirical approach.
+empirical analyses. We briefly describe results from the simulation
+approach and then discuss our empirical approach.
 
@@
 these explanatory variables were predictive of the minimum time series
-length required (\textbf{supp mat}). Some differences if \(T_{min}\) did
-occur between biological class (\textbf{supp mat}), but this can be
-attributed to the amount populations are known to fluctuate (i.e.~the
-variability in population size).
+length required (Fig. \ref{fig:biological_correlates}). Some differences
+if \(T_{min}\) did occur between biological class (Fig.
+\ref{fig:class}), but this can be attributed to the amount populations
+are known to fluctuate (i.e.~the variability in population size).
 
@@
 
-In the supplmentary material, we explore several additional questions
+In the supplementary material, we explore several additional questions
 related to statistical power and estimating \(T_{min}\). First, we
@@
 linear regression could be used. More specifically, we show how the same
-approach desribed here can be used to determine the minimum time
+approach described here can be used to determine the minimum time
 required to estimate a population's geometric growth rate (Figs.
@@
 lemon shark population dynamics in the Bahamas (White, Nagy, and Gruber
-2014). We found that over 27 years of continous monitoring were needed
+2014). We found that over 27 years of continuous monitoring were needed
 in this particular scenario (Fig. \ref{fig:shark_example}. Lastly, we
@@
 Here, we focus on an empirical approach to estimate the minimum time
-series length requried to assess changes in abundance over time. We
+series length required to assess changes in abundance over time. We
 examined long time series and then subsampled each to determine the
@@
 power for linear regression. Statistical power is important as it
-provides on information as to the neccessary samples required to
+provides on information as to the necessary samples required to
 determine a significant trend (Legg and Nagy 2006). We found that for
 878 populations, at least 10-15 years of continuous monitoring were
-neccessary (Fig. \ref{fig:min_time_dist}). This timeframe is in line
+necessary (Fig. \ref{fig:min_time_dist}). This time frame is in line
 with past work on a short-lived snail species (Rueda-Cediel et al. 2015)
@@
 used generalized additive models to detect population trends and came up
-with a similar, but quantivaely different value for \(T_{min}\) (Fig.
+with a similar, but quantitatively different value for \(T_{min}\) (Fig.
 \ref{fig:min_time_dist_gam}). An additional complication is that for our
@@
 limited number of species. We show that to identify long-term changes in
-abundance, 10-15 years of continuos monitoring are required (Fig.
+abundance, 10-15 years of continuous monitoring are required (Fig.
 \ref{fig:min_time_dist}). In line with theoretical predictions
@@
 In the supporting material, we provide an expanded methods sections,
-additional figures, minimum time calculations for determining exponetial
-growth, simulations with a more complicated population model, minimum
-time calculations for determining long-term growth rates, and the use of
-generalized additive models to identify population trends. All code and
-data can be found at www.github.com/erwhite1
+additional figures, minimum time calculations for determining
+exponential growth, simulations with a more complicated population
+model, minimum time calculations for determining long-term growth rates,
+and the use of generalized additive models to identify population
+trends. All code and data can be found at www.github.com/erwhite1
 
@@
 ed. Princeton, New Jersey: Princeton University Press.
+
+\hypertarget{ref-Cohen1992}{}
+Cohen, Jacob. 1992. ``A power primer.'' \emph{Psychological Bulletin}
+112 (1): 155--59.
+doi:\href{https://doi.org/10.1037/0033-2909.112.1.155}{10.1037/0033-2909.112.1.155}.
 
@@
 \hypertarget{ref-Johnson2015}{}
-Johnson, Paul C.D., Sarah J.E. Barry, Heather M. Ferguson, and Pie
-Müller. 2015. ``Power analysis for generalized linear mixed models in
-ecology and evolution.'' \emph{Methods in Ecology and Evolution} 6 (2):
-133--42.
+Johnson, Paul C, Sarah J Barry, Heather M Ferguson, and Pie Müller.
+2015. ``Power analysis for generalized linear mixed models in ecology
+and evolution.'' \emph{Methods in Ecology and Evolution} 6 (2): 133--42.
 doi:\href{https://doi.org/10.1111/2041-210X.12306}{10.1111/2041-210X.12306}.
@@
 \hypertarget{ref-Morrison2008}{}
-Morrison, Sharn, and David S. Hik. 2008. ``When? Where? And for how
+Morrison, Shawm, and David S. Hik. 2008. ``When? Where? And for how
 long? Census design considerations for an Alpine Lagomorph, the Collared

### 2017-08-14 writing/manuscript_draft.Rmd
added stacked barplot for density plot

@@
 geometry: margin=1in
-header-includes: \usepackage{float} \usepackage{setspace}\doublespacing \usepackage{lineno}
-  \usepackage[round]{natbib} \bibpunct[; ]{(}{)}{,}{a}{}{;} \linenumbers
+header-includes: \usepackage{float} \usepackage{setspace}\doublespacing \usepackage{lineno} \usepackage[firstinits=true,style=authoryear-comp,natbib=true,doi=false,isbn=false,url=false,uniquename=false,uniquelist=false,sorting=none]{biblatex}
+
+ \linenumbers
 
@@
 
-To be submitted to: <!--\emph{Biological Conservation} (Short Communication)-->
+To be submitted to: \emph{Biological Conservation} (Short Communication)
 \vspace{3 mm}
@@
 Long-term census data are necessary to better understand ecological processes and make management decisions.
-However, census data are often expensive, requiring a lot of people-hours and equipment.
+However, census data are often expensive, requiring a lot of time and resource.
 Little work has addressed the question, when is a time series of census data long enough to address a question of interest?
@@
 Using a simulation approach, we demonstrate how the minimum time series length required increases with weaker trends in abundance and with higher population variability.
-In addition, we examine 868 populations of vertebrate species to determine the minimum time required to detect changes in their abundance.
+In addition, we examine 878 populations of vertebrate species to determine the minimum time required to detect changes in their abundance.
 We show that 10-15 years of continuous monitoring are required in order to achieve a high level of statistical power.
@@
 <!--Perhaps surprisingly, the minimum time required did not correlate well with biological explanatory variables, like body size or generation length.--> These results point to the importance of sampling populations over long periods of time.
-<!--Here, We stress the need for verifying a study has enough statistical power to accurately determine long-term changes in population abundance.--> Most studies, especially those less than 10-15 years, are probably under-powered and potentially misleading.
+We stress the need for verifying a study has enough statistical power to accurately determine long-term population trends.
+Most studies, especially those less than 10-15 years, are probably under-powered and potentially misleading.
 
@@
 Too short of a time series may lead to wrong conclusions given large natural year-to-year variability (citations).
-From a management perspective, it is important to understand when a trend in abundance over time is actually meaningful or not.
-The IUCN Red List Categories and Criteria suggest under Criterion A2, a species qualifies as vulnerable if it has experienced a 30% decline over 10 years, or 3 generations [@IUCN2012].
-For both scientific and management questions, because sampling is typically expensive, we do not want to sample for longer than is necessary [@Gerber1999].
+From a management perspective, it is important to understand when population trend over time is actually meaningful or not.
+The IUCN Red List Categories and Criteria suggest, under Criterion A2, a species qualifies as vulnerable if it has experienced a 30% decline over 10 years, or 3 generations [@IUCN2012].
+For both scientific and management questions, because sampling is typically expensive, we also do not want to sample for longer than is necessary [@Gerber1999].
 
@@
 They found that for this short-lived organism, 15 years was adequate to assess long-term trends in abundance.
-However, these studies, and other past work, have typically focused either on theoretical aspects of monitoring design or focused on single species.
+However, these studies, and other past work, have typically focused either on theoretical aspects of monitoring design or focused on only a few species.
 
-In this paper, we use both simulations and a database of empirical time series to determine the minimum number of years required to address various questions.
-More specifically, we estimate the minimum time series length required to assess long-term changes in abundance via simple linear regression.
-We examine 878 population time series to estimate this minimum time series length required for detecting population trends.
+In this paper, we use both simulations and a database of empirical time series to determine the minimum number of years required to address several questions.
+In the main manuscript, we estimate the minimum time series length required to assess long-term changes in abundance via simple linear regression.
+We then examine 878 population time series to estimate this minimum time series length required for detecting population trends.
+We expand on the results in the supplmentary material by examining other questions of interest.
 
@@
 One approach to determining the minimum time series length needed is through repetitive simulations of a population model [@Gerrodette1987].
-This is the same approach one might use in sample size calculations for any experimental design too complicated for simple power analyses [@Johnson2015] (and cite Bolker).
-We only briefly discuss this approach as it has been described elsewhere (citations).
-Essentially, we take a stochastic population model and simulate it for a number of years repetitively.
+This is the same approach one might use in sample size calculations for any experimental design too complicated for simple power analyses [@Bolker2008;@Johnson2015].
+We only briefly discuss this approach as it has been described elsewhere.
+Essentially, we use a population model and simulate it for a number of years repetitively.
 We can then determine the minimum time series required to achieve designed levels of significance and power for a particular question.
 This approach requires us to determine values for our model parameters (e.g.
-growth rate).
+population variability).
 As an example, we can take the following population model for population size $N$ at time $t$:
@@
 $$
-where $\epsilon$ is a normally-distributed random noise term with mean $\mu$ and standard deviation $\sigma$. The rate of growth $r$ is also the trend strength of increase or decrease.
+where $\epsilon$ is a normally-distributed random noise term with mean $\mu$ and standard deviation $\sigma$.
+The rate of growth $r$ is also the trend strength of the increase or decrease (i.e.
+the rate of increase).
 
 By simulating this model repetitively for different lengths of time, we can determine the minimum time series required.
-Here statistical power is simply the fraction of simulations with significant trends in abundance.
+Here, statistical power is simply the fraction of simulations with significant trends in abundance.
 A significant trend in abundance is determined by estimating the significance of the slope parameter from simple linear regression.
@@
 \ref{fig:theoretical_approach}b).
-Once power is greater than 0.8 (the dotted line), that is the minimum time required ($T_{min}$).
+Where power is greater than 0.8 (the dotted line), that is the minimum time required ($T_{min}$).
 
@@
 For each time series, we also calculated variables of interest like variance in population size, long-term trend in abundance (slope coefficient from simple linear regression), and temporal autocorrelation.
-Lastly, we also calculated the minimum number of years required (10 years or 3 generations) under IUCN Red List Criterion A2 to classify a species as vulnerable [@IUCN2012].
+<!--Lastly, we also calculated the minimum number of years required (10 years or 3 generations) under IUCN Red List Criterion A2 to classify a species as vulnerable [@IUCN2012].-->
 
@@
 ```
-For a subset of populations in the database ($n =$ `r nrow(LPI_pop_info)`), we were able to connect a database on biological characteristics [@Myhrvold2015] like body size and generation length.
+For a subset of populations in the database ($n =$ `r nrow(LPI_pop_info)`), we were able to connect a database on biological characteristics [@Myhrvold2015], including body size and generation length.
 All `r nrow(LPI_pop_info)` populations are within the Aves class.
@@
 For example, a time series of 40 years would have 39 possible contiguous subsamples of length 2, 38 possible contiguous subsamples of length 3, and continuing until 1 possible contiguous subsample of length 40 [@Gerber1999; @Giron-Nava2017].
-Then we determined the $T_{min}$ required to be confident (i.e.
+Then, we determine the $T_{min}$ required to be confident (i.e.
 have high enough statistical power) in an estimate of the slope coefficient from linear regression.
@@
 
-We determined the minimum time series length required to address a particular question of interest.
-More specifically we asked, what is the minimum time series length required to determine, via linear regression, the long-term trend in population abundance?
-The minimum time series length required is that which had high enough statistical power.
+We determined the minimum time series length ($T_{min}$) required to address a particular question of interest.
+More specifically we asked, what is the minimum time series length required to determine, via linear regression, the long-term population trend?
+Here, the minimum time series length required had high enough statistical power (greater than 0.8) for a set significance level of 0.05.
 We investigated this question through two approaches: simulation-based and empirical analyses.
@@
 
-We constructed a general population model where the trend over time could be a model parameter.
-We chose a simple model, but any other population model could be used.
-The specific model choice should be tailored to the population of interest.
+We constructed a general population model where the trend (i.e.
+slope coefficient) over time could be a model parameter.
+We chose a simple model, but any other population model could be used (Fig.
+\ref{fig:min_time_dist_log_pop}).
+Ideally, the specific model choice should be tailored to the population of interest.
 We then ran simulated time series of different lengths.
-From these simulations we determined the minimum time series length required to achieve a certain level of power (which we set to 0.8).
-We were also able to alter the parameter values (trend strength and population variability) of the population model to see how this altered $T_{min}$.
+From these simulations we determined the minimum time series length required to achieve a certain level of statistical power.
+We were also able to alter the parameter values (trend strength and population variability) of the population model to see how this altered $T_{min}$ (Fig.
+\ref{fig:theoretical_approach}).
 
-In line with past work, we found the $T_{min}$ increases (i.e. more time is required) with decreases in trend strength and with increases in variability (citation).
+In line with past work [@Gerrodette1987], we found the $T_{min}$ increases (i.e.
+more time is required) with decreases in trend strength and with increases in variability (Fig.
+\ref{fig:theoretical_approach}).
 
-## Empirical test
+## Empirical approach
 
@@
 
-plot(density(pop_info$min_time_for_power),las=1,ylab='',xlab='',cex.axis=1.2,main='')
+#plot(density(pop_info$min_time_for_power),las=1,ylab='',xlab='',cex.axis=1.2,main='')
+
+counts <- table(pop_info$class,pop_info$min_time_for_power)
+barplot(counts, main=" ",
+        xlab=" ", ylab="", col=1:4,
+        legend = rownames(counts),ylim=c(0,50),xlim=c(0,38),
+        las=1,cex.names =1.2,cex.axis = 1.2,density=30,
+        angle=c(20,40,60,80))
+
 mtext(text = 'Frequency',2,line=3,cex=1.2)
@@
 \ref{fig:min_time_dist}).
-In addition, as expected based on theoretical results (cite papers), the minimum time series length required is strongly correlated with trend strength, population variance, and autocorrelation (Fig \ref{fig:correlates}).
-Using a generalized linear model, with a Poisson error structure, all three of these explanatory variables were significant and had large effect sizes (see Fig.
-SX).
+In addition, as expected based on our simulation results and those of others [@Rhodes2011], the minimum time series length required is strongly correlated with trend strength, population variance, and autocorrelation (Fig \ref{fig:correlates}).
+Using a generalized linear model, with a Poisson error structure, all three of these explanatory variables were significant and had large effect sizes (see Table S1).
 Combined, trend strength, population variance, and autocorrelation account for about `r round(100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance),1)` of the explained deviance, which is similar to $R^2$ [@Zuur2009], in minimum time series length required.
@@
 \ref{fig:biological_correlates}).
-Some differences if $T_{min}$ did occur between biological class (Fig.
+Some differences in $T_{min}$ did occur between biological class (Fig.
 \ref{fig:class}), but this can be attributed to the amount populations are known to fluctuate (i.e.
@@
 We explored two different approaches to estimate the minimum time series length required to address a particular question of interest.
-We specifically asked, what is the minimum time series length required to determine long-term changes in abundance using linear regression?
+We specifically asked, what is the minimum time series length required to determine long-term population trends using linear regression?
+This is one of the simplest questions one could ask of a time series.
+We would expect more involved questions (e.g.
+detecting chaos) to require a larger minimum time series length.
 The simulation-based approach has been suggested by others, especially in situations more complicated than that suited for classic power analysis [@Gerrodette1987;@Johnson2015;@Bolker2008].
@@
 
-# Figures
-
-stuff here \ref{fig:empirical_approach_example}
-
 \clearpage

### 2017-08-14 writing/manuscript_draft.tex
added stacked barplot for density plot

@@
 \usepackage{setspace}\doublespacing \usepackage{lineno}
-\usepackage[round]{natbib} \bibpunct[; ]{(}{)}{,}{a}{}{;} \linenumbers
+\usepackage[firstinits=true,style=authoryear-comp,natbib=true,doi=false,isbn=false,url=false,uniquename=false,uniquelist=false,sorting=none]{biblatex}
+\linenumbers
 
@@
 processes and make management decisions. However, census data are often
-expensive, requiring a lot of people-hours and equipment. Little work
-has addressed the question, when is a time series of census data long
-enough to address a question of interest? Here, we explore two
-approaches to address this question: simulations and an empirical
-approach. We specifically determine the minimum time series length
-required to estimate significant increases or decreases in population
-abundance. Importantly, we examine the ability to detect a trend with a
-set level of statistical power, which is often neglected in ecological
-time series analyses. Using a simulation approach, we demonstrate how
-the minimum time series length required increases with weaker trends in
-abundance and with higher population variability. In addition, we
-examine 868 populations of vertebrate species to determine the minimum
-time required to detect changes in their abundance. We show that 10-15
-years of continuous monitoring are required in order to achieve a high
-level of statistical power. Similar to simulation-approaches, the
-minimum time required for field census data strongly depends on trend
-strength, population variability, and temporal autocorrelation. These
-results point to the importance of sampling populations over long
-periods of time. Most studies, especially those less than 10-15 years,
-are probably under-powered and potentially misleading.
+expensive, requiring a lot of time and resource. Little work has
+addressed the question, when is a time series of census data long enough
+to address a question of interest? Here, we explore two approaches to
+address this question: simulations and an empirical approach. We
+specifically determine the minimum time series length required to
+estimate significant increases or decreases in population abundance.
+Importantly, we examine the ability to detect a trend with a set level
+of statistical power, which is often neglected in ecological time series
+analyses. Using a simulation approach, we demonstrate how the minimum
+time series length required increases with weaker trends in abundance
+and with higher population variability. In addition, we examine 878
+populations of vertebrate species to determine the minimum time required
+to detect changes in their abundance. We show that 10-15 years of
+continuous monitoring are required in order to achieve a high level of
+statistical power. Similar to simulation-approaches, the minimum time
+required for field census data strongly depends on trend strength,
+population variability, and temporal autocorrelation. These results
+point to the importance of sampling populations over long periods of
+time. We stress the need for verifying a study has enough statistical
+power to accurately determine long-term population trends. Most studies,
+especially those less than 10-15 years, are probably under-powered and
+potentially misleading.
+
+  \renewcommand{\thesection}{\arabic{section}}
 
@@
 (citations). From a management perspective, it is important to
-understand when a trend in abundance over time is actually meaningful or
-not. The IUCN Red List Categories and Criteria suggest under Criterion
+understand when population trend over time is actually meaningful or
+not. The IUCN Red List Categories and Criteria suggest, under Criterion
 A2, a species qualifies as vulnerable if it has experienced a 30\%
 decline over 10 years, or 3 generations (IUCN 2012). For both scientific
-and management questions, because sampling is typically expensive, we do
-not want to sample for longer than is necessary (Gerber, DeMaster, and
-Kareiva 1999).
+and management questions, because sampling is typically expensive, we
+also do not want to sample for longer than is necessary (Gerber,
+DeMaster, and Kareiva 1999).
 
@@
 typically focused either on theoretical aspects of monitoring design or
-focused on single species.
+focused on only a few species.
 
@@
 series to determine the minimum number of years required to address
-various questions. More specifically, we estimate the minimum time
+several questions. In the main manuscript, we estimate the minimum time
 series length required to assess long-term changes in abundance via
-simple linear regression. We examine 878 population time series to
+simple linear regression. We then examine 878 population time series to
 estimate this minimum time series length required for detecting
-population trends.
+population trends. We expand on the results in the supplmentary material
+by examining other questions of interest.
 
@@
 This is the same approach one might use in sample size calculations for
-any experimental design too complicated for simple power analyses (P. C.
-Johnson et al. 2015) (and cite Bolker). We only briefly discuss this
-approach as it has been described elsewhere (citations). Essentially, we
-take a stochastic population model and simulate it for a number of years
-repetitively. We can then determine the minimum time series required to
-achieve designed levels of significance and power for a particular
-question. This approach requires us to determine values for our model
-parameters (e.g.~growth rate). As an example, we can take the following
+any experimental design too complicated for simple power analyses
+(Bolker 2008; P. C. Johnson et al. 2015). We only briefly discuss this
+approach as it has been described elsewhere. Essentially, we use a
+population model and simulate it for a number of years repetitively. We
+can then determine the minimum time series required to achieve designed
+levels of significance and power for a particular question. This
+approach requires us to determine values for our model parameters
+(e.g.~population variability). As an example, we can take the following
 population model for population size \(N\) at time \(t\): \[
@@
 mean \(\mu\) and standard deviation \(\sigma\). The rate of growth \(r\)
-is also the trend strength of increase or decrease.
+is also the trend strength of the increase or decrease (i.e.~the rate of
+increase).
 
 By simulating this model repetitively for different lengths of time, we
-can determine the minimum time series required. Here statistical power
+can determine the minimum time series required. Here, statistical power
 is simply the fraction of simulations with significant trends in
@@
 clear that statistical power increases quickly with increases in length
-of time sampled (Fig. \ref{fig:theoretical_approach}b). Once power is
+of time sampled (Fig. \ref{fig:theoretical_approach}b). Where power is
 greater than 0.8 (the dotted line), that is the minimum time required
@@
 (slope coefficient from simple linear regression), and temporal
-autocorrelation. Lastly, we also calculated the minimum number of years
-required (10 years or 3 generations) under IUCN Red List Criterion A2 to
-classify a species as vulnerable (IUCN 2012).
+autocorrelation.
 
@@
 to connect a database on biological characteristics (Myhrvold et al.
-2015) like body size and generation length. All 547 populations are
-within the Aves class. In the supplementary material, we explore how the
-minimum time required does not strongly correlate with any biological
-characteristics (Fig. \ref{fig:biological_correlates}).
+2015), including body size and generation length. All 547 populations
+are within the Aves class. In the supplementary material, we explore how
+the minimum time required does not strongly correlate with any
+biological characteristics (Fig. \ref{fig:biological_correlates}).
 
@@
 and continuing until 1 possible contiguous subsample of length 40
-(Gerber, DeMaster, and Kareiva 1999; Giron-Nava et al. 2017). Then we
-determined the \(T_{min}\) required to be confident (i.e.~have high
+(Gerber, DeMaster, and Kareiva 1999; Giron-Nava et al. 2017). Then, we
+determine the \(T_{min}\) required to be confident (i.e.~have high
 enough statistical power) in an estimate of the slope coefficient from
@@
 
-We determined the minimum time series length required to address a
-particular question of interest. More specifically we asked, what is the
-minimum time series length required to determine, via linear regression,
-the long-term trend in population abundance? The minimum time series
-length required is that which had high enough statistical power. We
-investigated this question through two approaches: simulation-based and
-empirical analyses. We briefly describe results from the simulation
-approach and then discuss our empirical approach.
+We determined the minimum time series length (\(T_{min}\)) required to
+address a particular question of interest. More specifically we asked,
+what is the minimum time series length required to determine, via linear
+regression, the long-term population trend? Here, the minimum time
+series length required had high enough statistical power (greater than
+0.8) for a set significance level of 0.05. We investigated this question
+through two approaches: simulation-based and empirical analyses. We
+briefly describe results from the simulation approach and then discuss
+our empirical approach.
 
@@
 
-We constructed a general population model where the trend over time
-could be a model parameter. We chose a simple model, but any other
-population model could be used. The specific model choice should be
-tailored to the population of interest. We then ran simulated time
-series of different lengths. From these simulations we determined the
-minimum time series length required to achieve a certain level of power
-(which we set to 0.8). We were also able to alter the parameter values
+We constructed a general population model where the trend (i.e.~slope
+coefficient) over time could be a model parameter. We chose a simple
+model, but any other population model could be used (Fig.
+\ref{fig:min_time_dist_log_pop}). Ideally, the specific model choice
+should be tailored to the population of interest. We then ran simulated
+time series of different lengths. From these simulations we determined
+the minimum time series length required to achieve a certain level of
+statistical power. We were also able to alter the parameter values
 (trend strength and population variability) of the population model to
-see how this altered \(T_{min}\).
+see how this altered \(T_{min}\) (Fig. \ref{fig:theoretical_approach}).
 
-In line with past work, we found the \(T_{min}\) increases (i.e.~more
-time is required) with decreases in trend strength and with increases in
-variability (citation).
+In line with past work (Gerrodette 1987), we found the \(T_{min}\)
+increases (i.e.~more time is required) with decreases in trend strength
+and with increases in variability (Fig. \ref{fig:theoretical_approach}).
 
-\subsection{Empirical test}\label{empirical-test}
+\subsection{Empirical approach}\label{empirical-approach-1}
 
@@
 wide distribution (Fig. \ref{fig:min_time_dist}). In addition, as
-expected based on theoretical results (cite papers), the minimum time
-series length required is strongly correlated with trend strength,
-population variance, and autocorrelation (Fig \ref{fig:correlates}).
-Using a generalized linear model, with a Poisson error structure, all
-three of these explanatory variables were significant and had large
-effect sizes (see Fig. SX). Combined, trend strength, population
-variance, and autocorrelation account for about 72.6 of the explained
-deviance, which is similar to \(R^2\) (Zuur et al. 2009), in minimum
-time series length required.
+expected based on our simulation results and those of others (Rhodes and
+Jonzen 2011), the minimum time series length required is strongly
+correlated with trend strength, population variance, and autocorrelation
+(Fig \ref{fig:correlates}). Using a generalized linear model, with a
+Poisson error structure, all three of these explanatory variables were
+significant and had large effect sizes (see Table S1). Combined, trend
+strength, population variance, and autocorrelation account for about
+72.6 of the explained deviance, which is similar to \(R^2\) (Zuur et al.
+2009), in minimum time series length required.
 
@@
 length required (Fig. \ref{fig:biological_correlates}). Some differences
-if \(T_{min}\) did occur between biological class (Fig.
+in \(T_{min}\) did occur between biological class (Fig.
 \ref{fig:class}), but this can be attributed to the amount populations
@@
 specifically asked, what is the minimum time series length required to
-determine long-term changes in abundance using linear regression? The
-simulation-based approach has been suggested by others, especially in
-situations more complicated than that suited for classic power analysis
-(Gerrodette 1987; P. C. Johnson et al. 2015; Bolker 2008). Our
-simulations support past work that longer time series are needed when
-the trend strength (i.e.~rate of increase or decrease) is weak or when
-population variability is high (Gerrodette 1987).
+determine long-term population trends using linear regression? This is
+one of the simplest questions one could ask of a time series. We would
+expect more involved questions (e.g.~detecting chaos) to require a
+larger minimum time series length. The simulation-based approach has
+been suggested by others, especially in situations more complicated than
+that suited for classic power analysis (Gerrodette 1987; P. C. Johnson
+et al. 2015; Bolker 2008). Our simulations support past work that longer
+time series are needed when the trend strength (i.e.~rate of increase or
+decrease) is weak or when population variability is high (Gerrodette
+1987).
 
@@
 \hypertarget{ref-Johnson2015}{}
-Johnson, Paul C, Sarah J Barry, Heather M Ferguson, and Pie Müller.
+Johnson, Paul C, Sarah JE Barry, Heather M Ferguson, and Pie Müller.
 2015. ``Power analysis for generalized linear mixed models in ecology
@@
 Ecology with R}. New York: Springer.
-
-\section{Figures}\label{figures}
-
-stuff here \ref{fig:empirical_approach_example}
 

### 2017-08-15 writing/manuscript_draft.Rmd
major revisions to introduction

@@
 
-Title: Statistical power and detecting population trends: the need for long-term monitoring programs
+Title: Minimum time required to detect population trends: the need for long-term monitoring programs
 \vspace{7 mm}
@@
 
-To be submitted to: \emph{Biological Conservation} (Short Communication)
+To be submitted to: \emph{Biological Conservation}
 \vspace{3 mm}
@@
 Long-term census data are necessary to better understand ecological processes and make management decisions.
-However, census data are often expensive, requiring a lot of time and resource.
-Little work has addressed the question, when is a time series of census data long enough to address a question of interest?
-Here, we explore two approaches to address this question: simulations and an empirical approach.
-We specifically determine the minimum time series length required to estimate significant increases or decreases in population abundance.
+However, census data are often expensive, requiring a lot of time and resources.
+Little work has addressed the question: when is a time series of census data long enough to address a question of interest?
+<!--Here, we explore two approaches to address this question: simulations and an empirical approach.--> We determine the minimum time series length required to estimate significant increases or decreases in population abundance.
 Importantly, we examine the ability to detect a trend with a set level of statistical power, which is often neglected in ecological time series analyses.
@@
 <!--Perhaps surprisingly, the minimum time required did not correlate well with biological explanatory variables, like body size or generation length.--> These results point to the importance of sampling populations over long periods of time.
-We stress the need for verifying a study has enough statistical power to accurately determine long-term population trends.
-Most studies, especially those less than 10-15 years, are probably under-powered and potentially misleading.
+<!--We stress the need for verifying a study has enough statistical power to accurately determine long-term population trends.--> Time series less than 10-15 years are likely under-powered and potentially misleading.
 
@@
 
-Observational studies and population censuses have become a cornerstone of modern ecological research and conservation biology [@Magurran2010; @Hughes2017].
-Long-term data are necessary to both understand population dynamics and to assess extinction risk of species.
-Even though many time series may now be considered "long-term", most are still short.
+Observational studies and population time series have become a cornerstone of modern ecological research and conservation biology [@Magurran2010; @Hughes2017].
+Long-term data are necessary to both understand population dynamics and to assess species extinction risk (citations).
+Even though many time series may now be considered "long-term" (e.g.
+continuous plankton recorder @Giron-Nava2017), most are still short.
 Time series are typically short for a variety of reasons.
@@
 In addition, short funding cycles make it difficult to examine populations over longer periods of time [@Hughes2017].
-There are, of course, notable exceptions where long-term monitoring has occurred, for example the continuous plankton recorder [@Giron-Nava2017].
 
-Naturally, one might ask, how long of a time series is actually necessary?
+How long of a time series is actually necessary?
 This question has important implications for both research and management.
-Scientifically, we need to know the time series length required to address a specific question.
-Too short of a time series may lead to wrong conclusions given large natural year-to-year variability (citations).
-From a management perspective, it is important to understand when population trend over time is actually meaningful or not.
+Scientists need to know the time series length required to address a specific question.
+A short time series may lead to wrong conclusions given large natural year-to-year variability (citations).
+Managers need to know when action is needed for a population.
+Therefore, they must understand when population trend over time is actually meaningful.
 The IUCN Red List Categories and Criteria suggest, under Criterion A2, a species qualifies as vulnerable if it has experienced a 30% decline over 10 years, or 3 generations [@IUCN2012].
@@
 
-In determining the number of samples required for a particular experiment four quantities are intricately linked: statistical power (1-$\beta$), effect size, $\alpha$, and sample size [@Legg2006].
+An important step in experimental design is to determine the number of samples required.
+For any particular experiment four quantities are intricately linked: significance level ($\alpha$), statistical power, effect size, and sample size [@Legg2006].
 The exact relationship between these quantities depends on the specific statistical test.
-Formally, statistical power ($1-\beta$) is one minus the probability of a type II error ($\beta$), or false negative.
-In other words, it is a measure of how well you could detect a trend given it is actually significant.
-Prior to an experiment, one could set appropriate levels of power, significance threshold ($\alpha$), and the effect size to estimate the sample size required for the experiment.
-This approach, however, is not straight-forward for a time series, or more complicated scenarios [@Johnson2015], as data points are clearly non-independent.
+A type I error is a false positive, or incorrect rejection of a true null hypothesis.
+For example, if a time series was assessed as significantly increasing or decreasing---when there was no true significant trend---this would be a false positive.
+The false positive rate, or significance level ($\alpha$) is often set at 0.05.
+A type II error ($\beta$) is failure to detect a true trend, or failure to reject a false null hypothesis.
+Formally, statistical power ($1-\beta$) is one minus the probability of a type II error ($\beta$).
+The effect size is a measure of the difference between two groups.
+Prior to an experiment, one could set appropriate levels of power, significance level, and the effect size to estimate the sample size required for the experiment.
+This approach, however, is not straight-forward for a time series, or more complicated scenarios [@Johnson2015], as data are clearly non-independent.
 
-For time series data, two general approaches can be used.
-First, as can be done with experimental data, we can relax the definition of statistical power [@Bolker2008].
-Instead of using classic statistical tests (e.g.
-t-test), and corresponding formulas for determining sample size, simulation approaches can be used [@Gerrodette1987; @Bolker2008; @Johnson2015].
-Simple models can be simulated with parameter values corresponding to a population of interest.
-Out of a large number of simulations, power can be calculated as the fraction of simulations that meet some criteria.
+For time series data, two general approaches to estimating sample size are appropriate.
+<!--Typically, calculations of sample size are given by a formula specific to standard statistical tests (e.g.
+t-test).
+However, many ecological scenarios require the use of simulations [@Bolker2008; @Johnson2015].
+As a first approach,-->Simulations can be designed for a specific population and question [@Bolker2008; @Johnson2015].
+Simple models can be simulated with parameter values corresponding to a population of interest [@Gerrodette1987].
+Statistical power can then be calculated as the fraction of simulations that meet some criteria.
 The specific criteria depend on the question at hand.
-Then, time series simulations of different lengths are run to determine the minimum time series length required to be confident that you will be able to answer your question [@Bolker2008].
-In addition to using simulated data, empirical time series can also be used.
-We usually do not have multiple replicates of similar populations, but we can subsample an empirical time series [@Gerber1999].
+For example, given a time series, when is the slope from linear regression significantly different from zero?
+In other words, when is the time series significantly increasing or decreasing?
+It is then possible to determine how power changes with a variable of interest.
+For example, time series are simulated for different lengths of time.
+From these simulations, the minimum time series length required is calculated to meet certain levels of statistical significance and power [@Bolker2008].
+
+In addition to using simulations, empirical time series can also be used.
+Multiple replicates of similar populations are usually not available, but it is possible to subsample an empirical time series [@Gerber1999].
 Subsamples of different lengths can then be evaluated to see which fraction of subsamples meet some criteria, again a measure of statistical power.
@@
 They found that for this short-lived organism, 15 years was adequate to assess long-term trends in abundance.
+In addition, @Hatch2003 examined seabird populations and determined the minimum time required to achieve high statistical power ranged between 11 and 69 years.
 However, these studies, and other past work, have typically focused either on theoretical aspects of monitoring design or focused on only a few species.
 
-In this paper, we use both simulations and a database of empirical time series to determine the minimum number of years required to address several questions.
-In the main manuscript, we estimate the minimum time series length required to assess long-term changes in abundance via simple linear regression.
-We then examine 878 population time series to estimate this minimum time series length required for detecting population trends.
-We expand on the results in the supplmentary material by examining other questions of interest.
+We use both simulations and a database of empirical time series to determine the minimum number of years required to address several questions.
+We estimate the minimum time series length required ($T_{min}$) to assess long-term changes in abundance via simple linear regression.
+First, we estimate $T_{min}$ using a simulation approach.
+Then we examine 878 population time series to estimate $T_{min}$.
+In the supplementary material, we determine $T_{min}$ for related ideas: calculating long-term growth rates, using more complicated population models, varying statistical level and power, and the use of generalized additive models.
 
@@
         legend = rownames(counts),ylim=c(0,50),xlim=c(0,38),
-        las=1,cex.names =1.2,cex.axis = 1.2,density=30,
+        las=1,cex.names =1.2,cex.axis = 1.2,density=35,
         angle=c(20,40,60,80))
@@
 
-```{r,echo=F, fig.cap='Minimum time required to estimate change in abundance correlated with (a) trend strength (absolute value of slope coefficient estimated from linear regression), (b) population variance (interannual variability in population size), and (c) temporal autocorrelation.\\label{fig:correlates}',fig.height=5}
+```{r,echo=F, fig.cap='Minimum time required to estimate change in abundance correlated with (a) trend strength (absolute value of slope coefficient estimated from linear regression), (b) population variance (interannual variability in population size), and (c) temporal autocorrelation.\\label{fig:correlates}',fig.height=7, fig.width=4}
 
 # Plot minimum time required against trend strength, population variance, and autocorrelation  
-  par(mfrow=c(3,1),mar=c(4,2,2,1),oma=c(0,4.5,0,0.5))
+  par(mfrow=c(3,1),mar=c(3,2,2,1),oma=c(1,4.5,0,0.5))
   plot(abs(pop_info$overall_trend),pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
@@
     plot(pop_info$variance,pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
-    mtext('population variance',1,line=3,cex=1.3)
+    mtext('population variance',1,line=3,cex=1.2)
     mtext('(b)',3,line=-1.5,adj = 0.985,cex=1.2,font = 2)

### 2017-08-15 writing/manuscript_draft.tex
major revisions to introduction

@@
 
-Title: Statistical power and detecting population trends: the need for
+Title: Minimum time required to detect population trends: the need for
 long-term monitoring programs \vspace{7 mm}
@@
 
-To be submitted to: \vspace{3 mm}
+To be submitted to: \emph{Biological Conservation} \vspace{3 mm}
 
@@
 processes and make management decisions. However, census data are often
-expensive, requiring a lot of time and resource. Little work has
-addressed the question, when is a time series of census data long enough
-to address a question of interest? Here, we explore two approaches to
-address this question: simulations and an empirical approach. We
-specifically determine the minimum time series length required to
-estimate significant increases or decreases in population abundance.
-Importantly, we examine the ability to detect a trend with a set level
-of statistical power, which is often neglected in ecological time series
-analyses. Using a simulation approach, we demonstrate how the minimum
-time series length required increases with weaker trends in abundance
-and with higher population variability. In addition, we examine 878
-populations of vertebrate species to determine the minimum time required
-to detect changes in their abundance. We show that 10-15 years of
-continuous monitoring are required in order to achieve a high level of
-statistical power. Similar to simulation-approaches, the minimum time
-required for field census data strongly depends on trend strength,
-population variability, and temporal autocorrelation. These results
-point to the importance of sampling populations over long periods of
-time. We stress the need for verifying a study has enough statistical
-power to accurately determine long-term population trends. Most studies,
-especially those less than 10-15 years, are probably under-powered and
-potentially misleading.
-
-  \renewcommand{\thesection}{\arabic{section}}
+expensive, requiring a lot of time and resources. Little work has
+addressed the question: when is a time series of census data long enough
+to address a question of interest? We determine the minimum time series
+length required to estimate significant increases or decreases in
+population abundance. Importantly, we examine the ability to detect a
+trend with a set level of statistical power, which is often neglected in
+ecological time series analyses. Using a simulation approach, we
+demonstrate how the minimum time series length required increases with
+weaker trends in abundance and with higher population variability. In
+addition, we examine 878 populations of vertebrate species to determine
+the minimum time required to detect changes in their abundance. We show
+that 10-15 years of continuous monitoring are required in order to
+achieve a high level of statistical power. Similar to
+simulation-approaches, the minimum time required for field census data
+strongly depends on trend strength, population variability, and temporal
+autocorrelation. These results point to the importance of sampling
+populations over long periods of time. Time series less than 10-15 years
+are likely under-powered and potentially misleading.
 
@@
 
-Observational studies and population censuses have become a cornerstone
-of modern ecological research and conservation biology (Magurran et al.
-2010; Hughes et al. 2017). Long-term data are necessary to both
-understand population dynamics and to assess extinction risk of species.
-Even though many time series may now be considered ``long-term'', most
-are still short. Time series are typically short for a variety of
-reasons. They are often coupled with an experiment, which may only last
-a couple of years. In addition, short funding cycles make it difficult
-to examine populations over longer periods of time (Hughes et al. 2017).
-There are, of course, notable exceptions where long-term monitoring has
-occurred, for example the continuous plankton recorder (Giron-Nava et
+Observational studies and population time series have become a
+cornerstone of modern ecological research and conservation biology
+(Magurran et al. 2010; Hughes et al. 2017). Long-term data are necessary
+to both understand population dynamics and to assess species extinction
+risk (citations). Even though many time series may now be considered
+``long-term'' (e.g.~continuous plankton recorder Giron-Nava et al.
+(2017)), most are still short. Time series are typically short for a
+variety of reasons. They are often coupled with an experiment, which may
+only last a couple of years. In addition, short funding cycles make it
+difficult to examine populations over longer periods of time (Hughes et
 al. 2017).
 
-Naturally, one might ask, how long of a time series is actually
-necessary? This question has important implications for both research
-and management. Scientifically, we need to know the time series length
-required to address a specific question. Too short of a time series may
-lead to wrong conclusions given large natural year-to-year variability
-(citations). From a management perspective, it is important to
-understand when population trend over time is actually meaningful or
-not. The IUCN Red List Categories and Criteria suggest, under Criterion
-A2, a species qualifies as vulnerable if it has experienced a 30\%
-decline over 10 years, or 3 generations (IUCN 2012). For both scientific
-and management questions, because sampling is typically expensive, we
-also do not want to sample for longer than is necessary (Gerber,
-DeMaster, and Kareiva 1999).
+How long of a time series is actually necessary? This question has
+important implications for both research and management. Scientists need
+to know the time series length required to address a specific question.
+A short time series may lead to wrong conclusions given large natural
+year-to-year variability (citations). Managers need to know when action
+is needed for a population. Therefore, they must understand when
+population trend over time is actually meaningful. The IUCN Red List
+Categories and Criteria suggest, under Criterion A2, a species qualifies
+as vulnerable if it has experienced a 30\% decline over 10 years, or 3
+generations (IUCN 2012). For both scientific and management questions,
+because sampling is typically expensive, we also do not want to sample
+for longer than is necessary (Gerber, DeMaster, and Kareiva 1999).
 
-In determining the number of samples required for a particular
-experiment four quantities are intricately linked: statistical power
-(1-\(\beta\)), effect size, \(\alpha\), and sample size (Legg and Nagy
-2006). The exact relationship between these quantities depends on the
-specific statistical test. Formally, statistical power (\(1-\beta\)) is
-one minus the probability of a type II error (\(\beta\)), or false
-negative. In other words, it is a measure of how well you could detect a
-trend given it is actually significant. Prior to an experiment, one
-could set appropriate levels of power, significance threshold
-(\(\alpha\)), and the effect size to estimate the sample size required
-for the experiment. This approach, however, is not straight-forward for
-a time series, or more complicated scenarios (P. C. Johnson et al.
-2015), as data points are clearly non-independent.
+An important step in experimental design is to determine the number of
+samples required. For any particular experiment four quantities are
+intricately linked: significance level (\(\alpha\)), statistical power,
+effect size, and sample size (Legg and Nagy 2006). The exact
+relationship between these quantities depends on the specific
+statistical test. A type I error is a false positive, or incorrect
+rejection of a true null hypothesis. For example, if a time series was
+assessed as significantly increasing or decreasing---when there was no
+true significant trend---this would be a false positive. The false
+positive rate, or significance level (\(\alpha\)) is often set at 0.05.
+A type II error (\(\beta\)) is failure to detect a true trend, or
+failure to reject a false null hypothesis. Formally, statistical power
+(\(1-\beta\)) is one minus the probability of a type II error
+(\(\beta\)). The effect size is a measure of the difference between two
+groups. Prior to an experiment, one could set appropriate levels of
+power, significance level, and the effect size to estimate the sample
+size required for the experiment. This approach, however, is not
+straight-forward for a time series, or more complicated scenarios (P. C.
+Johnson et al. 2015), as data are clearly non-independent.
 
-For time series data, two general approaches can be used. First, as can
-be done with experimental data, we can relax the definition of
-statistical power (Bolker 2008). Instead of using classic statistical
-tests (e.g.~t-test), and corresponding formulas for determining sample
-size, simulation approaches can be used (Gerrodette 1987; Bolker 2008;
-P. C. Johnson et al. 2015). Simple models can be simulated with
-parameter values corresponding to a population of interest. Out of a
-large number of simulations, power can be calculated as the fraction of
-simulations that meet some criteria. The specific criteria depend on the
-question at hand. Then, time series simulations of different lengths are
-run to determine the minimum time series length required to be confident
-that you will be able to answer your question (Bolker 2008). In addition
-to using simulated data, empirical time series can also be used. We
-usually do not have multiple replicates of similar populations, but we
-can subsample an empirical time series (Gerber, DeMaster, and Kareiva
-1999). Subsamples of different lengths can then be evaluated to see
-which fraction of subsamples meet some criteria, again a measure of
-statistical power. Similar to the simulation approach, we can use these
-measures of power to determine the minimum time series required for a
-particular question of interest.
+For time series data, two general approaches to estimating sample size
+are appropriate. Simulations can be designed for a specific population
+and question (Bolker 2008; P. C. Johnson et al. 2015). Simple models can
+be simulated with parameter values corresponding to a population of
+interest (Gerrodette 1987). Statistical power can then be calculated as
+the fraction of simulations that meet some criteria. The specific
+criteria depend on the question at hand. For example, given a time
+series, when is the slope from linear regression significantly different
+from zero? In other words, when is the time series significantly
+increasing or decreasing? It is then possible to determine how power
+changes with a variable of interest. For example, time series are
+simulated for different lengths of time. From these simulations, the
+minimum time series length required is calculated to meet certain levels
+of statistical significance and power (Bolker 2008).
+
+In addition to using simulations, empirical time series can also be
+used. Multiple replicates of similar populations are usually not
+available, but it is possible to subsample an empirical time series
+(Gerber, DeMaster, and Kareiva 1999). Subsamples of different lengths
+can then be evaluated to see which fraction of subsamples meet some
+criteria, again a measure of statistical power. Similar to the
+simulation approach, we can use these measures of power to determine the
+minimum time series required for a particular question of interest.
 
@@
 short-lived organism, 15 years was adequate to assess long-term trends
-in abundance. However, these studies, and other past work, have
-typically focused either on theoretical aspects of monitoring design or
-focused on only a few species.
+in abundance. In addition, Hatch (2003) examined seabird populations and
+determined the minimum time required to achieve high statistical power
+ranged between 11 and 69 years. However, these studies, and other past
+work, have typically focused either on theoretical aspects of monitoring
+design or focused on only a few species.
 
-In this paper, we use both simulations and a database of empirical time
-series to determine the minimum number of years required to address
-several questions. In the main manuscript, we estimate the minimum time
-series length required to assess long-term changes in abundance via
-simple linear regression. We then examine 878 population time series to
-estimate this minimum time series length required for detecting
-population trends. We expand on the results in the supplmentary material
-by examining other questions of interest.
+We use both simulations and a database of empirical time series to
+determine the minimum number of years required to address several
+questions. We estimate the minimum time series length required
+(\(T_{min}\)) to assess long-term changes in abundance via simple linear
+regression. First, we estimate \(T_{min}\) using a simulation approach.
+Then we examine 878 population time series to estimate \(T_{min}\). In
+the supplementary material, we determine \(T_{min}\) for related ideas:
+calculating long-term growth rates, using more complicated population
+models, varying statistical level and power, and the use of generalized
+additive models.
 

### 2017-08-16 writing/manuscript_draft.Rmd
updated fig captions and figs

@@
 As an example, we can take the following population model for population size $N$ at time $t$:
-$$
+\begin{equation}
 N(t + 1) = N(t) + r(t) \mbox{ with } r(t) \sim N(\mu, \sigma)
-$$
+\label{sim_model}
+\end{equation}
 where $\epsilon$ is a normally-distributed random noise term with mean $\mu$ and standard deviation $\sigma$.
@@
 the rate of increase).
+It is important to note that any population model could be subsituted for equation \ref{sim_model}, as we do in the supplementary material (Figs.
+\ref{fig:min_time_dist_log_pop}, \ref{fig:shark_example}).
 
-By simulating this model repetitively for different lengths of time, we can determine the minimum time series required.
+We can determine the minimum time series required by simulating this model repetitively for different lengths of time.
 Here, statistical power is simply the fraction of simulations with significant trends in abundance.
@@
 Statistical power of 0.8 would indicate that, if there was a true trend in abundance, we would have a 0.8 probability of detecting the trend.
+We also tested the effect of varying both the significance level and statistical power (Fig.
+\ref{fig:min_time_vs_alpha_beta}).
 
@@
 
-```{r,echo=F,eval=T,fig.cap='(a) example of simulated time series for 40 time periods and (b) statistical power versus the simulated time series length.
-The horizontal, dashed line is our desired statistical power of 0.8.
-The vertical, dashed line is the minimum time required to achieve our desired statistical power.
-(c) Minimum time required ($T_{min}$) for simulations with different values of the trend strength ($r$) and (d) different levels of population variability ($\\sigma$).
+```{r,echo=F,eval=T,fig.cap='(a) Example of a simulated time series for 40 time periods.
+(b) Statistical power versus the simulated time series length.
+The horizontal, dashed line is the desired statistical power of 0.8.
+The vertical, dashed line is the minimum time required to achieve the desired statistical power.
+(c) Minimum time required ($T_{min}$) for simulations with different values of the trend strength ($r$).
+(d) Minimum time required for different levels of population variability ($\\sigma$).
 In each case, the minimum time required is the minimum number of years to achieve 0.8 statistical power given a significance level of 0.05.\\label{fig:theoretical_approach}'}
@@
   # Plotting commands
-  par(mfrow=c(2,2),oma=c(1,2,0.5,0.5),mar=c(4,5,2.5,0.5))
+  par(mfrow=c(2,2),oma=c(1.5,2,0.5,0.5),mar=c(3,5,2.5,0.5))
     matplot(simulations,type='l',col=rgb(0.5,0.5,0.5,0.5),lty=1,ylab='',xlab='',las=1,cex.axis=1.2)
@@
     plot(years_to_sample,power,type='l',lwd=2,las=1,ylab='',xlab='',cex.axis=1.2)
-      abline(v=min_time_needed,lty=2,lwd=2,col='red')
-      abline(h=0.8,lty=2,lwd=2,col='red')
+      abline(v=min_time_needed,lty=2,lwd=2,col='black')
+      abline(h=0.8,lty=2,lwd=2,col='black')
       mtext(text = 'time sampled' ,side = 1,cex = 1.2,line=3)
@@
 
-```{r,echo=F,fig.cap='Distribution of the minimum time required in order to detect a significant trend (at the 0.05 level) in abundance given power of 0.8.\\label{fig:min_time_dist}'}
+```{r,echo=F,fig.cap='Distributions of the minimum time required for four different biological classes.
+The minimum time required calculation corresponds to a significant level of 0.05 and statistical power of 0.8.\\label{fig:min_time_dist}',fig.width=6}
 # Pull in analysis of average min time needed for **empirical species** (save analysis as Rdata file or something)
@@
 barplot(counts, main=" ",
-        xlab=" ", ylab="", col=1:4,
-        legend = rownames(counts),ylim=c(0,50),xlim=c(0,38),
-        las=1,cex.names =1.2,cex.axis = 1.2,density=35,
+        xlab=" ", ylab="", col=gray.colors(4),
+        ylim=c(0,50),xlim=c(0,40),
+        las=1,cex.names =1.2,cex.axis = 1.2,#density=35,
         angle=c(20,40,60,80))
+legend('topright',legend=rownames(counts),col = gray.colors(4),pch = 15,bty ="n")
 
@@
 
-```{r,echo=F, fig.cap='Minimum time required to estimate change in abundance correlated with (a) trend strength (absolute value of slope coefficient estimated from linear regression), (b) population variance (interannual variability in population size), and (c) temporal autocorrelation.\\label{fig:correlates}',fig.height=7, fig.width=4}
+```{r,echo=F, fig.cap='Minimum time required to estimate change in abundance correlated with (a) trend strength (absolute value of slope coefficient estimated from linear regression), (b) population variance (interannual variability in population size), and (c) temporal lag-1 autocorrelation.\\label{fig:correlates}',fig.height=7, fig.width=4}
 

### 2017-08-16 writing/manuscript_draft.tex
updated fig captions and figs

@@
 (e.g.~population variability). As an example, we can take the following
-population model for population size \(N\) at time \(t\): \[
+population model for population size \(N\) at time \(t\):
+
+\begin{equation}
 N(t + 1) = N(t) + r(t) \mbox{ with } r(t) \sim N(\mu, \sigma)
-\] where \(\epsilon\) is a normally-distributed random noise term with
-mean \(\mu\) and standard deviation \(\sigma\). The rate of growth \(r\)
-is also the trend strength of the increase or decrease (i.e.~the rate of
-increase).
+\label{sim_model}
+\end{equation}
 
-By simulating this model repetitively for different lengths of time, we
-can determine the minimum time series required. Here, statistical power
-is simply the fraction of simulations with significant trends in
+where \(\epsilon\) is a normally-distributed random noise term with mean
+\(\mu\) and standard deviation \(\sigma\). The rate of growth \(r\) is
+also the trend strength of the increase or decrease (i.e.~the rate of
+increase). It is important to note that any population model could be
+subsituted for equation \ref{sim_model}, as we do in the supplementary
+material (Figs. \ref{fig:min_time_dist_log_pop},
+\ref{fig:shark_example}).
+
+We can determine the minimum time series required by simulating this
+model repetitively for different lengths of time. Here, statistical
+power is simply the fraction of simulations with significant trends in
 abundance. A significant trend in abundance is determined by estimating
@@
 indicate that, if there was a true trend in abundance, we would have a
-0.8 probability of detecting the trend.
+0.8 probability of detecting the trend. We also tested the effect of
+varying both the significance level and statistical power (Fig.
+\ref{fig:min_time_vs_alpha_beta}).
 
@@
 \includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-1-1.pdf}
-\caption{(a) example of simulated time series for 40 time periods and
-(b) statistical power versus the simulated time series length. The
-horizontal, dashed line is our desired statistical power of 0.8. The
-vertical, dashed line is the minimum time required to achieve our
+\caption{(a) Example of a simulated time series for 40 time periods. (b)
+Statistical power versus the simulated time series length. The
+horizontal, dashed line is the desired statistical power of 0.8. The
+vertical, dashed line is the minimum time required to achieve the
 desired statistical power. (c) Minimum time required (\(T_{min}\)) for
-simulations with different values of the trend strength (\(r\)) and (d)
-different levels of population variability (\(\sigma\)). In each case,
-the minimum time required is the minimum number of years to achieve 0.8
-statistical power given a significance level of
-0.05.\label{fig:theoretical_approach}}
+simulations with different values of the trend strength (\(r\)). (d)
+Minimum time required for different levels of population variability
+(\(\sigma\)). In each case, the minimum time required is the minimum
+number of years to achieve 0.8 statistical power given a significance
+level of 0.05.\label{fig:theoretical_approach}}
 \end{figure}
@@
 \includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-4-1.pdf}
-\caption{Distribution of the minimum time required in order to detect a
-significant trend (at the 0.05 level) in abundance given power of
+\caption{Distributions of the minimum time required for four different
+biological classes. The minimum time required calculation corresponds to
+a significant level of 0.05 and statistical power of
 0.8.\label{fig:min_time_dist}}
@@
 estimated from linear regression), (b) population variance (interannual
-variability in population size), and (c) temporal
+variability in population size), and (c) temporal lag-1
 autocorrelation.\label{fig:correlates}}

### 2017-08-16 writing/manuscript_draft.Rmd
updated methods section

@@
 We only briefly discuss this approach as it has been described elsewhere.
-Essentially, we use a population model and simulate it for a number of years repetitively.
-We can then determine the minimum time series required to achieve designed levels of significance and power for a particular question.
+Essentially, we use a population model and repetitively simulate it for a number of years.
 This approach requires us to determine values for our model parameters (e.g.
@@
 As an example, we can take the following population model for population size $N$ at time $t$:
+
 \begin{equation}
@@
 
-We can determine the minimum time series required by simulating this model repetitively for different lengths of time.
-Here, statistical power is simply the fraction of simulations with significant trends in abundance.
-A significant trend in abundance is determined by estimating the significance of the slope parameter from simple linear regression.
-Then, the minimum time required ($T_{min}$) is the number of years required to obtain statistical power above some pre-determined threshold (historically at 0.8, @Cohen1992).
+Statistical power is then the fraction of simulations that meet some criteria.
+Here, we use whether the slope parameter from linear regression is significant.
+Statistical power changes with the simulation length (Fig.
+\ref{fig:theoretical_approach}b).
+Then, the minimum time required ($T_{min}$) is the number of years needed to obtain statistical power above some threshold (historically at 0.8, @Cohen1992).
 Statistical power of 0.8 would indicate that, if there was a true trend in abundance, we would have a 0.8 probability of detecting the trend.
@@
 ## Data source
-We use a database of `r length(table(dat$ID))` population time series compiled in [@Keith2015].
+We use a database of `r length(table(dat$ID))` population time series compiled in [@Keith2015] to compare the predictability of growth rates amongst populations.
 The data are originally from the Global Population Dynamics Database [@GPDD2010].
 We filtered out short time series (less than `r pop_info$series_length[1]` years), and those with missing data, and were left with `r nrow(pop_info)` time series.
-The database includes information on `r length(table(pop_info$Binomial))` vertebrate species with a focus on mammals, birds, and fish.
-The database also includes information on generation length and census specifications.
+The data includes information on `r length(table(pop_info$Binomial))` vertebrate species with a focus on mammals, birds, and fish.
+The data also includes information on generation length and census specifications.
 For each time series, we also calculated variables of interest like variance in population size, long-term trend in abundance (slope coefficient from simple linear regression), and temporal autocorrelation.
@@
 ```
-For a subset of populations in the database ($n =$ `r nrow(LPI_pop_info)`), we were able to connect a database on biological characteristics [@Myhrvold2015], including body size and generation length.
-All `r nrow(LPI_pop_info)` populations are within the Aves class.
-In the supplementary material, we explore how the minimum time required does not strongly correlate with any biological characteristics (Fig.
+For a subset of populations ($n =$ `r nrow(LPI_pop_info)`), we had information on biological characteristics from another paper [@Myhrvold2015], including body size and generation length.
+All `r nrow(LPI_pop_info)` populations were birds.
+We examine how the minimum time required is related to these biological characteristics (Fig.
 \ref{fig:biological_correlates}).
@@
 We first take all possible contiguous subsamples of each time series.
-For example, a time series of 40 years would have 39 possible contiguous subsamples of length 2, 38 possible contiguous subsamples of length 3, and continuing until 1 possible contiguous subsample of length 40 [@Gerber1999; @Giron-Nava2017].
-Then, we determine the $T_{min}$ required to be confident (i.e.
-have high enough statistical power) in an estimate of the slope coefficient from linear regression.
-Therefore, we run a linear regression for each subsampled time series.
-Then, we determine the fraction of subsamples of a particular length that have estimated slope coefficients which are statistically different from zero given the long-term, or ``true'', time series also has a significant slope.
-This is a measure of statistical power (often denoted as $1 - \beta$, where $\beta$ is the probability of a type II error).
-Lastly, we determine which subsample length is required to guarantee we have above a certain threshold of statistical power, which we will assume is 0.8 by convention [@Cohen1992].
+For example, a time series of 35 years would have 34 possible contiguous subsamples of length 2, 34 possible contiguous subsamples of length 3, and continuing until 1 possible contiguous subsample of length 35 [@Gerber1999; @Giron-Nava2017].
+Next, we run a linear regression for each subsampled time series.
+Then, we determine the fraction of subsamples of a particular length that have estimated slope coefficients which are statistically different from zero.
+We only look at the fraction of samples where the long-term, or ``true'', time series also has a significant slope.
+This fraction is a measure of statistical power.
+Lastly, we determine which subsample length is required to achieve a certain threshold of statistical power (0.8, @Cohen1992).
+The subsampled length meeting these criteria is the minimum time series length required ($T_{min}$).
 

### 2017-08-16 writing/manuscript_draft.tex
updated methods section

@@
 approach as it has been described elsewhere. Essentially, we use a
-population model and simulate it for a number of years repetitively. We
-can then determine the minimum time series required to achieve designed
-levels of significance and power for a particular question. This
-approach requires us to determine values for our model parameters
+population model and repetitively simulate it for a number of years.
+This approach requires us to determine values for our model parameters
 (e.g.~population variability). As an example, we can take the following
@@
 
-We can determine the minimum time series required by simulating this
-model repetitively for different lengths of time. Here, statistical
-power is simply the fraction of simulations with significant trends in
-abundance. A significant trend in abundance is determined by estimating
-the significance of the slope parameter from simple linear regression.
-Then, the minimum time required (\(T_{min}\)) is the number of years
-required to obtain statistical power above some pre-determined threshold
-(historically at 0.8, Cohen (1992)). Statistical power of 0.8 would
-indicate that, if there was a true trend in abundance, we would have a
-0.8 probability of detecting the trend. We also tested the effect of
-varying both the significance level and statistical power (Fig.
-\ref{fig:min_time_vs_alpha_beta}).
+Statistical power is then the fraction of simulations that meet some
+criteria. Here, we use whether the slope parameter from linear
+regression is significant. Statistical power changes with the simulation
+length (Fig. \ref{fig:theoretical_approach}b). Then, the minimum time
+required (\(T_{min}\)) is the number of years needed to obtain
+statistical power above some threshold (historically at 0.8, Cohen
+(1992)). Statistical power of 0.8 would indicate that, if there was a
+true trend in abundance, we would have a 0.8 probability of detecting
+the trend. We also tested the effect of varying both the significance
+level and statistical power (Fig. \ref{fig:min_time_vs_alpha_beta}).
 
@@
 We use a database of 2444 population time series compiled in (Keith et
-al. 2015). The data are originally from the Global Population Dynamics
+al. 2015) to compare the predictability of growth rates amongst
+populations. The data are originally from the Global Population Dynamics
 Database (NERC Centre for Population Biology 2010). We filtered out
 short time series (less than 35 years), and those with missing data, and
-were left with 878 time series. The database includes information on 478
-vertebrate species with a focus on mammals, birds, and fish. The
-database also includes information on generation length and census
+were left with 878 time series. The data includes information on 478
+vertebrate species with a focus on mammals, birds, and fish. The data
+also includes information on generation length and census
 specifications. For each time series, we also calculated variables of
@@
 
-For a subset of populations in the database (\(n =\) 547), we were able
-to connect a database on biological characteristics (Myhrvold et al.
-2015), including body size and generation length. All 547 populations
-are within the Aves class. In the supplementary material, we explore how
-the minimum time required does not strongly correlate with any
+For a subset of populations (\(n =\) 547), we had information on
+biological characteristics from another paper (Myhrvold et al. 2015),
+including body size and generation length. All 547 populations were
+birds. We examine how the minimum time required is related to these
 biological characteristics (Fig. \ref{fig:biological_correlates}).
@@
 We first take all possible contiguous subsamples of each time series.
-For example, a time series of 40 years would have 39 possible contiguous
-subsamples of length 2, 38 possible contiguous subsamples of length 3,
-and continuing until 1 possible contiguous subsample of length 40
-(Gerber, DeMaster, and Kareiva 1999; Giron-Nava et al. 2017). Then, we
-determine the \(T_{min}\) required to be confident (i.e.~have high
-enough statistical power) in an estimate of the slope coefficient from
-linear regression. Therefore, we run a linear regression for each
-subsampled time series. Then, we determine the fraction of subsamples of
-a particular length that have estimated slope coefficients which are
-statistically different from zero given the long-term, or ``true'', time
-series also has a significant slope. This is a measure of statistical
-power (often denoted as \(1 - \beta\), where \(\beta\) is the
-probability of a type II error). Lastly, we determine which subsample
-length is required to guarantee we have above a certain threshold of
-statistical power, which we will assume is 0.8 by convention (Cohen
-1992).
+For example, a time series of 35 years would have 34 possible contiguous
+subsamples of length 2, 34 possible contiguous subsamples of length 3,
+and continuing until 1 possible contiguous subsample of length 35
+(Gerber, DeMaster, and Kareiva 1999; Giron-Nava et al. 2017). Next, we
+run a linear regression for each subsampled time series. Then, we
+determine the fraction of subsamples of a particular length that have
+estimated slope coefficients which are statistically different from
+zero. We only look at the fraction of samples where the long-term, or
+``true'', time series also has a significant slope. This fraction is a
+measure of statistical power. Lastly, we determine which subsample
+length is required to achieve a certain threshold of statistical power
+(0.8, Cohen (1992)). The subsampled length meeting these criteria is the
+minimum time series length required (\(T_{min}\)).
 

### 2017-08-16 writing/manuscript_draft.Rmd
small changes in refs of figures

@@
 
+We also explored how the simulation approach can be applied to more biologically-realistic population models (Fig.
+\ref{fig:shark_example}).
+More specifically, we determine the minimum time required to estimate long-term population trends using a stochastic, age-structured model of lemon shark population dynamics in the Bahamas [@White2014].
+We found that over 27 years of continuous monitoring were needed in this particular scenario (Fig.
+\ref{fig:shark_example}).
+Similar to the simulation approach decribed above, the minimum time required for the lemon shark population was strongly dependent on model parameters.
+
 <!-- code to pull in databases-->
@@
 The subsampled length meeting these criteria is the minimum time series length required ($T_{min}$).
+
+In the supplementary material, we show how the same approach described here can be used to determine the minimum time required to estimate a population's geometric growth rate (Figs.
+\ref{fig:growth_rate},\ref{fig:min_time_growth_dist}).
+We also determine the minimum time required to estimate long-term trends according to generalized additive models, instead of the simple linear models used here (Fig.
+\ref{fig:gam_example}).
+We found that although we obtain a similar distribution of minimum times required for GAMs, the minimum time required for GAMs is on average shorter than for linear regression (Fig.
+\ref{fig:min_time_dist_gam}).
 
@@
 We limited our analyses to populations with at least 35 years of continuous sampling.
+We examined the minimum time required to estimate long-term trends via linear regression.
 
@@
 
+<!--Integrate with other sections
 ## Questions in the supplmentary material
@@
 We expand on these results in the supplementary material.
+-->
 
@@
 Therefore, life-history characteristics may be more important for other questions (e.g.
-extinction risk) where estimates minimum time required are needed.
+species extinction risk) where estimates minimum time required are needed.
 
@@
 Dallas and E.
-Malcolm-White<!--X anonymous reviewers -->for their helpful comments.
+Malcolm-White <!--X anonymous reviewers -->for their helpful comments.
 

### 2017-08-16 writing/manuscript_draft.tex
small changes in refs of figures

@@
 
+We also explored how the simulation approach can be applied to more
+biologically-realistic population models (Fig. \ref{fig:shark_example}).
+More specifically, we determine the minimum time required to estimate
+long-term population trends using a stochastic, age-structured model of
+lemon shark population dynamics in the Bahamas (White, Nagy, and Gruber
+2014). We found that over 27 years of continuous monitoring were needed
+in this particular scenario (Fig. \ref{fig:shark_example}). Similar to
+the simulation approach decribed above, the minimum time required for
+the lemon shark population was strongly dependent on model parameters.
+
 \subsection{Data source}\label{data-source}
@@
 minimum time series length required (\(T_{min}\)).
+
+In the supplementary material, we show how the same approach described
+here can be used to determine the minimum time required to estimate a
+population's geometric growth rate (Figs.
+\ref{fig:growth_rate},\ref{fig:min_time_growth_dist}). We also determine
+the minimum time required to estimate long-term trends according to
+generalized additive models, instead of the simple linear models used
+here (Fig. \ref{fig:gam_example}). We found that although we obtain a
+similar distribution of minimum times required for GAMs, the minimum
+time required for GAMs is on average shorter than for linear regression
+(Fig. \ref{fig:min_time_dist_gam}).
 
@@
 variety of life history characteristics. We limited our analyses to
-populations with at least 35 years of continuous sampling.
+populations with at least 35 years of continuous sampling. We examined
+the minimum time required to estimate long-term trends via linear
+regression.
 
@@
 are known to fluctuate (i.e.~the variability in population size).
-
-\subsection{Questions in the supplmentary
-material}\label{questions-in-the-supplmentary-material}
-
-In the supplementary material, we explore several additional questions
-related to statistical power and estimating \(T_{min}\). First, we
-showed that other metrics besides estimating the slope coefficient from
-linear regression could be used. More specifically, we show how the same
-approach described here can be used to determine the minimum time
-required to estimate a population's geometric growth rate (Figs.
-\ref{fig:growth_rate},\ref{fig:min_time_growth_dist}). We also explored
-how the simulation approach can be applied to more
-biologically-realistic population models (Fig. \ref{fig:shark_example}).
-More specifically, we determine the minimum time required to estimate
-long-term population trends using a stochastic, age-structured model of
-lemon shark population dynamics in the Bahamas (White, Nagy, and Gruber
-2014). We found that over 27 years of continuous monitoring were needed
-in this particular scenario (Fig. \ref{fig:shark_example}. Lastly, we
-determine the minimum time required to estimate long-term trends
-according to generalized additive models, instead of the simple linear
-models used here (Fig. \ref{fig:gam_example}). We found that although we
-obtain a similar distribution of minimum times required for GAMs, the
-minimum time required for GAMs is on average shorter than for linear
-regression (Fig. \ref{fig:min_time_dist_gam}). We expand on these
-results in the supplementary material.
 
@@
 over time. Therefore, life-history characteristics may be more important
-for other questions (e.g.~extinction risk) where estimates minimum time
-required are needed.
+for other questions (e.g.~species extinction risk) where estimates
+minimum time required are needed.
 

### 2017-08-17 writing/manuscript_draft.Rmd
rewrote results section

@@
 geometry: margin=1in
-header-includes: \usepackage{float} \usepackage{setspace}\doublespacing \usepackage{lineno} \usepackage[firstinits=true,style=authoryear-comp,natbib=true,doi=false,isbn=false,url=false,uniquename=false,uniquelist=false,sorting=none]{biblatex}
+header-includes: \usepackage{float} \usepackage{lineno} \usepackage[firstinits=true,style=authoryear-comp,natbib=true,doi=false,isbn=false,url=false,uniquename=false,uniquelist=false,sorting=none]{biblatex}
 
@@
 
-Addresses:\emph{$^1$ Center for Population Biology, University of California-Davis, Davis, California 95616 USA}
+Address:\emph{
+    \\$^1$Center for Population Biology \\
+    University of California, Davis \\
+    2320 Storer Hall \\
+		University of California, Davis \\
+		One Shields Avenue Davis, CA, USA}
 \vspace{3 mm}
 
-$^2$Corresponding author: eawhite@ucdavis.edu
+$^2$Corresponding author: eawhite@ucdavis.edu, 1 480 203 7931
 \vspace{3 mm}
 
-Number of words: X,XXX (less than 4,000 words or 20 pages)
+Number of words (including 300 words for figures): X,XXX 
+Number of figures:
+Number of figures in supplementary material:
 \vspace{3 mm}
@@
 Available as a preprint at: www.peerj.com
+\vspace{3 mm}
+
+Color does not need to be used for any figures in print
 \vspace{3 mm}
@@
 
-# Abstract
+\begin{abstract}
 Long-term census data are necessary to better understand ecological processes and make management decisions.
@@
 <!--We stress the need for verifying a study has enough statistical power to accurately determine long-term population trends.--> Time series less than 10-15 years are likely under-powered and potentially misleading.
+\end{abstract}
 
@@
 
-We also explored how the simulation approach can be applied to more biologically-realistic population models (Fig.
-\ref{fig:shark_example}).
-More specifically, we determine the minimum time required to estimate long-term population trends using a stochastic, age-structured model of lemon shark population dynamics in the Bahamas [@White2014].
-We found that over 27 years of continuous monitoring were needed in this particular scenario (Fig.
-\ref{fig:shark_example}).
-Similar to the simulation approach decribed above, the minimum time required for the lemon shark population was strongly dependent on model parameters.
-
 <!-- code to pull in databases-->
@@
 We determined the minimum time series length ($T_{min}$) required to address a particular question of interest.
-More specifically we asked, what is the minimum time series length required to determine, via linear regression, the long-term population trend?
-Here, the minimum time series length required had high enough statistical power (greater than 0.8) for a set significance level of 0.05.
-We investigated this question through two approaches: simulation-based and empirical analyses.
+What is the minimum time series length required to determine, via linear regression, the long-term population trend?
+Here, the minimum time series length required had high enough statistical power ($1-\beta$ greater than 0.8) for a set significance level ($\alpha$) of 0.05.
+It is also possible to alter $1-\beta$ and $\alpha$.
+Predictably, as we increased our level of power or decreased $\alpha$, $T_{min}$ increased (Fig.
+\ref{fig:min_time_vs_alpha_beta}).
+We investigated this question through two approaches.
 We briefly describe results from the simulation approach and then discuss our empirical approach.
@@
 
-We constructed a general population model where the trend (i.e.
+We constructed a general population model where the trend strength (i.e.
 slope coefficient) over time could be a model parameter.
-We chose a simple model, but any other population model could be used (Fig.
+We then ran simulated time series of different lengths.
+From these simulations we determined the minimum time series length required to achieve a certain level of statistical power.
+In line with past work [@Gerrodette1987], we found the $T_{min}$ increases (i.e.
+more time is required) with decreases in trend strength and with increases in population variability (Figs.
+\ref{fig:theoretical_approach}c,d).
+
+We chose a simple model, but any other population model could be used (see example in Fig.
 \ref{fig:min_time_dist_log_pop}).
 Ideally, the specific model choice should be tailored to the population of interest.
-We then ran simulated time series of different lengths.
-From these simulations we determined the minimum time series length required to achieve a certain level of statistical power.
-We were also able to alter the parameter values (trend strength and population variability) of the population model to see how this altered $T_{min}$ (Fig.
-\ref{fig:theoretical_approach}).
-
-In line with past work [@Gerrodette1987], we found the $T_{min}$ increases (i.e.
-more time is required) with decreases in trend strength and with increases in variability (Fig.
-\ref{fig:theoretical_approach}).
+We explored how the simulation approach can be applied to more biologically-realistic population models (Fig.
+\ref{fig:shark_example}).
+More specifically, we determined the minimum time required to estimate long-term population trends using a stochastic, age-structured model of lemon shark population dynamics in the Bahamas [@White2014].
+We found that over 27 years of continuous monitoring were needed in this particular scenario (Fig.
+\ref{fig:shark_example}).
+Similar to the simulation approach decribed above, the minimum time required for the lemon shark population was strongly dependent on model parameters.
 
@@
 We examined a database of `r nrow(pop_info)` separate population time series representing `r length(table(pop_info$Binomial))` species.
-This database consists of species with a variety of life history characteristics.
+This database consists of vertebrate species with a variety of life history characteristics (Fig.
+\ref{fig:biological_correlates}).
 We limited our analyses to populations with at least 35 years of continuous sampling.
-We examined the minimum time required to estimate long-term trends via linear regression.
+We then examined the minimum time required to estimate long-term trends via linear regression.
 
@@
 \ref{fig:min_time_dist}).
-In addition, as expected based on our simulation results and those of others [@Rhodes2011], the minimum time series length required is strongly correlated with trend strength, population variance, and autocorrelation (Fig \ref{fig:correlates}).
-Using a generalized linear model, with a Poisson error structure, all three of these explanatory variables were significant and had large effect sizes (see Table S1).
-Combined, trend strength, population variance, and autocorrelation account for about `r round(100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance),1)` of the explained deviance, which is similar to $R^2$ [@Zuur2009], in minimum time series length required.
-
-We also examined a subset of populations where more explanatory variables were known (e.g.
-body mass, longevity, location).
+Estimates of $T_{min}$ varied between biological class (Fig.
+\ref{fig:min_time_dist}).
+Ray-finned fish (class Actinopterygii) typically had estimates of $T_{min}$ over 20 years.
+Birds (class Aves) had a much wider distribution of $T_{min}$, but usually required less years of sampling.
+Differences between these classes were explained by differences in interannual variability in population size (Fig.
+\ref{fig:class}).
+We also examined a subset of populations where life-history characteristics were known.
 None of these explanatory variables were predictive of the minimum time series length required (Fig.
 \ref{fig:biological_correlates}).
-Some differences in $T_{min}$ did occur between biological class (Fig.
-\ref{fig:class}), but this can be attributed to the amount populations are known to fluctuate (i.e.
-the variability in population size).
+
+The minimum time series length required was strongly correlated with trend strength, population variance, and autocorrelation (Fig.
+\ref{fig:correlates}).
+This is in line based on our simulations and those of others [@Rhodes2011].
+Using a generalized linear model, with a Poisson error structure, all three of these explanatory variables were significant and had large effect sizes (see Table A1).
+Combined, trend strength, population variance, and autocorrelation account for about `r round(100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance),1)`% of the explained deviance, which is similar to $R^2$ [@Zuur2009], in minimum time series length required.
+
+Lastly, we tested model sensitivity by using generalized additive models (GAMs) instead of simple linear regression.
+Again, we examined the minimum time required to estimate long-term population trends (Fig.
+\ref{fig:gam_example}).
+We found that although we obtain a similar distribution of minimum times required for GAMs, the minimum time required for GAMs is on average shorter than for linear regression (Fig.
+\ref{fig:min_time_dist_gam}).
 
@@
 ```
-
-<!--Integrate with other sections
-## Questions in the supplmentary material
-
-In the supplementary material, we explore several additional questions related to statistical power and estimating $T_{min}$.
-First, we showed that other metrics besides estimating the slope coefficient from linear regression could be used.
-More specifically, we show how the same approach described here can be used to determine the minimum time required to estimate a population's geometric growth rate (Figs.
-\ref{fig:growth_rate},\ref{fig:min_time_growth_dist}).
-We also explored how the simulation approach can be applied to more biologically-realistic population models (Fig.
-\ref{fig:shark_example}).
-More specifically, we determine the minimum time required to estimate long-term population trends using a stochastic, age-structured model of lemon shark population dynamics in the Bahamas [@White2014].
-We found that over 27 years of continuous monitoring were needed in this particular scenario (Fig.
-\ref{fig:shark_example}.
-Lastly, we determine the minimum time required to estimate long-term trends according to generalized additive models, instead of the simple linear models used here (Fig.
-\ref{fig:gam_example}).
-We found that although we obtain a similar distribution of minimum times required for GAMs, the minimum time required for GAMs is on average shorter than for linear regression (Fig.
-\ref{fig:min_time_dist_gam}).
-We expand on these results in the supplementary material.
--->
 
@@
  \setcounter{figure}{0}
- \renewcommand{\thefigure}{S\arabic{figure}}
+ \renewcommand{\thefigure}{A\arabic{figure}}
 

### 2017-08-17 writing/manuscript_draft.tex
rewrote results section

@@
 
-\usepackage{float}
-\usepackage{setspace}\doublespacing \usepackage{lineno}
+\usepackage{float} \usepackage{lineno}
 \usepackage[firstinits=true,style=authoryear-comp,natbib=true,doi=false,isbn=false,url=false,uniquename=false,uniquelist=false,sorting=none]{biblatex}
@@
 
-Addresses:\emph{$^1$ Center for Population Biology, University of California-Davis, Davis, California 95616 USA}
-\vspace{3 mm}
+Address:\emph{
+    \\$^1$Center for Population Biology \\
+    University of California, Davis \\
+    2320 Storer Hall \\
+        University of California, Davis \\
+        One Shields Avenue Davis, CA, USA} \vspace{3 mm}
 
 \(^2\)Corresponding author:
-\href{mailto:eawhite@ucdavis.edu}{\nolinkurl{eawhite@ucdavis.edu}}
-\vspace{3 mm}
+\href{mailto:eawhite@ucdavis.edu}{\nolinkurl{eawhite@ucdavis.edu}}, 1
+480 203 7931 \vspace{3 mm}
 
-Number of words: X,XXX (less than 4,000 words or 20 pages) \vspace{3 mm}
+Number of words (including 300 words for figures): X,XXX Number of
+figures: Number of figures in supplementary material: \vspace{3 mm}
 
@@
 Available as a preprint at: www.peerj.com \vspace{3 mm}
+
+Color does not need to be used for any figures in print \vspace{3 mm}
 
@@
 
-\section{Abstract}\label{abstract}
-
-Long-term census data are necessary to better understand ecological
-processes and make management decisions. However, census data are often
-expensive, requiring a lot of time and resources. Little work has
-addressed the question: when is a time series of census data long enough
-to address a question of interest? We determine the minimum time series
-length required to estimate significant increases or decreases in
-population abundance. Importantly, we examine the ability to detect a
-trend with a set level of statistical power, which is often neglected in
-ecological time series analyses. Using a simulation approach, we
-demonstrate how the minimum time series length required increases with
-weaker trends in abundance and with higher population variability. In
-addition, we examine 878 populations of vertebrate species to determine
-the minimum time required to detect changes in their abundance. We show
-that 10-15 years of continuous monitoring are required in order to
-achieve a high level of statistical power. Similar to
-simulation-approaches, the minimum time required for field census data
-strongly depends on trend strength, population variability, and temporal
-autocorrelation. These results point to the importance of sampling
-populations over long periods of time. Time series less than 10-15 years
-are likely under-powered and potentially misleading.
+\begin{abstract}
+Long-term census data are necessary to better understand ecological processes and make management decisions.
+However, census data are often expensive, requiring a lot of time and resources.
+Little work has addressed the question: when is a time series of census data long enough to address a question of interest?
+<!--Here, we explore two approaches to address this question: simulations and an empirical approach.--> We determine the minimum time series length required to estimate significant increases or decreases in population abundance.
+Importantly, we examine the ability to detect a trend with a set level of statistical power, which is often neglected in ecological time series analyses.
+Using a simulation approach, we demonstrate how the minimum time series length required increases with weaker trends in abundance and with higher population variability.
+In addition, we examine 878 populations of vertebrate species to determine the minimum time required to detect changes in their abundance.
+We show that 10-15 years of continuous monitoring are required in order to achieve a high level of statistical power.
+Similar to simulation-approaches, the minimum time required for field census data strongly depends on trend strength, population variability, and temporal autocorrelation.
+<!--Perhaps surprisingly, the minimum time required did not correlate well with biological explanatory variables, like body size or generation length.--> These results point to the importance of sampling populations over long periods of time.
+<!--We stress the need for verifying a study has enough statistical power to accurately determine long-term population trends.--> Time series less than 10-15 years are likely under-powered and potentially misleading.
+\end{abstract}
 
@@
 
-We also explored how the simulation approach can be applied to more
-biologically-realistic population models (Fig. \ref{fig:shark_example}).
-More specifically, we determine the minimum time required to estimate
-long-term population trends using a stochastic, age-structured model of
-lemon shark population dynamics in the Bahamas (White, Nagy, and Gruber
-2014). We found that over 27 years of continuous monitoring were needed
-in this particular scenario (Fig. \ref{fig:shark_example}). Similar to
-the simulation approach decribed above, the minimum time required for
-the lemon shark population was strongly dependent on model parameters.
-
 \subsection{Data source}\label{data-source}
@@
 We determined the minimum time series length (\(T_{min}\)) required to
-address a particular question of interest. More specifically we asked,
-what is the minimum time series length required to determine, via linear
-regression, the long-term population trend? Here, the minimum time
-series length required had high enough statistical power (greater than
-0.8) for a set significance level of 0.05. We investigated this question
-through two approaches: simulation-based and empirical analyses. We
-briefly describe results from the simulation approach and then discuss
-our empirical approach.
+address a particular question of interest. What is the minimum time
+series length required to determine, via linear regression, the
+long-term population trend? Here, the minimum time series length
+required had high enough statistical power (\(1-\beta\) greater than
+0.8) for a set significance level (\(\alpha\)) of 0.05. It is also
+possible to alter \(1-\beta\) and \(\alpha\). Predictably, as we
+increased our level of power or decreased \(\alpha\), \(T_{min}\)
+increased (Fig. \ref{fig:min_time_vs_alpha_beta}). We investigated this
+question through two approaches. We briefly describe results from the
+simulation approach and then discuss our empirical approach.
 
@@
 
-We constructed a general population model where the trend (i.e.~slope
-coefficient) over time could be a model parameter. We chose a simple
-model, but any other population model could be used (Fig.
-\ref{fig:min_time_dist_log_pop}). Ideally, the specific model choice
-should be tailored to the population of interest. We then ran simulated
-time series of different lengths. From these simulations we determined
-the minimum time series length required to achieve a certain level of
-statistical power. We were also able to alter the parameter values
-(trend strength and population variability) of the population model to
-see how this altered \(T_{min}\) (Fig. \ref{fig:theoretical_approach}).
+We constructed a general population model where the trend strength
+(i.e.~slope coefficient) over time could be a model parameter. We then
+ran simulated time series of different lengths. From these simulations
+we determined the minimum time series length required to achieve a
+certain level of statistical power. In line with past work (Gerrodette
+1987), we found the \(T_{min}\) increases (i.e.~more time is required)
+with decreases in trend strength and with increases in population
+variability (Figs. \ref{fig:theoretical_approach}c,d).
 
-In line with past work (Gerrodette 1987), we found the \(T_{min}\)
-increases (i.e.~more time is required) with decreases in trend strength
-and with increases in variability (Fig. \ref{fig:theoretical_approach}).
+We chose a simple model, but any other population model could be used
+(see example in Fig. \ref{fig:min_time_dist_log_pop}). Ideally, the
+specific model choice should be tailored to the population of interest.
+We explored how the simulation approach can be applied to more
+biologically-realistic population models (Fig. \ref{fig:shark_example}).
+More specifically, we determined the minimum time required to estimate
+long-term population trends using a stochastic, age-structured model of
+lemon shark population dynamics in the Bahamas (White, Nagy, and Gruber
+2014). We found that over 27 years of continuous monitoring were needed
+in this particular scenario (Fig. \ref{fig:shark_example}). Similar to
+the simulation approach decribed above, the minimum time required for
+the lemon shark population was strongly dependent on model parameters.
 
@@
 We examined a database of 878 separate population time series
-representing 478 species. This database consists of species with a
-variety of life history characteristics. We limited our analyses to
-populations with at least 35 years of continuous sampling. We examined
-the minimum time required to estimate long-term trends via linear
+representing 478 species. This database consists of vertebrate species
+with a variety of life history characteristics (Fig.
+\ref{fig:biological_correlates}). We limited our analyses to populations
+with at least 35 years of continuous sampling. We then examined the
+minimum time required to estimate long-term trends via linear
 regression.
@@
 series length required (\(T_{min}\)) of 16.5 (\(\sigma\)=8.4), with a
-wide distribution (Fig. \ref{fig:min_time_dist}). In addition, as
-expected based on our simulation results and those of others (Rhodes and
-Jonzen 2011), the minimum time series length required is strongly
-correlated with trend strength, population variance, and autocorrelation
-(Fig \ref{fig:correlates}). Using a generalized linear model, with a
-Poisson error structure, all three of these explanatory variables were
-significant and had large effect sizes (see Table S1). Combined, trend
-strength, population variance, and autocorrelation account for about
-72.6 of the explained deviance, which is similar to \(R^2\) (Zuur et al.
-2009), in minimum time series length required.
+wide distribution (Fig. \ref{fig:min_time_dist}). Estimates of
+\(T_{min}\) varied between biological class (Fig.
+\ref{fig:min_time_dist}). Ray-finned fish (class Actinopterygii)
+typically had estimates of \(T_{min}\) over 20 years. Birds (class Aves)
+had a much wider distribution of \(T_{min}\), but usually required less
+years of sampling. Differences between these classes were explained by
+differences in interannual variability in population size (Fig.
+\ref{fig:class}). We also examined a subset of populations where
+life-history characteristics were known. None of these explanatory
+variables were predictive of the minimum time series length required
+(Fig. \ref{fig:biological_correlates}).
 
-We also examined a subset of populations where more explanatory
-variables were known (e.g.~body mass, longevity, location). None of
-these explanatory variables were predictive of the minimum time series
-length required (Fig. \ref{fig:biological_correlates}). Some differences
-in \(T_{min}\) did occur between biological class (Fig.
-\ref{fig:class}), but this can be attributed to the amount populations
-are known to fluctuate (i.e.~the variability in population size).
+The minimum time series length required was strongly correlated with
+trend strength, population variance, and autocorrelation (Fig.
+\ref{fig:correlates}). This is in line based on our simulations and
+those of others (Rhodes and Jonzen 2011). Using a generalized linear
+model, with a Poisson error structure, all three of these explanatory
+variables were significant and had large effect sizes (see Table A1).
+Combined, trend strength, population variance, and autocorrelation
+account for about 72.6\% of the explained deviance, which is similar to
+\(R^2\) (Zuur et al. 2009), in minimum time series length required.
+
+Lastly, we tested model sensitivity by using generalized additive models
+(GAMs) instead of simple linear regression. Again, we examined the
+minimum time required to estimate long-term population trends (Fig.
+\ref{fig:gam_example}). We found that although we obtain a similar
+distribution of minimum times required for GAMs, the minimum time
+required for GAMs is on average shorter than for linear regression (Fig.
+\ref{fig:min_time_dist_gam}).
 
@@
 group at the University of California, Davis for their insight. We would
-also like to thank T. Dallas and E. Malcolm-Whitefor their helpful
+also like to thank T. Dallas and E. Malcolm-White for their helpful
 comments.
@@
 
-\setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}
+\setcounter{figure}{0} \renewcommand{\thefigure}{A\arabic{figure}}
 

### 2017-08-17 writing/supp_mat_draft.Rmd
small changes in refs of figures
rewrote results section

@@
     \renewcommand{\thesection}{S\arabic{section}}  
-    \renewcommand{\thetable}{S\arabic{table}}  
-    \renewcommand{\thefigure}{S\arabic{figure}}
+    \renewcommand{\thetable}{A\arabic{table}}  
+    \renewcommand{\thefigure}{A\arabic{figure}}
     \usepackage[round]{natbib} \bibpunct[; ]{(}{)}{,}{a}{}{;}
@@
 
-Code for all the figures and tables can be found at (https://github.com/erwhite1).
+Data and code for all the figures and tables can be found at (https://github.com/erwhite1).
 
@@
 Here, we use a generalized linear model framework with a Poisson error structure to determine drivers of the minimum time required.
-In figure, \ref{fig:poisson_model} we show a set of residual plots for the regression.
+In figure \ref{fig:poisson_model}, we show a set of residual plots for the regression.
 We then show the coefficient estimates and levels of significance in table \ref{table:model_output}.

### 2017-08-17 writing/supp_mat_draft.tex
rewrote results section

@@
 \renewcommand{\thesection}{S\arabic{section}}
-\renewcommand{\thetable}{S\arabic{table}}
-\renewcommand{\thefigure}{S\arabic{figure}} \usepackage[round]{natbib}
+\renewcommand{\thetable}{A\arabic{table}}
+\renewcommand{\thefigure}{A\arabic{figure}} \usepackage[round]{natbib}
 \bibpunct[; ]{(}{)}{,}{a}{}{;}
@@
 
-Code for all the figures and tables can be found at
+Data and code for all the figures and tables can be found at
 (\url{https://github.com/erwhite1}).
@@
 framework with a Poisson error structure to determine drivers of the
-minimum time required. In figure, \ref{fig:poisson_model} we show a set
+minimum time required. In figure \ref{fig:poisson_model}, we show a set
 of residual plots for the regression. We then show the coefficient

### 2017-08-19 writing/manuscript_draft.Rmd
Final version of manuscript for PeerJ preprint

@@
 
-Running head: 
-\vspace{3 mm}
-
 Title: Minimum time required to detect population trends: the need for long-term monitoring programs
@@
 
-Author(s): \textsc{Easton R. White$^{1,2}$} 
+Author: \textsc{Easton R. White$^{1,2}$} 
 \vspace{3 mm}
@@
 
-$^2$Corresponding author: eawhite@ucdavis.edu, 1 480 203 7931
+$^2$Corresponding author: eawhite@ucdavis.edu <!--, 1 480 203 7931-->
 \vspace{3 mm}
 
-Number of words (including 300 words for figures): X,XXX 
-Number of figures:
-Number of figures in supplementary material:
+Number of words (including 300 words per figure):  5900 
+\vspace{1 mm}
+
+Number of figures: 3 
+\vspace{1 mm}
+
+Supplementary material: 11 figures and 1 table 
 \vspace{3 mm}
 
-To be submitted to: \emph{Biological Conservation}
+Data and code for all the figures and tables can be found at (https://github.com/erwhite1/time-series-project).
+
+<!--To be submitted to: \emph{Biological Conservation}
 \vspace{3 mm}
@@
 Color does not need to be used for any figures in print
-\vspace{3 mm}
-
-Keywords: ecological time series, experimental design, monitoring, power analysis, statistical power, sampling design
+\vspace{3 mm}-->
 
@@
 \begin{abstract}
-Long-term census data are necessary to better understand ecological processes and make management decisions.
-However, census data are often expensive, requiring a lot of time and resources.
-Little work has addressed the question: when is a time series of census data long enough to address a question of interest?
-<!--Here, we explore two approaches to address this question: simulations and an empirical approach.--> We determine the minimum time series length required to estimate significant increases or decreases in population abundance.
-Importantly, we examine the ability to detect a trend with a set level of statistical power, which is often neglected in ecological time series analyses.
-Using a simulation approach, we demonstrate how the minimum time series length required increases with weaker trends in abundance and with higher population variability.
-In addition, we examine 878 populations of vertebrate species to determine the minimum time required to detect changes in their abundance.
-We show that 10-15 years of continuous monitoring are required in order to achieve a high level of statistical power.
-Similar to simulation-approaches, the minimum time required for field census data strongly depends on trend strength, population variability, and temporal autocorrelation.
-<!--Perhaps surprisingly, the minimum time required did not correlate well with biological explanatory variables, like body size or generation length.--> These results point to the importance of sampling populations over long periods of time.
-<!--We stress the need for verifying a study has enough statistical power to accurately determine long-term population trends.--> Time series less than 10-15 years are likely under-powered and potentially misleading.
+Long-term time series are necessary to better understand population dynamics, assess species' conservation status, and make management decisions.
+However, population data are often expensive, requiring a lot of time and resources.
+When is a population time series long enough to address a question of interest?
+We determine the minimum time series length required to detect significant increases or decreases in population abundance.
+To address this question, we use simulation methods and examine 878 populations of vertebrate species.
+Here we show that 15-20 years of continuous monitoring are required in order to achieve a high level of statistical power.
+For both simulations and the time series data, the minimum time required depends on trend strength, population variability, and temporal autocorrelation.
+These results point to the importance of sampling populations over long periods of time.
+We argue that statistical power needs to be considered in monitoring program design and evaluation.
+Time series less than 15-20 years are likely underpowered and potentially misleading.
 \end{abstract}
+
+Keywords: ecological time series, experimental design, monitoring, power analysis, statistical power, sampling design
 
@@
 Observational studies and population time series have become a cornerstone of modern ecological research and conservation biology [@Magurran2010; @Hughes2017].
-Long-term data are necessary to both understand population dynamics and to assess species extinction risk (citations).
+Long-term data are necessary to both understand population dynamics and to assess species extinction risk.
 Even though many time series may now be considered "long-term" (e.g.
-continuous plankton recorder @Giron-Nava2017), most are still short.
-Time series are typically short for a variety of reasons.
+continuous plankton recorder, @Giron-Nava2017), most are still short.
+Time series are typically short for a variety of reasons [@Field2007].
 They are often coupled with an experiment, which may only last a couple of years.
@@
 How long of a time series is actually necessary?
-This question has important implications for both research and management.
+This question has important implications for both research and management [@Nichols2006].
 Scientists need to know the time series length required to address a specific question.
-A short time series may lead to wrong conclusions given large natural year-to-year variability (citations).
+A short time series may lead to wrong conclusions given large natural year-to-year variability.
 Managers need to know when action is needed for a population.
@@
 The IUCN Red List Categories and Criteria suggest, under Criterion A2, a species qualifies as vulnerable if it has experienced a 30% decline over 10 years, or 3 generations [@IUCN2012].
-For both scientific and management questions, because sampling is typically expensive, we also do not want to sample for longer than is necessary [@Gerber1999].
+For both scientific and management questions, because sampling is typically expensive, we also do not want to sample for longer than is necessary.
+For example, @Gerber1999 investigated the minimum time series required to estimate population growth of the endangered, but recovering, eastern North Pacific gray whale (\emph{Eschrichtius robustus}).
+They used a long-term census to retroactively determine the minimum time series required to assess threat status.
+They found that only 11 years were needed, eight years before the delisting decision was made.
+This highlights the importance of estimating the minimum time series required as an earlier decision would have saved time and money [@Gerber1999].
 
@@
 For example, if a time series was assessed as significantly increasing or decreasing---when there was no true significant trend---this would be a false positive.
-The false positive rate, or significance level ($\alpha$) is often set at 0.05.
-A type II error ($\beta$) is failure to detect a true trend, or failure to reject a false null hypothesis.
+The false positive rate, or significance level ($\alpha$) is often set at 0.05 (although this is purely historical, @Mapstone1995).
+A type II error ($\beta$) is a failure to detect a true trend, or failure to reject a false null hypothesis.
 Formally, statistical power ($1-\beta$) is one minus the probability of a type II error ($\beta$).
@@
 It is then possible to determine how power changes with a variable of interest.
-For example, time series are simulated for different lengths of time.
+For example, time series can be simulated for different lengths of time.
 From these simulations, the minimum time series length required is calculated to meet certain levels of statistical significance and power [@Bolker2008].
@@
 Subsamples of different lengths can then be evaluated to see which fraction of subsamples meet some criteria, again a measure of statistical power.
-Similar to the simulation approach, we can use these measures of power to determine the minimum time series required for a particular question of interest.
+Similar to the simulation approach, these measures of power can be used to determine the minimum time series required for a particular question of interest.
 
-Past work has investigated questions related to the minimum time series required to estimate trends in population size over time [@Giron-Nava2017].
+Past work has investigated questions related to the minimum time series required to estimate trends in population size over time [@Wagner2009; @Giron-Nava2017].
 For example, @Rhodes2011 examined the optimal allocation of effort between spatial and temporal replicates.
@@
 They found that for this short-lived organism, 15 years was adequate to assess long-term trends in abundance.
-In addition, @Hatch2003 examined seabird populations and determined the minimum time required to achieve high statistical power ranged between 11 and 69 years.
+@Hatch2003 examined seabird populations and determined the minimum time required to achieve high statistical power ranged between 11 and 69 years.
 However, these studies, and other past work, have typically focused either on theoretical aspects of monitoring design or focused on only a few species.
 
-We use both simulations and a database of empirical time series to determine the minimum number of years required to address several questions.
+We use both simulations and empirical time series to determine the minimum number of years required to address several questions.
 We estimate the minimum time series length required ($T_{min}$) to assess long-term changes in abundance via simple linear regression.
@@
 \begin{equation}
-N(t + 1) = N(t) + r(t) \mbox{ with } r(t) \sim N(\mu, \sigma)
+N(t + 1) = N(t) + r(t) + \epsilon \mbox{ with } \epsilon \sim N(\mu, \sigma)
 \label{sim_model}
@@
 the rate of increase).
-It is important to note that any population model could be subsituted for equation \ref{sim_model}, as we do in the supplementary material (Figs.
+It is important to note that any population model could be substituted for equation \ref{sim_model}, as we do in the supplementary material (Figs.
 \ref{fig:min_time_dist_log_pop}, \ref{fig:shark_example}).
@@
 Statistical power is then the fraction of simulations that meet some criteria.
-Here, we use whether the slope parameter from linear regression is significant.
-Statistical power changes with the simulation length (Fig.
-\ref{fig:theoretical_approach}b).
-Then, the minimum time required ($T_{min}$) is the number of years needed to obtain statistical power above some threshold (historically at 0.8, @Cohen1992).
+Here, our criteria is whether the slope parameter from linear regression is significant at the 0.05 threshold.
 Statistical power of 0.8 would indicate that, if there was a true trend in abundance, we would have a 0.8 probability of detecting the trend.
@@
 ## Data source
-We use a database of `r length(table(dat$ID))` population time series compiled in [@Keith2015] to compare the predictability of growth rates amongst populations.
+We use a database of `r length(table(dat$ID))` population time series compiled in [@Keith2015]; they compared the predictability of growth rates among populations.
 The data are originally from the Global Population Dynamics Database [@GPDD2010].
@@
 For each time series, we also calculated variables of interest like variance in population size, long-term trend in abundance (slope coefficient from simple linear regression), and temporal autocorrelation.
+All analyses were conducted in R [@RCoreTeam2016].
 <!--Lastly, we also calculated the minimum number of years required (10 years or 3 generations) under IUCN Red List Criterion A2 to classify a species as vulnerable [@IUCN2012].-->
@@
 ## Empirical approach
-
-<!--Our objective is to determine the minimum time series length ($T_{min}$) required to answer a specific question of interest.
-As we describe below, the minimum time series length calculation is particular to each question.
-Intuitively, we might expect a longer minimum time series length ($T_{min}$) required given more complicated questions (citations).-->
 
@@
 \ref{fig:gam_example}).
-We found that although we obtain a similar distribution of minimum times required for GAMs, the minimum time required for GAMs is on average shorter than for linear regression (Fig.
-\ref{fig:min_time_dist_gam}).
 
@@
 \ref{fig:min_time_vs_alpha_beta}).
-We investigated this question through two approaches.
+We estimated $T_{min}$ using two approaches.
 We briefly describe results from the simulation approach and then discuss our empirical approach.
@@
 \ref{fig:shark_example}).
-Similar to the simulation approach decribed above, the minimum time required for the lemon shark population was strongly dependent on model parameters.
+Similar to the simulation approach described above, the minimum time required for the lemon shark population was strongly dependent on model parameters.
 
@@
 ```{r,echo=F,fig.cap='Distributions of the minimum time required for four different biological classes.
-The minimum time required calculation corresponds to a significant level of 0.05 and statistical power of 0.8.\\label{fig:min_time_dist}',fig.width=6}
+The minimum time required calculation corresponds to a significance level of 0.05 and statistical power of 0.8.\\label{fig:min_time_dist}',fig.width=6}
 # Pull in analysis of average min time needed for **empirical species** (save analysis as Rdata file or something)
@@
 
-```{r,echo=F, fig.cap='Minimum time required to estimate change in abundance correlated with (a) trend strength (absolute value of slope coefficient estimated from linear regression), (b) population variance (interannual variability in population size), and (c) temporal lag-1 autocorrelation.\\label{fig:correlates}',fig.height=7, fig.width=4}
+```{r,echo=F, fig.cap='Minimum time required to estimate change in abundance correlated with (a) trend strength (absolute value of slope coefficient estimated from linear regression), (b) population variance (inter-annual variability in population size), and (c) temporal lag-1 autocorrelation.\\label{fig:correlates}',fig.height=7, fig.width=4}
 
@@
 Birds (class Aves) had a much wider distribution of $T_{min}$, but usually required less years of sampling.
-Differences between these classes were explained by differences in interannual variability in population size (Fig.
+Differences between these classes were explained by differences in inter-annual variability in population size (Fig.
 \ref{fig:class}).
@@
 Using a generalized linear model, with a Poisson error structure, all three of these explanatory variables were significant and had large effect sizes (see Table A1).
-Combined, trend strength, population variance, and autocorrelation account for about `r round(100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance),1)`% of the explained deviance, which is similar to $R^2$ [@Zuur2009], in minimum time series length required.
+Combined, trend strength, population variance, and autocorrelation account for about `r round(100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance),1)`% of the explained deviance [@Zuur2009] in minimum time series length required.
 
@@
 
-We explored two different approaches to estimate the minimum time series length required to address a particular question of interest.
-We specifically asked, what is the minimum time series length required to determine long-term population trends using linear regression?
+We explored two approaches to estimate the minimum time series length required to address a particular question of interest.
+We asked, what is the minimum time series length required to determine long-term population trends using linear regression?
 This is one of the simplest questions one could ask of a time series.
-We would expect more involved questions (e.g.
-detecting chaos) to require a larger minimum time series length.
 The simulation-based approach has been suggested by others, especially in situations more complicated than that suited for classic power analysis [@Gerrodette1987;@Johnson2015;@Bolker2008].
@@
 rate of increase or decrease) is weak or when population variability is high [@Gerrodette1987].
+We also showed how the simulation model can be altered for a particular population (Fig.
+\ref{fig:shark_example}) or question (Figs.
+\ref{fig:min_time_dist_log_pop},\ref{fig:gam_example}).
 
 Here, we focus on an empirical approach to estimate the minimum time series length required to assess changes in abundance over time.
-We examined long time series and then subsampled each to determine the minimum time required to achieve a desired level of significance and power for linear regression.
+We examined `r nrow(pop_info)` population time series (all longer than 35 years).
+We then subsampled each to determine the minimum time required to achieve a desired significance level and power for linear regression.
 Statistical power is important as it provides on information as to the necessary samples required to determine a significant trend [@Legg2006].
-We found that for `r nrow(pop_info)` populations, at least 10-15 years of continuous monitoring were necessary (Fig.
+We found that at least 15-20 years of continuous monitoring were typically necessary (Fig.
 \ref{fig:min_time_dist}).
 This time frame is in line with past work on a short-lived snail species [@Rueda-Cediel2015] and a long-lived whale species [@Gerber1999].
+@Hatch2003 used seabird monitoring data to estimate minimum sampling requirements.
+He found that the time required ranged from 11 to 69 years depending on species, trend strength, and study design.
+
 In line with theoretical predictions [@Rhodes2011], we also found $T_{min}$ was strongly correlated with the trend strength, variability in population size, and temporal autocorrelation (Fig.
@@
 We initially hypothesized that species with longer lifespans or generation times may require a longer sampling period.
-Our result could have been a result of at least two effects.
-First, the database we used may not include a diverse enough set of species with different life history traits.
+Our result could have been a result of at least two factors.
+First, the data we used may not include a diverse enough set of species with different life history traits.
 Second, the question we poised, whether a population is increasing or decreasing, was specifically concerned with trends in population density over time.
-Therefore, life-history characteristics may be more important for other questions (e.g.
-species extinction risk) where estimates minimum time required are needed.
+Therefore, life-history characteristics may be more important for other questions, like estimating species extinction risk [@Hutchings2012].
 
-There has been interest in determining the minimum time series length in the bird research community [@Hatch2003; @Bart2004; @Seavy2007].
-For example, @Hatch2003 used seabird monitoring data to estimate the minimum time needed to achieve appropriate levels of power.
-He found that the minimum time required ranged between 11 and 69 years depending on species, trend strength, and study design.
-Similarly, @Bart2004 recommended a standard of 80% power to detect a 50% decline over 20 years in landbird populations.
+<!-- Similarly, @Bart2004 recommended a standard of 80% power to detect a 50% decline over 20 years in landbird populations.
 They found that only about 42% of populations in the Breeding Bird Survey met their criteria.
@@
 Therefore, they cautioned against only using power analyses to design monitoring schemes and instead argued for metrics that would increase precision.
-For example, @Seavy2007 suggest improving randomization, reducing bias, and increases detection probability when designing and evaluating monitoring programs.
+For example, @Seavy2007 suggest improving randomization, reducing bias, and increases detection probability when designing and evaluating monitoring programs.-->
 
@@
 They found that censuses less than 5 years may be misleading and that extrapolating from one population to another, even when nearby geographically, may be untenable.
+
+@Seavy2007 asked whether statistical power was even a useful framework for assessing long-term population trends.
+They used 24 years of census data on Red-tailed Tropicbirds (\emph{Phaethon rubricauda}) in Hawaii and showed that to detect a 50% decline over 10 years almost always resulting in high statistical power (above 0.8).
+Therefore, they cautioned against only using power analyses to design monitoring schemes and instead argued for metrics that would increase precision.
+For example, @Seavy2007 suggest improving randomization, reducing bias, and increases detection probability when designing and evaluating monitoring programs.
+We agree that power analyses should not be the only consideration when designing monitoring schemes.
+However, unlike @Seavy2007, our results indicate that longer than 10 years is often needed to achieve high statistical power.
+Therefore, the design of monitoring programs should include calculations of statistical power, the allocation of sampling in space versus time [@Rhodes2011], and metrics to increase precision.
+Ideally, a formal decision analysis to evaluate these different factors would be conducted to design or assess any monitoring program [@Hauser2006;@McDonald-Madden2010].
 
@@
 First, $T_{min}$ is particular to the specific question of interest.
-For example, in the supplementary material, we used generalized additive models to detect population trends and came up with a similar, but quantitatively different value for $T_{min}$ (Fig.
-\ref{fig:min_time_dist_gam}).
 An additional complication is that for our empirical approach, the subsampling of the full time series allows for estimates of power, but the individual subsamples are clearly not independent of one another.
@@
 
-We use a database of `r nrow(pop_info)` populations to determine the minimum time series length required to determine population trends.
+We use a database of `r nrow(pop_info)` populations to determine the minimum time series length required to detect population trends.
 This goes beyond previous work that either focused on theoretical investigations or a limited number of species.
-We show that to identify long-term changes in abundance, 10-15 years of continuous monitoring are required (Fig.
+We show that to identify long-term changes in abundance, 15-20 years of continuous monitoring are often required (Fig.
 \ref{fig:min_time_dist}).
@@
 \ref{fig:correlates}).
-Our work implies that for many populations, census data less than 10-15 years are probably not reliable.
-This stresses the importance of long-term monitoring programs.
+Our work implies that for many populations, time series less than 15-20 years are probably not reliable for detecting population trends.
+This result highlights the importance of long-term monitoring programs.
 From both a scientific and management perspective estimates of $T_{min}$ are important.
-If a time series is too short we lack statistical power to reliably detect long-term population trends.
+If a time series is too short, we lack statistical power to reliably detect long-term population trends.
 In addition, a time series that is too long may be a poor use of already limited funds [@Gerber1999].
@@
 In the supporting material, we provide an expanded methods sections, additional figures, minimum time calculations for determining exponential growth, simulations with a more complicated population model, minimum time calculations for determining long-term growth rates, and the use of generalized additive models to identify population trends.
-All code and data can be found at www.github.com/erwhite1
+All code and data can be found at https://github.com/erwhite1/time-series-project
 

### 2017-08-19 writing/manuscript_draft.tex
Final version of manuscript for PeerJ preprint

@@
 
-Running head: \vspace{3 mm}
-
 Title: Minimum time required to detect population trends: the need for
@@
 
-Author(s): \textsc{Easton R. White$^{1,2}$} \vspace{3 mm}
+Author: \textsc{Easton R. White$^{1,2}$} \vspace{3 mm}
 
@@
 \(^2\)Corresponding author:
-\href{mailto:eawhite@ucdavis.edu}{\nolinkurl{eawhite@ucdavis.edu}}, 1
-480 203 7931 \vspace{3 mm}
+\href{mailto:eawhite@ucdavis.edu}{\nolinkurl{eawhite@ucdavis.edu}}
+\vspace{3 mm}
 
-Number of words (including 300 words for figures): X,XXX Number of
-figures: Number of figures in supplementary material: \vspace{3 mm}
+Number of words (including 300 words per figure): 5900 \vspace{1 mm}
 
-To be submitted to: \emph{Biological Conservation} \vspace{3 mm}
+Number of figures: 3 \vspace{1 mm}
 
-Available as a preprint at: www.peerj.com \vspace{3 mm}
+Supplementary material: 11 figures and 1 table \vspace{3 mm}
 
-Color does not need to be used for any figures in print \vspace{3 mm}
-
-Keywords: ecological time series, experimental design, monitoring, power
-analysis, statistical power, sampling design
+Data and code for all the figures and tables can be found at
+(\url{https://github.com/erwhite1/time-series-project}).
 
@@
 \begin{abstract}
-Long-term census data are necessary to better understand ecological processes and make management decisions.
-However, census data are often expensive, requiring a lot of time and resources.
-Little work has addressed the question: when is a time series of census data long enough to address a question of interest?
-<!--Here, we explore two approaches to address this question: simulations and an empirical approach.--> We determine the minimum time series length required to estimate significant increases or decreases in population abundance.
-Importantly, we examine the ability to detect a trend with a set level of statistical power, which is often neglected in ecological time series analyses.
-Using a simulation approach, we demonstrate how the minimum time series length required increases with weaker trends in abundance and with higher population variability.
-In addition, we examine 878 populations of vertebrate species to determine the minimum time required to detect changes in their abundance.
-We show that 10-15 years of continuous monitoring are required in order to achieve a high level of statistical power.
-Similar to simulation-approaches, the minimum time required for field census data strongly depends on trend strength, population variability, and temporal autocorrelation.
-<!--Perhaps surprisingly, the minimum time required did not correlate well with biological explanatory variables, like body size or generation length.--> These results point to the importance of sampling populations over long periods of time.
-<!--We stress the need for verifying a study has enough statistical power to accurately determine long-term population trends.--> Time series less than 10-15 years are likely under-powered and potentially misleading.
+Long-term time series are necessary to better understand population dynamics, assess species' conservation status, and make management decisions.
+However, population data are often expensive, requiring a lot of time and resources.
+When is a population time series long enough to address a question of interest?
+We determine the minimum time series length required to detect significant increases or decreases in population abundance.
+To address this question, we use simulation methods and examine 878 populations of vertebrate species.
+Here we show that 15-20 years of continuous monitoring are required in order to achieve a high level of statistical power.
+For both simulations and the time series data, the minimum time required depends on trend strength, population variability, and temporal autocorrelation.
+These results point to the importance of sampling populations over long periods of time.
+We argue that statistical power needs to be considered in monitoring program design and evaluation.
+Time series less than 15-20 years are likely underpowered and potentially misleading.
 \end{abstract}
+
+Keywords: ecological time series, experimental design, monitoring, power
+analysis, statistical power, sampling design
 
@@
 to both understand population dynamics and to assess species extinction
-risk (citations). Even though many time series may now be considered
-``long-term'' (e.g.~continuous plankton recorder Giron-Nava et al.
-(2017)), most are still short. Time series are typically short for a
-variety of reasons. They are often coupled with an experiment, which may
-only last a couple of years. In addition, short funding cycles make it
-difficult to examine populations over longer periods of time (Hughes et
-al. 2017).
+risk. Even though many time series may now be considered ``long-term''
+(e.g.~continuous plankton recorder, Giron-Nava et al. (2017)), most are
+still short. Time series are typically short for a variety of reasons
+(Field et al. 2007). They are often coupled with an experiment, which
+may only last a couple of years. In addition, short funding cycles make
+it difficult to examine populations over longer periods of time (Hughes
+et al. 2017).
 
 How long of a time series is actually necessary? This question has
-important implications for both research and management. Scientists need
-to know the time series length required to address a specific question.
-A short time series may lead to wrong conclusions given large natural
-year-to-year variability (citations). Managers need to know when action
-is needed for a population. Therefore, they must understand when
-population trend over time is actually meaningful. The IUCN Red List
-Categories and Criteria suggest, under Criterion A2, a species qualifies
-as vulnerable if it has experienced a 30\% decline over 10 years, or 3
-generations (IUCN 2012). For both scientific and management questions,
-because sampling is typically expensive, we also do not want to sample
-for longer than is necessary (Gerber, DeMaster, and Kareiva 1999).
+important implications for both research and management (Nichols and
+Williams 2006). Scientists need to know the time series length required
+to address a specific question. A short time series may lead to wrong
+conclusions given large natural year-to-year variability. Managers need
+to know when action is needed for a population. Therefore, they must
+understand when population trend over time is actually meaningful. The
+IUCN Red List Categories and Criteria suggest, under Criterion A2, a
+species qualifies as vulnerable if it has experienced a 30\% decline
+over 10 years, or 3 generations (IUCN 2012). For both scientific and
+management questions, because sampling is typically expensive, we also
+do not want to sample for longer than is necessary. For example, Gerber,
+DeMaster, and Kareiva (1999) investigated the minimum time series
+required to estimate population growth of the endangered, but
+recovering, eastern North Pacific gray whale
+(\emph{Eschrichtius robustus}). They used a long-term census to
+retroactively determine the minimum time series required to assess
+threat status. They found that only 11 years were needed, eight years
+before the delisting decision was made. This highlights the importance
+of estimating the minimum time series required as an earlier decision
+would have saved time and money (Gerber, DeMaster, and Kareiva 1999).
 
@@
 true significant trend---this would be a false positive. The false
-positive rate, or significance level (\(\alpha\)) is often set at 0.05.
-A type II error (\(\beta\)) is failure to detect a true trend, or
-failure to reject a false null hypothesis. Formally, statistical power
-(\(1-\beta\)) is one minus the probability of a type II error
-(\(\beta\)). The effect size is a measure of the difference between two
-groups. Prior to an experiment, one could set appropriate levels of
-power, significance level, and the effect size to estimate the sample
-size required for the experiment. This approach, however, is not
-straight-forward for a time series, or more complicated scenarios (P. C.
-Johnson et al. 2015), as data are clearly non-independent.
+positive rate, or significance level (\(\alpha\)) is often set at 0.05
+(although this is purely historical, Mapstone (1995)). A type II error
+(\(\beta\)) is a failure to detect a true trend, or failure to reject a
+false null hypothesis. Formally, statistical power (\(1-\beta\)) is one
+minus the probability of a type II error (\(\beta\)). The effect size is
+a measure of the difference between two groups. Prior to an experiment,
+one could set appropriate levels of power, significance level, and the
+effect size to estimate the sample size required for the experiment.
+This approach, however, is not straight-forward for a time series, or
+more complicated scenarios (P. C. Johnson et al. 2015), as data are
+clearly non-independent.
 
@@
 increasing or decreasing? It is then possible to determine how power
-changes with a variable of interest. For example, time series are
+changes with a variable of interest. For example, time series can be
 simulated for different lengths of time. From these simulations, the
@@
 criteria, again a measure of statistical power. Similar to the
-simulation approach, we can use these measures of power to determine the
-minimum time series required for a particular question of interest.
+simulation approach, these measures of power can be used to determine
+the minimum time series required for a particular question of interest.
 
 Past work has investigated questions related to the minimum time series
-required to estimate trends in population size over time (Giron-Nava et
-al. 2017). For example, Rhodes and Jonzen (2011) examined the optimal
-allocation of effort between spatial and temporal replicates. Using
-simple populations models, they found that the allocation of effort
-depends on environmental variation, spatial and temporal
-autocorrelation, and observer error. Rueda-Cediel et al. (2015) also
-used a modeling approach, but parameterized a model specifically for a
-threatened snail, \emph{Tasmaphena lamproides}. They found that for this
-short-lived organism, 15 years was adequate to assess long-term trends
-in abundance. In addition, Hatch (2003) examined seabird populations and
-determined the minimum time required to achieve high statistical power
-ranged between 11 and 69 years. However, these studies, and other past
-work, have typically focused either on theoretical aspects of monitoring
-design or focused on only a few species.
+required to estimate trends in population size over time (Wagner,
+Vandergoot, and Tyson 2009; Giron-Nava et al. 2017). For example, Rhodes
+and Jonzen (2011) examined the optimal allocation of effort between
+spatial and temporal replicates. Using simple populations models, they
+found that the allocation of effort depends on environmental variation,
+spatial and temporal autocorrelation, and observer error. Rueda-Cediel
+et al. (2015) also used a modeling approach, but parameterized a model
+specifically for a threatened snail, \emph{Tasmaphena lamproides}. They
+found that for this short-lived organism, 15 years was adequate to
+assess long-term trends in abundance. Hatch (2003) examined seabird
+populations and determined the minimum time required to achieve high
+statistical power ranged between 11 and 69 years. However, these
+studies, and other past work, have typically focused either on
+theoretical aspects of monitoring design or focused on only a few
+species.
 
-We use both simulations and a database of empirical time series to
-determine the minimum number of years required to address several
-questions. We estimate the minimum time series length required
-(\(T_{min}\)) to assess long-term changes in abundance via simple linear
-regression. First, we estimate \(T_{min}\) using a simulation approach.
-Then we examine 878 population time series to estimate \(T_{min}\). In
-the supplementary material, we determine \(T_{min}\) for related ideas:
-calculating long-term growth rates, using more complicated population
-models, varying statistical level and power, and the use of generalized
-additive models.
+We use both simulations and empirical time series to determine the
+minimum number of years required to address several questions. We
+estimate the minimum time series length required (\(T_{min}\)) to assess
+long-term changes in abundance via simple linear regression. First, we
+estimate \(T_{min}\) using a simulation approach. Then we examine 878
+population time series to estimate \(T_{min}\). In the supplementary
+material, we determine \(T_{min}\) for related ideas: calculating
+long-term growth rates, using more complicated population models,
+varying statistical level and power, and the use of generalized additive
+models.
 
@@
 increase). It is important to note that any population model could be
-subsituted for equation \ref{sim_model}, as we do in the supplementary
+substituted for equation \ref{sim_model}, as we do in the supplementary
 material (Figs. \ref{fig:min_time_dist_log_pop},
@@
 Statistical power is then the fraction of simulations that meet some
-criteria. Here, we use whether the slope parameter from linear
-regression is significant. Statistical power changes with the simulation
-length (Fig. \ref{fig:theoretical_approach}b). Then, the minimum time
-required (\(T_{min}\)) is the number of years needed to obtain
-statistical power above some threshold (historically at 0.8, Cohen
-(1992)). Statistical power of 0.8 would indicate that, if there was a
-true trend in abundance, we would have a 0.8 probability of detecting
-the trend. We also tested the effect of varying both the significance
-level and statistical power (Fig. \ref{fig:min_time_vs_alpha_beta}).
+criteria. Here, our criteria is whether the slope parameter from linear
+regression is significant at the 0.05 threshold. Statistical power of
+0.8 would indicate that, if there was a true trend in abundance, we
+would have a 0.8 probability of detecting the trend. We also tested the
+effect of varying both the significance level and statistical power
+(Fig. \ref{fig:min_time_vs_alpha_beta}).
 
@@
 We use a database of 2444 population time series compiled in (Keith et
-al. 2015) to compare the predictability of growth rates amongst
+al. 2015); they compared the predictability of growth rates among
 populations. The data are originally from the Global Population Dynamics
@@
 (slope coefficient from simple linear regression), and temporal
-autocorrelation.
+autocorrelation. All analyses were conducted in R (R Core Team 2016).
 
@@
 generalized additive models, instead of the simple linear models used
-here (Fig. \ref{fig:gam_example}). We found that although we obtain a
-similar distribution of minimum times required for GAMs, the minimum
-time required for GAMs is on average shorter than for linear regression
-(Fig. \ref{fig:min_time_dist_gam}).
+here (Fig. \ref{fig:gam_example}).
 
@@
 increased our level of power or decreased \(\alpha\), \(T_{min}\)
-increased (Fig. \ref{fig:min_time_vs_alpha_beta}). We investigated this
-question through two approaches. We briefly describe results from the
+increased (Fig. \ref{fig:min_time_vs_alpha_beta}). We estimated
+\(T_{min}\) using two approaches. We briefly describe results from the
 simulation approach and then discuss our empirical approach.
@@
 in this particular scenario (Fig. \ref{fig:shark_example}). Similar to
-the simulation approach decribed above, the minimum time required for
+the simulation approach described above, the minimum time required for
 the lemon shark population was strongly dependent on model parameters.
@@
 biological classes. The minimum time required calculation corresponds to
-a significant level of 0.05 and statistical power of
+a significance level of 0.05 and statistical power of
 0.8.\label{fig:min_time_dist}}
@@
 correlated with (a) trend strength (absolute value of slope coefficient
-estimated from linear regression), (b) population variance (interannual
+estimated from linear regression), (b) population variance (inter-annual
 variability in population size), and (c) temporal lag-1
@@
 years of sampling. Differences between these classes were explained by
-differences in interannual variability in population size (Fig.
+differences in inter-annual variability in population size (Fig.
 \ref{fig:class}). We also examined a subset of populations where
@@
 Combined, trend strength, population variance, and autocorrelation
-account for about 72.6\% of the explained deviance, which is similar to
-\(R^2\) (Zuur et al. 2009), in minimum time series length required.
+account for about 72.6\% of the explained deviance (Zuur et al. 2009) in
+minimum time series length required.
 
@@
 
-We explored two different approaches to estimate the minimum time series
-length required to address a particular question of interest. We
-specifically asked, what is the minimum time series length required to
-determine long-term population trends using linear regression? This is
-one of the simplest questions one could ask of a time series. We would
-expect more involved questions (e.g.~detecting chaos) to require a
-larger minimum time series length. The simulation-based approach has
-been suggested by others, especially in situations more complicated than
-that suited for classic power analysis (Gerrodette 1987; P. C. Johnson
-et al. 2015; Bolker 2008). Our simulations support past work that longer
-time series are needed when the trend strength (i.e.~rate of increase or
-decrease) is weak or when population variability is high (Gerrodette
-1987).
+We explored two approaches to estimate the minimum time series length
+required to address a particular question of interest. We asked, what is
+the minimum time series length required to determine long-term
+population trends using linear regression? This is one of the simplest
+questions one could ask of a time series. The simulation-based approach
+has been suggested by others, especially in situations more complicated
+than that suited for classic power analysis (Gerrodette 1987; P. C.
+Johnson et al. 2015; Bolker 2008). Our simulations support past work
+that longer time series are needed when the trend strength (i.e.~rate of
+increase or decrease) is weak or when population variability is high
+(Gerrodette 1987). We also showed how the simulation model can be
+altered for a particular population (Fig. \ref{fig:shark_example}) or
+question (Figs. \ref{fig:min_time_dist_log_pop},\ref{fig:gam_example}).
 
@@
 series length required to assess changes in abundance over time. We
-examined long time series and then subsampled each to determine the
-minimum time required to achieve a desired level of significance and
-power for linear regression. Statistical power is important as it
-provides on information as to the necessary samples required to
-determine a significant trend (Legg and Nagy 2006). We found that for
-878 populations, at least 10-15 years of continuous monitoring were
-necessary (Fig. \ref{fig:min_time_dist}). This time frame is in line
-with past work on a short-lived snail species (Rueda-Cediel et al. 2015)
-and a long-lived whale species (Gerber, DeMaster, and Kareiva 1999). In
-line with theoretical predictions (Rhodes and Jonzen 2011), we also
+examined 878 population time series (all longer than 35 years). We then
+subsampled each to determine the minimum time required to achieve a
+desired significance level and power for linear regression. Statistical
+power is important as it provides on information as to the necessary
+samples required to determine a significant trend (Legg and Nagy 2006).
+We found that at least 15-20 years of continuous monitoring were
+typically necessary (Fig. \ref{fig:min_time_dist}). This time frame is
+in line with past work on a short-lived snail species (Rueda-Cediel et
+al. 2015) and a long-lived whale species (Gerber, DeMaster, and Kareiva
+1999). Hatch (2003) used seabird monitoring data to estimate minimum
+sampling requirements. He found that the time required ranged from 11 to
+69 years depending on species, trend strength, and study design.
+
+In line with theoretical predictions (Rhodes and Jonzen 2011), we also
 found \(T_{min}\) was strongly correlated with the trend strength,
@@
 require a longer sampling period. Our result could have been a result of
-at least two effects. First, the database we used may not include a
-diverse enough set of species with different life history traits.
-Second, the question we poised, whether a population is increasing or
-decreasing, was specifically concerned with trends in population density
-over time. Therefore, life-history characteristics may be more important
-for other questions (e.g.~species extinction risk) where estimates
-minimum time required are needed.
-
-There has been interest in determining the minimum time series length in
-the bird research community (Hatch 2003; Bart et al. 2004; Seavy and
-Reynolds 2007). For example, Hatch (2003) used seabird monitoring data
-to estimate the minimum time needed to achieve appropriate levels of
-power. He found that the minimum time required ranged between 11 and 69
-years depending on species, trend strength, and study design. Similarly,
-Bart et al. (2004) recommended a standard of 80\% power to detect a 50\%
-decline over 20 years in landbird populations. They found that only
-about 42\% of populations in the Breeding Bird Survey met their
-criteria. Seavy and Reynolds (2007) asked whether statistical power was
-even a useful framework for assessing long-term population trends. They
-used 24 years of census data on Red-tailed Tropicbirds
-(\emph{Phaethon rubricauda}) in Hawaii and showed that to detect a 50\%
-decline over 10 years almost always resulting in high statistical power
-(above 0.8). Therefore, they cautioned against only using power analyses
-to design monitoring schemes and instead argued for metrics that would
-increase precision. For example, Seavy and Reynolds (2007) suggest
-improving randomization, reducing bias, and increases detection
-probability when designing and evaluating monitoring programs.
+at least two factors. First, the data we used may not include a diverse
+enough set of species with different life history traits. Second, the
+question we poised, whether a population is increasing or decreasing,
+was specifically concerned with trends in population density over time.
+Therefore, life-history characteristics may be more important for other
+questions, like estimating species extinction risk (J. A. Hutchings et
+al. 2012).
 
@@
 
+Seavy and Reynolds (2007) asked whether statistical power was even a
+useful framework for assessing long-term population trends. They used 24
+years of census data on Red-tailed Tropicbirds
+(\emph{Phaethon rubricauda}) in Hawaii and showed that to detect a 50\%
+decline over 10 years almost always resulting in high statistical power
+(above 0.8). Therefore, they cautioned against only using power analyses
+to design monitoring schemes and instead argued for metrics that would
+increase precision. For example, Seavy and Reynolds (2007) suggest
+improving randomization, reducing bias, and increases detection
+probability when designing and evaluating monitoring programs. We agree
+that power analyses should not be the only consideration when designing
+monitoring schemes. However, unlike Seavy and Reynolds (2007), our
+results indicate that longer than 10 years is often needed to achieve
+high statistical power. Therefore, the design of monitoring programs
+should include calculations of statistical power, the allocation of
+sampling in space versus time (Rhodes and Jonzen 2011), and metrics to
+increase precision. Ideally, a formal decision analysis to evaluate
+these different factors would be conducted to design or assess any
+monitoring program (Hauser, Pople, and Possingham 2006; McDonald-Madden
+et al. 2010).
+
 \subsection{Limitations}\label{limitations}
@@
 length required. First, \(T_{min}\) is particular to the specific
-question of interest. For example, in the supplementary material, we
-used generalized additive models to detect population trends and came up
-with a similar, but quantitatively different value for \(T_{min}\) (Fig.
-\ref{fig:min_time_dist_gam}). An additional complication is that for our
+question of interest. An additional complication is that for our
 empirical approach, the subsampling of the full time series allows for
@@
 We use a database of 878 populations to determine the minimum time
-series length required to determine population trends. This goes beyond
+series length required to detect population trends. This goes beyond
 previous work that either focused on theoretical investigations or a
 limited number of species. We show that to identify long-term changes in
-abundance, 10-15 years of continuous monitoring are required (Fig.
+abundance, 15-20 years of continuous monitoring are often required (Fig.
 \ref{fig:min_time_dist}). In line with theoretical predictions
@@
 in population size, and the temporal autocorrelation (Fig.
-\ref{fig:correlates}). Our work implies that for many populations,
-census data less than 10-15 years are probably not reliable. This
-stresses the importance of long-term monitoring programs. From both a
-scientific and management perspective estimates of \(T_{min}\) are
-important. If a time series is too short we lack statistical power to
-reliably detect long-term population trends. In addition, a time series
-that is too long may be a poor use of already limited funds (Gerber,
-DeMaster, and Kareiva 1999). Future work should examine other species,
-with a wider range of life history characteristics. In addition, similar
-approaches can be used to determine the minimum time series length
-required to address additional questions of interest.
+\ref{fig:correlates}). Our work implies that for many populations, time
+series less than 15-20 years are probably not reliable for detecting
+population trends. This result highlights the importance of long-term
+monitoring programs. From both a scientific and management perspective
+estimates of \(T_{min}\) are important. If a time series is too short,
+we lack statistical power to reliably detect long-term population
+trends. In addition, a time series that is too long may be a poor use of
+already limited funds (Gerber, DeMaster, and Kareiva 1999). Future work
+should examine other species, with a wider range of life history
+characteristics. In addition, similar approaches can be used to
+determine the minimum time series length required to address additional
+questions of interest.
 
@@
 and the use of generalized additive models to identify population
-trends. All code and data can be found at www.github.com/erwhite1
+trends. All code and data can be found at
+\url{https://github.com/erwhite1/time-series-project}
 
@@
 \hypertarget{refs}{}
-\hypertarget{ref-Bart2004}{}
-Bart, Jonathon, Kenneth P. Burnham, Erica H. Dunn, Charles M. Francis,
-and C. John Ralph. 2004. ``Goals and strategies for estimating trends in
-landbird abundance.'' \emph{Journal of Wildlife Management} 68 (3):
-611--26.
-
 \hypertarget{ref-Bolker2008}{}
@@
 doi:\href{https://doi.org/10.1037/0033-2909.112.1.155}{10.1037/0033-2909.112.1.155}.
+
+\hypertarget{ref-Field2007}{}
+Field, Scott A, Patrick J O Connor, Andrew J Tyre, and Hugh P
+Possingham. 2007. ``Making monitoring meaningful.'' \emph{Austral
+Ecology} 32: 485--91.
+doi:\href{https://doi.org/10.1111/j.1442-9993.2007.01715.x}{10.1111/j.1442-9993.2007.01715.x}.
 
@@
 
+\hypertarget{ref-Hauser2006}{}
+Hauser, Cindy E., Anthony R. Pople, and Hugh P. Possingham. 2006.
+``Should managed populations be monitored every year?'' \emph{Ecological
+Applications} 16 (2): 807--19.
+
 \hypertarget{ref-Hughes2017}{}
@@
 
+\hypertarget{ref-Hutchings2012}{}
+Hutchings, Jeffrey A, Ransom A Myers, Verónica B García, Luis O
+Lucifora, and Anna Kuparinen. 2012. ``Life-history correlates of
+extinction risk and recovery potential.'' \emph{Ecological Applications}
+22 (4): 1061--7.
+
 \hypertarget{ref-IUCN2012}{}
@@
 \hypertarget{ref-Johnson2015}{}
-Johnson, Paul C, Sarah JE Barry, Heather M Ferguson, and Pie Müller.
+Johnson, Paul CD, Sarah JE Barry, Heather M Ferguson, and Pie Müller.
 2015. ``Power analysis for generalized linear mixed models in ecology
@@
 
+\hypertarget{ref-Mapstone1995}{}
+Mapstone, Bruce D. 1995. ``Scalable decision rules for environmental
+impact studies : effect Size , type I , and type II errors.''
+\emph{Ecological Applications} 5 (2): 401--10.
+
+\hypertarget{ref-McDonald-Madden2010}{}
+McDonald-Madden, Eve, Peter W J Baxter, Richard A. Fuller, Tara G.
+Martin, Edward T. Game, Jensen Montambault, and Hugh P. Possingham.
+2010. ``Monitoring does not always count.'' \emph{Trends in Ecology and
+Evolution} 25 (10): 547--50.
+doi:\href{https://doi.org/10.1016/j.tree.2010.07.002}{10.1016/j.tree.2010.07.002}.
+
 \hypertarget{ref-Morrison2008}{}
@@
 Population Dynamics Database Version 2.''
+
+\hypertarget{ref-Nichols2006}{}
+Nichols, James D., and Bryon K. Williams. 2006. ``Monitoring for
+conservation.'' \emph{Trends in Ecology and Evolution} 21 (12): 668--73.
+doi:\href{https://doi.org/10.1016/j.tree.2006.08.007}{10.1016/j.tree.2006.08.007}.
+
+\hypertarget{ref-RCoreTeam2016}{}
+R Core Team. 2016. ``R: A language and environment for statistical
+computing.'' Vienna, Austria: R Foundation for Statistical Computing.
+\url{https://www.r-project.org/}.
 
@@
 doi:\href{https://doi.org/10.1016/j.biocon.2007.08.007}{10.1016/j.biocon.2007.08.007}.
+
+\hypertarget{ref-Wagner2009}{}
+Wagner, Tyler, Christopher S. Vandergoot, and Jeff Tyson. 2009.
+``Evaluating the power to detect temporal trends in fishery-independent
+surveys - A case study based on gill nets set in the Ohio waters of Lake
+Erie for walleyes.'' \emph{North American Journal of Fisheries
+Management} 29: 805--16.
+doi:\href{https://doi.org/10.1577/M08-197.1}{10.1577/M08-197.1}.
 

### 2017-08-19 writing/supp_mat_draft.Rmd
Final version of manuscript for PeerJ preprint

@@
 \begin{center}
- \textbf{Supplementary Material: Article title here}
+ \textbf{Supplementary Material: Minimum time required to detect population trends: the need for long-term monitoring programs}
 
@@
 
-Data and code for all the figures and tables can be found at (https://github.com/erwhite1).
+Data and code for all the figures and tables can be found at (https://github.com/erwhite1/time-series-project). All analyses were run using R [@RCoreTeam2016].
 
@@
 As an example, we examine a 35-year time series of Bigeye tuna (\emph{Thunnus obesus}), one species in the Global Population Dynamics Database [@GPDD2010].
-<!--Simple linear regression indicates a significant increase for this population with an estimated intercept of `r as.character(round(as.numeric(coef(model_example)[1])))` and a slope coefficient of `r as.numeric(coef(model_example)[2])`.--> We assume that this significant increase over 35 years is in fact the ``true trend".
-In statistical jargon, the 35-year trend is an effect that is actually present, we can reject the null hypothesis of no trend.
+Simple linear regression indicates a significant decrease for this population with an estimated slope coefficient of `r as.numeric(coef(model_example)[2])`.
+We assume that this significant increase over 35 years is in fact the ``true trend".
+In statistical jargon, the 35-year trend is an effect that is actually present; we can reject the null hypothesis of no trend.
 We can then use this as a benchmark to see if subsamples of the time series also indicate a significant increase.
-
-We first extract all contiguous subsamples of the time series.
-This leads to 34 two-year subsamples, 33 two-year subsamples, and so forth until a single 35-year subsample.
-We can call each set of subsamples, of the same length, a set.
-For each subsample, we conduct linear regression and extract model coefficients and p-values.
-Then, the fraction of subsamples within a set that show significant trends (significant slope coefficient) is the statistical power.
-It is important to note that we only consider subsamples to be significant if they are significant in the same direction as the complete 35-year time series.
-In other words, we are conducting a one-tailed test.
-
-We can then plot statistical power as a function of time series length (Fig.
-\ref{fig:empirical_approach_example}b).
-As expected, we can see that power increases with the more years that are sampled.
-Then, we determine an appropriate level of statistical power that we find acceptable.
-Traditionally, this has been at 0.8, however, this is purely historical.
-Statistical power of 0.8 implies if a true trend is present, or a real change in abundance, then we will detect this trend 0.8 fraction of the time.
 
@@
 
-With statistical power of 0.8, we then determine the minimum time series length ($T_{min}$) required to achieve that level of statistical power.
+We then perform a subsampling routine to estimate the minimum time required $T_{min}$ (similar to @Gerber1999). This is the same routine we used for the results in the main manuscript.
+\begin{enumerate}
+  \item We first extract all contiguous subsamples of the time series. This leads to 34 two-year subsamples, 33 two-year subsamples, and so forth until a single 35-year subsample.  
+  \item For each subsample, we conduct linear regression and extract model coefficients and p-values.
+\item We can call each set of subsamples, of the same length, a set.
+The fraction of subsamples within a set that show significant trends (significant slope coefficient $\alpha$ less than 0.05) is the statistical power.
+It is important to note that we only consider subsamples to be significant if they are significant in the same direction as the complete 35-year time series.
+In other words, we are conducting a one-tailed test.
+  \item We can then plot statistical power as a function of time series length (Fig.
+\ref{fig:empirical_approach_example}b).
+As expected, we can see that power increases with the more years that are sampled.
+  \item  Then, we determine an appropriate level of statistical power that we find acceptable.
+Traditionally, this has been at 0.8, however, this is purely historical.
+Statistical power of 0.8 implies if a true trend is present, or there is a real change in abundance, then we will detect this trend 0.8 fraction of the time.
+  \item We then determine the minimum time series length ($T_{min}$) required to achieve that level of statistical power.
 Here, $T_{min}$ is the first point in Fig.
@@
 In this example, $T_{min}$ is `r head(which(power_example>0.8),1)`.
-Therefore, a minimum of 22 years of continuous monitoring are required to have 0.8 statistical power in determining long-term changes in abundance.
+\end{enumerate}
+
+Therefore, a minimum of `r head(which(power_example>0.8),1)` years of continuous monitoring are required (for 0.8 statistical power at 0.05 significance level) to determine long-term changes in abundance.
 
@@
 
-## Statistical model of minimum time required and time series characteristics
-
-In the main text, we explained how the minimum time required strongly correlated with the trend strength, temporal autocorrelation, and variance in population size.
-Here, we use a generalized linear model framework with a Poisson error structure to determine drivers of the minimum time required.
-In figure \ref{fig:poisson_model}, we show a set of residual plots for the regression.
-We then show the coefficient estimates and levels of significance in table \ref{table:model_output}.
+## Predictors of minimum time required
 
@@
 ```
+
+In the main text, we explained how the minimum time required strongly correlated with the trend strength, temporal autocorrelation, and variance in population size.
+Here, we use a generalized linear model framework with a Poisson error structure to determine explanatory variables of the minimum time required.
+We use the same `r nrow(pop_info)` populations as in the main text.
+In figure \ref{fig:poisson_model}, we show a set of residual plots for the regression.
+We then show the coefficient estimates and levels of significance in table \ref{table:model_output}.
 
@@
 In the main manuscript, we examined characteristics of the time series itself, like variability, autocorrelation, and the trend in abundance over time.
-Here, we combined our time series database with a database on life history characteristics of amniotes from @Myhrvold2015.
-There was life history information available for `r nrow(LPI_pop_info)` populations representing `r length(table(LPI_pop_info$Binomial))` different species, all of which were in the Aves class.
-
-We then correlated minimum time required for each population with its corresponding life history characteristics.
-In figure \ref{fig:biological_correlates} we examined minimum time required versus generation length (years), litter size (n), adult body mass (grams), maximum longevity (years), egg mass (grams), and incubation (days).
-None of these variables had much explanatory power in accounting for the variance in the minimum time required.
-see table \ref{table:model_output}}
+Here, we combined our time series data with a data on life history characteristics of amniotes from @Myhrvold2015.
+There was life history information available for `r nrow(LPI_pop_info)` populations representing `r length(table(LPI_pop_info$Binomial))` different species, all of which were birds (Aves class).
 
@@
 
+#######
+
+poisson_model <-glm(min_time_for_power~ gen_len + litter_or_clutch_size_n + adult_body_mass_g +maximum_longevity_y + egg_mass_g + incubation_d,family = 'poisson',data=LPI_pop_info)
+
+#  par(mfrow=c(2,2),mar=c(5,5,1,2),oma=c(0,0,0.5,0))
+#    plot(poisson_model,cex.lab=1.2,las=1,cex.axis=1.2,main=' ',pch=16, col=rgb(0.5,0.5,0.5,0.4))
+#R^2 calculation (pg. 218 Zuur et al. 2009 mixed effects book)
+R2_fullmodel <- 100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance)
 ```
+
+We then correlated minimum time required for each population with its corresponding life history characteristics.
+In figure \ref{fig:biological_correlates} we examined minimum time required versus generation length (years), litter size (n), adult body mass (grams), maximum longevity (years), egg mass (grams), and incubation (days).
+None of these variables had much explanatory power in accounting for the variance in the minimum time required.
+We ran a generalized linear model (with a Poisson error structure) and found that these six variables only accounted for `r round(100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance),2)`% of the explained deviance in minimum time required.
 
@@
 
-When we examined differences in the minimum time required there were similar patterns with biological classes.
+Minimum time required differed for different biological classes.
 For instance, more time is required for species within the Actinopterygii class compared to other species (Fig.
 \ref{fig:class}a).
-These differences between biological classes can be explained by differences in population variability, with species in Actinopterygii have larger variability in population size from year-to-year.
+These differences between biological classes can be explained by differences in population variability, with species in the Actinopterygii class experiencing larger inter-annual variability in population size (Fig.
+\ref{fig:class}c).
 
@@
 # Biological correlates plots
-par(mfrow=c(2,2),oma=c(8,1,1.5,0.5),mar=c(0,4,0,1))
+par(mfrow=c(2,2),oma=c(8,1,1.5,0.5),mar=c(0.5,4,0,1))
 
@@
 
-Estimates of $T_{min}$ depend strongly on the values used for the significance level ($\alpha$) and the probability of type II error ($\beta$), both of which are set by the practitioner.
-Here we explore how estimates of $T_{min}$ are affected by changes in each of these parameters.
+Estimates of $T_{min}$ depend strongly on the values used for the statistical significance level ($\alpha$) and the probability of type II error ($\beta$), both of which are set by the practitioner.
+Again, statistical power is $1-\beta$.
+Here we used simulations of the model described in the main manuscript.
+The model simulates linear trends in population abundance.
+We explored how estimates of $T_{min}$ are affected by changes in each of these parameters.
 We see that the minimum time required increases with increases in statistical power or decreases with increases in the threshold for statistical significance (Fig \ref{fig:min_time_vs_alpha_beta}).
@@
 
-In the main text we evaluated the minimum time required to determine long-term trends in abundance via linear regression.
+In the main text, we evaluated the minimum time required to determine long-term trends in abundance via linear regression.
 This process examined linear trends in abundance over time.
@@
 
-We see that the distribution of $T_{min}$ is almost identical to that in the main manuscript.
+We see that the distribution of $T_{min}$ is almost identical to that in the main manuscript (Fig.
+\ref{fig:min_time_dist_log_pop}).
 This is perhaps not surprising as most time series that would significantly increase or decrease linearly would probably also significantly increase or decrease at an exponential rate.
-Further, the calculations here and in the main manuscript both use linear regression, and both estimate the same number of parameters.
+Further, the calculations here and in the main manuscript both use linear regression.
+Therefore both calculations estimate the same number of parameters.
 
@@
 
-In the main text, we showed how a simple population model could be simulated repeatedly to estimate the power obtained with time series of increasing length.
+In the main text, we showed how a simple population model could be simulated repeatedly to estimate the power obtained with time series of increasing length [@Bolker2008;@Johnson2015].
 The model in the main text simulated linear population growth with only a slope coefficient, y-intercept, and noise parameter required.
@@
 Here, we use the same routine as the main text, but simulate from a more biologically-realistic population model.
-We simulate the model described in @White2014.
-The model is a stochastic, age-structured population model that includes density-dependence for lemon sharks in Bimini, Bahamas.
+We use the model described in @White2014.
+The model is a stochastic, age-structured population model that includes density-dependence for lemon sharks (\emph{Negaprion brevirostris}) in Bimini, Bahamas.
 
@@
 \ref{fig:shark_example}).
-In this example, the minimum time required ($T_{min}$) to obtain at least 0.8 statistical power, given a significance level of 0.05, is `r years_to_sample[tail(which(shark_power<0.8),1)+1]` years.
+In this example, the minimum time required ($T_{min}$) t0 obtain at least 0.8 statistical power, given a significance level of 0.05, is `r years_to_sample[tail(which(shark_power<0.8),1)+1]` years.
 
@@
 
-We applied the same calculations to `r nrow(pop_info)` populations, as we did in the main text.
+We applied the same calculations to `r nrow(pop_info)` population time series.
 We then obtain a distribution of the minimum time required to measure the ``true" long-term growth rate (Fig.
@@
 This approach allowed us to identify increases or decreases, but a linear model may not always be a good fit.
-Generalized additive models (GAMs) are more general than general linear models and allow more flexible.
+Generalized additive models (GAMs) are more general than general linear models and allow more flexibility [@Wood2006].
 GAMs are models where a response variable depends on unknown smooth functions of explanatory variables.
 GAMs, therefore, can identify relationships between response variables and explanatory variables that are non-linear and perhaps more complicated.
-The downside of GAMs is they typically require more data and are also prone to overfitting (citations).
+The downside of GAMs is they typically require more data and are also prone to overfitting.
 
 Here, we conduct the same analyses in the main text, but instead calculate the minimum time series required to detect trends over time according to a GAM model.
-We hypothesize that GAMs should require less time to detect a trend as they are more flexible than linear regression.
+We hypothesized that GAMs should require less time to detect a trend as they are more flexible than linear regression.
 We provide an example that shows statistical power increases with more time sampled (Fig.

### 2017-08-19 writing/supp_mat_draft.tex
Final version of manuscript for PeerJ preprint

@@
 Data and code for all the figures and tables can be found at
-(\url{https://github.com/erwhite1}).
+(\url{https://github.com/erwhite1/time-series-project}). All analyses
+were run using R (R Core Team 2016).
 
@@
 Population Dynamics Database (NERC Centre for Population Biology 2010).
-We assume that this significant increase over 35 years is in fact the
-``true trend``. In statistical jargon, the 35-year trend is an effect
-that is actually present, we can reject the null hypothesis of no trend.
-We can then use this as a benchmark to see if subsamples of the time
-series also indicate a significant increase.
-
-We first extract all contiguous subsamples of the time series. This
-leads to 34 two-year subsamples, 33 two-year subsamples, and so forth
-until a single 35-year subsample. We can call each set of subsamples, of
-the same length, a set. For each subsample, we conduct linear regression
-and extract model coefficients and p-values. Then, the fraction of
-subsamples within a set that show significant trends (significant slope
-coefficient) is the statistical power. It is important to note that we
-only consider subsamples to be significant if they are significant in
-the same direction as the complete 35-year time series. In other words,
-we are conducting a one-tailed test.
-
-We can then plot statistical power as a function of time series length
-(Fig. \ref{fig:empirical_approach_example}b). As expected, we can see
-that power increases with the more years that are sampled. Then, we
-determine an appropriate level of statistical power that we find
-acceptable. Traditionally, this has been at 0.8, however, this is purely
-historical. Statistical power of 0.8 implies if a true trend is present,
-or a real change in abundance, then we will detect this trend 0.8
-fraction of the time.
+Simple linear regression indicates a significant decrease for this
+population with an estimated slope coefficient of -0.0189581. We assume
+that this significant increase over 35 years is in fact the ``true
+trend``. In statistical jargon, the 35-year trend is an effect that is
+actually present; we can reject the null hypothesis of no trend. We can
+then use this as a benchmark to see if subsamples of the time series
+also indicate a significant increase.
 
@@
 
-With statistical power of 0.8, we then determine the minimum time series
-length (\(T_{min}\)) required to achieve that level of statistical
-power. Here, \(T_{min}\) is the first point in Fig.
-\ref{fig:empirical_approach_example}b where following points are also
-above 0.8. In this example, \(T_{min}\) is 22. Therefore, a minimum of
-22 years of continuous monitoring are required to have 0.8 statistical
-power in determining long-term changes in abundance.
+We then perform a subsampling routine to estimate the minimum time
+required \(T_{min}\) (similar to Gerber, DeMaster, and Kareiva (1999)).
+This is the same routine we used for the results in the main manuscript.
+
+\begin{enumerate}
+  \item We first extract all contiguous subsamples of the time series. This leads to 34 two-year subsamples, 33 two-year subsamples, and so forth until a single 35-year subsample.  
+  \item For each subsample, we conduct linear regression and extract model coefficients and p-values.
+\item We can call each set of subsamples, of the same length, a set.
+The fraction of subsamples within a set that show significant trends (significant slope coefficient $\alpha$ less than 0.05) is the statistical power.
+It is important to note that we only consider subsamples to be significant if they are significant in the same direction as the complete 35-year time series.
+In other words, we are conducting a one-tailed test.
+  \item We can then plot statistical power as a function of time series length (Fig.
+\ref{fig:empirical_approach_example}b).
+As expected, we can see that power increases with the more years that are sampled.
+  \item  Then, we determine an appropriate level of statistical power that we find acceptable.
+Traditionally, this has been at 0.8, however, this is purely historical.
+Statistical power of 0.8 implies if a true trend is present, or there is a real change in abundance, then we will detect this trend 0.8 fraction of the time.
+  \item We then determine the minimum time series length ($T_{min}$) required to achieve that level of statistical power.
+Here, $T_{min}$ is the first point in Fig.
+\ref{fig:empirical_approach_example}b where following points are also above 0.8.
+In this example, $T_{min}$ is 22.
+\end{enumerate}
+
+Therefore, a minimum of 22 years of continuous monitoring are required
+(for 0.8 statistical power at 0.05 significance level) to determine
+long-term changes in abundance.
 
@@
 
-\subsection{Statistical model of minimum time required and time series
-characteristics}\label{statistical-model-of-minimum-time-required-and-time-series-characteristics}
-
-In the main text, we explained how the minimum time required strongly
-correlated with the trend strength, temporal autocorrelation, and
-variance in population size. Here, we use a generalized linear model
-framework with a Poisson error structure to determine drivers of the
-minimum time required. In figure \ref{fig:poisson_model}, we show a set
-of residual plots for the regression. We then show the coefficient
-estimates and levels of significance in table \ref{table:model_output}.
+\subsection{Predictors of minimum time
+required}\label{predictors-of-minimum-time-required}
 
@@
 \end{figure}
+
+In the main text, we explained how the minimum time required strongly
+correlated with the trend strength, temporal autocorrelation, and
+variance in population size. Here, we use a generalized linear model
+framework with a Poisson error structure to determine explanatory
+variables of the minimum time required. We use the same 878 populations
+as in the main text. In figure \ref{fig:poisson_model}, we show a set of
+residual plots for the regression. We then show the coefficient
+estimates and levels of significance in table \ref{table:model_output}.
 
@@
 like variability, autocorrelation, and the trend in abundance over time.
-Here, we combined our time series database with a database on life
-history characteristics of amniotes from Myhrvold et al. (2015). There
-was life history information available for 547 populations representing
-315 different species, all of which were in the Aves class.
-
-We then correlated minimum time required for each population with its
-corresponding life history characteristics. In figure
-\ref{fig:biological_correlates} we examined minimum time required versus
-generation length (years), litter size (n), adult body mass (grams),
-maximum longevity (years), egg mass (grams), and incubation (days). None
-of these variables had much explanatory power in accounting for the
-variance in the minimum time required. see table
-\ref{table:model_output}\}
+Here, we combined our time series data with a data on life history
+characteristics of amniotes from Myhrvold et al. (2015). There was life
+history information available for 547 populations representing 315
+different species, all of which were birds (Aves class).
 
@@
 
+We then correlated minimum time required for each population with its
+corresponding life history characteristics. In figure
+\ref{fig:biological_correlates} we examined minimum time required versus
+generation length (years), litter size (n), adult body mass (grams),
+maximum longevity (years), egg mass (grams), and incubation (days). None
+of these variables had much explanatory power in accounting for the
+variance in the minimum time required. We ran a generalized linear model
+(with a Poisson error structure) and found that these six variables only
+accounted for 6.87\% of the explained deviance in minimum time required.
+
 \pagebreak
@@
 
-When we examined differences in the minimum time required there were
-similar patterns with biological classes. For instance, more time is
-required for species within the Actinopterygii class compared to other
-species (Fig. \ref{fig:class}a). These differences between biological
-classes can be explained by differences in population variability, with
-species in Actinopterygii have larger variability in population size
-from year-to-year.
+Minimum time required differed for different biological classes. For
+instance, more time is required for species within the Actinopterygii
+class compared to other species (Fig. \ref{fig:class}a). These
+differences between biological classes can be explained by differences
+in population variability, with species in the Actinopterygii class
+experiencing larger inter-annual variability in population size (Fig.
+\ref{fig:class}c).
 
@@
 Estimates of \(T_{min}\) depend strongly on the values used for the
-significance level (\(\alpha\)) and the probability of type II error
-(\(\beta\)), both of which are set by the practitioner. Here we explore
-how estimates of \(T_{min}\) are affected by changes in each of these
-parameters. We see that the minimum time required increases with
-increases in statistical power or decreases with increases in the
-threshold for statistical significance (Fig
+statistical significance level (\(\alpha\)) and the probability of type
+II error (\(\beta\)), both of which are set by the practitioner. Again,
+statistical power is \(1-\beta\). Here we used simulations of the model
+described in the main manuscript. The model simulates linear trends in
+population abundance. We explored how estimates of \(T_{min}\) are
+affected by changes in each of these parameters. We see that the minimum
+time required increases with increases in statistical power or decreases
+with increases in the threshold for statistical significance (Fig
 \ref{fig:min_time_vs_alpha_beta}).
@@
 
-In the main text we evaluated the minimum time required to determine
+In the main text, we evaluated the minimum time required to determine
 long-term trends in abundance via linear regression. This process
@@
 We see that the distribution of \(T_{min}\) is almost identical to that
-in the main manuscript. This is perhaps not surprising as most time
-series that would significantly increase or decrease linearly would
-probably also significantly increase or decrease at an exponential rate.
-Further, the calculations here and in the main manuscript both use
-linear regression, and both estimate the same number of parameters.
+in the main manuscript (Fig. \ref{fig:min_time_dist_log_pop}). This is
+perhaps not surprising as most time series that would significantly
+increase or decrease linearly would probably also significantly increase
+or decrease at an exponential rate. Further, the calculations here and
+in the main manuscript both use linear regression. Therefore both
+calculations estimate the same number of parameters.
 
@@
 simulated repeatedly to estimate the power obtained with time series of
-increasing length. The model in the main text simulated linear
-population growth with only a slope coefficient, y-intercept, and noise
-parameter required. This model is purely phenomenological and does not
-include any species life history. Here, we use the same routine as the
-main text, but simulate from a more biologically-realistic population
-model. We simulate the model described in White, Nagy, and Gruber
-(2014). The model is a stochastic, age-structured population model that
-includes density-dependence for lemon sharks in Bimini, Bahamas.
+increasing length (Bolker 2008; Johnson et al. 2015). The model in the
+main text simulated linear population growth with only a slope
+coefficient, y-intercept, and noise parameter required. This model is
+purely phenomenological and does not include any species life history.
+Here, we use the same routine as the main text, but simulate from a more
+biologically-realistic population model. We use the model described in
+White, Nagy, and Gruber (2014). The model is a stochastic,
+age-structured population model that includes density-dependence for
+lemon sharks (\emph{Negaprion brevirostris}) in Bimini, Bahamas.
 
@@
 longer sampling time (Fig. \ref{fig:shark_example}). In this example,
-the minimum time required (\(T_{min}\)) to obtain at least 0.8
+the minimum time required (\(T_{min}\)) t0 obtain at least 0.8
 statistical power, given a significance level of 0.05, is 27 years.
@@
 
-We applied the same calculations to 1032 populations, as we did in the
-main text. We then obtain a distribution of the minimum time required to
-measure the ``true" long-term growth rate (Fig.
-\ref{fig:min_time_growth_dist}). We see a bimodal distribution with many
-populations required 30+ years to estimate the long-term growth rate.
-The large number of short years required is due to cases where the
-entire time series is consistently increasing or decreasing at the same
-rate.
+We applied the same calculations to 1032 population time series. We then
+obtain a distribution of the minimum time required to measure the
+``true" long-term growth rate (Fig. \ref{fig:min_time_growth_dist}). We
+see a bimodal distribution with many populations required 30+ years to
+estimate the long-term growth rate. The large number of short years
+required is due to cases where the entire time series is consistently
+increasing or decreasing at the same rate.
 
@@
 good fit. Generalized additive models (GAMs) are more general than
-general linear models and allow more flexible. GAMs are models where a
-response variable depends on unknown smooth functions of explanatory
-variables. GAMs, therefore, can identify relationships between response
-variables and explanatory variables that are non-linear and perhaps more
-complicated. The downside of GAMs is they typically require more data
-and are also prone to overfitting (citations).
+general linear models and allow more flexibility (Wood 2006). GAMs are
+models where a response variable depends on unknown smooth functions of
+explanatory variables. GAMs, therefore, can identify relationships
+between response variables and explanatory variables that are non-linear
+and perhaps more complicated. The downside of GAMs is they typically
+require more data and are also prone to overfitting.
 
@@
 calculate the minimum time series required to detect trends over time
-according to a GAM model. We hypothesize that GAMs should require less
+according to a GAM model. We hypothesized that GAMs should require less
 time to detect a trend as they are more flexible than linear regression.
@@
 \hypertarget{refs}{}
+\hypertarget{ref-Bolker2008}{}
+Bolker, Benjamin M. 2008. \emph{Ecological Models and Data in R}. 1st
+ed. Princeton, New Jersey: Princeton University Press.
+
+\hypertarget{ref-Gerber1999}{}
+Gerber, L R, D P DeMaster, and P M Kareiva. 1999. ``Gray whales and the
+value of monitering data in implementing the U.S. endangered species
+act.'' \emph{Conservation Biology} 13 (5): 1215--9.
+
+\hypertarget{ref-Johnson2015}{}
+Johnson, Paul CD, Sarah JE Barry, Heather M Ferguson, and Pie Müller.
+2015. ``Power analysis for generalized linear mixed models in ecology
+and evolution.'' \emph{Methods in Ecology and Evolution} 6 (2): 133--42.
+doi:\href{https://doi.org/10.1111/2041-210X.12306}{10.1111/2041-210X.12306}.
+
 \hypertarget{ref-Myhrvold2015}{}
@@
 
+\hypertarget{ref-RCoreTeam2016}{}
+R Core Team. 2016. ``R: A language and environment for statistical
+computing.'' Vienna, Austria: R Foundation for Statistical Computing.
+\url{https://www.r-project.org/}.
+
 \hypertarget{ref-White2014}{}
@@
 
+\hypertarget{ref-Wood2006}{}
+Wood, S.N. 2006. \emph{Generalized Additive Models: An Introduction with
+R}. Chapman Hall/CRC.
+
 \end{document}

### 2017-10-01 writing/manuscript_draft.Rmd
Version submitted to Biological Conservation and the second version to PeerJ preprints

@@
 ---
-
+<!--
 Title: Minimum time required to detect population trends: the need for long-term monitoring programs
@@
 
-$^2$Corresponding author: eawhite@ucdavis.edu <!--, 1 480 203 7931-->
+$^2$Corresponding author: eawhite@ucdavis.edu , 1 480 203 7931
 \vspace{3 mm}
 
-Number of words (including 300 words per figure):  5900 
+Number of words (including 300 words per figure):  5950 
 \vspace{1 mm}
@@
 Number of figures: 3 
+\vspace{1 mm}
+
+Number of references: 32
 \vspace{1 mm}
@@
 
-<!--To be submitted to: \emph{Biological Conservation}
+To be submitted to: \emph{Biological Conservation}
 \vspace{3 mm}
 
-Available as a preprint at: www.peerj.com
+Available as a preprint at: https://peerj.com/preprints/3168/
 \vspace{3 mm}
@@
 Color does not need to be used for any figures in print
-\vspace{3 mm}-->
+\vspace{3 mm}
 
@@
 
-\pagebreak
+\pagebreak-->
 
+ \textbf{Minimum time required to detect population trends: the need for long-term monitoring programs}
+ \vspace{3 mm}
+
+Easton R. White$^{*}$
+
+\begin{footnotesize} Center for Population Biology, University of California, Davis, California 95616 USA
+
+*eawhite@ucdavis.edu
+\end{footnotesize}
+
+\vspace{5 mm}
 \linenumbers
@@
 We argue that statistical power needs to be considered in monitoring program design and evaluation.
-Time series less than 15-20 years are likely underpowered and potentially misleading.
+Time series less than 15-20 years are likely under-powered and potentially misleading.
 \end{abstract}
@@
 Scientists need to know the time series length required to address a specific question.
-A short time series may lead to wrong conclusions given large natural year-to-year variability.
+A short time series may lead to wrong conclusions given large natural year-to-year variability [@McCain2016].
 Managers need to know when action is needed for a population.
-Therefore, they must understand when population trend over time is actually meaningful.
-The IUCN Red List Categories and Criteria suggest, under Criterion A2, a species qualifies as vulnerable if it has experienced a 30% decline over 10 years, or 3 generations [@IUCN2012].
+Therefore, managers must understand when population trend over time is actually meaningful.
+For example, the IUCN Red List Categories and Criteria suggest, under Criterion A2, a species qualifies as vulnerable if it has experienced a 30% decline over 10 years, or 3 generations [@IUCN2012].
 For both scientific and management questions, because sampling is typically expensive, we also do not want to sample for longer than is necessary.
@@
 Simple models can be simulated with parameter values corresponding to a population of interest [@Gerrodette1987].
-Statistical power can then be calculated as the fraction of simulations that meet some criteria.
+Statistical power is the proportion of simulations that meet some set of criteria.
 The specific criteria depend on the question at hand.
@@
 For example, time series can be simulated for different lengths of time.
-From these simulations, the minimum time series length required is calculated to meet certain levels of statistical significance and power [@Bolker2008].
+From these simulations, the minimum time series length required to meet certain levels of statistical significance and power is estimated [@Bolker2008].
 
 In addition to using simulations, empirical time series can also be used.
-Multiple replicates of similar populations are usually not available, but it is possible to subsample an empirical time series [@Gerber1999].
-Subsamples of different lengths can then be evaluated to see which fraction of subsamples meet some criteria, again a measure of statistical power.
-Similar to the simulation approach, these measures of power can be used to determine the minimum time series required for a particular question of interest.
+Multiple replicates of similar populations are usually not available, but it is possible to subsample an empirical time series [@Gerber1999; @Brashares2005].
+Subsamples of different lengths can then be evaluated to estimate the proportion of subsamples meeting some criteria, again a measure of statistical power.
+Similar to the simulation approach, this measures of power can be used to determine the minimum time series required for a particular question of interest.
 
@@
 Using simple populations models, they found that the allocation of effort depends on environmental variation, spatial and temporal autocorrelation, and observer error.
-@Rueda-Cediel2015 also used a modeling approach, but parameterized a model specifically for a threatened snail, \emph{Tasmaphena lamproides}.
+@Rueda-Cediel2015 also used a modeling approach, but parameterized a model specific for a threatened snail, \emph{Tasmaphena lamproides}.
 They found that for this short-lived organism, 15 years was adequate to assess long-term trends in abundance.
@@
 
-Statistical power is then the fraction of simulations that meet some criteria.
-Here, our criteria is whether the slope parameter from linear regression is significant at the 0.05 threshold.
+Statistical power is then the proportion of simulations that meet some criteria.
+Here, our criteria is whether the slope parameter from linear regression is significant at the 0.05 threshold with statistical power of 0.8.
 Statistical power of 0.8 would indicate that, if there was a true trend in abundance, we would have a 0.8 probability of detecting the trend.
@@
 The data includes information on `r length(table(pop_info$Binomial))` vertebrate species with a focus on mammals, birds, and fish.
-The data also includes information on generation length and census specifications.
+The data also includes information on generation length and survey specifications.
 For each time series, we also calculated variables of interest like variance in population size, long-term trend in abundance (slope coefficient from simple linear regression), and temporal autocorrelation.
@@
 Next, we run a linear regression for each subsampled time series.
-Then, we determine the fraction of subsamples of a particular length that have estimated slope coefficients which are statistically different from zero.
-We only look at the fraction of samples where the long-term, or ``true'', time series also has a significant slope.
-This fraction is a measure of statistical power.
+Then, we determine the proportion of subsamples of a particular length that have estimated slope coefficients which are statistically different from zero.
+We only look at the proportion of samples where the long-term, or ``true'', time series also has a significant slope.
+This proportion is a measure of statistical power.
 Lastly, we determine which subsample length is required to achieve a certain threshold of statistical power (0.8, @Cohen1992).
-The subsampled length meeting these criteria is the minimum time series length required ($T_{min}$).
+The minimum subsampled length that met these criteria is the minimum time series length required ($T_{min}$).
 
@@
 Ideally, the specific model choice should be tailored to the population of interest.
-We explored how the simulation approach can be applied to more biologically-realistic population models (Fig.
+We explored how the simulation approach could be applied to more biologically-realistic population models (Fig.
 \ref{fig:shark_example}).
@@
 
-```{r,echo=F,fig.cap='Distributions of the minimum time required for four different biological classes.
+```{r,echo=F,fig.cap='Distributions of the minimum time required for populations from four different biological classes.
 The minimum time required calculation corresponds to a significance level of 0.05 and statistical power of 0.8.\\label{fig:min_time_dist}',fig.width=6}
@@
 Therefore, life-history characteristics may be more important for other questions, like estimating species extinction risk [@Hutchings2012].
+For example, @Blanchard2007 used detailed simulations of spatially-distributed fisheries to compare survey designs.
+They found that statistical power depended on survey design, temperature preferences and the degree of population patchiness.
 
@@
 \ref{fig:correlates}).
-@Morrison2008 also studied the optimal allocation of sampling effort in space versus time, but used emprical data from a long-term census of the collared pika (\emph{Ochotona collaris}) found in the Yukon.
+@Morrison2008 also studied the optimal allocation of sampling effort in space versus time, but used emprical data from a long-term survey of the collared pika (\emph{Ochotona collaris}) found in the Yukon.
 They estimated long-term growth rates among three subpopulations over a 10-year period.
-They found that censuses less than 5 years may be misleading and that extrapolating from one population to another, even when nearby geographically, may be untenable.
+They found that surveys less than 5 years may be misleading and that extrapolating from one population to another, even when nearby geographically, may be untenable.
 
@@
 However, unlike @Seavy2007, our results indicate that longer than 10 years is often needed to achieve high statistical power.
-Therefore, the design of monitoring programs should include calculations of statistical power, the allocation of sampling in space versus time [@Rhodes2011], and metrics to increase precision.
-Ideally, a formal decision analysis to evaluate these different factors would be conducted to design or assess any monitoring program [@Hauser2006;@McDonald-Madden2010].
+
+The design of monitoring programs should include calculations of statistical power, the allocation of sampling in space versus time [@Rhodes2011], and metrics to increase precision.
+Ideally, a formal decision analysis to evaluate these different factors would be conducted to design or assess any monitoring program [@Hauser2006; @McDonald-Madden2010].
+This type of formal decision analysis would also include information on the costs of monitoring.
+These costs include the actual costs of sampling [@Brashares2005] and the ecological costs on inaction [@Thompson2000].
 
@@
 \ref{fig:min_time_vs_alpha_beta}).
-In an ideal setting, we would build a specific population model parameterized for each population of interest.
+In an ideal setting, we would build a specific population model parameterized for each population of interest [@McCain2016].
 Then, model simulations could be used to estimate the minimum time series required to address each specific question of interest.
@@
 \ref{fig:correlates}).
+
 Our work implies that for many populations, time series less than 15-20 years are probably not reliable for detecting population trends.
@@
 In addition, a time series that is too long may be a poor use of already limited funds [@Gerber1999].
+However, short time series data can not be used as an excuse for inaction.
+For example, @Thompson2000 demonstrate that long-term studies may be required to detect population trends for certain species.
+When a population trend is detected, it may be too late for management action.
+In these situations, the precautionary principle may be more appropriate [@Thompson2000].
 Future work should examine other species, with a wider range of life history characteristics.

### 2017-10-01 writing/manuscript_draft.tex
Version submitted to Biological Conservation and the second version to PeerJ preprints

@@
 
-Title: Minimum time required to detect population trends: the need for
-long-term monitoring programs \vspace{7 mm}
-
-Author: \textsc{Easton R. White$^{1,2}$} \vspace{3 mm}
-
-Address:\emph{
-    \\$^1$Center for Population Biology \\
-    University of California, Davis \\
-    2320 Storer Hall \\
-        University of California, Davis \\
-        One Shields Avenue Davis, CA, USA} \vspace{3 mm}
-
-\(^2\)Corresponding author:
-\href{mailto:eawhite@ucdavis.edu}{\nolinkurl{eawhite@ucdavis.edu}}
+\textbf{Minimum time required to detect population trends: the need for long-term monitoring programs}
 \vspace{3 mm}
 
-Number of words (including 300 words per figure): 5900 \vspace{1 mm}
+Easton R. White\(^{*}\)
 
-Number of figures: 3 \vspace{1 mm}
+\begin{footnotesize} Center for Population Biology, University of California, Davis, California 95616 USA
 
-Supplementary material: 11 figures and 1 table \vspace{3 mm}
+*eawhite@ucdavis.edu
+\end{footnotesize}
 
-Data and code for all the figures and tables can be found at
-(\url{https://github.com/erwhite1/time-series-project}).
-
-\vspace{3 mm}
-
-\pagebreak
+\vspace{5 mm}
 
@@
 to address a specific question. A short time series may lead to wrong
-conclusions given large natural year-to-year variability. Managers need
-to know when action is needed for a population. Therefore, they must
-understand when population trend over time is actually meaningful. The
-IUCN Red List Categories and Criteria suggest, under Criterion A2, a
-species qualifies as vulnerable if it has experienced a 30\% decline
-over 10 years, or 3 generations (IUCN 2012). For both scientific and
-management questions, because sampling is typically expensive, we also
-do not want to sample for longer than is necessary. For example, Gerber,
-DeMaster, and Kareiva (1999) investigated the minimum time series
-required to estimate population growth of the endangered, but
-recovering, eastern North Pacific gray whale
-(\emph{Eschrichtius robustus}). They used a long-term census to
-retroactively determine the minimum time series required to assess
-threat status. They found that only 11 years were needed, eight years
-before the delisting decision was made. This highlights the importance
-of estimating the minimum time series required as an earlier decision
-would have saved time and money (Gerber, DeMaster, and Kareiva 1999).
+conclusions given large natural year-to-year variability (McCain,
+Szewczyk, and Knight 2016). Managers need to know when action is needed
+for a population. Therefore, managers must understand when population
+trend over time is actually meaningful. For example, the IUCN Red List
+Categories and Criteria suggest, under Criterion A2, a species qualifies
+as vulnerable if it has experienced a 30\% decline over 10 years, or 3
+generations (IUCN 2012). For both scientific and management questions,
+because sampling is typically expensive, we also do not want to sample
+for longer than is necessary. For example, Gerber, DeMaster, and Kareiva
+(1999) investigated the minimum time series required to estimate
+population growth of the endangered, but recovering, eastern North
+Pacific gray whale (\emph{Eschrichtius robustus}). They used a long-term
+census to retroactively determine the minimum time series required to
+assess threat status. They found that only 11 years were needed, eight
+years before the delisting decision was made. This highlights the
+importance of estimating the minimum time series required as an earlier
+decision would have saved time and money (Gerber, DeMaster, and Kareiva
+1999).
 
@@
 be simulated with parameter values corresponding to a population of
-interest (Gerrodette 1987). Statistical power can then be calculated as
-the fraction of simulations that meet some criteria. The specific
-criteria depend on the question at hand. For example, given a time
-series, when is the slope from linear regression significantly different
-from zero? In other words, when is the time series significantly
-increasing or decreasing? It is then possible to determine how power
-changes with a variable of interest. For example, time series can be
-simulated for different lengths of time. From these simulations, the
-minimum time series length required is calculated to meet certain levels
-of statistical significance and power (Bolker 2008).
+interest (Gerrodette 1987). Statistical power is the proportion of
+simulations that meet some set of criteria. The specific criteria depend
+on the question at hand. For example, given a time series, when is the
+slope from linear regression significantly different from zero? In other
+words, when is the time series significantly increasing or decreasing?
+It is then possible to determine how power changes with a variable of
+interest. For example, time series can be simulated for different
+lengths of time. From these simulations, the minimum time series length
+required to meet certain levels of statistical significance and power is
+estimated (Bolker 2008).
 
@@
 available, but it is possible to subsample an empirical time series
-(Gerber, DeMaster, and Kareiva 1999). Subsamples of different lengths
-can then be evaluated to see which fraction of subsamples meet some
-criteria, again a measure of statistical power. Similar to the
-simulation approach, these measures of power can be used to determine
-the minimum time series required for a particular question of interest.
+(Gerber, DeMaster, and Kareiva 1999; Brashares and Sam 2005). Subsamples
+of different lengths can then be evaluated to estimate the proportion of
+subsamples meeting some criteria, again a measure of statistical power.
+Similar to the simulation approach, this measures of power can be used
+to determine the minimum time series required for a particular question
+of interest.
 
@@
 et al. (2015) also used a modeling approach, but parameterized a model
-specifically for a threatened snail, \emph{Tasmaphena lamproides}. They
+specific for a threatened snail, \emph{Tasmaphena lamproides}. They
 found that for this short-lived organism, 15 years was adequate to
@@
 
-Statistical power is then the fraction of simulations that meet some
+Statistical power is then the proportion of simulations that meet some
 criteria. Here, our criteria is whether the slope parameter from linear
-regression is significant at the 0.05 threshold. Statistical power of
-0.8 would indicate that, if there was a true trend in abundance, we
-would have a 0.8 probability of detecting the trend. We also tested the
-effect of varying both the significance level and statistical power
-(Fig. \ref{fig:min_time_vs_alpha_beta}).
+regression is significant at the 0.05 threshold with statistical power
+of 0.8. Statistical power of 0.8 would indicate that, if there was a
+true trend in abundance, we would have a 0.8 probability of detecting
+the trend. We also tested the effect of varying both the significance
+level and statistical power (Fig. \ref{fig:min_time_vs_alpha_beta}).
 
@@
 vertebrate species with a focus on mammals, birds, and fish. The data
-also includes information on generation length and census
+also includes information on generation length and survey
 specifications. For each time series, we also calculated variables of
@@
 run a linear regression for each subsampled time series. Then, we
-determine the fraction of subsamples of a particular length that have
+determine the proportion of subsamples of a particular length that have
 estimated slope coefficients which are statistically different from
-zero. We only look at the fraction of samples where the long-term, or
-``true'', time series also has a significant slope. This fraction is a
+zero. We only look at the proportion of samples where the long-term, or
+``true'', time series also has a significant slope. This proportion is a
 measure of statistical power. Lastly, we determine which subsample
 length is required to achieve a certain threshold of statistical power
-(0.8, Cohen (1992)). The subsampled length meeting these criteria is the
-minimum time series length required (\(T_{min}\)).
+(0.8, Cohen (1992)). The minimum subsampled length that met these
+criteria is the minimum time series length required (\(T_{min}\)).
 
@@
 specific model choice should be tailored to the population of interest.
-We explored how the simulation approach can be applied to more
+We explored how the simulation approach could be applied to more
 biologically-realistic population models (Fig. \ref{fig:shark_example}).
@@
 \includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-4-1.pdf}
-\caption{Distributions of the minimum time required for four different
-biological classes. The minimum time required calculation corresponds to
-a significance level of 0.05 and statistical power of
+\caption{Distributions of the minimum time required for populations from
+four different biological classes. The minimum time required calculation
+corresponds to a significance level of 0.05 and statistical power of
 0.8.\label{fig:min_time_dist}}
@@
 questions, like estimating species extinction risk (J. A. Hutchings et
-al. 2012).
+al. 2012). For example, Blanchard, Maxwell, and Jennings (2007) used
+detailed simulations of spatially-distributed fisheries to compare
+survey designs. They found that statistical power depended on survey
+design, temperature preferences and the degree of population patchiness.
 
@@
 studied the optimal allocation of sampling effort in space versus time,
-but used emprical data from a long-term census of the collared pika
+but used emprical data from a long-term survey of the collared pika
 (\emph{Ochotona collaris}) found in the Yukon. They estimated long-term
 growth rates among three subpopulations over a 10-year period. They
-found that censuses less than 5 years may be misleading and that
+found that surveys less than 5 years may be misleading and that
 extrapolating from one population to another, even when nearby
@@
 results indicate that longer than 10 years is often needed to achieve
-high statistical power. Therefore, the design of monitoring programs
-should include calculations of statistical power, the allocation of
-sampling in space versus time (Rhodes and Jonzen 2011), and metrics to
-increase precision. Ideally, a formal decision analysis to evaluate
-these different factors would be conducted to design or assess any
-monitoring program (Hauser, Pople, and Possingham 2006; McDonald-Madden
-et al. 2010).
+high statistical power.
+
+The design of monitoring programs should include calculations of
+statistical power, the allocation of sampling in space versus time
+(Rhodes and Jonzen 2011), and metrics to increase precision. Ideally, a
+formal decision analysis to evaluate these different factors would be
+conducted to design or assess any monitoring program (Hauser, Pople, and
+Possingham 2006; McDonald-Madden et al. 2010). This type of formal
+decision analysis would also include information on the costs of
+monitoring. These costs include the actual costs of sampling (Brashares
+and Sam 2005) and the ecological costs on inaction (Thompson et al.
+2000).
 
@@
 \ref{fig:min_time_vs_alpha_beta}). In an ideal setting, we would build a
-specific population model parameterized for each population of interest.
-Then, model simulations could be used to estimate the minimum time
-series required to address each specific question of interest. Clearly,
-this is not always practical, especially if conducting analyses for a
-wide array of species as we do here. In addition, our statistical models
-suggest that \(T_{min}\) does not correlate with any biological
-variables of interest, at least for the question of linear regression
-(Fig. \ref{fig:biological_correlates}). Therefore, it is not possible to
-use these results to predict \(T_{min}\) for another population, even if
-the population is of a species with a similar life-history to one in our
+specific population model parameterized for each population of interest
+(McCain, Szewczyk, and Knight 2016). Then, model simulations could be
+used to estimate the minimum time series required to address each
+specific question of interest. Clearly, this is not always practical,
+especially if conducting analyses for a wide array of species as we do
+here. In addition, our statistical models suggest that \(T_{min}\) does
+not correlate with any biological variables of interest, at least for
+the question of linear regression (Fig.
+\ref{fig:biological_correlates}). Therefore, it is not possible to use
+these results to predict \(T_{min}\) for another population, even if the
+population is of a species with a similar life-history to one in our
 database.
@@
 in population size, and the temporal autocorrelation (Fig.
-\ref{fig:correlates}). Our work implies that for many populations, time
-series less than 15-20 years are probably not reliable for detecting
-population trends. This result highlights the importance of long-term
-monitoring programs. From both a scientific and management perspective
-estimates of \(T_{min}\) are important. If a time series is too short,
-we lack statistical power to reliably detect long-term population
-trends. In addition, a time series that is too long may be a poor use of
-already limited funds (Gerber, DeMaster, and Kareiva 1999). Future work
-should examine other species, with a wider range of life history
-characteristics. In addition, similar approaches can be used to
-determine the minimum time series length required to address additional
-questions of interest.
+\ref{fig:correlates}).
+
+Our work implies that for many populations, time series less than 15-20
+years are probably not reliable for detecting population trends. This
+result highlights the importance of long-term monitoring programs. From
+both a scientific and management perspective estimates of \(T_{min}\)
+are important. If a time series is too short, we lack statistical power
+to reliably detect long-term population trends. In addition, a time
+series that is too long may be a poor use of already limited funds
+(Gerber, DeMaster, and Kareiva 1999). However, short time series data
+can not be used as an excuse for inaction. For example, Thompson et al.
+(2000) demonstrate that long-term studies may be required to detect
+population trends for certain species. When a population trend is
+detected, it may be too late for management action. In these situations,
+the precautionary principle may be more appropriate (Thompson et al.
+2000). Future work should examine other species, with a wider range of
+life history characteristics. In addition, similar approaches can be
+used to determine the minimum time series length required to address
+additional questions of interest.
 
@@
 \hypertarget{refs}{}
+\hypertarget{ref-Blanchard2007}{}
+Blanchard, Julia L, David L Maxwell, and Simon Jennings. 2007. ``Power
+of monitoring surveys to detect abundance trends in depleted
+populations: the effects of density-dependent habitat use, patchiness,
+and climate change.'' \emph{ICES Journal of Marine Science} 65 (1):
+111--20.
+
 \hypertarget{ref-Bolker2008}{}
@@
 ed. Princeton, New Jersey: Princeton University Press.
+
+\hypertarget{ref-Brashares2005}{}
+Brashares, Justin S, and Moses K Sam. 2005. ``How much is enough?
+Estimating the minimum sampling required for effective monitoring of
+African reserves.'' \emph{Biodiversity and Conservation} 14: 2709--22.
+doi:\href{https://doi.org/10.1007/s10531-005-8404-z}{10.1007/s10531-005-8404-z}.
 
@@
 
+\hypertarget{ref-McCain2016}{}
+McCain, Christy Marie, Tim Szewczyk, and Kevin Bracy Knight. 2016.
+``Population variability complicates the accurate detection of climate
+change responses.'' \emph{Global Change Biology} 22 (6): 2081--93.
+doi:\href{https://doi.org/10.1111/gcb.13211}{10.1111/gcb.13211}.
+
 \hypertarget{ref-McDonald-Madden2010}{}
@@
 doi:\href{https://doi.org/10.1016/j.biocon.2007.08.007}{10.1016/j.biocon.2007.08.007}.
+
+\hypertarget{ref-Thompson2000}{}
+Thompson, Paul M, Ben Wilson, Kate Grellier, and Philip S Hammond. 2000.
+``Combining power analysis and population viability analysis to compare
+traditional and precautionary approaches to conservation of coastal
+cetaceans.'' \emph{Conservation Biology} 14 (5): 1253--63.
 

### 2017-10-01 writing/supp_mat_draft.Rmd
Version submitted to Biological Conservation and the second version to PeerJ preprints

@@
 
-June 22, 2017
  \end{center}
@@
 \item We can call each set of subsamples, of the same length, a set.
-The fraction of subsamples within a set that show significant trends (significant slope coefficient $\alpha$ less than 0.05) is the statistical power.
+The proportion of subsamples within a set that show significant trends (significant slope coefficient $\alpha$ less than 0.05) is the statistical power.
 It is important to note that we only consider subsamples to be significant if they are significant in the same direction as the complete 35-year time series.
@@
 Traditionally, this has been at 0.8, however, this is purely historical.
-Statistical power of 0.8 implies if a true trend is present, or there is a real change in abundance, then we will detect this trend 0.8 fraction of the time.
+Statistical power of 0.8 implies if a true trend is present, or there is a real change in abundance, then we will detect this trend 0.8 proportion of the time.
   \item We then determine the minimum time series length ($T_{min}$) required to achieve that level of statistical power.
@@
 We simulate this model repetitively for different lengths of time series.
-Statistical power is the fraction of time series of a set length that showed significant trends in abundance, given a true trend in abundance.
+Statistical power is the proportion of time series of a set length that showed significant trends in abundance, given a true trend in abundance.
 *How do I determine significance here?

### 2018-02-19 writing/manuscript_draft.Rmd
first update of short format version of the article

@@
 header-includes: \usepackage{float} \usepackage{lineno} \usepackage[firstinits=true,style=authoryear-comp,natbib=true,doi=false,isbn=false,url=false,uniquename=false,uniquelist=false,sorting=none]{biblatex}
-
- \linenumbers
 
@@
 ---
-<!--
+
 Title: Minimum time required to detect population trends: the need for long-term monitoring programs
@@
 
-Number of words (including 300 words per figure):  5950 
+Number of words (including 300 words per figure):  6,950
 \vspace{1 mm}
 
-Number of figures: 3 
+Number of figures: 5 
 \vspace{1 mm}
 
-Number of references: 32
+Number of references: 34
 \vspace{1 mm}
 
-Supplementary material: 11 figures and 1 table 
+Supplementary material: 9 figures and 3 tables
 \vspace{3 mm}
@@
 
-\pagebreak-->
-
+<!-- Shorter title for PeerJ
  \textbf{Minimum time required to detect population trends: the need for long-term monitoring programs}
@@
 \vspace{5 mm}
+-->
 \linenumbers
@@
 When is a population time series long enough to address a question of interest?
-We determine the minimum time series length required to detect significant increases or decreases in population abundance.
-To address this question, we use simulation methods and examine 878 populations of vertebrate species.
-Here we show that 15-20 years of continuous monitoring are required in order to achieve a high level of statistical power.
+I determine the minimum time series length required to detect significant increases or decreases in population abundance.
+To address this question, I use simulation methods and examine 822 populations of vertebrate species.
+Here I show that on average 15.9 years of continuous monitoring are required in order to achieve a high level of statistical power.
+However, there is a wide distribution around this average, casting doubt on simple rules of thumb.
 For both simulations and the time series data, the minimum time required depends on trend strength, population variability, and temporal autocorrelation.
+However, there were no life-history traits (e.g.
+generation length) that were predictive of the minimum time required.
 These results point to the importance of sampling populations over long periods of time.
-We argue that statistical power needs to be considered in monitoring program design and evaluation.
-Time series less than 15-20 years are likely under-powered and potentially misleading.
+I argue that statistical power needs to be considered in monitoring program design and evaluation.
+Short time series are likely under-powered and potentially misleading.
 \end{abstract}
@@
 Therefore, managers must understand when population trend over time is actually meaningful.
-For example, the IUCN Red List Categories and Criteria suggest, under Criterion A2, a species qualifies as vulnerable if it has experienced a 30% decline over 10 years, or 3 generations [@IUCN2012].
+For example, the International Union for Conservation of Nature (IUCN) Red List Categories and Criteria suggest, under Criterion A2, a species qualifies as vulnerable if it has experienced a 30% decline over 10 years, or three generations [@IUCN2012].
 For both scientific and management questions, because sampling is typically expensive, we also do not want to sample for longer than is necessary.
@@
 This highlights the importance of estimating the minimum time series required as an earlier decision would have saved time and money [@Gerber1999].
+Further, waiting too long to make a decision can imperil a species where management action could have been taken earlier [@Martin2012; @Martin2017].
 
@@
 They found that for this short-lived organism, 15 years was adequate to assess long-term trends in abundance.
-@Hatch2003 examined seabird populations and determined the minimum time required to achieve high statistical power ranged between 11 and 69 years.
 However, these studies, and other past work, have typically focused either on theoretical aspects of monitoring design or focused on only a few species.
 
-We use both simulations and empirical time series to determine the minimum number of years required to address several questions.
-We estimate the minimum time series length required ($T_{min}$) to assess long-term changes in abundance via simple linear regression.
-First, we estimate $T_{min}$ using a simulation approach.
-Then we examine 878 population time series to estimate $T_{min}$.
-In the supplementary material, we determine $T_{min}$ for related ideas: calculating long-term growth rates, using more complicated population models, varying statistical level and power, and the use of generalized additive models.
+I use both simulations and empirical time series to determine the minimum number of years required to address several questions.
+I estimate the minimum time series length required ($T_{min}$) to assess long-term changes in abundance via simple linear regression.
+First, I estimate $T_{min}$ using a simulation approach.
+Then I examine 822 population time series to estimate $T_{min}$.
+In the supplementary material, I determine $T_{min}$ for related ideas: using more complicated population models, varying statistical level and power, and the use of generalized additive models.
 
@@
 This is the same approach one might use in sample size calculations for any experimental design too complicated for simple power analyses [@Bolker2008;@Johnson2015].
-We only briefly discuss this approach as it has been described elsewhere.
-Essentially, we use a population model and repetitively simulate it for a number of years.
-This approach requires us to determine values for our model parameters (e.g.
+I only briefly discuss this approach as it has been described elsewhere.
+Essentially, a population model is simulated repetitively for a number of years.
+This approach requires us to determine values for model parameters (e.g.
 population variability).
@@
 the rate of increase).
-It is important to note that any population model could be substituted for equation \ref{sim_model}, as we do in the supplementary material (Figs.
+It is important to note that any population model could be substituted for equation \ref{sim_model}, as in the supplementary material (Figs.
 \ref{fig:min_time_dist_log_pop}, \ref{fig:shark_example}).
@@
 Here, our criteria is whether the slope parameter from linear regression is significant at the 0.05 threshold with statistical power of 0.8.
-Statistical power of 0.8 would indicate that, if there was a true trend in abundance, we would have a 0.8 probability of detecting the trend.
-We also tested the effect of varying both the significance level and statistical power (Fig.
+Statistical power of 0.8 would indicate that, if there was a true trend in abundance, there would be a 0.8 probability of detecting the trend.
+Values of 0.05 for the significance level and 0.8 statistical power are historical and it is important to examine the effect of changing these values (Fig.
 \ref{fig:min_time_vs_alpha_beta}).
@@
 \ref{fig:theoretical_approach}b).
-Where power is greater than 0.8 (the dotted line), that is the minimum time required ($T_{min}$).
+Where power is greater than 0.8 (the dotted line), that is the minimum time required ($T_{min}$) to be confident in the detection of a long-term trend in abundance.
 
@@
 ## Data source
-We use a database of `r length(table(dat$ID))` population time series compiled in [@Keith2015]; they compared the predictability of growth rates among populations.
-The data are originally from the Global Population Dynamics Database [@GPDD2010].
-We filtered out short time series (less than `r pop_info$series_length[1]` years), and those with missing data, and were left with `r nrow(pop_info)` time series.
+I use a database of `r length(table(dat$ID))` population time series compiled in [@Keith2015]; they compared the predictability of growth rates among populations.
+The data are originally from the Global Population Dynamics Database [@GPDD2010] and several other sources [@Keith2015].
+I filtered out short time series (less than `r pop_info$series_length[1]` years), and those with missing data, leaving `r nrow(pop_info)` time series.
 The data includes information on `r length(table(pop_info$Binomial))` vertebrate species with a focus on mammals, birds, and fish.
 The data also includes information on generation length and survey specifications.
-For each time series, we also calculated variables of interest like variance in population size, long-term trend in abundance (slope coefficient from simple linear regression), and temporal autocorrelation.
+For each time series, I also calculate other variables of interest: coefficient of variance in population size, long-term trend in abundance (slope coefficient from simple linear regression), and temporal autocorrelation.
 All analyses were conducted in R [@RCoreTeam2016].
-<!--Lastly, we also calculated the minimum number of years required (10 years or 3 generations) under IUCN Red List Criterion A2 to classify a species as vulnerable [@IUCN2012].-->
 
@@
 ```
-For a subset of populations ($n =$ `r nrow(LPI_pop_info)`), we had information on biological characteristics from another paper [@Myhrvold2015], including body size and generation length.
+For a subset of populations ($n =$ `r nrow(LPI_pop_info)`), there is information on life-history traits available from another paper [@Myhrvold2015], including body size and generation length.
 All `r nrow(LPI_pop_info)` populations were birds.
-We examine how the minimum time required is related to these biological characteristics (Fig.
+I examine how the minimum time required is related to these life-history traits (Fig.
 \ref{fig:biological_correlates}).
@@
 
-We assume that each time series in our database is long enough to include all necessary information (e.g.
+I assume that each time series is long enough to include all necessary information (e.g.
 variability) about the population.
 In other words, each time series is a representative sample.
-We first take all possible contiguous subsamples of each time series.
+I first take all possible contiguous subsamples of each time series.
 For example, a time series of 35 years would have 34 possible contiguous subsamples of length 2, 34 possible contiguous subsamples of length 3, and continuing until 1 possible contiguous subsample of length 35 [@Gerber1999; @Giron-Nava2017].
-Next, we run a linear regression for each subsampled time series.
-Then, we determine the proportion of subsamples of a particular length that have estimated slope coefficients which are statistically different from zero.
-We only look at the proportion of samples where the long-term, or ``true'', time series also has a significant slope.
+Next, I run a linear regression for each subsampled time series.
+Then, I determine the proportion of subsamples of a particular length that have estimated slope coefficients which are statistically different from zero.
+I only look at the proportion of samples where the long-term, or ``true'', time series also has a significant slope.
 This proportion is a measure of statistical power.
-Lastly, we determine which subsample length is required to achieve a certain threshold of statistical power (0.8, @Cohen1992).
+Lastly, I determine which subsample length is required to achieve a certain threshold of statistical power (0.8, @Cohen1992).
 The minimum subsampled length that met these criteria is the minimum time series length required ($T_{min}$).
 
-In the supplementary material, we show how the same approach described here can be used to determine the minimum time required to estimate a population's geometric growth rate (Figs.
-\ref{fig:growth_rate},\ref{fig:min_time_growth_dist}).
-We also determine the minimum time required to estimate long-term trends according to generalized additive models, instead of the simple linear models used here (Fig.
+In the supplementary material, I show how the same approach described here for more complicated population models.
+I also determine the minimum time required to estimate long-term trends according to generalized additive models, instead of the simple linear models used here (Fig.
 \ref{fig:gam_example}).
@@
 
-We also examine how minimum time series calculations changed depending on the question of interest.
-In the supplementary material, we use similar calculations to determine the minimum time series required to estimate overall population growth rate (Fig.
+I also examine how minimum time series calculations changed depending on the question of interest.
+In the supplementary material, I use similar calculations to determine the minimum time series required to estimate overall population growth rate (Fig.
 \ref{fig:growth_rate}).
-We also explore how the method can be applied to more biologically-realistic population models (Fig.
+I also explore how the method can be applied to more biologically-realistic population models (Fig.
 \ref{fig:shark_example}.
-More specifically, we determine the minimum time required to estimate long-term population trends using a stochastic, age-structured model of lemon shark population dynamics in the Bahamas [@White2014].
-Lastly, we determine the minimum time required to estimate long-term trends according to generalized additive models, instead of the simple linear models used here.
+More specifically, I determine the minimum time required to estimate long-term population trends using a stochastic, age-structured model of lemon shark population dynamics in the Bahamas [@White2014].
+Lastly, I determine the minimum time required to estimate long-term trends according to generalized additive models, instead of the simple linear models used here.
 -->
@@
 
-We determined the minimum time series length ($T_{min}$) required to address a particular question of interest.
+I determined the minimum time series length ($T_{min}$) required to address a particular question of interest.
 What is the minimum time series length required to determine, via linear regression, the long-term population trend?
-Here, the minimum time series length required had high enough statistical power ($1-\beta$ greater than 0.8) for a set significance level ($\alpha$) of 0.05.
-It is also possible to alter $1-\beta$ and $\alpha$.
-Predictably, as we increased our level of power or decreased $\alpha$, $T_{min}$ increased (Fig.
+Here, the minimum time series length required had high enough statistical power (greater than 0.8) for a set significance level ($\alpha$) of 0.05.
+It is also possible to alter statistical power and $\alpha$.
+Predictably, with increased statistical power or decreased $\alpha$, $T_{min}$ increased (Fig.
 \ref{fig:min_time_vs_alpha_beta}).
-We estimated $T_{min}$ using two approaches.
-We briefly describe results from the simulation approach and then discuss our empirical approach.
+I then estimated $T_{min}$ using two approaches.
+I briefly describe results from the simulation approach and then discuss the empirical approach.
 
@@
 
-We constructed a general population model where the trend strength (i.e.
+I constructed a general population model where the trend strength (i.e.
 slope coefficient) over time could be a model parameter.
-We then ran simulated time series of different lengths.
-From these simulations we determined the minimum time series length required to achieve a certain level of statistical power.
-In line with past work [@Gerrodette1987], we found the $T_{min}$ increases (i.e.
+I then simulated time series of different lengths.
+From these simulations I determined the minimum time series length required to achieve a certain level of statistical power.
+In line with past work [@Gerrodette1987], I found the $T_{min}$ increases (i.e.
 more time is required) with decreases in trend strength and with increases in population variability (Figs.
@@
 
-We chose a simple model, but any other population model could be used (see example in Fig.
+I chose a simple model, but any other population model could be used (see example in Fig.
 \ref{fig:min_time_dist_log_pop}).
 Ideally, the specific model choice should be tailored to the population of interest.
-We explored how the simulation approach could be applied to more biologically-realistic population models (Fig.
+I explored how the simulation approach could be applied to more biologically-realistic population models (Fig.
 \ref{fig:shark_example}).
-More specifically, we determined the minimum time required to estimate long-term population trends using a stochastic, age-structured model of lemon shark population dynamics in the Bahamas [@White2014].
-We found that over 27 years of continuous monitoring were needed in this particular scenario (Fig.
+Specifically, I determined the minimum time required to estimate long-term population trends using a stochastic, age-structured model of lemon shark population dynamics in the Bahamas [@White2014].
+I found that over 27 years of continuous monitoring were needed in this particular scenario (Fig.
 \ref{fig:shark_example}).
@@
 
-```{r,echo=F,fig.cap='Distributions of the minimum time required for populations from four different biological classes.
-The minimum time required calculation corresponds to a significance level of 0.05 and statistical power of 0.8.\\label{fig:min_time_dist}',fig.width=6}
+```{r,echo=F,message=F,warning=F,fig.cap='(a) Distributions of the minimum time required for populations from four different biological classes.
+(b) Distribution of minimum time required for all populations regardless of biological class.
+The minimum time required calculation corresponds to a significance level of 0.05 and statistical power of 0.8.\\label{fig:min_time_dist}',fig.height=6}
 # Pull in analysis of average min time needed for **empirical species** (save analysis as Rdata file or something)
@@
 
-#only look at pops with significant long-term declines or increases
-#pop_info=pop_info[pop_info$trend_p_value<0.05,]
-#long_dat=subset(long_dat,long_dat$ID %in% pop_info$ID)
-#long_time_names=subset(long_time_names,long_time_names %in% pop_info$ID)
+require(ggplot2)
+plot1<- ggplot(aes(class,min_time_for_power),data=pop_info) + geom_violin() + geom_jitter(col=rgb(0.5,0.5,0.5,0.3),height = 0, width = 0.1) +
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(), axis.line = element_line(colour = "black"),axis.text = element_text(size=12),axis.title.x =element_text(size=16) ,axis.title.y =element_text(size=16) ) +
+ labs(x='biological class',y='minimum time required') +
+  geom_text(aes(x=4.4,y=33),label='(a)',size=6) +
+  scale_y_continuous(limits = c(0, 40))
 
-#plot(density(pop_info$min_time_for_power),las=1,ylab='',xlab='',cex.axis=1.2,main='')
+library(png)
+library(grid)
 
-counts <- table(pop_info$class,pop_info$min_time_for_power)
-barplot(counts, main=" ",
-        xlab=" ", ylab="", col=gray.colors(4),
-        ylim=c(0,50),xlim=c(0,40),
-        las=1,cex.names =1.2,cex.axis = 1.2,#density=35,
-        angle=c(20,40,60,80))
-legend('topright',legend=rownames(counts),col = gray.colors(4),pch = 15,bty ="n")
+fish <- annotation_custom(rasterGrob(readPNG("phylo-fish.png"), interpolate=TRUE), xmin=-2.5,ymin=35,ymax=40) 
+bird <- annotation_custom(rasterGrob(readPNG("phylo-bird.png"), interpolate=TRUE), xmin=-0.5, ymin=35) 
+shark <- annotation_custom(rasterGrob(readPNG("phylo-shark.png"), interpolate=TRUE), xmin=2.6,xmax=3.3, ymin=35) 
+mammal <- annotation_custom(rasterGrob(readPNG("phylo-mammal.png"), interpolate=TRUE), xmin=3.3, ymin=35) 
+plot1 <- plot1+ fish + bird + shark + mammal
 
-mtext(text = 'Frequency',2,line=3,cex=1.2)
-mtext(text = 'Minimum time required',1,line=3,cex=1.2)
+#counts <- table(pop_info$class,pop_info$min_time_for_power)
+plot2<- ggplot(aes(min_time_for_power),data=pop_info) + geom_bar() +  
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(), axis.line = element_line(colour = "black"),axis.text = element_text(size=12),axis.title.x =element_text(size=16) ,axis.title.y =element_text(size=16) ) +
+ labs(y='count',x='minimum time required')+
+  geom_text(aes(x=35,y=47),label='(b)',size=6)
+
+require(gridExtra)
+grid.arrange(plot1, plot2, nrow=2)
+
+# Code for barplot in my first submission
+# barplot(counts, main=" ",
+#         xlab=" ", ylab="", col=gray.colors(4),
+#         ylim=c(0,50),xlim=c(0,40),
+#         las=1,cex.names =1.2,cex.axis = 1.2,#density=35,
+#         angle=c(20,40,60,80))
+# legend('topright',legend=rownames(counts),col = gray.colors(4),pch = 15,bty ="n")
+# 
+# mtext(text = 'Frequency',2,line=3,cex=1.2)
+# mtext(text = 'Minimum time required',1,line=3,cex=1.2)
 ```
@@
 pop_info$abs_overall_trend = abs(pop_info$overall_trend)
-poisson_model <-glm(min_time_for_power~ abs_overall_trend + autocorrelation + variance + gen_len,family = 'poisson',data=pop_info)
-par(mfrow=c(2,2),mar=c(5,5,1,2),oma=c(0,0,0.5,0))
+time_series_correlates_poisson_model <-glm(min_time_for_power~ abs_overall_trend + autocorrelation + coefficient_variation,family = 'poisson',data=pop_info)
+#par(mfrow=c(2,2),mar=c(5,5,1,2),oma=c(0,0,0.5,0))
 #plot(poisson_model,cex.lab=1.2,las=1,cex.axis=1.2,main=' ')
@@
 #R^2 calculation (pg. 218 Zuur)
-R2 <- 100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance)
+R2 <- 100*(time_series_correlates_poisson_model$null.deviance - time_series_correlates_poisson_model$deviance)/(time_series_correlates_poisson_model$null.deviance)
 
@@
 
-```{r,echo=F, fig.cap='Minimum time required to estimate change in abundance correlated with (a) trend strength (absolute value of slope coefficient estimated from linear regression), (b) population variance (inter-annual variability in population size), and (c) temporal lag-1 autocorrelation.\\label{fig:correlates}',fig.height=7, fig.width=4}
+```{r,echo=F, fig.cap='Minimum time required to estimate change in abundance correlated with (a) trend strength (absolute value of slope coefficient estimated from linear regression), (b) coefficient of variation in interannual population size, and (c) temporal lag-1 autocorrelation.\\label{fig:correlates}',fig.height=7, fig.width=4}
 
@@
   par(mfrow=c(3,1),mar=c(3,2,2,1),oma=c(1,4.5,0,0.5))
-  plot(abs(pop_info$overall_trend),pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
+  plot(abs(pop_info$overall_trend),pop_info$min_time_for_power,pch=as.numeric(as.factor(pop_info$class))+14,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2,cex=1.3,ylim=c(0,35))
   mtext('| trend strength |',1,line=3,cex=1.2)
   mtext('(a)',3,line=-1.5,adj = 0.985,cex=1.2,font = 2)
-    plot(pop_info$variance,pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
-    mtext('population variance',1,line=3,cex=1.2)
+    plot(pop_info$coefficient_variation,pop_info$min_time_for_power,pch=as.numeric(as.factor(pop_info$class))+14,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2, xlim = c(0.2,2.2),cex=1.3,ylim=c(0,35))
+    mtext('coefficient of variation',1,line=3,cex=1.2)
     mtext('(b)',3,line=-1.5,adj = 0.985,cex=1.2,font = 2)
-      plot(pop_info$autocorrelation,pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
+      plot(pop_info$autocorrelation,pop_info$min_time_for_power,pch=as.numeric(as.factor(pop_info$class))+14,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2, xlim = c(-0.2,1),cex=1.3,ylim=c(0,35))
       mtext('autocorrelation',1,line=3,cex=1.2)
@@
 
+legend('bottomleft',legend=names(table(pop_info$class)),col = 'grey',pch = as.numeric(names(table(as.numeric(as.factor(pop_info$class))+14))),bty ="n",cex=1.2)
+
 ```
 
-We examined a database of `r nrow(pop_info)` separate population time series representing `r length(table(pop_info$Binomial))` species.
-This database consists of vertebrate species with a variety of life history characteristics (Fig.
+I examined a database of `r nrow(pop_info)` separate population time series representing `r length(table(pop_info$Binomial))` species.
+This database consisted of vertebrate species with a variety of life-history characteristics (Fig.
 \ref{fig:biological_correlates}).
-We limited our analyses to populations with at least 35 years of continuous sampling.
-We then examined the minimum time required to estimate long-term trends via linear regression.
+I limited analyses to populations with at least 35 years of continuous sampling.
+I then examined the minimum time required to estimate long-term trends via linear regression.
 
-Across all the populations we examined, we found an average minimum time series length required ($T_{min}$) of `r round(mean(pop_info$min_time_for_power),1)` ($\sigma$=`r round((var(pop_info$min_time_for_power))^0.5,1)`), with a wide distribution (Fig.
-\ref{fig:min_time_dist}).
+Across all the populations, I found an average minimum time series length required ($T_{min}$) of `r round(mean(pop_info$min_time_for_power),1)` (SD=`r round((var(pop_info$min_time_for_power))^0.5,1)`), with a wide distribution (Fig.
+\ref{fig:min_time_dist}b).
 Estimates of $T_{min}$ varied between biological class (Fig.
-\ref{fig:min_time_dist}).
+\ref{fig:min_time_dist}a).
 Ray-finned fish (class Actinopterygii) typically had estimates of $T_{min}$ over 20 years.
 Birds (class Aves) had a much wider distribution of $T_{min}$, but usually required less years of sampling.
-Differences between these classes were explained by differences in inter-annual variability in population size (Fig.
+Differences between these classes can be explained by differences in variability in population size and strength of trends in abundance (Fig.
 \ref{fig:class}).
-We also examined a subset of populations where life-history characteristics were known.
-None of these explanatory variables were predictive of the minimum time series length required (Fig.
-\ref{fig:biological_correlates}).
 
-The minimum time series length required was strongly correlated with trend strength, population variance, and autocorrelation (Fig.
+### Corrrelates for minimum time required
+
+The minimum time series length required was strongly correlated with trend strength (i.e.
+estimated slope coefficient from linear regression), coefficient of variation in population size, and autocorrelation in population size (Fig.
 \ref{fig:correlates}).
-This is in line based on our simulations and those of others [@Rhodes2011].
+This is in line based on simulations here and those of others [@Rhodes2011].
 Using a generalized linear model, with a Poisson error structure, all three of these explanatory variables were significant and had large effect sizes (see Table A1).
-Combined, trend strength, population variance, and autocorrelation account for about `r round(100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance),1)`% of the explained deviance [@Zuur2009] in minimum time series length required.
+Combined, trend strength, coefficient of variation in population, and autocorrelation account for `r round(100*(time_series_correlates_poisson_model$null.deviance - time_series_correlates_poisson_model$deviance)/(time_series_correlates_poisson_model$null.deviance),1)`% of the explained deviance [@Zuur2009] in minimum time series length required.
 
-Lastly, we tested model sensitivity by using generalized additive models (GAMs) instead of simple linear regression.
-Again, we examined the minimum time required to estimate long-term population trends (Fig.
+<!-- Minimum time required and biological correlates-->
+```{r, echo=F}
+load('../cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
+```
+
+For a subset of the populations I combined time series data with a data on life-history characteristics of amniotes [@Myhrvold2015].
+There was life-history information available for `r nrow(LPI_pop_info)` populations representing `r length(table(LPI_pop_info$Binomial))` different species, all of which were birds (Aves class).
+
+```{r, echo=F, fig.cap='Mimimum time required versus (a) generation length (years), (b) litter size (n), (c) log adult body mass (grams), (d) maximum longevity (years), and (e) incubation (days).
+The lines in each plot represent the best fit line from linear regression.\\label{fig:biological_correlates}',fig.height=6}
+
+load('../cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
+par(mfrow=c(3,2),oma=c(5,4,1,1),mar=c(5,3,0,0))
+#x_axis_name_list = c('generation length (years)','litter size (n)','adult body mass (grams)','maximum longevity (years)','egg mass (grams)','incubation (days)')
+
+  plot(LPI_pop_info$gen_len,LPI_pop_info$min_time_for_power,pch=as.numeric(as.factor(pop_info$class))+14,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2,cex=1.2)
+  abline(lm(LPI_pop_info$min_time_for_power~LPI_pop_info$gen_len),lwd=3)
+  mtext(text = 'generation length (years)',side = 1,line = 3,cex = 1.2)
+  mtext(text = "(a)",side = 3,line=-1.5,adj = 0.98,cex=1.2)
+
+  plot(LPI_pop_info$litter_or_clutch_size_n,LPI_pop_info$min_time_for_power,pch=as.numeric(as.factor(pop_info$class))+14,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2,cex=1.2)
+  abline(lm(LPI_pop_info$min_time_for_power~LPI_pop_info$litter_or_clutch_size_n),lwd=3)
+  mtext(text = 'litter size (n)',side = 1,line = 3,cex = 1.2)
+  mtext(text = "(b)",side = 3,line=-1.5,adj = 0.98,cex=1.2)
+
+    plot(log(LPI_pop_info$adult_body_mass_g),LPI_pop_info$min_time_for_power,pch=as.numeric(as.factor(pop_info$class))+14,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2,cex=1.2)
+  abline(lm(LPI_pop_info$min_time_for_power~LPI_pop_info$adult_body_mass_g),lwd=3)
+  mtext(text = 'log(adult mass (grams))',side = 1,line = 3,cex = 1.2)
+  mtext(text = "(c)",side = 3,line=-1.5,adj = 0.98,cex=1.2)
+
+  plot(LPI_pop_info$maximum_longevity_y,LPI_pop_info$min_time_for_power,pch=as.numeric(as.factor(pop_info$class))+14,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2,cex=1.2)
+  abline(lm(LPI_pop_info$min_time_for_power~LPI_pop_info$maximum_longevity_y),lwd=3)
+  mtext(text = 'maximum longevity (years)',side = 1,line = 3,cex = 1.2)
+  mtext(text = "(d)",side = 3,line=-1.5,adj = 0.98,cex=1.2)
+
+  plot(LPI_pop_info$incubation_d,LPI_pop_info$min_time_for_power,pch=as.numeric(as.factor(pop_info$class))+14,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2,cex=1.2)
+  abline(lm(LPI_pop_info$min_time_for_power~LPI_pop_info$incubation_d),lwd=3)
+  mtext(text = 'incubation (days)',side = 1,line = 3,cex = 1.2)
+  mtext(text = "(e)",side = 3,line=-1.5,adj = 0.98,cex=1.2)
+
+#  boxplot(LPI_pop_info$min_time_for_power~LPI_pop_info$trophic,pch=as.numeric(as.factor(pop_info$class))+14,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
+#  stripchart(LPI_pop_info$min_time_for_power~LPI_pop_info$trophic, vertical = TRUE,method = "jitter", add = TRUE, pch = as.numeric(as.factor(pop_info$class))+14, col = rgb(0.5,0.5,0.5,0.5))
+#  mtext(text = 'trophic position',side = 1,line = 3,cex = 1.2)
+#  mtext(text = "(f)",side = 3,line=-1.5,adj = 0.98,cex=1.2)
+
+plot(1, type="n", xlab="", ylab="", xaxt='n',yaxt='n',axes=FALSE)
+legend('center',legend=names(table(pop_info$class)),col = 'grey',pch = as.numeric(names(table(as.numeric(as.factor(pop_info$class))+14))),bty ="n",cex=1.4)
+
+mtext(text = 'minimum time required',side = 2,line = 2,cex = 1.4,outer = TRUE)
+
+#######
+
+biological_correlates_poisson_model <-glm(min_time_for_power~  litter_or_clutch_size_n  + gen_len + adult_body_mass_g + maximum_longevity_y  +incubation_d + trophic,family = 'poisson',data=LPI_pop_info)
+R2_biological_correlates_poisson_model <- 100*(biological_correlates_poisson_model$null.deviance - biological_correlates_poisson_model$deviance)/(biological_correlates_poisson_model$null.deviance)
+
+all_correlates_poisson_model <-glm(min_time_for_power~ abs(overall_trend) + coefficient_variation + autocorrelation +  litter_or_clutch_size_n  + gen_len + adult_body_mass_g + maximum_longevity_y  +incubation_d + trophic,family = 'poisson',data=LPI_pop_info)
+R2_all_correlates_poisson_model <- 100*(all_correlates_poisson_model$null.deviance - all_correlates_poisson_model$deviance)/(all_correlates_poisson_model$null.deviance)
+```
+
+Some life-history traits were significant predictors for the minimum time required (Fig.
+\ref{fig:biological_correlates}, Tables A2,A3).
+However, none of these life-history traits explained a large part of the variation in minimum time required.
+In a generalized linear model, all of the life-history traits described in figure \ref{fig:biological_correlates} account for only `r round(R2_biological_correlates_poisson_model,3)`% of the explained deviance in minimum time series length required.
+In addition, when accounting for trend strength, coefficient of variation, and autocorrelation, no life-history traits were significant predictors of the minimum time required (Table A3).
+
+### Evalulating the IUCN criteria
+
+```{r, echo=F,fig.cap='Minimum time required to achieve 0.8 statistical power versus the minimum time required under IUCN criteria A2 to classify a species as vulnerable.
+Each point represents a single population, all of which saw declines of 30% or greater over a 10 year period.\\label{fig:IUCN_analysis}'}
+
+declining_populations = subset(pop_info,pop_info$percent_change_10years>30 & pop_info$overall_trend<0)
+
+par(mfrow=c(1,1),oma=c(0,0,0,0),mar=c(5,5,0.5,0.5))
+
+shapes = as.numeric(as.factor(declining_populations$class))+14
+shapes[shapes==17]= 18
+plot(declining_populations$IUCN, declining_populations$min_time_for_power,
+       ylim=c(0,35),xlim=c(0,85),pch=shapes,
+       col=rgb(0.5,0.5,0.5,0.5),cex.lab=1.3,las=1,cex=1.5,
+       xlab = 'minimum time required under IUCN criteria',ylab='')
+mtext(text = 'minimum time required',side = 2,line = 3.8,outer = F,padj = 0.5,cex=1.3)
+mtext(text = 'for 0.8 statistical power',side = 2,line = 2.8,outer = F,padj = 0.5,cex=1.3)
+segments(0,0,100,100,lwd=2)
+
+#mtext(text = paste('n = ',nrow(declining_populations),sep=''),side = 3,line = -2,outer = F,at=80,cex=1.2)
+legend('bottomright',legend=names(table(declining_populations$class)),col = 'grey',pch = c(15,16,18),bty ="n",cex=1.2)
+
+whale <- readPNG('phylo-sperm-whale.png')
+rasterImage(whale,xleft=70,ybottom=12,xright=82,ytop=17)
+
+puffin <- readPNG('phylo-puffin.png')
+rasterImage(puffin,xleft=61,ybottom=30,xright=70,ytop=36)
+```
+
+The IUCN Red List Categories and Criteria suggest, under Criterion A2, a species qualifies as vulnerable if it has experienced a 30% decline over 10 years, or 3 generations (whichever is longer) [@IUCN2012].
+I examined a subset of populations with observed declines of 30% or greater over 10 years, qualifying all of them as vulnerable.
+This resulted in n = `r nrow(declining_populations)` populations of fish, birds, and a single mammal.
+I then compared the minimum time required to achieve 0.8 statistical power ($T_{min}$) to the minimum time required under the IUCN criteria (Fig.
+\ref{fig:IUCN_analysis}).
+For populations below the identity line in figure \ref{fig:IUCN_analysis}, IUCN criteria would require more sampling compared to estimates for $T_{min}$.
+Further, populations above the identity line are cases where the IUCN criteria would classify a population as vulnerable despite not having sampled enough years to achieve high statistical power (Fig.
+\ref{fig:IUCN_analysis}).
+The silhouettes on figure \ref{fig:IUCN_analysis} highlight that species with longer generation times typically have larger discrepancies between $T_{min}$ and the minimum time required for IUCN assessments (Fig.
+\ref{fig:IUCN_correlates}).
+
+### Sensitivity analysis
+Lastly, I tested model sensitivity by using generalized additive models (GAMs) instead of simple linear regression.
+Again, I examined the minimum time required to estimate long-term population trends (Fig.
 \ref{fig:gam_example}).
-We found that although we obtain a similar distribution of minimum times required for GAMs, the minimum time required for GAMs is on average shorter than for linear regression (Fig.
+I found that although I obtain a similar distribution of minimum times required for GAMs, the minimum time required for GAMs is on average shorter than for linear regression (Fig.
 \ref{fig:min_time_dist_gam}).
-
-```{r,echo=F,include=FALSE,eval=FALSE,fig.cap='Minimum time series required versus the generation length of each species\\label{fig:generation_length}'}
-par(mfrow=c(1,1))
-  plot(pop_info$gen_len,pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
-  mtext('generation time (years)',1,line=3,cex=1.2)
-  mtext('minimum time required',2,line=2.5,cex=1.2,outer = F)
-```
 
@@
 
-We explored two approaches to estimate the minimum time series length required to address a particular question of interest.
-We asked, what is the minimum time series length required to determine long-term population trends using linear regression?
+I explored two approaches to estimate the minimum time series length required to address a particular question of interest.
+I asked, what is the minimum time series length required to determine long-term population trends using linear regression?
 This is one of the simplest questions one could ask of a time series.
 The simulation-based approach has been suggested by others, especially in situations more complicated than that suited for classic power analysis [@Gerrodette1987;@Johnson2015;@Bolker2008].
-Our simulations support past work that longer time series are needed when the trend strength (i.e.
+My simulations support past work that longer time series are needed when the trend strength (i.e.
 rate of increase or decrease) is weak or when population variability is high [@Gerrodette1987].
-We also showed how the simulation model can be altered for a particular population (Fig.
+I also showed how the simulation model can be altered for a particular population (Fig.
 \ref{fig:shark_example}) or question (Figs.
@@
 
-Here, we focus on an empirical approach to estimate the minimum time series length required to assess changes in abundance over time.
-We examined `r nrow(pop_info)` population time series (all longer than 35 years).
-We then subsampled each to determine the minimum time required to achieve a desired significance level and power for linear regression.
+Here, I focus on an empirical approach to estimate the minimum time series length required to assess changes in abundance over time.
+I examined `r nrow(pop_info)` population time series (all longer than 35 years).
+I then subsampled each to determine the minimum time required to achieve a desired significance level and power for linear regression.
 Statistical power is important as it provides on information as to the necessary samples required to determine a significant trend [@Legg2006].
-We found that at least 15-20 years of continuous monitoring were typically necessary (Fig.
-\ref{fig:min_time_dist}).
-This time frame is in line with past work on a short-lived snail species [@Rueda-Cediel2015] and a long-lived whale species [@Gerber1999].
+I found that on average `r round(mean(pop_info$min_time_for_power),2)` years of continuous monitoring were typically necessary (Fig.
+\ref{fig:min_time_dist}b).
+However, the distribution of minimum time required was wide.
+This time-frame is in line with past work on a short-lived snail species [@Rueda-Cediel2015] and a long-lived whale species [@Gerber1999].
 @Hatch2003 used seabird monitoring data to estimate minimum sampling requirements.
@@
 
-In line with theoretical predictions [@Rhodes2011], we also found $T_{min}$ was strongly correlated with the trend strength, variability in population size, and temporal autocorrelation (Fig.
+In line with theoretical predictions [@Rhodes2011], I also found $T_{min}$ was strongly correlated with the trend strength, variability in population size, and temporal autocorrelation (Fig.
 \ref{fig:correlates}).
-Contrary to our prior expectations, we also found that $T_{min}$ did not correlate with any biological variables of interest (Fig.
+Contrary to my prior expectations, I also found that $T_{min}$ did not correlate with any life-history traits (Fig.
 \ref{fig:biological_correlates}).
-We initially hypothesized that species with longer lifespans or generation times may require a longer sampling period.
-Our result could have been a result of at least two factors.
-First, the data we used may not include a diverse enough set of species with different life history traits.
-Second, the question we poised, whether a population is increasing or decreasing, was specifically concerned with trends in population density over time.
+I initially hypothesized that species with longer lifespans or generation times may require a longer sampling period.
+This result could have been a result of at least two factors.
+First, the data I used may not include a diverse enough set of species with different life-history traits.
+Second, the question I poised, whether a population is increasing or decreasing, was specifically concerned with trends in population density over time.
 Therefore, life-history characteristics may be more important for other questions, like estimating species extinction risk [@Hutchings2012].
 For example, @Blanchard2007 used detailed simulations of spatially-distributed fisheries to compare survey designs.
-They found that statistical power depended on survey design, temperature preferences and the degree of population patchiness.
+They found that statistical power depended on survey design, temperature preferences, and the degree of population patchiness.
 
@@
 If spatial population dynamics were highly correlated, then it was better to sample more temporally, and vice versa.
-Our work supports this idea as populations with strong temporal autocorrelation needed less years of sampling (Fig.
+My work supports this idea as populations with strong temporal autocorrelation needed less years of sampling (Fig.
 \ref{fig:correlates}).
@@
 For example, @Seavy2007 suggest improving randomization, reducing bias, and increases detection probability when designing and evaluating monitoring programs.
-We agree that power analyses should not be the only consideration when designing monitoring schemes.
-However, unlike @Seavy2007, our results indicate that longer than 10 years is often needed to achieve high statistical power.
+I agree that power analyses should not be the only consideration when designing monitoring schemes.
+However, unlike @Seavy2007, my results indicate that longer than 10 years is often needed to achieve high statistical power.
+
+This paper also has practical implications for the IUCN Redlist criteria.
+IUCN criteria A2 suggests that species that have experienced 30% declines over 10 years (or three generations) should be listed as vulnerable [@IUCN2012].
+However, for the populations I examined, this criteria may be too simplistic (Fig.
+\ref{fig:IUCN_analysis}).
+For many populations, the IUCN criteria suggest more years than necessary are required to assess a population as vulnerable (points below diagonal line in Fig.
+\ref{fig:IUCN_analysis}).
+Conversely, for other populations the IUCN criteria suggest sampling times that are less than the minimum time required for statistical power.
+This suggests that the IUCN criteria are probably too simplistic as the minimum time required does not correlate with generation time (Fig.
+\ref{fig:biological_correlates}).
 
@@
 
-Our work has some limitations in determining the minimum time series length required.
+This paper has some limitations in determining the minimum time series length required.
 First, $T_{min}$ is particular to the specific question of interest.
-An additional complication is that for our empirical approach, the subsampling of the full time series allows for estimates of power, but the individual subsamples are clearly not independent of one another.
+An additional complication is that for the empirical approach, the subsampling of the full time series allows for estimates of power, but the individual subsamples are clearly not independent of one another.
 Further, estimates of $T_{min}$ depend on chosen values of $\alpha$ and $\beta$ (Fig.
 \ref{fig:min_time_vs_alpha_beta}).
-In an ideal setting, we would build a specific population model parameterized for each population of interest [@McCain2016].
+In an ideal setting, a specific population model would be parameterized for each population of interest [@McCain2016].
 Then, model simulations could be used to estimate the minimum time series required to address each specific question of interest.
-Clearly, this is not always practical, especially if conducting analyses for a wide array of species as we do here.
-In addition, our statistical models suggest that $T_{min}$ does not correlate with any biological variables of interest, at least for the question of linear regression (Fig.
+Clearly, this is not always practical, especially if conducting analyses for a wide array of species as I do here.
+In addition, the statistical models suggest that $T_{min}$ does not correlate with any life-history traits, at least for the question of linear regression (Fig.
 \ref{fig:biological_correlates}).
-Therefore, it is not possible to use these results to predict $T_{min}$ for another population, even if the population is of a species with a similar life-history to one in our database.
+Therefore, it is not possible to use these results to predict $T_{min}$ for another population, even if the population is of a species with a similar life-history to one in the database used.
 
@@
 
-We use a database of `r nrow(pop_info)` populations to determine the minimum time series length required to detect population trends.
+I used a database of `r nrow(pop_info)` populations to determine the minimum time series length required to detect population trends.
 This goes beyond previous work that either focused on theoretical investigations or a limited number of species.
-We show that to identify long-term changes in abundance, 15-20 years of continuous monitoring are often required (Fig.
+I show that to identify long-term changes in abundance, on average `r round(mean(pop_info$min_time_for_power),2)` years of continuous monitoring are often required (Fig.
 \ref{fig:min_time_dist}).
-In line with theoretical predictions [@Gerrodette1987], we also show that $T_{min}$ is strongly correlated with the long-term population trend (i.e.
+However, there is wide distribution of estimated minimum times.
+Therefore, it is probably not wise to use a simple threshold number of years in monitoring design.
+
+In line with theoretical predictions [@Gerrodette1987], I also show that $T_{min}$ is strongly correlated with the long-term population trend (i.e.
 rate of increase), variability in population size, and the temporal autocorrelation (Fig.
 \ref{fig:correlates}).
+Contrary to my initial hypotheses, minimum time required did not correlate with generation time or any other life-history traits (Fig.
+\ref{fig:biological_correlates}).
+This result argues against overly simplified measures of minimum sampling time based on generation length (Fig.
+\ref{fig:IUCN_analysis}).
 
-Our work implies that for many populations, time series less than 15-20 years are probably not reliable for detecting population trends.
+My work implies that for many populations, short time series are probably not reliable for detecting population trends.
 This result highlights the importance of long-term monitoring programs.
 From both a scientific and management perspective estimates of $T_{min}$ are important.
-If a time series is too short, we lack statistical power to reliably detect long-term population trends.
+If a time series is too short, we lack the statistical power to reliably detect long-term population trends.
 In addition, a time series that is too long may be a poor use of already limited funds [@Gerber1999].
-However, short time series data can not be used as an excuse for inaction.
-For example, @Thompson2000 demonstrate that long-term studies may be required to detect population trends for certain species.
+Further, more data is not always best in situations where management actions need to be taken [@Martin2012; @Martin2017].
 When a population trend is detected, it may be too late for management action.
 In these situations, the precautionary principle may be more appropriate [@Thompson2000].
-Future work should examine other species, with a wider range of life history characteristics.
+Future work should examine other species, with a wider range of life-history characteristics.
 In addition, similar approaches can be used to determine the minimum time series length required to address additional questions of interest.
@@
 
-In the supporting material, we provide an expanded methods sections, additional figures, minimum time calculations for determining exponential growth, simulations with a more complicated population model, minimum time calculations for determining long-term growth rates, and the use of generalized additive models to identify population trends.
+In the supporting material, I provide an expanded methods sections, additional figures, minimum time calculations for determining exponential growth, simulations with a more complicated population model, and the use of generalized additive models to identify population trends.
 All code and data can be found at https://github.com/erwhite1/time-series-project
@@
 ERW was partially supported by a National Science Foundation Graduate Fellowship.
-We would like to thank members of the Ecological Theory group at the University of California, Davis for their insight.
-We would also like to thank T.
+I would like to thank members of the Ecological Theory group at the University of California, Davis for their insight.
+I would also like to thank T.
 Dallas and E.
@@
 \centering
-\caption{Minimum time required versus (a) generation length (years), (b)
-litter size (n), (c) adult body mass (grams), (d) maximum longevity
-(years), (e) egg mass (grams), and (f) incubation (days). The lines in
-each plot represent the best fit line from linear
-regression.\label{fig:biological_correlates}}
-\end{figure}
-
-\begin{figure}[htbp]
-\centering
-\caption{(a) Minimum time required to estimate change in abundance for
-species class, (b) long-term trend (estimated slope coefficient) by
-species class, (c) inter-annual variability in population size by species
-class, and (d) temporal autocorrelation by species
-class.\label{fig:class}}
+\caption{(a) Minimum time required to estimate change in abundance by biological class, (b) long-term trend (estimated slope coefficient) by class, (c) coefficient of variation in population size by class, and (d) temporal autocorrelation by class.\label{fig:class}}
 \end{figure}
@@
 (\(1-\beta\)).\label{fig:min_time_vs_alpha_beta}}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\caption{The difference between minimum time estimates is the minimum time required to achieve 0.8 statistical power versus the minimum time required under IUCN criteria A2 to classify a species as vulnerable.
+Each point represents a single population, all of which saw declines of 30\% or greater over a 10 year period.
+(a) Difference between minimum time estimates versus the coefficient of variation in population size.
+(b) Difference between minimum time estimates versus the generation length in years.\label{fig:IUCN_correlates}}
 \end{figure}
@@
 
+<!--
 \begin{figure}[htbp]
@@
 \end{figure}
+-->
 
@@
 \end{figure}
+
+\vspace{1 cm}
+
+Table A1: Output of generalized linear model to examine time series characteristics as correlates of the minimum time required for determining long-term population trends.
+
+\vspace{1 cm}
+
+Table A2: Output of generalized linear model to examine life-history trait correlates of the minimum time required for determine long-term population trends.
+
+\vspace{1 cm}
+
+Table A3: Output of generalized linear model to examine both time series characteristics and life-history trait correlates of the minimum time required for determine long-term population trends.

### 2018-02-19 writing/manuscript_draft.tex
first update of short format version of the article

@@
 \usepackage{amssymb,amsmath}
+\usepackage{setspace}  
 \usepackage{ifxetex,ifluatex}
@@
 \usepackage[firstinits=true,style=authoryear-comp,natbib=true,doi=false,isbn=false,url=false,uniquename=false,uniquelist=false,sorting=none]{biblatex}
-\linenumbers
 
@@
 
-\textbf{Minimum time required to detect population trends: the need for long-term monitoring programs}
+Title: Minimum time required to detect population trends: the need for
+long-term monitoring programs \vspace{7 mm}
+
+Author: \textsc{Easton R. White$^{1,2}$} \vspace{3 mm}
+
+Address:\emph{
+    \\$^1$Center for Population Biology \\
+    University of California, Davis \\
+    2320 Storer Hall \\
+        University of California, Davis \\
+        One Shields Avenue Davis, CA, USA} \vspace{3 mm}
+
+\(^2\)Corresponding author:
+\href{mailto:eawhite@ucdavis.edu}{\nolinkurl{eawhite@ucdavis.edu}} , 1
+480 203 7931 \vspace{3 mm}
+
+Number of words (including 300 words per figure): 6,950 \vspace{1 mm}
+
+Number of figures: 5 \vspace{1 mm}
+
+Number of references: 34 \vspace{1 mm}
+
+Supplementary material: 9 figures and 3 tables \vspace{3 mm}
+
+Data and code for all the figures and tables can be found at
+(\url{https://github.com/erwhite1/time-series-project}).
+
+To be submitted to: \emph{Biological Conservation} \vspace{3 mm}
+
+Available as a preprint at: \url{https://peerj.com/preprints/3168/}
 \vspace{3 mm}
 
-Easton R. White\(^{*}\)
+Color does not need to be used for any figures in print \vspace{3 mm}
 
-\begin{footnotesize} Center for Population Biology, University of California, Davis, California 95616 USA
+\vspace{3 mm}
 
-*eawhite@ucdavis.edu
-\end{footnotesize}
-
-\vspace{5 mm}
+\begin{spacing}{2}
 
@@
 for a population. Therefore, managers must understand when population
-trend over time is actually meaningful. For example, the IUCN Red List
-Categories and Criteria suggest, under Criterion A2, a species qualifies
-as vulnerable if it has experienced a 30\% decline over 10 years, or 3
-generations (IUCN 2012). For both scientific and management questions,
-because sampling is typically expensive, we also do not want to sample
-for longer than is necessary. For example, Gerber, DeMaster, and Kareiva
-(1999) investigated the minimum time series required to estimate
-population growth of the endangered, but recovering, eastern North
-Pacific gray whale (\emph{Eschrichtius robustus}). They used a long-term
-census to retroactively determine the minimum time series required to
-assess threat status. They found that only 11 years were needed, eight
-years before the delisting decision was made. This highlights the
-importance of estimating the minimum time series required as an earlier
-decision would have saved time and money (Gerber, DeMaster, and Kareiva
-1999).
+trend over time is actually meaningful. For example, the International
+Union for Conservation of Nature (IUCN) Red List Categories and Criteria
+suggest, under Criterion A2, a species qualifies as vulnerable if it has
+experienced a 30\% decline over 10 years, or three generations (IUCN
+2012). For both scientific and management questions, because sampling is
+typically expensive, we also do not want to sample for longer than is
+necessary. For example, Gerber, DeMaster, and Kareiva (1999)
+investigated the minimum time series required to estimate population
+growth of the endangered, but recovering, eastern North Pacific gray
+whale (\emph{Eschrichtius robustus}). They used a long-term census to
+retroactively determine the minimum time series required to assess
+threat status. They found that only 11 years were needed, eight years
+before the delisting decision was made. This highlights the importance
+of estimating the minimum time series required as an earlier decision
+would have saved time and money (Gerber, DeMaster, and Kareiva 1999).
+Further, waiting too long to make a decision can imperil a species where
+management action could have been taken earlier (Martin et al. 2012;
+Martin et al. 2017).
 
@@
 found that for this short-lived organism, 15 years was adequate to
-assess long-term trends in abundance. Hatch (2003) examined seabird
-populations and determined the minimum time required to achieve high
-statistical power ranged between 11 and 69 years. However, these
-studies, and other past work, have typically focused either on
-theoretical aspects of monitoring design or focused on only a few
-species.
+assess long-term trends in abundance. However, these studies, and other
+past work, have typically focused either on theoretical aspects of
+monitoring design or focused on only a few species.
 
-We use both simulations and empirical time series to determine the
-minimum number of years required to address several questions. We
+I use both simulations and empirical time series to determine the
+minimum number of years required to address several questions. I
 estimate the minimum time series length required (\(T_{min}\)) to assess
-long-term changes in abundance via simple linear regression. First, we
-estimate \(T_{min}\) using a simulation approach. Then we examine 878
+long-term changes in abundance via simple linear regression. First, I
+estimate \(T_{min}\) using a simulation approach. Then I examine 822
 population time series to estimate \(T_{min}\). In the supplementary
-material, we determine \(T_{min}\) for related ideas: calculating
-long-term growth rates, using more complicated population models,
-varying statistical level and power, and the use of generalized additive
-models.
+material, I determine \(T_{min}\) for related ideas: using more
+complicated population models, varying statistical level and power, and
+the use of generalized additive models.
 
@@
 any experimental design too complicated for simple power analyses
-(Bolker 2008; P. C. Johnson et al. 2015). We only briefly discuss this
-approach as it has been described elsewhere. Essentially, we use a
-population model and repetitively simulate it for a number of years.
-This approach requires us to determine values for our model parameters
-(e.g.~population variability). As an example, we can take the following
-population model for population size \(N\) at time \(t\):
+(Bolker 2008; P. C. Johnson et al. 2015). I only briefly discuss this
+approach as it has been described elsewhere. Essentially, a population
+model is simulated repetitively for a number of years. This approach
+requires us to determine values for model parameters (e.g.~population
+variability). As an example, we can take the following population model
+for population size \(N\) at time \(t\):
 
@@
 increase). It is important to note that any population model could be
-substituted for equation \ref{sim_model}, as we do in the supplementary
+substituted for equation \ref{sim_model}, as in the supplementary
 material (Figs. \ref{fig:min_time_dist_log_pop},
@@
 of 0.8. Statistical power of 0.8 would indicate that, if there was a
-true trend in abundance, we would have a 0.8 probability of detecting
-the trend. We also tested the effect of varying both the significance
-level and statistical power (Fig. \ref{fig:min_time_vs_alpha_beta}).
+true trend in abundance, there would be a 0.8 probability of detecting
+the trend. Values of 0.05 for the significance level and 0.8 statistical
+power are historical and it is important to examine the effect of
+changing these values (Fig. \ref{fig:min_time_vs_alpha_beta}).
 
@@
 greater than 0.8 (the dotted line), that is the minimum time required
-(\(T_{min}\)).
+(\(T_{min}\)) to be confident in the detection of a long-term trend in
+abundance.
 
@@
 
-We use a database of 2444 population time series compiled in (Keith et
+I use a database of 2444 population time series compiled in (Keith et
 al. 2015); they compared the predictability of growth rates among
 populations. The data are originally from the Global Population Dynamics
-Database (NERC Centre for Population Biology 2010). We filtered out
-short time series (less than 35 years), and those with missing data, and
-were left with 878 time series. The data includes information on 478
-vertebrate species with a focus on mammals, birds, and fish. The data
-also includes information on generation length and survey
-specifications. For each time series, we also calculated variables of
-interest like variance in population size, long-term trend in abundance
-(slope coefficient from simple linear regression), and temporal
-autocorrelation. All analyses were conducted in R (R Core Team 2016).
+Database (NERC Centre for Population Biology 2010) and several other
+sources (Keith et al. 2015). I filtered out short time series (less than
+35 years), and those with missing data, leaving 822 time series. The
+data includes information on 477 vertebrate species with a focus on
+mammals, birds, and fish. The data also includes information on
+generation length and survey specifications. For each time series, I
+also calculate other variables of interest: coefficient of variance in
+population size, long-term trend in abundance (slope coefficient from
+simple linear regression), and temporal autocorrelation. All analyses
+were conducted in R (R Core Team 2016).
 
-For a subset of populations (\(n =\) 547), we had information on
-biological characteristics from another paper (Myhrvold et al. 2015),
+For a subset of populations (\(n =\) 547), there is information on
+life-history traits available from another paper (Myhrvold et al. 2015),
 including body size and generation length. All 547 populations were
-birds. We examine how the minimum time required is related to these
-biological characteristics (Fig. \ref{fig:biological_correlates}).
+birds. I examine how the minimum time required is related to these
+life-history traits (Fig. \ref{fig:biological_correlates}).
 
@@
 
-We assume that each time series in our database is long enough to
-include all necessary information (e.g.~variability) about the
-population. In other words, each time series is a representative sample.
-We first take all possible contiguous subsamples of each time series.
-For example, a time series of 35 years would have 34 possible contiguous
-subsamples of length 2, 34 possible contiguous subsamples of length 3,
-and continuing until 1 possible contiguous subsample of length 35
-(Gerber, DeMaster, and Kareiva 1999; Giron-Nava et al. 2017). Next, we
-run a linear regression for each subsampled time series. Then, we
-determine the proportion of subsamples of a particular length that have
-estimated slope coefficients which are statistically different from
-zero. We only look at the proportion of samples where the long-term, or
-``true'', time series also has a significant slope. This proportion is a
-measure of statistical power. Lastly, we determine which subsample
-length is required to achieve a certain threshold of statistical power
-(0.8, Cohen (1992)). The minimum subsampled length that met these
-criteria is the minimum time series length required (\(T_{min}\)).
+I assume that each time series is long enough to include all necessary
+information (e.g.~variability) about the population. In other words,
+each time series is a representative sample. I first take all possible
+contiguous subsamples of each time series. For example, a time series of
+35 years would have 34 possible contiguous subsamples of length 2, 34
+possible contiguous subsamples of length 3, and continuing until 1
+possible contiguous subsample of length 35 (Gerber, DeMaster, and
+Kareiva 1999; Giron-Nava et al. 2017). Next, I run a linear regression
+for each subsampled time series. Then, I determine the proportion of
+subsamples of a particular length that have estimated slope coefficients
+which are statistically different from zero. I only look at the
+proportion of samples where the long-term, or ``true'', time series also
+has a significant slope. This proportion is a measure of statistical
+power. Lastly, I determine which subsample length is required to achieve
+a certain threshold of statistical power (0.8, Cohen (1992)). The
+minimum subsampled length that met these criteria is the minimum time
+series length required (\(T_{min}\)).
 
-In the supplementary material, we show how the same approach described
-here can be used to determine the minimum time required to estimate a
-population's geometric growth rate (Figs.
-\ref{fig:growth_rate},\ref{fig:min_time_growth_dist}). We also determine
-the minimum time required to estimate long-term trends according to
+In the supplementary material, I show how the same approach described
+here for more complicated population models. I also determine the
+minimum time required to estimate long-term trends according to
 generalized additive models, instead of the simple linear models used
@@
 
-We determined the minimum time series length (\(T_{min}\)) required to
+I determined the minimum time series length (\(T_{min}\)) required to
 address a particular question of interest. What is the minimum time
@@
 long-term population trend? Here, the minimum time series length
-required had high enough statistical power (\(1-\beta\) greater than
-0.8) for a set significance level (\(\alpha\)) of 0.05. It is also
-possible to alter \(1-\beta\) and \(\alpha\). Predictably, as we
-increased our level of power or decreased \(\alpha\), \(T_{min}\)
-increased (Fig. \ref{fig:min_time_vs_alpha_beta}). We estimated
-\(T_{min}\) using two approaches. We briefly describe results from the
-simulation approach and then discuss our empirical approach.
+required had high enough statistical power (greater than 0.8) for a set
+significance level (\(\alpha\)) of 0.05. It is also possible to alter
+statistical power and \(\alpha\). Predictably, with increased
+statistical power or decreased \(\alpha\), \(T_{min}\) increased (Fig.
+\ref{fig:min_time_vs_alpha_beta}). I then estimated \(T_{min}\) using
+two approaches. I briefly describe results from the simulation approach
+and then discuss the empirical approach.
 
@@
 
-We constructed a general population model where the trend strength
-(i.e.~slope coefficient) over time could be a model parameter. We then
-ran simulated time series of different lengths. From these simulations
-we determined the minimum time series length required to achieve a
-certain level of statistical power. In line with past work (Gerrodette
-1987), we found the \(T_{min}\) increases (i.e.~more time is required)
-with decreases in trend strength and with increases in population
-variability (Figs. \ref{fig:theoretical_approach}c,d).
+I constructed a general population model where the trend strength
+(i.e.~slope coefficient) over time could be a model parameter. I then
+simulated time series of different lengths. From these simulations I
+determined the minimum time series length required to achieve a certain
+level of statistical power. In line with past work (Gerrodette 1987), I
+found the \(T_{min}\) increases (i.e.~more time is required) with
+decreases in trend strength and with increases in population variability
+(Figs. \ref{fig:theoretical_approach}c,d).
 
-We chose a simple model, but any other population model could be used
+I chose a simple model, but any other population model could be used
 (see example in Fig. \ref{fig:min_time_dist_log_pop}). Ideally, the
 specific model choice should be tailored to the population of interest.
-We explored how the simulation approach could be applied to more
+I explored how the simulation approach could be applied to more
 biologically-realistic population models (Fig. \ref{fig:shark_example}).
-More specifically, we determined the minimum time required to estimate
+Specifically, I determined the minimum time required to estimate
 long-term population trends using a stochastic, age-structured model of
 lemon shark population dynamics in the Bahamas (White, Nagy, and Gruber
-2014). We found that over 27 years of continuous monitoring were needed
+2014). I found that over 27 years of continuous monitoring were needed
 in this particular scenario (Fig. \ref{fig:shark_example}). Similar to
@@
 \includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-4-1.pdf}
-\caption{Distributions of the minimum time required for populations from
-four different biological classes. The minimum time required calculation
-corresponds to a significance level of 0.05 and statistical power of
-0.8.\label{fig:min_time_dist}}
+\caption{(a) Distributions of the minimum time required for populations
+from four different biological classes. (b) Distribution of minimum time
+required for all populations regardless of biological class. The minimum
+time required calculation corresponds to a significance level of 0.05
+and statistical power of 0.8.\label{fig:min_time_dist}}
 \end{figure}
@@
 correlated with (a) trend strength (absolute value of slope coefficient
-estimated from linear regression), (b) population variance (inter-annual
-variability in population size), and (c) temporal lag-1
+estimated from linear regression), (b) coefficient of variation in
+interannual population size, and (c) temporal lag-1
 autocorrelation.\label{fig:correlates}}
@@
 
-We examined a database of 878 separate population time series
-representing 478 species. This database consists of vertebrate species
-with a variety of life history characteristics (Fig.
-\ref{fig:biological_correlates}). We limited our analyses to populations
-with at least 35 years of continuous sampling. We then examined the
-minimum time required to estimate long-term trends via linear
-regression.
+I examined a database of 822 separate population time series
+representing 477 species. This database consisted of vertebrate species
+with a variety of life-history characteristics (Fig.
+\ref{fig:biological_correlates}). I limited analyses to populations with
+at least 35 years of continuous sampling. I then examined the minimum
+time required to estimate long-term trends via linear regression.
 
-Across all the populations we examined, we found an average minimum time
-series length required (\(T_{min}\)) of 16.5 (\(\sigma\)=8.4), with a
-wide distribution (Fig. \ref{fig:min_time_dist}). Estimates of
-\(T_{min}\) varied between biological class (Fig.
-\ref{fig:min_time_dist}). Ray-finned fish (class Actinopterygii)
-typically had estimates of \(T_{min}\) over 20 years. Birds (class Aves)
-had a much wider distribution of \(T_{min}\), but usually required less
-years of sampling. Differences between these classes were explained by
-differences in inter-annual variability in population size (Fig.
-\ref{fig:class}). We also examined a subset of populations where
-life-history characteristics were known. None of these explanatory
-variables were predictive of the minimum time series length required
-(Fig. \ref{fig:biological_correlates}).
+Across all the populations, I found an average minimum time series
+length required (\(T_{min}\)) of 15.9 (SD=8.3), with a wide distribution
+(Fig. \ref{fig:min_time_dist}b). Estimates of \(T_{min}\) varied between
+biological class (Fig. \ref{fig:min_time_dist}a). Ray-finned fish (class
+Actinopterygii) typically had estimates of \(T_{min}\) over 20 years.
+Birds (class Aves) had a much wider distribution of \(T_{min}\), but
+usually required less years of sampling. Differences between these
+classes can be explained by differences in variability in population
+size and strength of trends in abundance (Fig. \ref{fig:class}).
+
+\subsubsection{Corrrelates for minimum time
+required}\label{corrrelates-for-minimum-time-required}
 
 The minimum time series length required was strongly correlated with
-trend strength, population variance, and autocorrelation (Fig.
-\ref{fig:correlates}). This is in line based on our simulations and
-those of others (Rhodes and Jonzen 2011). Using a generalized linear
-model, with a Poisson error structure, all three of these explanatory
-variables were significant and had large effect sizes (see Table A1).
-Combined, trend strength, population variance, and autocorrelation
-account for about 72.6\% of the explained deviance (Zuur et al. 2009) in
-minimum time series length required.
+trend strength (i.e.~estimated slope coefficient from linear
+regression), coefficient of variation in population size, and
+autocorrelation in population size (Fig. \ref{fig:correlates}). This is
+in line based on simulations here and those of others (Rhodes and Jonzen
+2011). Using a generalized linear model, with a Poisson error structure,
+all three of these explanatory variables were significant and had large
+effect sizes (see Table A1). Combined, trend strength, coefficient of
+variation in population, and autocorrelation account for 75.1\% of the
+explained deviance (Zuur et al. 2009) in minimum time series length
+required.
 
-Lastly, we tested model sensitivity by using generalized additive models
-(GAMs) instead of simple linear regression. Again, we examined the
+For a subset of the populations I combined time series data with a data
+on life-history characteristics of amniotes (Myhrvold et al. 2015).
+There was life-history information available for 547 populations
+representing 315 different species, all of which were birds (Aves
+class).
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-7-1.pdf}
+\caption{Mimimum time required versus (a) generation length (years), (b)
+litter size (n), (c) log adult body mass (grams), (d) maximum longevity
+(years), and (e) incubation (days). The lines in each plot represent the
+best fit line from linear regression.\label{fig:biological_correlates}}
+\end{figure}
+
+Some life-history traits were significant predictors for the minimum
+time required (Fig. \ref{fig:biological_correlates}, Tables A2,A3).
+However, none of these life-history traits explained a large part of the
+variation in minimum time required. In a generalized linear model, all
+of the life-history traits described in figure
+\ref{fig:biological_correlates} account for only 5.99\% of the explained
+deviance in minimum time series length required. In addition, when
+accounting for trend strength, coefficient of variation, and
+autocorrelation, no life-history traits were significant predictors of
+the minimum time required (Table A3).
+
+\subsubsection{Evalulating the IUCN
+criteria}\label{evalulating-the-iucn-criteria}
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-8-1.pdf}
+\caption{Minimum time required to achieve 0.8 statistical power versus
+the minimum time required under IUCN criteria A2 to classify a species
+as vulnerable. Each point represents a single population, all of which
+saw declines of 30\% or greater over a 10 year
+period.\label{fig:IUCN_analysis}}
+\end{figure}
+
+The IUCN Red List Categories and Criteria suggest, under Criterion A2, a
+species qualifies as vulnerable if it has experienced a 30\% decline
+over 10 years, or 3 generations (whichever is longer) (IUCN 2012). I
+examined a subset of populations with observed declines of 30\% or
+greater over 10 years, qualifying all of them as vulnerable. This
+resulted in n = 162 populations of fish, birds, and a single mammal. I
+then compared the minimum time required to achieve 0.8 statistical power
+(\(T_{min}\)) to the minimum time required under the IUCN criteria (Fig.
+\ref{fig:IUCN_analysis}). For populations below the identity line in
+figure \ref{fig:IUCN_analysis}, IUCN criteria would require more
+sampling compared to estimates for \(T_{min}\). Further, populations
+above the identity line are cases where the IUCN criteria would classify
+a population as vulnerable despite not having sampled enough years to
+achieve high statistical power (Fig. \ref{fig:IUCN_analysis}). The
+silhouettes on figure \ref{fig:IUCN_analysis} highlight that species
+with longer generation times typically have larger discrepancies between
+\(T_{min}\) and the minimum time required for IUCN assessments (Fig.
+\ref{fig:IUCN_correlates}).
+
+\subsubsection{Sensitivity analysis}\label{sensitivity-analysis}
+
+Lastly, I tested model sensitivity by using generalized additive models
+(GAMs) instead of simple linear regression. Again, I examined the
 minimum time required to estimate long-term population trends (Fig.
-\ref{fig:gam_example}). We found that although we obtain a similar
+\ref{fig:gam_example}). I found that although I obtain a similar
 distribution of minimum times required for GAMs, the minimum time
@@
 
-We explored two approaches to estimate the minimum time series length
-required to address a particular question of interest. We asked, what is
+I explored two approaches to estimate the minimum time series length
+required to address a particular question of interest. I asked, what is
 the minimum time series length required to determine long-term
@@
 than that suited for classic power analysis (Gerrodette 1987; P. C.
-Johnson et al. 2015; Bolker 2008). Our simulations support past work
-that longer time series are needed when the trend strength (i.e.~rate of
+Johnson et al. 2015; Bolker 2008). My simulations support past work that
+longer time series are needed when the trend strength (i.e.~rate of
 increase or decrease) is weak or when population variability is high
-(Gerrodette 1987). We also showed how the simulation model can be
-altered for a particular population (Fig. \ref{fig:shark_example}) or
-question (Figs. \ref{fig:min_time_dist_log_pop},\ref{fig:gam_example}).
+(Gerrodette 1987). I also showed how the simulation model can be altered
+for a particular population (Fig. \ref{fig:shark_example}) or question
+(Figs. \ref{fig:min_time_dist_log_pop},\ref{fig:gam_example}).
 
-Here, we focus on an empirical approach to estimate the minimum time
-series length required to assess changes in abundance over time. We
-examined 878 population time series (all longer than 35 years). We then
+Here, I focus on an empirical approach to estimate the minimum time
+series length required to assess changes in abundance over time. I
+examined 822 population time series (all longer than 35 years). I then
 subsampled each to determine the minimum time required to achieve a
@@
 samples required to determine a significant trend (Legg and Nagy 2006).
-We found that at least 15-20 years of continuous monitoring were
-typically necessary (Fig. \ref{fig:min_time_dist}). This time frame is
-in line with past work on a short-lived snail species (Rueda-Cediel et
-al. 2015) and a long-lived whale species (Gerber, DeMaster, and Kareiva
+I found that on average 15.91 years of continuous monitoring were
+typically necessary (Fig. \ref{fig:min_time_dist}b). However, the
+distribution of minimum time required was wide. This time-frame is in
+line with past work on a short-lived snail species (Rueda-Cediel et al.
+2015) and a long-lived whale species (Gerber, DeMaster, and Kareiva
 1999). Hatch (2003) used seabird monitoring data to estimate minimum
@@
 
-In line with theoretical predictions (Rhodes and Jonzen 2011), we also
+In line with theoretical predictions (Rhodes and Jonzen 2011), I also
 found \(T_{min}\) was strongly correlated with the trend strength,
 variability in population size, and temporal autocorrelation (Fig.
-\ref{fig:correlates}). Contrary to our prior expectations, we also found
-that \(T_{min}\) did not correlate with any biological variables of
-interest (Fig. \ref{fig:biological_correlates}). We initially
-hypothesized that species with longer lifespans or generation times may
-require a longer sampling period. Our result could have been a result of
-at least two factors. First, the data we used may not include a diverse
-enough set of species with different life history traits. Second, the
-question we poised, whether a population is increasing or decreasing,
-was specifically concerned with trends in population density over time.
-Therefore, life-history characteristics may be more important for other
-questions, like estimating species extinction risk (J. A. Hutchings et
-al. 2012). For example, Blanchard, Maxwell, and Jennings (2007) used
-detailed simulations of spatially-distributed fisheries to compare
-survey designs. They found that statistical power depended on survey
-design, temperature preferences and the degree of population patchiness.
+\ref{fig:correlates}). Contrary to my prior expectations, I also found
+that \(T_{min}\) did not correlate with any life-history traits (Fig.
+\ref{fig:biological_correlates}). I initially hypothesized that species
+with longer lifespans or generation times may require a longer sampling
+period. This result could have been a result of at least two factors.
+First, the data I used may not include a diverse enough set of species
+with different life-history traits. Second, the question I poised,
+whether a population is increasing or decreasing, was specifically
+concerned with trends in population density over time. Therefore,
+life-history characteristics may be more important for other questions,
+like estimating species extinction risk (J. A. Hutchings et al. 2012).
+For example, Blanchard, Maxwell, and Jennings (2007) used detailed
+simulations of spatially-distributed fisheries to compare survey
+designs. They found that statistical power depended on survey design,
+temperature preferences, and the degree of population patchiness.
 
@@
 spatial population dynamics were highly correlated, then it was better
-to sample more temporally, and vice versa. Our work supports this idea
-as populations with strong temporal autocorrelation needed less years of
+to sample more temporally, and vice versa. My work supports this idea as
+populations with strong temporal autocorrelation needed less years of
 sampling (Fig. \ref{fig:correlates}). Morrison and Hik (2008) also
@@
 improving randomization, reducing bias, and increases detection
-probability when designing and evaluating monitoring programs. We agree
+probability when designing and evaluating monitoring programs. I agree
 that power analyses should not be the only consideration when designing
-monitoring schemes. However, unlike Seavy and Reynolds (2007), our
+monitoring schemes. However, unlike Seavy and Reynolds (2007), my
 results indicate that longer than 10 years is often needed to achieve
 high statistical power.
+
+This paper also has practical implications for the IUCN Redlist
+criteria. IUCN criteria A2 suggests that species that have experienced
+30\% declines over 10 years (or three generations) should be listed as
+vulnerable (IUCN 2012). However, for the populations I examined, this
+criteria may be too simplistic (Fig. \ref{fig:IUCN_analysis}). For many
+populations, the IUCN criteria suggest more years than necessary are
+required to assess a population as vulnerable (points below diagonal
+line in Fig. \ref{fig:IUCN_analysis}). Conversely, for other populations
+the IUCN criteria suggest sampling times that are less than the minimum
+time required for statistical power. This suggests that the IUCN
+criteria are probably too simplistic as the minimum time required does
+not correlate with generation time (Fig.
+\ref{fig:biological_correlates}).
 
@@
 
-Our work has some limitations in determining the minimum time series
+This paper has some limitations in determining the minimum time series
 length required. First, \(T_{min}\) is particular to the specific
-question of interest. An additional complication is that for our
+question of interest. An additional complication is that for the
 empirical approach, the subsampling of the full time series allows for
@@
 chosen values of \(\alpha\) and \(\beta\) (Fig.
-\ref{fig:min_time_vs_alpha_beta}). In an ideal setting, we would build a
-specific population model parameterized for each population of interest
+\ref{fig:min_time_vs_alpha_beta}). In an ideal setting, a specific
+population model would be parameterized for each population of interest
 (McCain, Szewczyk, and Knight 2016). Then, model simulations could be
@@
 specific question of interest. Clearly, this is not always practical,
-especially if conducting analyses for a wide array of species as we do
-here. In addition, our statistical models suggest that \(T_{min}\) does
-not correlate with any biological variables of interest, at least for
-the question of linear regression (Fig.
-\ref{fig:biological_correlates}). Therefore, it is not possible to use
-these results to predict \(T_{min}\) for another population, even if the
-population is of a species with a similar life-history to one in our
-database.
+especially if conducting analyses for a wide array of species as I do
+here. In addition, the statistical models suggest that \(T_{min}\) does
+not correlate with any life-history traits, at least for the question of
+linear regression (Fig. \ref{fig:biological_correlates}). Therefore, it
+is not possible to use these results to predict \(T_{min}\) for another
+population, even if the population is of a species with a similar
+life-history to one in the database used.
 
@@
 
-We use a database of 878 populations to determine the minimum time
+I used a database of 822 populations to determine the minimum time
 series length required to detect population trends. This goes beyond
 previous work that either focused on theoretical investigations or a
-limited number of species. We show that to identify long-term changes in
-abundance, 15-20 years of continuous monitoring are often required (Fig.
-\ref{fig:min_time_dist}). In line with theoretical predictions
-(Gerrodette 1987), we also show that \(T_{min}\) is strongly correlated
-with the long-term population trend (i.e.~rate of increase), variability
-in population size, and the temporal autocorrelation (Fig.
-\ref{fig:correlates}).
+limited number of species. I show that to identify long-term changes in
+abundance, on average 15.91 years of continuous monitoring are often
+required (Fig. \ref{fig:min_time_dist}). However, there is wide
+distribution of estimated minimum times. Therefore, it is probably not
+wise to use a simple threshold number of years in monitoring design.
 
-Our work implies that for many populations, time series less than 15-20
-years are probably not reliable for detecting population trends. This
-result highlights the importance of long-term monitoring programs. From
-both a scientific and management perspective estimates of \(T_{min}\)
-are important. If a time series is too short, we lack statistical power
+In line with theoretical predictions (Gerrodette 1987), I also show that
+\(T_{min}\) is strongly correlated with the long-term population trend
+(i.e.~rate of increase), variability in population size, and the
+temporal autocorrelation (Fig. \ref{fig:correlates}). Contrary to my
+initial hypotheses, minimum time required did not correlate with
+generation time or any other life-history traits (Fig.
+\ref{fig:biological_correlates}). This result argues against overly
+simplified measures of minimum sampling time based on generation length
+(Fig. \ref{fig:IUCN_analysis}).
+
+My work implies that for many populations, short time series are
+probably not reliable for detecting population trends. This result
+highlights the importance of long-term monitoring programs. From both a
+scientific and management perspective estimates of \(T_{min}\) are
+important. If a time series is too short, we lack the statistical power
 to reliably detect long-term population trends. In addition, a time
 series that is too long may be a poor use of already limited funds
-(Gerber, DeMaster, and Kareiva 1999). However, short time series data
-can not be used as an excuse for inaction. For example, Thompson et al.
-(2000) demonstrate that long-term studies may be required to detect
-population trends for certain species. When a population trend is
-detected, it may be too late for management action. In these situations,
-the precautionary principle may be more appropriate (Thompson et al.
-2000). Future work should examine other species, with a wider range of
-life history characteristics. In addition, similar approaches can be
+(Gerber, DeMaster, and Kareiva 1999). Further, more data is not always
+best in situations where management actions need to be taken (Martin et
+al. 2012; Martin et al. 2017). When a population trend is detected, it
+may be too late for management action. In these situations, the
+precautionary principle may be more appropriate (Thompson et al. 2000).
+Future work should examine other species, with a wider range of
+life-history characteristics. In addition, similar approaches can be
 used to determine the minimum time series length required to address
@@
 
-In the supporting material, we provide an expanded methods sections,
+In the supporting material, I provide an expanded methods sections,
 additional figures, minimum time calculations for determining
 exponential growth, simulations with a more complicated population
-model, minimum time calculations for determining long-term growth rates,
-and the use of generalized additive models to identify population
+model, and the use of generalized additive models to identify population
 trends. All code and data can be found at
@@
 ERW was partially supported by a National Science Foundation Graduate
-Fellowship. We would like to thank members of the Ecological Theory
-group at the University of California, Davis for their insight. We would
-also like to thank T. Dallas and E. Malcolm-White for their helpful
-comments.
+Fellowship. I would like to thank members of the Ecological Theory group
+at the University of California, Davis for their insight. I would also
+like to thank T. Dallas and E. Malcolm-White for their helpful comments.
 
@@
 
+\hypertarget{ref-Martin2017}{}
+Martin, Tara G., Abbey E. Camaclang, Hugh P. Possingham, Lynn A.
+Maguire, and Iadine Chadès. 2017. ``Timing of Protection of Critical
+Habitat Matters.'' \emph{Conservation Letters} 10 (3): 308--16.
+doi:\href{https://doi.org/10.1111/conl.12266}{10.1111/conl.12266}.
+
+\hypertarget{ref-Martin2012}{}
+Martin, Tara G., Simon Nally, Andrew A. Burbidge, Sophie Arnall, Stephen
+T. Garnett, Matt W. Hayward, Linda F. Lumsden, Peter Menkhorst, Eve
+Mcdonald-Madden, and Hugh P. Possingham. 2012. ``Acting fast helps avoid
+extinction.'' \emph{Conservation Letters} 5 (4): 274--80.
+doi:\href{https://doi.org/10.1111/j.1755-263X.2012.00239.x}{10.1111/j.1755-263X.2012.00239.x}.
+
 \hypertarget{ref-McCain2016}{}
@@
 
-\clearpage
-
-\section{Supplementary material}\label{supplementary-material}
-
-\setcounter{figure}{0} \renewcommand{\thefigure}{A\arabic{figure}}
-
-\begin{figure}[htbp]
-\centering
-\caption{(a) Population size of Bigeye tuna (\emph{Thunnus obesus}) over
-time. The line is the best fit line from linear regression. (b)
-Statistical power for different subsets of the time series in panel
-a.\label{fig:empirical_approach_example}}
-\end{figure}
-
-\begin{figure}[htbp]
-\centering
-\caption{Output of generalized linear model with a Poisson error
-structure for predicting the minimum time required with explanatory
-variables of the absolute value of the slope coefficient (or trend
-strength), temporal autocorrelation, and variability in population
-size.\label{fig:poisson_model}}
-\end{figure}
-
-\begin{figure}[htbp]
-\centering
-\caption{Minimum time required versus (a) generation length (years), (b)
-litter size (n), (c) adult body mass (grams), (d) maximum longevity
-(years), (e) egg mass (grams), and (f) incubation (days). The lines in
-each plot represent the best fit line from linear
-regression.\label{fig:biological_correlates}}
-\end{figure}
-
-\begin{figure}[htbp]
-\centering
-\caption{(a) Minimum time required to estimate change in abundance for
-species class, (b) long-term trend (estimated slope coefficient) by
-species class, (c) inter-annual variability in population size by species
-class, and (d) temporal autocorrelation by species
-class.\label{fig:class}}
-\end{figure}
-
-\begin{figure}[htbp]
-\centering
-\caption{Minimum time required to assess long-term trends in abundance
-for values of statistical significance (\(\alpha\)) and power
-(\(1-\beta\)).\label{fig:min_time_vs_alpha_beta}}
-\end{figure}
-
-\begin{figure}[htbp]
-\centering
-\caption{Distribution of the minimum time required in order to detect a
-significant trend (at the 0.05 level) in log(abundance) given power of
-0.8.\label{fig:min_time_dist_log_pop}}
-\end{figure}
-
-\begin{figure}[htbp]
-\centering
-\caption{Statistical power for different length of time series
-simulations for a lemon shark population in Bimini,
-Bahamas.\label{fig:shark_example}}
-\end{figure}
-
-\begin{figure}[htbp]
-\centering
-\caption{Example of calculating minimum time required for growth rate
-estimation. (a) European herring gull (\emph{Larus argentatus}) scaled
-density over time, (b) mean and standard deviation of growth rate for
-subsamples of entire time series, and (c) the percent error between mean
-estimated growth rate and the true long-term growth rate. The vertical
-bar denotes the minimum time required to estimate growth rate within
-20\% error.\label{fig:growth_rate}}
-\end{figure}
-
-\begin{figure}[htbp]
-\centering
-\caption{Histogram of the minimum time required in order to estimate the
-long-term growth rate within 20\%
-error.\label{fig:min_time_growth_dist}}
-\end{figure}
-
-\begin{figure}[htbp]
-\centering
-\caption{(a) Time series for Bigeye tuna (\emph{Thunnus obesus}) with
-corresponding fitted GAM model in red and (b) statistical power as a
-function of the number of years sampled. The horizontal line at 0.8
-indicates the minimum threshold for statistical power and the vertical
-line denotes the minimum time required to achieve 0.8 statistical
-power.\label{fig:gam_example}}
-\end{figure}
-
-\begin{figure}[htbp]
-\centering
-\caption{Distribution of the minimum time required in order to detect a
-significant trend (at the 0.05 level) in abundance according to a GAM
-model given statistical power of 0.8. The smoothing parameter was set to
-3 for each population.\label{fig:min_time_dist_gam}}
-\end{figure}
-
+\end{spacing}
 \end{document}

### 2018-02-19 writing/manuscript_draft_short_version.Rmd
first update of short format version of the article

--- 2018-02-19 writing/manuscript_draft.Rmd
@@
 geometry: margin=1in
-header-includes: \usepackage{float} \usepackage{lineno} \usepackage[firstinits=true,style=authoryear-comp,natbib=true,doi=false,isbn=false,url=false,uniquename=false,uniquelist=false,sorting=none]{biblatex}
-
+header-includes: \usepackage{float} \usepackage{lineno} \usepackage{setspace}\doublespacing \usepackage[firstinits=true,style=authoryear-comp,natbib=true,doi=false,isbn=false,url=false,uniquename=false,uniquelist=false,sorting=none]{biblatex}
 output:
@@
   word_document: default
+editor_options: 
+  chunk_output_type: console
 ---
+Running head: Power to detect population trends [33 of 40 characters]
 
@@
 $^2$Corresponding author: eawhite@ucdavis.edu , 1 480 203 7931
-\vspace{3 mm}
-
-Number of words (including 300 words per figure):  6,950
 \vspace{1 mm}
 
-Number of figures: 5 
+To be submitted to: \emph{Ecology} as a Statistical Report
 \vspace{1 mm}
 
-Number of references: 34
+Number of pages: 20  
+\vspace{1 mm}
+
+Number of figures: 4
+\vspace{1 mm}
+
+Number of references: 33
 \vspace{1 mm}
@@
 Supplementary material: 9 figures and 3 tables
-\vspace{3 mm}
+\vspace{2 mm}
 
 Data and code for all the figures and tables can be found at (https://github.com/erwhite1/time-series-project).
-
-To be submitted to: \emph{Biological Conservation}
-\vspace{3 mm}
+\vspace{2 mm}
 
 Available as a preprint at: https://peerj.com/preprints/3168/
-\vspace{3 mm}
+\vspace{2 mm}
 
-Color does not need to be used for any figures in print
-\vspace{3 mm}
-
-\vspace{3 mm}
+\pagebreak
 
@@
 I determine the minimum time series length required to detect significant increases or decreases in population abundance.
-To address this question, I use simulation methods and examine 822 populations of vertebrate species.
-Here I show that on average 15.9 years of continuous monitoring are required in order to achieve a high level of statistical power.
+To address this question, I examine 822 populations of vertebrate species.
+I show that on average 15.9 years of continuous monitoring are required in order to achieve a high level of statistical power.
 However, there is a wide distribution around this average, casting doubt on simple rules of thumb.
-For both simulations and the time series data, the minimum time required depends on trend strength, population variability, and temporal autocorrelation.
+The minimum time required depends on trend strength, population variability, and temporal autocorrelation.
 However, there were no life-history traits (e.g.
@@
 Long-term data are necessary to both understand population dynamics and to assess species extinction risk.
-Even though many time series may now be considered "long-term" (e.g.
-continuous plankton recorder, @Giron-Nava2017), most are still short.
-Time series are typically short for a variety of reasons [@Field2007].
-They are often coupled with an experiment, which may only last a couple of years.
-In addition, short funding cycles make it difficult to examine populations over longer periods of time [@Hughes2017].
+Some time series may now be considered "long-term" (e.g.
+continuous plankton recorder, @Giron-Nava2017), but most are still short.
+Time series are typically short due to short funding cycles and typical experimental time-frames [@Field2007;@Hughes2017].
 
@@
 This question has important implications for both research and management [@Nichols2006].
-Scientists need to know the time series length required to address a specific question.
 A short time series may lead to wrong conclusions given large natural year-to-year variability [@McCain2016].
 Managers need to know when action is needed for a population.
-Therefore, managers must understand when population trend over time is actually meaningful.
+Therefore, managers must understand when a population trend over time is actually meaningful.
 For example, the International Union for Conservation of Nature (IUCN) Red List Categories and Criteria suggest, under Criterion A2, a species qualifies as vulnerable if it has experienced a 30% decline over 10 years, or three generations [@IUCN2012].
-For both scientific and management questions, because sampling is typically expensive, we also do not want to sample for longer than is necessary.
+Sampling is typically expensive, therefore, we also do not want to sample for longer than is necessary.
 For example, @Gerber1999 investigated the minimum time series required to estimate population growth of the endangered, but recovering, eastern North Pacific gray whale (\emph{Eschrichtius robustus}).
@@
 This highlights the importance of estimating the minimum time series required as an earlier decision would have saved time and money [@Gerber1999].
-Further, waiting too long to make a decision can imperil a species where management action could have been taken earlier [@Martin2012; @Martin2017].
+Further, waiting too long to decide an action can imperil a species where management action could have been taken earlier [@Martin2012; @Martin2017].
+
+Past work has investigated questions related to the minimum time series required to estimate trends in population size over time [@Wagner2009; @Giron-Nava2017].
+For example, @Rhodes2011 examined the optimal allocation of effort between spatial and temporal replicates.
+Using simple populations models, they found that the allocation of effort depends on environmental variation, spatial and temporal autocorrelation, and observer error.
+@Rueda-Cediel2015 also used a modeling approach, but parameterized a model specific for a threatened snail, \emph{Tasmaphena lamproides}.
+They found that for this short-lived organism, 15 years was adequate to assess long-term trends in abundance.
+However, these studies, and other past work, have typically focused either on theoretical aspects of monitoring design or focused on only a few species.
+
+I estimate the minimum time series length required ($T_{min}$) to assess long-term changes in abundance via simple linear regression.
+Statistical power is not a new tool [@Cohen1992], but it is still under-appreciated in ecological research [@Legg2006].
+Therefore, I only briefly describe the idea of statistical power for time series analyses using simulations.
+I then examine 822 population time series to estimate $T_{min}$.
+
+<!--
+
+For time series data, two general approaches to estimating sample size are appropriate.
+Simulations can be designed for a specific population and question [@Bolker2008; @Johnson2015].
+Simple models can be simulated with parameter values corresponding to a population of interest [@Gerrodette1987].
+Statistical power is the proportion of simulations that then meet some set of criteria.
+The specific criteria depend on the question at hand.
+For example, given a time series, when is the slope from linear regression significantly different from zero?
+In other words, when is the time series significantly increasing or decreasing?
+It is then possible to determine how power changes with a variable of interest.
+For example, time series can be simulated for different lengths of time.
+From these simulations, the minimum time series length required to meet certain levels of statistical significance and power is estimated [@Bolker2008].
 
@@
 
-For time series data, two general approaches to estimating sample size are appropriate.
-<!--Typically, calculations of sample size are given by a formula specific to standard statistical tests (e.g.
-t-test).
-However, many ecological scenarios require the use of simulations [@Bolker2008; @Johnson2015].
-As a first approach,-->Simulations can be designed for a specific population and question [@Bolker2008; @Johnson2015].
-Simple models can be simulated with parameter values corresponding to a population of interest [@Gerrodette1987].
-Statistical power is the proportion of simulations that meet some set of criteria.
-The specific criteria depend on the question at hand.
-For example, given a time series, when is the slope from linear regression significantly different from zero?
-In other words, when is the time series significantly increasing or decreasing?
-It is then possible to determine how power changes with a variable of interest.
-For example, time series can be simulated for different lengths of time.
-From these simulations, the minimum time series length required to meet certain levels of statistical significance and power is estimated [@Bolker2008].
-
 In addition to using simulations, empirical time series can also be used.
@@
 
-Past work has investigated questions related to the minimum time series required to estimate trends in population size over time [@Wagner2009; @Giron-Nava2017].
-For example, @Rhodes2011 examined the optimal allocation of effort between spatial and temporal replicates.
-Using simple populations models, they found that the allocation of effort depends on environmental variation, spatial and temporal autocorrelation, and observer error.
-@Rueda-Cediel2015 also used a modeling approach, but parameterized a model specific for a threatened snail, \emph{Tasmaphena lamproides}.
-They found that for this short-lived organism, 15 years was adequate to assess long-term trends in abundance.
-However, these studies, and other past work, have typically focused either on theoretical aspects of monitoring design or focused on only a few species.
-
-I use both simulations and empirical time series to determine the minimum number of years required to address several questions.
-I estimate the minimum time series length required ($T_{min}$) to assess long-term changes in abundance via simple linear regression.
-First, I estimate $T_{min}$ using a simulation approach.
-Then I examine 822 population time series to estimate $T_{min}$.
-In the supplementary material, I determine $T_{min}$ for related ideas: using more complicated population models, varying statistical level and power, and the use of generalized additive models.
+-->
 
 # Methods
+
+## Statistical power in time series
+
+An important step in experimental design is to determine the number of samples required.
+For any particular experiment four quantities are intricately linked: significance level ($\alpha$), statistical power, effect size, and sample size [@Legg2006].
+The exact relationship between these quantities depends on the specific statistical test.
+A type I error is a false positive, or incorrect rejection of a true null hypothesis.
+For example, if a time series was assessed as significantly increasing or decreasing---when there was no true significant trend---this would be a false positive.
+The false positive rate, or significance level ($\alpha$) is often set at 0.05 (although this is purely historical, @Mapstone1995).
+A type II error ($\beta$) is a failure to detect a true trend, or failure to reject a false null hypothesis.
+Formally, statistical power ($1-\beta$) is one minus the probability of a type II error ($\beta$).
+The effect size is a estimate of the strength of a particular phenomenon.
+Prior to an experiment, one could set appropriate levels of power, significance level, and the effect size to estimate the sample size required for the experiment.
+This approach, however, is not straight-forward for a time series, or more complicated scenarios [@Johnson2015], as data are clearly non-independent.
 
@@
 This is the same approach one might use in sample size calculations for any experimental design too complicated for simple power analyses [@Bolker2008;@Johnson2015].
-I only briefly discuss this approach as it has been described elsewhere.
-Essentially, a population model is simulated repetitively for a number of years.
-This approach requires us to determine values for model parameters (e.g.
-population variability).
+Essentially, a population model, with a selected set of parameter values, is simulated repetitively for a number of years.
 As an example, we can take the following population model for population size $N$ at time $t$:
-
 \begin{equation}
@@
 where $\epsilon$ is a normally-distributed random noise term with mean $\mu$ and standard deviation $\sigma$.
-The rate of growth $r$ is also the trend strength of the increase or decrease (i.e.
-the rate of increase).
-It is important to note that any population model could be substituted for equation \ref{sim_model}, as in the supplementary material (Figs.
-\ref{fig:min_time_dist_log_pop}, \ref{fig:shark_example}).
+The rate of growth ($r$) is the trend strength of the increase or decrease (i.e.
+the estimated slope from linear regression).
+It is important to note that any population model could be used here.
+As an example, in the supplementary material (Fig.
+\ref{fig:shark_example}), an age-structured model of population dynamics is used in place of equation \ref{sim_model} [@White2014].
 
-Statistical power is then the proportion of simulations that meet some criteria.
-Here, our criteria is whether the slope parameter from linear regression is significant at the 0.05 threshold with statistical power of 0.8.
+Statistical power is the proportion of simulations where the slope parameter from linear regression is significant at the 0.05 threshold.
 Statistical power of 0.8 would indicate that, if there was a true trend in abundance, there would be a 0.8 probability of detecting the trend.
-Values of 0.05 for the significance level and 0.8 statistical power are historical and it is important to examine the effect of changing these values (Fig.
+Values of 0.05 for the significance level and 0.8 statistical power are historical [@Cohen1992] and it is important to examine the effect of changing these values (Fig.
 \ref{fig:min_time_vs_alpha_beta}).
@@
 In Fig.
-\ref{fig:theoretical_approach}a, a number of simulated time series are shown for a set number of time periods ($t=40$).
-It is clear that statistical power increases quickly with increases in length of time sampled (Fig.
+\ref{fig:theoretical_approach}a, the model in equation \ref{sim_model} is simulated repetitively.
+Statistical power increases with increases in the length of time sampled (Fig.
 \ref{fig:theoretical_approach}b).
 Where power is greater than 0.8 (the dotted line), that is the minimum time required ($T_{min}$) to be confident in the detection of a long-term trend in abundance.
+As shown previously [@Rhodes2011; @Rueda-Cediel2015], statistical power increases with larger trend strength and lower population variability (Fig.
+\ref{fig:theoretical_approach}c,d).
+Simulation approaches can be useful before designing a monitoring program or when a realistic model exists for the population in question.
+However, it is also informative to examine how statistical power changes with sampling effort in empirical time series.
+
+<!-- code to pull in databases-->
+```{r, echo=F}
+# Pull in original database from Keith et al. 2015 and the cleaned version after our series of filtering steps
+
+#Load keith et al. database (from GPDD database)
+  load('../data/Keith_et_al_2015_data.Rdata')
+
+#Load our cleaned data (see other code for cleaning steps)
+  load('../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata') #Do i need to filter out non-significant populations as well (where no trend was present)
+ # pop_info=pop_info[pop_info$trend_p_value<0.05,]
+ # long_dat=subset(long_dat,long_dat$ID %in% pop_info$ID)
+ # long_time_names=subset(long_time_names,long_time_names %in% pop_info$ID)
+```
+
+## Data source
+I use a database of `r length(table(dat$ID))` population time series compiled by @Keith2015.
+The data are originally from the Global Population Dynamics Database [@GPDD2010] and several other sources [@Keith2015].
+I filtered out short time series (less than `r pop_info$series_length[1]` years), and those with missing data, leaving `r nrow(pop_info)` time series.
+The data includes information on `r length(table(pop_info$Binomial))` vertebrate species with a focus on mammals, birds, and fish.
+<!-- The data also includes information on generation length and survey specifications.
+For each time series, I also calculate other variables of interest: coefficient of variance in population size, long-term trend in abundance (slope coefficient from simple linear regression), and temporal autocorrelation.-->
+
+```{r,echo=F}
+# Pull in results of biological characteristics comparison to min time series length required
+load('../cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
+```
+<!--For a subset of populations ($n =$ `r nrow(LPI_pop_info)`), there is information on life-history traits available from another paper [@Myhrvold2015], including body size and generation length.
+All `r nrow(LPI_pop_info)` populations were birds.
+I examine how the minimum time required is related to these life-history traits (Fig.
+\ref{fig:correlates}d-h).-->
+
+## Empirical approach
+
+I assume that each time series is long enough to include all necessary information (e.g.
+variability) about the population.
+In other words, each time series is a representative sample.
+I first take all possible contiguous subsamples of each time series.
+For example, a time series of 35 years would have 34 possible contiguous subsamples of length 2, 34 possible contiguous subsamples of length 3, and continuing until 1 possible contiguous subsample of length 35 [@Gerber1999; @Brashares2005; @Giron-Nava2017].
+In line with the simulation approach, I determine the proportion of subsamples of a particular length that have estimated slope coefficients statistically different from zero.
+This proportion is a measure of statistical power.
+Lastly, I determine which subsample length is required to achieve a certain threshold of statistical power (0.8, @Cohen1992).
+The minimum subsampled length that met these criteria is the minimum time series length required ($T_{min}$).
+All analyses were conducted in R [@RCoreTeam2016].
+
+<!--
+In the supplementary material, I show how the same approach described here for more complicated population models [@White2014].
+I also determine the minimum time required to estimate long-term trends according to generalized additive models, instead of the simple linear models used here (Fig.
+\ref{fig:gam_example}).
+-->
+
+<!--
+## Questions addressed in the appendix
+
+I also examine how minimum time series calculations changed depending on the question of interest.
+In the supplementary material, I use similar calculations to determine the minimum time series required to estimate overall population growth rate (Fig.
+\ref{fig:growth_rate}).
+I also explore how the method can be applied to more biologically-realistic population models (Fig.
+\ref{fig:shark_example}.
+More specifically, I determine the minimum time required to estimate long-term population trends using a stochastic, age-structured model of lemon shark population dynamics in the Bahamas [@White2014].
+Lastly, I determine the minimum time required to estimate long-term trends according to generalized additive models, instead of the simple linear models used here.
+-->
+
+# Results
+
+<!--
+I determined the minimum time series length ($T_{min}$) required to determine, via linear regression, long-term population trends.
+Here, the minimum time series length required had high enough statistical power (greater than 0.8) for a set significance level ($\alpha$) of 0.05.
+It is also possible to alter statistical power and $\alpha$.
+Predictably, with increased statistical power or decreased $\alpha$, $T_{min}$ increased (Fig.
+\ref{fig:min_time_vs_alpha_beta}).
+I then estimated $T_{min}$ for each population time series.
+-->
+
+<!--
+## Simulation approach
+
+I constructed a general population model where the trend strength (i.e.
+slope coefficient) over time could be a model parameter.
+I then simulated time series of different lengths.
+From these simulations I determined the minimum time series length required to achieve a certain level of statistical power.
+In line with past work [@Gerrodette1987], I found the $T_{min}$ increases (i.e.
+more time is required) with decreases in trend strength and with increases in population variability (Figs.
+\ref{fig:theoretical_approach}c,d).
+
+I chose a simple model, but any other population model could be used (see example in Fig.
+\ref{fig:min_time_dist_log_pop}).
+Ideally, the specific model choice should be tailored to the population of interest.
+I explored how the simulation approach could be applied to more biologically-realistic population models (Fig.
+\ref{fig:shark_example}).
+Specifically, I determined the minimum time required to estimate long-term population trends using a stochastic, age-structured model of lemon shark population dynamics in the Bahamas [@White2014].
+I found that over 27 years of continuous monitoring were needed in this particular scenario (Fig.
+\ref{fig:shark_example}).
+Similar to the simulation approach described above, the minimum time required for the lemon shark population was strongly dependent on model parameters.
+-->
+
+```{r poisson_model, echo=F,include=F}
+# I should put this in the appendix
+#Poisson regression
+pop_info$abs_overall_trend = abs(pop_info$overall_trend)
+time_series_correlates_poisson_model <-glm(min_time_for_power~ abs_overall_trend + autocorrelation + coefficient_variation,family = 'poisson',data=pop_info)
+#par(mfrow=c(2,2),mar=c(5,5,1,2),oma=c(0,0,0.5,0))
+#plot(poisson_model,cex.lab=1.2,las=1,cex.axis=1.2,main=' ')
+
+#R^2 calculation (pg. 218 Zuur)
+R2 <- 100*(time_series_correlates_poisson_model$null.deviance - time_series_correlates_poisson_model$deviance)/(time_series_correlates_poisson_model$null.deviance)
+
+```
+
+I examined a database of `r nrow(pop_info)` separate population time series representing `r length(table(pop_info$Binomial))` species.
+This database consisted of vertebrate species with a variety of life-history characteristics.
+Across all the populations, I found an average minimum time series length required ($T_{min}$) of `r round(mean(pop_info$min_time_for_power),1)` (SD=`r round((var(pop_info$min_time_for_power))^0.5,1)`), with a wide distribution (Fig.
+\ref{fig:min_time_dist}b).
+Estimates of $T_{min}$ varied between biological class (Fig.
+\ref{fig:min_time_dist}a).
+Ray-finned fish (class Actinopterygii) typically had estimates of $T_{min}$ over 20 years.
+Birds (class Aves) had a much wider distribution of $T_{min}$, but usually required less years of sampling.
+Differences between these classes can be explained by differences in variability in population size and strength of trends in abundance (Fig.
+\ref{fig:class}).
+
+## Corrrelates for minimum time required
+
+The minimum time series length required was strongly correlated with trend strength (i.e.
+estimated slope coefficient from linear regression), coefficient of variation in population size, and autocorrelation in population size (Figs.
+\ref{fig:correlates}a-c).
+All three of these explanatory variables were significant and had large effect sizes (see Table A1).
+Combined, trend strength, coefficient of variation in population, and autocorrelation account for `r round(100*(time_series_correlates_poisson_model$null.deviance - time_series_correlates_poisson_model$deviance)/(time_series_correlates_poisson_model$null.deviance),1)`% of the explained deviance [@Zuur2009] in minimum time series length required.
+
+<!-- Minimum time required and biological correlates-->
+```{r, echo=F}
+load('../cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
+
+biological_correlates_poisson_model <-glm(min_time_for_power~  litter_or_clutch_size_n  + gen_len + adult_body_mass_g + maximum_longevity_y  +incubation_d + trophic,family = 'poisson',data=LPI_pop_info)
+R2_biological_correlates_poisson_model <- 100*(biological_correlates_poisson_model$null.deviance - biological_correlates_poisson_model$deviance)/(biological_correlates_poisson_model$null.deviance)
+
+all_correlates_poisson_model <-glm(min_time_for_power~ abs(overall_trend) + coefficient_variation + autocorrelation +  litter_or_clutch_size_n  + gen_len + adult_body_mass_g + maximum_longevity_y  +incubation_d + trophic,family = 'poisson',data=LPI_pop_info)
+R2_all_correlates_poisson_model <- 100*(all_correlates_poisson_model$null.deviance - all_correlates_poisson_model$deviance)/(all_correlates_poisson_model$null.deviance)
+```
+
+There was life-history information available [@Myhrvold2015] for `r nrow(LPI_pop_info)` populations representing `r length(table(LPI_pop_info$Binomial))` different species, all of which were birds (Aves class).
+Some life-history traits were significant predictors for the minimum time required (Figs.
+\ref{fig:correlates}d-h, Tables A2,A3).
+However, even combined, all five of the life-history traits accounted for only `r round(R2_biological_correlates_poisson_model,3)`% of the explained deviance in minimum time series length required.
+In addition, when accounting for trend strength, coefficient of variation, and autocorrelation, no life-history traits were significant predictors of the minimum time required (Table A3).
+
+## Evalulating the IUCN criteria
+
+```{r, echo=F}
+declining_populations = subset(pop_info,pop_info$percent_change_10years>30 & pop_info$overall_trend<0)
+```
+
+I examined a subset of populations with observed declines of 30% or greater over 10 years, qualifying all of them as vulnerable under IUCN Criterion A2 [@IUCN2012].
+This resulted in n = `r nrow(declining_populations)` populations.
+I then compared the minimum time required to achieve 0.8 statistical power ($T_{min}$) to the minimum time required under the IUCN criteria (Fig.
+\ref{fig:IUCN_analysis}).
+For populations below the identity line in figure \ref{fig:IUCN_analysis}, IUCN criteria would require more sampling compared to estimates for $T_{min}$.
+Further, populations above the identity line are cases where the IUCN criteria would classify a population as vulnerable despite not having sampled enough years to achieve high statistical power (Fig.
+\ref{fig:IUCN_analysis}).
+The silhouettes on figure \ref{fig:IUCN_analysis} highlight that species with long generation times had larger discrepancies between $T_{min}$ and the minimum time required for IUCN assessments (Fig.
+\ref{fig:IUCN_correlates}).
+
+<!--
+### Sensitivity analysis
+Lastly, I tested model sensitivity by using generalized additive models (GAMs) instead of simple linear regression.
+Again, I examined the minimum time required to estimate long-term population trends (Fig.
+\ref{fig:gam_example}).
+I found that although I obtain a similar distribution of minimum times required for GAMs, the minimum time required for GAMs is on average shorter than for linear regression (Fig.
+\ref{fig:min_time_dist_gam}).
+-->
+
+# Discussion
+
+<!--I examined the minimum time series length required to address a particular question of interest: what is the minimum time series length required to determine long-term population trends using linear regression?
+This is one of the simplest questions one could ask of a time series.
+The simulation-based approach has been suggested by others, especially in situations more complicated than that suited for classic power analysis [@Gerrodette1987;@Johnson2015;@Bolker2008].
+My simulations support past work that longer time series are needed when the trend strength (i.e.
+rate of increase or decrease) is weak or when population variability is high [@Gerrodette1987].
+I also showed how the simulation model can be altered for a particular population (Fig.
+\ref{fig:shark_example}) or question (Figs.
+\ref{fig:min_time_dist_log_pop},\ref{fig:gam_example}).-->
+
+I examined `r nrow(pop_info)` population time series.
+I then subsampled each to determine the minimum time required to achieve a desired significance level and power for linear regression.
+Statistical power is important as it provides information on the necessary number of samples required to determine a significant trend [@Legg2006].
+I found that on average `r round(mean(pop_info$min_time_for_power),2)` years of continuous monitoring were typically necessary (Fig.
+\ref{fig:min_time_dist}b).
+However, the distribution of minimum time required was wide.
+Therefore, simplistic rules of thumb are probably not appropriate.
+This time-frame is in line with past work on a short-lived snail species [@Rueda-Cediel2015] and a long-lived whale species [@Gerber1999].
+@Hatch2003 used seabird monitoring data to estimate minimum sampling requirements.
+He found that the time required ranged from 11 to 69 years depending on species, trend strength, and study design.
+
+In line with theoretical predictions [@Rhodes2011], I also found $T_{min}$ was strongly correlated with the trend strength, variability in population size, and temporal autocorrelation (Fig.
+\ref{fig:correlates}).
+Contrary to my prior expectations, I also found that $T_{min}$ did not correlate with any life-history traits (Fig.
+\ref{fig:correlates}d-h).
+I initially hypothesized that species with longer lifespans or generation times may require a longer sampling period.
+This result could have been a result of at least two factors.
+First, the data I used may not include a diverse enough set of species with different life-history traits.
+Second, the question I posed, whether a population is increasing or decreasing, was specifically concerned with population trends over time.
+Therefore, life-history characteristics may be more important for other questions more closely tied to species biology.
+For example, @Blanchard2007 used detailed simulations of spatially-distributed fisheries to compare survey designs.
+They found that statistical power depended on survey design, temperature preferences, and the degree of population patchiness.
+
+<!-- Similarly, @Bart2004 recommended a standard of 80% power to detect a 50% decline over 20 years in landbird populations.
+They found that only about 42% of populations in the Breeding Bird Survey met their criteria.
+@Seavy2007 asked whether statistical power was even a useful framework for assessing long-term population trends.
+They used 24 years of census data on Red-tailed Tropicbirds (\emph{Phaethon rubricauda}) in Hawaii and showed that to detect a 50% decline over 10 years almost always resulting in high statistical power (above 0.8).
+Therefore, they cautioned against only using power analyses to design monitoring schemes and instead argued for metrics that would increase precision.
+For example, @Seavy2007 suggest improving randomization, reducing bias, and increases detection probability when designing and evaluating monitoring programs.-->
+
+An important related idea is the optimal allocation of sampling effort in space versus time.
+In a theoretical investigation of this question, @Rhodes2011 found that the optimal allocation of sampling depended strongly on temporal and spatial autocorrelation.
+If spatial population dynamics were highly correlated, then it was better to sample more temporally, and vice versa.
+My work supports this idea as populations with strong temporal autocorrelation needed less years of sampling (Fig.
+\ref{fig:correlates}).
+@Morrison2008 also studied the optimal allocation of sampling effort in space versus time, but used emprical data from a long-term survey of the collared pika (\emph{Ochotona collaris}) in the Yukon.
+They found that surveys less than 5 years may be misleading and that extrapolating from one population to another, even when nearby geographically, may be untenable.
+
+@Seavy2007 asked whether statistical power was even a useful framework for assessing long-term population trends.
+They used 24 years of census data on Red-tailed Tropicbirds (\emph{Phaethon rubricauda}) in Hawaii and showed that to detect a 50% decline over 10 years almost always resulted in high statistical power (above 0.8).
+Therefore, they cautioned against only using power analyses to design monitoring schemes and instead argued for metrics that would increase precision: improving randomization, reducing bias, and increasing detection probability.
+Power analyses should not be the only consideration when designing monitoring schemes.
+However, unlike @Seavy2007, the results here indicate that longer than 10 years is often needed to achieve high statistical power.
+
+As a test of the consequences of not conducting power analyses, I compared the results here to the IUCN Redlist criteria.
+IUCN criteria A2 suggests that species that have experienced 30% declines over 10 years (or three generations) should be listed as vulnerable [@IUCN2012].
+However, for the populations I examined, this criteria may be too simplistic (Fig.
+\ref{fig:IUCN_analysis}).
+For many populations, the IUCN criteria suggest more years than necessary are required to assess a population as vulnerable (points below diagonal line in Fig.
+\ref{fig:IUCN_analysis}).
+Conversely, for other populations the IUCN criteria suggest sampling times that are less than the minimum time required for statistical power.
+This suggests that the IUCN criteria are probably too simplistic as the minimum time required does not correlate with generation time or other biological covariates (Fig.
+\ref{fig:correlates}d-h).
+
+The design of monitoring programs should include calculations of statistical power, the allocation of sampling in space versus time [@Rhodes2011], and metrics to increase precision [@Seavy2007].
+Ideally, a formal decision analysis to evaluate these different factors would be conducted to design or assess any monitoring program [@Hauser2006; @McDonald-Madden2010].
+This type of formal decision analysis would also include information on the costs of monitoring.
+These costs include the actual costs of sampling [@Brashares2005] and the ecological costs on inaction [@Thompson2000].
+
+<!--
+@Rueda-Cediel2015 found that 10-15 years was adequate to characterize population trends in a short lived animal. They used a matrix model of a particular species...
+
+@Gerber1999 investigated the minimum time series required to estimate population growth of an endagered, but recovering, species.
+They used a long-term census to retroactively determine the minimum time series required to assess threat status.
+They found that only XX years were needed, XX years before the delisting decision was actually made.
+This highlights the importance of estimating the minimunm time series required as an earlier decision would have saved time and money [@Gerber1999].
+-->
+
+## Limitations
+
+This paper has some limitations in determining the minimum time series length required.
+First, $T_{min}$ is particular to the specific question of interest.
+An additional complication is that for the empirical approach, the subsampling of the full time series allows for estimates of power, but the individual subsamples are clearly not independent of one another.
+In an ideal setting, a specific population model would be parameterized for each population of interest.
+Then, model simulations could be used to estimate the minimum time series required to address each specific question of interest.
+Clearly, this is not always practical, especially if conducting analyses for a wide array of species.
+<!--In addition, the statistical models suggest that $T_{min}$ does not correlate with any life-history traits, at least for the question of linear regression (Fig.
+\ref{fig:correlates}d-h).
+Therefore, it is not possible to use these results to predict $T_{min}$ for another population, even if the population is of a species with a similar life-history to one in the database used.-->
+
+## Conclusions
+
+Power analyses are not a novel aspect of ecological research [@Legg2006].
+However, power analyses are still underutilized, especially in the context of time series analyses.
+I used a database of `r nrow(pop_info)` populations to determine the minimum time series length required to detect population trends.
+This goes beyond previous work that either focused on theoretical investigations or a limited number of species.
+I show that to identify long-term changes in abundance, on average `r round(mean(pop_info$min_time_for_power),2)` years of continuous monitoring are often required (Fig.
+\ref{fig:min_time_dist}).
+However, there is wide distribution of estimated minimum times.
+Therefore, it is probably not wise to use a simple threshold number of years in monitoring design.
+Further, contrary to my initial hypotheses, minimum time required did not correlate with generation time or any other life-history traits (Fig.
+\ref{fig:correlates}d-h).
+This result argues against overly simplified measures of minimum sampling time based on generation length or other life-history traits (Fig.
+\ref{fig:IUCN_analysis}).
+
+My work implies that for many populations, short time series are probably not reliable for detecting population trends.
+This result highlights the importance of long-term monitoring programs.
+From both a scientific and management perspective, estimates of $T_{min}$ are important.
+If a time series is too short, we lack the statistical power to reliably detect long-term population trends.
+In addition, a time series that is too long may be a poor use of already limited funds [@Gerber1999].
+Further, more data is not always best in situations where management actions need to be taken [@Martin2012; @Martin2017].
+When a population trend is detected, it may be too late for management action.
+In these situations, the precautionary principle may be more appropriate [@Thompson2000].
+Future work should examine other species, with a wider range of life-history characteristics.
+In addition, similar approaches can be used to determine the minimum time series length required to address additional questions of interest.
+
+<!--
+## Time series length recommendations from other papers
+
+- greater than 5 years [@Morrison2008]
+- more than 10-15 years [@Rueda-Cediel2015]
+- general advice (variability, correlations) [@Rhodes2011]
+- more than 10 years [@Gerber1999]
+- specific recommendations depend on amount of power and strength of trend (Bart 2004, Hatch 2003) [@Seavy2007]
+- paper on syncrony from James
+-->
+
+# Supporting Information
+
+In the supporting material, I provide an expanded methods sections, additional figures, minimum time calculations for determining exponential growth, simulations with a more complicated population model, and the use of generalized additive models to identify population trends.
+All code and data can be found at https://github.com/erwhite1/time-series-project
+
+# Acknowledgements
+
+ERW was partially supported by a National Science Foundation Graduate Fellowship.
+I would like to thank members of the Ecological Theory group at the University of California, Davis for their insight.
+<!--I would also like to thank T.
+Dallas.
+E.
+Malcolm-White and X anonymous reviewers for their helpful comments.
+-->
+
+# References
+
+<div id="refs"></div>
+
+\clearpage
+
+\pagebreak
+
+# Figure captions
+
+Figure 1: (a) Example of a simulated time series for 40 time periods.
+(b) Statistical power versus the simulated time series length.
+The horizontal, dashed line is the desired statistical power of 0.8.
+The vertical, dashed line is the minimum time required to achieve the desired statistical power.
+(c) Minimum time required ($T_{min}$) for simulations with different values of the trend strength ($r$) and $\sigma = 5.0$.
+(d) Minimum time required for different levels of population variability ($\sigma$) and $r=1.5$.
+In each case, the minimum time required is the minimum number of years to achieve 0.8 statistical power given a significance level of 0.05..
+
+Figure 2: (a) Distributions of the minimum time required for populations from four different biological classes.
+(b) Distribution of minimum time required for all populations regardless of biological class.
+The minimum time required calculation corresponds to a significance level of 0.05 and statistical power of 0.8.
+
+Figure 3: Minimum time required to estimate change in abundance correlated with (a) trend strength (absolute value of slope coefficient estimated from linear regression), (b) coefficient of variation in interannual population size, (c) temporal lag-1 autocorrelation, (d) generation length (years), (e) litter size (n), (f) log adult body mass (grams), (g) maximum longevity (years), and (h) incubation (days).
+The lines in each plot represent the best fit line from linear regression.
+
+Figure 4: Minimum time required to achieve 0.8 statistical power versus the minimum time required under IUCN criteria A2 to classify a species as vulnerable.
+Each point represents a single population, all of which saw declines of 30% or greater over a 10 year period.
+The silhouettes highlight that species with longer generation times typically have larger discrepancies between $T_{min}$ and the minimum time required for IUCN assessments.
+
+\pagebreak
 
@@
 The vertical, dashed line is the minimum time required to achieve the desired statistical power.
-(c) Minimum time required ($T_{min}$) for simulations with different values of the trend strength ($r$).
-(d) Minimum time required for different levels of population variability ($\\sigma$).
+(c) Minimum time required ($T_{min}$) for simulations with different values of the trend strength ($r$) and $\\sigma = 5.0$.
+(d) Minimum time required for different levels of population variability ($\\sigma$) and $r=1.5$.
 In each case, the minimum time required is the minimum number of years to achieve 0.8 statistical power given a significance level of 0.05.\\label{fig:theoretical_approach}'}
@@
 
-<!-- code to pull in databases-->
-```{r, echo=F}
-# Pull in original database from Keith et al. 2015 and the cleaned version after our series of filtering steps
-
-#Load keith et al. database (from GPDD database)
-  load('../data/Keith_et_al_2015_data.Rdata')
-
-#Load our cleaned data (see other code for cleaning steps)
-  load('../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata') #Do i need to filter out non-significant populations as well (where no trend was present)
- # pop_info=pop_info[pop_info$trend_p_value<0.05,]
- # long_dat=subset(long_dat,long_dat$ID %in% pop_info$ID)
- # long_time_names=subset(long_time_names,long_time_names %in% pop_info$ID)
-```
-
-## Data source
-I use a database of `r length(table(dat$ID))` population time series compiled in [@Keith2015]; they compared the predictability of growth rates among populations.
-The data are originally from the Global Population Dynamics Database [@GPDD2010] and several other sources [@Keith2015].
-I filtered out short time series (less than `r pop_info$series_length[1]` years), and those with missing data, leaving `r nrow(pop_info)` time series.
-The data includes information on `r length(table(pop_info$Binomial))` vertebrate species with a focus on mammals, birds, and fish.
-The data also includes information on generation length and survey specifications.
-For each time series, I also calculate other variables of interest: coefficient of variance in population size, long-term trend in abundance (slope coefficient from simple linear regression), and temporal autocorrelation.
-All analyses were conducted in R [@RCoreTeam2016].
-
-```{r,echo=F}
-# Pull in results of biological characteristics comparison to min time series length required
-load('../cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
-```
-For a subset of populations ($n =$ `r nrow(LPI_pop_info)`), there is information on life-history traits available from another paper [@Myhrvold2015], including body size and generation length.
-All `r nrow(LPI_pop_info)` populations were birds.
-I examine how the minimum time required is related to these life-history traits (Fig.
-\ref{fig:biological_correlates}).
-
-## Empirical approach
-
-I assume that each time series is long enough to include all necessary information (e.g.
-variability) about the population.
-In other words, each time series is a representative sample.
-I first take all possible contiguous subsamples of each time series.
-For example, a time series of 35 years would have 34 possible contiguous subsamples of length 2, 34 possible contiguous subsamples of length 3, and continuing until 1 possible contiguous subsample of length 35 [@Gerber1999; @Giron-Nava2017].
-Next, I run a linear regression for each subsampled time series.
-Then, I determine the proportion of subsamples of a particular length that have estimated slope coefficients which are statistically different from zero.
-I only look at the proportion of samples where the long-term, or ``true'', time series also has a significant slope.
-This proportion is a measure of statistical power.
-Lastly, I determine which subsample length is required to achieve a certain threshold of statistical power (0.8, @Cohen1992).
-The minimum subsampled length that met these criteria is the minimum time series length required ($T_{min}$).
-
-In the supplementary material, I show how the same approach described here for more complicated population models.
-I also determine the minimum time required to estimate long-term trends according to generalized additive models, instead of the simple linear models used here (Fig.
-\ref{fig:gam_example}).
-
-<!--
-## Questions addressed in the appendix
-
-I also examine how minimum time series calculations changed depending on the question of interest.
-In the supplementary material, I use similar calculations to determine the minimum time series required to estimate overall population growth rate (Fig.
-\ref{fig:growth_rate}).
-I also explore how the method can be applied to more biologically-realistic population models (Fig.
-\ref{fig:shark_example}.
-More specifically, I determine the minimum time required to estimate long-term population trends using a stochastic, age-structured model of lemon shark population dynamics in the Bahamas [@White2014].
-Lastly, I determine the minimum time required to estimate long-term trends according to generalized additive models, instead of the simple linear models used here.
--->
-
-# Results
-
-I determined the minimum time series length ($T_{min}$) required to address a particular question of interest.
-What is the minimum time series length required to determine, via linear regression, the long-term population trend?
-Here, the minimum time series length required had high enough statistical power (greater than 0.8) for a set significance level ($\alpha$) of 0.05.
-It is also possible to alter statistical power and $\alpha$.
-Predictably, with increased statistical power or decreased $\alpha$, $T_{min}$ increased (Fig.
-\ref{fig:min_time_vs_alpha_beta}).
-I then estimated $T_{min}$ using two approaches.
-I briefly describe results from the simulation approach and then discuss the empirical approach.
-
-## Simulation approach
-
-I constructed a general population model where the trend strength (i.e.
-slope coefficient) over time could be a model parameter.
-I then simulated time series of different lengths.
-From these simulations I determined the minimum time series length required to achieve a certain level of statistical power.
-In line with past work [@Gerrodette1987], I found the $T_{min}$ increases (i.e.
-more time is required) with decreases in trend strength and with increases in population variability (Figs.
-\ref{fig:theoretical_approach}c,d).
-
-I chose a simple model, but any other population model could be used (see example in Fig.
-\ref{fig:min_time_dist_log_pop}).
-Ideally, the specific model choice should be tailored to the population of interest.
-I explored how the simulation approach could be applied to more biologically-realistic population models (Fig.
-\ref{fig:shark_example}).
-Specifically, I determined the minimum time required to estimate long-term population trends using a stochastic, age-structured model of lemon shark population dynamics in the Bahamas [@White2014].
-I found that over 27 years of continuous monitoring were needed in this particular scenario (Fig.
-\ref{fig:shark_example}).
-Similar to the simulation approach described above, the minimum time required for the lemon shark population was strongly dependent on model parameters.
-
-## Empirical approach
+\pagebreak
 
@@
  load("../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
+library(viridis)
+pop_info$color = as.numeric(as.factor(pop_info$class))
+pop_info$color = viridis_pal(alpha=0.6,option='viridis')(5)[pop_info$color]
+pop_info$shape = as.numeric(as.factor(pop_info$color)) + 14
 
 require(ggplot2)
-plot1<- ggplot(aes(class,min_time_for_power),data=pop_info) + geom_violin() + geom_jitter(col=rgb(0.5,0.5,0.5,0.3),height = 0, width = 0.1) +
+plot1<- ggplot(aes(class,min_time_for_power),data=pop_info) + geom_violin() + geom_jitter(col=pop_info$color,height = 0, width = 0.1,pch=pop_info$shape) +
   theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(), axis.line = element_line(colour = "black"),axis.text = element_text(size=12),axis.title.x =element_text(size=16) ,axis.title.y =element_text(size=16) ) +
@@
 
-```{r poisson_model, echo=F,include=F}
-# I should put this in the appendix
-#Poisson regression
-pop_info$abs_overall_trend = abs(pop_info$overall_trend)
-time_series_correlates_poisson_model <-glm(min_time_for_power~ abs_overall_trend + autocorrelation + coefficient_variation,family = 'poisson',data=pop_info)
-#par(mfrow=c(2,2),mar=c(5,5,1,2),oma=c(0,0,0.5,0))
-#plot(poisson_model,cex.lab=1.2,las=1,cex.axis=1.2,main=' ')
+\clearpage
 
-#R^2 calculation (pg. 218 Zuur)
-R2 <- 100*(time_series_correlates_poisson_model$null.deviance - time_series_correlates_poisson_model$deviance)/(time_series_correlates_poisson_model$null.deviance)
-
-```
-
+<!-- Individual figures for correlates, they are now combined into one
 ```{r,echo=F, fig.cap='Minimum time required to estimate change in abundance correlated with (a) trend strength (absolute value of slope coefficient estimated from linear regression), (b) coefficient of variation in interannual population size, and (c) temporal lag-1 autocorrelation.\\label{fig:correlates}',fig.height=7, fig.width=4}
@@
 # Plot minimum time required against trend strength, population variance, and autocorrelation
+
   par(mfrow=c(3,1),mar=c(3,2,2,1),oma=c(1,4.5,0,0.5))
@@
 
-I examined a database of `r nrow(pop_info)` separate population time series representing `r length(table(pop_info$Binomial))` species.
-This database consisted of vertebrate species with a variety of life-history characteristics (Fig.
-\ref{fig:biological_correlates}).
-I limited analyses to populations with at least 35 years of continuous sampling.
-I then examined the minimum time required to estimate long-term trends via linear regression.
-
-Across all the populations, I found an average minimum time series length required ($T_{min}$) of `r round(mean(pop_info$min_time_for_power),1)` (SD=`r round((var(pop_info$min_time_for_power))^0.5,1)`), with a wide distribution (Fig.
-\ref{fig:min_time_dist}b).
-Estimates of $T_{min}$ varied between biological class (Fig.
-\ref{fig:min_time_dist}a).
-Ray-finned fish (class Actinopterygii) typically had estimates of $T_{min}$ over 20 years.
-Birds (class Aves) had a much wider distribution of $T_{min}$, but usually required less years of sampling.
-Differences between these classes can be explained by differences in variability in population size and strength of trends in abundance (Fig.
-\ref{fig:class}).
-
-### Corrrelates for minimum time required
-
-The minimum time series length required was strongly correlated with trend strength (i.e.
-estimated slope coefficient from linear regression), coefficient of variation in population size, and autocorrelation in population size (Fig.
-\ref{fig:correlates}).
-This is in line based on simulations here and those of others [@Rhodes2011].
-Using a generalized linear model, with a Poisson error structure, all three of these explanatory variables were significant and had large effect sizes (see Table A1).
-Combined, trend strength, coefficient of variation in population, and autocorrelation account for `r round(100*(time_series_correlates_poisson_model$null.deviance - time_series_correlates_poisson_model$deviance)/(time_series_correlates_poisson_model$null.deviance),1)`% of the explained deviance [@Zuur2009] in minimum time series length required.
-
-<!-- Minimum time required and biological correlates-->
-```{r, echo=F}
-load('../cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
-```
-
-For a subset of the populations I combined time series data with a data on life-history characteristics of amniotes [@Myhrvold2015].
-There was life-history information available for `r nrow(LPI_pop_info)` populations representing `r length(table(LPI_pop_info$Binomial))` different species, all of which were birds (Aves class).
+\pagebreak
 
@@
 ```
+-->
 
-Some life-history traits were significant predictors for the minimum time required (Fig.
-\ref{fig:biological_correlates}, Tables A2,A3).
-However, none of these life-history traits explained a large part of the variation in minimum time required.
-In a generalized linear model, all of the life-history traits described in figure \ref{fig:biological_correlates} account for only `r round(R2_biological_correlates_poisson_model,3)`% of the explained deviance in minimum time series length required.
-In addition, when accounting for trend strength, coefficient of variation, and autocorrelation, no life-history traits were significant predictors of the minimum time required (Table A3).
+```{r,echo=F, fig.cap='Minimum time required to estimate change in abundance correlated with (a) trend strength (absolute value of slope coefficient estimated from linear regression), (b) coefficient of variation in interannual population size, (c) temporal lag-1 autocorrelation, (d) generation length (years), (e) litter size (n), (f) log adult body mass (grams), (g) maximum longevity (years), and (h) incubation (days).
+The lines in each plot represent the best fit line from linear regression.\\label{fig:correlates}',fig.height=7}
 
-### Evalulating the IUCN criteria
+# Plot minimum time required against trend strength, population variance, and autocorrelation  
+ load("../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
+library(viridis)
+pop_info$color = as.numeric(as.factor(pop_info$class))
+pop_info$color = viridis_pal(alpha=0.3,option='viridis')(5)[pop_info$color]
+pop_info$shape = as.numeric(as.factor(pop_info$color)) + 14
+
+  par(mfcol=c(4,2),mar=c(3.5,2,2.5,1),oma=c(1,4.5,0,0.5))
+  plot(abs(pop_info$overall_trend),pop_info$min_time_for_power,pch=pop_info$shape,col=pop_info$color,las=1,ylab='',xlab='',cex.axis=1.2,cex=1.3,ylim=c(0,35))
+  mtext('| trend strength |',1,line=3,cex=1.2)
+  mtext('A',3,line=0,adj = 0.0,cex=1.2)
+    plot(pop_info$coefficient_variation,pop_info$min_time_for_power,pch=pop_info$shape,col=pop_info$color,las=1,ylab='',xlab='',cex.axis=1.2, xlim = c(0.2,2.2),cex=1.3,ylim=c(0,35))
+    mtext('coefficient of variation',1,line=3,cex=1.2)
+    mtext('B',3,line=0,adj = 0.0,cex=1.2)
+      plot(pop_info$autocorrelation,pop_info$min_time_for_power,pch=pop_info$shape,col=pop_info$color,las=1,ylab='',xlab='',cex.axis=1.2, xlim = c(-0.5,1),cex=1.3,ylim=c(0,35))
+      mtext('autocorrelation',1,line=3,cex=1.2)
+      mtext('C',3,line=0,adj = 0.0,cex=1.2)
+ #      mtext('minimum time required',2,line=2.5,cex=1.2,outer = T)
+
+legend('bottomleft',legend=c('Chondrichthyes','Aves','Actinopterygii','Mammalia'),pch=as.numeric(names(table(pop_info$shape))),col=c(names(table(pop_info$color))),bty ="n",cex=1.2,pt.cex = 1.4)
+
+# Biological correlates
+
+load('../cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
+#par(mfrow=c(3,2),oma=c(5,4,1,1),mar=c(5,3,0,0))
+#x_axis_name_list = c('generation length (years)','litter size (n)','adult body mass (grams)','maximum longevity (years)','egg mass (grams)','incubation (days)')
+
+LPI_pop_info$color = '#3B528B66'
+LPI_pop_info$shape = 16
+
+  plot(pop_info$gen_len,pop_info$min_time_for_power,pch=pop_info$shape,col=pop_info$color,las=1,ylab='',xlab='',cex.axis=1.2,cex=1.2)
+  abline(lm(LPI_pop_info$min_time_for_power~LPI_pop_info$gen_len),lwd=3)
+  mtext(text = 'generation length (years)',side = 1,line = 3,cex = 1.2)
+  mtext('D',3,line=0,adj = 0.0,cex=1.2)
+
+  plot(LPI_pop_info$litter_or_clutch_size_n,LPI_pop_info$min_time_for_power,pch=LPI_pop_info$shape,col=LPI_pop_info$color,las=1,ylab='',xlab='',cex.axis=1.2,cex=1.2)
+  abline(lm(LPI_pop_info$min_time_for_power~LPI_pop_info$litter_or_clutch_size_n),lwd=3)
+  mtext(text = 'litter size (n)',side = 1,line = 3,cex = 1.2)
+  mtext('E',3,line=0,adj = 0.0,cex=1.2)
+
+    plot(log(LPI_pop_info$adult_body_mass_g),LPI_pop_info$min_time_for_power,pch=LPI_pop_info$shape,col=LPI_pop_info$color,las=1,ylab='',xlab='',cex.axis=1.2,cex=1.2)
+  abline(lm(LPI_pop_info$min_time_for_power~LPI_pop_info$adult_body_mass_g),lwd=3)
+  mtext(text = 'log(adult mass (grams))',side = 1,line = 3,cex = 1.2)
+  mtext('F',3,line=0,adj = 0.0,cex=1.2)
+
+  plot(LPI_pop_info$maximum_longevity_y,LPI_pop_info$min_time_for_power,pch=LPI_pop_info$shape,col=LPI_pop_info$color,las=1,ylab='',xlab='',cex.axis=1.2,cex=1.2)
+  abline(lm(LPI_pop_info$min_time_for_power~LPI_pop_info$maximum_longevity_y),lwd=3)
+  mtext(text = 'maximum longevity (years)',side = 1,line = 3,cex = 1.2)
+  mtext('G',3,line=0,adj = 0.0,cex=1.2)
+
+  plot(LPI_pop_info$incubation_d,LPI_pop_info$min_time_for_power,pch=LPI_pop_info$shape,col=LPI_pop_info$color,las=1,ylab='',xlab='',cex.axis=1.2,cex=1.2)
+  abline(lm(LPI_pop_info$min_time_for_power~LPI_pop_info$incubation_d),lwd=3)
+  mtext(text = 'incubation (days)',side = 1,line = 3,cex = 1.2)
+  mtext('H',3,line=0,adj = 0.0,cex=1.2)
+
+#  boxplot(LPI_pop_info$min_time_for_power~LPI_pop_info$trophic,pch=as.numeric(as.factor(pop_info$class))+14,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
+#  stripchart(LPI_pop_info$min_time_for_power~LPI_pop_info$trophic, vertical = TRUE,method = "jitter", add = TRUE, pch = as.numeric(as.factor(pop_info$class))+14, col = rgb(0.5,0.5,0.5,0.5))
+#  mtext(text = 'trophic position',side = 1,line = 3,cex = 1.2)
+#  mtext(text = "(f)",side = 3,line=-1.5,adj = 0.98,cex=1.2)
+
+#plot(1, type="n", xlab="", ylab="", xaxt='n',yaxt='n',axes=FALSE)
+#legend('center',legend=names(table(pop_info$class)),col = 'grey',pch = as.numeric(names(table(as.numeric(as.factor(pop_info$class))+14))),bty ="n",cex=1.4)
+
+mtext(text = 'minimum time required',side = 2,line = 2,cex = 1.4,outer = TRUE)
+
+#######
+
+biological_correlates_poisson_model <-glm(min_time_for_power~  litter_or_clutch_size_n  + gen_len + adult_body_mass_g + maximum_longevity_y  +incubation_d + trophic,family = 'poisson',data=LPI_pop_info)
+R2_biological_correlates_poisson_model <- 100*(biological_correlates_poisson_model$null.deviance - biological_correlates_poisson_model$deviance)/(biological_correlates_poisson_model$null.deviance)
+
+all_correlates_poisson_model <-glm(min_time_for_power~ abs(overall_trend) + coefficient_variation + autocorrelation +  litter_or_clutch_size_n  + gen_len + adult_body_mass_g + maximum_longevity_y  +incubation_d + trophic,family = 'poisson',data=LPI_pop_info)
+R2_all_correlates_poisson_model <- 100*(all_correlates_poisson_model$null.deviance - all_correlates_poisson_model$deviance)/(all_correlates_poisson_model$null.deviance)
+
+```
+
+\clearpage
 
 ```{r, echo=F,fig.cap='Minimum time required to achieve 0.8 statistical power versus the minimum time required under IUCN criteria A2 to classify a species as vulnerable.
-Each point represents a single population, all of which saw declines of 30% or greater over a 10 year period.\\label{fig:IUCN_analysis}'}
+Each point represents a single population, all of which saw declines of 30% or greater over a 10 year period.
+The silhouettes highlight that species with longer generation times typically have larger discrepancies between $T_{min}$ and the minimum time required for IUCN assessments.\\label{fig:IUCN_analysis}'}
 
 declining_populations = subset(pop_info,pop_info$percent_change_10years>30 & pop_info$overall_trend<0)
+#declining_populations$color = as.numeric(as.factor(declining_populations$class))
+#declining_populations$color = viridis_pal(alpha=0.4,option='viridis')(5)[declining_populations$color]
+#declining_populations$shape = as.numeric(as.factor(declining_populations$color)) + 14
 
@@
 
-shapes = as.numeric(as.factor(declining_populations$class))+14
-shapes[shapes==17]= 18
 plot(declining_populations$IUCN, declining_populations$min_time_for_power,
-       ylim=c(0,35),xlim=c(0,85),pch=shapes,
-       col=rgb(0.5,0.5,0.5,0.5),cex.lab=1.3,las=1,cex=1.5,
+       ylim=c(0,35),xlim=c(0,85),col=declining_populations$color,
+       pch=declining_populations$shape,cex.lab=1.3,las=1,cex=1.5,
        xlab = 'minimum time required under IUCN criteria',ylab='')
@@
 #mtext(text = paste('n = ',nrow(declining_populations),sep=''),side = 3,line = -2,outer = F,at=80,cex=1.2)
-legend('bottomright',legend=names(table(declining_populations$class)),col = 'grey',pch = c(15,16,18),bty ="n",cex=1.2)
+legend('bottomright',legend=c('Aves','Actinopterygii','Mammalia'),pch=as.numeric(names(table(declining_populations$shape))),col=c(names(table(declining_populations$color))),bty ="n",cex=1.2,pt.cex = 1.4)
 
@@
 ```
-
-The IUCN Red List Categories and Criteria suggest, under Criterion A2, a species qualifies as vulnerable if it has experienced a 30% decline over 10 years, or 3 generations (whichever is longer) [@IUCN2012].
-I examined a subset of populations with observed declines of 30% or greater over 10 years, qualifying all of them as vulnerable.
-This resulted in n = `r nrow(declining_populations)` populations of fish, birds, and a single mammal.
-I then compared the minimum time required to achieve 0.8 statistical power ($T_{min}$) to the minimum time required under the IUCN criteria (Fig.
-\ref{fig:IUCN_analysis}).
-For populations below the identity line in figure \ref{fig:IUCN_analysis}, IUCN criteria would require more sampling compared to estimates for $T_{min}$.
-Further, populations above the identity line are cases where the IUCN criteria would classify a population as vulnerable despite not having sampled enough years to achieve high statistical power (Fig.
-\ref{fig:IUCN_analysis}).
-The silhouettes on figure \ref{fig:IUCN_analysis} highlight that species with longer generation times typically have larger discrepancies between $T_{min}$ and the minimum time required for IUCN assessments (Fig.
-\ref{fig:IUCN_correlates}).
-
-### Sensitivity analysis
-Lastly, I tested model sensitivity by using generalized additive models (GAMs) instead of simple linear regression.
-Again, I examined the minimum time required to estimate long-term population trends (Fig.
-\ref{fig:gam_example}).
-I found that although I obtain a similar distribution of minimum times required for GAMs, the minimum time required for GAMs is on average shorter than for linear regression (Fig.
-\ref{fig:min_time_dist_gam}).
-
-# Discussion
-
-I explored two approaches to estimate the minimum time series length required to address a particular question of interest.
-I asked, what is the minimum time series length required to determine long-term population trends using linear regression?
-This is one of the simplest questions one could ask of a time series.
-The simulation-based approach has been suggested by others, especially in situations more complicated than that suited for classic power analysis [@Gerrodette1987;@Johnson2015;@Bolker2008].
-My simulations support past work that longer time series are needed when the trend strength (i.e.
-rate of increase or decrease) is weak or when population variability is high [@Gerrodette1987].
-I also showed how the simulation model can be altered for a particular population (Fig.
-\ref{fig:shark_example}) or question (Figs.
-\ref{fig:min_time_dist_log_pop},\ref{fig:gam_example}).
-
-Here, I focus on an empirical approach to estimate the minimum time series length required to assess changes in abundance over time.
-I examined `r nrow(pop_info)` population time series (all longer than 35 years).
-I then subsampled each to determine the minimum time required to achieve a desired significance level and power for linear regression.
-Statistical power is important as it provides on information as to the necessary samples required to determine a significant trend [@Legg2006].
-I found that on average `r round(mean(pop_info$min_time_for_power),2)` years of continuous monitoring were typically necessary (Fig.
-\ref{fig:min_time_dist}b).
-However, the distribution of minimum time required was wide.
-This time-frame is in line with past work on a short-lived snail species [@Rueda-Cediel2015] and a long-lived whale species [@Gerber1999].
-@Hatch2003 used seabird monitoring data to estimate minimum sampling requirements.
-He found that the time required ranged from 11 to 69 years depending on species, trend strength, and study design.
-
-In line with theoretical predictions [@Rhodes2011], I also found $T_{min}$ was strongly correlated with the trend strength, variability in population size, and temporal autocorrelation (Fig.
-\ref{fig:correlates}).
-Contrary to my prior expectations, I also found that $T_{min}$ did not correlate with any life-history traits (Fig.
-\ref{fig:biological_correlates}).
-I initially hypothesized that species with longer lifespans or generation times may require a longer sampling period.
-This result could have been a result of at least two factors.
-First, the data I used may not include a diverse enough set of species with different life-history traits.
-Second, the question I poised, whether a population is increasing or decreasing, was specifically concerned with trends in population density over time.
-Therefore, life-history characteristics may be more important for other questions, like estimating species extinction risk [@Hutchings2012].
-For example, @Blanchard2007 used detailed simulations of spatially-distributed fisheries to compare survey designs.
-They found that statistical power depended on survey design, temperature preferences, and the degree of population patchiness.
-
-<!-- Similarly, @Bart2004 recommended a standard of 80% power to detect a 50% decline over 20 years in landbird populations.
-They found that only about 42% of populations in the Breeding Bird Survey met their criteria.
-@Seavy2007 asked whether statistical power was even a useful framework for assessing long-term population trends.
-They used 24 years of census data on Red-tailed Tropicbirds (\emph{Phaethon rubricauda}) in Hawaii and showed that to detect a 50% decline over 10 years almost always resulting in high statistical power (above 0.8).
-Therefore, they cautioned against only using power analyses to design monitoring schemes and instead argued for metrics that would increase precision.
-For example, @Seavy2007 suggest improving randomization, reducing bias, and increases detection probability when designing and evaluating monitoring programs.-->
-
-An important related question, is the optimal allocation of sampling effort in space versus time.
-In a theoretical investigation of this question, @Rhodes2011 found that the optimal allocation of sampling depended strongly on temporal and spatial autocorrelation.
-If spatial population dynamics were highly correlated, then it was better to sample more temporally, and vice versa.
-My work supports this idea as populations with strong temporal autocorrelation needed less years of sampling (Fig.
-\ref{fig:correlates}).
-@Morrison2008 also studied the optimal allocation of sampling effort in space versus time, but used emprical data from a long-term survey of the collared pika (\emph{Ochotona collaris}) found in the Yukon.
-They estimated long-term growth rates among three subpopulations over a 10-year period.
-They found that surveys less than 5 years may be misleading and that extrapolating from one population to another, even when nearby geographically, may be untenable.
-
-@Seavy2007 asked whether statistical power was even a useful framework for assessing long-term population trends.
-They used 24 years of census data on Red-tailed Tropicbirds (\emph{Phaethon rubricauda}) in Hawaii and showed that to detect a 50% decline over 10 years almost always resulting in high statistical power (above 0.8).
-Therefore, they cautioned against only using power analyses to design monitoring schemes and instead argued for metrics that would increase precision.
-For example, @Seavy2007 suggest improving randomization, reducing bias, and increases detection probability when designing and evaluating monitoring programs.
-I agree that power analyses should not be the only consideration when designing monitoring schemes.
-However, unlike @Seavy2007, my results indicate that longer than 10 years is often needed to achieve high statistical power.
-
-This paper also has practical implications for the IUCN Redlist criteria.
-IUCN criteria A2 suggests that species that have experienced 30% declines over 10 years (or three generations) should be listed as vulnerable [@IUCN2012].
-However, for the populations I examined, this criteria may be too simplistic (Fig.
-\ref{fig:IUCN_analysis}).
-For many populations, the IUCN criteria suggest more years than necessary are required to assess a population as vulnerable (points below diagonal line in Fig.
-\ref{fig:IUCN_analysis}).
-Conversely, for other populations the IUCN criteria suggest sampling times that are less than the minimum time required for statistical power.
-This suggests that the IUCN criteria are probably too simplistic as the minimum time required does not correlate with generation time (Fig.
-\ref{fig:biological_correlates}).
-
-The design of monitoring programs should include calculations of statistical power, the allocation of sampling in space versus time [@Rhodes2011], and metrics to increase precision.
-Ideally, a formal decision analysis to evaluate these different factors would be conducted to design or assess any monitoring program [@Hauser2006; @McDonald-Madden2010].
-This type of formal decision analysis would also include information on the costs of monitoring.
-These costs include the actual costs of sampling [@Brashares2005] and the ecological costs on inaction [@Thompson2000].
-
-<!--
-@Rueda-Cediel2015 found that 10-15 years was adequate to characterize population trends in a short lived animal. They used a matrix model of a particular species...
-
-@Gerber1999 investigated the minimum time series required to estimate population growth of an endagered, but recovering, species.
-They used a long-term census to retroactively determine the minimum time series required to assess threat status.
-They found that only XX years were needed, XX years before the delisting decision was actually made.
-This highlights the importance of estimating the minimunm time series required as an earlier decision would have saved time and money [@Gerber1999].
--->
-
-## Limitations
-
-This paper has some limitations in determining the minimum time series length required.
-First, $T_{min}$ is particular to the specific question of interest.
-An additional complication is that for the empirical approach, the subsampling of the full time series allows for estimates of power, but the individual subsamples are clearly not independent of one another.
-Further, estimates of $T_{min}$ depend on chosen values of $\alpha$ and $\beta$ (Fig.
-\ref{fig:min_time_vs_alpha_beta}).
-In an ideal setting, a specific population model would be parameterized for each population of interest [@McCain2016].
-Then, model simulations could be used to estimate the minimum time series required to address each specific question of interest.
-Clearly, this is not always practical, especially if conducting analyses for a wide array of species as I do here.
-In addition, the statistical models suggest that $T_{min}$ does not correlate with any life-history traits, at least for the question of linear regression (Fig.
-\ref{fig:biological_correlates}).
-Therefore, it is not possible to use these results to predict $T_{min}$ for another population, even if the population is of a species with a similar life-history to one in the database used.
-
-## Conclusions
-
-I used a database of `r nrow(pop_info)` populations to determine the minimum time series length required to detect population trends.
-This goes beyond previous work that either focused on theoretical investigations or a limited number of species.
-I show that to identify long-term changes in abundance, on average `r round(mean(pop_info$min_time_for_power),2)` years of continuous monitoring are often required (Fig.
-\ref{fig:min_time_dist}).
-However, there is wide distribution of estimated minimum times.
-Therefore, it is probably not wise to use a simple threshold number of years in monitoring design.
-
-In line with theoretical predictions [@Gerrodette1987], I also show that $T_{min}$ is strongly correlated with the long-term population trend (i.e.
-rate of increase), variability in population size, and the temporal autocorrelation (Fig.
-\ref{fig:correlates}).
-Contrary to my initial hypotheses, minimum time required did not correlate with generation time or any other life-history traits (Fig.
-\ref{fig:biological_correlates}).
-This result argues against overly simplified measures of minimum sampling time based on generation length (Fig.
-\ref{fig:IUCN_analysis}).
-
-My work implies that for many populations, short time series are probably not reliable for detecting population trends.
-This result highlights the importance of long-term monitoring programs.
-From both a scientific and management perspective estimates of $T_{min}$ are important.
-If a time series is too short, we lack the statistical power to reliably detect long-term population trends.
-In addition, a time series that is too long may be a poor use of already limited funds [@Gerber1999].
-Further, more data is not always best in situations where management actions need to be taken [@Martin2012; @Martin2017].
-When a population trend is detected, it may be too late for management action.
-In these situations, the precautionary principle may be more appropriate [@Thompson2000].
-Future work should examine other species, with a wider range of life-history characteristics.
-In addition, similar approaches can be used to determine the minimum time series length required to address additional questions of interest.
-
-<!--
-## Time series length recommendations from other papers
-
-- greater than 5 years [@Morrison2008]
-- more than 10-15 years [@Rueda-Cediel2015]
-- general advice (variability, correlations) [@Rhodes2011]
-- more than 10 years [@Gerber1999]
-- specific recommendations depend on amount of power and strength of trend (Bart 2004, Hatch 2003) [@Seavy2007]
-- paper on syncrony from James
--->
-
-# Supporting Information
-
-In the supporting material, I provide an expanded methods sections, additional figures, minimum time calculations for determining exponential growth, simulations with a more complicated population model, and the use of generalized additive models to identify population trends.
-All code and data can be found at https://github.com/erwhite1/time-series-project
-
-# Acknowledgements
-
-ERW was partially supported by a National Science Foundation Graduate Fellowship.
-I would like to thank members of the Ecological Theory group at the University of California, Davis for their insight.
-I would also like to thank T.
-Dallas and E.
-Malcolm-White <!--X anonymous reviewers -->for their helpful comments.
-
-# References
-
-<div id="refs"></div>
 

### 2018-02-19 writing/manuscript_draft_short_version.tex
first update of short format version of the article

--- 2018-02-19 writing/manuscript_draft.tex
@@
 \usepackage{amssymb,amsmath}
-\usepackage{setspace}  
 \usepackage{ifxetex,ifluatex}
@@
 \usepackage{float} \usepackage{lineno}
-\usepackage[firstinits=true,style=authoryear-comp,natbib=true,doi=false,isbn=false,url=false,uniquename=false,uniquelist=false,sorting=none]{biblatex}
+\usepackage{setspace}\doublespacing \usepackage[firstinits=true,style=authoryear-comp,natbib=true,doi=false,isbn=false,url=false,uniquename=false,uniquelist=false,sorting=none]{biblatex}
+
+\usepackage{titlesec}
+
+\titlespacing\section{0pt}{-5pt plus 1pt minus 1pt}{-5pt plus 1pt minus 1pt}
+\titlespacing\subsection{0pt}{-5pt plus 1pt minus 1pt}{-5pt plus 1pt minus 1pt}
+\titlespacing\subsubsection{0pt}{-5pt plus 1pt minus 1pt}{-5pt plus 1pt minus 1pt}
 
 \begin{document}
+
+Running head: Power to detect population trends {[}33 of 40
+characters{]}
 
@@
 \href{mailto:eawhite@ucdavis.edu}{\nolinkurl{eawhite@ucdavis.edu}} , 1
-480 203 7931 \vspace{3 mm}
+480 203 7931 \vspace{1 mm}
 
-Number of words (including 300 words per figure): 6,950 \vspace{1 mm}
+To be submitted to: \emph{Ecology} as a Statistical Report \vspace{1 mm}
 
-Number of figures: 5 \vspace{1 mm}
+Number of pages: 20\\
+\vspace{1 mm}
 
-Number of references: 34 \vspace{1 mm}
+Number of figures: 4 \vspace{1 mm}
 
-Supplementary material: 9 figures and 3 tables \vspace{3 mm}
+Number of references: 33 \vspace{1 mm}
+
+Supplementary material: 9 figures and 3 tables \vspace{2 mm}
 
 Data and code for all the figures and tables can be found at
-(\url{https://github.com/erwhite1/time-series-project}).
-
-To be submitted to: \emph{Biological Conservation} \vspace{3 mm}
+(\url{https://github.com/erwhite1/time-series-project}). \vspace{2 mm}
 
 Available as a preprint at: \url{https://peerj.com/preprints/3168/}
-\vspace{3 mm}
+\vspace{2 mm}
 
-Color does not need to be used for any figures in print \vspace{3 mm}
-
-\vspace{3 mm}
-
-\begin{spacing}{2}
+\pagebreak
 
@@
 to both understand population dynamics and to assess species extinction
-risk. Even though many time series may now be considered ``long-term''
-(e.g.~continuous plankton recorder, Giron-Nava et al. (2017)), most are
-still short. Time series are typically short for a variety of reasons
-(Field et al. 2007). They are often coupled with an experiment, which
-may only last a couple of years. In addition, short funding cycles make
-it difficult to examine populations over longer periods of time (Hughes
+risk. Some time series may now be considered ``long-term''
+(e.g.~continuous plankton recorder, Giron-Nava et al. (2017)), but most
+are still short. Time series are typically short due to short funding
+cycles and typical experimental time-frames (Field et al. 2007; Hughes
 et al. 2017).
@@
 important implications for both research and management (Nichols and
-Williams 2006). Scientists need to know the time series length required
-to address a specific question. A short time series may lead to wrong
-conclusions given large natural year-to-year variability (McCain,
-Szewczyk, and Knight 2016). Managers need to know when action is needed
-for a population. Therefore, managers must understand when population
-trend over time is actually meaningful. For example, the International
-Union for Conservation of Nature (IUCN) Red List Categories and Criteria
-suggest, under Criterion A2, a species qualifies as vulnerable if it has
+Williams 2006). A short time series may lead to wrong conclusions given
+large natural year-to-year variability (McCain, Szewczyk, and Knight
+2016). Managers need to know when action is needed for a population.
+Therefore, managers must understand when a population trend over time is
+actually meaningful. For example, the International Union for
+Conservation of Nature (IUCN) Red List Categories and Criteria suggest,
+under Criterion A2, a species qualifies as vulnerable if it has
 experienced a 30\% decline over 10 years, or three generations (IUCN
-2012). For both scientific and management questions, because sampling is
-typically expensive, we also do not want to sample for longer than is
-necessary. For example, Gerber, DeMaster, and Kareiva (1999)
-investigated the minimum time series required to estimate population
-growth of the endangered, but recovering, eastern North Pacific gray
-whale (\emph{Eschrichtius robustus}). They used a long-term census to
-retroactively determine the minimum time series required to assess
-threat status. They found that only 11 years were needed, eight years
-before the delisting decision was made. This highlights the importance
-of estimating the minimum time series required as an earlier decision
-would have saved time and money (Gerber, DeMaster, and Kareiva 1999).
-Further, waiting too long to make a decision can imperil a species where
-management action could have been taken earlier (Martin et al. 2012;
-Martin et al. 2017).
+2012). Sampling is typically expensive, therefore, we also do not want
+to sample for longer than is necessary. For example, Gerber, DeMaster,
+and Kareiva (1999) investigated the minimum time series required to
+estimate population growth of the endangered, but recovering, eastern
+North Pacific gray whale (\emph{Eschrichtius robustus}). They used a
+long-term census to retroactively determine the minimum time series
+required to assess threat status. They found that only 11 years were
+needed, eight years before the delisting decision was made. This
+highlights the importance of estimating the minimum time series required
+as an earlier decision would have saved time and money (Gerber,
+DeMaster, and Kareiva 1999). Further, waiting too long to decide an
+action can imperil a species where management action could have been
+taken earlier (Martin et al. 2012; Martin et al. 2017).
+
+Past work has investigated questions related to the minimum time series
+required to estimate trends in population size over time (Wagner,
+Vandergoot, and Tyson 2009; Giron-Nava et al. 2017). For example, Rhodes
+and Jonzen (2011) examined the optimal allocation of effort between
+spatial and temporal replicates. Using simple populations models, they
+found that the allocation of effort depends on environmental variation,
+spatial and temporal autocorrelation, and observer error. Rueda-Cediel
+et al. (2015) also used a modeling approach, but parameterized a model
+specific for a threatened snail, \emph{Tasmaphena lamproides}. They
+found that for this short-lived organism, 15 years was adequate to
+assess long-term trends in abundance. However, these studies, and other
+past work, have typically focused either on theoretical aspects of
+monitoring design or focused on only a few species.
+
+I estimate the minimum time series length required (\(T_{min}\)) to
+assess long-term changes in abundance via simple linear regression.
+Statistical power is not a new tool (Cohen 1992), but it is still
+under-appreciated in ecological research (Legg and Nagy 2006).
+Therefore, I only briefly describe the idea of statistical power for
+time series analyses using simulations. I then examine 822 population
+time series to estimate \(T_{min}\).
+
+\section{Methods}\label{methods}
+
+\subsection{Statistical power in time
+series}\label{statistical-power-in-time-series}
 
@@
 minus the probability of a type II error (\(\beta\)). The effect size is
-a measure of the difference between two groups. Prior to an experiment,
-one could set appropriate levels of power, significance level, and the
-effect size to estimate the sample size required for the experiment.
-This approach, however, is not straight-forward for a time series, or
-more complicated scenarios (P. C. Johnson et al. 2015), as data are
-clearly non-independent.
-
-For time series data, two general approaches to estimating sample size
-are appropriate. Simulations can be designed for a specific population
-and question (Bolker 2008; P. C. Johnson et al. 2015). Simple models can
-be simulated with parameter values corresponding to a population of
-interest (Gerrodette 1987). Statistical power is the proportion of
-simulations that meet some set of criteria. The specific criteria depend
-on the question at hand. For example, given a time series, when is the
-slope from linear regression significantly different from zero? In other
-words, when is the time series significantly increasing or decreasing?
-It is then possible to determine how power changes with a variable of
-interest. For example, time series can be simulated for different
-lengths of time. From these simulations, the minimum time series length
-required to meet certain levels of statistical significance and power is
-estimated (Bolker 2008).
-
-In addition to using simulations, empirical time series can also be
-used. Multiple replicates of similar populations are usually not
-available, but it is possible to subsample an empirical time series
-(Gerber, DeMaster, and Kareiva 1999; Brashares and Sam 2005). Subsamples
-of different lengths can then be evaluated to estimate the proportion of
-subsamples meeting some criteria, again a measure of statistical power.
-Similar to the simulation approach, this measures of power can be used
-to determine the minimum time series required for a particular question
-of interest.
-
-Past work has investigated questions related to the minimum time series
-required to estimate trends in population size over time (Wagner,
-Vandergoot, and Tyson 2009; Giron-Nava et al. 2017). For example, Rhodes
-and Jonzen (2011) examined the optimal allocation of effort between
-spatial and temporal replicates. Using simple populations models, they
-found that the allocation of effort depends on environmental variation,
-spatial and temporal autocorrelation, and observer error. Rueda-Cediel
-et al. (2015) also used a modeling approach, but parameterized a model
-specific for a threatened snail, \emph{Tasmaphena lamproides}. They
-found that for this short-lived organism, 15 years was adequate to
-assess long-term trends in abundance. However, these studies, and other
-past work, have typically focused either on theoretical aspects of
-monitoring design or focused on only a few species.
-
-I use both simulations and empirical time series to determine the
-minimum number of years required to address several questions. I
-estimate the minimum time series length required (\(T_{min}\)) to assess
-long-term changes in abundance via simple linear regression. First, I
-estimate \(T_{min}\) using a simulation approach. Then I examine 822
-population time series to estimate \(T_{min}\). In the supplementary
-material, I determine \(T_{min}\) for related ideas: using more
-complicated population models, varying statistical level and power, and
-the use of generalized additive models.
-
-\section{Methods}\label{methods}
+a estimate of the strength of a particular phenomenon. Prior to an
+experiment, one could set appropriate levels of power, significance
+level, and the effect size to estimate the sample size required for the
+experiment. This approach, however, is not straight-forward for a time
+series, or more complicated scenarios (P. C. Johnson et al. 2015), as
+data are clearly non-independent.
 
@@
 any experimental design too complicated for simple power analyses
-(Bolker 2008; P. C. Johnson et al. 2015). I only briefly discuss this
-approach as it has been described elsewhere. Essentially, a population
-model is simulated repetitively for a number of years. This approach
-requires us to determine values for model parameters (e.g.~population
-variability). As an example, we can take the following population model
-for population size \(N\) at time \(t\):
+(Bolker 2008; P. C. Johnson et al. 2015). Essentially, a population
+model, with a selected set of parameter values, is simulated
+repetitively for a number of years. As an example, we can take the
+following population model for population size \(N\) at time \(t\):
 
@@
 where \(\epsilon\) is a normally-distributed random noise term with mean
-\(\mu\) and standard deviation \(\sigma\). The rate of growth \(r\) is
-also the trend strength of the increase or decrease (i.e.~the rate of
-increase). It is important to note that any population model could be
-substituted for equation \ref{sim_model}, as in the supplementary
-material (Figs. \ref{fig:min_time_dist_log_pop},
-\ref{fig:shark_example}).
+\(\mu\) and standard deviation \(\sigma\). The rate of growth (\(r\)) is
+the trend strength of the increase or decrease (i.e.~the estimated slope
+from linear regression). It is important to note that any population
+model could be used here. As an example, in the supplementary material
+(Fig. \ref{fig:shark_example}), an age-structured model of population
+dynamics is used in place of equation \ref{sim_model} (White, Nagy, and
+Gruber 2014).
 
-Statistical power is then the proportion of simulations that meet some
-criteria. Here, our criteria is whether the slope parameter from linear
-regression is significant at the 0.05 threshold with statistical power
-of 0.8. Statistical power of 0.8 would indicate that, if there was a
-true trend in abundance, there would be a 0.8 probability of detecting
-the trend. Values of 0.05 for the significance level and 0.8 statistical
-power are historical and it is important to examine the effect of
+Statistical power is the proportion of simulations where the slope
+parameter from linear regression is significant at the 0.05 threshold.
+Statistical power of 0.8 would indicate that, if there was a true trend
+in abundance, there would be a 0.8 probability of detecting the trend.
+Values of 0.05 for the significance level and 0.8 statistical power are
+historical (Cohen 1992) and it is important to examine the effect of
 changing these values (Fig. \ref{fig:min_time_vs_alpha_beta}).
 
-In Fig. \ref{fig:theoretical_approach}a, a number of simulated time
-series are shown for a set number of time periods (\(t=40\)). It is
-clear that statistical power increases quickly with increases in length
-of time sampled (Fig. \ref{fig:theoretical_approach}b). Where power is
-greater than 0.8 (the dotted line), that is the minimum time required
-(\(T_{min}\)) to be confident in the detection of a long-term trend in
-abundance.
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-1-1.pdf}
-\caption{(a) Example of a simulated time series for 40 time periods. (b)
-Statistical power versus the simulated time series length. The
-horizontal, dashed line is the desired statistical power of 0.8. The
-vertical, dashed line is the minimum time required to achieve the
-desired statistical power. (c) Minimum time required (\(T_{min}\)) for
-simulations with different values of the trend strength (\(r\)). (d)
-Minimum time required for different levels of population variability
-(\(\sigma\)). In each case, the minimum time required is the minimum
-number of years to achieve 0.8 statistical power given a significance
-level of 0.05.\label{fig:theoretical_approach}}
-\end{figure}
+In Fig. \ref{fig:theoretical_approach}a, the model in equation
+\ref{sim_model} is simulated repetitively. Statistical power increases
+with increases in the length of time sampled (Fig.
+\ref{fig:theoretical_approach}b). Where power is greater than 0.8 (the
+dotted line), that is the minimum time required (\(T_{min}\)) to be
+confident in the detection of a long-term trend in abundance. As shown
+previously (Rhodes and Jonzen 2011; Rueda-Cediel et al. 2015),
+statistical power increases with larger trend strength and lower
+population variability (Fig. \ref{fig:theoretical_approach}c,d).
+Simulation approaches can be useful before designing a monitoring
+program or when a realistic model exists for the population in question.
+However, it is also informative to examine how statistical power changes
+with sampling effort in empirical time series.
 
@@
 
-I use a database of 2444 population time series compiled in (Keith et
-al. 2015); they compared the predictability of growth rates among
-populations. The data are originally from the Global Population Dynamics
+I use a database of 2444 population time series compiled by Keith et al.
+(2015). The data are originally from the Global Population Dynamics
 Database (NERC Centre for Population Biology 2010) and several other
@@
 data includes information on 477 vertebrate species with a focus on
-mammals, birds, and fish. The data also includes information on
-generation length and survey specifications. For each time series, I
-also calculate other variables of interest: coefficient of variance in
-population size, long-term trend in abundance (slope coefficient from
-simple linear regression), and temporal autocorrelation. All analyses
-were conducted in R (R Core Team 2016).
-
-For a subset of populations (\(n =\) 547), there is information on
-life-history traits available from another paper (Myhrvold et al. 2015),
-including body size and generation length. All 547 populations were
-birds. I examine how the minimum time required is related to these
-life-history traits (Fig. \ref{fig:biological_correlates}).
+mammals, birds, and fish.
 
@@
 possible contiguous subsample of length 35 (Gerber, DeMaster, and
-Kareiva 1999; Giron-Nava et al. 2017). Next, I run a linear regression
-for each subsampled time series. Then, I determine the proportion of
-subsamples of a particular length that have estimated slope coefficients
-which are statistically different from zero. I only look at the
-proportion of samples where the long-term, or ``true'', time series also
-has a significant slope. This proportion is a measure of statistical
-power. Lastly, I determine which subsample length is required to achieve
-a certain threshold of statistical power (0.8, Cohen (1992)). The
-minimum subsampled length that met these criteria is the minimum time
-series length required (\(T_{min}\)).
-
-In the supplementary material, I show how the same approach described
-here for more complicated population models. I also determine the
-minimum time required to estimate long-term trends according to
-generalized additive models, instead of the simple linear models used
-here (Fig. \ref{fig:gam_example}).
+Kareiva 1999; Brashares and Sam 2005; Giron-Nava et al. 2017). In line
+with the simulation approach, I determine the proportion of subsamples
+of a particular length that have estimated slope coefficients
+statistically different from zero. This proportion is a measure of
+statistical power. Lastly, I determine which subsample length is
+required to achieve a certain threshold of statistical power (0.8, Cohen
+(1992)). The minimum subsampled length that met these criteria is the
+minimum time series length required (\(T_{min}\)). All analyses were
+conducted in R (R Core Team 2016).
 
@@
 
-I determined the minimum time series length (\(T_{min}\)) required to
-address a particular question of interest. What is the minimum time
-series length required to determine, via linear regression, the
-long-term population trend? Here, the minimum time series length
-required had high enough statistical power (greater than 0.8) for a set
-significance level (\(\alpha\)) of 0.05. It is also possible to alter
-statistical power and \(\alpha\). Predictably, with increased
-statistical power or decreased \(\alpha\), \(T_{min}\) increased (Fig.
-\ref{fig:min_time_vs_alpha_beta}). I then estimated \(T_{min}\) using
-two approaches. I briefly describe results from the simulation approach
-and then discuss the empirical approach.
-
-\subsection{Simulation approach}\label{simulation-approach-1}
-
-I constructed a general population model where the trend strength
-(i.e.~slope coefficient) over time could be a model parameter. I then
-simulated time series of different lengths. From these simulations I
-determined the minimum time series length required to achieve a certain
-level of statistical power. In line with past work (Gerrodette 1987), I
-found the \(T_{min}\) increases (i.e.~more time is required) with
-decreases in trend strength and with increases in population variability
-(Figs. \ref{fig:theoretical_approach}c,d).
-
-I chose a simple model, but any other population model could be used
-(see example in Fig. \ref{fig:min_time_dist_log_pop}). Ideally, the
-specific model choice should be tailored to the population of interest.
-I explored how the simulation approach could be applied to more
-biologically-realistic population models (Fig. \ref{fig:shark_example}).
-Specifically, I determined the minimum time required to estimate
-long-term population trends using a stochastic, age-structured model of
-lemon shark population dynamics in the Bahamas (White, Nagy, and Gruber
-2014). I found that over 27 years of continuous monitoring were needed
-in this particular scenario (Fig. \ref{fig:shark_example}). Similar to
-the simulation approach described above, the minimum time required for
-the lemon shark population was strongly dependent on model parameters.
-
-\subsection{Empirical approach}\label{empirical-approach-1}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-4-1.pdf}
-\caption{(a) Distributions of the minimum time required for populations
-from four different biological classes. (b) Distribution of minimum time
-required for all populations regardless of biological class. The minimum
-time required calculation corresponds to a significance level of 0.05
-and statistical power of 0.8.\label{fig:min_time_dist}}
-\end{figure}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
-\caption{Minimum time required to estimate change in abundance
-correlated with (a) trend strength (absolute value of slope coefficient
-estimated from linear regression), (b) coefficient of variation in
-interannual population size, and (c) temporal lag-1
-autocorrelation.\label{fig:correlates}}
-\end{figure}
-
 I examined a database of 822 separate population time series
 representing 477 species. This database consisted of vertebrate species
-with a variety of life-history characteristics (Fig.
-\ref{fig:biological_correlates}). I limited analyses to populations with
-at least 35 years of continuous sampling. I then examined the minimum
-time required to estimate long-term trends via linear regression.
-
-Across all the populations, I found an average minimum time series
-length required (\(T_{min}\)) of 15.9 (SD=8.3), with a wide distribution
-(Fig. \ref{fig:min_time_dist}b). Estimates of \(T_{min}\) varied between
+with a variety of life-history characteristics. Across all the
+populations, I found an average minimum time series length required
+(\(T_{min}\)) of 15.9 (SD=8.3), with a wide distribution (Fig.
+\ref{fig:min_time_dist}b). Estimates of \(T_{min}\) varied between
 biological class (Fig. \ref{fig:min_time_dist}a). Ray-finned fish (class
@@
 
-\subsubsection{Corrrelates for minimum time
+\subsection{Corrrelates for minimum time
 required}\label{corrrelates-for-minimum-time-required}
@@
 regression), coefficient of variation in population size, and
-autocorrelation in population size (Fig. \ref{fig:correlates}). This is
-in line based on simulations here and those of others (Rhodes and Jonzen
-2011). Using a generalized linear model, with a Poisson error structure,
-all three of these explanatory variables were significant and had large
+autocorrelation in population size (Figs. \ref{fig:correlates}a-c). All
+three of these explanatory variables were significant and had large
 effect sizes (see Table A1). Combined, trend strength, coefficient of
@@
 
-For a subset of the populations I combined time series data with a data
-on life-history characteristics of amniotes (Myhrvold et al. 2015).
-There was life-history information available for 547 populations
-representing 315 different species, all of which were birds (Aves
-class).
+There was life-history information available (Myhrvold et al. 2015) for
+547 populations representing 315 different species, all of which were
+birds (Aves class). Some life-history traits were significant predictors
+for the minimum time required (Figs. \ref{fig:correlates}d-h, Tables
+A2,A3). However, even combined, all five of the life-history traits
+accounted for only 5.99\% of the explained deviance in minimum time
+series length required. In addition, when accounting for trend strength,
+coefficient of variation, and autocorrelation, no life-history traits
+were significant predictors of the minimum time required (Table A3).
 
-\begin{figure}[htbp]
-\centering
-\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-7-1.pdf}
-\caption{Mimimum time required versus (a) generation length (years), (b)
-litter size (n), (c) log adult body mass (grams), (d) maximum longevity
-(years), and (e) incubation (days). The lines in each plot represent the
-best fit line from linear regression.\label{fig:biological_correlates}}
-\end{figure}
-
-Some life-history traits were significant predictors for the minimum
-time required (Fig. \ref{fig:biological_correlates}, Tables A2,A3).
-However, none of these life-history traits explained a large part of the
-variation in minimum time required. In a generalized linear model, all
-of the life-history traits described in figure
-\ref{fig:biological_correlates} account for only 5.99\% of the explained
-deviance in minimum time series length required. In addition, when
-accounting for trend strength, coefficient of variation, and
-autocorrelation, no life-history traits were significant predictors of
-the minimum time required (Table A3).
-
-\subsubsection{Evalulating the IUCN
+\subsection{Evalulating the IUCN
 criteria}\label{evalulating-the-iucn-criteria}
 
-\begin{figure}[htbp]
-\centering
-\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-8-1.pdf}
-\caption{Minimum time required to achieve 0.8 statistical power versus
-the minimum time required under IUCN criteria A2 to classify a species
-as vulnerable. Each point represents a single population, all of which
-saw declines of 30\% or greater over a 10 year
-period.\label{fig:IUCN_analysis}}
-\end{figure}
-
-The IUCN Red List Categories and Criteria suggest, under Criterion A2, a
-species qualifies as vulnerable if it has experienced a 30\% decline
-over 10 years, or 3 generations (whichever is longer) (IUCN 2012). I
-examined a subset of populations with observed declines of 30\% or
-greater over 10 years, qualifying all of them as vulnerable. This
-resulted in n = 162 populations of fish, birds, and a single mammal. I
-then compared the minimum time required to achieve 0.8 statistical power
+I examined a subset of populations with observed declines of 30\% or
+greater over 10 years, qualifying all of them as vulnerable under IUCN
+Criterion A2 (IUCN 2012). This resulted in n = 162 populations. I then
+compared the minimum time required to achieve 0.8 statistical power
 (\(T_{min}\)) to the minimum time required under the IUCN criteria (Fig.
@@
 silhouettes on figure \ref{fig:IUCN_analysis} highlight that species
-with longer generation times typically have larger discrepancies between
-\(T_{min}\) and the minimum time required for IUCN assessments (Fig.
+with long generation times had larger discrepancies between \(T_{min}\)
+and the minimum time required for IUCN assessments (Fig.
 \ref{fig:IUCN_correlates}).
-
-\subsubsection{Sensitivity analysis}\label{sensitivity-analysis}
-
-Lastly, I tested model sensitivity by using generalized additive models
-(GAMs) instead of simple linear regression. Again, I examined the
-minimum time required to estimate long-term population trends (Fig.
-\ref{fig:gam_example}). I found that although I obtain a similar
-distribution of minimum times required for GAMs, the minimum time
-required for GAMs is on average shorter than for linear regression (Fig.
-\ref{fig:min_time_dist_gam}).
 
@@
 
-I explored two approaches to estimate the minimum time series length
-required to address a particular question of interest. I asked, what is
-the minimum time series length required to determine long-term
-population trends using linear regression? This is one of the simplest
-questions one could ask of a time series. The simulation-based approach
-has been suggested by others, especially in situations more complicated
-than that suited for classic power analysis (Gerrodette 1987; P. C.
-Johnson et al. 2015; Bolker 2008). My simulations support past work that
-longer time series are needed when the trend strength (i.e.~rate of
-increase or decrease) is weak or when population variability is high
-(Gerrodette 1987). I also showed how the simulation model can be altered
-for a particular population (Fig. \ref{fig:shark_example}) or question
-(Figs. \ref{fig:min_time_dist_log_pop},\ref{fig:gam_example}).
-
-Here, I focus on an empirical approach to estimate the minimum time
-series length required to assess changes in abundance over time. I
-examined 822 population time series (all longer than 35 years). I then
-subsampled each to determine the minimum time required to achieve a
-desired significance level and power for linear regression. Statistical
-power is important as it provides on information as to the necessary
-samples required to determine a significant trend (Legg and Nagy 2006).
-I found that on average 15.91 years of continuous monitoring were
-typically necessary (Fig. \ref{fig:min_time_dist}b). However, the
-distribution of minimum time required was wide. This time-frame is in
-line with past work on a short-lived snail species (Rueda-Cediel et al.
-2015) and a long-lived whale species (Gerber, DeMaster, and Kareiva
-1999). Hatch (2003) used seabird monitoring data to estimate minimum
-sampling requirements. He found that the time required ranged from 11 to
-69 years depending on species, trend strength, and study design.
+I examined 822 population time series. I then subsampled each to
+determine the minimum time required to achieve a desired significance
+level and power for linear regression. Statistical power is important as
+it provides information on the necessary number of samples required to
+determine a significant trend (Legg and Nagy 2006). I found that on
+average 15.91 years of continuous monitoring were typically necessary
+(Fig. \ref{fig:min_time_dist}b). However, the distribution of minimum
+time required was wide. Therefore, simplistic rules of thumb are
+probably not appropriate. This time-frame is in line with past work on a
+short-lived snail species (Rueda-Cediel et al. 2015) and a long-lived
+whale species (Gerber, DeMaster, and Kareiva 1999). Hatch (2003) used
+seabird monitoring data to estimate minimum sampling requirements. He
+found that the time required ranged from 11 to 69 years depending on
+species, trend strength, and study design.
 
@@
 that \(T_{min}\) did not correlate with any life-history traits (Fig.
-\ref{fig:biological_correlates}). I initially hypothesized that species
-with longer lifespans or generation times may require a longer sampling
+\ref{fig:correlates}d-h). I initially hypothesized that species with
+longer lifespans or generation times may require a longer sampling
 period. This result could have been a result of at least two factors.
 First, the data I used may not include a diverse enough set of species
-with different life-history traits. Second, the question I poised,
+with different life-history traits. Second, the question I posed,
 whether a population is increasing or decreasing, was specifically
-concerned with trends in population density over time. Therefore,
-life-history characteristics may be more important for other questions,
-like estimating species extinction risk (J. A. Hutchings et al. 2012).
-For example, Blanchard, Maxwell, and Jennings (2007) used detailed
-simulations of spatially-distributed fisheries to compare survey
-designs. They found that statistical power depended on survey design,
-temperature preferences, and the degree of population patchiness.
+concerned with population trends over time. Therefore, life-history
+characteristics may be more important for other questions more closely
+tied to species biology. For example, Blanchard, Maxwell, and Jennings
+(2007) used detailed simulations of spatially-distributed fisheries to
+compare survey designs. They found that statistical power depended on
+survey design, temperature preferences, and the degree of population
+patchiness.
 
-An important related question, is the optimal allocation of sampling
-effort in space versus time. In a theoretical investigation of this
-question, Rhodes and Jonzen (2011) found that the optimal allocation of
-sampling depended strongly on temporal and spatial autocorrelation. If
-spatial population dynamics were highly correlated, then it was better
-to sample more temporally, and vice versa. My work supports this idea as
+An important related idea is the optimal allocation of sampling effort
+in space versus time. In a theoretical investigation of this question,
+Rhodes and Jonzen (2011) found that the optimal allocation of sampling
+depended strongly on temporal and spatial autocorrelation. If spatial
+population dynamics were highly correlated, then it was better to sample
+more temporally, and vice versa. My work supports this idea as
 populations with strong temporal autocorrelation needed less years of
@@
 but used emprical data from a long-term survey of the collared pika
-(\emph{Ochotona collaris}) found in the Yukon. They estimated long-term
-growth rates among three subpopulations over a 10-year period. They
-found that surveys less than 5 years may be misleading and that
-extrapolating from one population to another, even when nearby
-geographically, may be untenable.
+(\emph{Ochotona collaris}) in the Yukon. They found that surveys less
+than 5 years may be misleading and that extrapolating from one
+population to another, even when nearby geographically, may be
+untenable.
 
@@
 (\emph{Phaethon rubricauda}) in Hawaii and showed that to detect a 50\%
-decline over 10 years almost always resulting in high statistical power
+decline over 10 years almost always resulted in high statistical power
 (above 0.8). Therefore, they cautioned against only using power analyses
 to design monitoring schemes and instead argued for metrics that would
-increase precision. For example, Seavy and Reynolds (2007) suggest
-improving randomization, reducing bias, and increases detection
-probability when designing and evaluating monitoring programs. I agree
-that power analyses should not be the only consideration when designing
-monitoring schemes. However, unlike Seavy and Reynolds (2007), my
-results indicate that longer than 10 years is often needed to achieve
-high statistical power.
+increase precision: improving randomization, reducing bias, and
+increasing detection probability. Power analyses should not be the only
+consideration when designing monitoring schemes. However, unlike Seavy
+and Reynolds (2007), the results here indicate that longer than 10 years
+is often needed to achieve high statistical power.
 
-This paper also has practical implications for the IUCN Redlist
-criteria. IUCN criteria A2 suggests that species that have experienced
-30\% declines over 10 years (or three generations) should be listed as
-vulnerable (IUCN 2012). However, for the populations I examined, this
-criteria may be too simplistic (Fig. \ref{fig:IUCN_analysis}). For many
-populations, the IUCN criteria suggest more years than necessary are
-required to assess a population as vulnerable (points below diagonal
-line in Fig. \ref{fig:IUCN_analysis}). Conversely, for other populations
-the IUCN criteria suggest sampling times that are less than the minimum
-time required for statistical power. This suggests that the IUCN
-criteria are probably too simplistic as the minimum time required does
-not correlate with generation time (Fig.
-\ref{fig:biological_correlates}).
+As a test of the consequences of not conducting power analyses, I
+compared the results here to the IUCN Redlist criteria. IUCN criteria A2
+suggests that species that have experienced 30\% declines over 10 years
+(or three generations) should be listed as vulnerable (IUCN 2012).
+However, for the populations I examined, this criteria may be too
+simplistic (Fig. \ref{fig:IUCN_analysis}). For many populations, the
+IUCN criteria suggest more years than necessary are required to assess a
+population as vulnerable (points below diagonal line in Fig.
+\ref{fig:IUCN_analysis}). Conversely, for other populations the IUCN
+criteria suggest sampling times that are less than the minimum time
+required for statistical power. This suggests that the IUCN criteria are
+probably too simplistic as the minimum time required does not correlate
+with generation time or other biological covariates (Fig.
+\ref{fig:correlates}d-h).
 
@@
 statistical power, the allocation of sampling in space versus time
-(Rhodes and Jonzen 2011), and metrics to increase precision. Ideally, a
-formal decision analysis to evaluate these different factors would be
-conducted to design or assess any monitoring program (Hauser, Pople, and
-Possingham 2006; McDonald-Madden et al. 2010). This type of formal
-decision analysis would also include information on the costs of
-monitoring. These costs include the actual costs of sampling (Brashares
-and Sam 2005) and the ecological costs on inaction (Thompson et al.
-2000).
+(Rhodes and Jonzen 2011), and metrics to increase precision (Seavy and
+Reynolds 2007). Ideally, a formal decision analysis to evaluate these
+different factors would be conducted to design or assess any monitoring
+program (Hauser, Pople, and Possingham 2006; McDonald-Madden et al.
+2010). This type of formal decision analysis would also include
+information on the costs of monitoring. These costs include the actual
+costs of sampling (Brashares and Sam 2005) and the ecological costs on
+inaction (Thompson et al. 2000).
 
@@
 estimates of power, but the individual subsamples are clearly not
-independent of one another. Further, estimates of \(T_{min}\) depend on
-chosen values of \(\alpha\) and \(\beta\) (Fig.
-\ref{fig:min_time_vs_alpha_beta}). In an ideal setting, a specific
-population model would be parameterized for each population of interest
-(McCain, Szewczyk, and Knight 2016). Then, model simulations could be
-used to estimate the minimum time series required to address each
-specific question of interest. Clearly, this is not always practical,
-especially if conducting analyses for a wide array of species as I do
-here. In addition, the statistical models suggest that \(T_{min}\) does
-not correlate with any life-history traits, at least for the question of
-linear regression (Fig. \ref{fig:biological_correlates}). Therefore, it
-is not possible to use these results to predict \(T_{min}\) for another
-population, even if the population is of a species with a similar
-life-history to one in the database used.
+independent of one another. In an ideal setting, a specific population
+model would be parameterized for each population of interest. Then,
+model simulations could be used to estimate the minimum time series
+required to address each specific question of interest. Clearly, this is
+not always practical, especially if conducting analyses for a wide array
+of species.
 
@@
 
-I used a database of 822 populations to determine the minimum time
-series length required to detect population trends. This goes beyond
-previous work that either focused on theoretical investigations or a
-limited number of species. I show that to identify long-term changes in
-abundance, on average 15.91 years of continuous monitoring are often
-required (Fig. \ref{fig:min_time_dist}). However, there is wide
-distribution of estimated minimum times. Therefore, it is probably not
-wise to use a simple threshold number of years in monitoring design.
-
-In line with theoretical predictions (Gerrodette 1987), I also show that
-\(T_{min}\) is strongly correlated with the long-term population trend
-(i.e.~rate of increase), variability in population size, and the
-temporal autocorrelation (Fig. \ref{fig:correlates}). Contrary to my
-initial hypotheses, minimum time required did not correlate with
+Power analyses are not a novel aspect of ecological research (Legg and
+Nagy 2006). However, power analyses are still underutilized, especially
+in the context of time series analyses. I used a database of 822
+populations to determine the minimum time series length required to
+detect population trends. This goes beyond previous work that either
+focused on theoretical investigations or a limited number of species. I
+show that to identify long-term changes in abundance, on average 15.91
+years of continuous monitoring are often required (Fig.
+\ref{fig:min_time_dist}). However, there is wide distribution of
+estimated minimum times. Therefore, it is probably not wise to use a
+simple threshold number of years in monitoring design. Further, contrary
+to my initial hypotheses, minimum time required did not correlate with
 generation time or any other life-history traits (Fig.
-\ref{fig:biological_correlates}). This result argues against overly
-simplified measures of minimum sampling time based on generation length
-(Fig. \ref{fig:IUCN_analysis}).
+\ref{fig:correlates}d-h). This result argues against overly simplified
+measures of minimum sampling time based on generation length or other
+life-history traits (Fig. \ref{fig:IUCN_analysis}).
 
@@
 highlights the importance of long-term monitoring programs. From both a
-scientific and management perspective estimates of \(T_{min}\) are
+scientific and management perspective, estimates of \(T_{min}\) are
 important. If a time series is too short, we lack the statistical power
@@
 Fellowship. I would like to thank members of the Ecological Theory group
-at the University of California, Davis for their insight. I would also
-like to thank T. Dallas and E. Malcolm-White for their helpful comments.
+at the University of California, Davis for their insight.
 
@@
 doi:\href{https://doi.org/10.1093/biosci/biw185}{10.1093/biosci/biw185}.
-
-\hypertarget{ref-Hutchings2012}{}
-Hutchings, Jeffrey A, Ransom A Myers, Verónica B García, Luis O
-Lucifora, and Anna Kuparinen. 2012. ``Life-history correlates of
-extinction risk and recovery potential.'' \emph{Ecological Applications}
-22 (4): 1061--7.
 
@@
 
-\end{spacing}
+\clearpage
+
+\pagebreak
+
+\section{Figure captions}\label{figure-captions}
+
+Figure 1: (a) Example of a simulated time series for 40 time periods.
+(b) Statistical power versus the simulated time series length. The
+horizontal, dashed line is the desired statistical power of 0.8. The
+vertical, dashed line is the minimum time required to achieve the
+desired statistical power. (c) Minimum time required (\(T_{min}\)) for
+simulations with different values of the trend strength (\(r\)) and
+\(\sigma = 5.0\). (d) Minimum time required for different levels of
+population variability (\(\sigma\)) and \(r=1.5\). In each case, the
+minimum time required is the minimum number of years to achieve 0.8
+statistical power given a significance level of 0.05..
+
+Figure 2: (a) Distributions of the minimum time required for populations
+from four different biological classes. (b) Distribution of minimum time
+required for all populations regardless of biological class. The minimum
+time required calculation corresponds to a significance level of 0.05
+and statistical power of 0.8.
+
+Figure 3: Minimum time required to estimate change in abundance
+correlated with (a) trend strength (absolute value of slope coefficient
+estimated from linear regression), (b) coefficient of variation in
+interannual population size, (c) temporal lag-1 autocorrelation, (d)
+generation length (years), (e) litter size (n), (f) log adult body mass
+(grams), (g) maximum longevity (years), and (h) incubation (days). The
+lines in each plot represent the best fit line from linear regression.
+
+Figure 4: Minimum time required to achieve 0.8 statistical power versus
+the minimum time required under IUCN criteria A2 to classify a species
+as vulnerable. Each point represents a single population, all of which
+saw declines of 30\% or greater over a 10 year period. The silhouettes
+highlight that species with longer generation times typically have
+larger discrepancies between \(T_{min}\) and the minimum time required
+for IUCN assessments.
+
+\pagebreak
+
+\begin{figure}
+\centering
+\includegraphics{manuscript_draft_short_version_files/figure-latex/unnamed-chunk-5-1.pdf}
+\caption{(a) Example of a simulated time series for 40 time periods. (b)
+Statistical power versus the simulated time series length. The
+horizontal, dashed line is the desired statistical power of 0.8. The
+vertical, dashed line is the minimum time required to achieve the
+desired statistical power. (c) Minimum time required (\(T_{min}\)) for
+simulations with different values of the trend strength (\(r\)) and
+\(\sigma = 5.0\). (d) Minimum time required for different levels of
+population variability (\(\sigma\)) and \(r=1.5\). In each case, the
+minimum time required is the minimum number of years to achieve 0.8
+statistical power given a significance level of
+0.05.\label{fig:theoretical_approach}}
+\end{figure}
+
+\pagebreak
+
+\begin{figure}
+\centering
+\includegraphics{manuscript_draft_short_version_files/figure-latex/unnamed-chunk-6-1.pdf}
+\caption{(a) Distributions of the minimum time required for populations
+from four different biological classes. (b) Distribution of minimum time
+required for all populations regardless of biological class. The minimum
+time required calculation corresponds to a significance level of 0.05
+and statistical power of 0.8.\label{fig:min_time_dist}}
+\end{figure}
+
+\clearpage
+
+\begin{figure}
+\centering
+\includegraphics{manuscript_draft_short_version_files/figure-latex/unnamed-chunk-9-1.pdf}
+\caption{Minimum time required to estimate change in abundance
+correlated with (a) trend strength (absolute value of slope coefficient
+estimated from linear regression), (b) coefficient of variation in
+interannual population size, (c) temporal lag-1 autocorrelation, (d)
+generation length (years), (e) litter size (n), (f) log adult body mass
+(grams), (g) maximum longevity (years), and (h) incubation (days). The
+lines in each plot represent the best fit line from linear
+regression.\label{fig:correlates}}
+\end{figure}
+
+\clearpage
+
+\begin{figure}
+\centering
+\includegraphics{manuscript_draft_short_version_files/figure-latex/unnamed-chunk-10-1.pdf}
+\caption{Minimum time required to achieve 0.8 statistical power versus
+the minimum time required under IUCN criteria A2 to classify a species
+as vulnerable. Each point represents a single population, all of which
+saw declines of 30\% or greater over a 10 year period. The silhouettes
+highlight that species with longer generation times typically have
+larger discrepancies between \(T_{min}\) and the minimum time required
+for IUCN assessments.\label{fig:IUCN_analysis}}
+\end{figure}
+
+\clearpage
+
+\section{Supplementary material}\label{supplementary-material}
+
+\setcounter{figure}{0} \renewcommand{\thefigure}{A\arabic{figure}}
+
+\begin{figure}[htbp]
+\centering
+\caption{(a) Population size of Bigeye tuna (\emph{Thunnus obesus}) over
+time. The line is the best fit line from linear regression. (b)
+Statistical power for different subsets of the time series in panel
+a.\label{fig:empirical_approach_example}}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\caption{Output of generalized linear model with a Poisson error
+structure for predicting the minimum time required with explanatory
+variables of the absolute value of the slope coefficient (or trend
+strength), temporal autocorrelation, and variability in population
+size.\label{fig:poisson_model}}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\caption{(a) Minimum time required to estimate change in abundance by biological class, (b) long-term trend (estimated slope coefficient) by class, (c) coefficient of variation in population size by class, and (d) temporal autocorrelation by class.\label{fig:class}}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\caption{Minimum time required to assess long-term trends in abundance
+for values of statistical significance (\(\alpha\)) and power
+(\(1-\beta\)).\label{fig:min_time_vs_alpha_beta}}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\caption{The difference between minimum time estimates is the minimum time required to achieve 0.8 statistical power versus the minimum time required under IUCN criteria A2 to classify a species as vulnerable.
+Each point represents a single population, all of which saw declines of 30\% or greater over a 10 year period.
+(a) Difference between minimum time estimates versus the coefficient of variation in population size.
+(b) Difference between minimum time estimates versus the generation length in years.\label{fig:IUCN_correlates}}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\caption{Distribution of the minimum time required in order to detect a
+significant trend (at the 0.05 level) in log(abundance) given power of
+0.8.\label{fig:min_time_dist_log_pop}}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\caption{Statistical power for different length of time series
+simulations for a lemon shark population in Bimini,
+Bahamas.\label{fig:shark_example}}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\caption{(a) Time series for Bigeye tuna (\emph{Thunnus obesus}) with
+corresponding fitted GAM model in red and (b) statistical power as a
+function of the number of years sampled. The horizontal line at 0.8
+indicates the minimum threshold for statistical power and the vertical
+line denotes the minimum time required to achieve 0.8 statistical
+power.\label{fig:gam_example}}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\caption{Distribution of the minimum time required in order to detect a
+significant trend (at the 0.05 level) in abundance according to a GAM
+model given statistical power of 0.8. The smoothing parameter was set to
+3 for each population.\label{fig:min_time_dist_gam}}
+\end{figure}
+
+\vspace{1 cm}
+
+Table A1: Output of generalized linear model to examine time series
+characteristics as correlates of the minimum time required for
+determining long-term population trends.
+
+\vspace{1 cm}
+
+Table A2: Output of generalized linear model to examine life-history
+trait correlates of the minimum time required for determine long-term
+population trends.
+
+\vspace{1 cm}
+
+Table A3: Output of generalized linear model to examine both time series
+characteristics and life-history trait correlates of the minimum time
+required for determine long-term population trends.
+
 \end{document}

### 2018-02-19 writing/supp_mat_draft.Rmd
first update of short format version of the article

@@
 
-Easton R. White$^{*}$
+Easton R. White
 \vspace{3 mm}
 
-Address: \\ \emph{Center for Population Biology, University of California-Davis, Davis, California 95616 USA}
+Center for Population Biology, University of California, Davis, California 95616 USA
 
-*Corresponding author: eawhite@ucdavis.edu
+eawhite@ucdavis.edu
 
@@
 
-Here, we illustrate how we performed the subsampling and power calculations for a specific population.
-We provide more detail here than in the main manuscript.
-As an example, we examine a 35-year time series of Bigeye tuna (\emph{Thunnus obesus}), one species in the Global Population Dynamics Database [@GPDD2010].
+Here, I illustrate how we performed the subsampling and power calculations for a specific population.
+As an example, I examine a 35-year time series of Bigeye tuna (\emph{Thunnus obesus}), one species in the Global Population Dynamics Database [@GPDD2010].
 Simple linear regression indicates a significant decrease for this population with an estimated slope coefficient of `r as.numeric(coef(model_example)[2])`.
-We assume that this significant increase over 35 years is in fact the ``true trend".
+I assume that this significant increase over 35 years is in fact the ``true trend".
 In statistical jargon, the 35-year trend is an effect that is actually present; we can reject the null hypothesis of no trend.
@@
 # could insert all possible regressions of length n
-source("../scripts/calculate_power_metric.R",chdir=T)
+source("../scripts/calculate_power_metric.R")
     power_example=calculate_power_for_subsamples(pop$popvalue,0.05)
@@
 
-We then perform a subsampling routine to estimate the minimum time required $T_{min}$ (similar to @Gerber1999). This is the same routine we used for the results in the main manuscript.
+I then perform a subsampling routine to estimate the minimum time required $T_{min}$ (similar to @Gerber1999). This is the same routine I used for the results in the main manuscript.
 \begin{enumerate}
@@
 
+\pagebreak
+
 # Additional results from the main manuscript
 
-## Predictors of minimum time required
+## Correlates of minimum time required
 
-```{r poisson_model, echo=F,include=T,fig.cap='Output of generalized linear model with a Poisson error structure for predicting the minimum time required with explanatory variables of the absolute value of the slope coefficient (or trend strength), temporal autocorrelation, and variability in population size.\\label{fig:poisson_model}'}
+```{r poisson_model, echo=F,fig.cap='Output of generalized linear model with a Poisson error structure for predicting the minimum time required with explanatory variables of the absolute value of the slope coefficient (or trend strength), temporal autocorrelation, and coefficient of variation of population size.\\label{fig:poisson_model}'}
 
 load("../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
+load("../cleaned-data/combined_databases_with_min_time_linear_regression.Rdata")
 
@@
 pop_info$abs_overall_trend = abs(pop_info$overall_trend)
-poisson_model <-glm(min_time_for_power~ abs_overall_trend + autocorrelation + variance,family = 'poisson',data=pop_info)
+poisson_model <-glm(min_time_for_power~ abs_overall_trend + autocorrelation + coefficient_variation,family = 'poisson',data=pop_info)
 
@@
 #R^2 calculation (pg. 218 Zuur et al. 2009 mixed effects book)
-R2_fullmodel <- 100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance)
 
-poisson_model_no_variance <-glm(min_time_for_power~ abs_overall_trend + autocorrelation,family = 'poisson',data=pop_info)
-R2_no_variance <- 100*(poisson_model_no_variance$null.deviance - poisson_model_no_variance$deviance)/(poisson_model_no_variance$null.deviance)
+time_series_correlates_poisson_model <-glm(min_time_for_power~ abs_overall_trend + autocorrelation + coefficient_variation,family = 'poisson',data=pop_info)
+R2_time_series_correlates <- 100*(time_series_correlates_poisson_model$null.deviance - time_series_correlates_poisson_model$deviance)/(time_series_correlates_poisson_model$null.deviance)
 
-poisson_model_no_trend <-glm(min_time_for_power~ autocorrelation + variance,family = 'poisson',data=pop_info)
-R2_no_trend <- 100*(poisson_model_no_trend$null.deviance - poisson_model_no_trend$deviance)/(poisson_model_no_trend$null.deviance)
+biological_correlates_poisson_model <-glm(min_time_for_power~  litter_or_clutch_size_n  + gen_len + adult_body_mass_g + maximum_longevity_y  +incubation_d + trophic,family = 'poisson',data=LPI_pop_info)
+R2_biological_correlates_poisson_model <- 100*(biological_correlates_poisson_model$null.deviance - biological_correlates_poisson_model$deviance)/(biological_correlates_poisson_model$null.deviance)
+
+all_correlates_poisson_model <-glm(min_time_for_power~ abs(overall_trend) + coefficient_variation + autocorrelation +  litter_or_clutch_size_n  + gen_len + adult_body_mass_g + maximum_longevity_y  +incubation_d + trophic,family = 'poisson',data=LPI_pop_info)
+R2_all_correlates_poisson_model <- 100*(all_correlates_poisson_model$null.deviance - all_correlates_poisson_model$deviance)/(all_correlates_poisson_model$null.deviance)
 
@@
 
-In the main text, we explained how the minimum time required strongly correlated with the trend strength, temporal autocorrelation, and variance in population size.
-Here, we use a generalized linear model framework with a Poisson error structure to determine explanatory variables of the minimum time required.
-We use the same `r nrow(pop_info)` populations as in the main text.
-In figure \ref{fig:poisson_model}, we show a set of residual plots for the regression.
-We then show the coefficient estimates and levels of significance in table \ref{table:model_output}.
+In the main text, I explained how the minimum time required strongly correlated with the trend strength, temporal autocorrelation, and coefficient of variation in population size.
+Here, I use a generalized linear model framework with a Poisson error structure to determine explanatory variables of the minimum time required.
+I use the same `r nrow(pop_info)` populations as in the main text.
+In figure \ref{fig:poisson_model}, I show a set of residual plots for the regression analysis.
+I then show the coefficient estimates and levels of significance in table \ref{table:model_summary_time_series_correlates}.
 
 ```{r,echo=F,message=F,warning=F}
-
 require(knitr)
-knitr::kable(summary(poisson_model)$coef,caption = 'Output of generalized linear model to examine predictors of the minimum time required for determing long-term population trends.\\label{table:model_output}')
-
+knitr::kable(summary(time_series_correlates_poisson_model)$coef,caption = 'Output of generalized linear model to examine time series characteristics as correlates of the minimum time required for determining long-term population trends.\\label{table:model_summary_time_series_correlates}')
 ```
 
-Our model with trend strength, autocorrelation, variance, and generation length accounted for `r round(R2_fullmodel,2)`% of the variation in the minimum time required (Table \ref{table:model_output}).
-However, we also found trend strength and variance to be strongly correlated with one another.
-Therefore, we ran two additional models with either trend strength or variance, but not both together.
-This resulted in lower explained deviance (analogue to $R^2$) of `r round(R2_no_variance,2)`% and `r round(R2_no_trend,2)`%, respectively.
+```{r,echo=F,message=F,warning=F}
+require(knitr)
+knitr::kable(summary(biological_correlates_poisson_model)$coef,caption = 'Output of generalized linear model to examine life-history trait correlates of the minimum time required for determining long-term population trends.\\label{table:model_summary_biological_correlates}')
+```
+
+```{r,echo=F,message=F,warning=F}
+require(knitr)
+knitr::kable(summary(all_correlates_poisson_model)$coef,caption = 'Output of generalized linear model to examine both time series characteristics and life-history trait correlates of the minimum time required for determining long-term population trends.\\label{table:model_summary_all_correlates}')
+```
+
+The model with trend strength, autocorrelation, and the coefficient of variation in population size `r round(R2_time_series_correlates,2)`% of the variation in the minimum time required (Table \ref{table:model_summary_time_series_correlates}).
+
+For a subset of populations, I examined life-history correlates as potential explanatory variables for the minimum time required.
+Table \ref{table:model_summary_biological_correlates} and \ref{table:model_summary_all_correlates} summarize results for a model with only life-history traits and an additional model with life-history traits and time series characteristics combined, respectively.
+The model with all life-history traits explained only `r round(R2_biological_correlates_poisson_model,2)`% of the variation in the minimum time required (Table \ref{table:model_summary_biological_correlates}).
+The model with all life-history traits and time series characteristics explained `r round(R2_all_correlates_poisson_model,2)`% of the variation in the minimum time required (Table \ref{table:model_summary_all_correlates}).
+When accounting for characteristics of the time series, no life-history variables were significant.
 
@@
 
+<!--
 ## Minimum time required and biological correlates
@@
 
-In the main manuscript, we examined the minimum time required to detect a significant trend in abundance over time using linear regression.
+In the main manuscript, I examined the minimum time required to detect a significant trend in abundance over time using linear regression.
 As detailed in the main manuscript the minimum time required was around 15, but there was a wide distribution.
-Therefore, we were interested in potential explanatory variables of the minimum time required.
-In the main manuscript, we examined characteristics of the time series itself, like variability, autocorrelation, and the trend in abundance over time.
-Here, we combined our time series data with a data on life history characteristics of amniotes from @Myhrvold2015.
+Therefore, I were interested in potential explanatory variables of the minimum time required.
+In the main manuscript, I examined characteristics of the time series itself, like variability, autocorrelation, and the trend in abundance over time.
+Here, I combined our time series data with a data on life history characteristics of amniotes from @Myhrvold2015.
 There was life history information available for `r nrow(LPI_pop_info)` populations representing `r length(table(LPI_pop_info$Binomial))` different species, all of which were birds (Aves class).
@@
 
-We then correlated minimum time required for each population with its corresponding life history characteristics.
-In figure \ref{fig:biological_correlates} we examined minimum time required versus generation length (years), litter size (n), adult body mass (grams), maximum longevity (years), egg mass (grams), and incubation (days).
+I then correlated minimum time required for each population with its corresponding life history characteristics.
+In figure \ref{fig:biological_correlates} I examined minimum time required versus generation length (years), litter size (n), adult body mass (grams), maximum longevity (years), egg mass (grams), and incubation (days).
 None of these variables had much explanatory power in accounting for the variance in the minimum time required.
 We ran a generalized linear model (with a Poisson error structure) and found that these six variables only accounted for `r round(100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance),2)`% of the explained deviance in minimum time required.
+-->
 
@@
 \ref{fig:class}c).
+In addition, populations in the Actinopterygii class generally had weaker trends in abundance (i.e.
+smaller slope coefficients) compared to the Aves and Mammalia populations.
 
-```{r,echo=F,fig.cap='(a) Minimum time required to estimate change in abundance for species class, (b) long-term trend (estimated slope coefficient) by species class, (c) interannual variability in population size by species class, and (d) temporal autocorrelation by species class.\\label{fig:class}',fig.height=6}
+```{r,echo=F,fig.cap='(a) Minimum time required to estimate change in abundance by biological class, (b) long-term trend (estimated slope coefficient) by class, (c) coefficient of variation in population size by class, and (d) temporal autocorrelation by class.\\label{fig:class}',fig.height=6}
 
@@
 
-boxplot(pop_info$min_time_for_power~pop_info$class, lwd = 1, ylab = 'minimum time required',col='white',cex.lab=1.4,cex.axis=1.2,las=2,xaxt="n")
+boxplot(pop_info$min_time_for_power~pop_info$class, lwd = 1, ylab = 'minimum time required',col='white',cex.lab=1.4,cex.axis=1.2,las=2,xaxt="n",outline=FALSE)
 mtext(text = '(a)',side = 3,line = -1.2,adj=0.95,font = 2,cex=1.2)
@@
 
-boxplot(abs(pop_info$overall_trend)~pop_info$class, lwd = 1, ylab = ' ',col='white',cex.lab=1.4,cex.axis=1.2,las=2,xaxt="n")
+boxplot(abs(pop_info$overall_trend)~pop_info$class, lwd = 1, ylab = ' ',col='white',cex.lab=1.4,cex.axis=1.2,las=2,xaxt="n",outline=FALSE)
 stripchart(abs(pop_info$overall_trend)~pop_info$class, vertical = TRUE,method = "jitter", add = TRUE, pch = 20, col = rgb(0.5,0.5,0.5,0.5))
@@
 
-boxplot(pop_info$variance~pop_info$class, lwd = 1, ylab = 'population variance',col='white',cex.lab=1.4,cex.axis=1.2,las=2)
-stripchart(pop_info$variance~pop_info$class, vertical = TRUE,method = "jitter", add = TRUE, pch = 20, col = rgb(0.5,0.5,0.5,0.5))
+boxplot(pop_info$coefficient_variation~pop_info$class, lwd = 1, ylab = 'coefficient of variation',col='white',cex.lab=1.4,cex.axis=1.2,las=2,outline=FALSE)
+stripchart(pop_info$coefficient_variation~pop_info$class, vertical = TRUE,method = "jitter", add = TRUE, pch = 20, col = rgb(0.5,0.5,0.5,0.5))
 mtext(text = '(c)',side = 3,line = -1.2,adj=0.95,font = 2,cex=1.2)
 
-boxplot(pop_info$autocorrelation~pop_info$class, lwd = 1, ylab = 'autocorrelation',col='white',cex.lab=1.4,cex.axis=1.2,las=2,ylim=c(-0.2,1.2))
+boxplot(pop_info$autocorrelation~pop_info$class, lwd = 1, ylab = 'autocorrelation',col='white',cex.lab=1.4,cex.axis=1.2,las=2,ylim=c(-0.2,1.2),outline=FALSE)
 stripchart(pop_info$autocorrelation~pop_info$class, vertical = TRUE,method = "jitter", add = TRUE, pch = 20, col = rgb(0.5,0.5,0.5,0.5))
@@
 
+## Additional results for IUCN analysis
+
+In the main text, I presented results detailing the difference between the minimum time required to achieve a high level of statistical power and the number of years suggested by IUCN based on generation length to assess a species as vulnerable.
+There were deviations between the two approaches.
+These deviations were mostly explained by the differences in inter-annual population variability (coefficient of variation) and generation time (Fig.
+\ref{fig:IUCN_correlates}).
+
+```{r, echo=F,message=F,warning=F,fig.cap='The difference between minimum time estimates is the minimum time required to achieve 0.8 statistical power versus the minimum time required under IUCN criteria A2 to classify a species as vulnerable.
+Each point represents a single population, all of which saw declines of 30% or greater over a 10 year period.
+(a) Difference between minimum time estimates versus the coefficient of variation in population size.
+(b) Difference between minimum time estimates versus the generation length in years.\\label{fig:IUCN_correlates}',fig.height=6}
+
+library(viridis)
+pop_info$color = as.numeric(as.factor(pop_info$class))
+pop_info$color = viridis_pal(alpha=0.3,option='viridis')(5)[pop_info$color]
+pop_info$shape = as.numeric(as.factor(pop_info$color)) + 14
+
+declining_populations = subset(pop_info,pop_info$percent_change_10years>30 & pop_info$overall_trend<0)
+
+par(mfrow=c(2,1),oma=c(0,5,0,0),mar=c(5,0.5,0.5,0.5))
+error=(declining_populations$IUCN - declining_populations$min_time_for_power)
+plot(declining_populations$coefficient_variation,error,
+     ylab = '', las=1,
+     xlab = 'coefficient of variation', col=declining_populations$color,
+       pch=declining_populations$shape)
+abline(lm(error~declining_populations$coefficient_variation),lwd=2)
+mtext(text = '(a)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
+
+plot(declining_populations$gen_len,error,
+     ylab = '', las=1,
+     xlab = 'generation length (years)', col=declining_populations$color,
+       pch=declining_populations$shape)
+abline(lm(error~declining_populations$gen_len),lwd=2)
+mtext(text = '(b)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
+
+mtext(text = 'Difference between minimum time estimates',side = 2,
+      outer = T,at=0.6,line = 2,cex = 1.2)
+
+legend('bottomright',legend=c('Aves','Actinopterygii','Mammalia'),pch=as.numeric(names(table(declining_populations$shape))),col=c(names(table(declining_populations$color))),bty ="n",cex=1.2,pt.cex = 1.4)
+```
+
+\pagebreak
+
 # Minimum time calculations testing exponential growth
 
-In the main text, we evaluated the minimum time required to determine long-term trends in abundance via linear regression.
+In the main text, I evaluated the minimum time required to determine long-term trends in abundance via linear regression.
 This process examined linear trends in abundance over time.
-Here we examine the minimum time required to estimate long-term trends that are either exponential growth or decay.
-We use the same methods as described in the previous sections, but we take the $log$ of population density, or abundance.
+Here I examine the minimum time required to estimate long-term trends that are either exponential growth or decay.
+I use the same methods as described in the previous sections, but we take the $log$ of population density, or abundance.
 
@@
 # source('../scripts/calculate_power_metric.R')
-# 
-# for (j in 1:nrow(pop_info)){
-#     pop= subset(long_dat,long_dat$ID==pop_info$ID[j])
-#     pop = pop[(nrow(pop)-34):nrow(pop),]
-#     pop$popvalue=as.numeric(pop$popvalue)
-#     pop$popvalue=pop$popvalue/max(pop$popvalue) #should I standaridize this in a better way
-#     ## Take the log of population size and add a small number
-#     pop$popvalue = log(pop$popvalue + 0.001)
-#     
-#     pop_info$year[j] = pop$year[1]
-#     pop_info$variance[j]= var(pop$popvalue)
-#     pop_info$overall_trend[j] = calculate_slope(pop$popvalue)
-#     pop_info$autocorrelation[j] = acf(pop$popvalue,plot = FALSE)$acf[2]
-#     pop_info$trend_p_value[j]=calculate_p_value(pop$popvalue)
-#     pop_info$popvalue = pop$popvalue[1] #initial abundance
-#     pop_info$series_length = nrow(pop)
-#     
-# 
-#      pop_info$min_time_for_power[j] = min_time_needed(pop$popvalue,0.05,0.8)
-#     print(paste(j,':',pop_info$min_time_for_power[j],sep=' '))
-#     
-# }
-# 
-# pop_info=pop_info[pop_info$trend_p_value<0.05,]
-# long_dat=subset(long_dat,long_dat$ID %in% pop_info$ID)
-# long_time_names=subset(long_time_names,long_time_names %in% pop_info$ID)
-# 
+
 # save(pop_info,long_dat,long_time_names,file = '../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression_for_log_pop.Rdata')
@@
 
-plot(density(pop_info$min_time_for_power),las=1,ylab='',xlab='',cex.axis=1.2,main='')
-mtext(text = 'Frequency',2,line=3,cex=1.2)
-mtext(text = 'Minimum time required',1,line=3,cex=1.2)
+#plot(density(pop_info$min_time_for_power),las=1,ylab='',xlab='',cex.axis=1.2,main='')
+#mtext(text = 'Frequency',2,line=3,cex=1.2)
+#mtext(text = 'Minimum time required',1,line=3,cex=1.2)
+require(ggplot2)
+ggplot(aes(min_time_for_power),data=pop_info) + geom_bar() +  
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(), axis.line = element_line(colour = "black"),axis.text = element_text(size=12),axis.title.x =element_text(size=16) ,axis.title.y =element_text(size=16) ) +
+ labs(y='count',x='minimum time required')
 
@@
 
-In the main text, we showed how a simple population model could be simulated repeatedly to estimate the power obtained with time series of increasing length [@Bolker2008;@Johnson2015].
+In the main text, I showed how a simple population model could be simulated repeatedly to estimate the power obtained with time series of increasing length [@Bolker2008;@Johnson2015].
 The model in the main text simulated linear population growth with only a slope coefficient, y-intercept, and noise parameter required.
-This model is purely phenomenological and does not include any species life history.
-Here, we use the same routine as the main text, but simulate from a more biologically-realistic population model.
-We use the model described in @White2014.
+This model is purely phenomenological and does not include any species life-history information.
+Here, I use the same routine as the main text, but simulate from a more biologically-realistic population model.
+I use the model described in @White2014.
 The model is a stochastic, age-structured population model that includes density-dependence for lemon sharks (\emph{Negaprion brevirostris}) in Bimini, Bahamas.
 
-```{r, echo=F,fig.cap='Statistical power for different length of time series simulations for a lemon shark population in Bimini, Bahamas.\\label{fig:shark_example}'}
+```{r, echo=F,fig.height=3,fig.cap='Statistical power for different length of time series simulations for a lemon shark population in Bimini, Bahamas.\\label{fig:shark_example}'}
 
+par(mfrow=c(1,2),oma=c(0.5,0.5,0.5,0.5),mar=c(4,4,0,0))
+#### Simulate lemon shark model (See White et al. 2014)
+total_lemons=NULL
+IC= c(30,30,10,15,15,5,5,3,3,3,3,3,2,2,2,2,2,2,1,1,1,1,2,2,2,2)
+no.runs=50
+max.time=50
+for (g in 1:(no.runs)){
+#set up matrix (Lemons) to track number of individuals in each age class over time and matrix (Stable_age) to track the age distribution through time
+	Lemons = matrix(IC,nrow= 26,ncol=max.time)
+	Stable_age =matrix(0,nrow=26,ncol=max.time)
+	#initialize matrices
+	Lemons[1:26,1] = IC
+	#keep track of Juveniles, Subadults, and Adults
+	J =  sum(Lemons[1:3,1])    #Juveniles and newborns in lagoon
+	S =  sum(Lemons[4:12,1])   #subadults who have left the lagoon and cannot reproduce
+	A =  sum(Lemons[13:26,1])  #adults who can reproduce
+
+	#initialize parameters (these are described in the paper)
+	b=6.087   #Number of offspring per female 8.3 from Feldheim 2002, 6.7 from 
+	h=1   	  #hill coefficient
+	k=100     #half saturation constant (Estimated from Gruber 2001)
+	u=0.17    #adult mortality rate, 0.15 is the default rate
+
+		for (j in 2:(max.time)){
+				Lemons[1,j]= sum(rpois(sum(Lemons[13:26,j-1])/4,b))  #use 
+				Lemons[2,j] = Lemons[1,j-1] - rbinom (1, Lemons[1,j-1],(Lemons[1,j-1]^h)/(k+Lemons[1,j-1]^h))
+				Lemons[3,j] = Lemons[2,j-1] - rbinom (1, Lemons[2,j-1],u) 
+				Lemons[4,j] = Lemons[3,j-1] - rbinom (1, Lemons[3,j-1], u)
+					for (i in 5:25){
+					    Lemons[i,j] = Lemons[i-1,j-1] - rbinom (1, Lemons[i-1,j-1], u)
+					}
+					Lemons[26,j] = Lemons[25,j-1] - rbinom (1, Lemons[25,j-1], 1)
+		          J = c(J, sum(Lemons[1:3,j]))
+			 	  S = c(S, sum(Lemons[4:12,j]))
+			  	  A = c(A, sum(Lemons[13:26,j]))
+		}#end of time loop
+
+total_lemons = rbind(total_lemons,colSums(Lemons))
+}#end of trial loop 
+matplot(t(total_lemons),type='l',col=1,lty=1,las=1,ylab='population size',xlab='time (years)',cex.lab=1.2)
+mtext(text = '(a)',side = 3,line = -1.2,adj=0.95,font = 2,cex=1.2)
+###
 #source('../scripts/LemonShark_PopulationModel.R')
@@
 load('../analysis-outputs/example_power_for_shark_model.Rdata')
-par(mfrow=c(1,1),oma=c(3,3,0.5,0.5),mar=c(2,2,0,0))
-   plot(years_to_sample,shark_power,type='l',lwd=2,las=1,ylab='',xlab='',cex.axis=1.2)
+   plot(years_to_sample,shark_power,type='l',lwd=2,las=1,
+        ylab='',xlab='',cex.axis=1, ylim = c(0,1.05))
    abline(v=years_to_sample[tail(which(shark_power<0.8),1)+1],lty=2,lwd=2,col='red')
@@
    mtext(text = 'time sampled' ,side = 1,cex = 1.2,line=3)
-      mtext(text = 'statistical power' ,side = 2,cex = 1.2,line=3.5)
-      #mtext(text = '(b)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
+   mtext(text = 'statistical power' ,side = 2,cex = 1.2,line=2.5)
+   mtext(text = '(b)',side = 3,line = -1.2,adj=0.95,font = 2,cex=1.2)
 
@@
 
-Here, we parameterize the model for a situation where adult mortality rate is high enough to cause a population decline.
-In the same way we simulated models in the main text, we simulate this more biologically-realistic model for different lengths of time.
-For each length of time, we calculate the statistical power.
+Here, I parameterize the model for a situation where adult mortality rate is high enough to cause a population decline.
+As in the main text, I simulated this more biologically-realistic model for different lengths of time.
+For each length of time, I calculated the statistical power.
 Similar to results with the simpler model, statistical power generally increases with longer sampling time (Fig.
 \ref{fig:shark_example}).
-In this example, the minimum time required ($T_{min}$) t0 obtain at least 0.8 statistical power, given a significance level of 0.05, is `r years_to_sample[tail(which(shark_power<0.8),1)+1]` years.
+In this example, the minimum time required ($T_{min}$) to obtain at least 0.8 statistical power, given a significance level of 0.05, is `r years_to_sample[tail(which(shark_power<0.8),1)+1]` years.
 
@@
 
+<!-- REMOVING from paper as it takes away from main message
 # Minimum time required to estimate geometric growth rate
@@
 
-```{r, echo=F, fig.cap='Example of calculating minimum time required for growth rate estimation.
+```{r, echo=F,eval=F, fig.cap='Example of calculating minimum time required for growth rate estimation.
 (a) European herring gull (\\emph{Larus argentatus}) scaled density over time, (b) mean and standard deviation of growth rate for subsamples of entire time series, and (c) the percent error between mean estimated growth rate and the true long-term growth rate.
@@
   source('../scripts/percent_error.R')
-  aaa=calculate_average_geometric_growth(pop$popvalue)
-  plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='',xlab='',pch=16,las=1)
+  aaa=calculate_average_geometric_growth(pop$popvalue + 0.0001)
+  plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='',xlab='',pch=16,las=1,ylim=c(0,1))
     #abline(lm(pop$popvalue~pop$year),lwd=2,col='red')
@@
 
-  plot(rowMeans(aaa,na.rm=T),cex.axis=1.2,pch=16,ylab='',xlab='',las=1)
+  plot(rowMeans(aaa,na.rm=T),cex.axis=1.2,pch=16,ylab='',xlab='',las=1,ylim=c(0.8,1))
     mtext(text = 'years sampled',1,line=3,cex=1.4)
@@
 
-    arrows(1:34,rowMeans(aaa,na.rm=T),1:34,rowMeans(aaa,na.rm=T)+sd(apply(aaa,MARGIN = 1,FUN=var,na.rm=T),na.rm=T),angle = 90,length = 0.1)
-    arrows(1:34,rowMeans(aaa,na.rm=T),1:34,rowMeans(aaa,na.rm=T)-sd(apply(aaa,MARGIN = 1,FUN=var,na.rm=T),na.rm=T),angle = 90,length = 0.1)
+    #arrows(1:34,rowMeans(aaa,na.rm=T),1:34,rowMeans(aaa,na.rm=T)+sd(apply(aaa,MARGIN = 1,FUN=var,na.rm=T),na.rm=T),angle = 90,length = 0.1)
+    #arrows(1:34,rowMeans(aaa,na.rm=T),1:34,rowMeans(aaa,na.rm=T)-sd(apply(aaa,MARGIN = 1,FUN=var,na.rm=T),na.rm=T),angle = 90,length = 0.1)
     mtext(text = '(b)',side = 3,line = -1.2,adj=0.95,font = 2,cex=1.2)
 
-    plot(percent_error(rowMeans(aaa,na.rm=T),rowMeans(aaa,na.rm=T)[34]),cex.axis=1.2,pch=16,ylab='',xlab='',las=1)
+    plot(percent_error(rowMeans(aaa,na.rm=T),rowMeans(aaa,na.rm=T)[34]),cex.axis=1.2,pch=16,ylab='',xlab='',las=1, ylim = c(0,10))
     mtext(text = 'years sampled',1,line=3,cex=1.4)
@@
 
-```{r, echo=F,eval=T,fig.cap='Histogram of the minimum time required in order to estimate the long-term growth rate within 20% error.\\label{fig:min_time_growth_dist}'}
+```{r, echo=F,eval=F,fig.cap='Histogram of the minimum time required in order to estimate the long-term growth rate within 20% error.\\label{fig:min_time_growth_dist}'}
 # Growth rates calculations for each species
-# load('../cleaned-data/cleaned_timeseries_database.Rdata')
+load('../cleaned-data/cleaned_timeseries_database.Rdata')
 # 
@@
 #     pop$popvalue=as.numeric(pop$popvalue)
-#     pop$popvalue=pop$popvalue/max(pop$popvalue) #should I standaridize this in a better way
+#     pop$popvalue=(pop$popvalue - min(pop$popvalue))/(max(pop$popvalue) - min(pop$popvalue))
+#     pop$popvalue = pop$popvalue + 0.001
 # 
@@
 #       pop_info$min_time_growth_rate[j] = tail(which(percent_error(rowMeans(aaa,na.rm=T),rowMeans(aaa,na.rm=T)[34])>0.2),1)+1}
-#     #print(paste(j,':',pop_info$min_time_for_power[j],sep=' '))
+#     print(paste(j,':',pop_info$min_time_for_power[j],sep=' '))
 #  }
+# 
 # save(pop_info,file = '../analysis-outputs/min_time_for_growth_rate0_2.Rdata')
@@
 # Create plot
-par(mfrow=c(1,1))
-hist(pop_info$min_time_growth_rate,las=1,ylab='',xlab='',cex.axis=1.2,main='',breaks=20,xlim=c(0,40),ylim=c(0,500))
-mtext(text = 'frequency',2,line=3,cex=1.2)
-mtext(text = 'minimum time required',1,line=3,cex=1.2)
+#par(mfrow=c(1,1))
+#hist(pop_info$min_time_growth_rate,las=1,ylab='',xlab='',cex.axis=1.2,main='',breaks=20,xlim=c(0,40),ylim=c(0,500))
+#mtext(text = 'frequency',2,line=3,cex=1.2)
+#mtext(text = 'minimum time required',1,line=3,cex=1.2)
+
+ggplot(aes(min_time_growth_rate),data=pop_info) + geom_bar() +  
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(), axis.line = element_line(colour = "black"),axis.text = element_text(size=12),axis.title.x =element_text(size=16) ,axis.title.y =element_text(size=16) ) +
+ labs(y='count',x='minimum time required')
 
@@
 \clearpage 
+-->
+
+\pagebreak
 
@@
 
-In the main text, we examined the minimum time required to identify a trend in abundance via linear regression.
+In the main text, I examined the minimum time required to identify a trend in abundance via linear regression.
 This approach allowed us to identify increases or decreases, but a linear model may not always be a good fit.
@@
 
-Here, we conduct the same analyses in the main text, but instead calculate the minimum time series required to detect trends over time according to a GAM model.
-We hypothesized that GAMs should require less time to detect a trend as they are more flexible than linear regression.
-We provide an example that shows statistical power increases with more time sampled (Fig.
+Here, I conduct the same analyses in the main text, but instead calculate the minimum time series required to detect trends over time according to a GAM model.
+I hypothesized that GAMs should require less time to detect a trend as they are more flexible than linear regression.
+I provide an example that shows statistical power increases with more time sampled (Fig.
 \ref{fig:gam_example}).
 
-```{r,echo=F,message=F,warning=F,fig.cap='(a) Time series for Bigeye tuna (\\emph{Thunnus obesus}) with corresponding fitted GAM model in red and (b) statistical power as a function of the number of years sampled.
+```{r,echo=F,message=F,warning=F,fig.cap='(a) Time series for Bigeye tuna (\\emph{Thunnus obesus}) with corresponding fitted GAM model in red (with a smoothing parameter of 3) and (b) statistical power as a function of the number of years sampled.
 The horizontal ine at 0.8 indicates the minimum threshold for statistical power and the vertical line denotes the minimum time required to achieve 0.8 statistical power.\\label{fig:gam_example}'}
@@
     pop = subset(long_dat,long_dat$ID=='7558') #13192 for Harp seals, 7558 for Thunnus obesus
+    pop$popvalue=(pop$popvalue - min(pop$popvalue))/(max(pop$popvalue) - min(pop$popvalue)) 
     plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='',xlab='time (years)',pch=16,las=1)
@@
 #     pop$popvalue=as.numeric(pop$popvalue)
-#     pop$popvalue=pop$popvalue/max(pop$popvalue) #should I standaridize this in a better way
+#     pop$popvalue=(pop$popvalue - min(pop$popvalue))/(max(pop$popvalue) - min(pop$popvalue))
 # 
@@
 # Create plot
-par(mfrow=c(1,1))
-min_time_gam = pop_info$min_time_for_power_gam[pop_info$min_time_for_power_gam!=-999]
-plot(density(min_time_gam),las=1,ylab='',xlab='',cex.axis=1.2,main='',xlim=c(0,40))
-mtext(text = 'Frequency',2,line=3,cex=1.2)
-mtext(text = 'Minimum time required',1,line=3,cex=1.2)
+#par(mfrow=c(1,1))
+pop_info = subset(pop_info,pop_info$min_time_for_power_gam!=-999)
+#plot(density(min_time_gam),las=1,ylab='',xlab='',cex.axis=1.2,main='',xlim=c(0,40))
+#mtext(text = 'Frequency',2,line=3,cex=1.2)
+#mtext(text = 'Minimum time required',1,line=3,cex=1.2)
+require(ggplot2)
+ggplot(aes(min_time_for_power_gam),data=pop_info) + geom_bar() +  
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(), axis.line = element_line(colour = "black"),axis.text = element_text(size=12),axis.title.x =element_text(size=16) ,axis.title.y =element_text(size=16) ) +
+ labs(y='count',x='minimum time required')
 ```
 
-We then fit GAM models for `r length(pop_info$min_time_for_power_gam[pop_info$min_time_for_power_gam!=-999])` populations.
+We then fit GAM models for `r nrow(pop_info)` populations.
 We found a similar distribution of minimum time required as in the main text for linear regression (Fig.
 \ref{fig:min_time_dist_gam}).
-However, in line with our hypothesis, the GAM models did result in a lower mean minimum time required of `r round(mean(pop_info$min_time_for_power_gam[pop_info$min_time_for_power_gam!=-999]),2)` years compared to the results from the main text of `r round(mean(main_text$min_time_for_power),2)` years.
+However, in line with our hypothesis, the GAM models did result in a lower mean minimum time required of `r round(mean(pop_info$min_time_for_power_gam),2)` years compared to the results from the main text of `r round(mean(main_text$min_time_for_power),2)` years.
 

### 2018-02-19 writing/supp_mat_draft.tex
first update of short format version of the article

@@
 
-Here, we illustrate how we performed the subsampling and power
-calculations for a specific population. We provide more detail here than
-in the main manuscript. As an example, we examine a 35-year time series
-of Bigeye tuna (\emph{Thunnus obesus}), one species in the Global
-Population Dynamics Database (NERC Centre for Population Biology 2010).
-Simple linear regression indicates a significant decrease for this
-population with an estimated slope coefficient of -0.0189581. We assume
-that this significant increase over 35 years is in fact the ``true
-trend``. In statistical jargon, the 35-year trend is an effect that is
-actually present; we can reject the null hypothesis of no trend. We can
-then use this as a benchmark to see if subsamples of the time series
-also indicate a significant increase.
+Here, I illustrate how we performed the subsampling and power
+calculations for a specific population. As an example, I examine a
+35-year time series of Bigeye tuna (\emph{Thunnus obesus}), one species
+in the Global Population Dynamics Database (NERC Centre for Population
+Biology 2010). Simple linear regression indicates a significant decrease
+for this population with an estimated slope coefficient of -0.0271717. I
+assume that this significant increase over 35 years is in fact the
+``true trend``. In statistical jargon, the 35-year trend is an effect
+that is actually present; we can reject the null hypothesis of no trend.
+We can then use this as a benchmark to see if subsamples of the time
+series also indicate a significant increase.
 
-\begin{figure}[htbp]
+\begin{figure}
 \centering
@@
 
-We then perform a subsampling routine to estimate the minimum time
+I then perform a subsampling routine to estimate the minimum time
 required \(T_{min}\) (similar to Gerber, DeMaster, and Kareiva (1999)).
-This is the same routine we used for the results in the main manuscript.
+This is the same routine I used for the results in the main manuscript.
 
@@
 
+\pagebreak
+
 \section{Additional results from the main
@@
 
-\subsection{Predictors of minimum time
-required}\label{predictors-of-minimum-time-required}
+\subsection{Correlates of minimum time
+required}\label{correlates-of-minimum-time-required}
 
-\begin{figure}[htbp]
+\begin{figure}
 \centering
@@
 variables of the absolute value of the slope coefficient (or trend
-strength), temporal autocorrelation, and variability in population
-size.\label{fig:poisson_model}}
+strength), temporal autocorrelation, and coefficient of variation of
+population size.\label{fig:poisson_model}}
 \end{figure}
 
-In the main text, we explained how the minimum time required strongly
+In the main text, I explained how the minimum time required strongly
 correlated with the trend strength, temporal autocorrelation, and
-variance in population size. Here, we use a generalized linear model
-framework with a Poisson error structure to determine explanatory
-variables of the minimum time required. We use the same 878 populations
-as in the main text. In figure \ref{fig:poisson_model}, we show a set of
-residual plots for the regression. We then show the coefficient
-estimates and levels of significance in table \ref{table:model_output}.
+coefficient of variation in population size. Here, I use a generalized
+linear model framework with a Poisson error structure to determine
+explanatory variables of the minimum time required. I use the same 822
+populations as in the main text. In figure \ref{fig:poisson_model}, I
+show a set of residual plots for the regression analysis. I then show
+the coefficient estimates and levels of significance in table
+\ref{table:model_summary_time_series_correlates}.
 
 \begin{longtable}[]{@{}lrrrr@{}}
-\caption{Output of generalized linear model to examine predictors of the
-minimum time required for determing long-term population
-trends.\label{table:model_output}}\tabularnewline
+\caption{Output of generalized linear model to examine time series
+characteristics as correlates of the minimum time required for
+determining long-term population
+trends.\label{table:model_summary_time_series_correlates}}\tabularnewline
 \toprule
@@
 \endhead
-(Intercept) & 3.7405513 & 0.0242742 & 154.09595 & 0\tabularnewline
-abs\_overall\_trend & -104.0533454 & 3.0941097 & -33.62949 &
-0\tabularnewline
-autocorrelation & -0.4899533 & 0.0431769 & -11.34757 & 0\tabularnewline
-variance & 18.7366110 & 0.6404458 & 29.25557 & 0\tabularnewline
+(Intercept) & 4.2658002 & 0.0369853 & 115.337838 &
+0.0000000\tabularnewline
+abs\_overall\_trend & -70.0413235 & 2.0027094 & -34.973283 &
+0.0000000\tabularnewline
+autocorrelation & 0.0601596 & 0.0550282 & 1.093249 &
+0.2742843\tabularnewline
+coefficient\_variation & -0.1928445 & 0.0387392 & -4.978022 &
+0.0000006\tabularnewline
 \bottomrule
@@
 
-Our model with trend strength, autocorrelation, variance, and generation
-length accounted for 72.45\% of the variation in the minimum time
-required (Table \ref{table:model_output}). However, we also found trend
-strength and variance to be strongly correlated with one another.
-Therefore, we ran two additional models with either trend strength or
-variance, but not both together. This resulted in lower explained
-deviance (analogue to \(R^2\)) of 53.03\% and 45.03\%, respectively.
+\begin{longtable}[]{@{}lrrrr@{}}
+\caption{Output of generalized linear model to examine life-history
+trait correlates of the minimum time required for determining long-term
+population
+trends.\label{table:model_summary_biological_correlates}}\tabularnewline
+\toprule
+& Estimate & Std. Error & z value &
+Pr(\textgreater{}\textbar{}z\textbar{})\tabularnewline
+\midrule
+\endfirsthead
+\toprule
+& Estimate & Std. Error & z value &
+Pr(\textgreater{}\textbar{}z\textbar{})\tabularnewline
+\midrule
+\endhead
+(Intercept) & 2.8131074 & 0.1297488 & 21.681181 &
+0.0000000\tabularnewline
+litter\_or\_clutch\_size\_n & 0.0473894 & 0.0058279 & 8.131439 &
+0.0000000\tabularnewline
+gen\_len & 0.0640424 & 0.0087382 & 7.328999 & 0.0000000\tabularnewline
+adult\_body\_mass\_g & -0.0000809 & 0.0000177 & -4.563594 &
+0.0000050\tabularnewline
+maximum\_longevity\_y & -0.0074775 & 0.0019534 & -3.827907 &
+0.0001292\tabularnewline
+incubation\_d & -0.0127828 & 0.0031681 & -4.034864 &
+0.0000546\tabularnewline
+trophic2 & -0.2872678 & 0.1267799 & -2.265878 & 0.0234589\tabularnewline
+trophic2.5 & -0.3604948 & 0.1260791 & -2.859274 &
+0.0042461\tabularnewline
+trophic3 & -0.2936987 & 0.1274698 & -2.304064 & 0.0212190\tabularnewline
+\bottomrule
+\end{longtable}
+
+\begin{longtable}[]{@{}lrrrr@{}}
+\caption{Output of generalized linear model to examine both time series
+characteristics and life-history trait correlates of the minimum time
+required for determining long-term population
+trends.\label{table:model_summary_all_correlates}}\tabularnewline
+\toprule
+& Estimate & Std. Error & z value &
+Pr(\textgreater{}\textbar{}z\textbar{})\tabularnewline
+\midrule
+\endfirsthead
+\toprule
+& Estimate & Std. Error & z value &
+Pr(\textgreater{}\textbar{}z\textbar{})\tabularnewline
+\midrule
+\endhead
+(Intercept) & 4.3236120 & 0.1361628 & 31.7532519 &
+0.0000000\tabularnewline
+abs(overall\_trend) & -70.8497548 & 2.8424505 & -24.9255898 &
+0.0000000\tabularnewline
+coefficient\_variation & -0.2089905 & 0.0531591 & -3.9314122 &
+0.0000844\tabularnewline
+autocorrelation & -0.0038698 & 0.0790400 & -0.0489597 &
+0.9609515\tabularnewline
+litter\_or\_clutch\_size\_n & 0.0008944 & 0.0058450 & 0.1530248 &
+0.8783787\tabularnewline
+gen\_len & 0.0087069 & 0.0088167 & 0.9875484 & 0.3233738\tabularnewline
+adult\_body\_mass\_g & -0.0000214 & 0.0000174 & -1.2311406 &
+0.2182703\tabularnewline
+maximum\_longevity\_y & -0.0023131 & 0.0019647 & -1.1773197 &
+0.2390679\tabularnewline
+incubation\_d & -0.0004509 & 0.0031244 & -0.1443118 &
+0.8852542\tabularnewline
+trophic2 & 0.0302480 & 0.1273238 & 0.2375677 & 0.8122164\tabularnewline
+trophic2.5 & -0.0101150 & 0.1263433 & -0.0800595 &
+0.9361899\tabularnewline
+trophic3 & -0.0049709 & 0.1276051 & -0.0389552 &
+0.9689261\tabularnewline
+\bottomrule
+\end{longtable}
+
+The model with trend strength, autocorrelation, and the coefficient of
+variation in population size 75.13\% of the variation in the minimum
+time required (Table \ref{table:model_summary_time_series_correlates}).
+
+For a subset of populations, I examined life-history correlates as
+potential explanatory variables for the minimum time required. Table
+\ref{table:model_summary_biological_correlates} and
+\ref{table:model_summary_all_correlates} summarize results for a model
+with only life-history traits and an additional model with life-history
+traits and time series characteristics combined, respectively. The model
+with all life-history traits explained only 5.99\% of the variation in
+the minimum time required (Table
+\ref{table:model_summary_biological_correlates}). The model with all
+life-history traits and time series characteristics explained 77.26\% of
+the variation in the minimum time required (Table
+\ref{table:model_summary_all_correlates}). When accounting for
+characteristics of the time series, no life-history variables were
+significant.
 
 \pagebreak
-
-\subsection{Minimum time required and biological
-correlates}\label{minimum-time-required-and-biological-correlates}
-
-In the main manuscript, we examined the minimum time required to detect
-a significant trend in abundance over time using linear regression. As
-detailed in the main manuscript the minimum time required was around 15,
-but there was a wide distribution. Therefore, we were interested in
-potential explanatory variables of the minimum time required. In the
-main manuscript, we examined characteristics of the time series itself,
-like variability, autocorrelation, and the trend in abundance over time.
-Here, we combined our time series data with a data on life history
-characteristics of amniotes from Myhrvold et al. (2015). There was life
-history information available for 547 populations representing 315
-different species, all of which were birds (Aves class).
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
-\caption{Mimimum time required versus (a) generation length (years), (b)
-litter size (n), (c) adult body mass (grams), (d) maximum longevity
-(years), (e) egg mass (grams), and (f) incubation (days). The lines in
-each plot represent the best fit line from linear
-regression.\label{fig:biological_correlates}}
-\end{figure}
-
-We then correlated minimum time required for each population with its
-corresponding life history characteristics. In figure
-\ref{fig:biological_correlates} we examined minimum time required versus
-generation length (years), litter size (n), adult body mass (grams),
-maximum longevity (years), egg mass (grams), and incubation (days). None
-of these variables had much explanatory power in accounting for the
-variance in the minimum time required. We ran a generalized linear model
-(with a Poisson error structure) and found that these six variables only
-accounted for 6.87\% of the explained deviance in minimum time required.
 
@@
 experiencing larger inter-annual variability in population size (Fig.
-\ref{fig:class}c).
+\ref{fig:class}c). In addition, populations in the Actinopterygii class
+generally had weaker trends in abundance (i.e.~smaller slope
+coefficients) compared to the Aves and Mammalia populations.
 
-\begin{figure}[htbp]
+\begin{figure}
 \centering
-\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-6-1.pdf}
-\caption{(a) Minimum time required to estimate change in abundance for
-species class, (b) long-term trend (estimated slope coefficient) by
-species class, (c) interannual variability in population size by species
-class, and (d) temporal autocorrelation by species
-class.\label{fig:class}}
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-8-1.pdf}
+\caption{(a) Minimum time required to estimate change in abundance by
+biological class, (b) long-term trend (estimated slope coefficient) by
+class, (c) coefficient of variation in population size by class, and (d)
+temporal autocorrelation by class.\label{fig:class}}
 \end{figure}
@@
 
-\begin{figure}[htbp]
+\begin{figure}
 \centering
-\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-7-1.pdf}
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-9-1.pdf}
 \caption{Minimum time required to assess long-term trends in abundance
@@
 
+\subsection{Additional results for IUCN
+analysis}\label{additional-results-for-iucn-analysis}
+
+In the main text, I presented results detailing the difference between
+the minimum time required to achieve a high level of statistical power
+and the number of years suggested by IUCN based on generation length to
+assess a species as vulnerable. There were deviations between the two
+approaches. These deviations were mostly explained by the differences in
+inter-annual population variability (coefficient of variation) and
+generation time (Fig. \ref{fig:IUCN_correlates}).
+
+\begin{figure}
+\centering
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-10-1.pdf}
+\caption{The difference between minimum time estimates is the minimum
+time required to achieve 0.8 statistical power versus the minimum time
+required under IUCN criteria A2 to classify a species as vulnerable.
+Each point represents a single population, all of which saw declines of
+30\% or greater over a 10 year period. (a) Difference between minimum
+time estimates versus the coefficient of variation in population size.
+(b) Difference between minimum time estimates versus the generation
+length in years.\label{fig:IUCN_correlates}}
+\end{figure}
+
+\pagebreak
+
 \section{Minimum time calculations testing exponential
@@
 
-In the main text, we evaluated the minimum time required to determine
+In the main text, I evaluated the minimum time required to determine
 long-term trends in abundance via linear regression. This process
-examined linear trends in abundance over time. Here we examine the
+examined linear trends in abundance over time. Here I examine the
 minimum time required to estimate long-term trends that are either
-exponential growth or decay. We use the same methods as described in the
+exponential growth or decay. I use the same methods as described in the
 previous sections, but we take the \(log\) of population density, or
@@
 
-\begin{figure}[htbp]
+\begin{figure}
 \centering
-\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-8-1.pdf}
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-11-1.pdf}
 \caption{Distribution of the minimum time required in order to detect a
@@
 
-In the main text, we showed how a simple population model could be
+In the main text, I showed how a simple population model could be
 simulated repeatedly to estimate the power obtained with time series of
@@
 coefficient, y-intercept, and noise parameter required. This model is
-purely phenomenological and does not include any species life history.
-Here, we use the same routine as the main text, but simulate from a more
-biologically-realistic population model. We use the model described in
-White, Nagy, and Gruber (2014). The model is a stochastic,
+purely phenomenological and does not include any species life-history
+information. Here, I use the same routine as the main text, but simulate
+from a more biologically-realistic population model. I use the model
+described in White, Nagy, and Gruber (2014). The model is a stochastic,
 age-structured population model that includes density-dependence for
@@
 
-\begin{figure}[htbp]
+\begin{figure}
 \centering
-\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-9-1.pdf}
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-12-1.pdf}
 \caption{Statistical power for different length of time series
@@
 
-Here, we parameterize the model for a situation where adult mortality
-rate is high enough to cause a population decline. In the same way we
-simulated models in the main text, we simulate this more
-biologically-realistic model for different lengths of time. For each
-length of time, we calculate the statistical power. Similar to results
-with the simpler model, statistical power generally increases with
-longer sampling time (Fig. \ref{fig:shark_example}). In this example,
-the minimum time required (\(T_{min}\)) t0 obtain at least 0.8
-statistical power, given a significance level of 0.05, is 27 years.
+Here, I parameterize the model for a situation where adult mortality
+rate is high enough to cause a population decline. As in the main text,
+I simulated this more biologically-realistic model for different lengths
+of time. For each length of time, I calculated the statistical power.
+Similar to results with the simpler model, statistical power generally
+increases with longer sampling time (Fig. \ref{fig:shark_example}). In
+this example, the minimum time required (\(T_{min}\)) to obtain at least
+0.8 statistical power, given a significance level of 0.05, is 27 years.
 
-\section{Minimum time required to estimate geometric growth
-rate}\label{minimum-time-required-to-estimate-geometric-growth-rate}
-
-Instead of detecting a trend over time with linear regression, we could
-also calculate the geometric growth rate of the population. In figure
-\ref{fig:growth_rate}, we show how to calculate growth rates for
-subsamples of a time series. First, we created subsamples of each
-possible length from the full 35 year time series, as we did in the main
-next. Next, we calculated the mean and standard deviation of growth
-rates for each possible time series length (Fig.
-\ref{fig:growth_rate}b). Lastly, we calculated the percent error
-\(\mbox{percent error} = 100 \times \left| \frac{\mbox{observed} - \mbox{theoretical}}{\mbox{theoretical}} \right|\)
-between the mean of each time series length (observed) and the overall
-population growth rate (theoretical). In Fig. \ref{fig:growth_rate}c),
-we show the percent error as a function of time series length. Here we
-define the minimum time required as the minimum number of years to
-achieve less than 20\% error.
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-10-1.pdf}
-\caption{Example of calculating minimum time required for growth rate
-estimation. (a) European herring gull (\emph{Larus argentatus}) scaled
-density over time, (b) mean and standard deviation of growth rate for
-subsamples of entire time series, and (c) the percent error between mean
-estimated growth rate and the true long-term growth rate. The vertical
-bar denotes the minimum time required to estimate growth rate within
-20\% error.\label{fig:growth_rate}}
-\end{figure}
-
-We applied the same calculations to 1032 population time series. We then
-obtain a distribution of the minimum time required to measure the
-``true" long-term growth rate (Fig. \ref{fig:min_time_growth_dist}). We
-see a bimodal distribution with many populations required 30+ years to
-estimate the long-term growth rate. The large number of short years
-required is due to cases where the entire time series is consistently
-increasing or decreasing at the same rate.
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-11-1.pdf}
-\caption{Histogram of the minimum time required in order to estimate the
-long-term growth rate within 20\%
-error.\label{fig:min_time_growth_dist}}
-\end{figure}
-
-\clearpage
+\pagebreak
 
@@
 
-In the main text, we examined the minimum time required to identify a
+In the main text, I examined the minimum time required to identify a
 trend in abundance via linear regression. This approach allowed us to
@@
 
-Here, we conduct the same analyses in the main text, but instead
+Here, I conduct the same analyses in the main text, but instead
 calculate the minimum time series required to detect trends over time
-according to a GAM model. We hypothesized that GAMs should require less
+according to a GAM model. I hypothesized that GAMs should require less
 time to detect a trend as they are more flexible than linear regression.
-We provide an example that shows statistical power increases with more
+I provide an example that shows statistical power increases with more
 time sampled (Fig. \ref{fig:gam_example}).
 
-\begin{figure}[htbp]
+\begin{figure}
 \centering
-\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-12-1.pdf}
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-15-1.pdf}
 \caption{(a) Time series for Bigeye tuna (\emph{Thunnus obesus}) with
-corresponding fitted GAM model in red and (b) statistical power as a
-function of the number of years sampled. The horizontal ine at 0.8
-indicates the minimum threshold for statistical power and the vertical
-line denotes the minimum time required to achieve 0.8 statistical
-power.\label{fig:gam_example}}
+corresponding fitted GAM model in red (with a smoothing parameter of 3)
+and (b) statistical power as a function of the number of years sampled.
+The horizontal ine at 0.8 indicates the minimum threshold for
+statistical power and the vertical line denotes the minimum time
+required to achieve 0.8 statistical power.\label{fig:gam_example}}
 \end{figure}
 
-\begin{figure}[htbp]
+\begin{figure}
 \centering
-\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-13-1.pdf}
+\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-16-1.pdf}
 \caption{Distribution of the minimum time required in order to detect a
@@
 
-We then fit GAM models for 851 populations. We found a similar
+We then fit GAM models for 799 populations. We found a similar
 distribution of minimum time required as in the main text for linear
@@
 hypothesis, the GAM models did result in a lower mean minimum time
-required of 14.65 years compared to the results from the main text of
-16.45 years.
+required of 14.12 years compared to the results from the main text of
+15.91 years.
 
@@
 
-\hypertarget{ref-Myhrvold2015}{}
-Myhrvold, Nathan P., Elita Baldridge, Benjamin Chan, Dhileep Sivam,
-Daniel L. Freeman, and S.K. Morgan Ernest. 2015. ``An amniote
-life-history database to perform comparative analyses with birds,
-mammals, and reptiles.'' \emph{Ecology} 96 (11): 3109.
-
 \hypertarget{ref-GPDD2010}{}

### 2018-02-21 writing/manuscript_draft_short_version.Rmd
changes to short format version of the article

@@
 bibliography: White_bib.bib
+csl: ecology-letters.csl
 fontsize: 12pt
 geometry: margin=1in
-header-includes: \usepackage{float} \usepackage{lineno} \usepackage{setspace}\doublespacing \usepackage[firstinits=true,style=authoryear-comp,natbib=true,doi=false,isbn=false,url=false,uniquename=false,uniquelist=false,sorting=none]{biblatex}
+header-includes: 
+  \usepackage{float} 
+  \usepackage{lineno} 
+  \usepackage{setspace}\doublespacing 
+  \usepackage[round]{natbib} \bibpunct[; ]{(}{)}{,}{a}{}{,}
 output:
@@
     keep_tex: yes
-    number_sections: yes
+    number_sections: no
   word_document: default
@@
 ---
-Running head: Power to detect population trends [33 of 40 characters]
+<!--Running head: Power to detect population trends-->
 
@@
 
-To be submitted to: \emph{Ecology} as a Statistical Report
+To be submitted to: \emph{BioScience} as a Forum
 \vspace{1 mm}
 
-Number of pages: 20  
+Number of words: 4,400 (21 pages)  
 \vspace{1 mm}
@@
 
-Number of references: 33
+Number of references: 34
 \vspace{1 mm}
@@
 
-Data and code for all the figures and tables can be found at (https://github.com/erwhite1/time-series-project).
+All data and code is freely available at: (https://github.com/erwhite1/time-series-project).
 \vspace{2 mm}
@@
 However, population data are often expensive, requiring a lot of time and resources.
-When is a population time series long enough to address a question of interest?
-I determine the minimum time series length required to detect significant increases or decreases in population abundance.
-To address this question, I examine 822 populations of vertebrate species.
+What is the minimum population time series length required to detect significant trends in abundance?
+I first present an overview of the theory and past work that has tried to address this question.
+As a test of these approaches, I then examine 822 populations of vertebrate species.
 I show that on average 15.9 years of continuous monitoring are required in order to achieve a high level of statistical power.
-However, there is a wide distribution around this average, casting doubt on simple rules of thumb.
-The minimum time required depends on trend strength, population variability, and temporal autocorrelation.
-However, there were no life-history traits (e.g.
-generation length) that were predictive of the minimum time required.
-These results point to the importance of sampling populations over long periods of time.
-I argue that statistical power needs to be considered in monitoring program design and evaluation.
+However, there is a wide distribution around this average, casting doubt on commonly used, simple rules of thumb.
+These results point to the importance of long-term monitoring.
+I argue that statistical power needs to be considered more often in monitoring programs.
 Short time series are likely under-powered and potentially misleading.
@@
 
-Keywords: ecological time series, experimental design, monitoring, power analysis, statistical power, sampling design
+Keywords: ecological time series, experimental design, population monitoring, statistical power, sampling design
 
@@
 
-Observational studies and population time series have become a cornerstone of modern ecological research and conservation biology [@Magurran2010; @Hughes2017].
+Observational studies and population time series have become a cornerstone of modern ecological research and conservation biology [@Peters2010; @Magurran2010; @Hughes2017].
 Long-term data are necessary to both understand population dynamics and to assess species extinction risk.
@@
 They found that for this short-lived organism, 15 years was adequate to assess long-term trends in abundance.
-However, these studies, and other past work, have typically focused either on theoretical aspects of monitoring design or focused on only a few species.
+However, these studies, and other past work, have typically been only on theoretical aspects of monitoring design or focused on a few species.
 
-I estimate the minimum time series length required ($T_{min}$) to assess long-term changes in abundance via simple linear regression.
-Statistical power is not a new tool [@Cohen1992], but it is still under-appreciated in ecological research [@Legg2006].
-Therefore, I only briefly describe the idea of statistical power for time series analyses using simulations.
-I then examine 822 population time series to estimate $T_{min}$.
+Statistical power is not a new tool [@Gibbs1998; @Cohen1992], but it is still under-appreciated in ecological research [@Legg2006].
+Therefore, I begin by reviewing key concepts of power analyses in relation to time series analysis.
+I then explain how simulation approaches have been used to estimate the minimum time required to estimate long-term population trends.
+Lastly, I take an empirical approach to estimate the minimum time required for 822 animal populations.
 
@@
 
-# Methods
-
-## Statistical power in time series
+# Statistical power in time series analyses
 
@@
 The effect size is a estimate of the strength of a particular phenomenon.
+
 Prior to an experiment, one could set appropriate levels of power, significance level, and the effect size to estimate the sample size required for the experiment.
@@
 
-## Simulation approach
+In the context of time series data, sample size can be the number of study sites surveyed, frequency of surveys per year, and the number of years surveyed.
+For example, @Gibbs1998 examined how many times within a year a population needs to be sampled to ensure high statistical power.
+They found that the sampling intensity within a year differed greatly depending on species, because of differences in population variability.
+I explore similar methods, but I focus on the number of years required to estimate trends in abundance.
 
-One approach to determining the minimum time series length needed is through repetitive simulations of a population model [@Gerrodette1987].
+<!--For time series data, simulations can be designed for a specific population and question [@Bolker2008; @Johnson2015].
+Simple models can be simulated with parameter values corresponding to a population of interest [@Gerrodette1987].
+Statistical power is the proportion of simulations that then meet some set of criteria.
+The specific criteria depend on the question at hand.
+For example, given a time series, when is the slope from linear regression significantly different from zero?
+In other words, when is the time series significantly increasing or decreasing?
+It is then possible to determine how power changes with a variable of interest.
+For example, time series can be simulated for different lengths of time.
+From these simulations, the minimum time series length required to meet certain levels of statistical significance and power is estimated [@Bolker2008].
+-->
+
+# Simulation approach
+
+One approach to determining the minimum time series length needed is through repetitive simulations of a population model [@Gerrodette1987;@Gibbs1998].
 This is the same approach one might use in sample size calculations for any experimental design too complicated for simple power analyses [@Bolker2008;@Johnson2015].
@@
 the estimated slope from linear regression).
-It is important to note that any population model could be used here.
-As an example, in the supplementary material (Fig.
-\ref{fig:shark_example}), an age-structured model of population dynamics is used in place of equation \ref{sim_model} [@White2014].
 
-Statistical power is the proportion of simulations where the slope parameter from linear regression is significant at the 0.05 threshold.
+With regard to detecting time series trends, statistical power is the proportion of simulations where the slope parameter from linear regression is significant at the 0.05 threshold.
 Statistical power of 0.8 would indicate that, if there was a true trend in abundance, there would be a 0.8 probability of detecting the trend.
@@
 In Fig.
-\ref{fig:theoretical_approach}a, the model in equation \ref{sim_model} is simulated repetitively.
+\ref{fig:theoretical_approach}a, the model in equation \ref{sim_model} was simulated repetitively.
 Statistical power increases with increases in the length of time sampled (Fig.
@@
 However, it is also informative to examine how statistical power changes with sampling effort in empirical time series.
+
+It is important to note that any population model could be used here (see example in Fig.
+\ref{fig:min_time_dist_log_pop}).
+Ideally, the specific model choice should be tailored to the population of interest.
+As an example, I determined the minimum time required to estimate long-term population trends using a stochastic, age-structured model of lemon shark population dynamics in the Bahamas [@White2014].
+I found that over 27 years of continuous monitoring were needed in this particular scenario (Fig.
+\ref{fig:shark_example}).
+Not surprisingly, the minimum time required for the lemon shark population was strongly dependent on model parameters.
+Similarly, @Rueda-Cediel2015 used a matrix model parameterized for a particular snail species.
+They used the model to argue that only 10-15 years were needed to accurately assess trends in abundance.
 
@@
 
-## Data source
-I use a database of `r length(table(dat$ID))` population time series compiled by @Keith2015.
+# Empirical approach
+
+As an empirical test of these ideas, I used a database of `r length(table(dat$ID))` population time series compiled by @Keith2015.
 The data are originally from the Global Population Dynamics Database [@GPDD2010] and several other sources [@Keith2015].
@@
 
-## Empirical approach
-
-I assume that each time series is long enough to include all necessary information (e.g.
+I assumed that each time series was long enough to include all necessary information (e.g.
 variability) about the population.
-In other words, each time series is a representative sample.
-I first take all possible contiguous subsamples of each time series.
-For example, a time series of 35 years would have 34 possible contiguous subsamples of length 2, 34 possible contiguous subsamples of length 3, and continuing until 1 possible contiguous subsample of length 35 [@Gerber1999; @Brashares2005; @Giron-Nava2017].
-In line with the simulation approach, I determine the proportion of subsamples of a particular length that have estimated slope coefficients statistically different from zero.
+In other words, each time series was a representative sample.
+I first took all possible contiguous subsamples of each time series.
+For example, a time series of 35 years had 34 possible contiguous subsamples of length 2, 34 possible contiguous subsamples of length 3, and continuing until 1 possible contiguous subsample of length 35 [@Gerber1999; @Brashares2005; @Giron-Nava2017].
+In line with the simulation approach, I determined the proportion of subsamples of a particular length that had estimated slope coefficients statistically different from zero.
 This proportion is a measure of statistical power.
-Lastly, I determine which subsample length is required to achieve a certain threshold of statistical power (0.8, @Cohen1992).
-The minimum subsampled length that met these criteria is the minimum time series length required ($T_{min}$).
-All analyses were conducted in R [@RCoreTeam2016].
+Lastly, I determined which subsample length is required to achieve a certain threshold of statistical power (0.8, @Cohen1992).
+The minimum subsampled length that met these criteria was the minimum time series length required ($T_{min}$).
+All analyses were conducted in R [@RCoreTeam2017].
 
@@
 
-# Results
+<!--RESULTS-->
 
@@
 
-I examined a database of `r nrow(pop_info)` separate population time series representing `r length(table(pop_info$Binomial))` species.
-This database consisted of vertebrate species with a variety of life-history characteristics.
-Across all the populations, I found an average minimum time series length required ($T_{min}$) of `r round(mean(pop_info$min_time_for_power),1)` (SD=`r round((var(pop_info$min_time_for_power))^0.5,1)`), with a wide distribution (Fig.
+## Estimates of the minimum time required
+
+<!--I examined a database of `r nrow(pop_info)` separate population time series representing `r length(table(pop_info$Binomial))` species.
+This database consisted of vertebrate species with a variety of life-history characteristics.--> Across all the populations, I found an average minimum time series length required ($T_{min}$) of `r round(mean(pop_info$min_time_for_power),1)` (SD=`r round((var(pop_info$min_time_for_power))^0.5,1)`), with a wide distribution (Fig.
 \ref{fig:min_time_dist}b).
@@
 \ref{fig:class}).
+
+This time-frame is in line with past work on a short-lived snail species [@Rueda-Cediel2015] and a long-lived whale species [@Gerber1999].
+@Hatch2003 used seabird monitoring data to estimate minimum sampling requirements.
+He found that the time required ranged from 11 to 69 years depending on species, trend strength, and study design.
 
@@
 
+I initially hypothesized that species with longer lifespans or generation times may require a longer sampling period.
+This result could have been a result of at least two factors.
+First, the data I used may not include a diverse enough set of species with different life-history traits.
+Second, the question I posed, whether a population is increasing or decreasing, was specifically concerned with population trends over time.
+Therefore, life-history characteristics may be more important for other questions more closely tied to species biology.
+For example, @Blanchard2007 used detailed simulations of spatially-distributed fisheries to compare survey designs.
+They found that statistical power depended on survey design, temperature preferences, and the degree of population patchiness.
+
 ## Evalulating the IUCN criteria
@@
 
+For the populations I examined, the IUCN criteria may be overly simplistic (Fig.
+\ref{fig:IUCN_analysis}).
+For many populations, the IUCN criteria suggest more years than necessary are required to assess a population as vulnerable.
+Conversely, for other populations the IUCN criteria suggest sampling times that are less than the minimum time required for statistical power.
+This suggests that the IUCN criteria are probably too simplistic as the minimum time required does not correlate with generation time or other biological covariates (Fig.
+\ref{fig:correlates}d-h).
+
 <!--
@@
 
-# Discussion
+# Related questions
 
@@
 
+<!--
 I examined `r nrow(pop_info)` population time series.
@@
 
+-->
+
 <!-- Similarly, @Bart2004 recommended a standard of 80% power to detect a 50% decline over 20 years in landbird populations.
@@
 If spatial population dynamics were highly correlated, then it was better to sample more temporally, and vice versa.
-My work supports this idea as populations with strong temporal autocorrelation needed less years of sampling (Fig.
+The empirical data supports this idea as populations with strong temporal autocorrelation needed less years of sampling (Fig.
 \ref{fig:correlates}).
@@
 
+<!--
 As a test of the consequences of not conducting power analyses, I compared the results here to the IUCN Redlist criteria.
@@
 \ref{fig:correlates}d-h).
-
-The design of monitoring programs should include calculations of statistical power, the allocation of sampling in space versus time [@Rhodes2011], and metrics to increase precision [@Seavy2007].
-Ideally, a formal decision analysis to evaluate these different factors would be conducted to design or assess any monitoring program [@Hauser2006; @McDonald-Madden2010].
-This type of formal decision analysis would also include information on the costs of monitoring.
-These costs include the actual costs of sampling [@Brashares2005] and the ecological costs on inaction [@Thompson2000].
+-->
 
@@
 
-## Limitations
+# Limitations
 
 This paper has some limitations in determining the minimum time series length required.
-First, $T_{min}$ is particular to the specific question of interest.
+First, the minimum time estimated is particular to the specific question of interest.
 An additional complication is that for the empirical approach, the subsampling of the full time series allows for estimates of power, but the individual subsamples are clearly not independent of one another.
@@
 
-## Conclusions
+# Conclusions
 
@@
 \ref{fig:correlates}d-h).
-This result argues against overly simplified measures of minimum sampling time based on generation length or other life-history traits (Fig.
+This result argues against overly simplified measures of minimum sampling time based on generation length or other life-history traits, like those of the IUCN criteria (Fig.
 \ref{fig:IUCN_analysis}).
 
-My work implies that for many populations, short time series are probably not reliable for detecting population trends.
+The design of monitoring programs should include calculations of statistical power, the allocation of sampling in space versus time [@Larsen2001; @Rhodes2011], and metrics to increase precision [@Seavy2007].
+Ideally, a formal decision analysis to evaluate these different factors would be conducted to design or assess any monitoring program [@Hauser2006; @McDonald-Madden2010].
+This type of formal decision analysis would also include information on the costs of monitoring.
+These costs include the actual costs of sampling [@Brashares2005] and the ecological costs of inaction [@Thompson2000].
+
+For many populations, short time series are probably not reliable for detecting population trends.
 This result highlights the importance of long-term monitoring programs.
@@
     matplot(simulations,type='l',col=rgb(0.5,0.5,0.5,0.5),lty=1,ylab='',xlab='',las=1,cex.axis=1.2)
-      mtext(text = 'time' ,side = 1,cex = 1.2,line=3)
-      mtext(text = 'population size' ,side = 2,cex = 1.2,line=3.5)
+      mtext(text = 'Time' ,side = 1,cex = 1.2,line=3)
+      mtext(text = 'Population size' ,side = 2,cex = 1.2,line=3.5)
       mtext(text = '(a)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
@@
       abline(h=0.8,lty=2,lwd=2,col='black')
-      mtext(text = 'time sampled' ,side = 1,cex = 1.2,line=3)
-      mtext(text = 'statistical power' ,side = 2,cex = 1.2,line=3)
+      mtext(text = 'Time sampled' ,side = 1,cex = 1.2,line=3)
+      mtext(text = 'Statistical power' ,side = 2,cex = 1.2,line=3)
       mtext(text = '(b)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
@@
       plot(parameter_combinations$r,parameter_combinations$min_time_required,type='l',lwd=2,las=1,ylab='',xlab='',cex.axis=1.2,ylim=c(0,50))
-      mtext(text = 'trend strength (slope coefficient)' ,side = 1,cex = 1.2,line=3)
+      mtext(text = 'Trend strength (slope coefficient)' ,side = 1,cex = 1.2,line=3)
       mtext(text = expression(T[min]) ,side = 2,cex = 1.2,line=3)
@@
       plot(parameter_combinations$sigma,parameter_combinations$min_time_required,type='l',lwd=2,las=1,ylab='',xlab='',cex.axis=1.2,ylim=c(0,50))
-      mtext(text = 'population variability' ,side = 1,cex = 1.2,line=3)
+      mtext(text = 'Population variability' ,side = 1,cex = 1.2,line=3)
       mtext(text = expression(T[min]) ,side = 2,cex = 1.2,line=3)
@@
   theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(), axis.line = element_line(colour = "black"),axis.text = element_text(size=12),axis.title.x =element_text(size=16) ,axis.title.y =element_text(size=16) ) +
- labs(x='biological class',y='minimum time required') +
+ labs(x='Biological class',y='Minimum time required') +
   geom_text(aes(x=4.4,y=33),label='(a)',size=6) +
@@
 
-fish <- annotation_custom(rasterGrob(readPNG("phylo-fish.png"), interpolate=TRUE), xmin=-2.5,ymin=35,ymax=40) 
-bird <- annotation_custom(rasterGrob(readPNG("phylo-bird.png"), interpolate=TRUE), xmin=-0.5, ymin=35) 
-shark <- annotation_custom(rasterGrob(readPNG("phylo-shark.png"), interpolate=TRUE), xmin=2.6,xmax=3.3, ymin=35) 
-mammal <- annotation_custom(rasterGrob(readPNG("phylo-mammal.png"), interpolate=TRUE), xmin=3.3, ymin=35) 
+fish <- annotation_custom(rasterGrob(readPNG("phylo_pics/phylo-fish.png"), interpolate=TRUE), xmin=-2.5,ymin=35,ymax=40) 
+bird <- annotation_custom(rasterGrob(readPNG("phylo_pics/phylo-bird.png"), interpolate=TRUE), xmin=-0.5, ymin=35) 
+shark <- annotation_custom(rasterGrob(readPNG("phylo_pics/phylo-shark.png"), interpolate=TRUE), xmin=2.6,xmax=3.3, ymin=35) 
+mammal <- annotation_custom(rasterGrob(readPNG("phylo_pics/phylo-mammal.png"), interpolate=TRUE), xmin=3.3, ymin=35) 
 plot1 <- plot1+ fish + bird + shark + mammal
@@
   theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(), axis.line = element_line(colour = "black"),axis.text = element_text(size=12),axis.title.x =element_text(size=16) ,axis.title.y =element_text(size=16) ) +
- labs(y='count',x='minimum time required')+
+ labs(y='Count',x='Minimum time required')+
   geom_text(aes(x=35,y=47),label='(b)',size=6)
@@
   plot(abs(pop_info$overall_trend),pop_info$min_time_for_power,pch=pop_info$shape,col=pop_info$color,las=1,ylab='',xlab='',cex.axis=1.2,cex=1.3,ylim=c(0,35))
-  mtext('| trend strength |',1,line=3,cex=1.2)
+  mtext('| Trend strength |',1,line=3,cex=1.2)
   mtext('A',3,line=0,adj = 0.0,cex=1.2)
     plot(pop_info$coefficient_variation,pop_info$min_time_for_power,pch=pop_info$shape,col=pop_info$color,las=1,ylab='',xlab='',cex.axis=1.2, xlim = c(0.2,2.2),cex=1.3,ylim=c(0,35))
-    mtext('coefficient of variation',1,line=3,cex=1.2)
+    mtext('Coefficient of variation',1,line=3,cex=1.2)
     mtext('B',3,line=0,adj = 0.0,cex=1.2)
       plot(pop_info$autocorrelation,pop_info$min_time_for_power,pch=pop_info$shape,col=pop_info$color,las=1,ylab='',xlab='',cex.axis=1.2, xlim = c(-0.5,1),cex=1.3,ylim=c(0,35))
-      mtext('autocorrelation',1,line=3,cex=1.2)
+      mtext('Autocorrelation',1,line=3,cex=1.2)
       mtext('C',3,line=0,adj = 0.0,cex=1.2)
@@
   abline(lm(LPI_pop_info$min_time_for_power~LPI_pop_info$gen_len),lwd=3)
-  mtext(text = 'generation length (years)',side = 1,line = 3,cex = 1.2)
+  mtext(text = 'Generation length (years)',side = 1,line = 3,cex = 1.2)
   mtext('D',3,line=0,adj = 0.0,cex=1.2)
@@
   abline(lm(LPI_pop_info$min_time_for_power~LPI_pop_info$litter_or_clutch_size_n),lwd=3)
-  mtext(text = 'litter size (n)',side = 1,line = 3,cex = 1.2)
+  mtext(text = 'Litter size (n)',side = 1,line = 3,cex = 1.2)
   mtext('E',3,line=0,adj = 0.0,cex=1.2)
@@
   abline(lm(LPI_pop_info$min_time_for_power~LPI_pop_info$adult_body_mass_g),lwd=3)
-  mtext(text = 'log(adult mass (grams))',side = 1,line = 3,cex = 1.2)
+  mtext(text = 'log(adult mass) (grams)',side = 1,line = 3,cex = 1.2)
   mtext('F',3,line=0,adj = 0.0,cex=1.2)
@@
   abline(lm(LPI_pop_info$min_time_for_power~LPI_pop_info$maximum_longevity_y),lwd=3)
-  mtext(text = 'maximum longevity (years)',side = 1,line = 3,cex = 1.2)
+  mtext(text = 'Maximum longevity (years)',side = 1,line = 3,cex = 1.2)
   mtext('G',3,line=0,adj = 0.0,cex=1.2)
@@
   abline(lm(LPI_pop_info$min_time_for_power~LPI_pop_info$incubation_d),lwd=3)
-  mtext(text = 'incubation (days)',side = 1,line = 3,cex = 1.2)
+  mtext(text = 'Incubation (days)',side = 1,line = 3,cex = 1.2)
   mtext('H',3,line=0,adj = 0.0,cex=1.2)
@@
 
-mtext(text = 'minimum time required',side = 2,line = 2,cex = 1.4,outer = TRUE)
+mtext(text = 'Minimum time required',side = 2,line = 2,cex = 1.4,outer = TRUE)
 
@@
        pch=declining_populations$shape,cex.lab=1.3,las=1,cex=1.5,
-       xlab = 'minimum time required under IUCN criteria',ylab='')
-mtext(text = 'minimum time required',side = 2,line = 3.8,outer = F,padj = 0.5,cex=1.3)
+       xlab = 'Minimum time required under IUCN criteria',ylab='')
+mtext(text = 'Minimum time required',side = 2,line = 3.8,outer = F,padj = 0.5,cex=1.3)
 mtext(text = 'for 0.8 statistical power',side = 2,line = 2.8,outer = F,padj = 0.5,cex=1.3)
@@
 
-whale <- readPNG('phylo-sperm-whale.png')
+whale <- readPNG('phylo_pics/phylo-sperm-whale.png')
 rasterImage(whale,xleft=70,ybottom=12,xright=82,ytop=17)
 
-puffin <- readPNG('phylo-puffin.png')
+puffin <- readPNG('phylo_pics/phylo-puffin.png')
 rasterImage(puffin,xleft=61,ybottom=30,xright=70,ytop=36)

### 2018-02-21 writing/manuscript_draft_short_version.tex
changes to short format version of the article

@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{5}
+\setcounter{secnumdepth}{0}
 % Redefines (sub)paragraphs to behave more like sections
@@
 \usepackage{float} \usepackage{lineno}
-\usepackage{setspace}\doublespacing \usepackage[firstinits=true,style=authoryear-comp,natbib=true,doi=false,isbn=false,url=false,uniquename=false,uniquelist=false,sorting=none]{biblatex}
-
-\usepackage{titlesec}
-
-\titlespacing\section{0pt}{-5pt plus 1pt minus 1pt}{-5pt plus 1pt minus 1pt}
-\titlespacing\subsection{0pt}{-5pt plus 1pt minus 1pt}{-5pt plus 1pt minus 1pt}
-\titlespacing\subsubsection{0pt}{-5pt plus 1pt minus 1pt}{-5pt plus 1pt minus 1pt}
+\usepackage{setspace}\doublespacing \usepackage[round]{natbib}
+\bibpunct[; ]{(}{)}{,}{a}{}{,}
 
 \begin{document}
-
-Running head: Power to detect population trends {[}33 of 40
-characters{]}
 
@@
 
-To be submitted to: \emph{Ecology} as a Statistical Report \vspace{1 mm}
+To be submitted to: \emph{BioScience} as a Forum \vspace{1 mm}
 
-Number of pages: 20\\
+Number of words: 4,400 (21 pages)\\
 \vspace{1 mm}
@@
 
-Number of references: 33 \vspace{1 mm}
+Number of references: 34 \vspace{1 mm}
 
@@
 
-Data and code for all the figures and tables can be found at
+All data and code is freely available at:
 (\url{https://github.com/erwhite1/time-series-project}). \vspace{2 mm}
@@
 
-Keywords: ecological time series, experimental design, monitoring, power
-analysis, statistical power, sampling design
+Keywords: ecological time series, experimental design, population
+monitoring, statistical power, sampling design
 
@@
 cornerstone of modern ecological research and conservation biology
-(Magurran et al. 2010; Hughes et al. 2017). Long-term data are necessary
-to both understand population dynamics and to assess species extinction
-risk. Some time series may now be considered ``long-term''
-(e.g.~continuous plankton recorder, Giron-Nava et al. (2017)), but most
-are still short. Time series are typically short due to short funding
-cycles and typical experimental time-frames (Field et al. 2007; Hughes
-et al. 2017).
+(Magurran \emph{et al.} 2010, Peters 2010, Hughes \emph{et al.} 2017).
+Long-term data are necessary to both understand population dynamics and
+to assess species extinction risk. Some time series may now be
+considered ``long-term'' (e.g.~continuous plankton recorder, Giron-Nava
+\emph{et al.} (2017)), but most are still short. Time series are
+typically short due to short funding cycles and typical experimental
+time-frames (Field \emph{et al.} 2007, Hughes \emph{et al.} 2017).
 
 How long of a time series is actually necessary? This question has
-important implications for both research and management (Nichols and
+important implications for both research and management (Nichols \&
 Williams 2006). A short time series may lead to wrong conclusions given
-large natural year-to-year variability (McCain, Szewczyk, and Knight
-2016). Managers need to know when action is needed for a population.
-Therefore, managers must understand when a population trend over time is
-actually meaningful. For example, the International Union for
-Conservation of Nature (IUCN) Red List Categories and Criteria suggest,
-under Criterion A2, a species qualifies as vulnerable if it has
-experienced a 30\% decline over 10 years, or three generations (IUCN
-2012). Sampling is typically expensive, therefore, we also do not want
-to sample for longer than is necessary. For example, Gerber, DeMaster,
-and Kareiva (1999) investigated the minimum time series required to
-estimate population growth of the endangered, but recovering, eastern
-North Pacific gray whale (\emph{Eschrichtius robustus}). They used a
-long-term census to retroactively determine the minimum time series
-required to assess threat status. They found that only 11 years were
-needed, eight years before the delisting decision was made. This
-highlights the importance of estimating the minimum time series required
-as an earlier decision would have saved time and money (Gerber,
-DeMaster, and Kareiva 1999). Further, waiting too long to decide an
-action can imperil a species where management action could have been
-taken earlier (Martin et al. 2012; Martin et al. 2017).
+large natural year-to-year variability (McCain \emph{et al.} 2016).
+Managers need to know when action is needed for a population. Therefore,
+managers must understand when a population trend over time is actually
+meaningful. For example, the International Union for Conservation of
+Nature (IUCN) Red List Categories and Criteria suggest, under Criterion
+A2, a species qualifies as vulnerable if it has experienced a 30\%
+decline over 10 years, or three generations (IUCN 2012). Sampling is
+typically expensive, therefore, we also do not want to sample for longer
+than is necessary. For example, Gerber \emph{et al.} (1999) investigated
+the minimum time series required to estimate population growth of the
+endangered, but recovering, eastern North Pacific gray whale
+(\emph{Eschrichtius robustus}). They used a long-term census to
+retroactively determine the minimum time series required to assess
+threat status. They found that only 11 years were needed, eight years
+before the delisting decision was made. This highlights the importance
+of estimating the minimum time series required as an earlier decision
+would have saved time and money (Gerber \emph{et al.} 1999). Further,
+waiting too long to decide an action can imperil a species where
+management action could have been taken earlier (Martin \emph{et al.}
+2012, 2017).
 
 Past work has investigated questions related to the minimum time series
-required to estimate trends in population size over time (Wagner,
-Vandergoot, and Tyson 2009; Giron-Nava et al. 2017). For example, Rhodes
-and Jonzen (2011) examined the optimal allocation of effort between
+required to estimate trends in population size over time (Wagner
+\emph{et al.} 2009, Giron-Nava \emph{et al.} 2017). For example, Rhodes
+\& Jonzen (2011) examined the optimal allocation of effort between
 spatial and temporal replicates. Using simple populations models, they
@@
 spatial and temporal autocorrelation, and observer error. Rueda-Cediel
-et al. (2015) also used a modeling approach, but parameterized a model
-specific for a threatened snail, \emph{Tasmaphena lamproides}. They
-found that for this short-lived organism, 15 years was adequate to
+\emph{et al.} (2015) also used a modeling approach, but parameterized a
+model specific for a threatened snail, \emph{Tasmaphena lamproides}.
+They found that for this short-lived organism, 15 years was adequate to
 assess long-term trends in abundance. However, these studies, and other
-past work, have typically focused either on theoretical aspects of
-monitoring design or focused on only a few species.
+past work, have typically been only on theoretical aspects of monitoring
+design or focused on a few species.
 
-I estimate the minimum time series length required (\(T_{min}\)) to
-assess long-term changes in abundance via simple linear regression.
-Statistical power is not a new tool (Cohen 1992), but it is still
-under-appreciated in ecological research (Legg and Nagy 2006).
-Therefore, I only briefly describe the idea of statistical power for
-time series analyses using simulations. I then examine 822 population
-time series to estimate \(T_{min}\).
+Statistical power is not a new tool (Cohen 1992, Gibbs \emph{et al.}
+1998), but it is still under-appreciated in ecological research (Legg \&
+Nagy 2006). Therefore, I begin by reviewing key concepts of power
+analyses in relation to time series analysis. I then explain how
+simulation approaches have been used to estimate the minimum time
+required to estimate long-term population trends. Lastly, I take an
+empirical approach to estimate the minimum time required for 822 animal
+populations.
 
-\section{Methods}\label{methods}
-
-\subsection{Statistical power in time
-series}\label{statistical-power-in-time-series}
+\section{Statistical power in time series
+analyses}\label{statistical-power-in-time-series-analyses}
 
@@
 intricately linked: significance level (\(\alpha\)), statistical power,
-effect size, and sample size (Legg and Nagy 2006). The exact
-relationship between these quantities depends on the specific
-statistical test. A type I error is a false positive, or incorrect
-rejection of a true null hypothesis. For example, if a time series was
-assessed as significantly increasing or decreasing---when there was no
-true significant trend---this would be a false positive. The false
-positive rate, or significance level (\(\alpha\)) is often set at 0.05
-(although this is purely historical, Mapstone (1995)). A type II error
-(\(\beta\)) is a failure to detect a true trend, or failure to reject a
-false null hypothesis. Formally, statistical power (\(1-\beta\)) is one
-minus the probability of a type II error (\(\beta\)). The effect size is
-a estimate of the strength of a particular phenomenon. Prior to an
-experiment, one could set appropriate levels of power, significance
-level, and the effect size to estimate the sample size required for the
-experiment. This approach, however, is not straight-forward for a time
-series, or more complicated scenarios (P. C. Johnson et al. 2015), as
-data are clearly non-independent.
+effect size, and sample size (Legg \& Nagy 2006). The exact relationship
+between these quantities depends on the specific statistical test. A
+type I error is a false positive, or incorrect rejection of a true null
+hypothesis. For example, if a time series was assessed as significantly
+increasing or decreasing---when there was no true significant
+trend---this would be a false positive. The false positive rate, or
+significance level (\(\alpha\)) is often set at 0.05 (although this is
+purely historical, Mapstone (1995)). A type II error (\(\beta\)) is a
+failure to detect a true trend, or failure to reject a false null
+hypothesis. Formally, statistical power (\(1-\beta\)) is one minus the
+probability of a type II error (\(\beta\)). The effect size is a
+estimate of the strength of a particular phenomenon.
 
-\subsection{Simulation approach}\label{simulation-approach}
+Prior to an experiment, one could set appropriate levels of power,
+significance level, and the effect size to estimate the sample size
+required for the experiment. This approach, however, is not
+straight-forward for a time series, or more complicated scenarios
+(Johnson \emph{et al.} 2015), as data are clearly non-independent.
+
+In the context of time series data, sample size can be the number of
+study sites surveyed, frequency of surveys per year, and the number of
+years surveyed. For example, Gibbs \emph{et al.} (1998) examined how
+many times within a year a population needs to be sampled to ensure high
+statistical power. They found that the sampling intensity within a year
+differed greatly depending on species, because of differences in
+population variability. I explore similar methods, but I focus on the
+number of years required to estimate trends in abundance.
+
+\section{Simulation approach}\label{simulation-approach}
 
 One approach to determining the minimum time series length needed is
-through repetitive simulations of a population model (Gerrodette 1987).
-This is the same approach one might use in sample size calculations for
-any experimental design too complicated for simple power analyses
-(Bolker 2008; P. C. Johnson et al. 2015). Essentially, a population
-model, with a selected set of parameter values, is simulated
-repetitively for a number of years. As an example, we can take the
-following population model for population size \(N\) at time \(t\):
+through repetitive simulations of a population model (Gerrodette 1987,
+Gibbs \emph{et al.} 1998). This is the same approach one might use in
+sample size calculations for any experimental design too complicated for
+simple power analyses (Bolker 2008, Johnson \emph{et al.} 2015).
+Essentially, a population model, with a selected set of parameter
+values, is simulated repetitively for a number of years. As an example,
+we can take the following population model for population size \(N\) at
+time \(t\):
 
@@
 the trend strength of the increase or decrease (i.e.~the estimated slope
-from linear regression). It is important to note that any population
-model could be used here. As an example, in the supplementary material
-(Fig. \ref{fig:shark_example}), an age-structured model of population
-dynamics is used in place of equation \ref{sim_model} (White, Nagy, and
-Gruber 2014).
+from linear regression).
 
-Statistical power is the proportion of simulations where the slope
-parameter from linear regression is significant at the 0.05 threshold.
-Statistical power of 0.8 would indicate that, if there was a true trend
-in abundance, there would be a 0.8 probability of detecting the trend.
-Values of 0.05 for the significance level and 0.8 statistical power are
-historical (Cohen 1992) and it is important to examine the effect of
-changing these values (Fig. \ref{fig:min_time_vs_alpha_beta}).
+With regard to detecting time series trends, statistical power is the
+proportion of simulations where the slope parameter from linear
+regression is significant at the 0.05 threshold. Statistical power of
+0.8 would indicate that, if there was a true trend in abundance, there
+would be a 0.8 probability of detecting the trend. Values of 0.05 for
+the significance level and 0.8 statistical power are historical (Cohen
+1992) and it is important to examine the effect of changing these values
+(Fig. \ref{fig:min_time_vs_alpha_beta}).
 
 In Fig. \ref{fig:theoretical_approach}a, the model in equation
-\ref{sim_model} is simulated repetitively. Statistical power increases
+\ref{sim_model} was simulated repetitively. Statistical power increases
 with increases in the length of time sampled (Fig.
@@
 confident in the detection of a long-term trend in abundance. As shown
-previously (Rhodes and Jonzen 2011; Rueda-Cediel et al. 2015),
+previously (Rhodes \& Jonzen 2011, Rueda-Cediel \emph{et al.} 2015),
 statistical power increases with larger trend strength and lower
@@
 
-\subsection{Data source}\label{data-source}
+It is important to note that any population model could be used here
+(see example in Fig. \ref{fig:min_time_dist_log_pop}). Ideally, the
+specific model choice should be tailored to the population of interest.
+As an example, I determined the minimum time required to estimate
+long-term population trends using a stochastic, age-structured model of
+lemon shark population dynamics in the Bahamas (White \emph{et al.}
+2014). I found that over 27 years of continuous monitoring were needed
+in this particular scenario (Fig. \ref{fig:shark_example}). Not
+surprisingly, the minimum time required for the lemon shark population
+was strongly dependent on model parameters. Similarly, Rueda-Cediel
+\emph{et al.} (2015) used a matrix model parameterized for a particular
+snail species. They used the model to argue that only 10-15 years were
+needed to accurately assess trends in abundance.
 
-I use a database of 2444 population time series compiled by Keith et al.
-(2015). The data are originally from the Global Population Dynamics
-Database (NERC Centre for Population Biology 2010) and several other
-sources (Keith et al. 2015). I filtered out short time series (less than
-35 years), and those with missing data, leaving 822 time series. The
-data includes information on 477 vertebrate species with a focus on
-mammals, birds, and fish.
+\section{Empirical approach}\label{empirical-approach}
 
-\subsection{Empirical approach}\label{empirical-approach}
+As an empirical test of these ideas, I used a database of 2444
+population time series compiled by Keith \emph{et al.} (2015). The data
+are originally from the Global Population Dynamics Database (NERC Centre
+for Population Biology 2010) and several other sources (Keith \emph{et
+al.} 2015). I filtered out short time series (less than 35 years), and
+those with missing data, leaving 822 time series. The data includes
+information on 477 vertebrate species with a focus on mammals, birds,
+and fish.
 
-I assume that each time series is long enough to include all necessary
+I assumed that each time series was long enough to include all necessary
 information (e.g.~variability) about the population. In other words,
-each time series is a representative sample. I first take all possible
+each time series was a representative sample. I first took all possible
 contiguous subsamples of each time series. For example, a time series of
-35 years would have 34 possible contiguous subsamples of length 2, 34
-possible contiguous subsamples of length 3, and continuing until 1
-possible contiguous subsample of length 35 (Gerber, DeMaster, and
-Kareiva 1999; Brashares and Sam 2005; Giron-Nava et al. 2017). In line
-with the simulation approach, I determine the proportion of subsamples
-of a particular length that have estimated slope coefficients
-statistically different from zero. This proportion is a measure of
-statistical power. Lastly, I determine which subsample length is
-required to achieve a certain threshold of statistical power (0.8, Cohen
-(1992)). The minimum subsampled length that met these criteria is the
-minimum time series length required (\(T_{min}\)). All analyses were
-conducted in R (R Core Team 2016).
+35 years had 34 possible contiguous subsamples of length 2, 34 possible
+contiguous subsamples of length 3, and continuing until 1 possible
+contiguous subsample of length 35 (Gerber \emph{et al.} 1999, Brashares
+\& Sam 2005, Giron-Nava \emph{et al.} 2017). In line with the simulation
+approach, I determined the proportion of subsamples of a particular
+length that had estimated slope coefficients statistically different
+from zero. This proportion is a measure of statistical power. Lastly, I
+determined which subsample length is required to achieve a certain
+threshold of statistical power (0.8, Cohen (1992)). The minimum
+subsampled length that met these criteria was the minimum time series
+length required (\(T_{min}\)). All analyses were conducted in R (R Core
+Team 2017).
 
-\section{Results}\label{results}
+\subsection{Estimates of the minimum time
+required}\label{estimates-of-the-minimum-time-required}
 
-I examined a database of 822 separate population time series
-representing 477 species. This database consisted of vertebrate species
-with a variety of life-history characteristics. Across all the
-populations, I found an average minimum time series length required
-(\(T_{min}\)) of 15.9 (SD=8.3), with a wide distribution (Fig.
-\ref{fig:min_time_dist}b). Estimates of \(T_{min}\) varied between
+Across all the populations, I found an average minimum time series
+length required (\(T_{min}\)) of 15.9 (SD=8.3), with a wide distribution
+(Fig. \ref{fig:min_time_dist}b). Estimates of \(T_{min}\) varied between
 biological class (Fig. \ref{fig:min_time_dist}a). Ray-finned fish (class
@@
 size and strength of trends in abundance (Fig. \ref{fig:class}).
+
+This time-frame is in line with past work on a short-lived snail species
+(Rueda-Cediel \emph{et al.} 2015) and a long-lived whale species (Gerber
+\emph{et al.} 1999). Hatch (2003) used seabird monitoring data to
+estimate minimum sampling requirements. He found that the time required
+ranged from 11 to 69 years depending on species, trend strength, and
+study design.
 
@@
 variation in population, and autocorrelation account for 75.1\% of the
-explained deviance (Zuur et al. 2009) in minimum time series length
-required.
+explained deviance (Zuur \emph{et al.} 2009) in minimum time series
+length required.
 
-There was life-history information available (Myhrvold et al. 2015) for
-547 populations representing 315 different species, all of which were
-birds (Aves class). Some life-history traits were significant predictors
-for the minimum time required (Figs. \ref{fig:correlates}d-h, Tables
-A2,A3). However, even combined, all five of the life-history traits
-accounted for only 5.99\% of the explained deviance in minimum time
-series length required. In addition, when accounting for trend strength,
-coefficient of variation, and autocorrelation, no life-history traits
-were significant predictors of the minimum time required (Table A3).
+There was life-history information available (Myhrvold \emph{et al.}
+2015) for 547 populations representing 315 different species, all of
+which were birds (Aves class). Some life-history traits were significant
+predictors for the minimum time required (Figs. \ref{fig:correlates}d-h,
+Tables A2,A3). However, even combined, all five of the life-history
+traits accounted for only 5.99\% of the explained deviance in minimum
+time series length required. In addition, when accounting for trend
+strength, coefficient of variation, and autocorrelation, no life-history
+traits were significant predictors of the minimum time required (Table
+A3).
+
+I initially hypothesized that species with longer lifespans or
+generation times may require a longer sampling period. This result could
+have been a result of at least two factors. First, the data I used may
+not include a diverse enough set of species with different life-history
+traits. Second, the question I posed, whether a population is increasing
+or decreasing, was specifically concerned with population trends over
+time. Therefore, life-history characteristics may be more important for
+other questions more closely tied to species biology. For example,
+Blanchard \emph{et al.} (2007) used detailed simulations of
+spatially-distributed fisheries to compare survey designs. They found
+that statistical power depended on survey design, temperature
+preferences, and the degree of population patchiness.
 
@@
 
-\section{Discussion}\label{discussion}
+For the populations I examined, the IUCN criteria may be overly
+simplistic (Fig. \ref{fig:IUCN_analysis}). For many populations, the
+IUCN criteria suggest more years than necessary are required to assess a
+population as vulnerable. Conversely, for other populations the IUCN
+criteria suggest sampling times that are less than the minimum time
+required for statistical power. This suggests that the IUCN criteria are
+probably too simplistic as the minimum time required does not correlate
+with generation time or other biological covariates (Fig.
+\ref{fig:correlates}d-h).
 
-I examined 822 population time series. I then subsampled each to
-determine the minimum time required to achieve a desired significance
-level and power for linear regression. Statistical power is important as
-it provides information on the necessary number of samples required to
-determine a significant trend (Legg and Nagy 2006). I found that on
-average 15.91 years of continuous monitoring were typically necessary
-(Fig. \ref{fig:min_time_dist}b). However, the distribution of minimum
-time required was wide. Therefore, simplistic rules of thumb are
-probably not appropriate. This time-frame is in line with past work on a
-short-lived snail species (Rueda-Cediel et al. 2015) and a long-lived
-whale species (Gerber, DeMaster, and Kareiva 1999). Hatch (2003) used
-seabird monitoring data to estimate minimum sampling requirements. He
-found that the time required ranged from 11 to 69 years depending on
-species, trend strength, and study design.
-
-In line with theoretical predictions (Rhodes and Jonzen 2011), I also
-found \(T_{min}\) was strongly correlated with the trend strength,
-variability in population size, and temporal autocorrelation (Fig.
-\ref{fig:correlates}). Contrary to my prior expectations, I also found
-that \(T_{min}\) did not correlate with any life-history traits (Fig.
-\ref{fig:correlates}d-h). I initially hypothesized that species with
-longer lifespans or generation times may require a longer sampling
-period. This result could have been a result of at least two factors.
-First, the data I used may not include a diverse enough set of species
-with different life-history traits. Second, the question I posed,
-whether a population is increasing or decreasing, was specifically
-concerned with population trends over time. Therefore, life-history
-characteristics may be more important for other questions more closely
-tied to species biology. For example, Blanchard, Maxwell, and Jennings
-(2007) used detailed simulations of spatially-distributed fisheries to
-compare survey designs. They found that statistical power depended on
-survey design, temperature preferences, and the degree of population
-patchiness.
+\section{Related questions}\label{related-questions}
 
@@
 in space versus time. In a theoretical investigation of this question,
-Rhodes and Jonzen (2011) found that the optimal allocation of sampling
+Rhodes \& Jonzen (2011) found that the optimal allocation of sampling
 depended strongly on temporal and spatial autocorrelation. If spatial
 population dynamics were highly correlated, then it was better to sample
-more temporally, and vice versa. My work supports this idea as
-populations with strong temporal autocorrelation needed less years of
-sampling (Fig. \ref{fig:correlates}). Morrison and Hik (2008) also
+more temporally, and vice versa. The empirical data supports this idea
+as populations with strong temporal autocorrelation needed less years of
+sampling (Fig. \ref{fig:correlates}). Morrison \& Hik (2008) also
 studied the optimal allocation of sampling effort in space versus time,
@@
 
-Seavy and Reynolds (2007) asked whether statistical power was even a
+Seavy \& Reynolds (2007) asked whether statistical power was even a
 useful framework for assessing long-term population trends. They used 24
@@
 consideration when designing monitoring schemes. However, unlike Seavy
-and Reynolds (2007), the results here indicate that longer than 10 years
+\& Reynolds (2007), the results here indicate that longer than 10 years
 is often needed to achieve high statistical power.
 
-As a test of the consequences of not conducting power analyses, I
-compared the results here to the IUCN Redlist criteria. IUCN criteria A2
-suggests that species that have experienced 30\% declines over 10 years
-(or three generations) should be listed as vulnerable (IUCN 2012).
-However, for the populations I examined, this criteria may be too
-simplistic (Fig. \ref{fig:IUCN_analysis}). For many populations, the
-IUCN criteria suggest more years than necessary are required to assess a
-population as vulnerable (points below diagonal line in Fig.
-\ref{fig:IUCN_analysis}). Conversely, for other populations the IUCN
-criteria suggest sampling times that are less than the minimum time
-required for statistical power. This suggests that the IUCN criteria are
-probably too simplistic as the minimum time required does not correlate
-with generation time or other biological covariates (Fig.
-\ref{fig:correlates}d-h).
-
-The design of monitoring programs should include calculations of
-statistical power, the allocation of sampling in space versus time
-(Rhodes and Jonzen 2011), and metrics to increase precision (Seavy and
-Reynolds 2007). Ideally, a formal decision analysis to evaluate these
-different factors would be conducted to design or assess any monitoring
-program (Hauser, Pople, and Possingham 2006; McDonald-Madden et al.
-2010). This type of formal decision analysis would also include
-information on the costs of monitoring. These costs include the actual
-costs of sampling (Brashares and Sam 2005) and the ecological costs on
-inaction (Thompson et al. 2000).
-
-\subsection{Limitations}\label{limitations}
+\section{Limitations}\label{limitations}
 
 This paper has some limitations in determining the minimum time series
-length required. First, \(T_{min}\) is particular to the specific
-question of interest. An additional complication is that for the
-empirical approach, the subsampling of the full time series allows for
-estimates of power, but the individual subsamples are clearly not
+length required. First, the minimum time estimated is particular to the
+specific question of interest. An additional complication is that for
+the empirical approach, the subsampling of the full time series allows
+for estimates of power, but the individual subsamples are clearly not
 independent of one another. In an ideal setting, a specific population
@@
 
-\subsection{Conclusions}\label{conclusions}
+\section{Conclusions}\label{conclusions}
 
-Power analyses are not a novel aspect of ecological research (Legg and
+Power analyses are not a novel aspect of ecological research (Legg \&
 Nagy 2006). However, power analyses are still underutilized, especially
@@
 measures of minimum sampling time based on generation length or other
-life-history traits (Fig. \ref{fig:IUCN_analysis}).
+life-history traits, like those of the IUCN criteria (Fig.
+\ref{fig:IUCN_analysis}).
 
-My work implies that for many populations, short time series are
-probably not reliable for detecting population trends. This result
-highlights the importance of long-term monitoring programs. From both a
-scientific and management perspective, estimates of \(T_{min}\) are
-important. If a time series is too short, we lack the statistical power
-to reliably detect long-term population trends. In addition, a time
-series that is too long may be a poor use of already limited funds
-(Gerber, DeMaster, and Kareiva 1999). Further, more data is not always
-best in situations where management actions need to be taken (Martin et
-al. 2012; Martin et al. 2017). When a population trend is detected, it
-may be too late for management action. In these situations, the
-precautionary principle may be more appropriate (Thompson et al. 2000).
-Future work should examine other species, with a wider range of
-life-history characteristics. In addition, similar approaches can be
-used to determine the minimum time series length required to address
-additional questions of interest.
+The design of monitoring programs should include calculations of
+statistical power, the allocation of sampling in space versus time
+(Larsen \emph{et al.} 2001, Rhodes \& Jonzen 2011), and metrics to
+increase precision (Seavy \& Reynolds 2007). Ideally, a formal decision
+analysis to evaluate these different factors would be conducted to
+design or assess any monitoring program (Hauser \emph{et al.} 2006,
+McDonald-Madden \emph{et al.} 2010). This type of formal decision
+analysis would also include information on the costs of monitoring.
+These costs include the actual costs of sampling (Brashares \& Sam 2005)
+and the ecological costs of inaction (Thompson \emph{et al.} 2000).
+
+For many populations, short time series are probably not reliable for
+detecting population trends. This result highlights the importance of
+long-term monitoring programs. From both a scientific and management
+perspective, estimates of \(T_{min}\) are important. If a time series is
+too short, we lack the statistical power to reliably detect long-term
+population trends. In addition, a time series that is too long may be a
+poor use of already limited funds (Gerber \emph{et al.} 1999). Further,
+more data is not always best in situations where management actions need
+to be taken (Martin \emph{et al.} 2012, 2017). When a population trend
+is detected, it may be too late for management action. In these
+situations, the precautionary principle may be more appropriate
+(Thompson \emph{et al.} 2000). Future work should examine other species,
+with a wider range of life-history characteristics. In addition, similar
+approaches can be used to determine the minimum time series length
+required to address additional questions of interest.
 
@@
 \hypertarget{ref-Blanchard2007}{}
-Blanchard, Julia L, David L Maxwell, and Simon Jennings. 2007. ``Power
-of monitoring surveys to detect abundance trends in depleted
-populations: the effects of density-dependent habitat use, patchiness,
-and climate change.'' \emph{ICES Journal of Marine Science} 65 (1):
-111--20.
+Blanchard, J.L., Maxwell, D.L. \& Jennings, S. (2007). Power of
+monitoring surveys to detect abundance trends in depleted populations:
+the effects of density-dependent habitat use, patchiness, and climate
+change. \emph{ICES Journal of Marine Science}, 65, 111--120.
 
 \hypertarget{ref-Bolker2008}{}
-Bolker, Benjamin M. 2008. \emph{Ecological Models and Data in R}. 1st
-ed. Princeton, New Jersey: Princeton University Press.
+Bolker, B.M. (2008). \emph{Ecological Models and Data in R}. 1st edn.
+Princeton University Press, Princeton, New Jersey.
 
 \hypertarget{ref-Brashares2005}{}
-Brashares, Justin S, and Moses K Sam. 2005. ``How much is enough?
-Estimating the minimum sampling required for effective monitoring of
-African reserves.'' \emph{Biodiversity and Conservation} 14: 2709--22.
-doi:\href{https://doi.org/10.1007/s10531-005-8404-z}{10.1007/s10531-005-8404-z}.
+Brashares, J.S. \& Sam, M.K. (2005). How much is enough? Estimating the
+minimum sampling required for effective monitoring of African reserves.
+\emph{Biodiversity and Conservation}, 14, 2709--2722.
 
 \hypertarget{ref-Cohen1992}{}
-Cohen, Jacob. 1992. ``A power primer.'' \emph{Psychological Bulletin}
-112 (1): 155--59.
-doi:\href{https://doi.org/10.1037/0033-2909.112.1.155}{10.1037/0033-2909.112.1.155}.
+Cohen, J. (1992). A power primer. \emph{Psychological Bulletin}, 112,
+155--159.
 
 \hypertarget{ref-Field2007}{}
-Field, Scott A, Patrick J O Connor, Andrew J Tyre, and Hugh P
-Possingham. 2007. ``Making monitoring meaningful.'' \emph{Austral
-Ecology} 32: 485--91.
-doi:\href{https://doi.org/10.1111/j.1442-9993.2007.01715.x}{10.1111/j.1442-9993.2007.01715.x}.
+Field, S.A., Connor, P.J.O., Tyre, A.J. \& Possingham, H.P. (2007).
+Making monitoring meaningful. \emph{Austral Ecology}, 32, 485--491.
 
 \hypertarget{ref-Gerber1999}{}
-Gerber, L R, D P DeMaster, and P M Kareiva. 1999. ``Gray whales and the
-value of monitering data in implementing the U.S. endangered species
-act.'' \emph{Conservation Biology} 13 (5): 1215--9.
+Gerber, L.R., DeMaster, D.P. \& Kareiva, P.M. (1999). Gray whales and
+the value of monitering data in implementing the U.S. endangered species
+act. \emph{Conservation Biology}, 13, 1215--1219.
 
 \hypertarget{ref-Gerrodette1987}{}
-Gerrodette, Tim. 1987. ``A power analysis for detecting trends.''
-doi:\href{https://doi.org/10.2307/1939220}{10.2307/1939220}.
+Gerrodette, T. (1987). A power analysis for detecting trends.
+
+\hypertarget{ref-Gibbs1998}{}
+Gibbs, J.P., Droege, S. \& Eagle, P. (1998). Monitoring populations of
+plants and animals. \emph{BioScience}, 48, 935--940.
 
 \hypertarget{ref-Giron-Nava2017}{}
-Giron-Nava, Alfredo, Chase C James, Andrew F Johnson, David Dannecker,
-Bethany Kolody, Adrienne Lee, Maitreyi Nagarkar, et al. 2017.
-``Quantitative argument for long-term ecological monitoring.''
-\emph{Marine Ecology Progress Series} 572: 269--74.
+Giron-Nava, A., James, C.C., Johnson, A.F., Dannecker, D., Kolody, B. \&
+Lee, A. \emph{et al.} (2017). Quantitative argument for long-term
+ecological monitoring. \emph{Marine Ecology Progress Series}, 572,
+269--274.
 
 \hypertarget{ref-Hatch2003}{}
-Hatch, S A. 2003. ``Statistical power for detecting trends with
-applications to seabirds monitoring.'' \emph{Biological Conservation}
-111: 317--29.
+Hatch, S.A. (2003). Statistical power for detecting trends with
+applications to seabirds monitoring. \emph{Biological Conservation},
+111, 317--329.
 
 \hypertarget{ref-Hauser2006}{}
-Hauser, Cindy E., Anthony R. Pople, and Hugh P. Possingham. 2006.
-``Should managed populations be monitored every year?'' \emph{Ecological
-Applications} 16 (2): 807--19.
+Hauser, C.E., Pople, A.R. \& Possingham, H.P. (2006). Should managed
+populations be monitored every year? \emph{Ecological Applications}, 16,
+807--819.
 
 \hypertarget{ref-Hughes2017}{}
-Hughes, Brent B, Rodrigo Beas-luna, Allison K Barner, Kimberly Brewitt,
-Daniel R Brumbaugh, Elizabeth B. Cerny-Chipman, Sarah L. Close, et al.
-2017. ``Long-term studies contribute disproportionately to ecology and
-policy.'' \emph{BioScience} 67 (3): 271--81.
-doi:\href{https://doi.org/10.1093/biosci/biw185}{10.1093/biosci/biw185}.
+Hughes, B.B., Beas-luna, R., Barner, A.K., Brewitt, K., Brumbaugh, D.R.
+\& Cerny-Chipman, E.B. \emph{et al.} (2017). Long-term studies
+contribute disproportionately to ecology and policy. \emph{BioScience},
+67, 271--281.
 
 \hypertarget{ref-IUCN2012}{}
-IUCN. 2012. ``IUCN Red List Categories and Criteria: Version 3.1.''
-doi:\href{https://doi.org/10.9782-8317-0633-5}{10.9782-8317-0633-5}.
+IUCN. (2012). IUCN Red List Categories and Criteria: Version 3.1.
 
 \hypertarget{ref-Johnson2015}{}
-Johnson, Paul CD, Sarah JE Barry, Heather M Ferguson, and Pie Müller.
-2015. ``Power analysis for generalized linear mixed models in ecology
-and evolution.'' \emph{Methods in Ecology and Evolution} 6 (2): 133--42.
-doi:\href{https://doi.org/10.1111/2041-210X.12306}{10.1111/2041-210X.12306}.
+Johnson, P.C., Barry, S.J., Ferguson, H.M. \& Müller, P. (2015). Power
+analysis for generalized linear mixed models in ecology and evolution.
+\emph{Methods in Ecology and Evolution}, 6, 133--142.
 
 \hypertarget{ref-Keith2015}{}
-Keith, David, H. Resit Akçakaya, Stuart H.M. Butchart, Ben Collen,
-Nicholas K. Dulvy, Elizabeth E. Holmes, Jeffrey A. Hutchings, et al.
-2015. ``Temporal correlations in population trends: Conservation
-implications from time-series analysis of diverse animal taxa.''
-\emph{Biological Conservation} 192. Elsevier B.V.: 247--57.
-doi:\href{https://doi.org/10.1016/j.biocon.2015.09.021}{10.1016/j.biocon.2015.09.021}.
+Keith, D., Akçakaya, H.R., Butchart, S.H., Collen, B., Dulvy, N.K. \&
+Holmes, E.E. \emph{et al.} (2015). Temporal correlations in population
+trends: Conservation implications from time-series analysis of diverse
+animal taxa. \emph{Biological Conservation}, 192, 247--257.
+
+\hypertarget{ref-Larsen2001}{}
+Larsen, D.P., Kincaid, T.M., Jacobs, S.E. \& Urquhart, N.S. (2001).
+Designs for Evaluating Local and Regional Scale Trends.
+\emph{BioScience}, 51, 1069.
 
 \hypertarget{ref-Legg2006}{}
-Legg, Colin J, and Laszlo Nagy. 2006. ``Why most conservation monitoring
-is, but need not be, a waste of time.'' \emph{Journal of Environmental
-Management} 78: 194--99.
-doi:\href{https://doi.org/10.1016/j.jenvman.2005.04.016}{10.1016/j.jenvman.2005.04.016}.
+Legg, C.J. \& Nagy, L. (2006). Why most conservation monitoring is, but
+need not be, a waste of time. \emph{Journal of Environmental
+Management}, 78, 194--199.
 
 \hypertarget{ref-Magurran2010}{}
-Magurran, Anne E, Stephen R Baillie, Stephen T Buckland, Jan Mcp Dick,
-David A Elston, E Marian Scott, Rognvald I Smith, Paul J Somerfield, and
-Allan D Watt. 2010. ``Long-term datasets in biodiversity research and
-monitoring : assessing change in ecological communities through time.''
-\emph{Trends in Ecology and Evolution} 25: 574--82.
-doi:\href{https://doi.org/10.1016/j.tree.2010.06.016}{10.1016/j.tree.2010.06.016}.
+Magurran, A.E., Baillie, S.R., Buckland, S.T., Dick, J.M., Elston, D.A.
+\& Scott, E.M. \emph{et al.} (2010). Long-term datasets in biodiversity
+research and monitoring : assessing change in ecological communities
+through time. \emph{Trends in Ecology and Evolution}, 25, 574--582.
 
 \hypertarget{ref-Mapstone1995}{}
-Mapstone, Bruce D. 1995. ``Scalable decision rules for environmental
-impact studies : effect Size , type I , and type II errors.''
-\emph{Ecological Applications} 5 (2): 401--10.
+Mapstone, B.D. (1995). Scalable decision rules for environmental impact
+studies : effect Size , type I , and type II errors. \emph{Ecological
+Applications}, 5, 401--410.
 
 \hypertarget{ref-Martin2017}{}
-Martin, Tara G., Abbey E. Camaclang, Hugh P. Possingham, Lynn A.
-Maguire, and Iadine Chadès. 2017. ``Timing of Protection of Critical
-Habitat Matters.'' \emph{Conservation Letters} 10 (3): 308--16.
-doi:\href{https://doi.org/10.1111/conl.12266}{10.1111/conl.12266}.
+Martin, T.G., Camaclang, A.E., Possingham, H.P., Maguire, L.A. \&
+Chadès, I. (2017). Timing of Protection of Critical Habitat Matters.
+\emph{Conservation Letters}, 10, 308--316.
 
 \hypertarget{ref-Martin2012}{}
-Martin, Tara G., Simon Nally, Andrew A. Burbidge, Sophie Arnall, Stephen
-T. Garnett, Matt W. Hayward, Linda F. Lumsden, Peter Menkhorst, Eve
-Mcdonald-Madden, and Hugh P. Possingham. 2012. ``Acting fast helps avoid
-extinction.'' \emph{Conservation Letters} 5 (4): 274--80.
-doi:\href{https://doi.org/10.1111/j.1755-263X.2012.00239.x}{10.1111/j.1755-263X.2012.00239.x}.
+Martin, T.G., Nally, S., Burbidge, A.A., Arnall, S., Garnett, S.T. \&
+Hayward, M.W. \emph{et al.} (2012). Acting fast helps avoid extinction.
+\emph{Conservation Letters}, 5, 274--280.
 
 \hypertarget{ref-McCain2016}{}
-McCain, Christy Marie, Tim Szewczyk, and Kevin Bracy Knight. 2016.
-``Population variability complicates the accurate detection of climate
-change responses.'' \emph{Global Change Biology} 22 (6): 2081--93.
-doi:\href{https://doi.org/10.1111/gcb.13211}{10.1111/gcb.13211}.
+McCain, C.M., Szewczyk, T. \& Knight, K.B. (2016). Population
+variability complicates the accurate detection of climate change
+responses. \emph{Global Change Biology}, 22, 2081--2093.
 
 \hypertarget{ref-McDonald-Madden2010}{}
-McDonald-Madden, Eve, Peter W J Baxter, Richard A. Fuller, Tara G.
-Martin, Edward T. Game, Jensen Montambault, and Hugh P. Possingham.
-2010. ``Monitoring does not always count.'' \emph{Trends in Ecology and
-Evolution} 25 (10): 547--50.
-doi:\href{https://doi.org/10.1016/j.tree.2010.07.002}{10.1016/j.tree.2010.07.002}.
+McDonald-Madden, E., Baxter, P.W.J., Fuller, R.A., Martin, T.G., Game,
+E.T. \& Montambault, J. \emph{et al.} (2010). Monitoring does not always
+count. \emph{Trends in Ecology and Evolution}, 25, 547--550.
 
 \hypertarget{ref-Morrison2008}{}
-Morrison, Shawm, and David S. Hik. 2008. ``When? Where? And for how
-long? Census design considerations for an Alpine Lagomorph, the Collared
-pika.'' In \emph{Lagomorph Biology}, 103--13. Springer Berlin
-Heidelberg.
-doi:\href{https://doi.org/10.1007/978-3-540-72446-9}{10.1007/978-3-540-72446-9}.
+Morrison, S. \& Hik, D.S. (2008). When? Where? And for how long? Census
+design considerations for an Alpine Lagomorph, the Collared pika. In:
+\emph{Lagomorph biology}. Springer Berlin Heidelberg, pp. 103--113.
 
 \hypertarget{ref-Myhrvold2015}{}
-Myhrvold, Nathan P., Elita Baldridge, Benjamin Chan, Dhileep Sivam,
-Daniel L. Freeman, and S.K. Morgan Ernest. 2015. ``An amniote
-life-history database to perform comparative analyses with birds,
-mammals, and reptiles.'' \emph{Ecology} 96 (11): 3109.
+Myhrvold, N.P., Baldridge, E., Chan, B., Sivam, D., Freeman, D.L. \&
+Ernest, S.M. (2015). An amniote life-history database to perform
+comparative analyses with birds, mammals, and reptiles. \emph{Ecology},
+96, 3109.
 
 \hypertarget{ref-GPDD2010}{}
-NERC Centre for Population Biology, Imperial College. 2010. ``The Global
-Population Dynamics Database Version 2.''
+NERC Centre for Population Biology, I.C. (2010). The Global Population
+Dynamics Database Version 2.
 
 \hypertarget{ref-Nichols2006}{}
-Nichols, James D., and Bryon K. Williams. 2006. ``Monitoring for
-conservation.'' \emph{Trends in Ecology and Evolution} 21 (12): 668--73.
-doi:\href{https://doi.org/10.1016/j.tree.2006.08.007}{10.1016/j.tree.2006.08.007}.
+Nichols, J.D. \& Williams, B.K. (2006). Monitoring for conservation.
+\emph{Trends in Ecology and Evolution}, 21, 668--673.
 
-\hypertarget{ref-RCoreTeam2016}{}
-R Core Team. 2016. ``R: A language and environment for statistical
-computing.'' Vienna, Austria: R Foundation for Statistical Computing.
-\url{https://www.r-project.org/}.
+\hypertarget{ref-Peters2010}{}
+Peters, D.P. (2010). Accessible ecology: Synthesis of the long, deep,
+and broad. \emph{Trends in Ecology and Evolution}, 25, 592--601.
+
+\hypertarget{ref-RCoreTeam2017}{}
+R Core Team. (2017). R: A language and environment for statistical
+computing.
 
 \hypertarget{ref-Rhodes2011}{}
-Rhodes, Jonathan R., and Niclas Jonzen. 2011. ``Monitoring temporal
-trends in spatially structured populations: how should sampling effort
-be allocated between space and time?'' \emph{Ecography} 34 (6): 1040--8.
-doi:\href{https://doi.org/10.1111/j.1600-0587.2011.06370.x}{10.1111/j.1600-0587.2011.06370.x}.
+Rhodes, J.R. \& Jonzen, N. (2011). Monitoring temporal trends in
+spatially structured populations: how should sampling effort be
+allocated between space and time? \emph{Ecography}, 34, 1040--1048.
 
 \hypertarget{ref-Rueda-Cediel2015}{}
-Rueda-Cediel, Pamela, Kurt E Anderson, Tracey J Regan, Janet Franklin,
-and M Regan. 2015. ``Combined influences of model choice, data quality,
-and data quantity when estimating population trends.'' \emph{PLoSONE} 10
-(7): e0132255.
-doi:\href{https://doi.org/10.1371/journal.pone.0132255}{10.1371/journal.pone.0132255}.
+Rueda-Cediel, P., Anderson, K.E., Regan, T.J., Franklin, J. \& Regan, M.
+(2015). Combined influences of model choice, data quality, and data
+quantity when estimating population trends. \emph{PLoSONE}, 10,
+e0132255.
 
 \hypertarget{ref-Seavy2007}{}
-Seavy, Nathaniel E., and Michelle H. Reynolds. 2007. ``Is statistical
-power to detect trends a good assessment of population monitoring?''
-\emph{Biological Conservation} 140 (1-2): 187--91.
-doi:\href{https://doi.org/10.1016/j.biocon.2007.08.007}{10.1016/j.biocon.2007.08.007}.
+Seavy, N.E. \& Reynolds, M.H. (2007). Is statistical power to detect
+trends a good assessment of population monitoring? \emph{Biological
+Conservation}, 140, 187--191.
 
 \hypertarget{ref-Thompson2000}{}
-Thompson, Paul M, Ben Wilson, Kate Grellier, and Philip S Hammond. 2000.
-``Combining power analysis and population viability analysis to compare
+Thompson, P.M., Wilson, B., Grellier, K. \& Hammond, P.S. (2000).
+Combining power analysis and population viability analysis to compare
 traditional and precautionary approaches to conservation of coastal
-cetaceans.'' \emph{Conservation Biology} 14 (5): 1253--63.
+cetaceans. \emph{Conservation Biology}, 14, 1253--1263.
 
 \hypertarget{ref-Wagner2009}{}
-Wagner, Tyler, Christopher S. Vandergoot, and Jeff Tyson. 2009.
-``Evaluating the power to detect temporal trends in fishery-independent
-surveys - A case study based on gill nets set in the Ohio waters of Lake
-Erie for walleyes.'' \emph{North American Journal of Fisheries
-Management} 29: 805--16.
-doi:\href{https://doi.org/10.1577/M08-197.1}{10.1577/M08-197.1}.
+Wagner, T., Vandergoot, C.S. \& Tyson, J. (2009). Evaluating the power
+to detect temporal trends in fishery-independent surveys - A case study
+based on gill nets set in the Ohio waters of Lake Erie for walleyes.
+\emph{North American Journal of Fisheries Management}, 29, 805--816.
 
 \hypertarget{ref-White2014}{}
-White, Easton R, John D Nagy, and Samuel H Gruber. 2014. ``Modeling the
-population dynamics of lemon sharks.'' \emph{Biology Direct} 9 (23):
-1--18.
+White, E.R., Nagy, J.D. \& Gruber, S.H. (2014). Modeling the population
+dynamics of lemon sharks. \emph{Biology Direct}, 9, 1--18.
 
 \hypertarget{ref-Zuur2009}{}
-Zuur, Alain F., Elena N. Ieno, Neil J. Walker, Anatoly A. Saveliev, and
-Graham M. Smith. 2009. \emph{Mixed Effects Models and Extensions in
-Ecology with R}. New York: Springer.
+Zuur, A.F., Ieno, E.N., Walker, N.J., Saveliev, A.A. \& Smith, G.M.
+(2009). \emph{Mixed Effects Models and Extensions in Ecology with R}.
+Springer, New York.
 
